\NormalFont\ShortTitle{2 John}
{\MT John’s Second Letter

\par }\OneChap {\PP \VerseOne{1}The elder, to the chosen lady and her children, whom I love in truth, and not I only, but also all those who know the truth,
\VS{2}for the truth’s sake, which remains in us, and it will be with us forever:
\VS{3}Grace, mercy, and peace will be with us, from God the Father, and from the Lord Jesus Christ, the Son of the Father, in truth and love.
\par }{\PP \VS{4}I rejoice greatly that I have found some of your children walking in truth, even as we have been commanded by the Father.
\VS{5}Now I beg you, dear lady, not as though I wrote to you a new commandment, but that which we had from the beginning, that we love one another.
\VS{6}This is love, that we should walk according to his commandments. This is the commandment, even as you heard from the beginning, that you should walk in it.
\VS{7}For many deceivers have gone out into the world, those who don’t confess that Jesus Christ came in the flesh. This is the deceiver and the Antichrist.
\VS{8}Watch yourselves, that we don’t lose the things which we have accomplished, but that we receive a full reward.
\VS{9}Whoever transgresses and doesn’t remain in the teaching of Christ, doesn’t have God. He who remains in the teaching has both the Father and the Son.
\VS{10}If anyone comes to you, and doesn’t bring this teaching, don’t receive him into your house, and don’t welcome him,
\VS{11}for he who welcomes him participates in his evil deeds.
\par }{\PP \VS{12}Having many things to write to you, I don’t want to do so with paper and ink, but I hope to come to you, and to speak face to face, that our joy may be made full.
\VS{13}The children of your chosen sister greet you. Amen.
\par }