\NormalFont\ShortTitle{Daniel (Greek)}
{\MTB The Book of
\par }{\MT Daniel
\par }{\MTB with Greek Portions

\par }\ChapOne{1}{\PP \VerseOne{1}In the third year of the reign of Jehoiakim king of Judah Nebuchadnezzar king of Babylon came to Jerusalem and besieged it.
\VS{2}The Lord\FTNT{*}{{\FR 1:2 }
The word translated “Lord” is “Adonai.”} gave Jehoiakim king of Judah into his hand, with part of the vessels of the house of God;
\FTNT{†}{{\FR 1:2 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).} and he carried them into the land of Shinar to the house of his god. He brought the vessels into the treasure house of his god.
\par }{\PP \VS{3}The king spoke to Ashpenaz the master of his eunuchs, that he should bring in some of the children of Israel, even of the royal offspring\FTNT{‡}{{\FR 1:3 }
or, seed} and of the nobles;
\VS{4}youths in whom was no defect, but well-favored, and skillful in all wisdom, and endowed with knowledge, and understanding science, and who had the ability to stand in the king’s palace; and that he should teach them the learning and the language of the Chaldeans.
\VS{5}The king appointed for them a daily portion of the king’s dainties, and of the wine which he drank, and that they should be nourished three years; that at its end they should stand before the king.
\par }{\PP \VS{6}Now among these were of the children of Judah: Daniel, Hananiah, Mishael, and Azariah.
\VS{7}The prince of the eunuchs gave names to them: to Daniel he gave the name Belteshazzar; to Hananiah, Shadrach; to Mishael, Meshach; and to Azariah, Abednego.
\par }{\PP \VS{8}But Daniel purposed in his heart that he would not defile himself with the king’s dainties, nor with the wine which he drank. Therefore he requested of the prince of the eunuchs that he might not defile himself.
\VS{9}Now God made Daniel find kindness and compassion in the sight of the prince of the eunuchs.
\VS{10}The prince of the eunuchs said to Daniel, “I fear my lord the king, who has appointed your food and your drink. For why should he see your faces worse looking than the youths who are of your own age? Then you would endanger my head with the king.”
\par }{\PP \VS{11}Then Daniel said to the steward whom the prince of the eunuchs had appointed over Daniel, Hananiah, Mishael, and Azariah:
\VS{12}“Test your servants, I beg you, ten days; and let them give us vegetables to eat, and water to drink.
\VS{13}Then let our faces be examined before you, and the face of the youths who eat of the king’s dainties; and as you see, deal with your servants.”
\VS{14}So he listened to them in this matter, and tested them for ten days.
\par }{\PP \VS{15}At the end of ten days, their faces appeared fairer, and they were fatter in flesh, than all the youths who ate of the king’s dainties.
\VS{16}So the steward took away their dainties, and the wine that they would drink, and gave them vegetables.
\par }{\PP \VS{17}Now as for these four youths, God gave them knowledge and skill in all learning and wisdom; and Daniel had understanding in all visions and dreams.
\par }{\PP \VS{18}At the end of the days which the king had appointed for bringing them in, the prince of the eunuchs brought them in before Nebuchadnezzar.
\VS{19}The king talked with them; and among them all was found no one like Daniel, Hananiah, Mishael, and Azariah. Therefore stood they before the king.
\VS{20}In every matter of wisdom and understanding, concerning which the king inquired of them, he found them ten times better than all the magicians and enchanters who were in all his realm.
\par }{\PP \VS{21}Daniel continued even to the first year of king Cyrus.

\par }\Chap{2}{\PP \VerseOne{1}In the second year of the reign of Nebuchadnezzar, Nebuchadnezzar dreamed dreams; and his spirit was troubled, and his sleep went from him.
\VS{2}Then the king commanded that the magicians, the enchanters, the sorcerers, and the Chaldeans be called to tell the king his dreams. So they came in and stood before the king.
\VS{3}The king said to them, “I have dreamed a dream, and my spirit is troubled to know the dream.”
\par }{\PP \VS{4}Then the Chaldeans spoke to the king in the Syrian language, “O king, live forever! Tell your servants the dream, and we will show the interpretation.”
\par }{\PP \VS{5}The king answered the Chaldeans, “The thing has gone from me. If you don’t make known to me the dream and its interpretation, you will be cut in pieces, and your houses will be made a dunghill.
\VS{6}But if you show the dream and its interpretation, you will receive of me gifts and rewards and great honor. Therefore show me the dream and its interpretation.”
\par }{\PP \VS{7}They answered the second time and said, “Let the king tell his servants the dream, and we will show the interpretation.”
\par }{\PP \VS{8}The king answered, “I know of a certainty that you are trying to gain time, because you see the thing has gone from me.
\VS{9}But if you don’t make known to me the dream, there is but one law for you; for you have prepared lying and corrupt words to speak before me, until the situation changes. Therefore tell me the dream, and I will know that you can show me its interpretation.”
\par }{\PP \VS{10}The Chaldeans answered before the king, and said, “There is not a man on the earth who can show the king’s matter, because no king, lord, or ruler, has asked such a thing of any magician, or enchanter, or Chaldean.
\VS{11}It is a rare thing that the king requires, and there is no other who can show it before the king, except the gods, whose dwelling is not with flesh.”
\par }{\PP \VS{12}Because of this, the king was angry and very furious, and commanded that all the wise men of Babylon be destroyed.
\VS{13}So the decree went out, and the wise men were to be slain. They sought Daniel and his companions to be slain.
\par }{\PP \VS{14}Then Daniel returned answer with counsel and prudence to Arioch the captain of the king’s guard, who had gone out to kill the wise men of Babylon.
\VS{15}He answered Arioch the king’s captain, “Why is the decree so urgent from the king?” Then Arioch made the thing known to Daniel.
\VS{16}Daniel went in, and desired of the king that he would appoint him a time, and he would show the king the interpretation.
\par }{\PP \VS{17}Then Daniel went to his house and made the thing known to Hananiah, Mishael, and Azariah, his companions:
\VS{18}that they would desire mercies of the God of heaven concerning this secret; that Daniel and his companions would not perish with the rest of the wise men of Babylon.
\VS{19}Then the secret was revealed to Daniel in a vision of the night. Then Daniel blessed the God of heaven.
\VS{20}Daniel answered,
\par }{\Q “Blessed be the name of God forever and ever;
\par }{\QB for wisdom and might are his.
\par }{\Q \VS{21}He changes the times and the seasons.
\par }{\QB He removes kings, and sets up kings.
\par }{\Q He gives wisdom to the wise,
\par }{\QB and knowledge to those who have understanding.
\par }{\Q \VS{22}He reveals the deep and secret things.
\par }{\QB He knows what is in the darkness,
\par }{\QB and the light dwells with him.
\par }{\Q \VS{23}I thank you, and praise you,
\par }{\QB you God of my fathers,
\par }{\Q who have given me wisdom and might,
\par }{\QB and have now made known to me what we desired of you;
\par }{\QB for you have made known to us the king’s matter.”
\par }{\PP \VS{24}Therefore Daniel went in to Arioch, whom the king had appointed to destroy the wise men of Babylon. He went and said this to him: “Don’t destroy the wise men of Babylon. Bring me in before the king, and I will show to the king the interpretation.”
\par }{\PP \VS{25}Then Arioch brought in Daniel before the king in haste, and said this to him: “I have found a man of the children of the captivity of Judah who will make known to the king the interpretation.”
\par }{\PP \VS{26}The king answered Daniel, whose name was Belteshazzar, “Are you able to make known to me the dream which I have seen, and its interpretation?”
\par }{\PP \VS{27}Daniel answered before the king, and said, “The secret which the king has demanded can’t be shown to the king by wise men, enchanters, magicians, or soothsayers;
\VS{28}but there is a God in heaven who reveals secrets, and he has made known to king Nebuchadnezzar what will be in the latter days. Your dream, and the visions of your head on your bed, are these:
\par }{\PP \VS{29}“As for you, O king, your thoughts came on your bed, what should happen hereafter; and he who reveals secrets has made known to you what will happen.
\VS{30}But as for me, this secret is not revealed to me for any wisdom that I have more than any living, but to the intent that the interpretation may be made known to the king, and that you may know the thoughts of your heart.
\par }{\PP \VS{31}“You, O king, saw, and behold,\FTNT{*}{{\FR 2:31 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} a great image. This image, which was mighty, and whose brightness was excellent, stood before you; and its appearance was terrifying.
\VS{32}As for this image, its head was of fine gold, its breast and its arms of silver, its belly and its thighs of bronze,
\VS{33}its legs of iron, its feet part of iron, and part of clay.
\VS{34}You saw until a stone was cut out without hands, which struck the image on its feet that were of iron and clay, and broke them in pieces.
\VS{35}Then the iron, the clay, the bronze, the silver, and the gold were broken in pieces together, and became like the chaff of the summer threshing floors. The wind carried them away, so that no place was found for them. The stone that struck the image became a great mountain, and filled the whole earth.
\par }{\PP \VS{36}“This is the dream; and we will tell its interpretation before the king.
\VS{37}You, O king, are king of kings, to whom the God of heaven has given the kingdom, the power, the strength, and the glory.
\VS{38}Wherever the children of men dwell, he has given the animals of the field and the birds of the sky into your hand, and has made you rule over them all. You are the head of gold.
\par }{\PP \VS{39}“After you, another kingdom will arise that is inferior to you; and another third kingdom of bronze, which will rule over all the earth.
\VS{40}The fourth kingdom will be strong as iron, because iron breaks in pieces and subdues all things; and as iron that crushes all these, it will break in pieces and crush.
\VS{41}Whereas you saw the feet and toes, part of potters’ clay, and part of iron, it will be a divided kingdom; but there will be in it of the strength of the iron, because you saw the iron mixed with miry clay.
\VS{42}As the toes of the feet were part of iron, and part of clay, so the kingdom will be partly strong, and partly broken.
\VS{43}Whereas you saw the iron mixed with miry clay, they will mingle themselves with the seed of men; but they won’t cling to one another, even as iron does not mix with clay.
\par }{\PP \VS{44}“In the days of those kings the God of heaven will set up a kingdom which will never be destroyed, nor will its sovereignty be left to another people; but it will break in pieces and consume all these kingdoms, and it will stand forever.
\VS{45}Because you saw that a stone was cut out of the mountain without hands, and that it broke in pieces the iron, the bronze, the clay, the silver, and the gold; the great God has made known to the king what will happen hereafter. The dream is certain, and its interpretation sure.”
\par }{\PP \VS{46}Then king Nebuchadnezzar fell on his face, worshiped Daniel, and commanded that they should offer an offering and sweet odors to him.
\VS{47}The king answered to Daniel, and said, “Of a truth your God is the God of gods, and the Lord of kings, and a revealer of secrets, since you have been able to reveal this secret.”
\par }{\PP \VS{48}Then the king made Daniel great, and gave him many great gifts, and made him rule over the whole province of Babylon, and to be chief governor over all the wise men of Babylon.
\VS{49}Daniel requested of the king, and he appointed Shadrach, Meshach, and Abednego over the affairs of the province of Babylon; but Daniel was in the king’s gate.

\par }\Chap{3}{\PP \VerseOne{1}Nebuchadnezzar the king made an image of gold, whose height was sixty cubits,\FTNT{*}{{\FR 3:1 }
A cubit is the length from the tip of the middle finger to the elbow on a man’s arm, or about 18 inches or 46 centimeters.} and its width six cubits. He set it up in the plain of Dura, in the province of Babylon.
\VS{2}Then Nebuchadnezzar the king sent to gather together the local governors, the deputies, and the governors, the judges, the treasurers, the counselors, the sheriffs, and all the rulers of the provinces, to come to the dedication of the image which Nebuchadnezzar the king had set up.
\VS{3}Then the local governors, the deputies, and the governors, the judges, the treasurers, the counselors, the sheriffs, and all the rulers of the provinces, were gathered together to the dedication of the image that Nebuchadnezzar the king had set up; and they stood before the image that Nebuchadnezzar had set up.
\par }{\PP \VS{4}Then the herald cried aloud, “To you it is commanded, peoples, nations, and languages,
\VS{5}that whenever you hear the sound of the horn, flute, zither, lyre, harp, pipe, and all kinds of music, you fall down and worship the golden image that Nebuchadnezzar the king has set up.
\VS{6}Whoever doesn’t fall down and worship shall be cast into the middle of a burning fiery furnace the same hour.”
\par }{\PP \VS{7}Therefore at that time, when all the peoples heard the sound of the horn, flute, zither, lyre, harp, pipe, and all kinds of music, all the peoples, the nations, and the languages, fell down and worshiped the golden image that Nebuchadnezzar the king had set up.
\par }{\PP \VS{8}Therefore at that time certain Chaldeans came near, and brought accusation against the Jews.
\VS{9}They answered Nebuchadnezzar the king, “O king, live for ever!
\VS{10}You, O king, have made a decree, that every man that hears the sound of the horn, flute, zither, lyre, harp, pipe, and all kinds of music, shall fall down and worship the golden image;
\VS{11}and whoever doesn’t fall down and worship shall be cast into the middle of a burning fiery furnace.
\VS{12}There are certain Jews whom you have appointed over the affairs of the province of Babylon: Shadrach, Meshach, and Abednego. These men, O king, have not respected you. They don’t serve your gods, and don’t worship the golden image which you have set up.”
\par }{\PP \VS{13}Then Nebuchadnezzar in rage and fury commanded that Shadrach, Meshach, and Abednego be brought. Then these men were brought before the king.
\VS{14}Nebuchadnezzar answered them, “Is it on purpose, Shadrach, Meshach, and Abednego, that you don’t serve my god, nor worship the golden image which I have set up?
\VS{15}Now if you are ready whenever you hear the sound of the horn, flute, zither, lyre, harp, pipe, and all kinds of music to fall down and worship the image which I have made, good; but if you don’t worship, you shall be cast the same hour into the middle of a burning fiery furnace. Who is that god that will deliver you out of my hands?”
\par }{\PP \VS{16}Shadrach, Meshach, and Abednego answered the king, “Nebuchadnezzar, we have no need to answer you in this matter.
\VS{17}If it happens, our God whom we serve is able to deliver us from the burning fiery furnace; and he will deliver us out of your hand, O king.
\VS{18}But if not, let it be known to you, O king, that we will not serve your gods or worship the golden image which you have set up.”
\par }{\PP \VS{19}Then Nebuchadnezzar was full of fury, and the form of his appearance was changed against Shadrach, Meshach, and Abednego. He spoke, and commanded that they should heat the furnace seven times more than it was usually heated.
\VS{20}He commanded certain mighty men who were in his army to bind Shadrach, Meshach, and Abednego, and to cast them into the burning fiery furnace.
\VS{21}Then these men were bound in their pants, their tunics, and their mantles, and their other clothes, and were cast into the middle of the burning fiery furnace.
\VS{22}Therefore because the king’s commandment was urgent, and the furnace exceedingly hot, the flame of the fire killed those men who took up Shadrach, Meshach, and Abednego.
\VS{23}These three men, Shadrach, Meshach, and Abednego, fell down bound into the middle of the burning fiery furnace.
\par }{\SH THE SONG OF THE THREE HOLY CHILDREN
\FTNT{†}{{\FR 3:23 }

{\BK{The Song of the Three Holy Children}} is an addition to
{\BK{Daniel}} found in the Greek Septuagint but not found in the traditional Hebrew text of
{\BK{Daniel}}. This portion is recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches. It is found inserted between Daniel 3:23 and Daniel 3:24 of the traditional Hebrew Bible. Here, the verses after 23 from the Hebrew Bible are numbered starting at 91 to make room for these verses.}
\par }{\PP \VS{24}They walked in the midst of the fire, praising God, and blessing the Lord.
\VS{25}Then Azarias stood, and prayed like this. Opening his mouth in the midst of the fire he said:
\VS{26}“Blessed are you, O Lord, you God of our fathers! Your name is worthy to be praised and glorified for evermore;
\VS{27}for you are righteous in all the things that you have done. Yes, all your works are true. Your ways are right, and all your judgments are truth.
\VS{28}In all the things that you have brought upon us, and upon the holy city of our fathers, Jerusalem, you have executed true judgments. For according to truth and justice you have brought all these things upon us because of our sins.
\VS{29}For we have sinned and committed iniquity, in departing from you.
\VS{30}In all things we have trespassed, and not obeyed your commandments, or kept them. We haven’t done as you have commanded us, that it might go well with us.
\VS{31}Therefore all that you have brought upon us, and everything that you have done to us, you have done in true judgement.
\VS{32}You delivered us into the hands of lawless enemies, most hateful rebels, and to an unjust king who is the most wicked in all the world.
\VS{33}And now we can’t open our mouth. Shame and reproach have come on your servants and those who worship you.
\VS{34}Don’t utterly deliver us up, for your name’s sake. Don’t annul your covenant.
\VS{35}Don’t cause your mercy to depart from us, for the sake of Abraham who is loved by you, and for the sake of Isaac your servant, and Israel your holy one,
\VS{36}to whom you promised that you would multiply their offspring as the stars of the sky, and as the sand that is on the sea shore.
\VS{37}For we, O Lord, have become less than any nation, and are kept under this day in all the world because of our sins.
\VS{38}There isn’t at this time prince, or prophet, or leader, or burnt offering, or sacrifice, or oblation, or incense, or place to offer before you, and to find mercy.
\VS{39}Nevertheless in a contrite heart and a humble spirit let us be accepted,
\VS{40}like the burnt offerings of rams and bullocks, and like ten thousands of fat lambs. So let our sacrifice be in your sight this day, that we may wholly go after you, for they shall not be ashamed who put their trust in you.
\VS{41}And now we follow you with all our heart. We fear you, and seek your face.
\VS{42}Put us not to shame; but deal with us after your kindness, and according to the multitude of your mercy.
\VS{43}Deliver us also according to your marvelous works, and give glory to your name, O Lord. Let all those who harm your servants be confounded.
\VS{44}Let them be ashamed of all their power and might, and let their strength be broken.
\VS{45}Let them know that you are the Lord, the only God, and glorious over the whole world.”
\par }{\PP \VS{46}The king’s servants who put them in didn’t stop making the furnace hot with naphtha, pitch, tinder, and small wood,
\VS{47}so that the flame streamed out forty nine cubits above the furnace.
\VS{48}It spread and burned those Chaldeans whom it found around the furnace.
\VS{49}But the angel of the Lord came down into the furnace together with Azarias and his fellows, and he struck the flame of the fire out of the furnace,
\VS{50}and made the midst of the furnace as it had been a moist whistling wind, so that the fire didn’t touch them at all. It neither hurt nor troubled them.
\par }{\PP \VS{51}Then the three, as out of one mouth, praised, and glorified, and blessed God in the furnace, saying,
\VS{52}“Blessed are you, O Lord, you God of our fathers, to be praised and exalted above all forever!
\VS{53}Blessed is your glorious and holy name, to be praised and exalted above all forever!
\VS{54}Blessed are you in the temple of your holy glory, to be praised and glorified above all forever!
\VS{55}Blessed are you that behold the depths, and sit upon the cherubim, to be praised and exalted above all forever.
\VS{56}Blessed are you on the throne of your kingdom, to be praised and extolled above all forever!
\VS{57}Blessed are you in the firmament of heaven, to be praised and glorified forever!
\par }{\PP \VS{58}O all you works of the Lord, bless the Lord! Praise and exalt him above all forever!
\VS{59}O you heavens, bless the Lord! Praise and exalt him above all for ever!
\VS{60}O you angels of the Lord, bless the Lord! Praise and exalt him above all forever!
\VS{61}O all you waters that are above the sky, bless the Lord! Praise and exalt him above all forever!
\VS{62}O all you powers of the Lord, bless the Lord! Praise and exalt him above all forever!
\VS{63}O you sun and moon, bless the Lord! Praise and exalt him above all forever!
\VS{64}O you stars of heaven, bless the Lord! Praise and exalt him above all forever!
\VS{65}O every shower and dew, bless the Lord! Praise and exalt him above all forever!
\VS{66}O all you winds, bless the Lord! Praise and exalt him above all forever!
\VS{67}O you fire and heat, bless the Lord! Praise and exalt him above all forever!
\VS{68}O you dews and storms of snow, bless the Lord! Praise and exalt him above all forever!
\VS{69}O you nights and days, bless the Lord! Praise and exalt him above all forever!
\VS{70}O you light and darkness, bless the Lord! Praise and exalt him above all forever!
\VS{71}O you cold and heat, bless the Lord! Praise and exalt him above all forever!
\VS{72}O you frost and snow, bless the Lord! Praise and exalt him above all forever!
\VS{73}O you lightnings and clouds, bless the Lord! Praise and exalt him above all forever!
\VS{74}O let the earth bless the Lord! Let it praise and exalt him above all forever!
\VS{75}O you mountains and hills, bless the Lord! Praise and exalt him above all forever!
\VS{76}O all you things that grow on the earth, bless the Lord! Praise and exalt him above all forever!
\VS{77}\FTNT{‡}{{\FR 3:77 }
Some authorities transpose this verse and the next one.}O sea and rivers, bless the Lord! Praise and exalt him above all forever!
\VS{78}O you springs, bless the Lord! Praise and exalt him above all forever!
\VS{79}O you whales and all that move in the waters, bless the Lord! Praise and exalt him above all forever!
\VS{80}O all you fowls of the air, bless the Lord! Praise and exalt him above all forever!
\VS{81}O all you beasts and cattle, bless the Lord! Praise and exalt him above all forever!
\VS{82}O you children of men, bless the Lord! Praise and exalt him above all forever!
\VS{83}O let Israel bless the Lord! Praise and exalt him above all forever.
\VS{84}O you priests of the Lord, bless the Lord! Praise and exalt him above all forever!
\VS{85}O you servants of the Lord, bless the Lord! Praise and exalt him above all forever!
\VS{86}O you spirits and souls of the righteous, bless the Lord! Praise and exalt him above all forever!
\VS{87}O you who are holy and humble of heart, bless the Lord! Praise and exalt him above all forever!
\VS{88}O Hananiah, Mishael, and Azariah, bless the Lord! Praise and exalt him above all forever; for he has rescued us from Hades, and saved us from the hand of death! He has delivered us out of the midst of the furnace and burning flame. He has delivered us out of the midst of the fire.
\VS{89}O give thanks to the Lord, for he is good; for his mercy is forever.
\VS{90}O all you who worship the Lord, bless the God of gods, praise him, and give him thanks; for his mercy is forever!”
\par }{\SH Deliverance from the Furnace
\par }{\PP \VS{91}\FTNT{§}{{\FR 3:91 }
Verses 91-97 were numbered 24-30 in the traditional Hebrew text of Daniel.} Then Nebuchadnezzar the king was astonished, and rose up in haste. He spoke and said to his counselors, “Didn’t we cast three men bound into the middle of the fire?”
\par }{\PP They answered the king, “True, O king.”
\par }{\PP \VS{92}He answered, “Look, I see four men loose, walking in the middle of the fire, and they are unharmed. The appearance of the fourth is like a son of the gods.”
\par }{\PP \VS{93}Then Nebuchadnezzar came near to the mouth of the burning fiery furnace. He spoke and said, “Shadrach, Meshach, and Abednego, you servants of the Most High God, come out, and come here!”
\par }{\PP Then Shadrach, Meshach, and Abednego came out of the middle of the fire.
\VS{94}The local governors, the deputies, and the governors, and the king’s counselors, being gathered together, saw these men, that the fire had no power on their bodies. The hair of their head wasn’t singed. Their pants weren’t changed, the smell of fire wasn’t even on them.
\par }{\PP \VS{95}Nebuchadnezzar spoke and said, “Blessed be the God of Shadrach, Meshach, and Abednego, who has sent his angel, and delivered his servants who trusted in him, and have changed the king’s word, and have yielded their bodies, that they might not serve nor worship any god, except their own God.
\VS{96}Therefore I make a decree, that every people, nation, and language, which speak anything evil against the God of Shadrach, Meshach, and Abednego, shall be cut in pieces, and their houses shall be made a dunghill; because there is no other god who is able to deliver like this.”
\par }{\PP \VS{97}Then the king promoted Shadrach, Meshach, and Abednego in the province of Babylon.

\par }\Chap{4}{\PI \VerseOne{1}Nebuchadnezzar the king,
\par }{\PI to all the peoples, nations, and languages, who dwell in all the earth:
\par }{\PI Peace be multiplied to you.
\par }{\PI \VS{2}It has seemed good to me to show the signs and wonders that the Most High God has worked toward me.
\par }{\Q \VS{3}How great are his signs!
\par }{\QB How mighty are his wonders!
\par }{\Q His kingdom is an everlasting kingdom.
\par }{\QB His dominion is from generation to generation.
\par }{\PI \VS{4}I, Nebuchadnezzar, was at rest in my house, and flourishing in my palace.
\VS{5}I saw a dream which made me afraid; and the thoughts on my bed and the visions of my head troubled me.
\VS{6}Therefore I made a decree to bring in all the wise men of Babylon before me, that they might make known to me the interpretation of the dream.
\VS{7}Then the magicians, the enchanters, the Chaldeans, and the soothsayers came in; and I told the dream before them; but they didn’t make known to me its interpretation.
\VS{8}But at the last Daniel came in before me, whose name was Belteshazzar, according to the name of my god, and in whom is the spirit of the holy gods. I told the dream before him, saying,
\par }{\PI \VS{9}“Belteshazzar, master of the magicians, because I know that the spirit of the holy gods is in you, and no secret troubles you, tell me the visions of my dream that I have seen, and its interpretation.
\VS{10}Thus were the visions of my head on my bed: I saw, and behold, a tree in the middle of the earth; and its height was great.
\VS{11}The tree grew, and was strong, and its height reached to the sky, and its sight to the end of all the earth.
\VS{12}The leaves of it were beautiful, and it had much fruit, and in it was food for all. The animals of the field had shade under it, and the birds of the sky lived in its branches, and all flesh was fed from it.
\par }{\PI \VS{13}“I saw in the visions of my head on my bed, and behold, a watcher and a holy one came down from the sky.
\VS{14}He cried aloud, and said this, ‘Cut down the tree, and cut off its branches! Shake off its leaves, and scatter its fruit! Let the animals get away from under it, and the fowls from its branches.
\VS{15}Nevertheless leave the stump of its roots in the earth, even with a band of iron and bronze, in the tender grass of the field; and let it be wet with the dew of the sky. Let his portion be with the animals in the grass of the earth.
\VS{16}Let his heart be changed from man’s, and let an animal’s heart be given to him. Then let seven times pass over him.
\par }{\PI \VS{17}“ ‘The sentence is by the decree of the watchers, and the demand by the word of the holy ones; to the intent that the living may know that the Most High rules in the kingdom of men, and gives it to whomever he will, and sets up over it the lowest of men.’
\par }{\PI \VS{18}“This dream I, king Nebuchadnezzar, have seen; and you, Belteshazzar, declare the interpretation, because all the wise men of my kingdom are not able to make known to me the interpretation; but you are able; for the spirit of the holy gods is in you.”
\par }{\PI \VS{19}Then Daniel, whose name was Belteshazzar, was stricken mute for a while, and his thoughts troubled him. The king answered, “Belteshazzar, don’t let the dream, or the interpretation, trouble you.”
\par }{\PI Belteshazzar answered, “My lord, may the dream be for those who hate you, and its interpretation to your adversaries.
\VS{20}The tree that you saw, which grew, and was strong, whose height reached to the sky, and its sight to all the earth;
\VS{21}whose leaves were beautiful, and its fruit plentiful, and in it was food for all; under which the animals of the field lived, and on whose branches the birds of the sky had their habitation:
\VS{22}it is you, O king, that have grown and become strong; for your greatness has grown, and reaches to the sky, and your dominion to the end of the earth.
\par }{\PI \VS{23}“Whereas the king saw a watcher and a holy one coming down from the sky, and saying, ‘Cut down the tree, and destroy it; nevertheless leave the stump of its roots in the earth, even with a band of iron and bronze, in the tender grass of the field, and let it be wet with the dew of the sky. Let his portion be with the animals of the field, until seven times pass over him.’
\par }{\PI \VS{24}“This is the interpretation, O king, and it is the decree of the Most High, which has come on my lord the king:
\VS{25}that you shall be driven from men, and your dwelling shall be with the animals of the field. You shall be made to eat grass as oxen, and shall be wet with the dew of the sky, and seven times shall pass over you; until you know that the Most High rules in the kingdom of men, and gives it to whomever he will.
\VS{26}Whereas they commanded to leave the stump of the roots of the tree; your kingdom shall be sure to you, after that you will have known that the heavens do rule.
\VS{27}Therefore, O king, let my counsel be acceptable to you, and break off your sins by righteousness, and your iniquities by showing mercy to the poor. Perhaps there may be a lengthening of your tranquility.”
\par }{\PI \VS{28}All this came on the king Nebuchadnezzar.
\VS{29}At the end of twelve months he was walking in the royal palace of Babylon.
\VS{30}The king spoke and said, “Is not this great Babylon, which I have built for the royal dwelling place, by the might of my power and for the glory of my majesty?”
\par }{\PI \VS{31}While the word was in the king’s mouth, a voice came from the sky, saying, “O king Nebuchadnezzar, to you it is spoken: ‘The kingdom has departed from you.
\VS{32}You shall be driven from men; and your dwelling shall be with the animals of the field. You shall be made to eat grass as oxen. Seven times shall pass over you, until you know that the Most High rules in the kingdom of men, and gives it to whomever he will.’ ”
\par }{\PI \VS{33}This was fulfilled the same hour on Nebuchadnezzar. He was driven from men, and ate grass as oxen, and his body was wet with the dew of the sky, until his hair had grown like eagles’ feathers, and his nails like birds’ claws.
\par }{\PI \VS{34}At the end of the days I, Nebuchadnezzar, lifted up my eyes to heaven, and my understanding returned to me, and I blessed the Most High, and I praised and honored him who lives forever;
\par }{\Q for his dominion is an everlasting dominion,
\par }{\QB and his kingdom from generation to generation.
\par }{\Q \VS{35}All the inhabitants of the earth are reputed as nothing;
\par }{\QB and he does according to his will in the army of heaven,
\par }{\QB and among the inhabitants of the earth;
\par }{\Q and no one can stop his hand,
\par }{\QB or ask him, “What are you doing?”
\par }{\PI \VS{36}At the same time my understanding returned to me; and for the glory of my kingdom, my majesty and brightness returned to me. My counselors and my lords sought me; and I was established in my kingdom, and excellent greatness was added to me.
\VS{37}Now I, Nebuchadnezzar, praise and extol and honor the King of heaven; for all his works are truth, and his ways justice; and those who walk in pride he is able to abase.

\par }\Chap{5}{\PP \VerseOne{1}Belshazzar the king made a great feast to a thousand of his lords, and drank wine before the thousand.
\VS{2}Belshazzar, while he tasted the wine, commanded that the golden and silver vessels which Nebuchadnezzar his father had taken out of the temple which was in Jerusalem be brought to him; that the king and his lords, his wives and his concubines, might drink from them.
\VS{3}Then they brought the golden vessels that were taken out of the temple of God’s house which was at Jerusalem; and the king and his lords, his wives and his concubines, drank from them.
\VS{4}They drank wine, and praised the gods of gold, and of silver, of bronze, of iron, of wood, and of stone.
\par }{\PP \VS{5}In the same hour, the fingers of a man’s hand came out and wrote near the lamp stand on the plaster of the wall of the king’s palace. The king saw the part of the hand that wrote.
\VS{6}Then the king’s face was changed in him, and his thoughts troubled him; and the joints of his thighs were loosened, and his knees struck one against another.
\par }{\PP \VS{7}The king cried aloud to bring in the enchanters, the Chaldeans, and the soothsayers. The king spoke and said to the wise men of Babylon, “Whoever reads this writing, and shows me its interpretation, shall be clothed with purple, and have a chain of gold about his neck, and shall be the third ruler in the kingdom.”
\par }{\PP \VS{8}Then all the king’s wise men came in; but they could not read the writing, and couldn’t make known to the king the interpretation.
\VS{9}Then king Belshazzar was greatly troubled, and his face was changed in him, and his lords were perplexed.
\par }{\PP \VS{10}The queen by reason of the words of the king and his lords came into the banquet house. The queen spoke and said, “O king, live forever; don’t let your thoughts trouble you, nor let your face be changed.
\VS{11}There is a man in your kingdom, in whom is the spirit of the holy gods; and in the days of your father light and understanding and wisdom, like the wisdom of the gods, were found in him. The king, Nebuchadnezzar, your father, yes, the king, your father, made him master of the magicians, enchanters, Chaldeans, and soothsayers;
\VS{12}because an excellent spirit, knowledge, understanding, interpreting of dreams, showing of dark sentences, and dissolving of doubts were found in the same Daniel, whom the king named Belteshazzar. Now let Daniel be called, and he will show the interpretation.”
\par }{\PP \VS{13}Then Daniel was brought in before the king. The king spoke and said to Daniel, “Are you that Daniel of the children of the captivity of Judah, whom the king my father brought out of Judah?
\VS{14}I have heard of you, that the spirit of the gods is in you, and that light, understanding, and excellent wisdom are found in you.
\VS{15}Now the wise men, the enchanters, have been brought in before me, that they should read this writing, and make known to me its interpretation; but they could not show the interpretation of the thing.
\VS{16}But I have heard of you, that you can give interpretations, and dissolve doubts. Now if you can read the writing, and make known to me its interpretation, you shall be clothed with purple, and have a chain of gold around your neck, and shall be the third ruler in the kingdom.”
\par }{\PP \VS{17}Then Daniel answered before the king, “Let your gifts be to yourself, and give your rewards to another. Nevertheless, I will read the writing to the king, and make known to him the interpretation.
\par }{\PP \VS{18}“To you, king, the Most High God gave Nebuchadnezzar your father the kingdom, and greatness, and glory, and majesty.
\VS{19}Because of the greatness that he gave him, all the peoples, nations, and languages trembled and feared before him. He killed whom he wanted to, and he kept alive whom he wanted to. He raised up whom he wanted to, and he put down whom he wanted to.
\VS{20}But when his heart was lifted up, and his spirit was hardened so that he dealt proudly, he was deposed from his kingly throne, and they took his glory from him.
\VS{21}He was driven from the sons of men, and his heart was made like the animals’, and his dwelling was with the wild donkeys. He was fed with grass like oxen, and his body was wet with the dew of the sky; until he knew that the Most High God rules in the kingdom of men, and that he sets up over it whomever he will.
\par }{\PP \VS{22}“You, his son, Belshazzar, have not humbled your heart, though you knew all this,
\VS{23}but have lifted up yourself against the Lord of heaven; and they have brought the vessels of his house before you, and you and your lords, your wives, and your concubines, have drunk wine from them. You have praised the gods of silver and gold, of bronze, iron, wood, and stone, which don’t see, or hear, or know; and you have not glorified the God in whose hand your breath is, and whose are all your ways.
\VS{24}Then the part of the hand was sent from before him, and this writing was inscribed.
\par }{\PP \VS{25}“This is the writing that was inscribed: ‘MENE, MENE, TEKEL, UPHARSIN.’
\par }{\PP \VS{26}“This is the interpretation of the thing:
\par }{\MM MENE: God has counted your kingdom, and brought it to an end.
\par }{\MM \VS{27}TEKEL: you are weighed in the balances, and are found wanting.
\par }{\MM \VS{28}PERES: your kingdom is divided, and given to the Medes and Persians.”
\par }{\PP \VS{29}Then Belshazzar commanded, and they clothed Daniel with purple, and put a chain of gold about his neck, and made proclamation concerning him, that he should be the third ruler in the kingdom.
\par }{\PP \VS{30}In that night Belshazzar the Chaldean King was slain.
\VS{31}Darius the Mede received the kingdom, being about sixty-two years old.

\par }\Chap{6}{\PP \VerseOne{1}It pleased Darius to set over the kingdom one hundred twenty local governors, who should be throughout the whole kingdom;
\VS{2}and over them three presidents, of whom Daniel was one; that these local governors might give account to them, and that the king should suffer no loss.
\VS{3}Then this Daniel was distinguished above the presidents and the local governors, because an excellent spirit was in him; and the king thought to set him over the whole realm.
\par }{\PP \VS{4}Then the presidents and the local governors sought to find occasion against Daniel as touching the kingdom; but they could find no occasion or fault, because he was faithful. There wasn’t any error or fault found in him.
\VS{5}Then these men said, “We won’t find any occasion against this Daniel, unless we find it against him concerning the law of his God.”
\par }{\PP \VS{6}Then these presidents and local governors assembled together to the king, and said this to him, “King Darius, live forever!
\VS{7}All the presidents of the kingdom, the deputies and the local governors, the counselors and the governors, have consulted together to establish a royal statute, and to make a strong decree, that whoever asks a petition of any god or man for thirty days, except of you, O king, he shall be cast into the den of lions.
\VS{8}Now, O king, establish the decree, and sign the writing, that it not be changed, according to the law of the Medes and Persians, which doesn’t alter.”
\VS{9}Therefore king Darius signed the writing and the decree.
\par }{\PP \VS{10}When Daniel knew that the writing was signed, he went into his house (now his windows were open in his room toward Jerusalem) and he kneeled on his knees three times a day, and prayed, and gave thanks before his God, as he did before.
\VS{11}Then these men assembled together, and found Daniel making petition and supplication before his God.
\VS{12}Then they came near, and spoke before the king concerning the king’s decree: “Haven’t you signed a decree that every man who makes a petition to any god or man within thirty days, except to you, O king, shall be cast into the den of lions?”
\par }{\PP The king answered, “This thing is true, according to the law of the Medes and Persians, which doesn’t alter.”
\par }{\PP \VS{13}Then they answered and said before the king, “That Daniel, who is of the children of the captivity of Judah, doesn’t respect you, O king, nor the decree that you have signed, but makes his petition three times a day.”
\VS{14}Then the king, when he heard these words, was very displeased, and set his heart on Daniel to deliver him; and he labored until the going down of the sun to rescue him.
\par }{\PP \VS{15}Then these men assembled together to the king, and said to the king, “Know, O king, that it is a law of the Medes and Persians, that no decree nor statute which the king establishes may be changed.”
\par }{\PP \VS{16}Then the king commanded, and they brought Daniel, and cast him into the den of lions. The king spoke and said to Daniel, “Your God whom you serve continually, he will deliver you.”
\par }{\PP \VS{17}A stone was brought, and laid on the mouth of the den; and the king sealed it with his own signet, and with the signet of his lords; that nothing might be changed concerning Daniel.
\VS{18}Then the king went to his palace, and passed the night fasting. No musical instruments were brought before him; and his sleep fled from him.
\par }{\PP \VS{19}Then the king arose very early in the morning, and went in haste to the den of lions.
\VS{20}When he came near to the den to Daniel, he cried with a troubled voice. The king spoke and said to Daniel, “Daniel, servant of the living God, is your God, whom you serve continually, able to deliver you from the lions?”
\par }{\PP \VS{21}Then Daniel said to the king, “O king, live forever!
\VS{22}My God has sent his angel, and has shut the lions’ mouths, and they have not hurt me; because as before him innocence was found in me; and also before you, O king, I have done no harm.”
\par }{\PP \VS{23}Then the king was exceedingly glad, and commanded that they should take Daniel up out of the den. So Daniel was taken up out of the den, and no kind of harm was found on him, because he had trusted in his God.
\par }{\PP \VS{24}The king commanded, and they brought those men who had accused Daniel, and they cast them into the den of lions, them, their children, and their wives; and the lions mauled them, and broke all their bones in pieces, before they came to the bottom of the den.
\par }{\PP \VS{25}Then king Darius wrote to all the peoples, nations, and languages, who dwell in all the earth:
\par }{\PI “Peace be multiplied to you.
\par }{\PI \VS{26}“I make a decree that in all the dominion of my kingdom men tremble and fear before the God of Daniel;
\par }{\Q “for he is the living God,
\par }{\QB and steadfast forever.
\par }{\Q His kingdom is that which will not be destroyed.
\par }{\QB His dominion will be even to the end.
\par }{\Q \VS{27}He delivers and rescues.
\par }{\QB He works signs and wonders in heaven and in earth,
\par }{\QB who has delivered Daniel from the power of the lions.”
\par }{\PP \VS{28}So this Daniel prospered in the reign of Darius, and in the reign of Cyrus the Persian.

\par }\Chap{7}{\PP \VerseOne{1}In the first year of Belshazzar king of Babylon Daniel had a dream and visions of his head on his bed. Then he wrote the dream and told the sum of the matters.
\par }{\PP \VS{2}Daniel spoke and said, “I saw in my vision by night, and, behold, the four winds of the sky broke out on the great sea.
\VS{3}Four great animals came up from the sea, different from one another.
\par }{\PP \VS{4}“The first was like a lion, and had eagle’s wings. I watched until its wings were plucked, and it was lifted up from the earth, and made to stand on two feet as a man. A man’s heart was given to it.
\par }{\PP \VS{5}“Behold, there was another animal, a second, like a bear. It was raised up on one side, and three ribs were in its mouth between its teeth. They said this to it: ‘Arise! Devour much flesh!’
\par }{\PP \VS{6}“After this I saw, and behold, another, like a leopard, which had on its back four wings of a bird. The animal also had four heads; and dominion was given to it.
\par }{\PP \VS{7}“After this I saw in the night visions, and, behold, there was a fourth animal, awesome and powerful, and exceedingly strong. It had great iron teeth. It devoured and broke in pieces, and stamped the residue with its feet. It was different from all the animals that were before it. It had ten horns.
\par }{\PP \VS{8}“I considered the horns, and behold, there came up among them another horn, a little one, before which three of the first horns were plucked up by the roots: and behold, in this horn were eyes like the eyes of a man, and a mouth speaking great things.
\par }{\Q \VS{9}“I watched until thrones were placed,
\par }{\QB and one who was ancient of days sat.
\par }{\Q His clothing was white as snow,
\par }{\QB and the hair of his head like pure wool.
\par }{\Q His throne was fiery flames,
\par }{\QB and its wheels burning fire.
\par }{\Q \VS{10}A fiery stream issued and came out from before him.
\par }{\QB Thousands of thousands ministered to him.
\par }{\QB Ten thousand times ten thousand stood before him.
\par }{\Q The judgment was set.
\par }{\QB The books were opened.
\par }{\PP \VS{11}“I watched at that time because of the voice of the great words which the horn spoke. I watched even until the animal was slain, and its body destroyed, and it was given to be burned with fire.
\VS{12}As for the rest of the animals, their dominion was taken away; yet their lives were prolonged for a season and a time.
\par }{\PP \VS{13}“I saw in the night visions, and behold, there came with the clouds of the sky one like a son of man, and he came even to the ancient of days, and they brought him near before him.
\VS{14}Dominion was given him, and glory, and a kingdom, that all the peoples, nations, and languages should serve him. His dominion is an everlasting dominion, which will not pass away, and his kingdom that which will not be destroyed.
\par }{\PP \VS{15}“As for me, Daniel, my spirit was grieved within my body, and the visions of my head troubled me.
\VS{16}I came near to one of those who stood by, and asked him the truth concerning all this.
\par }{\PP “So he told me, and made me know the interpretation of the things.
\VS{17}‘These great animals, which are four, are four kings, who will arise out of the earth.
\VS{18}But the saints of the Most High will receive the kingdom, and possess the kingdom forever, even forever and ever.’
\par }{\PP \VS{19}“Then I desired to know the truth concerning the fourth animal, which was different from all of them, exceedingly terrible, whose teeth were of iron, and its nails of bronze; which devoured, broke in pieces, and stamped the residue with its feet;
\VS{20}and concerning the ten horns that were on its head, and the other horn which came up, and before which three fell, even that horn that had eyes, and a mouth that spoke great things, whose look was more stout than its fellows.
\VS{21}I saw, and the same horn made war with the saints, and prevailed against them,
\VS{22}until the ancient of days came, and judgment was given to the saints of the Most High, and the time came that the saints possessed the kingdom.
\par }{\PP \VS{23}“Thus he said, ‘The fourth animal will be a fourth kingdom on earth, which will be different from all the kingdoms, and will devour the whole earth, and will tread it down, and break it in pieces.
\VS{24}As for the ten horns, ten kings will arise out of this kingdom. Another will arise after them; and he will be different from the former, and he will put down three kings.
\VS{25}He will speak words against the Most High, and will wear out the saints of the Most High. He will plan to change the times and the law; and they will be given into his hand until a time and times and half a time.
\par }{\PP \VS{26}“ ‘But the judgment will be set, and they will take away his dominion, to consume and to destroy it to the end.
\VS{27}The kingdom and the dominion, and the greatness of the kingdoms under the whole sky, will be given to the people of the saints of the Most High. His kingdom is an everlasting kingdom, and all dominions will serve and obey him.’
\par }{\PP \VS{28}“Here is the end of the matter. As for me, Daniel, my thoughts much troubled me, and my face was changed in me; but I kept the matter in my heart.”

\par }\Chap{8}{\PP \VerseOne{1}In the third year of the reign of king Belshazzar a vision appeared to me, even to me, Daniel, after that which appeared to me at the first.
\VS{2}I saw the vision. Now it was so, that when I saw, I was in the citadel of Susa, which is in the province of Elam. I saw in the vision, and I was by the river Ulai.
\VS{3}Then I lifted up my eyes, and saw, and behold, there stood before the river a ram which had two horns. The two horns were high; but one was higher than the other, and the higher came up last.
\VS{4}I saw the ram pushing westward, northward, and southward. No animals could stand before him. There wasn’t any who could deliver out of his hand; but he did according to his will, and magnified himself.
\par }{\PP \VS{5}As I was considering, behold, a male goat came from the west over the surface of the whole earth, and didn’t touch the ground. The goat had a notable horn between his eyes.
\VS{6}He came to the ram that had the two horns, which I saw standing before the river, and ran on him in the fury of his power.
\VS{7}I saw him come close to the ram, and he was moved with anger against him, and struck the ram, and broke his two horns. There was no power in the ram to stand before him; but he cast him down to the ground, and trampled on him. There was no one who could deliver the ram out of his hand.
\VS{8}The male goat magnified himself exceedingly. When he was strong, the great horn was broken; and instead of it there came up four notable horns toward the four winds of the sky.
\par }{\PP \VS{9}Out of one of them came out a little horn, which grew exceedingly great, toward the south, and toward the east, and toward the glorious land.
\VS{10}It grew great, even to the army of the sky; and it cast down some of the army and of the stars to the ground, and trampled on them.
\VS{11}Yes, it magnified itself, even to the prince of the army; and it took away from him the continual burnt offering, and the place of his sanctuary was cast down.
\VS{12}The army was given over to it together with the continual burnt offering through disobedience. It cast down truth to the ground, and it did its pleasure and prospered.
\par }{\PP \VS{13}Then I heard a holy one speaking; and another holy one said to that certain one who spoke, “How long will the vision about the continual burnt offering, and the disobedience that makes desolate, to give both the sanctuary and the army to be trodden under foot be?”
\par }{\PP \VS{14}He said to me, “To two thousand and three hundred evenings and mornings. Then the sanctuary will be cleansed.”
\par }{\PP \VS{15}When I, even I Daniel, had seen the vision, I sought to understand it. Then behold, there stood before me something like the appearance of a man.
\VS{16}I heard a man’s voice between the banks of the Ulai, which called, and said, “Gabriel, make this man understand the vision.”
\par }{\PP \VS{17}So he came near where I stood; and when he came, I was frightened, and fell on my face; but he said to me, “Understand, son of man; for the vision belongs to the time of the end.”
\par }{\PP \VS{18}Now as he was speaking with me, I fell into a deep sleep with my face toward the ground; but he touched me, and set me upright.
\par }{\PP \VS{19}He said, “Behold, I will make you know what will be in the latter time of the indignation; for it belongs to the appointed time of the end.
\VS{20}The ram which you saw, that had the two horns, they are the kings of Media and Persia.
\VS{21}The rough male goat is the king of Greece. The great horn that is between his eyes is the first king.
\VS{22}As for that which was broken, in the place where four stood up, four kingdoms will stand up out of the nation, but not with his power.
\par }{\PP \VS{23}“In the latter time of their kingdom, when the transgressors have come to the full, a king of fierce face, and understanding dark sentences, will stand up.
\VS{24}His power will be mighty, but not by his own power. He will destroy awesomely, and will prosper in what he does. He will destroy the mighty ones and the holy people.
\VS{25}Through his policy he will cause deceit to prosper in his hand. He will magnify himself in his heart, and he will destroy many in their security. He will also stand up against the prince of princes; but he will be broken without hand.
\par }{\PP \VS{26}“The vision of the evenings and mornings which has been told is true; but seal up the vision, for it belongs to many days to come.”
\par }{\PP \VS{27}I, Daniel, fainted, and was sick for some days. Then I rose up, and did the king’s business. I wondered at the vision, but no one understood it.

\par }\Chap{9}{\PP \VerseOne{1}In the first year of Darius the son of Ahasuerus, of the offspring of the Medes, who was made king over the realm of the Chaldeans,
\VS{2}in the first year of his reign I, Daniel, understood by the books the number of the years about which Yahweh’s word\FTNT{*}{{\FR 9:2 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.
} came to Jeremiah the prophet, for the accomplishing of the desolations of Jerusalem, even seventy years.
\VS{3}I set my face to the Lord God, to seek by prayer and petitions, with fasting and sackcloth and ashes.
\par }{\PP \VS{4}I prayed to Yahweh my God, and made confession, and said,
\par }{\PI “Oh, Lord, the great and dreadful God, who keeps covenant and loving kindness with those who love him and keep his commandments,
\VS{5}we have sinned, and have dealt perversely, and have done wickedly, and have rebelled, even turning aside from your precepts and from your ordinances.
\VS{6}We haven’t listened to your servants the prophets, who spoke in your name to our kings, our princes, and our fathers, and to all the people of the land.
\par }{\PI \VS{7}“Lord, righteousness belongs to you, but to us confusion of face, as it is today; to the men of Judah, and to the inhabitants of Jerusalem, and to all Israel, who are near, and who are far off, through all the countries where you have driven them, because of their trespass that they have trespassed against you.
\VS{8}Lord, to us belongs confusion of face, to our kings, to our princes, and to our fathers, because we have sinned against you.
\VS{9}To the Lord our God belong mercies and forgiveness; for we have rebelled against him.
\VS{10}We haven’t obeyed Yahweh our God’s voice, to walk in his laws, which he set before us by his servants the prophets.
\VS{11}Yes, all Israel have transgressed your law, turning aside, that they should not obey your voice.
\par }{\PI “Therefore the curse and the oath written in the law of Moses the servant of God has been poured out on us; for we have sinned against him.
\VS{12}He has confirmed his words, which he spoke against us, and against our judges who judged us, by bringing on us a great evil; for under the whole sky, such has not been done as has been done to Jerusalem.
\VS{13}As it is written in the law of Moses, all this evil has come on us. Yet we have not entreated the favor of Yahweh our God, that we should turn from our iniquities and have discernment in your truth.
\VS{14}Therefore Yahweh has watched over the evil, and brought it on us; for Yahweh our God is righteous in all his works which he does, and we have not obeyed his voice.
\par }{\PI \VS{15}“Now, Lord our God, who has brought your people out of the land of Egypt with a mighty hand, and have gotten yourself renown, as it is today; we have sinned. We have done wickedly.
\VS{16}Lord, according to all your righteousness, let your anger and please let your wrath be turned away from your city Jerusalem, your holy mountain; because for our sins, and for the iniquities of our fathers, Jerusalem and your people have become a reproach to all who are around us.
\par }{\PI \VS{17}“Now therefore, our God, listen to the prayer of your servant, and to his petitions, and cause your face to shine on your sanctuary that is desolate, for the Lord’s sake.
\VS{18}My God, turn your ear, and hear. Open your eyes, and see our desolations, and the city which is called by your name; for we do not present our petitions before you for our righteousness, but for your great mercies’ sake.
\VS{19}Lord, hear. Lord, forgive. Lord, listen and do. Don’t defer, for your own sake, my God, because your city and your people are called by your name.”
\par }{\BB \par }{\PP \VS{20}While I was speaking, and praying, and confessing my sin and the sin of my people Israel, and presenting my supplication before Yahweh my God for the holy mountain of my God;
\VS{21}yes, while I was speaking in prayer, the man Gabriel, whom I had seen in the vision at the beginning, being caused to fly swiftly, touched me about the time of the evening offering.
\VS{22}He instructed me and talked with me, and said, “Daniel, I have now come to give you wisdom and understanding.
\VS{23}At the beginning of your petitions the commandment went out, and I have come to tell you; for you are greatly beloved. Therefore consider the matter, and understand the vision.
\par }{\PP \VS{24}“Seventy weeks are decreed on your people and on your holy city, to finish disobedience, and to make an end of sins, and to make reconciliation for iniquity, and to bring in everlasting righteousness, and to seal up vision and prophecy, and to anoint the most holy.
\par }{\PP \VS{25}“Know therefore and discern that from the going out of the commandment to restore and to build Jerusalem to the Anointed One,\FTNT{†}{{\FR 9:25 }
“Anointed One” can also be translated “Messiah” (same as “Christ”).} the prince, will be seven weeks and sixty-two weeks. It will be built again, with street and moat, even in troubled times.
\VS{26}After the sixty-two weeks the Anointed One\FTNT{‡}{{\FR 9:26 }
“Anointed One” can also be translated “Messiah” (same as “Christ”).} will be cut off, and will have nothing. The people of the prince who come will destroy the city and the sanctuary. Its end will be with a flood, and war will be even to the end. Desolations are determined.
\VS{27}He will make a firm covenant with many for one week. In the middle of the week he will cause the sacrifice and the offering to cease. On the wing of abominations will come one who makes desolate; and even to the full end, and that determined, wrath will be poured out on the desolate.”

\par }\Chap{10}{\PP \VerseOne{1}In the third year of Cyrus king of Persia a thing was revealed to Daniel, whose name was called Belteshazzar; and the thing was true, even a great warfare. He understood the thing, and had understanding of the vision.
\par }{\PP \VS{2}In those days I, Daniel, was mourning three whole weeks.
\VS{3}I ate no pleasant bread. No meat or wine came into my mouth. I didn’t anoint myself at all, until three whole weeks were fulfilled.
\par }{\PP \VS{4}In the twenty-fourth day of the first month, as I was by the side of the great river, which is Hiddekel,
\VS{5}I lifted up my eyes, and looked, and behold, there was a man clothed in linen, whose thighs were adorned with pure gold of Uphaz.
\VS{6}His body also was like beryl, and his face as the appearance of lightning, and his eyes as flaming torches. His arms and his feet were like burnished bronze. The voice of his words was like the voice of a multitude.
\par }{\PP \VS{7}I, Daniel, alone saw the vision; for the men who were with me didn’t see the vision; but a great quaking fell on them, and they fled to hide themselves.
\VS{8}So I was left alone, and saw this great vision. No strength remained in me; for my face grew deathly pale, and I retained no strength.
\VS{9}Yet I heard the voice of his words. When I heard the voice of his words, then I fell into a deep sleep on my face, with my face toward the ground.
\par }{\PP \VS{10}Behold, a hand touched me, which set me on my knees and on the palms of my hands.
\VS{11}He said to me, “Daniel, you greatly beloved man, understand the words that I speak to you, and stand upright; for I have been sent to you, now.” When he had spoken this word to me, I stood trembling.
\par }{\PP \VS{12}Then he said to me, “Don’t be afraid, Daniel; for from the first day that you set your heart to understand, and to humble yourself before your God, your words were heard. I have come for your words’ sake.
\VS{13}But the prince of the kingdom of Persia withstood me twenty-one days; but, behold, Michael, one of the chief princes, came to help me because I remained there with the kings of Persia.
\VS{14}Now I have come to make you understand what will happen to your people in the latter days; for the vision is yet for many days.”
\par }{\PP \VS{15}When he had spoken these words to me, I set my face toward the ground, and was mute.
\VS{16}Behold, one in the likeness of the sons of men touched my lips. Then I opened my mouth, and spoke and said to him who stood before me, “My lord, by reason of the vision my sorrows have overtaken me, and I retain no strength.
\VS{17}For how can the servant of this my lord talk with this my lord? For as for me, immediately there remained no strength in me. There was no breath left in me.”
\par }{\PP \VS{18}Then one like the appearance of a man touched me again, and he strengthened me.
\VS{19}He said, “Greatly beloved man, don’t be afraid. Peace be to you. Be strong. Yes, be strong.”
\par }{\PP When he spoke to me, I was strengthened, and said, “Let my lord speak; for you have strengthened me.”
\par }{\PP \VS{20}Then he said, “Do you know why I have come to you? Now I will return to fight with the prince of Persia. When I go out, behold, the prince of Greece will come.
\VS{21}But I will tell you that which is inscribed in the writing of truth. There is no one who holds with me against these, but Michael your prince.

\par }\Chap{11}{\PP \VerseOne{1}“As for me, in the first year of Darius the Mede, I stood up to confirm and strengthen him.
\par }{\PP \VS{2}“Now I will show you the truth. Behold, three more kings will stand up in Persia; and the fourth will be far richer than all of them. When he has grown strong through his riches, he will stir up all against the realm of Greece.
\VS{3}A mighty king will stand up, who will rule with great dominion, and do according to his will.
\VS{4}When he stands up, his kingdom will be broken, and will be divided toward the four winds of the sky, but not to his posterity, nor according to his dominion with which he ruled; for his kingdom will be plucked up, even for others besides these.
\par }{\PP \VS{5}“The king of the south will be strong. One of his princes will become stronger than him, and have dominion. His dominion will be a great dominion.
\VS{6}At the end of years they will join themselves together; and the daughter of the king of the south will come to the king of the north to make an agreement; but she will not retain the strength of her arm. He will also not stand, nor will his arm; but she will be given up, with those who brought her, and he who became the father of her, and he who strengthened her in those times.
\par }{\PP \VS{7}“But out of a shoot from her roots one will stand up in his place, who will come to the army, and will enter into the fortress of the king of the north, and will deal against them, and will prevail.
\VS{8}He will also carry their gods, with their molten images, and with their goodly vessels of silver and of gold, captive into Egypt. He will refrain some years from the king of the north.
\VS{9}He will come into the realm of the king of the south, but he will return into his own land.
\VS{10}His sons will wage war, and will assemble a multitude of great forces, which will come on, and overflow, and pass through. They will return and wage war, even to his fortress.
\par }{\PP \VS{11}“The king of the south will be moved with anger, and will come out and fight with him, even with the king of the north. He will send out a great multitude, and the multitude will be given into his hand.
\VS{12}The multitude will be lifted up, and his heart will be exalted. He will cast down tens of thousands, but he won’t prevail.
\VS{13}The king of the north will return, and will send out a multitude greater than the former. He will come on at the end of the times, even of years, with a great army and with much substance.
\par }{\PP \VS{14}“In those times many will stand up against the king of the south. Also the children of the violent among your people will lift themselves up to establish the vision; but they will fall.
\VS{15}So the king of the north will come and cast up a mound, and take a well-fortified city. The forces of the south won’t stand, neither will his chosen people, neither will there be any strength to stand.
\VS{16}But he who comes against him will do according to his own will, and no one will stand before him. He will stand in the glorious land, and destruction will be in his hand.
\VS{17}He will set his face to come with the strength of his whole kingdom, and with him equitable conditions. He will perform them. He will give him the daughter of women, to corrupt her; but she will not stand, and won’t be for him.
\VS{18}After this he will turn his face to the islands, and will take many; but a prince will cause the reproach offered by him to cease. Yes, moreover, he will cause his reproach to turn on him.
\VS{19}Then he will turn his face toward the fortresses of his own land; but he will stumble and fall, and won’t be found.
\par }{\PP \VS{20}“Then one who will cause a tax collector to pass through the kingdom to maintain its glory will stand up in his place; but within few days he shall be destroyed, not in anger, and not in battle.
\par }{\PP \VS{21}“In his place a contemptible person will stand up, to whom they had not given the honor of the kingdom; but he will come in time of security, and will obtain the kingdom by flatteries.
\VS{22}The overwhelming forces will be overwhelmed from before him, and will be broken. Yes, also the prince of the covenant.
\VS{23}After the treaty made with him he will work deceitfully; for he will come up, and will become strong, with a small people.
\VS{24}In time of security he will come even on the fattest places of the province. He will do that which his fathers have not done, nor his fathers’ fathers. He will scatter among them prey, plunder, and substance. Yes, he will devise his plans against the strongholds, even for a time.
\par }{\PP \VS{25}“He will stir up his power and his courage against the king of the south with a great army; and the king of the south will wage war in battle with an exceedingly great and mighty army; but he won’t stand; for they will devise plans against him.
\VS{26}Yes, those who eat of his dainties will destroy him, and his army will be swept away. Many will fall down slain.
\VS{27}As for both these kings, their hearts will be to do mischief, and they will speak lies at one table; but it won’t prosper, for the end will still be at the appointed time.
\VS{28}Then he will return into his land with great wealth. His heart will be against the holy covenant. He will take action, and return to his own land.
\par }{\PP \VS{29}“He will return at the appointed time, and come into the south; but it won’t be in the latter time as it was in the former.
\VS{30}For ships of Kittim will come against him. Therefore he will be grieved, and will return, and have indignation against the holy covenant, and will take action. He will even return, and have regard to those who forsake the holy covenant.
\par }{\PP \VS{31}“Forces will stand on his part, and they will profane the sanctuary, even the fortress, and will take away the continual burnt offering. Then they will set up the abomination that makes desolate.
\VS{32}He will corrupt those who do wickedly against the covenant by flatteries; but the people who know their God will be strong, and take action.
\par }{\PP \VS{33}“Those who are wise among the people will instruct many; yet they will fall by the sword and by flame, by captivity and by plunder, many days.
\VS{34}Now when they fall, they will be helped with a little help; but many will join themselves to them with flatteries.
\VS{35}Some of those who are wise will fall, to refine them, and to purify, and to make them white, even to the time of the end; because it is yet for the time appointed.
\par }{\PP \VS{36}“The king will do according to his will. He will exalt himself, and magnify himself above every god, and will speak marvelous things against the God of gods. He will prosper until the indignation is accomplished; for that which is determined will be done.
\VS{37}He won’t regard the gods of his fathers, or the desire of women, or regard any god; for he will magnify himself above all.
\VS{38}But in his place he will honor the god of fortresses. He will honor a god whom his fathers didn’t know with gold, silver, and with precious stones and pleasant things.
\VS{39}He will deal with the strongest fortresses by the help of a foreign god. He will increase with glory whoever acknowledges him. He will cause them to rule over many, and will divide the land for a price.
\par }{\PP \VS{40}“At the time of the end the king of the south will contend with him; and the king of the north will come against him like a whirlwind, with chariots, with horsemen, and with many ships. He will enter into the countries, and will overflow and pass through.
\VS{41}He will enter also into the glorious land, and many countries will be overthrown; but these will be delivered out of his hand: Edom, Moab, and the chief of the children of Ammon.
\VS{42}He will also stretch out his hand on the countries. The land of Egypt won’t escape.
\VS{43}But he will have power over the treasures of gold and of silver, and over all the precious things of Egypt. The Libyans and the Ethiopians will be at his steps.
\VS{44}But news out of the east and out of the north will trouble him; and he will go out with great fury to destroy and utterly to sweep away many.
\VS{45}He will plant the tents of his palace between the sea and the glorious holy mountain; yet he will come to his end, and no one will help him.

\par }\Chap{12}{\PP \VerseOne{1}“At that time Michael will stand up, the great prince who stands for the children of your people; and there will be a time of trouble, such as never was since there was a nation even to that same time. At that time your people will be delivered, everyone who is found written in the book.
\VS{2}Many of those who sleep in the dust of the earth will awake, some to everlasting life, and some to shame and everlasting contempt.
\VS{3}Those who are wise will shine as the brightness of the expanse. Those who turn many to righteousness will shine as the stars forever and ever.
\VS{4}But you, Daniel, shut up the words, and seal the book, even to the time of the end. Many will run back and forth, and knowledge will be increased.”
\par }{\PP \VS{5}Then I, Daniel, looked, and behold, two others stood, one on the river bank on this side, and the other on the river bank on that side.
\VS{6}One said to the man clothed in linen, who was above the waters of the river, “How long will it be to the end of these wonders?”
\par }{\PP \VS{7}I heard the man clothed in linen, who was above the waters of the river, when he held up his right hand and his left hand to heaven, and swore by him who lives forever that it will be for a time, times, and a half; and when they have finished breaking in pieces the power of the holy people, all these things will be finished.
\par }{\PP \VS{8}I heard, but I didn’t understand. Then I said, “My lord, what will be the outcome of these things?”
\par }{\PP \VS{9}He said, “Go your way, Daniel; for the words are shut up and sealed until the time of the end.
\VS{10}Many will purify themselves, and make themselves white, and be refined; but the wicked will do wickedly; and none of the wicked will understand; but those who are wise will understand.
\par }{\PP \VS{11}“From the time that the continual burnt offering is taken away, and the abomination that makes desolate set up, there will be one thousand two hundred ninety days.
\VS{12}Blessed is he who waits, and comes to the one thousand three hundred thirty-five days.
\par }{\PP \VS{13}“But go your way until the end; for you will rest, and will stand in your inheritance at the end of the days.”

\par }\Chap{13}{\SH THE HISTORY OF SUSANNA
\par }{\PP \VerseOne{1}\FTNT{*}{{\FR 13:1 }

{\BK{The History of Susanna}} is translated from chapter 13 of
{\BK{Daniel}} in the Greek Septuagint. It is not found in the traditional Hebrew text of
{\BK{Daniel}}.
{\BK{The History of Susanna}} is recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches.}A man lived in Babylon, and his name was Joakim.
\VS{2}He took a wife, whose name was Susanna, the daughter of Helkias, a very fair woman, and one who feared the Lord.
\VS{3}Her parents were also righteous, and taught their daughter according to the law of Moses.
\VS{4}Now Joakim was a great rich man, and had a fair garden joining to his house. The Jews used to come to him, because he was more honorable than all others.
\VS{5}The same year two of the elders of the people were appointed to be judges, such as the Lord spoke of, that wickedness came from Babylon from elders who were judges, who were supposed to govern the people.
\VS{6}These were often at Joakim’s house. All that had any suits in law came to them.
\par }{\PP \VS{7}When the people departed away at noon, Susanna went into her husband’s garden to walk.
\VS{8}The two elders saw her going in every day, and walking; and they were inflamed with lust for her.
\VS{9}They perverted their own mind and turned away their eyes, that they might not look to heaven, nor remember just judgments.
\VS{10}And although they both were wounded with lust for her, yet dared not show the other his grief.
\VS{11}For they were ashamed to declare their lust, that they desired to have to do with her.
\VS{12}Yet they watched jealously from day to day to see her.
\VS{13}The one said to the other, “Let’s go home, now; for it is dinner time.”
\VS{14}So when they had gone out, they parted company, and turning back again, they came to the same place. After they had asked one another the cause, they acknowledged their lust. Then they appointed a time both together, when they might find her alone.
\par }{\PP \VS{15}It happened, as they watched on an opportune day, she went in as before with only two maids, and she desired to wash herself in the garden; for it was hot.
\VS{16}There was nobody there except the two elders who had hid themselves and watched her.
\VS{17}Then she said to her maids, “Bring me oil and washing balls, and shut the garden doors, that I may wash myself.”
\VS{18}They did as she asked them, and shut the garden doors, and went out themselves at the side doors to fetch the things that she had commanded them. They didn’t see the elders, because they were hidden.
\VS{19}Now when the maids had gone out, the two elders rose up, and ran to her, saying,
\VS{20}“Behold, the garden doors are shut, that no man can see us, and we are in love with you. Therefore consent to us, and lie with us.
\VS{21}If you will not, we will testify against you, that a young man was with you; therefore you sent your maids away from you.”
\par }{\PP \VS{22}Then Susanna sighed, and said, “I am trapped; for if I do this thing, it is death to me. If I don’t do it, I can’t escape your hands.
\VS{23}It is better for me to fall into your hands, and not do it, than to sin in the sight of the Lord.”
\VS{24}With that Susanna cried with a loud voice; and the two elders cried out against her.
\VS{25}Then one of them ran and opened the garden doors.
\par }{\PP \VS{26}So when the servants of the house heard the cry in the garden, they rushed in at the side door, to see what had happened to her.
\VS{27}But when the elders had told their tale, the servants were greatly ashamed; for there was never such a report made of Susanna.
\par }{\PP \VS{28}It came to pass on the next day, when the people assembled to her husband Joakim, the two elders came full of their wicked intent against Susanna to put her to death,
\VS{29}and said before the people, “Send for Susanna, the daughter of Helkias, Joakim’s wife.” So they sent;
\VS{30}and she came with her father and mother, her children, and all her kindred.
\VS{31}Now Susanna was a very delicate woman, and beautiful to behold.
\VS{32}These wicked men commanded her to be unveiled, for she was veiled, that they might be filled with her beauty.
\VS{33}Therefore her friends and all who saw her wept.
\VS{34}Then the two elders stood up in the midst of the people, and laid their hands upon her head.
\VS{35}She, weeping, looked up toward heaven; for her heart trusted in the Lord.
\par }{\PP \VS{36}The elders said, “As we walked in the garden alone, this woman came in with two maids, shut the garden doors, and sent the maids away.
\VS{37}Then a young man, who was hidden there, came to her and lay with her.
\VS{38}And we, being in a corner of the garden, saw this wickedness, and ran to them.
\VS{39}And when we saw them together, we couldn’t hold the man; for he was stronger than we, and opened the doors, and leaped out.
\VS{40}But having taken this woman, we asked who the young man was, but she would not tell us. We testify these things.
\par }{\PP \VS{41}Then the assembly believed them, as those who were elders of the people and judges; so they condemned her to death.
\par }{\PP \VS{42}Then Susanna cried out with a loud voice, and said, “O everlasting God, you know the secrets, and know all things before they happen,
\VS{43}you know that they have testified falsly against me. Behold, I must die, even though I never did such things as these men have maliciously invented against me.”
\par }{\PP \VS{44}The Lord heard her voice.
\VS{45}Therefore when she was led away to be put to death, God raised up the holy spirit of a young youth, whose name was Daniel.
\VS{46}He cried with a loud voice, “I am clear from the blood of this woman!”
\par }{\PP \VS{47}Then all the people turned them toward him, and said, “What do these words that you have spoken mean?”
\par }{\PP \VS{48}So he, standing in the midst of them, said, “Are you all such fools, you sons of Israel, that without examination or knowledge of the truth you have condemned a daughter of Israel?
\VS{49}Return again to the place of judgement; for these have testified falsly against her.”
\par }{\PP \VS{50}Therefore all the people turned again in haste, and the elders said to him, “Come, sit down among us, and show it to us, seeing God has given you the honor of an elder.”
\par }{\PP \VS{51}Then Daniel said to them, “Put them far apart from each another, and I will examine them.”
\VS{52}So when they were put apart one from another, he called one of them, and said to him, “O you who have become old in wickedness, now your sins have returned which you have committed before,
\VS{53}in pronouncing unjust judgement, condemning the innocent, and letting the guilty go free; although the Lord says, ‘You shall not kill the innocent and righteous.’
\VS{54}Now then, if you saw her, tell me, under which tree did you see them companying together?”
\par }{\PP He answered, “Under a mastick tree.”
\par }{\PP \VS{55}And Daniel said, “You have certainly lied against your own head; for even now the angel of God has received the sentence of God and will cut you in two.”
\VS{56}So he put him aside, and commanded to bring the other, and said to him, “O you seed of Canaan, and not of Judah, beauty has deceived you, and lust has perverted your heart.
\VS{57}Thus you have dealt with the daughters of Israel, and they for fear were intimate with you; but the daughter of Judah would not tolerate your wickedness.
\VS{58}Now therefore tell me, under which tree did you take them being intimate together?”
\par }{\PP Heanswered, “Under an evergreen oak tree.”
\par }{\PP \VS{59}Then Daniel said to him, You have also certainly lied against your own head; for the angel of God waits with the sword to cut you in two, that he may destroy you.”
\par }{\PP \VS{60}With that all the assembly cried out with a loud voice, and blessed God, who saves those who hope in him.
\VS{61}Then they arose against the two elders, for Daniel had convicted them of false testimony out of their own mouth.
\VS{62}According to the law of Moses they did to them in such sort as they maliciously intended to do to their neighbor. They put them to death, and the innocent blood was saved the same day.
\VS{63}Therefore Helkias and his wife praised God for their daughter Susanna, with Joakim her husband, and all the kindred, because there was no dishonesty found in her.
\VS{64}And from that day forth was Daniel had in great reputation in the sight of the people.

\par }\Chap{14}{\SH Bel and the Dragon
\par }{\PP \VerseOne{1}\FTNT{*}{{\FR 14:1 }

{\BK{Bel and the Dragon}} is translated from chapter 14 of
{\BK{Daniel}} in the Greek Septuagint. It is not found in the traditional Hebrew text of
{\BK{Daniel}}.
{\BK{Bel and the Dragon}} is recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches.}King Astyages was gathered to his fathers, and Cyrus the Persian received his kingdom.
\VS{2}Daniel lived with the king, and was honored above all his friends.
\VS{3}Now the Babylonians had an idol, called Bel, and there were spent upon him every day twelve great measures of fine flour, and forty sheep, and six firkins\FTNT{†}{{\FR 14:3 }
a firkin is about 41 liters or 11 gallons.} of wine.
\VS{4}And the king honored it, and went daily to worship it; but Daniel worshiped his own God. The king said to him, “Why don’t you worship Bel?”
\par }{\PP \VS{5}He said, “Because I may not honor idols made with hands, but only the living God, who has created the sky and the earth, and has sovereignty over all flesh.”
\par }{\PP \VS{6}Then the king said to him, “Don’t you that Bel is a living god? Don’t you see how much he eats and drinks every day?”
\par }{\PP \VS{7}Then Daniel laughed, and said, “O king, don’t be deceived; for this is just clay inside, and brass outside, and never ate or drank anything.”
\par }{\PP \VS{8}So the king was angry, and called for his priests, and said to them, “If you don’t tell me who this is who devours these expenses, you shall die.
\VS{9}But if you can show me that Bel devours them, then Daniel shall die; for he has spoken blasphemy against Bel.”
\par }{\PP Daniel said to the king, “Let it be according to your word.”
\par }{\PP \VS{10}Now there were seventy priests of Bel, besides their wives and children. The king went with Daniel into Bel’s temple.
\VS{11}So Bel’s priests said, “Behold, we will leave; but you, O king, set out the meat, and mix the wine and set it out, shut the door securely, and seal it with your own signet.
\VS{12}When you come in the morning, if you find not that Bel has eaten up everything, we will suffer death, or else Daniel, who speaks falsely against us.”
\par }{\PP \VS{13}They weren’t concerned, for under the table they had made a secret entrance, whereby they entered in continually, and consumed those things.
\VS{14}It happened, when they had gone out, the king set the meat before Bel. Now Daniel had commanded his servants to bring ashes, and they scattered them all over the temple in the presence of the king alone. Then they went out, shut the door, sealed it with the king’s signet, and so departed.
\par }{\PP \VS{15}Now in the night, the priests came with their wives and children, as they usually did, and ate and drank it all.
\VS{16}In the morning, the king arose, and Daniel with him.
\VS{17}The king said, “Daniel, are the seals whole?”
\par }{\PP He said, “Yes, O king, they are whole.”
\par }{\PP \VS{18}And as soon as he had opened the door, the king looked at the table, and cried with a loud voice, “You are great, O Bel, and with you is no deceit at all!”
\par }{\PP \VS{19}Then Daniel laughed, and held the king that he should not go in, and said, “Behold now the pavement, and mark well whose footsteps these are.”
\par }{\PP \VS{20}The king said, “I see the footsteps of men, women, and children.” Then the king was angry,
\VS{21}and took the priests with their wives and children, who showed him the secret doors, where they came in, and consumed the things that were on the table.
\VS{22}Therefore the king killed them, and delivered Bel into Daniel’s power, who overthrew him and his temple.
\par }{\PP \VS{23}In that same place there was a great dragon, which the people of Babylon worshiped.
\VS{24}The king said to Daniel, “Will you also say that this is of brass? Behold, he lives, and eats and drinks. You can’t say that he is no living god. Therefore worship him.”
\par }{\PP \VS{25}Then Daniel said, “I will worship the Lord my God; for he is a living God.
\VS{26}But allow me, O king, and I will kill this dragon without sword or staff.”
\par }{\PP The king said, “I allow you.”
\par }{\PP \VS{27}Then Daniel took pitch, fat, and hair, and melted them together, and made lumps of them. He put these in the dragon’s mouth, so the dragon ate and burst apart. Daniel said, “Behold, these are the gods you all worship.”
\par }{\PP \VS{28}When the people of Babylon heard that, they took great indignation, and conspired against the king, saying, “The king has become a Jew. He has pulled down Bel, slain the dragon, and put the priests to the sword.”
\VS{29}So they came to the king, and said, “Deliver Daniel to us, or else we will destroy you and your house.”
\par }{\PP \VS{30}Now when the king saw that they trapped him, being constrained, the king delivered Daniel to them.
\VS{31}They cast him into the lion’s den, where he was six days.
\VS{32}There were seven lions in the den, and they had been giving them two carcasses and two sheep every day, which then were not given to them, intending that they would devour Daniel.
\par }{\PP \VS{33}Now there was in Jewry the prophet
\FTNT{‡}{{\FR 14:33 }
Gr.
Ambakoum.
}Habakkuk, who had made stew, and had broken bread into a bowl. He was going into the field to bring it to the reapers.
\VS{34}But the angel of the Lord said to Habakkuk, “Go carry the dinner that you have into Babylon to Daniel, in the lions’ den.”
\par }{\PP \VS{35}Habakkuk said, “Lord, I never saw Babylon. I don’t know where the den is.”
\par }{\PP \VS{36}Then the angel of the Lord took him by the crown, and lifted him up by the hair of his head, and with the blast of his breath set him in Babylon over the den.
\par }{\PP \VS{37}Habakkuk cried, saying, “O Daniel, Daniel, take the dinner which God has sent you.”
\par }{\PP \VS{38}Daniel said, “You have remembered me, O God! You haven’t forsaken those who love you!”
\VS{39}So Daniel arose and ate; and the angel of God set Habakkuk in his own place again immediately.
\VS{40}On the seventh day, the king came to mourn for Daniel. When he came to the den, he looked in, and, behold, Daniel was sitting.
\VS{41}Then the king cried with a loud voice, saying, “Great are you, O Lord, you God of Daniel, and there is none other beside you!”
\VS{42}So he drew him out, and cast those that were the cause of his destruction into the den; and they were devoured in a moment before his face.
\par }