\NormalFont\ShortTitle{Isaiah}
{\MTB The Book of the Prophet
\par }{\MT Isaiah

\par }\ChapOne{1}{\PP \VerseOne{1}The vision of Isaiah the son of Amoz, which he saw concerning Judah and Jerusalem, in the days of Uzziah, Jotham, Ahaz, and Hezekiah, kings of Judah.
\par }{\Q \VS{2}Hear, heavens,
\par }{\QB and listen, earth; for Yahweh\FTNT{*}{{\FR 1:2 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.} has spoken:
\par }{\Q “I have nourished and brought up children
\par }{\QB and they have rebelled against me.
\par }{\Q \VS{3}The ox knows his owner,
\par }{\QB and the donkey his master’s crib;
\par }{\QB but Israel doesn’t know.
\par }{\QB My people don’t consider.”
\par }{\Q \VS{4}Ah sinful nation,
\par }{\QB a people loaded with iniquity,
\par }{\QB offspring\FTNT{†}{{\FR 1:4 }
or, seed} of evildoers,
\par }{\QB children who deal corruptly!
\par }{\Q They have forsaken Yahweh.
\par }{\QB They have despised the Holy One of Israel.
\par }{\QB They are estranged and backward.
\par }{\Q \VS{5}Why should you be beaten more,
\par }{\QB that you revolt more and more?
\par }{\Q The whole head is sick,
\par }{\QB and the whole heart faint.
\par }{\Q \VS{6}From the sole of the foot even to the head there is no soundness in it:
\par }{\QB wounds, welts, and open sores.
\par }{\QB They haven’t been closed, bandaged, or soothed with oil.
\par }{\Q \VS{7}Your country is desolate.
\par }{\QB Your cities are burned with fire.
\par }{\QB Strangers devour your land in your presence
\par }{\QB and it is desolate,
\par }{\QB as overthrown by strangers.
\par }{\Q \VS{8}The daughter of Zion is left like a shelter in a vineyard,
\par }{\QB like a hut in a field of melons,
\par }{\QB like a besieged city.
\par }{\Q \VS{9}Unless Yahweh of Armies had left to us a very small remnant,
\par }{\QB we would have been as Sodom.
\par }{\QB We would have been like Gomorrah.
\par }{\BB \par }{\Q \VS{10}Hear Yahweh’s word, you rulers of Sodom!
\par }{\QB Listen to the law of our God,\FTNT{‡}{{\FR 1:10 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).} you people of Gomorrah!
\par }{\Q \VS{11}“What are the multitude of your sacrifices to me?”, says Yahweh.
\par }{\QB “I have had enough of the burnt offerings of rams
\par }{\QB and the fat of fed animals.
\par }{\QB I don’t delight in the blood of bulls,
\par }{\QB or of lambs,
\par }{\QB or of male goats.
\par }{\Q \VS{12}When you come to appear before me,
\par }{\QB who has required this at your hand, to trample my courts?
\par }{\Q \VS{13}Bring no more vain offerings.
\par }{\QB Incense is an abomination to me.
\par }{\QB New moons, Sabbaths, and convocations:
\par }{\QB I can’t stand evil assemblies.
\par }{\Q \VS{14}My soul hates your New Moons and your appointed feasts.
\par }{\QB They are a burden to me.
\par }{\QB I am weary of bearing them.
\par }{\Q \VS{15}When you spread out your hands, I will hide my eyes from you.
\par }{\QB Yes, when you make many prayers, I will not hear.
\par }{\QB Your hands are full of blood.
\par }{\Q \VS{16}Wash yourselves. Make yourself clean.
\par }{\QB Put away the evil of your doings from before my eyes.
\par }{\QB Cease to do evil.
\par }{\Q \VS{17}Learn to do well.
\par }{\QB Seek justice.
\par }{\QB Relieve the oppressed.
\par }{\Q Defend the fatherless.
\par }{\QB Plead for the widow.”
\par }{\BB \par }{\Q \VS{18}“Come now, and let’s reason together,” says Yahweh:
\par }{\QB “Though your sins are as scarlet, they shall be as white as snow.
\par }{\QB Though they are red like crimson, they shall be as wool.
\par }{\Q \VS{19}If you are willing and obedient,
\par }{\QB you will eat the good of the land;
\par }{\QB \VS{20}but if you refuse and rebel, you will be devoured with the sword;
\par }{\QB for the mouth of Yahweh has spoken it.”
\par }{\BB \par }{\Q \VS{21}How the faithful city has become a prostitute!
\par }{\QB She was full of justice.
\par }{\Q Righteousness lodged in her,
\par }{\QB but now there are murderers.
\par }{\Q \VS{22}Your silver has become dross,
\par }{\QB your wine mixed with water.
\par }{\Q \VS{23}Your princes are rebellious and companions of thieves.
\par }{\QB Everyone loves bribes and follows after rewards.
\par }{\QB They don’t defend the fatherless,
\par }{\QB neither does the cause of the widow come to them.
\par }{\BB \par }{\Q \VS{24}Therefore the Lord,\FTNT{§}{{\FR 1:24 }
The word translated “Lord” is “Adonai.”
} Yahweh of Armies,
\par }{\QB the Mighty One of Israel, says:
\par }{\Q “Ah, I will get relief from my adversaries,
\par }{\QB and avenge myself on my enemies.
\par }{\Q \VS{25}I will turn my hand on you,
\par }{\QB thoroughly purge away your dross,
\par }{\QB and will take away all your tin.\FTNT{*}{{\FR 1:25 }
tin is a metal that is separated from silver during the refining and purification process.}
\par }{\Q \VS{26}I will restore your judges as at the first,
\par }{\QB and your counselors as at the beginning.
\par }{\Q Afterward you shall be called ‘The city of righteousness,
\par }{\QB a faithful town.’
\par }{\Q \VS{27}Zion shall be redeemed with justice,
\par }{\QB and her converts with righteousness.
\par }{\Q \VS{28}But the destruction of transgressors and sinners shall be together,
\par }{\QB and those who forsake Yahweh shall be consumed.
\par }{\Q \VS{29}For they shall be ashamed of the oaks which you have desired,
\par }{\QB and you shall be confounded for the gardens that you have chosen.
\par }{\Q \VS{30}For you shall be as an oak whose leaf fades,
\par }{\QB and as a garden that has no water.
\par }{\Q \VS{31}The strong will be like tinder,
\par }{\QB and his work like a spark.
\par }{\Q They will both burn together,
\par }{\QB and no one will quench them.”

\par }\Chap{2}{\PP \VerseOne{1}This is what Isaiah the son of Amoz saw concerning Judah and Jerusalem.
\par }{\Q \VS{2}It shall happen in the latter days, that the mountain of Yahweh’s house shall be established on the top of the mountains,
\par }{\QB and shall be raised above the hills;
\par }{\QB and all nations shall flow to it.
\par }{\Q \VS{3}Many peoples shall go and say,
\par }{\QB “Come, let’s go up to the mountain of Yahweh,
\par }{\QB to the house of the God of Jacob;
\par }{\QB and he will teach us of his ways,
\par }{\QB and we will walk in his paths.”
\par }{\Q For the law shall go out of Zion,
\par }{\QB and Yahweh’s word from Jerusalem.
\par }{\Q \VS{4}He will judge between the nations,
\par }{\QB and will decide concerning many peoples.
\par }{\QB They shall beat their swords into plowshares,
\par }{\QB and their spears into pruning hooks.
\par }{\Q Nation shall not lift up sword against nation,
\par }{\QB neither shall they learn war any more.
\par }{\BB \par }{\Q \VS{5}House of Jacob, come, and let’s walk in the light of Yahweh.
\par }{\Q \VS{6}For you have forsaken your people, the house of Jacob,
\par }{\QB because they are filled from the east,
\par }{\QB with those who practice divination like the Philistines,
\par }{\QB and they clasp hands with the children of foreigners.
\par }{\Q \VS{7}Their land is full of silver and gold,
\par }{\QB neither is there any end of their treasures.
\par }{\Q Their land also is full of horses,
\par }{\QB neither is there any end of their chariots.
\par }{\Q \VS{8}Their land also is full of idols.
\par }{\QB They worship the work of their own hands,
\par }{\QB that which their own fingers have made.
\par }{\Q \VS{9}Man is brought low,
\par }{\QB and mankind is humbled;
\par }{\QB therefore don’t forgive them.
\par }{\Q \VS{10}Enter into the rock,
\par }{\QB and hide in the dust,
\par }{\Q from before the terror of Yahweh,
\par }{\QB and from the glory of his majesty.
\par }{\Q \VS{11}The lofty looks of man will be brought low,
\par }{\QB the arrogance of men will be bowed down,
\par }{\QB and Yahweh alone will be exalted in that day.
\par }{\BB \par }{\Q \VS{12}For there will be a day of Yahweh of Armies for all that is proud and arrogant,
\par }{\QB and for all that is lifted up;
\par }{\QB and it shall be brought low:
\par }{\QB \VS{13}for all the cedars of Lebanon, that are high and lifted up,
\par }{\QB for all the oaks of Bashan,
\par }{\QB \VS{14}for all the high mountains,
\par }{\QB for all the hills that are lifted up,
\par }{\QB \VS{15}for every lofty tower,
\par }{\QB for every fortified wall,
\par }{\QB \VS{16}for all the ships of Tarshish,
\par }{\QB and for all pleasant imagery.
\par }{\Q \VS{17}The loftiness of man shall be bowed down,
\par }{\QB and the arrogance of men shall be brought low;
\par }{\QB and Yahweh alone shall be exalted in that day.
\par }{\Q \VS{18}The idols shall utterly pass away.
\par }{\Q \VS{19}Men shall go into the caves of the rocks,
\par }{\QB and into the holes of the earth,
\par }{\QB from before the terror of Yahweh,
\par }{\QB and from the glory of his majesty,
\par }{\QB when he arises to shake the earth mightily.
\par }{\Q \VS{20}In that day, men shall cast away their idols of silver
\par }{\QB and their idols of gold,
\par }{\QB which have been made for themselves to worship,
\par }{\QB to the moles and to the bats,
\par }{\QB \VS{21}to go into the caverns of the rocks,
\par }{\QB and into the clefts of the ragged rocks,
\par }{\QB from before the terror of Yahweh,
\par }{\QB and from the glory of his majesty,
\par }{\QB when he arises to shake the earth mightily.
\par }{\Q \VS{22}Stop trusting in man, whose breath is in his nostrils;
\par }{\QB for of what account is he?

\par }\PoetryChap{3}{\Q \VerseOne{1}For, behold,\FTNT{*}{{\FR 3:1 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} the Lord, Yahweh of Armies, takes away from Jerusalem and from Judah supply and support,
\par }{\QB the whole supply of bread,
\par }{\QB and the whole supply of water;
\par }{\QB \VS{2}the mighty man,
\par }{\QB the man of war,
\par }{\QB the judge,
\par }{\QB the prophet,
\par }{\QB the diviner,
\par }{\QB the elder,
\par }{\QB \VS{3}the captain of fifty,
\par }{\QB the honorable man,
\par }{\QB the counselor,
\par }{\QB the skilled craftsman,
\par }{\QB and the clever enchanter.
\par }{\Q \VS{4}I will give boys to be their princes,
\par }{\QB and children shall rule over them.
\par }{\Q \VS{5}The people will be oppressed,
\par }{\QB everyone by another,
\par }{\QB and everyone by his neighbor.
\par }{\Q The child will behave himself proudly against the old man,
\par }{\QB and the wicked against the honorable.
\par }{\Q \VS{6}Indeed a man shall take hold of his brother in the house of his father, saying,
\par }{\QB “You have clothing, you be our ruler,
\par }{\QB and let this ruin be under your hand.”
\par }{\Q \VS{7}In that day he will cry out, saying, “I will not be a healer;
\par }{\QB for in my house is neither bread nor clothing.
\par }{\QB You shall not make me ruler of the people.”
\par }{\Q \VS{8}For Jerusalem is ruined, and Judah is fallen;
\par }{\QB because their tongue and their doings are against Yahweh,
\par }{\Q to provoke the eyes of his glory.
\par }{\Q \VS{9}The look of their faces testify against them.
\par }{\QB They parade their sin like Sodom.
\par }{\QB They don’t hide it.
\par }{\QB Woe to their soul!
\par }{\QB For they have brought disaster upon themselves.
\par }{\Q \VS{10}Tell the righteous “Good!”
\par }{\QB For they shall eat the fruit of their deeds.
\par }{\Q \VS{11}Woe to the wicked!
\par }{\QB Disaster is upon them;
\par }{\QB for the deeds of his hands will be paid back to him.
\par }{\Q \VS{12}As for my people, children are their oppressors,
\par }{\QB and women rule over them.
\par }{\QB My people, those who lead you cause you to err,
\par }{\QB and destroy the way of your paths.
\par }{\BB \par }{\Q \VS{13}Yahweh stands up to contend,
\par }{\QB and stands to judge the peoples.
\par }{\Q \VS{14}Yahweh will enter into judgment with the elders of his people
\par }{\QB and their leaders:
\par }{\QB “It is you who have eaten up the vineyard.
\par }{\QB The plunder of the poor is in your houses.
\par }{\QB \VS{15}What do you mean that you crush my people,
\par }{\QB and grind the face of the poor?” says the Lord, Yahweh of Armies.
\par }{\BB \par }{\Q \VS{16}Moreover Yahweh said, “Because the daughters of Zion are arrogant,
\par }{\QB and walk with outstretched necks and flirting eyes,
\par }{\QB walking to trip as they go,
\par }{\QB jingling ornaments on their feet;
\par }{\Q \VS{17}therefore the Lord brings sores on the crown of the head of the women of Zion,
\par }{\QB and Yahweh will make their scalps bald.”
\par }{\PP \VS{18}In that day the Lord will take away the beauty of their anklets, the headbands, the crescent necklaces,
\VS{19}the earrings, the bracelets, the veils,
\VS{20}the headdresses, the ankle chains, the sashes, the perfume containers, the charms,
\VS{21}the signet rings, the nose rings,
\VS{22}the fine robes, the capes, the cloaks, the purses,
\VS{23}the hand mirrors, the fine linen garments, the tiaras, and the shawls.
\par }{\Q \VS{24}It shall happen that instead of sweet spices, there shall be rottenness;
\par }{\QB instead of a belt, a rope;
\par }{\QB instead of well set hair, baldness;
\par }{\QB instead of a robe, a wearing of sackcloth;
\par }{\QB and branding instead of beauty.
\par }{\Q \VS{25}Your men shall fall by the sword,
\par }{\QB and your mighty in the war.
\par }{\Q \VS{26}Her gates shall lament and mourn.
\par }{\QB She shall be desolate and sit on the ground.

\par }\Chap{4}{\PP \VerseOne{1}Seven women shall take hold of one man in that day, saying, “We will eat our own bread, and wear our own clothing. Just let us be called by your name. Take away our reproach.”
\par }{\PP \VS{2}In that day, Yahweh’s branch will be beautiful and glorious, and the fruit of the land will be the beauty and glory of the survivors of Israel.
\VS{3}It will happen that he who is left in Zion and he who remains in Jerusalem shall be called holy, even everyone who is written among the living in Jerusalem,
\VS{4}when the Lord shall have washed away the filth of the daughters of Zion, and shall have purged the blood of Jerusalem from within it, by the spirit of justice and by the spirit of burning.
\VS{5}Yahweh will create over the whole habitation of Mount Zion and over her assemblies, a cloud and smoke by day, and the shining of a flaming fire by night, for over all the glory will be a canopy.
\VS{6}There will be a pavilion for a shade in the daytime from the heat, and for a refuge and for a shelter from storm and from rain.

\par }\PoetryChap{5}{\Q \VerseOne{1}Let me sing for my well beloved a song of my beloved about his vineyard.
\par }{\QB My beloved had a vineyard on a very fruitful hill.
\par }{\Q \VS{2}He dug it up,
\par }{\QB gathered out its stones,
\par }{\QB planted it with the choicest vine,
\par }{\QB built a tower in the middle of it,
\par }{\QB and also cut out a wine press in it.
\par }{\Q He looked for it to yield grapes,
\par }{\QB but it yielded wild grapes.
\par }{\BB \par }{\Q \VS{3}“Now, inhabitants of Jerusalem and men of Judah,
\par }{\QB please judge between me and my vineyard.
\par }{\Q \VS{4}What could have been done more to my vineyard, that I have not done in it?
\par }{\QB Why, when I looked for it to yield grapes, did it yield wild grapes?
\par }{\Q \VS{5}Now I will tell you what I will do to my vineyard.
\par }{\QB I will take away its hedge, and it will be eaten up.
\par }{\QB I will break down its wall, and it will be trampled down.
\par }{\Q \VS{6}I will lay it a wasteland.
\par }{\QB It won’t be pruned or hoed,
\par }{\QB but it will grow briers and thorns.
\par }{\QB I will also command the clouds that they rain no rain on it.”
\par }{\Q \VS{7}For the vineyard of Yahweh of Armies is the house of Israel,
\par }{\QB and the men of Judah his pleasant plant:
\par }{\QB and he looked for justice, but, behold, oppression;
\par }{\QB for righteousness, but, behold, a cry of distress.
\par }{\BB \par }{\Q \VS{8}Woe to those who join house to house,
\par }{\QB who lay field to field, until there is no room,
\par }{\QB and you are made to dwell alone in the middle of the land!
\par }{\Q \VS{9}In my ears, Yahweh of Armies says: “Surely many houses will be desolate,
\par }{\QB even great and beautiful, unoccupied.
\par }{\Q \VS{10}For ten acres\FTNT{*}{{\FR 5:10 }
literally, ten yokes, or the amount of land that ten yokes of oxen can plow in one day, which is about 10 acres or 4 hectares.} of vineyard shall yield one bath,\FTNT{†}{{\FR 5:10 }
1 bath is about 22 liters or 5.8 U. S. gallons}
\par }{\QB and a homer\FTNT{‡}{{\FR 5:10 }
1 homer is about 220 liters or 6 bushels} of seed shall yield an ephah.”\FTNT{§}{{\FR 5:10 }
1 ephah is about 22 liters or 0.6 bushels or about 2 pecks—only one tenth of what was sown.}
\par }{\Q \VS{11}Woe to those who rise up early in the morning, that they may follow strong drink,
\par }{\QB who stay late into the night, until wine inflames them!
\par }{\Q \VS{12}The harp, lyre, tambourine, and flute, with wine, are at their feasts;
\par }{\QB but they don’t respect the work of Yahweh,
\par }{\QB neither have they considered the operation of his hands.
\par }{\BB \par }{\Q \VS{13}Therefore my people go into captivity for lack of knowledge.
\par }{\QB Their honorable men are famished,
\par }{\QB and their multitudes are parched with thirst.
\par }{\Q \VS{14}Therefore Sheol\FTNT{*}{{\FR 5:14 }
Sheol is the place of the dead.} has enlarged its desire,
\par }{\QB and opened its mouth without measure;
\par }{\QB and their glory, their multitude, their pomp, and he who rejoices among them, descend into it.
\par }{\Q \VS{15}So man is brought low,
\par }{\QB mankind is humbled,
\par }{\QB and the eyes of the arrogant ones are humbled;
\par }{\Q \VS{16}but Yahweh of Armies is exalted in justice,
\par }{\QB and God the Holy One is sanctified in righteousness.
\par }{\Q \VS{17}Then the lambs will graze as in their pasture,
\par }{\QB and strangers will eat the ruins of the rich.
\par }{\BB \par }{\Q \VS{18}Woe to those who draw iniquity with cords of falsehood,
\par }{\QB and wickedness as with cart rope,
\par }{\Q \VS{19}who say, “Let him make haste, let him hasten his work, that we may see it;
\par }{\QB let the counsel of the Holy One of Israel draw near and come,
\par }{\QB that we may know it!”
\par }{\Q \VS{20}Woe to those who call evil good, and good evil;
\par }{\QB who put darkness for light,
\par }{\QB and light for darkness;
\par }{\Q who put bitter for sweet,
\par }{\QB and sweet for bitter!
\par }{\Q \VS{21}Woe to those who are wise in their own eyes,
\par }{\QB and prudent in their own sight!
\par }{\Q \VS{22}Woe to those who are mighty to drink wine,
\par }{\QB and champions at mixing strong drink;
\par }{\Q \VS{23}who acquit the guilty for a bribe,
\par }{\QB but deny justice for the innocent!
\par }{\BB \par }{\Q \VS{24}Therefore as the tongue of fire devours the stubble,
\par }{\QB and as the dry grass sinks down in the flame,
\par }{\QB so their root shall be as rottenness,
\par }{\QB and their blossom shall go up as dust,
\par }{\Q because they have rejected the law of Yahweh of Armies,
\par }{\QB and despised the word of the Holy One of Israel.
\par }{\Q \VS{25}Therefore Yahweh’s anger burns against his people,
\par }{\QB and he has stretched out his hand against them and has struck them.
\par }{\Q The mountains tremble,
\par }{\QB and their dead bodies are as refuse in the middle of the streets.
\par }{\Q For all this, his anger is not turned away,
\par }{\QB but his hand is still stretched out.
\par }{\BB \par }{\Q \VS{26}He will lift up a banner to the nations from far away,
\par }{\QB and he will whistle for them from the end of the earth.
\par }{\QB Behold, they will come speedily and swiftly.
\par }{\Q \VS{27}No one shall be weary nor stumble among them;
\par }{\QB no one shall slumber nor sleep,
\par }{\QB neither shall the belt of their waist be untied,
\par }{\QB nor the strap of their sandals be broken,
\par }{\Q \VS{28}whose arrows are sharp,
\par }{\QB and all their bows bent.
\par }{\Q Their horses’ hoofs will be like flint,
\par }{\QB and their wheels like a whirlwind.
\par }{\Q \VS{29}Their roaring will be like a lioness.
\par }{\QB They will roar like young lions.
\par }{\Q Yes, they shall roar,
\par }{\QB and seize their prey and carry it off,
\par }{\QB and there will be no one to deliver.
\par }{\Q \VS{30}They will roar against them in that day like the roaring of the sea.
\par }{\QB If one looks to the land, behold, darkness and distress.
\par }{\QB The light is darkened in its clouds.

\par }\Chap{6}{\PP \VerseOne{1}In the year that king Uzziah died, I saw the Lord sitting on a throne, high and lifted up; and his train filled the temple.
\VS{2}Above him stood the seraphim. Each one had six wings. With two he covered his face. With two he covered his feet. With two he flew.
\VS{3}One called to another, and said,
\par }{\Q “Holy, holy, holy, is Yahweh of Armies!
\par }{\QB The whole earth is full of his glory!”
\par }{\PP \VS{4}The foundations of the thresholds shook at the voice of him who called, and the house was filled with smoke.
\VS{5}Then I said, “Woe is me! For I am undone, because I am a man of unclean lips, and I dwell among a people of unclean lips: for my eyes have seen the King, Yahweh of Armies!”
\par }{\PP \VS{6}Then one of the seraphim flew to me, having a live coal in his hand, which he had taken with the tongs from off the altar.
\VS{7}He touched my mouth with it, and said, “Behold, this has touched your lips; and your iniquity is taken away, and your sin forgiven.”
\par }{\PP \VS{8}I heard the Lord’s voice, saying, “Whom shall I send, and who will go for us?”
\par }{\PP Then I said, “Here I am. Send me!”
\par }{\PP \VS{9}He said, “Go, and tell this people,
\par }{\Q ‘You hear indeed,
\par }{\QB but don’t understand.
\par }{\Q You see indeed,
\par }{\QB but don’t perceive.’
\par }{\Q \VS{10}Make the heart of this people fat.
\par }{\QB Make their ears heavy, and shut their eyes;
\par }{\Q lest they see with their eyes,
\par }{\QB hear with their ears,
\par }{\QB understand with their heart,
\par }{\QB and turn again, and be healed.”
\par }{\PP \VS{11}Then I said, “Lord, how long?”
\par }{\PP He answered,
\par }{\Q “Until cities are waste without inhabitant,
\par }{\QB houses without man,
\par }{\QB the land becomes utterly waste,
\par }{\QB \VS{12}and Yahweh has removed men far away,
\par }{\QB and the forsaken places are many within the land.
\par }{\Q \VS{13}If there is a tenth left in it,
\par }{\QB that also will in turn be consumed,
\par }{\Q as a terebinth, and as an oak, whose stump remains when they are cut down;
\par }{\QB so the holy seed is its stock.”

\par }\Chap{7}{\PP \VerseOne{1}In the days of Ahaz the son of Jotham, the son of Uzziah, king of Judah, Rezin the king of Syria, and Pekah the son of Remaliah, king of Israel, went up to Jerusalem to war against it, but could not prevail against it.
\VS{2}David’s house was told, “Syria is allied with Ephraim.” His heart trembled, and the heart of his people, as the trees of the forest tremble with the wind.
\par }{\PP \VS{3}Then Yahweh said to Isaiah, “Go out now to meet Ahaz, you, and Shearjashub your son, at the end of the conduit of the upper pool, on the highway of the fuller’s field.
\VS{4}Tell him, ‘Be careful, and keep calm. Don’t be afraid, neither let your heart be faint because of these two tails of smoking torches, for the fierce anger of Rezin and Syria, and of the son of Remaliah.
\VS{5}Because Syria, Ephraim, and the son of Remaliah, have plotted evil against you, saying,
\VS{6}“Let’s go up against Judah, and tear it apart, and let’s divide it among ourselves, and set up a king within it, even the son of Tabeel.”
\VS{7}This is what the Lord Yahweh says: “It shall not stand, neither shall it happen.”
\VS{8}For the head of Syria is Damascus, and the head of Damascus is Rezin. Within sixty-five years Ephraim shall be broken in pieces, so that it shall not be a people.
\VS{9}The head of Ephraim is Samaria, and the head of Samaria is Remaliah’s son. If you will not believe, surely you shall not be established.’ ”
\par }{\PP \VS{10}Yahweh spoke again to Ahaz, saying,
\VS{11}“Ask a sign of Yahweh your God; ask it either in the depth, or in the height above.”
\par }{\PP \VS{12}But Ahaz said, “I won’t ask. I won’t tempt Yahweh.”
\par }{\PP \VS{13}He said, “Listen now, house of David. Is it not enough for you to try the patience of men, that you will try the patience of my God also?
\VS{14}Therefore the Lord himself will give you a sign. Behold, the virgin will conceive, and bear a son, and shall call his name Immanuel.\FTNT{*}{{\FR 7:14 }
“Immanuel” means “God with us”.}
\VS{15}He shall eat butter and honey when he knows to refuse the evil and choose the good.
\VS{16}For before the child knows to refuse the evil and choose the good, the land whose two kings you abhor shall be forsaken.
\VS{17}Yahweh will bring on you, on your people, and on your father’s house, days that have not come, from the day that Ephraim departed from Judah, even the king of Assyria.
\par }{\PP \VS{18}It will happen in that day that Yahweh will whistle for the fly that is in the uttermost part of the rivers of Egypt, and for the bee that is in the land of Assyria.
\VS{19}They shall come, and shall all rest in the desolate valleys, in the clefts of the rocks, on all thorn hedges, and on all pastures.
\par }{\PP \VS{20}In that day the Lord will shave with a razor that is hired in the parts beyond the River, even with the king of Assyria, the head and the hair of the feet; and it shall also consume the beard.
\par }{\PP \VS{21}It shall happen in that day that a man shall keep alive a young cow, and two sheep.
\VS{22}It shall happen, that because of the abundance of milk which they shall give he shall eat butter; for everyone will eat butter and honey that is left within the land.
\par }{\PP \VS{23}It will happen in that day that every place where there were a thousand vines at a thousand silver shekels,\FTNT{†}{{\FR 7:23 }
A shekel is about 10 grams or about 0.35 ounces, so 1000 shekels is about 10 kilograms or 22 pounds.} shall be for briers and thorns.
\VS{24}People will go there with arrows and with bow, because all the land will be briers and thorns.
\VS{25}All the hills that were cultivated with the hoe, you shall not come there for fear of briers and thorns; but it shall be for the sending out of oxen, and for sheep to tread on.”

\par }\Chap{8}{\PP \VerseOne{1}Yahweh said to me, “Take a large tablet, and write on it with a man’s pen, ‘For Maher Shalal Hash Baz’;\FTNT{*}{{\FR 8:1 }
“Maher Shalal Hash Baz” means “quick to the plunder, swift to the prey”.}
\VS{2}and I will take for myself faithful witnesses to testify: Uriah the priest, and Zechariah the son of Jeberechiah.”
\par }{\PP \VS{3}I went to the prophetess, and she conceived, and bore a son. Then Yahweh said to me, “Call his name ‘Maher Shalal Hash Baz.’
\VS{4}For before the child knows how to say, ‘My father,’ and, ‘My mother,’ the riches of Damascus and the plunder of Samaria will be carried away by the king of Assyria.”
\par }{\PP \VS{5}Yahweh spoke to me yet again, saying,
\VS{6}“Because this people has refused the waters of Shiloah that go softly, and rejoice in Rezin and Remaliah’s son;
\VS{7}now therefore, behold, the Lord brings upon them the mighty flood waters of the River: the king of Assyria and all his glory. It will come up over all its channels, and go over all its banks.
\VS{8}It will sweep onward into Judah. It will overflow and pass through. It will reach even to the neck. The stretching out of its wings will fill the width of your land, Immanuel.
\VS{9}Make an uproar, you peoples, and be broken in pieces! Listen, all you from far countries: dress for battle, and be shattered! Dress for battle, and be shattered!
\VS{10}Take counsel together, and it will be brought to nothing; speak the word, and it will not stand, for God is with us.”
\VS{11}For Yahweh spoke this to me with a strong hand, and instructed me not to walk in the way of this people, saying,
\VS{12}“Don’t say, ‘A conspiracy!’ concerning all about which this people say, ‘A conspiracy!’ neither fear their threats, nor be terrorized.
\VS{13}Yahweh of Armies is who you must respect as holy. He is the one you must fear. He is the one you must dread.
\VS{14}He will be a sanctuary, but for both houses of Israel, he will be a stumbling stone and a rock that makes them fall. For the people of Jerusalem, he will be a trap and a snare.
\VS{15}Many will stumble over it, fall, be broken, be snared, and be captured.”
\par }{\PP \VS{16}Wrap up the covenant. Seal the law among my disciples.
\VS{17}I will wait for Yahweh, who hides his face from the house of Jacob, and I will look for him.
\VS{18}Behold, I and the children whom Yahweh has given me are for signs and for wonders in Israel from Yahweh of Armies, who dwells in Mount Zion.
\par }{\PP \VS{19}When they tell you, “Consult with those who have familiar spirits and with the wizards, who chirp and who mutter,” shouldn’t a people consult with their God? Should they consult the dead on behalf of the living?
\VS{20}Turn to the law and to the covenant! If they don’t speak according to this word, surely there is no morning for them.
\VS{21}They will pass through it, very distressed and hungry. It will happen that when they are hungry, they will worry, and curse by their king and by their God. They will turn their faces upward,
\VS{22}and look to the earth, and see distress, darkness, and the gloom of anguish. They will be driven into thick darkness.

\par }\Chap{9}{\PP \VerseOne{1}But there shall be no more gloom for her who was in anguish. In the former time, he brought into contempt the land of Zebulun and the land of Naphtali; but in the latter time he has made it glorious, by the way of the sea, beyond the Jordan, Galilee of the nations.
\par }{\PP \VS{2}The people who walked in darkness have seen a great light.
\par }{\QB The light has shined on those who lived in the land of the shadow of death.
\par }{\Q \VS{3}You have multiplied the nation.
\par }{\QB You have increased their joy.
\par }{\PP They rejoice before you according to the joy in harvest, as men rejoice when they divide the plunder.
\VS{4}For the yoke of his burden, and the staff of his shoulder, the rod of his oppressor, you have broken as in the day of Midian.
\VS{5}For all the armor of the armed man in the noisy battle, and the garments rolled in blood, will be for burning, fuel for the fire.
\VS{6}For a child is born to us. A son is given to us; and the government will be on his shoulders. His name will be called Wonderful Counselor, Mighty God, Everlasting Father, Prince of Peace.
\VS{7}Of the increase of his government and of peace there shall be no end, on David’s throne, and on his kingdom, to establish it, and to uphold it with justice and with righteousness from that time on, even forever. The zeal of Yahweh of Armies will perform this.
\par }{\PP \VS{8}The Lord sent a word into Jacob,
\par }{\QB and it falls on Israel.
\par }{\Q \VS{9}All the people will know,
\par }{\QB including Ephraim and the inhabitants of Samaria, who say in pride and in arrogance of heart,
\par }{\Q \VS{10}“The bricks have fallen,
\par }{\QB but we will build with cut stone.
\par }{\Q The sycamore fig trees have been cut down,
\par }{\QB but we will put cedars in their place.”
\par }{\Q \VS{11}Therefore Yahweh will set up on high against him the adversaries of Rezin,
\par }{\QB and will stir up his enemies,
\par }{\QB \VS{12}The Syrians in front,
\par }{\QB and the Philistines behind;
\par }{\QB and they will devour Israel with open mouth.
\par }{\Q For all this, his anger is not turned away,
\par }{\QB but his hand is stretched out still.
\par }{\BB \par }{\Q \VS{13}Yet the people have not turned to him who struck them,
\par }{\QB neither have they sought Yahweh of Armies.
\par }{\Q \VS{14}Therefore Yahweh will cut off from Israel head and tail,
\par }{\QB palm branch and reed, in one day.
\par }{\Q \VS{15}The elder and the honorable man is the head,
\par }{\QB and the prophet who teaches lies is the tail.
\par }{\Q \VS{16}For those who lead this people lead them astray;
\par }{\QB and those who are led by them are destroyed.
\par }{\Q \VS{17}Therefore the Lord will not rejoice over their young men,
\par }{\QB neither will he have compassion on their fatherless and widows;
\par }{\Q for everyone is profane and an evildoer,
\par }{\QB and every mouth speaks folly.
\par }{\Q For all this his anger is not turned away,
\par }{\QB but his hand is stretched out still.
\par }{\BB \par }{\Q \VS{18}For wickedness burns like a fire.
\par }{\QB It devours the briers and thorns;
\par }{\QB yes, it kindles in the thickets of the forest,
\par }{\QB and they roll upward in a column of smoke.
\par }{\Q \VS{19}Through Yahweh of Armies’ wrath, the land is burned up;
\par }{\QB and the people are the fuel for the fire.
\par }{\QB No one spares his brother.
\par }{\Q \VS{20}One will devour on the right hand, and be hungry;
\par }{\QB and he will eat on the left hand, and they will not be satisfied.
\par }{\Q Everyone will eat the flesh of his own arm:
\par }{\QB \VS{21}Manasseh, Ephraim; and Ephraim, Manasseh; and they together shall be against Judah.
\par }{\Q For all this his anger is not turned away,
\par }{\QB but his hand is stretched out still.

\par }\Chap{10}{\PP \VerseOne{1}Woe to those who decree unrighteous decrees, and to the writers who write oppressive decrees;
\VS{2}to deprive the needy of justice, and to rob the poor among my people of their rights, that widows may be their plunder, and that they may make the fatherless their prey!
\VS{3}What will you do in the day of visitation, and in the desolation which will come from afar? To whom will you flee for help? Where will you leave your wealth?
\par }{\PP \VS{4}They will only bow down under the prisoners,
\par }{\QB and will fall under the slain.
\par }{\Q For all this his anger is not turned away,
\par }{\QB but his hand is stretched out still.
\par }{\BB \par }{\PP \VS{5}Alas Assyrian, the rod of my anger, the staff in whose hand is my indignation!
\VS{6}I will send him against a profane nation, and against the people who anger me I will give him a command to take the plunder and to take the prey, and to tread them down like the mire of the streets.
\VS{7}However he doesn’t mean so, neither does his heart think so; but it is in his heart to destroy, and to cut off not a few nations.
\VS{8}For he says, “Aren’t all of my princes kings?
\VS{9}Isn’t Calno like Carchemish? Isn’t Hamath like Arpad? Isn’t Samaria like Damascus?”
\VS{10}As my hand has found the kingdoms of the idols, whose engraved images exceeded those of Jerusalem and of Samaria;
\VS{11}shall I not, as I have done to Samaria and her idols, so do to Jerusalem and her idols?
\par }{\PP \VS{12}Therefore it will happen that when the Lord has performed his whole work on Mount Zion and on Jerusalem, I will punish the fruit of the willful proud heart of the king of Assyria, and the insolence of his arrogant looks.
\VS{13}For he has said, “By the strength of my hand I have done it, and by my wisdom; for I have understanding. I have removed the boundaries of the peoples, and have robbed their treasures. Like a valiant man I have brought down their rulers.
\VS{14}My hand has found the riches of the peoples like a nest, and like one gathers eggs that are abandoned, I have gathered all the earth. There was no one who moved their wing, or that opened their mouth, or chirped.”
\par }{\PP \VS{15}Should an ax brag against him who chops with it? Should a saw exalt itself above him who saws with it? As if a rod should lift those who lift it up, or as if a staff should lift up someone who is not wood.
\VS{16}Therefore the Lord, Yahweh of Armies, will send among his fat ones leanness; and under his glory a burning will be kindled like the burning of fire.
\VS{17}The light of Israel will be for a fire, and his Holy One for a flame; and it will burn and devour his thorns and his briers in one day.
\VS{18}He will consume the glory of his forest, and of his fruitful field, both soul and body. It will be as when a standard bearer faints.
\VS{19}The remnant of the trees of his forest shall be few, so that a child could write their number.
\par }{\PP \VS{20}It will come to pass in that day that the remnant of Israel, and those who have escaped from the house of Jacob will no more again lean on him who struck them, but shall lean on Yahweh, the Holy One of Israel, in truth.
\VS{21}A remnant will return, even the remnant of Jacob, to the mighty God.
\VS{22}For though your people, Israel, are like the sand of the sea, only a remnant of them will return. A destruction is determined, overflowing with righteousness.
\VS{23}For the Lord, Yahweh of Armies, will make a full end, and that determined, throughout all the earth.
\par }{\PP \VS{24}Therefore the Lord, Yahweh of Armies, says “My people who dwell in Zion, don’t be afraid of the Assyrian, though he strike you with the rod, and lift up his staff against you, as Egypt did.
\VS{25}For yet a very little while, and the indignation against you will be accomplished, and my anger will be directed to his destruction.”
\VS{26}Yahweh of Armies will stir up a scourge against him, as in the slaughter of Midian at the rock of Oreb. His rod will be over the sea, and he will lift it up like he did against Egypt.
\VS{27}It will happen in that day that his burden will depart from off your shoulder, and his yoke from off your neck, and the yoke shall be destroyed because of the anointing oil.
\par }{\PP \VS{28}He has come to Aiath. He has passed through Migron. At Michmash he stores his baggage.
\VS{29}They have gone over the pass. They have taken up their lodging at Geba. Ramah trembles. Gibeah of Saul has fled.
\VS{30}Cry aloud with your voice, daughter of Gallim! Listen, Laishah! You poor Anathoth!
\VS{31}Madmenah is a fugitive. The inhabitants of Gebim flee for safety.
\VS{32}This very day he will halt at Nob. He shakes his hand at the mountain of the daughter of Zion, the hill of Jerusalem.
\par }{\PP \VS{33}Behold, the Lord, Yahweh of Armies, will lop the boughs with terror. The tall will be cut down, and the lofty will be brought low.
\VS{34}He will cut down the thickets of the forest with iron, and Lebanon will fall by the Mighty One.

\par }\PoetryChap{11}{\Q \VerseOne{1}A shoot will come out of the stock of Jesse,
\par }{\QB and a branch out of his roots will bear fruit.
\par }{\Q \VS{2}Yahweh’s Spirit will rest on him:
\par }{\QB the spirit of wisdom and understanding,
\par }{\QB the spirit of counsel and might,
\par }{\QB the spirit of knowledge and of the fear of Yahweh.
\par }{\Q \VS{3}His delight will be in the fear of Yahweh.
\par }{\Q He will not judge by the sight of his eyes,
\par }{\QB neither decide by the hearing of his ears;
\par }{\Q \VS{4}but he will judge the poor with righteousness,
\par }{\QB and decide with equity for the humble of the earth.
\par }{\Q He will strike the earth with the rod of his mouth;
\par }{\QB and with the breath of his lips he will kill the wicked.
\par }{\Q \VS{5}Righteousness will be the belt of his waist,
\par }{\QB and faithfulness the belt of his waist.
\par }{\BB \par }{\Q \VS{6}The wolf will live with the lamb,
\par }{\QB and the leopard will lie down with the young goat,
\par }{\QB the calf, the young lion, and the fattened calf together;
\par }{\QB and a little child will lead them.
\par }{\Q \VS{7}The cow and the bear will graze.
\par }{\QB Their young ones will lie down together.
\par }{\QB The lion will eat straw like the ox.
\par }{\Q \VS{8}The nursing child will play near a cobra’s hole,
\par }{\QB and the weaned child will put his hand on the viper’s den.
\par }{\Q \VS{9}They will not hurt nor destroy in all my holy mountain;
\par }{\QB for the earth will be full of the knowledge of Yahweh,
\par }{\QB as the waters cover the sea.
\par }{\BB \par }{\PP \VS{10}It will happen in that day that the nations will seek the root of Jesse, who stands as a banner of the peoples; and his resting place will be glorious.
\par }{\PP \VS{11}It will happen in that day that the Lord will set his hand again the second time to recover the remnant that is left of his people from Assyria, from Egypt, from Pathros, from Cush, from Elam, from Shinar, from Hamath, and from the islands of the sea.
\VS{12}He will set up a banner for the nations, and will assemble the outcasts of Israel, and gather together the dispersed of Judah from the four corners of the earth.
\VS{13}The envy also of Ephraim will depart, and those who persecute Judah will be cut off. Ephraim won’t envy Judah, and Judah won’t persecute Ephraim.
\VS{14}They will fly down on the shoulders of the Philistines on the west. Together they will plunder the children of the east. They will extend their power over Edom and Moab, and the children of Ammon will obey them.
\VS{15}Yahweh will utterly destroy the tongue of the Egyptian sea; and with his scorching wind he will wave his hand over the River, and will split it into seven streams, and cause men to march over in sandals.
\VS{16}There will be a highway for the remnant that is left of his people from Assyria, like there was for Israel in the day that he came up out of the land of Egypt.

\par }\Chap{12}{\PP \VerseOne{1}In that day you will say, “I will give thanks to you, Yahweh; for though you were angry with me, your anger has turned away and you comfort me.
\VS{2}Behold, God is my salvation. I will trust, and will not be afraid; for Yah, Yahweh, is my strength and song; and he has become my salvation.”
\VS{3}Therefore with joy you will draw water out of the wells of salvation.
\VS{4}In that day you will say, “Give thanks to Yahweh! Call on his name. Declare his doings among the peoples. Proclaim that his name is exalted!
\VS{5}Sing to Yahweh, for he has done excellent things! Let this be known in all the earth!
\VS{6}Cry aloud and shout, you inhabitant of Zion; for the Holy One of Israel is great among you!”

\par }\Chap{13}{\PP \VerseOne{1}The burden of Babylon, which Isaiah the son of Amoz saw.
\par }{\PP \VS{2}Set up a banner on the bare mountain! Lift up your voice to them! Wave your hand, that they may go into the gates of the nobles.
\VS{3}I have commanded my consecrated ones; yes, I have called my mighty men for my anger, even my proudly exulting ones.
\VS{4}The noise of a multitude is in the mountains, as of a great people; the noise of an uproar of the kingdoms of the nations gathered together! Yahweh of Armies is mustering the army for the battle.
\VS{5}They come from a far country, from the uttermost part of heaven, even Yahweh, and the weapons of his indignation, to destroy the whole land.
\par }{\PP \VS{6}Wail, for Yahweh’s day is at hand! It will come as destruction from the Almighty.
\VS{7}Therefore all hands will be feeble, and everyone’s heart will melt.
\VS{8}They will be dismayed. Pangs and sorrows will seize them. They will be in pain like a woman in labor. They will look in amazement one at another. Their faces will be faces of flame.
\VS{9}Behold, the day of Yahweh comes, cruel, with wrath and fierce anger; to make the land a desolation, and to destroy its sinners out of it.
\VS{10}For the stars of the sky and its constellations will not give their light. The sun will be darkened in its going out, and the moon will not cause its light to shine.
\VS{11}I will punish the world for their evil, and the wicked for their iniquity. I will cause the arrogance of the proud to cease, and will humble the arrogance of the terrible.
\VS{12}I will make people more rare than fine gold, even a person than the pure gold of Ophir.
\VS{13}Therefore I will make the heavens tremble, and the earth will be shaken out of its place in Yahweh of Armies’ wrath, and in the day of his fierce anger.
\VS{14}It will happen that like a hunted gazelle, and like sheep that no one gathers, they will each turn to their own people, and will each flee to their own land.
\VS{15}Everyone who is found will be thrust through. Everyone who is captured will fall by the sword.
\VS{16}Their infants also will be dashed in pieces before their eyes. Their houses will be ransacked, and their wives raped.
\par }{\PP \VS{17}Behold, I will stir up the Medes against them, who will not value silver, and as for gold, they will not delight in it.
\VS{18}Their bows will dash the young men in pieces; and they shall have no pity on the fruit of the womb. Their eyes will not spare children.
\VS{19}Babylon, the glory of kingdoms, the beauty of the Chaldeans’ pride, will be like when God overthrew Sodom and Gomorrah.
\VS{20}It will never be inhabited, neither will it be lived in from generation to generation. The Arabian will not pitch a tent there, neither will shepherds make their flocks lie down there.
\VS{21}But wild animals of the desert will lie there, and their houses will be full of jackals. Ostriches will dwell there, and wild goats will frolic there.
\VS{22}Wolves will cry in their fortresses, and jackals in the pleasant palaces. Her time is near to come, and her days will not be prolonged.

\par }\Chap{14}{\PP \VerseOne{1}For Yahweh will have compassion on Jacob, and will yet choose Israel, and set them in their own land. The foreigner will join himself with them, and they will unite with the house of Jacob.
\VS{2}The peoples will take them, and bring them to their place. The house of Israel will possess them in Yahweh’s land for servants and for handmaids. They will take as captives those whose captives they were; and they shall rule over their oppressors.
\par }{\PP \VS{3}It will happen in the day that Yahweh will give you rest from your sorrow, from your trouble, and from the hard service in which you were made to serve,
\VS{4}that you will take up this parable against the king of Babylon, and say, “How the oppressor has ceased! The golden city has ceased!”
\VS{5}Yahweh has broken the staff of the wicked, the scepter of the rulers,
\VS{6}who struck the peoples in wrath with a continual stroke, who ruled the nations in anger, with a persecution that no one restrained.
\VS{7}The whole earth is at rest, and is quiet. They break out in song.
\VS{8}Yes, the cypress trees rejoice with you, with the cedars of Lebanon, saying, “Since you are humbled, no lumberjack has come up against us.”
\VS{9}Sheol\FTNT{*}{{\FR 14:9 }
Sheol is the place of the dead.} from beneath has moved for you to meet you at your coming. It stirs up the departed spirits for you, even all the rulers of the earth. It has raised up from their thrones all the kings of the nations.
\VS{10}They all will answer and ask you, “Have you also become as weak as we are? Have you become like us?”
\VS{11}Your pomp is brought down to Sheol,\FTNT{†}{{\FR 14:11 }
Sheol is the place of the dead.
} with the sound of your stringed instruments. Maggots are spread out under you, and worms cover you.
\par }{\PP \VS{12}How you have fallen from heaven, shining one, son of the dawn! How you are cut down to the ground, who laid the nations low!
\VS{13}You said in your heart, “I will ascend into heaven! I will exalt my throne above the stars of God! I will sit on the mountain of assembly, in the far north!
\VS{14}I will ascend above the heights of the clouds! I will make myself like the Most High!”
\VS{15}Yet you shall be brought down to Sheol,\FTNT{‡}{{\FR 14:15 }
Sheol is the place of the dead.} to the depths of the pit.
\VS{16}Those who see you will stare at you. They will ponder you, saying, “Is this the man who made the earth to tremble, who shook kingdoms,
\VS{17}who made the world like a wilderness, and overthrew its cities, who didn’t release his prisoners to their home?”
\par }{\PP \VS{18}All the kings of the nations, sleep in glory, everyone in his own house.
\VS{19}But you are cast away from your tomb like an abominable branch, clothed with the slain, who are thrust through with the sword, who go down to the stones of the pit; like a dead body trodden under foot.
\VS{20}You will not join them in burial, because you have destroyed your land. You have killed your people. The offspring of evildoers will not be named forever.
\par }{\PP \VS{21}Prepare for slaughter of his children because of the iniquity of their fathers, that they not rise up and possess the earth, and fill the surface of the world with cities.
\VS{22}“I will rise up against them,” says Yahweh of Armies, “and cut off from Babylon name and remnant, and son and son’s son,” says Yahweh.
\VS{23}“I will also make it a possession for the porcupine, and pools of water. I will sweep it with the broom of destruction,” says Yahweh of Armies.
\par }{\PP \VS{24}Yahweh of Armies has sworn, saying, “Surely, as I have thought, so shall it happen; and as I have purposed, so shall it stand:
\VS{25}that I will break the Assyrian in my land, and tread him under foot on my mountains. Then his yoke will leave them, and his burden leave their shoulders.
\VS{26}This is the plan that is determined for the whole earth. This is the hand that is stretched out over all the nations.
\VS{27}For Yahweh of Armies has planned, and who can stop it? His hand is stretched out, and who can turn it back?”
\par }{\PP \VS{28}This burden was in the year that king Ahaz died.
\par }{\PP \VS{29}Don’t rejoice, O Philistia, all of you, because the rod that struck you is broken; for out of the serpent’s root an adder will emerge, and his fruit will be a fiery flying serpent.
\VS{30}The firstborn of the poor will eat, and the needy will lie down in safety; and I will kill your root with famine, and your remnant will be killed.
\par }{\PP \VS{31}Howl, gate! Cry, city! You are melted away, Philistia, all of you; for smoke comes out of the north, and there is no straggler in his ranks.
\par }{\PP \VS{32}What will they answer the messengers of the nation? That Yahweh has founded Zion, and in her the afflicted of his people will take refuge.

\par }\Chap{15}{\PP \VerseOne{1}The burden of Moab.
\par }{\PP For in a night, Ar of Moab is laid waste, and brought to nothing. For in a night Kir of Moab is laid waste, and brought to nothing.
\VS{2}They have gone up to Bayith, and to Dibon, to the high places, to weep. Moab wails over Nebo and over Medeba. Baldness is on all of their heads. Every beard is cut off.
\VS{3}In their streets, they clothe themselves in sackcloth. In their streets and on their housetops, everyone wails, weeping abundantly.
\VS{4}Heshbon cries out with Elealeh. Their voice is heard even to Jahaz. Therefore the armed men of Moab cry aloud. Their souls tremble within them.
\VS{5}My heart cries out for Moab! Her nobles flee to Zoar, to Eglath Shelishiyah; for they go up by the ascent of Luhith with weeping; for on the way to Horonaim, they raise up a cry of destruction.
\VS{6}For the waters of Nimrim will be desolate; for the grass has withered away, the tender grass fails, there is no green thing.
\VS{7}Therefore they will carry away the abundance they have gotten, and that which they have stored up, over the brook of the willows.
\VS{8}For the cry has gone around the borders of Moab, its wailing to Eglaim, and its wailing to Beer Elim.
\VS{9}For the waters of Dimon are full of blood; for I will bring yet more on Dimon, a lion on those of Moab who escape, and on the remnant of the land.

\par }\Chap{16}{\PP \VerseOne{1}Send the lambs for the ruler of the land from Selah to the wilderness, to the mountain of the daughter of Zion.
\VS{2}For it will be that as wandering birds, as a scattered nest, so will the daughters of Moab be at the fords of the Arnon.
\VS{3}Give counsel! Execute justice! Make your shade like the night in the middle of the noonday! Hide the outcasts! Don’t betray the fugitive!
\VS{4}Let my outcasts dwell with you! As for Moab, be a hiding place for him from the face of the destroyer. For the extortionist is brought to nothing. Destruction ceases. The oppressors are consumed out of the land.
\VS{5}A throne will be established in loving kindness. One will sit on it in truth, in the tent of David, judging, seeking justice, and swift to do righteousness.
\par }{\PP \VS{6}We have heard of the pride of Moab, that he is very proud; even of his arrogance, his pride, and his wrath. His boastings are nothing.
\VS{7}Therefore Moab will wail for Moab. Everyone will wail. You will mourn for the raisin cakes of Kir Hareseth, utterly stricken.
\VS{8}For the fields of Heshbon languish with the vine of Sibmah. The lords of the nations have broken down its choice branches, which reached even to Jazer, which wandered into the wilderness. Its shoots were spread abroad. They passed over the sea.
\VS{9}Therefore I will weep with the weeping of Jazer for the vine of Sibmah. I will water you with my tears, Heshbon, and Elealeh: for on your summer fruits and on your harvest the battle shout has fallen.
\VS{10}Gladness is taken away, and joy out of the fruitful field; and in the vineyards there will be no singing, neither joyful noise. Nobody will tread out wine in the presses. I have made the shouting stop.
\VS{11}Therefore my heart sounds like a harp for Moab, and my inward parts for Kir Heres.
\VS{12}It will happen that when Moab presents himself, when he wearies himself on the high place, and comes to his sanctuary to pray, that he will not prevail.
\par }{\PP \VS{13}This is the word that Yahweh spoke concerning Moab in time past.
\VS{14}But now Yahweh has spoken, saying, “Within three years, as a worker bound by contract would count them, the glory of Moab shall be brought into contempt, with all his great multitude; and the remnant will be very small and feeble.”

\par }\Chap{17}{\PP \VerseOne{1}The burden of Damascus.
\par }{\PP “Behold, Damascus is taken away from being a city, and it will be a ruinous heap.
\VS{2}The cities of Aroer are forsaken. They will be for flocks, which shall lie down, and no one shall make them afraid.
\VS{3}The fortress shall cease from Ephraim, and the kingdom from Damascus, and the remnant of Syria. They will be as the glory of the children of Israel,” says Yahweh of Armies.
\par }{\PP \VS{4}“It will happen in that day that the glory of Jacob will be made thin, and the fatness of his flesh will become lean.
\VS{5}It will be like when the harvester gathers the wheat, and his arm reaps the grain. Yes, it will be like when one gleans grain in the valley of Rephaim.
\VS{6}Yet gleanings will be left there, like the shaking of an olive tree, two or three olives in the top of the uppermost bough, four or five in the outermost branches of a fruitful tree,” says Yahweh, the God of Israel.
\VS{7}In that day, people will look to their Maker, and their eyes will have respect for the Holy One of Israel.
\VS{8}They will not look to the altars, the work of their hands; neither shall they respect that which their fingers have made, either the Asherah poles, or the incense altars.
\VS{9}In that day, their strong cities will be like the forsaken places in the woods and on the mountain top, which were forsaken from before the children of Israel; and it will be a desolation.
\VS{10}For you have forgotten the God of your salvation, and have not remembered the rock of your strength. Therefore you plant pleasant plants, and set out foreign seedlings.
\VS{11}In the day of your planting, you hedge it in. In the morning, you make your seed blossom, but the harvest flees away in the day of grief and of desperate sorrow.
\par }{\PP \VS{12}Ah, the uproar of many peoples, who roar like the roaring of the seas; and the rushing of nations, that rush like the rushing of mighty waters!
\VS{13}The nations will rush like the rushing of many waters: but he will rebuke them, and they will flee far off, and will be chased like the chaff of the mountains before the wind, and like the whirling dust before the storm.
\VS{14}At evening, behold, terror! Before the morning, they are no more. This is the portion of those who plunder us, and the lot of those who rob us.

\par }\Chap{18}{\PP \VerseOne{1}Ah, the land of the rustling of wings, which is beyond the rivers of Ethiopia;
\VS{2}that sends ambassadors by the sea, even in vessels of papyrus on the waters, saying, “Go, you swift messengers, to a nation tall and smooth, to a people awesome from their beginning onward, a nation that measures out and treads down, whose land the rivers divide!”
\VS{3}All you inhabitants of the world, and you dwellers on the earth, when a banner is lifted up on the mountains, look! When the trumpet is blown, listen!
\par }{\PP \VS{4}For Yahweh said to me, “I will be still, and I will see in my dwelling place, like clear heat in sunshine, like a cloud of dew in the heat of harvest.”
\VS{5}For before the harvest, when the blossom is over, and the flower becomes a ripening grape, he will cut off the sprigs with pruning hooks, and he will cut down and take away the spreading branches.
\VS{6}They will be left together for the ravenous birds of the mountains, and for the animals of the earth. The ravenous birds will eat them in the summer, and all the animals of the earth will eat them in the winter.
\VS{7}In that time, a present will be brought to Yahweh of Armies from a people tall and smooth, even from a people awesome from their beginning onward, a nation that measures out and treads down, whose land the rivers divide, to the place of the name of Yahweh of Armies, Mount Zion.

\par }\Chap{19}{\PP \VerseOne{1}The burden of Egypt.
\par }{\PP “Behold, Yahweh rides on a swift cloud, and comes to Egypt. The idols of Egypt will tremble at his presence; and the heart of Egypt will melt within it.
\VS{2}I will stir up the Egyptians against the Egyptians, and they will fight everyone against his brother, and everyone against his neighbor; city against city, and kingdom against kingdom.
\VS{3}The spirit of Egypt will fail within it. I will destroy its counsel. They will seek the idols, the charmers, those who have familiar spirits, and the wizards.
\VS{4}I will give over the Egyptians into the hand of a cruel lord. A fierce king will rule over them,” says the Lord, Yahweh of Armies.
\par }{\PP \VS{5}The waters will fail from the sea, and the river will be wasted and become dry.
\VS{6}The rivers will become foul. The streams of Egypt will be diminished and dried up. The reeds and flags will wither away.
\VS{7}The meadows by the Nile, by the brink of the Nile, and all the sown fields of the Nile, will become dry, be driven away, and be no more.
\VS{8}The fishermen will lament, and all those who fish in the Nile will mourn, and those who spread nets on the waters will languish.
\VS{9}Moreover those who work in combed flax, and those who weave white cloth, will be confounded.
\VS{10}The pillars will be broken in pieces. All those who work for hire will be grieved in soul.
\par }{\PP \VS{11}The princes of Zoan are utterly foolish. The counsel of the wisest counselors of Pharaoh has become stupid. How do you say to Pharaoh, “I am the son of the wise, the son of ancient kings?”
\VS{12}Where then are your wise men? Let them tell you now; and let them know what Yahweh of Armies has purposed concerning Egypt.
\VS{13}The princes of Zoan have become fools. The princes of Memphis are deceived. They have caused Egypt to go astray, who are the cornerstone of her tribes.
\VS{14}Yahweh has mixed a spirit of perverseness in the middle of her; and they have caused Egypt to go astray in all of its works, like a drunken man staggers in his vomit.
\VS{15}Neither shall there be any work for Egypt, which head or tail, palm branch or rush, may do.
\VS{16}In that day the Egyptians will be like women. They will tremble and fear because of the shaking of Yahweh of Armies’s hand, which he shakes over them.
\VS{17}The land of Judah will become a terror to Egypt. Everyone to whom mention is made of it will be afraid, because of the plans of Yahweh of Armies, which he determines against it.
\VS{18}In that day, there will be five cities in the land of Egypt that speak the language of Canaan, and swear to Yahweh of Armies. One will be called “The city of destruction.”
\par }{\PP \VS{19}In that day, there will be an altar to Yahweh in the middle of the land of Egypt, and a pillar to Yahweh at its border.
\VS{20}It will be for a sign and for a witness to Yahweh of Armies in the land of Egypt; for they will cry to Yahweh because of oppressors, and he will send them a savior and a defender, and he will deliver them.
\VS{21}Yahweh will be known to Egypt, and the Egyptians will know Yahweh in that day. Yes, they will worship with sacrifice and offering, and will vow a vow to Yahweh, and will perform it.
\VS{22}Yahweh will strike Egypt, striking and healing. They will return to Yahweh, and he will be entreated by them, and will heal them.
\par }{\PP \VS{23}In that day there will be a highway out of Egypt to Assyria, and the Assyrian shall come into Egypt, and the Egyptian into Assyria; and the Egyptians will worship with the Assyrians.
\par }{\PP \VS{24}In that day, Israel will be the third with Egypt and with Assyria, a blessing within the earth;
\VS{25}because Yahweh of Armies has blessed them, saying, “Blessed be Egypt my people, Assyria the work of my hands, and Israel my inheritance.”

\par }\Chap{20}{\PP \VerseOne{1}In the year that Tartan came to Ashdod, when Sargon the king of Assyria sent him, and he fought against Ashdod and took it;
\VS{2}at that time Yahweh spoke by Isaiah the son of Amoz, saying, “Go, and loosen the sackcloth from off your waist, and take your sandals from off your feet.” He did so, walking naked and barefoot.
\VS{3}Yahweh said, “As my servant Isaiah has walked naked and barefoot three years for a sign and a wonder concerning Egypt and concerning Ethiopia,
\VS{4}so the king of Assyria will lead away the captives of Egypt and the exiles of Ethiopia, young and old, naked and barefoot, and with buttocks uncovered, to the shame of Egypt.
\VS{5}They will be dismayed and confounded, because of Ethiopia their expectation, and of Egypt their glory.
\VS{6}The inhabitants of this coast land will say in that day, ‘Behold, this is our expectation, where we fled for help to be delivered from the king of Assyria. And we, how will we escape?’ ”

\par }\Chap{21}{\PP \VerseOne{1}The burden of the wilderness of the sea.
\par }{\PP As whirlwinds in the South sweep through, it comes from the wilderness, from an awesome land.
\VS{2}A grievous vision is declared to me. The treacherous man deals treacherously, and the destroyer destroys. Go up, Elam; attack! I have stopped all of Media’s sighing.
\VS{3}Therefore my thighs are filled with anguish. Pains have seized me, like the pains of a woman in labor. I am in so much pain that I can’t hear. I am so dismayed that I can’t see.
\VS{4}My heart flutters. Horror has frightened me. The twilight that I desired has been turned into trembling for me.
\VS{5}They prepare the table. They set the watch. They eat. They drink. Rise up, you princes, oil the shield!
\VS{6}For the Lord said to me, “Go, set a watchman. Let him declare what he sees.
\VS{7}When he sees a troop, horsemen in pairs, a troop of donkeys, a troop of camels, he shall listen diligently with great attentiveness.”
\VS{8}He cried like a lion: “Lord, I stand continually on the watchtower in the daytime, and every night I stay at my post.
\VS{9}Behold, here comes a troop of men, horsemen in pairs.” He answered, “Fallen, fallen is Babylon; and all the engraved images of her gods are broken to the ground.
\par }{\PP \VS{10}You are my threshing, and the grain of my floor!” That which I have heard from Yahweh of Armies, the God of Israel, I have declared to you.
\par }{\PP \VS{11}The burden of Dumah.
\par }{\PP One calls to me out of Seir, “Watchman, what of the night? Watchman, what of the night?”
\VS{12}The watchman said, “The morning comes, and also the night. If you will inquire, inquire. Come back again.”
\par }{\PP \VS{13}The burden on Arabia.
\par }{\PP In the forest in Arabia you will lodge, you caravans of Dedanites.
\VS{14}They brought water to him who was thirsty. The inhabitants of the land of Tema met the fugitives with their bread.
\VS{15}For they fled away from the swords, from the drawn sword, from the bent bow, and from the heat of battle.
\VS{16}For the Lord said to me, “Within a year, as a worker bound by contract would count it, all the glory of Kedar will fail,
\VS{17}and the residue of the number of the archers, the mighty men of the children of Kedar, will be few; for Yahweh, the God of Israel, has spoken it.”

\par }\Chap{22}{\PP \VerseOne{1}The burden of the valley of vision.
\par }{\PP What ails you now, that you have all gone up to the housetops?
\VS{2}You that are full of shouting, a tumultuous city, a joyous town; your slain are not slain with the sword, neither are they dead in battle.
\VS{3}All your rulers fled away together. They were bound by the archers. All who were found by you were bound together. They fled far away.
\VS{4}Therefore I said, “Look away from me. I will weep bitterly. Don’t labor to comfort me for the destruction of the daughter of my people.
\par }{\PP \VS{5}For it is a day of confusion, and of treading down, and of perplexity, from the Lord, Yahweh of Armies, in the valley of vision, a breaking down of the walls, and a crying to the mountains.”
\VS{6}Elam carried his quiver, with chariots of men and horsemen; and Kir uncovered the shield.
\VS{7}Your choicest valleys were full of chariots, and the horsemen set themselves in array at the gate.
\VS{8}He took away the covering of Judah; and you looked in that day to the armor in the house of the forest.
\VS{9}You saw the breaches of David’s city, that they were many; and you gathered together the waters of the lower pool.
\VS{10}You counted the houses of Jerusalem, and you broke down the houses to fortify the wall.
\VS{11}You also made a reservoir between the two walls for the water of the old pool. But you didn’t look to him who had done this, neither did you have respect for him who planed it long ago.
\par }{\PP \VS{12}In that day, the Lord, Yahweh of Armies, called to weeping, to mourning, to baldness, and to dressing in sackcloth;
\VS{13}and behold, joy and gladness, killing cattle and killing sheep, eating meat and drinking wine: “Let’s eat and drink, for tomorrow we will die.”
\VS{14}Yahweh of Armies revealed himself in my ears, “Surely this iniquity will not be forgiven you until you die,” says the Lord, Yahweh of Armies.
\par }{\PP \VS{15}The Lord, Yahweh of Armies says, “Go, get yourself to this treasurer, even to Shebna, who is over the house, and say,
\VS{16}‘What are you doing here? Who has you here, that you have dug out a tomb here?’ Cutting himself out a tomb on high, chiseling a habitation for himself in the rock!”
\VS{17}Behold, Yahweh will overcome you and hurl you away violently. Yes, he will grasp you firmly.
\VS{18}He will surely wind you around and around, and throw you like a ball into a large country. There you will die, and there the chariots of your glory will be, you shame of your lord’s house.
\VS{19}I will thrust you from your office. You will be pulled down from your station.
\par }{\PP \VS{20}It will happen in that day that I will call my servant Eliakim the son of Hilkiah,
\VS{21}and I will clothe him with your robe, and strengthen him with your belt. I will commit your government into his hand; and he will be a father to the inhabitants of Jerusalem, and to the house of Judah.
\VS{22}I will lay the key of David’s house on his shoulder. He will open, and no one will shut. He will shut, and no one will open.
\VS{23}I will fasten him like a nail in a sure place. He will be for a throne of glory to his father’s house.
\VS{24}They will hang on him all the glory of his father’s house, the offspring and the issue, every small vessel, from the cups even to all the pitchers.
\VS{25}“In that day,” says Yahweh of Armies, “the nail that was fastened in a sure place will give way. It will be cut down and fall. The burden that was on it will be cut off, for Yahweh has spoken it.”

\par }\Chap{23}{\PP \VerseOne{1}The burden of Tyre.
\par }{\PP Howl, you ships of Tarshish! For it is laid waste, so that there is no house, no entering in. From the land of Kittim it is revealed to them.
\VS{2}Be still, you inhabitants of the coast, you whom the merchants of Sidon that pass over the sea have replenished.
\VS{3}On great waters, the seed of the Shihor, the harvest of the Nile, was her revenue. She was the market of nations.
\VS{4}Be ashamed, Sidon; for the sea has spoken, the stronghold of the sea, saying, “I have not travailed, nor given birth, neither have I nourished young men, nor brought up virgins.”
\VS{5}When the report comes to Egypt, they will be in anguish at the report of Tyre.
\VS{6}Pass over to Tarshish! Wail, you inhabitants of the coast!
\VS{7}Is this your joyous city, whose antiquity is of ancient days, whose feet carried her far away to travel?
\par }{\PP \VS{8}Who has planned this against Tyre, the giver of crowns, whose merchants are princes, whose traders are the honorable of the earth?
\VS{9}Yahweh of Armies has planned it, to stain the pride of all glory, to bring into contempt all the honorable of the earth.
\VS{10}Pass through your land like the Nile, daughter of Tarshish. There is no restraint any more.
\VS{11}He has stretched out his hand over the sea. He has shaken the kingdoms. Yahweh has ordered the destruction of Canaan’s strongholds.
\VS{12}He said, “You shall rejoice no more, you oppressed virgin daughter of Sidon. Arise, pass over to Kittim. Even there you will have no rest.”
\par }{\PP \VS{13}Behold, the land of the Chaldeans. This people didn’t exist. The Assyrians founded it for those who dwell in the wilderness. They set up their towers. They overthrew its palaces. They made it a ruin.
\VS{14}Howl, you ships of Tarshish, for your stronghold is laid waste!
\VS{15}It will come to pass in that day that Tyre will be forgotten seventy years, according to the days of one king. After the end of seventy years it will be to Tyre like in the song of the prostitute.
\VS{16}Take a harp; go about the city, you prostitute that has been forgotten. Make sweet melody. Sing many songs, that you may be remembered.
\VS{17}It will happen after the end of seventy years that Yahweh will visit Tyre. She will return to her wages, and will play the prostitute with all the kingdoms of the world on the surface of the earth.
\VS{18}Her merchandise and her wages will be holiness to Yahweh. It will not be treasured nor laid up; for her merchandise will be for those who dwell before Yahweh, to eat sufficiently, and for durable clothing.

\par }\Chap{24}{\PP \VerseOne{1}Behold, Yahweh makes the earth empty, makes it waste, turns it upside down, and scatters its inhabitants.
\VS{2}It will be as with the people, so with the priest; as with the servant, so with his master; as with the maid, so with her mistress; as with the buyer, so with the seller; as with the creditor, so with the debtor; as with the taker of interest, so with the giver of interest.
\VS{3}The earth will be utterly emptied and utterly laid waste; for Yahweh has spoken this word.
\VS{4}The earth mourns and fades away. The world languishes and fades away. The lofty people of the earth languish.
\VS{5}The earth also is polluted under its inhabitants, because they have transgressed the laws, violated the statutes, and broken the everlasting covenant.
\VS{6}Therefore the curse has devoured the earth, and those who dwell therein are found guilty. Therefore the inhabitants of the earth are burned, and few men are left.
\VS{7}The new wine mourns. The vine languishes. All the merry-hearted sigh.
\VS{8}The mirth of tambourines ceases. The sound of those who rejoice ends. The joy of the harp ceases.
\VS{9}They will not drink wine with a song. Strong drink will be bitter to those who drink it.
\VS{10}The confused city is broken down. Every house is shut up, that no man may come in.
\VS{11}There is a crying in the streets because of the wine. All joy is darkened. The mirth of the land is gone.
\VS{12}The city is left in desolation, and the gate is struck with destruction.
\VS{13}For it will be so within the earth among the peoples, as the shaking of an olive tree, as the gleanings when the vintage is done.
\par }{\PP \VS{14}These shall lift up their voice. They will shout for the majesty of Yahweh. They cry aloud from the sea.
\VS{15}Therefore glorify Yahweh in the east, even the name of Yahweh, the God of Israel, in the islands of the sea!
\VS{16}From the uttermost part of the earth have we heard songs. Glory to the righteous!
\par }{\PP But I said, “I pine away! I pine away! woe is me!” The treacherous have dealt treacherously. Yes, the treacherous have dealt very treacherously.
\VS{17}Fear, the pit, and the snare, are on you who inhabit the earth.
\VS{18}It will happen that he who flees from the noise of the fear will fall into the pit; and he who comes up out of the middle of the pit will be taken in the snare; for the windows on high are opened, and the foundations of the earth tremble.
\VS{19}The earth is utterly broken. The earth is torn apart. The earth is shaken violently.
\VS{20}The earth will stagger like a drunken man, and will sway back and forth like a hammock. Its disobedience will be heavy on it, and it will fall and not rise again.
\par }{\PP \VS{21}It will happen in that day that Yahweh will punish the army of the high ones on high, and the kings of the earth on the earth.
\VS{22}They will be gathered together, as prisoners are gathered in the pit, and will be shut up in the prison; and after many days they will be visited.
\VS{23}Then the moon will be confounded, and the sun ashamed; for Yahweh of Armies will reign on Mount Zion, and in Jerusalem; and glory will be before his elders.

\par }\Chap{25}{\PP \VerseOne{1}Yahweh, you are my God. I will exalt you! I will praise your name, for you have done wonderful things, things planned long ago, in complete faithfulness and truth.
\VS{2}For you have made a city into a heap, a fortified city into a ruin, a palace of strangers to be no city. It will never be built.
\VS{3}Therefore a strong people will glorify you. A city of awesome nations will fear you.
\VS{4}For you have been a stronghold to the poor, a stronghold to the needy in his distress, a refuge from the storm, a shade from the heat, when the blast of the dreaded ones is like a storm against the wall.
\VS{5}As the heat in a dry place you will bring down the noise of strangers; as the heat by the shade of a cloud, the song of the dreaded ones will be brought low.
\par }{\PP \VS{6}In this mountain, Yahweh of Armies will make all peoples a feast of choice meat,\FTNT{*}{{\FR 25:6 }
literally, fat things} a feast of choice wines, of choice meat full of marrow, of well refined choice wines.
\VS{7}He will destroy in this mountain the surface of the covering that covers all peoples, and the veil that is spread over all nations.
\VS{8}He has swallowed up death forever! The Lord Yahweh will wipe away tears from off all faces. He will take the reproach of his people away from off all the earth, for Yahweh has spoken it.
\par }{\PP \VS{9}It shall be said in that day, “Behold, this is our God! We have waited for him, and he will save us! This is Yahweh! We have waited for him. We will be glad and rejoice in his salvation!”
\VS{10}For Yahweh’s hand will rest in this mountain.
\par }{\PP Moab will be trodden down in his place, even like straw is trodden down in the water of the dunghill.
\VS{11}He will spread out his hands in the middle of it, like one who swims spreads out hands to swim, but his pride will be humbled together with the craft of his hands.
\VS{12}He has brought the high fortress of your walls down, laid low, and brought to the ground, even to the dust.

\par }\Chap{26}{\PP \VerseOne{1}In that day, this song will be sung in the land of Judah:
\par }{\Q “We have a strong city.
\par }{\QB God appoints salvation for walls and bulwarks.
\par }{\Q \VS{2}Open the gates, that the righteous nation may enter:
\par }{\QB the one which keeps faith.
\par }{\Q \VS{3}You will keep whoever’s mind is steadfast in perfect peace,
\par }{\QB because he trusts in you.
\par }{\Q \VS{4}Trust in Yahweh forever;
\par }{\QB for in Yah, Yahweh, is an everlasting Rock.
\par }{\Q \VS{5}For he has brought down those who dwell on high, the lofty city.
\par }{\QB He lays it low.
\par }{\QB He lays it low even to the ground.
\par }{\QB He brings it even to the dust.
\par }{\Q \VS{6}The foot shall tread it down,
\par }{\QB even the feet of the poor
\par }{\QB and the steps of the needy.”
\par }{\Q \VS{7}The way of the just is uprightness.
\par }{\QB You who are upright make the path of the righteous level.
\par }{\BB \par }{\Q \VS{8}Yes, in the way of your judgments, Yahweh, we have waited for you.
\par }{\QB Your name and your renown are the desire of our soul.
\par }{\Q \VS{9}With my soul I have desired you in the night.
\par }{\QB Yes, with my spirit within me I will seek you earnestly;
\par }{\QB for when your judgments are in the earth, the inhabitants of the world learn righteousness.
\par }{\Q \VS{10}Let favor be shown to the wicked,
\par }{\QB yet he will not learn righteousness.
\par }{\Q In the land of uprightness he will deal wrongfully,
\par }{\QB and will not see Yahweh’s majesty.
\par }{\BB \par }{\Q \VS{11}Yahweh, your hand is lifted up, yet they don’t see;
\par }{\QB but they will see your zeal for the people, and be disappointed.
\par }{\QB Yes, fire will consume your adversaries.
\par }{\Q \VS{12}Yahweh, you will ordain peace for us,
\par }{\QB for you have also done all our work for us.
\par }{\Q \VS{13}Yahweh our God, other lords besides you have had dominion over us,
\par }{\QB but we will only acknowledge your name.
\par }{\Q \VS{14}The dead shall not live.
\par }{\QB The departed spirits shall not rise.
\par }{\Q Therefore you have visited and destroyed them,
\par }{\QB and caused all memory of them to perish.
\par }{\Q \VS{15}You have increased the nation, O Yahweh.
\par }{\QB You have increased the nation!
\par }{\Q You are glorified!
\par }{\QB You have enlarged all the borders of the land.
\par }{\BB \par }{\Q \VS{16}Yahweh, in trouble they have visited you.
\par }{\QB They poured out a prayer when your chastening was on them.
\par }{\Q \VS{17}Just as a woman with child, who draws near the time of her delivery,
\par }{\QB is in pain and cries out in her pangs,
\par }{\QB so we have been before you, Yahweh.
\par }{\Q \VS{18}We have been with child.
\par }{\QB We have been in pain.
\par }{\Q We gave birth, it seems, only to wind.
\par }{\QB We have not worked any deliverance in the earth;
\par }{\QB neither have the inhabitants of the world fallen.
\par }{\Q \VS{19}Your dead shall live.
\par }{\QB My dead bodies shall arise.
\par }{\Q Awake and sing, you who dwell in the dust;
\par }{\QB for your dew is like the dew of herbs,
\par }{\QB and the earth will cast out the departed spirits.
\par }{\BB \par }{\Q \VS{20}Come, my people, enter into your rooms,
\par }{\QB and shut your doors behind you.
\par }{\Q Hide yourself for a little moment,
\par }{\QB until the indignation is past.
\par }{\Q \VS{21}For, behold, Yahweh comes out of his place to punish the inhabitants of the earth for their iniquity.
\par }{\QB The earth also will disclose her blood, and will no longer cover her slain.

\par }\Chap{27}{\PP \VerseOne{1}In that day, Yahweh with his hard and great and strong sword will punish leviathan, the fleeing serpent, and leviathan the twisted serpent; and he will kill the dragon that is in the sea.
\par }{\PP \VS{2}In that day, sing to her, “A pleasant vineyard!
\VS{3}I, Yahweh, am its keeper. I will water it every moment. Lest anyone damage it, I will keep it night and day.
\VS{4}Wrath is not in me, but if I should find briers and thorns, I would do battle! I would march on them and I would burn them together.
\VS{5}Or else let him take hold of my strength, that he may make peace with me. Let him make peace with me.”
\par }{\PP \VS{6}In days to come, Jacob will take root. Israel will blossom and bud. They will fill the surface of the world with fruit.
\VS{7}Has he struck them as he struck those who struck them? Or are they killed like those who killed them were killed?
\VS{8}In measure, when you send them away, you contend with them. He has removed them with his rough blast in the day of the east wind.
\VS{9}Therefore by this the iniquity of Jacob will be forgiven, and this is all the fruit of taking away his sin: that he makes all the stones of the altar as chalk stones that are beaten in pieces, so that the Asherah poles and the incense altars shall rise no more.
\VS{10}For the fortified city is solitary, a habitation deserted and forsaken, like the wilderness. The calf will feed there, and there he will lie down, and consume its branches.
\VS{11}When its boughs are withered, they will be broken off. The women will come and set them on fire, for they are a people of no understanding. Therefore he who made them will not have compassion on them, and he who formed them will show them no favor.
\par }{\PP \VS{12}It will happen in that day that Yahweh will thresh from the flowing stream of the Euphrates to the brook of Egypt; and you will be gathered one by one, children of Israel.
\par }{\PP \VS{13}It will happen in that day that a great trumpet will be blown; and those who were ready to perish in the land of Assyria, and those who were outcasts in the land of Egypt, shall come; and they will worship Yahweh in the holy mountain at Jerusalem.

\par }\Chap{28}{\PP \VerseOne{1}Woe to the crown of pride of the drunkards of Ephraim, and to the fading flower of his glorious beauty, which is on the head of the fertile valley of those who are overcome with wine!
\VS{2}Behold, the Lord has a mighty and strong one. Like a storm of hail, a destroying storm, and like a storm of mighty waters overflowing, he will cast them down to the earth with his hand.
\VS{3}The crown of pride of the drunkards of Ephraim will be trodden under foot.
\VS{4}The fading flower of his glorious beauty, which is on the head of the fertile valley, shall be like the first-ripe fig before the summer; which someone picks and eats as soon as he sees it.
\VS{5}In that day, Yahweh of Armies will become a crown of glory and a diadem of beauty to the residue of his people,
\VS{6}and a spirit of justice to him who sits in judgment, and strength to those who turn back the battle at the gate.
\par }{\PP \VS{7}They also reel with wine, and stagger with strong drink. The priest and the prophet reel with strong drink. They are swallowed up by wine. They stagger with strong drink. They err in vision. They stumble in judgment.
\VS{8}For all tables are completely full of filthy vomit and filthiness.
\par }{\PP \VS{9}Whom will he teach knowledge? To whom will he explain the message? Those who are weaned from the milk, and drawn from the breasts?
\VS{10}For it is precept on precept, precept on precept; line on line, line on line; here a little, there a little.
\par }{\PP \VS{11}But he will speak to this nation with stammering lips and in another language,
\VS{12}to whom he said, “This is the resting place. Give rest to the weary,” and “This is the refreshing;” yet they would not hear.
\VS{13}Therefore Yahweh’s word will be to them precept on precept, precept on precept; line on line, line on line; here a little, there a little; that they may go, fall backward, be broken, be snared, and be taken.
\par }{\PP \VS{14}Therefore hear Yahweh’s word, you scoffers, that rule this people in Jerusalem:
\VS{15}“Because you have said, ‘We have made a covenant with death, and we are in agreement with Sheol.
\FTNT{*}{{\FR 28:15 }
Sheol is the place of the dead.} When the overflowing scourge passes through, it won’t come to us; for we have made lies our refuge, and we have hidden ourselves under falsehood.’ ”
\VS{16}Therefore the Lord Yahweh says, “Behold, I lay in Zion for a foundation a stone, a tried stone, a precious cornerstone of a sure foundation. He who believes shall not act hastily.
\VS{17}I will make justice the measuring line, and righteousness the plumb line. The hail will sweep away the refuge of lies, and the waters will overflow the hiding place.
\VS{18}Your covenant with death shall be annulled, and your agreement with Sheol
\FTNT{†}{{\FR 28:18 }
Sheol is the place of the dead.} shall not stand. When the overflowing scourge passes through, then you will be trampled down by it.
\VS{19}As often as it passes through, it will seize you; for morning by morning it will pass through, by day and by night; and it will be nothing but terror to understand the message.”
\VS{20}For the bed is too short to stretch out on, and the blanket is too narrow to wrap oneself in.
\VS{21}For Yahweh will rise up as on Mount Perazim. He will be angry as in the valley of Gibeon; that he may do his work, his unusual work, and bring to pass his act, his extraordinary act.
\VS{22}Now therefore don’t be scoffers, lest your bonds be made strong; for I have heard a decree of destruction from the Lord, Yahweh of Armies, on the whole earth.
\par }{\PP \VS{23}Give ear, and hear my voice! Listen, and hear my speech!
\VS{24}Does he who plows to sow plow continually? Does he keep turning the soil and breaking the clods?
\VS{25}When he has leveled its surface, doesn’t he plant the dill, and scatter the cumin seed, and put in the wheat in rows, the barley in the appointed place, and the spelt in its place?
\VS{26}For his God instructs him in right judgment and teaches him.
\VS{27}For the dill are not threshed with a sharp instrument, neither is a cart wheel turned over the cumin; but the dill is beaten out with a stick, and the cumin with a rod.
\VS{28}Bread flour must be ground; so he will not always be threshing it. Although he drives the wheel of his threshing cart over it, his horses don’t grind it.
\VS{29}This also comes out from Yahweh of Armies, who is wonderful in counsel, and excellent in wisdom.

\par }\Chap{29}{\PP \VerseOne{1}Woe to Ariel! Ariel, the city where David encamped! Add year to year; let the feasts come around;
\VS{2}then I will distress Ariel, and there will be mourning and lamentation. She shall be to me as an altar hearth.\FTNT{*}{{\FR 29:2 }
or, Ariel}
\VS{3}I will encamp against you all around you, and will lay siege against you with posted troops. I will raise siege works against you.
\VS{4}You will be brought down, and will speak out of the ground. Your speech will mumble out of the dust. Your voice will be as of one who has a familiar spirit, out of the ground, and your speech will whisper out of the dust.
\par }{\PP \VS{5}But the multitude of your foes will be like fine dust, and the multitude of the ruthless ones like chaff that blows away. Yes, it will be in an instant, suddenly.
\VS{6}She will be visited by Yahweh of Armies with thunder, with earthquake, with great noise, with whirlwind and storm, and with the flame of a devouring fire.
\VS{7}The multitude of all the nations that fight against Ariel, even all who fight against her and her stronghold, and who distress her, will be like a dream, a vision of the night.
\VS{8}It will be like when a hungry man dreams, and behold, he eats; but he awakes, and his hunger isn’t satisfied; or like when a thirsty man dreams, and behold, he drinks; but he awakes, and behold, he is faint, and he is still thirsty. The multitude of all the nations that fight against Mount Zion will be like that.
\par }{\PP \VS{9}Pause and wonder! Blind yourselves and be blind! They are drunken, but not with wine; they stagger, but not with strong drink.
\VS{10}For Yahweh has poured out on you a spirit of deep sleep, and has closed your eyes, the prophets; and he has covered your heads, the seers.
\VS{11}All vision has become to you like the words of a book that is sealed, which men deliver to one who is educated, saying, “Read this, please;” and he says, “I can’t, for it is sealed;”
\VS{12}and the book is delivered to one who is not educated, saying, “Read this, please;” and he says, “I can’t read.”
\par }{\PP \VS{13}The Lord said, “Because this people draws near with their mouth and honors me with their lips, but they have removed their heart far from me, and their fear of me is a commandment of men which has been taught;
\VS{14}therefore, behold, I will proceed to do a marvelous work among this people, even a marvelous work and a wonder; and the wisdom of their wise men will perish, and the understanding of their prudent men will be hidden.”
\par }{\PP \VS{15}Woe to those who deeply hide their counsel from Yahweh, and whose deeds are in the dark, and who say, “Who sees us?” and “Who knows us?”
\VS{16}You turn things upside down! Should the potter be thought to be like clay; that the thing made should say about him who made it, “He didn’t make me;” or the thing formed say of him who formed it, “He has no understanding?”
\par }{\PP \VS{17}Isn’t it yet a very little while, and Lebanon will be turned into a fruitful field, and the fruitful field will be regarded as a forest?
\VS{18}In that day, the deaf will hear the words of the book, and the eyes of the blind will see out of obscurity and out of darkness.
\VS{19}The humble also will increase their joy in Yahweh, and the poor among men will rejoice in the Holy One of Israel.
\VS{20}For the ruthless is brought to nothing, and the scoffer ceases, and all those who are alert to do evil are cut off—
\VS{21}who cause a person to be indicted by a word, and lay a snare for one who reproves in the gate, and who deprive the innocent of justice with false testimony.
\par }{\PP \VS{22}Therefore Yahweh, who redeemed Abraham, says concerning the house of Jacob: “Jacob shall no longer be ashamed, neither shall his face grow pale.
\VS{23}But when he sees his children, the work of my hands, in the middle of him, they will sanctify my name. Yes, they will sanctify the Holy One of Jacob, and will stand in awe of the God of Israel.
\VS{24}They also who err in spirit will come to understanding, and those who grumble will receive instruction.”

\par }\Chap{30}{\PP \VerseOne{1}“Woe to the rebellious children”, says Yahweh, “who take counsel, but not from me; and who make an alliance, but not with my Spirit, that they may add sin to sin,
\VS{2}who set out to go down into Egypt, and have not asked my advice, to strengthen themselves in the strength of Pharaoh, and to take refuge in the shadow of Egypt!
\VS{3}Therefore the strength of Pharaoh will be your shame, and the refuge in the shadow of Egypt your confusion.
\VS{4}For their princes are at Zoan, and their ambassadors have come to Hanes.
\VS{5}They shall all be ashamed because of a people that can’t profit them, that are not a help nor profit, but a shame, and also a reproach.”
\par }{\PP \VS{6}The burden of the animals of the South.
\par }{\PP Through the land of trouble and anguish, of the lioness and the lion, the viper and fiery flying serpent, they carry their riches on the shoulders of young donkeys, and their treasures on the humps of camels, to an unprofitable people.
\VS{7}For Egypt helps in vain, and to no purpose; therefore I have called her Rahab who sits still.
\VS{8}Now go, write it before them on a tablet, and inscribe it in a book, that it may be for the time to come forever and ever.
\VS{9}For it is a rebellious people, lying children, children who will not hear Yahweh’s law;
\VS{10}who tell the seers, “Don’t see!” and the prophets, “Don’t prophesy to us right things. Tell us pleasant things. Prophesy deceits.
\VS{11}Get out of the way. Turn away from the path. Cause the Holy One of Israel to cease from before us.”
\VS{12}Therefore the Holy One of Israel says, “Because you despise this word, and trust in oppression and perverseness, and rely on it,
\VS{13}therefore this iniquity shall be to you like a breach ready to fall, swelling out in a high wall, whose breaking comes suddenly in an instant.
\VS{14}He will break it as a potter’s vessel is broken, breaking it in pieces without sparing, so that there won’t be found among the broken pieces a piece good enough to take fire from the hearth, or to dip up water out of the cistern.”
\par }{\PP \VS{15}For thus said the Lord Yahweh, the Holy One of Israel, “You will be saved in returning and rest. Your strength will be in quietness and in confidence.” You refused,
\VS{16}but you said, “No, for we will flee on horses;” therefore you will flee; and, “We will ride on the swift;” therefore those who pursue you will be swift.
\VS{17}One thousand will flee at the threat of one. At the threat of five, you will flee until you are left like a beacon on the top of a mountain, and like a banner on a hill.
\par }{\PP \VS{18}Therefore Yahweh will wait, that he may be gracious to you; and therefore he will be exalted, that he may have mercy on you, for Yahweh is a God of justice. Blessed are all those who wait for him.
\VS{19}For the people will dwell in Zion at Jerusalem. You will weep no more. He will surely be gracious to you at the voice of your cry. When he hears you, he will answer you.
\VS{20}Though the Lord may give you the bread of adversity and the water of affliction, yet your teachers won’t be hidden any more, but your eyes will see your teachers;
\VS{21}and when you turn to the right hand, and when you turn to the left, your ears will hear a voice behind you, saying, “This is the way. Walk in it.”
\VS{22}You shall defile the overlaying of your engraved images of silver, and the plating of your molten images of gold. You shall cast them away as an unclean thing. You shall tell it, “Go away!”
\par }{\PP \VS{23}He will give the rain for your seed, with which you will sow the ground; and bread of the increase of the ground will be rich and plentiful. In that day, your livestock will feed in large pastures.
\VS{24}The oxen likewise and the young donkeys that till the ground will eat savory feed, which has been winnowed with the shovel and with the fork.
\VS{25}There will be brooks and streams of water on every lofty mountain and on every high hill in the day of the great slaughter, when the towers fall.
\VS{26}Moreover the light of the moon will be like the light of the sun, and the light of the sun will be seven times brighter, like the light of seven days, in the day that Yahweh binds up the fracture of his people, and heals the wound they were struck with.
\par }{\PP \VS{27}Behold, Yahweh’s name comes from far away, burning with his anger, and in thick rising smoke. His lips are full of indignation. His tongue is as a devouring fire.
\VS{28}His breath is as an overflowing stream that reaches even to the neck, to sift the nations with the sieve of destruction. A bridle that leads to ruin will be in the jaws of the peoples.
\VS{29}You will have a song, as in the night when a holy feast is kept, and gladness of heart, as when one goes with a flute to come to Yahweh’s mountain, to Israel’s Rock.
\VS{30}Yahweh will cause his glorious voice to be heard, and will show the descent of his arm, with the indignation of his anger, and the flame of a devouring fire, with a blast, storm, and hailstones.
\VS{31}For through Yahweh’s voice the Assyrian will be dismayed. He will strike him with his rod.
\VS{32}Every stroke of the rod of punishment, which Yahweh will lay on him, will be with the sound of tambourines and harps. He will fight with them in battles, brandishing weapons.
\VS{33}For his burning place has long been ready. Yes, for the king it is prepared. He has made its pyre deep and large with fire and much wood. Yahweh’s breath, like a stream of sulfur, kindles it.

\par }\PoetryChap{31}{\Q \VerseOne{1}Woe to those who go down to Egypt for help,
\par }{\QB and rely on horses,
\par }{\QB and trust in chariots because they are many,
\par }{\QB and in horsemen because they are very strong,
\par }{\QB but they don’t look to the Holy One of Israel,
\par }{\QB and they don’t seek Yahweh!
\par }{\Q \VS{2}Yet he also is wise, and will bring disaster,
\par }{\QB and will not call back his words, but will arise against the house of the evildoers,
\par }{\QB and against the help of those who work iniquity.
\par }{\Q \VS{3}Now the Egyptians are men, and not God;
\par }{\QB and their horses flesh, and not spirit.
\par }{\Q When Yahweh stretches out his hand, both he who helps shall stumble,
\par }{\QB and he who is helped shall fall,
\par }{\QB and they all shall be consumed together.
\par }{\BB \par }{\Q \VS{4}For Yahweh says to me,
\par }{\Q “As the lion and the young lion growling over his prey,
\par }{\QB if a multitude of shepherds is called together against him,
\par }{\QB will not be dismayed at their voice,
\par }{\QB nor abase himself for their noise,
\par }{\QB so Yahweh of Armies will come down to fight on Mount Zion and on its heights.
\par }{\Q \VS{5}As birds hovering, so Yahweh of Armies will protect Jerusalem.
\par }{\QB He will protect and deliver it.
\par }{\QB He will pass over and preserve it.”
\par }{\Q \VS{6}Return to him from whom you have deeply revolted, children of Israel.
\VS{7}For in that day everyone shall cast away his idols of silver and his idols of gold—sin which your own hands have made for you.
\par }{\PP \VS{8}“The Assyrian will fall by the sword, not of man;
\par }{\QB and the sword, not of mankind, shall devour him.
\par }{\Q He will flee from the sword,
\par }{\QB and his young men will become subject to forced labor.
\par }{\Q \VS{9}His rock will pass away by reason of terror,
\par }{\QB and his princes will be afraid of the banner,”
\par }{\Q says Yahweh, whose fire is in Zion,
\par }{\QB and his furnace in Jerusalem.

\par }\PoetryChap{32}{\Q \VerseOne{1}Behold, a king shall reign in righteousness,
\par }{\QB and princes shall rule in justice.
\par }{\Q \VS{2}A man shall be as a hiding place from the wind,
\par }{\QB and a covert from the storm,
\par }{\QB as streams of water in a dry place,
\par }{\QB as the shade of a large rock in a weary land.
\par }{\Q \VS{3}The eyes of those who see will not be dim,
\par }{\QB and the ears of those who hear will listen.
\par }{\Q \VS{4}The heart of the rash will understand knowledge,
\par }{\QB and the tongue of the stammerers will be ready to speak plainly.
\par }{\Q \VS{5}The fool will no longer be called noble,
\par }{\QB nor the scoundrel be highly respected.
\par }{\Q \VS{6}For the fool will speak folly,
\par }{\QB and his heart will work iniquity,
\par }{\QB to practice profanity,
\par }{\QB and to utter error against Yahweh,
\par }{\QB to make empty the soul of the hungry,
\par }{\QB and to cause the drink of the thirsty to fail.
\par }{\Q \VS{7}The ways of the scoundrel are evil.
\par }{\QB He devises wicked plans to destroy the humble with lying words,
\par }{\QB even when the needy speaks right.
\par }{\Q \VS{8}But the noble devises noble things;
\par }{\QB and he will continue in noble things.
\par }{\BB \par }{\Q \VS{9}Rise up, you women who are at ease! Hear my voice!
\par }{\QB You careless daughters, give ear to my speech!
\par }{\Q \VS{10}For days beyond a year you will be troubled, you careless women;
\par }{\QB for the vintage will fail.
\par }{\QB The harvest won’t come.
\par }{\Q \VS{11}Tremble, you women who are at ease!
\par }{\QB Be troubled, you careless ones!
\par }{\QB Strip yourselves, make yourselves naked,
\par }{\QB and put sackcloth on your waist.
\par }{\Q \VS{12}Beat your breasts for the pleasant fields,
\par }{\QB for the fruitful vine.
\par }{\Q \VS{13}Thorns and briers will come up on my people’s land;
\par }{\QB yes, on all the houses of joy in the joyous city.
\par }{\Q \VS{14}For the palace will be forsaken.
\par }{\QB The populous city will be deserted.
\par }{\QB The hill and the watchtower will be for dens forever,
\par }{\QB a delight for wild donkeys,
\par }{\QB a pasture of flocks,
\par }{\Q \VS{15}until the Spirit is poured on us from on high,
\par }{\QB and the wilderness becomes a fruitful field,
\par }{\QB and the fruitful field is considered a forest.
\par }{\BB \par }{\Q \VS{16}Then justice will dwell in the wilderness;
\par }{\QB and righteousness will remain in the fruitful field.
\par }{\Q \VS{17}The work of righteousness will be peace,
\par }{\QB and the effect of righteousness, quietness and confidence forever.
\par }{\Q \VS{18}My people will live in a peaceful habitation,
\par }{\QB in safe dwellings,
\par }{\QB and in quiet resting places,
\par }{\Q \VS{19}though hail flattens the forest,
\par }{\QB and the city is leveled completely.
\par }{\Q \VS{20}Blessed are you who sow beside all waters,
\par }{\QB who send out the feet of the ox and the donkey.

\par }\PoetryChap{33}{\Q \VerseOne{1}Woe to you who destroy, but you weren’t destroyed,
\par }{\QB and who betray, but nobody betrayed you!
\par }{\Q When you have finished destroying, you will be destroyed;
\par }{\QB and when you have finished betrayal, you will be betrayed.
\par }{\BB \par }{\Q \VS{2}Yahweh, be gracious to us. We have waited for you.
\par }{\QB Be our strength every morning,
\par }{\QB our salvation also in the time of trouble.
\par }{\Q \VS{3}At the noise of the thunder, the peoples have fled.
\par }{\QB When you lift yourself up, the nations are scattered.
\par }{\Q \VS{4}Your plunder will be gathered as the caterpillar gathers.
\par }{\QB Men will leap on it as locusts leap.
\par }{\Q \VS{5}Yahweh is exalted, for he dwells on high.
\par }{\QB He has filled Zion with justice and righteousness.
\par }{\Q \VS{6}There will be stability in your times, abundance of salvation, wisdom, and knowledge.
\par }{\QB The fear of Yahweh is your treasure.
\par }{\BB \par }{\Q \VS{7}Behold, their valiant ones cry outside;
\par }{\QB the ambassadors of peace weep bitterly.
\par }{\Q \VS{8}The highways are desolate.
\par }{\QB The traveling man ceases.
\par }{\QB The covenant is broken.
\par }{\QB He has despised the cities.
\par }{\QB He doesn’t respect man.
\par }{\Q \VS{9}The land mourns and languishes.
\par }{\QB Lebanon is confounded and withers away.
\par }{\QB Sharon is like a desert, and Bashan and Carmel are stripped bare.
\par }{\Q \VS{10}“Now I will arise,” says Yahweh.
\par }{\QB “Now I will lift myself up.
\par }{\QB Now I will be exalted.
\par }{\Q \VS{11}You will conceive chaff.
\par }{\QB You will give birth to stubble.
\par }{\QB Your breath is a fire that will devour you.
\par }{\Q \VS{12}The peoples will be like the burning of lime,
\par }{\QB like thorns that are cut down and burned in the fire.
\par }{\BB \par }{\Q \VS{13}Hear, you who are far off, what I have done;
\par }{\QB and, you who are near, acknowledge my might.”
\par }{\Q \VS{14}The sinners in Zion are afraid.
\par }{\QB Trembling has seized the godless ones.
\par }{\Q Who among us can live with the devouring fire?
\par }{\QB Who among us can live with everlasting burning?
\par }{\Q \VS{15}He who walks righteously
\par }{\QB and speaks blamelessly,
\par }{\QB he who despises the gain of oppressions,
\par }{\QB who gestures with his hands, refusing to take a bribe,
\par }{\QB who stops his ears from hearing of blood,
\par }{\QB and shuts his eyes from looking at evil—
\par }{\Q \VS{16}he will dwell on high.
\par }{\QB His place of defense will be the fortress of rocks.
\par }{\QB His bread will be supplied.
\par }{\QB His waters will be sure.
\par }{\BB \par }{\Q \VS{17}Your eyes will see the king in his beauty.
\par }{\QB They will see a distant land.
\par }{\Q \VS{18}Your heart will meditate on the terror.
\par }{\QB Where is he who counted?
\par }{\QB Where is he who weighed?
\par }{\QB Where is he who counted the towers?
\par }{\Q \VS{19}You will no longer see the fierce people,
\par }{\QB a people of a deep speech that you can’t comprehend,
\par }{\QB with a strange language that you can’t understand.
\par }{\Q \VS{20}Look at Zion, the city of our appointed festivals.
\par }{\QB Your eyes will see Jerusalem, a quiet habitation,
\par }{\QB a tent that won’t be removed.
\par }{\Q Its stakes will never be plucked up,
\par }{\QB nor will any of its cords be broken.
\par }{\Q \VS{21}But there Yahweh will be with us in majesty,
\par }{\QB a place of wide rivers and streams,
\par }{\QB in which no galley with oars will go,
\par }{\QB neither will any gallant ship pass by there.
\par }{\Q \VS{22}For Yahweh is our judge.
\par }{\QB Yahweh is our lawgiver.
\par }{\QB Yahweh is our king.
\par }{\QB He will save us.
\par }{\BB \par }{\Q \VS{23}Your rigging is untied.
\par }{\QB They couldn’t strengthen the foot of their mast.
\par }{\QB They couldn’t spread the sail.
\par }{\Q Then the prey of a great plunder was divided.
\par }{\QB The lame took the prey.
\par }{\BB \par }{\Q \VS{24}The inhabitant won’t say, “I am sick.”
\par }{\QB The people who dwell therein will be forgiven their iniquity.

\par }\PoetryChap{34}{\Q \VerseOne{1}Come near, you nations, to hear!
\par }{\QB Listen, you peoples.
\par }{\QB Let the earth and all it contains hear,
\par }{\QB the world, and everything that comes from it.
\par }{\Q \VS{2}For Yahweh is enraged against all the nations,
\par }{\QB and angry with all their armies.
\par }{\Q He has utterly destroyed them.
\par }{\QB He has given them over for slaughter.
\par }{\Q \VS{3}Their slain will also be cast out,
\par }{\QB and the stench of their dead bodies will come up.
\par }{\QB The mountains will melt in their blood.
\par }{\Q \VS{4}All of the army of the sky will be dissolved.
\par }{\QB The sky will be rolled up like a scroll,
\par }{\QB and all its armies will fade away,
\par }{\QB as a leaf fades from off a vine or a fig tree.
\par }{\Q \VS{5}For my sword has drunk its fill in the sky.
\par }{\QB Behold, it will come down on Edom,
\par }{\QB and on the people of my curse, for judgment.
\par }{\Q \VS{6}Yahweh’s sword is filled with blood.
\par }{\QB It is covered with fat, with the blood of lambs and goats,
\par }{\QB with the fat of the kidneys of rams;
\par }{\QB for Yahweh has a sacrifice in Bozrah,
\par }{\QB And a great slaughter in the land of Edom.
\par }{\Q \VS{7}The wild oxen will come down with them,
\par }{\QB and the young bulls with the mighty bulls;
\par }{\QB and their land will be drunken with blood,
\par }{\QB and their dust made greasy with fat.
\par }{\BB \par }{\Q \VS{8}For Yahweh has a day of vengeance,
\par }{\QB a year of recompense for the cause of Zion.
\par }{\Q \VS{9}Its streams will be turned into pitch,
\par }{\QB its dust into sulfur,
\par }{\QB And its land will become burning pitch.
\par }{\Q \VS{10}It won’t be quenched night or day.
\par }{\QB Its smoke will go up forever.
\par }{\QB From generation to generation, it will lie waste.
\par }{\QB No one will pass through it forever and ever.
\par }{\Q \VS{11}But the pelican and the porcupine will possess it.
\par }{\QB The owl and the raven will dwell in it.
\par }{\Q He will stretch the line of confusion over it,
\par }{\QB and the plumb line of emptiness.
\par }{\Q \VS{12}They shall call its nobles to the kingdom, but none shall be there;
\par }{\QB and all its princes shall be nothing.
\par }{\Q \VS{13}Thorns will come up in its palaces,
\par }{\QB nettles and thistles in its fortresses;
\par }{\QB and it will be a habitation of jackals,
\par }{\QB a court for ostriches.
\par }{\Q \VS{14}The wild animals of the desert will meet with the wolves,
\par }{\QB and the wild goat will cry to his fellow.
\par }{\Q Yes, the night creature\FTNT{*}{{\FR 34:14 }
literally, lilith, which could also be a night demon or night monster} shall settle there,
\par }{\QB and shall find herself a place of rest.
\par }{\Q \VS{15}The arrow snake will make her nest there,
\par }{\QB and lay, hatch, and gather under her shade.
\par }{\QB Yes, the kites will be gathered there, every one with her mate.
\par }{\BB \par }{\Q \VS{16}Search in the book of Yahweh, and read:
\par }{\QB not one of these will be missing.
\par }{\QB None will lack her mate.
\par }{\QB For my mouth has commanded,
\par }{\QB and his Spirit has gathered them.
\par }{\Q \VS{17}He has cast the lot for them,
\par }{\QB and his hand has divided it to them with a measuring line.
\par }{\QB They shall possess it forever.
\par }{\QB From generation to generation they will dwell in it.
\par }{\BB \par }
\PoetryChap{35}{\Q \VerseOne{1}The wilderness and the dry land will be glad.
\par }{\QB The desert will rejoice and blossom like a rose.
\par }{\Q \VS{2}It will blossom abundantly,
\par }{\QB and rejoice even with joy and singing.
\par }{\QB Lebanon’s glory will be given to it,
\par }{\QB the excellence of Carmel and Sharon.
\par }{\QB They will see Yahweh’s glory,
\par }{\QB the excellence of our God.
\par }{\BB \par }{\Q \VS{3}Strengthen the weak hands,
\par }{\QB and make the feeble knees firm.
\par }{\Q \VS{4}Tell those who have a fearful heart, “Be strong!
\par }{\QB Don’t be afraid!
\par }{\Q Behold, your God will come with vengeance, God’s retribution.
\par }{\QB He will come and save you.
\par }{\BB \par }{\Q \VS{5}Then the eyes of the blind will be opened,
\par }{\QB and the ears of the deaf will be unstopped.
\par }{\Q \VS{6}Then the lame man will leap like a deer,
\par }{\QB and the tongue of the mute will sing;
\par }{\QB for waters will break out in the wilderness,
\par }{\QB and streams in the desert.
\par }{\Q \VS{7}The burning sand will become a pool,
\par }{\QB and the thirsty ground springs of water.
\par }{\QB Grass with reeds and rushes will be in the habitation of jackals, where they lay.
\par }{\Q \VS{8}A highway will be there, a road,
\par }{\QB and it will be called “The Holy Way”.
\par }{\Q The unclean shall not pass over it,
\par }{\QB but it will be for those who walk in the Way.
\par }{\QB Wicked fools shall not go there.
\par }{\Q \VS{9}No lion will be there,
\par }{\QB nor will any ravenous animal go up on it.
\par }{\QB They will not be found there;
\par }{\QB but the redeemed will walk there.
\par }{\Q \VS{10}Then Yahweh’s ransomed ones will return,
\par }{\QB and come with singing to Zion;
\par }{\QB and everlasting joy will be on their heads.
\par }{\Q They will obtain gladness and joy,
\par }{\QB and sorrow and sighing will flee away.”

\par }\Chap{36}{\PP \VerseOne{1}Now in the fourteenth year of king Hezekiah, Sennacherib king of Assyria attacked all of the fortified cities of Judah and captured them.
\VS{2}The king of Assyria sent Rabshakeh from Lachish to Jerusalem to king Hezekiah with a large army. He stood by the aqueduct from the upper pool in the fuller’s field highway.
\VS{3}Then Eliakim the son of Hilkiah, who was over the household, and Shebna the scribe, and Joah, the son of Asaph, the recorder came out to him.
\par }{\PP \VS{4}Rabshakeh said to them, “Now tell Hezekiah, ‘The great king, the king of Assyria, says, “What confidence is this in which you trust?
\VS{5}I say that your counsel and strength for the war are only vain words. Now in whom do you trust, that you have rebelled against me?
\VS{6}Behold, you trust in the staff of this bruised reed, even in Egypt, which if a man leans on it, it will go into his hand and pierce it. So is Pharaoh king of Egypt to all who trust in him.
\VS{7}But if you tell me, ‘We trust in Yahweh our God,’ isn’t that he whose high places and whose altars Hezekiah has taken away, and has said to Judah and to Jerusalem, ‘You shall worship before this altar?’ ”
\VS{8}Now therefore, please make a pledge to my master the king of Assyria, and I will give you two thousand horses, if you are able on your part to set riders on them.
\VS{9}How then can you turn away the face of one captain of the least of my master’s servants, and put your trust in Egypt for chariots and for horsemen?
\VS{10}Have I come up now without Yahweh against this land to destroy it? Yahweh said to me, “Go up against this land, and destroy it.” ’ ”
\par }{\PP \VS{11}Then Eliakim, Shebna and Joah said to Rabshakeh, “Please speak to your servants in Aramaic, for we understand it. Don’t speak to us in the Jews’ language in the hearing of the people who are on the wall.”
\par }{\PP \VS{12}But Rabshakeh said, “Has my master sent me only to your master and to you, to speak these words, and not to the men who sit on the wall, who will eat their own dung and drink their own urine with you?”
\VS{13}Then Rabshakeh stood, and called out with a loud voice in the Jews’ language, and said, “Hear the words of the great king, the king of Assyria!
\VS{14}The king says, ‘Don’t let Hezekiah deceive you; for he will not be able to deliver you.
\VS{15}Don’t let Hezekiah make you trust in Yahweh, saying, “Yahweh will surely deliver us. This city won’t be given into the hand of the king of Assyria.” ’
\VS{16}Don’t listen to Hezekiah, for the king of Assyria says, ‘Make your peace with me, and come out to me; and each of you eat from his vine, and each one from his fig tree, and each one of you drink the waters of his own cistern;
\VS{17}until I come and take you away to a land like your own land, a land of grain and new wine, a land of bread and vineyards.
\VS{18}Beware lest Hezekiah persuade you, saying, “Yahweh will deliver us.” Have any of the gods of the nations delivered their lands from the hand of the king of Assyria?
\VS{19}Where are the gods of Hamath and Arpad? Where are the gods of Sepharvaim? Have they delivered Samaria from my hand?
\VS{20}Who are they among all the gods of these countries that have delivered their country out of my hand, that Yahweh should deliver Jerusalem out of my hand?’ ”
\par }{\PP \VS{21}But they remained silent, and said nothing in reply, for the king’s commandment was, “Don’t answer him.”
\par }{\PP \VS{22}Then Eliakim the son of Hilkiah, who was over the household, and Shebna the scribe, and Joah, the son of Asaph, the recorder, came to Hezekiah with their clothes torn, and told him the words of Rabshakeh.

\par }\Chap{37}{\PP \VerseOne{1}When king Hezekiah heard it, he tore his clothes, covered himself with sackcloth, and went into Yahweh’s house.
\VS{2}He sent Eliakim, who was over the household, and Shebna the scribe, and the elders of the priests, covered with sackcloth, to Isaiah the prophet, the son of Amoz.
\VS{3}They said to him, “Hezekiah says, ‘Today is a day of trouble, and of rebuke, and of rejection; for the children have come to the birth, and there is no strength to give birth.
\VS{4}It may be Yahweh your God will hear the words of Rabshakeh, whom the king of Assyria his master has sent to defy the living God, and will rebuke the words which Yahweh your God has heard. Therefore lift up your prayer for the remnant that is left.’ ”
\par }{\PP \VS{5}So the servants of king Hezekiah came to Isaiah.
\par }{\PP \VS{6}Isaiah said to them, “Tell your master, ‘Yahweh says, “Don’t be afraid of the words that you have heard, with which the servants of the king of Assyria have blasphemed me.
\VS{7}Behold, I will put a spirit in him and he will hear news, and will return to his own land. I will cause him to fall by the sword in his own land.” ’ ”
\par }{\PP \VS{8}So Rabshakeh returned, and found the king of Assyria warring against Libnah, for he had heard that he was departed from Lachish.
\VS{9}He heard news concerning Tirhakah king of Ethiopia, “He has come out to fight against you.” When he heard it, he sent messengers to Hezekiah, saying,
\VS{10}“Thus you shall speak to Hezekiah king of Judah, saying, ‘Don’t let your God in whom you trust deceive you, saying, “Jerusalem won’t be given into the hand of the king of Assyria.”
\VS{11}Behold, you have heard what the kings of Assyria have done to all lands, by destroying them utterly. Shall you be delivered?
\VS{12}Have the gods of the nations delivered them, which my fathers have destroyed, Gozan, Haran, Rezeph, and the children of Eden who were in Telassar?
\VS{13}Where is the king of Hamath, and the king of Arpad, and the king of the city of Sepharvaim, of Hena, and Ivvah?’ ”
\par }{\PP \VS{14}Hezekiah received the letter from the hand of the messengers and read it. Then Hezekiah went up to Yahweh’s house, and spread it before Yahweh.
\VS{15}Hezekiah prayed to Yahweh, saying,
\VS{16}“Yahweh of Armies, the God of Israel, who is enthroned among the cherubim, you are the God, even you alone, of all the kingdoms of the earth. You have made heaven and earth.
\VS{17}Turn your ear, Yahweh, and hear. Open your eyes, Yahweh, and behold. Hear all of the words of Sennacherib, who has sent to defy the living God.
\VS{18}Truly, Yahweh, the kings of Assyria have destroyed all the countries and their land,
\VS{19}and have cast their gods into the fire; for they were no gods, but the work of men’s hands, wood and stone; therefore they have destroyed them.
\VS{20}Now therefore, Yahweh our God, save us from his hand, that all the kingdoms of the earth may know that you are Yahweh, even you only.”
\par }{\PP \VS{21}Then Isaiah the son of Amoz sent to Hezekiah, saying, “Yahweh, the God of Israel says, ‘Because you have prayed to me against Sennacherib king of Assyria,
\VS{22}this is the word which Yahweh has spoken concerning him. The virgin daughter of Zion has despised you and ridiculed you. The daughter of Jerusalem has shaken her head at you.
\VS{23}Whom have you defied and blasphemed? Against whom have you exalted your voice and lifted up your eyes on high? Against the Holy One of Israel.
\VS{24}By your servants, you have defied the Lord, and have said, “With the multitude of my chariots I have come up to the height of the mountains, to the innermost parts of Lebanon. I will cut down its tall cedars and its choice cypress trees. I will enter into its farthest height, the forest of its fruitful field.
\VS{25}I have dug and drunk water, and with the sole of my feet I will dry up all the rivers of Egypt.”
\par }{\PP \VS{26}“ ‘Have you not heard how I have done it long ago, and formed it in ancient times? Now I have brought it to pass, that it should be yours to destroy fortified cities, turning them into ruinous heaps.
\VS{27}Therefore their inhabitants had little power. They were dismayed and confounded. They were like the grass of the field, and like the green herb, like the grass on the housetops, and like a field before its crop has grown.
\VS{28}But I know your sitting down, your going out, your coming in, and your raging against me.
\VS{29}Because of your raging against me, and because your arrogance has come up into my ears, therefore I will put my hook in your nose and my bridle in your lips, and I will turn you back by the way by which you came.
\par }{\PP \VS{30}“ ‘This shall be the sign to you. You will eat this year that which grows of itself, and in the second year that which springs from it; and in the third year sow and reap and plant vineyards, and eat their fruit.
\VS{31}The remnant that is escaped of the house of Judah will again take root downward, and bear fruit upward.
\VS{32}For out of Jerusalem a remnant will go out, and survivors will escape from Mount Zion. The zeal of Yahweh of Armies will perform this.’
\par }{\PP \VS{33}“Therefore Yahweh says concerning the king of Assyria, ‘He will not come to this city, nor shoot an arrow there, neither will he come before it with shield, nor cast up a mound against it.
\VS{34}He will return the way that he came, and he won’t come to this city,’ says Yahweh.
\VS{35}‘For I will defend this city to save it, for my own sake, and for my servant David’s sake.’ ”
\par }{\PP \VS{36}Then Yahweh’s angel went out and struck one hundred and eighty-five thousand men in the camp of the Assyrians. When men arose early in the morning, behold, these were all dead bodies.
\VS{37}So Sennacherib king of Assyria departed, went away, returned to Nineveh, and stayed there.
\VS{38}As he was worshiping in the house of Nisroch his god, Adrammelech and Sharezer his sons struck him with the sword; and they escaped into the land of Ararat. Esar Haddon his son reigned in his place.

\par }\Chap{38}{\PP \VerseOne{1}In those days Hezekiah was sick and near death. Isaiah the prophet, the son of Amoz, came to him, and said to him, “Yahweh says, ‘Set your house in order, for you will die, and not live.’ ”
\par }{\PP \VS{2}Then Hezekiah turned his face to the wall and prayed to Yahweh,
\VS{3}and said, “Remember now, Yahweh, I beg you, how I have walked before you in truth and with a perfect heart, and have done that which is good in your sight.” Then Hezekiah wept bitterly.
\par }{\PP \VS{4}Then Yahweh’s word came to Isaiah, saying,
\VS{5}“Go, and tell Hezekiah, ‘Yahweh, the God of David your father, says, “I have heard your prayer. I have seen your tears. Behold, I will add fifteen years to your life.
\VS{6}I will deliver you and this city out of the hand of the king of Assyria, and I will defend this city.
\VS{7}This shall be the sign to you from Yahweh, that Yahweh will do this thing that he has spoken.
\VS{8}Behold, I will cause the shadow on the sundial, which has gone down on the sundial of Ahaz with the sun, to return backward ten steps.” ’ ” So the sun returned ten steps on the sundial on which it had gone down.
\par }{\PP \VS{9}The writing of Hezekiah king of Judah, when he had been sick, and had recovered of his sickness.
\par }{\Q \VS{10}I said, “In the middle of my life I go into the gates of Sheol.\FTNT{*}{{\FR 38:10 }
Sheol is the place of the dead.}
\par }{\QB I am deprived of the residue of my years.”
\par }{\Q \VS{11}I said, “I won’t see Yah,
\par }{\QB Yah in the land of the living.
\par }{\QB I will see man no more with the inhabitants of the world.
\par }{\Q \VS{12}My dwelling is removed,
\par }{\QB and is carried away from me like a shepherd’s tent.
\par }{\Q I have rolled up my life like a weaver.
\par }{\QB He will cut me off from the loom.
\par }{\QB From day even to night you will make an end of me.
\par }{\Q \VS{13}I waited patiently until morning.
\par }{\QB He breaks all my bones like a lion.
\par }{\QB From day even to night you will make an end of me.
\par }{\Q \VS{14}I chattered like a swallow or a crane.
\par }{\QB I moaned like a dove.
\par }{\QB My eyes weaken looking upward.
\par }{\QB Lord, I am oppressed.
\par }{\QB Be my security.”
\par }{\Q \VS{15}What will I say?
\par }{\QB He has both spoken to me, and himself has done it.
\par }{\QB I will walk carefully all my years because of the anguish of my soul.
\par }{\Q \VS{16}Lord, men live by these things;
\par }{\QB and my spirit finds life in all of them:
\par }{\QB you restore me, and cause me to live.
\par }{\Q \VS{17}Behold, for peace I had great anguish,
\par }{\QB but you have in love for my soul delivered it from the pit of corruption;
\par }{\QB for you have cast all my sins behind your back.
\par }{\Q \VS{18}For Sheol\FTNT{†}{{\FR 38:18 }
Sheol is the place of the dead.} can’t praise you.
\par }{\QB Death can’t celebrate you.
\par }{\Q Those who go down into the pit can’t hope for your truth.
\par }{\Q \VS{19}The living, the living, he shall praise you, as I do today.
\par }{\QB The father shall make known your truth to the children.
\par }{\Q \VS{20}Yahweh will save me.
\par }{\QB Therefore we will sing my songs with stringed instruments all the days of our life in Yahweh’s house.
\par }{\PP \VS{21}Now Isaiah had said, “Let them take a cake of figs, and lay it for a poultice on the boil, and he shall recover.”
\VS{22}Hezekiah also had said, “What is the sign that I will go up to Yahweh’s house?”

\par }\Chap{39}{\PP \VerseOne{1}At that time, Merodach Baladan the son of Baladan, king of Babylon, sent letters and a present to Hezekiah; for he heard that he had been sick, and had recovered.
\VS{2}Hezekiah was pleased with them, and showed them the house of his precious things, the silver, the gold, the spices, and the precious oil, and all the house of his armor, and all that was found in his treasures. There was nothing in his house, nor in all his dominion, that Hezekiah didn’t show them.
\VS{3}Then Isaiah the prophet came to king Hezekiah, and asked him, “What did these men say? From where did they come to you?”
\par }{\PP Hezekiah said, “They have come from a country far from me, even from Babylon.”
\par }{\PP \VS{4}Then he asked, “What have they seen in your house?”
\par }{\PP Hezekiah answered, “They have seen all that is in my house. There is nothing among my treasures that I have not shown them.”
\par }{\PP \VS{5}Then Isaiah said to Hezekiah, “Hear the word of Yahweh of Armies:
\VS{6}‘Behold, the days are coming when all that is in your house, and that which your fathers have stored up until today, will be carried to Babylon. Nothing will be left,’ says Yahweh.
\VS{7}‘They will take away your sons who will issue from you, whom you shall father, and they will be eunuchs in the king of Babylon’s palace.’ ”
\par }{\PP \VS{8}Then Hezekiah said to Isaiah, “Yahweh’s word which you have spoken is good.” He said moreover, “For there will be peace and truth in my days.”

\par }\Chap{40}{\PP \VerseOne{1}“Comfort, comfort my people,” says your God.
\VS{2}“Speak comfortably to Jerusalem; and call out to her that her warfare is accomplished, that her iniquity is pardoned, that she has received of Yahweh’s hand double for all her sins.”
\par }{\Q \VS{3}The voice of one who calls out,
\par }{\QB “Prepare the way of Yahweh in the wilderness!
\par }{\QB Make a level highway in the desert for our God.
\par }{\Q \VS{4}Every valley shall be exalted,
\par }{\QB and every mountain and hill shall be made low.
\par }{\QB The uneven shall be made level,
\par }{\QB and the rough places a plain.
\par }{\Q \VS{5}Yahweh’s glory shall be revealed,
\par }{\QB and all flesh shall see it together;
\par }{\QB for the mouth of Yahweh has spoken it.”
\par }{\BB \par }{\Q \VS{6}The voice of one saying, “Cry!”
\par }{\QB One said, “What shall I cry?”
\par }{\Q “All flesh is like grass,
\par }{\QB and all its glory is like the flower of the field.
\par }{\Q \VS{7}The grass withers,
\par }{\QB the flower fades,
\par }{\QB because Yahweh’s breath blows on it.
\par }{\QB Surely the people are like grass.
\par }{\Q \VS{8}The grass withers,
\par }{\QB the flower fades;
\par }{\QB but the word of our God stands forever.”
\par }{\BB \par }{\Q \VS{9}You who tell good news to Zion, go up on a high mountain.
\par }{\QB You who tell good news to Jerusalem, lift up your voice with strength!
\par }{\QB Lift it up! Don’t be afraid!
\par }{\QB Say to the cities of Judah, “Behold, your God!”
\par }{\Q \VS{10}Behold, the Lord Yahweh will come as a mighty one,
\par }{\QB and his arm will rule for him.
\par }{\QB Behold, his reward is with him,
\par }{\QB and his recompense before him.
\par }{\Q \VS{11}He will feed his flock like a shepherd.
\par }{\QB He will gather the lambs in his arm,
\par }{\QB and carry them in his bosom.
\par }{\QB He will gently lead those who have their young.
\par }{\BB \par }{\Q \VS{12}Who has measured the waters in the hollow of his hand,
\par }{\QB and marked off the sky with his span,
\par }{\QB and calculated the dust of the earth in a measuring basket,
\par }{\QB and weighed the mountains in scales,
\par }{\QB and the hills in a balance?
\par }{\Q \VS{13}Who has directed Yahweh’s Spirit,
\par }{\QB or has taught him as his counselor?
\par }{\Q \VS{14}Who did he take counsel with,
\par }{\QB and who instructed him,
\par }{\QB and taught him in the path of justice,
\par }{\QB and taught him knowledge,
\par }{\QB and showed him the way of understanding?
\par }{\Q \VS{15}Behold, the nations are like a drop in a bucket,
\par }{\QB and are regarded as a speck of dust on a balance.
\par }{\QB Behold, he lifts up the islands like a very little thing.
\par }{\Q \VS{16}Lebanon is not sufficient to burn,
\par }{\QB nor its animals sufficient for a burnt offering.
\par }{\Q \VS{17}All the nations are like nothing before him.
\par }{\QB They are regarded by him as less than nothing, and vanity.
\par }{\BB \par }{\Q \VS{18}To whom then will you liken God?
\par }{\QB Or what likeness will you compare to him?
\par }{\Q \VS{19}A workman has cast an image,
\par }{\QB and the goldsmith overlays it with gold,
\par }{\QB and casts silver chains for it.
\par }{\Q \VS{20}He who is too impoverished for such an offering chooses a tree that will not rot.
\par }{\QB He seeks a skillful workman to set up a carved image for him that will not be moved.
\par }{\BB \par }{\Q \VS{21}Haven’t you known?
\par }{\QB Haven’t you heard?
\par }{\QB Haven’t you been told from the beginning?
\par }{\QB Haven’t you understood from the foundations of the earth?
\par }{\Q \VS{22}It is he who sits above the circle of the earth,
\par }{\QB and its inhabitants are like grasshoppers;
\par }{\QB who stretches out the heavens like a curtain,
\par }{\QB and spreads them out like a tent to dwell in,
\par }{\QB \VS{23}who brings princes to nothing,
\par }{\QB who makes the judges of the earth meaningless.
\par }{\Q \VS{24}They are planted scarcely.
\par }{\QB They are sown scarcely.
\par }{\QB Their stock has scarcely taken root in the ground.
\par }{\QB He merely blows on them, and they wither,
\par }{\QB and the whirlwind takes them away as stubble.
\par }{\BB \par }{\Q \VS{25}“To whom then will you liken me?
\par }{\QB Who is my equal?” says the Holy One.
\par }{\Q \VS{26}Lift up your eyes on high,
\par }{\QB and see who has created these,
\par }{\QB who brings out their army by number.
\par }{\QB He calls them all by name.
\par }{\QB by the greatness of his might,
\par }{\QB and because he is strong in power,
\par }{\QB not one is lacking.
\par }{\BB \par }{\Q \VS{27}Why do you say, Jacob,
\par }{\QB and speak, Israel,
\par }{\QB “My way is hidden from Yahweh,
\par }{\QB and the justice due me is disregarded by my God?”
\par }{\Q \VS{28}Haven’t you known?
\par }{\QB Haven’t you heard?
\par }{\QB The everlasting God, Yahweh,
\par }{\QB the Creator of the ends of the earth, doesn’t faint.
\par }{\QB He isn’t weary.
\par }{\QB His understanding is unsearchable.
\par }{\Q \VS{29}He gives power to the weak.
\par }{\QB He increases the strength of him who has no might.
\par }{\Q \VS{30}Even the youths faint and get weary,
\par }{\QB and the young men utterly fall;
\par }{\QB \VS{31}but those who wait for Yahweh will renew their strength.
\par }{\QB They will mount up with wings like eagles.
\par }{\QB They will run, and not be weary.
\par }{\QB They will walk, and not faint.

\par }\PoetryChap{41}{\Q \VerseOne{1}“Keep silent before me, islands,
\par }{\QB and let the peoples renew their strength.
\par }{\Q Let them come near,
\par }{\QB then let them speak.
\par }{\QB Let’s meet together for judgment.
\par }{\Q \VS{2}Who has raised up one from the east?
\par }{\QB Who called him to his foot in righteousness?
\par }{\QB He hands over nations to him
\par }{\QB and makes him rule over kings.
\par }{\QB He gives them like the dust to his sword,
\par }{\QB like the driven stubble to his bow.
\par }{\Q \VS{3}He pursues them
\par }{\QB and passes by safely,
\par }{\QB even by a way that he had not gone with his feet.
\par }{\Q \VS{4}Who has worked and done it,
\par }{\QB calling the generations from the beginning?
\par }{\QB I, Yahweh, the first, and with the last, I am he.”
\par }{\BB \par }{\Q \VS{5}The islands have seen, and fear.
\par }{\QB The ends of the earth tremble.
\par }{\QB They approach, and come.
\par }{\Q \VS{6}Everyone helps his neighbor.
\par }{\QB They say to their brothers, “Be strong!”
\par }{\Q \VS{7}So the carpenter encourages the goldsmith.
\par }{\QB He who smoothes with the hammer encourages him who strikes the anvil,
\par }{\QB saying of the soldering, “It is good;”
\par }{\QB and he fastens it with nails, that it might not totter.
\par }{\BB \par }{\Q \VS{8}“But you, Israel, my servant,
\par }{\QB Jacob whom I have chosen,
\par }{\QB the offspring of Abraham my friend,
\par }{\QB \VS{9}You whom I have taken hold of from the ends of the earth,
\par }{\QB and called from its corners,
\par }{\QB and said to you, ‘You are my servant, I have chosen you and have not cast you away.’
\par }{\Q \VS{10}Don’t you be afraid, for I am with you.
\par }{\QB Don’t be dismayed, for I am your God.
\par }{\QB I will strengthen you.
\par }{\QB Yes, I will help you.
\par }{\QB Yes, I will uphold you with the right hand of my righteousness.
\par }{\Q \VS{11}Behold, all those who are incensed against you will be disappointed and confounded.
\par }{\QB Those who strive with you will be like nothing, and shall perish.
\par }{\Q \VS{12}You will seek them, and won’t find them,
\par }{\QB even those who contend with you.
\par }{\QB Those who war against you will be as nothing,
\par }{\QB as a non-existent thing.
\par }{\Q \VS{13}For I, Yahweh your God, will hold your right hand,
\par }{\QB saying to you, ‘Don’t be afraid.
\par }{\QB I will help you.’
\par }{\Q \VS{14}Don’t be afraid, you worm Jacob,
\par }{\QB and you men of Israel.
\par }{\QB I will help you,” says Yahweh.
\par }{\QB “Your Redeemer is the Holy One of Israel.
\par }{\Q \VS{15}Behold, I have made you into a new sharp threshing instrument with teeth.
\par }{\QB You will thresh the mountains,
\par }{\QB and beat them small,
\par }{\QB and will make the hills like chaff.
\par }{\Q \VS{16}You will winnow them,
\par }{\QB and the wind will carry them away,
\par }{\QB and the whirlwind will scatter them.
\par }{\Q You will rejoice in Yahweh.
\par }{\QB You will glory in the Holy One of Israel.
\par }{\BB \par }{\Q \VS{17}The poor and needy seek water, and there is none.
\par }{\QB Their tongue fails for thirst.
\par }{\Q I, Yahweh, will answer them.
\par }{\QB I, the God of Israel, will not forsake them.
\par }{\Q \VS{18}I will open rivers on the bare heights,
\par }{\QB and springs in the middle of the valleys.
\par }{\QB I will make the wilderness a pool of water,
\par }{\QB and the dry land springs of water.
\par }{\Q \VS{19}I will put cedar, acacia, myrtle, and oil trees in the wilderness.
\par }{\QB I will set cypress trees, pine, and box trees together in the desert;
\par }{\QB \VS{20}that they may see, know, consider, and understand together,
\par }{\QB that Yahweh’s hand has done this,
\par }{\QB and the Holy One of Israel has created it.
\par }{\BB \par }{\Q \VS{21}Produce your cause,” says Yahweh.
\par }{\QB “Bring out your strong reasons!” says the King of Jacob.
\par }{\Q \VS{22}“Let them announce and declare to us what will happen!
\par }{\QB Declare the former things, what they are,
\par }{\QB that we may consider them, and know the latter end of them;
\par }{\QB or show us things to come.
\par }{\Q \VS{23}Declare the things that are to come hereafter,
\par }{\QB that we may know that you are gods.
\par }{\Q Yes, do good, or do evil,
\par }{\QB that we may be dismayed,
\par }{\QB and see it together.
\par }{\Q \VS{24}Behold, you are nothing,
\par }{\QB and your work is nothing.
\par }{\QB He who chooses you is an abomination.
\par }{\BB \par }{\Q \VS{25}“I have raised up one from the north, and he has come,
\par }{\QB from the rising of the sun, one who calls on my name,
\par }{\QB and he shall come on rulers as on mortar,
\par }{\QB and as the potter treads clay.
\par }{\Q \VS{26}Who has declared it from the beginning, that we may know?
\par }{\QB and before, that we may say, ‘He is right?’
\par }{\Q Surely, there is no one who declares.
\par }{\QB Surely, there is no one who shows.
\par }{\QB Surely, there is no one who hears your words.
\par }{\Q \VS{27}I am the first to say to Zion, ‘Behold, look at them;’
\par }{\QB and I will give one who brings good news to Jerusalem.
\par }{\Q \VS{28}When I look, there is no man,
\par }{\QB even among them there is no counselor who, when I ask of them, can answer a word.
\par }{\Q \VS{29}Behold, all of their deeds are vanity and nothing.
\par }{\QB Their molten images are wind and confusion.

\par }\PoetryChap{42}{\Q \VerseOne{1}“Behold, my servant, whom I uphold,
\par }{\QB my chosen, in whom my soul delights:
\par }{\QB I have put my Spirit on him.
\par }{\QB He will bring justice to the nations.
\par }{\Q \VS{2}He will not shout,
\par }{\QB nor raise his voice,
\par }{\QB nor cause it to be heard in the street.
\par }{\Q \VS{3}He won’t break a bruised reed.
\par }{\QB He won’t quench a dimly burning wick.
\par }{\QB He will faithfully bring justice.
\par }{\Q \VS{4}He will not fail nor be discouraged,
\par }{\QB until he has set justice in the earth,
\par }{\QB and the islands wait for his law.”
\par }{\BB \par }{\Q \VS{5}God Yahweh,
\par }{\QB he who created the heavens and stretched them out,
\par }{\QB he who spread out the earth and that which comes out of it,
\par }{\QB he who gives breath to its people and spirit to those who walk in it, says:
\par }{\Q \VS{6}“I, Yahweh, have called you in righteousness.
\par }{\QB I will hold your hand.
\par }{\QB I will keep you,
\par }{\QB and make you a covenant for the people,
\par }{\QB as a light for the nations,
\par }{\QB \VS{7}to open the blind eyes,
\par }{\QB to bring the prisoners out of the dungeon,
\par }{\QB and those who sit in darkness out of the prison.
\par }{\BB \par }{\Q \VS{8}“I am Yahweh.
\par }{\QB That is my name.
\par }{\QB I will not give my glory to another,
\par }{\QB nor my praise to engraved images.
\par }{\Q \VS{9}Behold, the former things have happened
\par }{\QB and I declare new things.
\par }{\QB I tell you about them before they come up.”
\par }{\BB \par }{\Q \VS{10}Sing to Yahweh a new song,
\par }{\QB and his praise from the end of the earth,
\par }{\QB you who go down to the sea,
\par }{\QB and all that is therein,
\par }{\QB the islands and their inhabitants.
\par }{\Q \VS{11}Let the wilderness and its cities raise their voices,
\par }{\QB with the villages that Kedar inhabits.
\par }{\QB Let the inhabitants of Sela sing.
\par }{\QB Let them shout from the top of the mountains!
\par }{\Q \VS{12}Let them give glory to Yahweh,
\par }{\QB and declare his praise in the islands.
\par }{\Q \VS{13}Yahweh will go out like a mighty man.
\par }{\QB He will stir up zeal like a man of war.
\par }{\QB He will raise a war cry.
\par }{\QB Yes, he will shout aloud.
\par }{\QB He will triumph over his enemies.
\par }{\BB \par }{\Q \VS{14}“I have been silent a long time.
\par }{\QB I have been quiet and restrained myself.
\par }{\QB Now I will cry out like a travailing woman. I will both gasp and pant.
\par }{\Q \VS{15}I will destroy mountains and hills,
\par }{\QB and dry up all their herbs.
\par }{\QB I will make the rivers islands,
\par }{\QB and will dry up the pools.
\par }{\Q \VS{16}I will bring the blind by a way that they don’t know.
\par }{\QB I will lead them in paths that they don’t know.
\par }{\QB I will make darkness light before them,
\par }{\QB and crooked places straight.
\par }{\QB I will do these things,
\par }{\QB and I will not forsake them.
\par }{\BB \par }{\Q \VS{17}“Those who trust in engraved images,
\par }{\QB who tell molten images,
\par }{\QB ‘You are our gods,’
\par }{\QB will be turned back.
\par }{\QB They will be utterly disappointed.
\par }{\BB \par }{\Q \VS{18}“Hear, you deaf,
\par }{\QB and look, you blind,
\par }{\QB that you may see.
\par }{\Q \VS{19}Who is blind, but my servant?
\par }{\QB Or who is as deaf as my messenger whom I send?
\par }{\Q Who is as blind as he who is at peace,
\par }{\QB and as blind as Yahweh’s servant?
\par }{\Q \VS{20}You see many things, but don’t observe.
\par }{\QB His ears are open, but he doesn’t listen.
\par }{\Q \VS{21}It pleased Yahweh, for his righteousness’ sake, to magnify the law
\par }{\QB and make it honorable.
\par }{\Q \VS{22}But this is a robbed and plundered people.
\par }{\QB All of them are snared in holes,
\par }{\QB and they are hidden in prisons.
\par }{\Q They have become captives, and no one delivers,
\par }{\QB and a plunder, and no one says, ‘Restore them!’
\par }{\BB \par }{\Q \VS{23}Who is there among you who will give ear to this?
\par }{\QB Who will listen and hear for the time to come?
\par }{\Q \VS{24}Who gave Jacob as plunder,
\par }{\QB and Israel to the robbers?
\par }{\QB Didn’t Yahweh, he against whom we have sinned?
\par }{\QB For they would not walk in his ways,
\par }{\QB and they disobeyed his law.
\par }{\Q \VS{25}Therefore he poured the fierceness of his anger on him,
\par }{\QB and the strength of battle.
\par }{\Q It set him on fire all around, but he didn’t know.
\par }{\QB It burned him, but he didn’t take it to heart.”

\par }\PoetryChap{43}{\Q \VerseOne{1}But now Yahweh who created you, Jacob,
\par }{\QB and he who formed you, Israel, says:
\par }{\Q “Don’t be afraid, for I have redeemed you.
\par }{\QB I have called you by your name.
\par }{\QB You are mine.
\par }{\Q \VS{2}When you pass through the waters, I will be with you,
\par }{\QB and through the rivers, they will not overflow you.
\par }{\Q When you walk through the fire, you will not be burned,
\par }{\QB and flame will not scorch you.
\par }{\Q \VS{3}For I am Yahweh your God,
\par }{\QB the Holy One of Israel,
\par }{\QB your Savior.
\par }{\Q I have given Egypt as your ransom,
\par }{\QB Ethiopia and Seba in your place.
\par }{\Q \VS{4}Since you have been precious and honored in my sight,
\par }{\QB and I have loved you,
\par }{\QB therefore I will give people in your place,
\par }{\QB and nations instead of your life.
\par }{\Q \VS{5}Don’t be afraid, for I am with you.
\par }{\QB I will bring your offspring from the east,
\par }{\QB and gather you from the west.
\par }{\Q \VS{6}I will tell the north, ‘Give them up!’
\par }{\QB and tell the south, ‘Don’t hold them back!
\par }{\QB Bring my sons from far away,
\par }{\QB and my daughters from the ends of the earth—
\par }{\Q \VS{7}everyone who is called by my name,
\par }{\QB and whom I have created for my glory,
\par }{\QB whom I have formed,
\par }{\QB yes, whom I have made.’ ”
\par }{\BB \par }{\Q \VS{8}Bring out the blind people who have eyes,
\par }{\QB and the deaf who have ears.
\par }{\Q \VS{9}Let all the nations be gathered together,
\par }{\QB and let the peoples be assembled.
\par }{\Q Who among them can declare this,
\par }{\QB and show us former things?
\par }{\Q Let them bring their witnesses, that they may be justified,
\par }{\QB or let them hear, and say, “That is true.”
\par }{\BB \par }{\Q \VS{10}“You are my witnesses,” says Yahweh,
\par }{\QB “With my servant whom I have chosen;
\par }{\QB that you may know and believe me,
\par }{\QB and understand that I am he.
\par }{\Q Before me there was no God formed,
\par }{\QB neither will there be after me.
\par }{\Q \VS{11}I myself am Yahweh.
\par }{\QB Besides me, there is no savior.
\par }{\Q \VS{12}I have declared, I have saved, and I have shown,
\par }{\QB and there was no strange god among you.
\par }{\Q Therefore you are my witnesses”,
\par }{\QB says Yahweh, “and I am God.
\par }{\Q \VS{13}Yes, since the day was, I am he.
\par }{\QB There is no one who can deliver out of my hand.
\par }{\QB I will work, and who can hinder it?”
\par }{\PP \VS{14}Yahweh, your Redeemer, the Holy One of Israel says: “For your sake, I have sent to Babylon, and I will bring all of them down as fugitives, even the Chaldeans, in the ships of their rejoicing.
\VS{15}I am Yahweh, your Holy One, the Creator of Israel, your King.”
\par }{\PP \VS{16}Yahweh, who makes a way in the sea,
\par }{\QB and a path in the mighty waters,
\par }{\Q \VS{17}who brings out the chariot and horse,
\par }{\QB the army and the mighty man
\par }{\QB (they lie down together, they shall not rise;
\par }{\QB they are extinct, they are quenched like a wick) says:
\par }{\Q \VS{18}“Don’t remember the former things,
\par }{\QB and don’t consider the things of old.
\par }{\Q \VS{19}Behold, I will do a new thing.
\par }{\QB It springs out now.
\par }{\QB Don’t you know it?
\par }{\Q I will even make a way in the wilderness,
\par }{\QB and rivers in the desert.
\par }{\Q \VS{20}The animals of the field, the jackals and the ostriches, shall honor me,
\par }{\QB because I give water in the wilderness and rivers in the desert,
\par }{\QB to give drink to my people, my chosen,
\par }{\QB \VS{21}the people which I formed for myself,
\par }{\QB that they might declare my praise.
\par }{\BB \par }{\Q \VS{22}Yet you have not called on me, Jacob;
\par }{\QB but you have been weary of me, Israel.
\par }{\Q \VS{23}You have not brought me any of your sheep for burnt offerings,
\par }{\QB neither have you honored me with your sacrifices.
\par }{\Q I have not burdened you with offerings,
\par }{\QB nor wearied you with frankincense.
\par }{\Q \VS{24}You have bought me no sweet cane with money,
\par }{\QB nor have you filled me with the fat of your sacrifices,
\par }{\Q but you have burdened me with your sins.
\par }{\QB You have wearied me with your iniquities.
\par }{\BB \par }{\Q \VS{25}I, even I, am he who blots out your transgressions for my own sake;
\par }{\QB and I will not remember your sins.
\par }{\Q \VS{26}Put me in remembrance.
\par }{\QB Let us plead together.
\par }{\Q Declare your case,
\par }{\QB that you may be justified.
\par }{\Q \VS{27}Your first father sinned,
\par }{\QB and your teachers have transgressed against me.
\par }{\Q \VS{28}Therefore I will profane the princes of the sanctuary;
\par }{\QB and I will make Jacob a curse,
\par }{\QB and Israel an insult.”

\par }\PoetryChap{44}{\Q \VerseOne{1}Yet listen now, Jacob my servant,
\par }{\QB and Israel, whom I have chosen.
\par }{\Q \VS{2}This is what Yahweh who made you,
\par }{\QB and formed you from the womb,
\par }{\QB who will help you says:
\par }{\Q “Don’t be afraid, Jacob my servant;
\par }{\QB and you, Jeshurun, whom I have chosen.
\par }{\Q \VS{3}For I will pour water on him who is thirsty,
\par }{\QB and streams on the dry ground.
\par }{\Q I will pour my Spirit on your descendants,
\par }{\QB and my blessing on your offspring:
\par }{\Q \VS{4}and they will spring up among the grass,
\par }{\QB as willows by the watercourses.
\par }{\Q \VS{5}One will say, ‘I am Yahweh’s;’
\par }{\QB and another will be called by the name of Jacob;
\par }{\QB and another will write with his hand ‘to Yahweh,’
\par }{\QB and honor the name of Israel.”
\par }{\BB \par }{\Q \VS{6}This is what Yahweh, the King of Israel,
\par }{\QB and his Redeemer, Yahweh of Armies, says:
\par }{\Q “I am the first, and I am the last;
\par }{\QB and besides me there is no God.
\par }{\Q \VS{7}Who is like me?
\par }{\QB Who will call,
\par }{\QB and will declare it,
\par }{\QB and set it in order for me,
\par }{\QB since I established the ancient people?
\par }{\Q Let them declare the things that are coming,
\par }{\QB and that will happen.
\par }{\Q \VS{8}Don’t fear,
\par }{\QB neither be afraid.
\par }{\Q Haven’t I declared it to you long ago,
\par }{\QB and shown it?
\par }{\Q You are my witnesses.
\par }{\QB Is there a God besides me?
\par }{\Q Indeed, there is not.
\par }{\QB I don’t know any other Rock.”
\par }{\BB \par }{\Q \VS{9}Everyone who makes a carved image is vain.
\par }{\QB The things that they delight in will not profit.
\par }{\QB Their own witnesses don’t see, nor know, that they may be disappointed.
\par }{\Q \VS{10}Who has fashioned a god,
\par }{\QB or molds an image that is profitable for nothing?
\par }{\Q \VS{11}Behold, all his fellows will be disappointed;
\par }{\QB and the workmen are mere men.
\par }{\Q Let them all be gathered together.
\par }{\QB Let them stand up.
\par }{\QB They will fear.
\par }{\QB They will be put to shame together.
\par }{\BB \par }{\Q \VS{12}The blacksmith takes an ax,
\par }{\QB works in the coals,
\par }{\QB fashions it with hammers,
\par }{\QB and works it with his strong arm.
\par }{\Q He is hungry,
\par }{\QB and his strength fails;
\par }{\Q he drinks no water,
\par }{\QB and is faint.
\par }{\Q \VS{13}The carpenter stretches out a line.
\par }{\QB He marks it out with a pencil.
\par }{\QB He shapes it with planes.
\par }{\QB He marks it out with compasses,
\par }{\QB and shapes it like the figure of a man,
\par }{\QB with the beauty of a man,
\par }{\QB to reside in a house.
\par }{\Q \VS{14}He cuts down cedars for himself,
\par }{\QB and takes the cypress and the oak,
\par }{\QB and strengthens for himself one among the trees of the forest.
\par }{\Q He plants a cypress tree,
\par }{\QB and the rain nourishes it.
\par }{\Q \VS{15}Then it will be for a man to burn;
\par }{\QB and he takes some of it, and warms himself.
\par }{\QB Yes, he burns it, and bakes bread.
\par }{\Q Yes, he makes a god, and worships it;
\par }{\QB he makes it a carved image, and falls down to it.
\par }{\Q \VS{16}He burns part of it in the fire.
\par }{\QB With part of it, he eats meat.
\par }{\QB He roasts a roast, and is satisfied.
\par }{\Q Yes, he warms himself,
\par }{\QB and says, “Aha! I am warm. I have seen the fire.”
\par }{\Q \VS{17}The rest of it he makes into a god,
\par }{\QB even his engraved image.
\par }{\Q He bows down to it and worships,
\par }{\QB and prays to it, and says, “Deliver me; for you are my god!”
\par }{\BB \par }{\Q \VS{18}They don’t know, neither do they consider:
\par }{\QB for he has shut their eyes, that they can’t see;
\par }{\QB and their hearts, that they can’t understand.
\par }{\Q \VS{19}No one thinks,
\par }{\QB neither is there knowledge nor understanding to say,
\par }{\QB “I have burned part of it in the fire.
\par }{\QB Yes, I have also baked bread on its coals.
\par }{\QB I have roasted meat and eaten it.
\par }{\QB Shall I make the rest of it into an abomination?
\par }{\QB Shall I bow down to a tree trunk?”
\par }{\Q \VS{20}He feeds on ashes.
\par }{\QB A deceived heart has turned him aside;
\par }{\QB and he can’t deliver his soul,
\par }{\QB nor say, “Isn’t there a lie in my right hand?”
\par }{\BB \par }{\Q \VS{21}Remember these things, Jacob and Israel;
\par }{\QB for you are my servant.
\par }{\QB I have formed you.
\par }{\QB You are my servant.
\par }{\QB Israel, you will not be forgotten by me.
\par }{\Q \VS{22}I have blotted out, as a thick cloud, your transgressions,
\par }{\QB and, as a cloud, your sins.
\par }{\QB Return to me, for I have redeemed you.
\par }{\BB \par }{\Q \VS{23}Sing, you heavens, for Yahweh has done it!
\par }{\QB Shout, you lower parts of the earth!
\par }{\QB Break out into singing, you mountains, O forest, all of your trees,
\par }{\QB for Yahweh has redeemed Jacob,
\par }{\QB and will glorify himself in Israel.
\par }{\BB \par }{\Q \VS{24}Yahweh, your Redeemer,
\par }{\QB and he who formed you from the womb says:
\par }{\Q “I am Yahweh, who makes all things;
\par }{\QB who alone stretches out the heavens;
\par }{\QB who spreads out the earth by myself;
\par }{\Q \VS{25}who frustrates the signs of the liars,
\par }{\QB and makes diviners mad;
\par }{\Q who turns wise men backward,
\par }{\QB and makes their knowledge foolish;
\par }{\Q \VS{26}who confirms the word of his servant,
\par }{\QB and performs the counsel of his messengers;
\par }{\Q who says of Jerusalem, ‘She will be inhabited;’
\par }{\QB and of the cities of Judah, ‘They will be built,’
\par }{\QB and ‘I will raise up its waste places;’
\par }{\Q \VS{27}who says to the deep, ‘Be dry,’
\par }{\QB and ‘I will dry up your rivers;’
\par }{\Q \VS{28}Who says of Cyrus, ‘He is my shepherd, and shall perform all my pleasure,’
\par }{\QB even saying of Jerusalem, ‘She will be built;’
\par }{\QB and of the temple, ‘Your foundation will be laid.’ ”

\par }\Chap{45}{\PP \VerseOne{1}Yahweh says to his anointed, to Cyrus, whose right hand I have held, to subdue nations before him, and strip kings of their armor; to open the doors before him, and the gates shall not be shut:
\par }{\Q \VS{2}“I will go before you
\par }{\QB and make the rough places smooth.
\par }{\Q I will break the doors of bronze in pieces
\par }{\QB and cut apart the bars of iron.
\par }{\Q \VS{3}I will give you the treasures of darkness
\par }{\QB and hidden riches of secret places,
\par }{\Q that you may know that it is I, Yahweh, who calls you by your name,
\par }{\QB even the God of Israel.
\par }{\Q \VS{4}For Jacob my servant’s sake,
\par }{\QB and Israel my chosen,
\par }{\Q I have called you by your name.
\par }{\QB I have given you a title,
\par }{\QB though you have not known me.
\par }{\Q \VS{5}I am Yahweh, and there is no one else.
\par }{\QB Besides me, there is no God.
\par }{\Q I will strengthen\FTNT{*}{{\FR 45:5 }
or, equip} you,
\par }{\QB though you have not known me,
\par }{\Q \VS{6}that they may know from the rising of the sun,
\par }{\QB and from the west,
\par }{\Q that there is no one besides me.
\par }{\QB I am Yahweh, and there is no one else.
\par }{\Q \VS{7}I form the light
\par }{\QB and create darkness.
\par }{\Q I make peace
\par }{\QB and create calamity.
\par }{\Q I am Yahweh,
\par }{\QB who does all these things.
\par }{\BB \par }{\Q \VS{8}Rain, you heavens, from above,
\par }{\QB and let the skies pour down righteousness.
\par }{\Q Let the earth open, that it may produce salvation,
\par }{\QB and let it cause righteousness to spring up with it.
\par }{\Q I, Yahweh, have created it.
\par }{\BB \par }{\Q \VS{9}Woe to him who strives with his Maker—
\par }{\QB a clay pot among the clay pots of the earth!
\par }{\Q Shall the clay ask him who fashions it, ‘What are you making?’
\par }{\QB or your work, ‘He has no hands?’
\par }{\Q \VS{10}Woe to him who says to a father, ‘What have you become the father of?’
\par }{\QB or to a mother, ‘What have you given birth to?’ ”
\par }{\BB \par }{\Q \VS{11}Yahweh, the Holy One of Israel
\par }{\QB and his Maker says:
\par }{\Q “You ask me about the things that are to come, concerning my sons,
\par }{\QB and you command me concerning the work of my hands!
\par }{\Q \VS{12}I have made the earth, and created man on it.
\par }{\QB I, even my hands, have stretched out the heavens.
\par }{\QB I have commanded all their army.
\par }{\Q \VS{13}I have raised him up in righteousness,
\par }{\QB and I will make all his ways straight.
\par }{\Q He shall build my city,
\par }{\QB and he shall let my exiles go free,
\par }{\QB not for price nor reward,” says Yahweh of Armies.
\par }{\BB \par }{\Q \VS{14}Yahweh says: “The labor of Egypt,
\par }{\QB and the merchandise of Ethiopia,
\par }{\QB and the Sabeans, men of stature, will come over to you,
\par }{\QB and they will be yours.
\par }{\Q They will go after you.
\par }{\QB They shall come over in chains.
\par }{\QB They will bow down to you.
\par }{\Q They will make supplication to you:
\par }{\QB ‘Surely God is in you; and there is no one else.
\par }{\QB There is no other god.
\par }{\Q \VS{15}Most certainly you are a God who has hidden yourself,
\par }{\QB God of Israel, the Savior.’ ”
\par }{\Q \VS{16}They will be disappointed,
\par }{\QB yes, confounded, all of them.
\par }{\QB Those who are makers of idols will go into confusion together.
\par }{\Q \VS{17}Israel will be saved by Yahweh with an everlasting salvation.
\par }{\QB You will not be disappointed nor confounded to ages everlasting.
\par }{\BB \par }{\Q \VS{18}For Yahweh who created the heavens,
\par }{\QB the God who formed the earth and made it,
\par }{\QB who established it and didn’t create it a waste,
\par }{\QB who formed it to be inhabited says:
\par }{\Q “I am Yahweh.
\par }{\QB There is no other.
\par }{\Q \VS{19}I have not spoken in secret,
\par }{\QB in a place of the land of darkness.
\par }{\Q I didn’t say to the offspring of Jacob, ‘Seek me in vain.’
\par }{\QB I, Yahweh, speak righteousness.
\par }{\QB I declare things that are right.
\par }{\BB \par }{\Q \VS{20}“Assemble yourselves and come.
\par }{\QB Draw near together, you who have escaped from the nations.
\par }{\Q Those have no knowledge who carry the wood of their engraved image,
\par }{\QB and pray to a god that can’t save.
\par }{\Q \VS{21}Declare and present it.
\par }{\QB Yes, let them take counsel together.
\par }{\Q Who has shown this from ancient time?
\par }{\QB Who has declared it of old?
\par }{\QB Haven’t I, Yahweh?
\par }{\Q There is no other God besides me, a just God and a Savior.
\par }{\QB There is no one besides me.
\par }{\BB \par }{\Q \VS{22}“Look to me, and be saved, all the ends of the earth;
\par }{\QB for I am God, and there is no other.
\par }{\Q \VS{23}I have sworn by myself.
\par }{\QB The word has gone out of my mouth in righteousness, and will not be revoked,
\par }{\Q that to me every knee shall bow,
\par }{\QB every tongue shall take an oath.
\par }{\Q \VS{24}They will say of me,
\par }{\QB ‘There is righteousness and strength only in Yahweh.’ ”
\par }{\Q Even to him will men come.
\par }{\QB All those who raged against him will be disappointed.
\par }{\Q \VS{25}All the offspring of Israel will be justified in Yahweh,
\par }{\QB and will rejoice!

\par }\PoetryChap{46}{\Q \VerseOne{1}Bel bows down.
\par }{\QB Nebo stoops.
\par }{\Q Their idols are carried by animals,
\par }{\QB and on the livestock.
\par }{\Q The things that you carried around are heavy loads,
\par }{\QB a burden for the weary.
\par }{\Q \VS{2}They stoop and they bow down together.
\par }{\QB They could not deliver the burden,
\par }{\QB but they have gone into captivity.
\par }{\BB \par }{\Q \VS{3}“Listen to me, house of Jacob,
\par }{\QB and all the remnant of the house of Israel,
\par }{\QB that have been carried from their birth,
\par }{\QB that have been carried from the womb.
\par }{\Q \VS{4}Even to old age I am he,
\par }{\QB and even to gray hairs I will carry you.
\par }{\Q I have made, and I will bear.
\par }{\QB Yes, I will carry, and will deliver.
\par }{\BB \par }{\Q \VS{5}“To whom will you compare me, and consider my equal,
\par }{\QB and compare me, as if we were the same?
\par }{\Q \VS{6}Some pour out gold from the bag,
\par }{\QB and weigh silver in the balance.
\par }{\Q They hire a goldsmith,
\par }{\QB and he makes it a god.
\par }{\Q They fall down—
\par }{\QB yes, they worship.
\par }{\Q \VS{7}They bear it on their shoulder.
\par }{\QB They carry it, and set it in its place, and it stands there.
\par }{\QB It cannot move from its place.
\par }{\Q Yes, one may cry to it, yet it can not answer.
\par }{\QB It cannot save him out of his trouble.
\par }{\BB \par }{\Q \VS{8}“Remember this, and show yourselves men.
\par }{\QB Bring it to mind again, you transgressors.
\par }{\Q \VS{9}Remember the former things of old:
\par }{\QB for I am God, and there is no other.
\par }{\QB I am God, and there is none like me.
\par }{\Q \VS{10}I declare the end from the beginning,
\par }{\QB and from ancient times things that are not yet done.
\par }{\Q I say: My counsel will stand,
\par }{\QB and I will do all that I please.
\par }{\Q \VS{11}I call a ravenous bird from the east,
\par }{\QB the man of my counsel from a far country.
\par }{\Q Yes, I have spoken.
\par }{\QB I will also bring it to pass.
\par }{\Q I have planned.
\par }{\QB I will also do it.
\par }{\BB \par }{\Q \VS{12}Listen to me, you stubborn-hearted,
\par }{\QB who are far from righteousness!
\par }{\Q \VS{13}I bring my righteousness near.
\par }{\QB It is not far off,
\par }{\QB and my salvation will not wait.
\par }{\Q I will grant salvation to Zion,
\par }{\QB my glory to Israel.
\par }{\BB \par }
\Chap{47}{\PP \VerseOne{1}“Come down and sit in the dust, virgin daughter of Babylon.
\par }{\QB Sit on the ground without a throne, daughter of the Chaldeans.
\par }{\QB For you will no longer be called tender and delicate.
\par }{\Q \VS{2}Take the millstones and grind flour.
\par }{\QB Remove your veil, lift up your skirt, uncover your legs,
\par }{\QB and wade through the rivers.
\par }{\Q \VS{3}Your nakedness will be uncovered.
\par }{\QB Yes, your shame will be seen.
\par }{\Q I will take vengeance,
\par }{\QB and will spare no one.”
\par }{\BB \par }{\Q \VS{4}Our Redeemer, Yahweh of Armies is his name,
\par }{\QB is the Holy One of Israel.
\par }{\BB \par }{\Q \VS{5}“Sit in silence, and go into darkness,
\par }{\QB daughter of the Chaldeans.
\par }{\Q For you shall no longer be called
\par }{\QB the mistress of kingdoms.
\par }{\Q \VS{6}I was angry with my people.
\par }{\QB I profaned my inheritance
\par }{\QB and gave them into your hand.
\par }{\Q You showed them no mercy.
\par }{\QB You laid a very heavy yoke on the aged.
\par }{\Q \VS{7}You said, ‘I will be a princess forever,’
\par }{\QB so that you didn’t lay these things to your heart,
\par }{\QB nor did you remember the results.
\par }{\BB \par }{\Q \VS{8}“Now therefore hear this, you who are given to pleasures,
\par }{\QB who sit securely,
\par }{\Q who say in your heart,
\par }{\QB ‘I am, and there is no one else besides me.
\par }{\Q I won’t sit as a widow,
\par }{\QB neither will I know the loss of children.’
\par }{\Q \VS{9}But these two things will come to you in a moment in one day,
\par }{\QB the loss of children and widowhood.
\par }{\Q They will come on you in their full measure,
\par }{\QB in the multitude of your sorceries,
\par }{\QB and the great abundance of your enchantments.
\par }{\Q \VS{10}For you have trusted in your wickedness.
\par }{\QB You have said, ‘No one sees me.’
\par }{\Q Your wisdom and your knowledge has perverted you.
\par }{\QB You have said in your heart, ‘I am, and there is no one else besides me.’
\par }{\Q \VS{11}Therefore disaster will come on you.
\par }{\QB You won’t know when it dawns.
\par }{\Q Mischief will fall on you.
\par }{\QB You won’t be able to put it away.
\par }{\Q Desolation will come on you suddenly,
\par }{\QB which you don’t understand.
\par }{\BB \par }{\Q \VS{12}“Stand now with your enchantments
\par }{\QB and with the multitude of your sorceries,
\par }{\QB in which you have labored from your youth,
\par }{\Q as if you might profit,
\par }{\QB as if you might prevail.
\par }{\Q \VS{13}You are wearied in the multitude of your counsels.
\par }{\QB Now let the astrologers, the stargazers, and the monthly prognosticators, stand up and save you from the things that will happen to you.
\par }{\Q \VS{14}Behold, they are like stubble.
\par }{\QB The fire will burn them.
\par }{\QB They won’t deliver themselves from the power of the flame.
\par }{\Q It won’t be a coal to warm at
\par }{\QB or a fire to sit by.
\par }{\Q \VS{15}The things that you labored in will be like this:
\par }{\QB those who have trafficked with you from your youth will each wander in his own way.
\par }{\QB There will be no one to save you.
\par }{\BB \par }
\PoetryChap{48}{\Q \VerseOne{1}“Hear this, house of Jacob,
\par }{\QB you who are called by the name of Israel,
\par }{\QB and have come out of the waters of Judah.
\par }{\Q You swear by Yahweh’s name,
\par }{\QB and make mention of the God of Israel,
\par }{\QB but not in truth, nor in righteousness—
\par }{\Q \VS{2}for they call themselves citizens of the holy city,
\par }{\QB and rely on the God of Israel;
\par }{\QB Yahweh of Armies is his name.
\par }{\Q \VS{3}I have declared the former things from of old.
\par }{\QB Yes, they went out of my mouth, and I revealed them.
\par }{\QB I did them suddenly, and they happened.
\par }{\Q \VS{4}Because I knew that you are obstinate,
\par }{\QB and your neck is an iron sinew,
\par }{\QB and your brow bronze;
\par }{\Q \VS{5}therefore I have declared it to you from of old;
\par }{\QB before it came to pass I showed it to you;
\par }{\QB lest you should say, ‘My idol has done them.
\par }{\QB My engraved image and my molten image has commanded them.’
\par }{\Q \VS{6}You have heard it.
\par }{\QB Now see all this.
\par }{\QB And you, won’t you declare it?
\par }{\BB \par }{\Q “I have shown you new things from this time,
\par }{\QB even hidden things, which you have not known.
\par }{\Q \VS{7}They are created now, and not from of old.
\par }{\QB Before today, you didn’t hear them,
\par }{\QB lest you should say, ‘Behold, I knew them.’
\par }{\Q \VS{8}Yes, you didn’t hear.
\par }{\QB Yes, you didn’t know.
\par }{\QB Yes, from of old your ear was not opened,
\par }{\Q for I knew that you dealt very treacherously,
\par }{\QB and were called a transgressor from the womb.
\par }{\Q \VS{9}For my name’s sake, I will defer my anger,
\par }{\QB and for my praise, I hold it back for you
\par }{\QB so that I don’t cut you off.
\par }{\Q \VS{10}Behold, I have refined you,
\par }{\QB but not as silver.
\par }{\QB I have chosen you in the furnace of affliction.
\par }{\Q \VS{11}For my own sake,
\par }{\QB for my own sake, I will do it;
\par }{\Q for how would my name be profaned?
\par }{\QB I will not give my glory to another.
\par }{\BB \par }{\Q \VS{12}“Listen to me, O Jacob,
\par }{\QB and Israel my called:
\par }{\Q I am he.
\par }{\QB I am the first.
\par }{\QB I am also the last.
\par }{\Q \VS{13}Yes, my hand has laid the foundation of the earth,
\par }{\QB and my right hand has spread out the heavens.
\par }{\QB when I call to them, they stand up together.
\par }{\BB \par }{\Q \VS{14}“Assemble yourselves, all of you, and hear!
\par }{\QB Who among them has declared these things?
\par }{\Q He whom Yahweh loves will do what he likes to Babylon,
\par }{\QB and his arm will be against the Chaldeans.
\par }{\Q \VS{15}I, even I, have spoken.
\par }{\QB Yes, I have called him.
\par }{\Q I have brought him
\par }{\QB and he shall make his way prosperous.
\par }{\PP \VS{16}“Come near to me and hear this:
\par }{\BB \par }{\Q “From the beginning I have not spoken in secret;
\par }{\QB from the time that it happened, I was there.”
\par }{\BB \par }{\Q Now the Lord Yahweh has sent me
\par }{\QB with his Spirit.
\par }{\BB \par }{\Q \VS{17}Yahweh,
\par }{\QB your Redeemer,
\par }{\QB the Holy One of Israel says:
\par }{\Q “I am Yahweh your God,
\par }{\QB who teaches you to profit,
\par }{\QB who leads you by the way that you should go.
\par }{\Q \VS{18}Oh that you had listened to my commandments!
\par }{\QB Then your peace would have been like a river
\par }{\QB and your righteousness like the waves of the sea.
\par }{\Q \VS{19}Your offspring also would have been as the sand
\par }{\QB and the descendants of your body like its grains.
\par }{\QB His name would not be cut off nor destroyed from before me.”
\par }{\BB \par }{\Q \VS{20}Leave Babylon!
\par }{\QB Flee from the Chaldeans!
\par }{\Q With a voice of singing announce this,
\par }{\QB tell it even to the end of the earth:
\par }{\QB say, “Yahweh has redeemed his servant Jacob!”
\par }{\Q \VS{21}They didn’t thirst when he led them through the deserts.
\par }{\QB He caused the waters to flow out of the rock for them.
\par }{\QB He also split the rock and the waters gushed out.
\par }{\BB \par }{\Q \VS{22}“There is no peace”, says Yahweh, “for the wicked.”

\par }\PoetryChap{49}{\Q \VerseOne{1}Listen, islands, to me.
\par }{\QB Listen, you peoples, from afar:
\par }{\Q Yahweh has called me from the womb;
\par }{\QB from the inside of my mother, he has mentioned my name.
\par }{\Q \VS{2}He has made my mouth like a sharp sword.
\par }{\QB He has hidden me in the shadow of his hand.
\par }{\Q He has made me a polished shaft.
\par }{\QB He has kept me close in his quiver.
\par }{\Q \VS{3}He said to me, “You are my servant,
\par }{\QB Israel, in whom I will be glorified.”
\par }{\Q \VS{4}But I said, “I have labored in vain.
\par }{\QB I have spent my strength in vain for nothing;
\par }{\Q yet surely the justice due to me is with Yahweh,
\par }{\QB and my reward with my God.”
\par }{\BB \par }{\Q \VS{5}Now Yahweh, he who formed me from the womb to be his servant,
\par }{\QB says to bring Jacob again to him,
\par }{\QB and to gather Israel to him,
\par }{\QB for I am honorable in Yahweh’s eyes,
\par }{\QB and my God has become my strength.
\par }{\Q \VS{6}Indeed, he says, “It is too light a thing that you should be my servant to raise up the tribes of Jacob,
\par }{\QB and to restore the preserved of Israel.
\par }{\Q I will also give you as a light to the nations,
\par }{\QB that you may be my salvation to the end of the earth.”
\par }{\Q \VS{7}Yahweh, the Redeemer of Israel, and his Holy One,
\par }{\QB says to him whom man despises, to him whom the nation abhors, to a servant of rulers:
\par }{\Q “Kings shall see and rise up,
\par }{\QB princes, and they shall worship,
\par }{\QB because of Yahweh who is faithful, even the Holy One of Israel, who has chosen you.”
\par }{\BB \par }{\Q \VS{8}Yahweh says, “I have answered you in an acceptable time.
\par }{\QB I have helped you in a day of salvation.
\par }{\Q I will preserve you and give you for a covenant of the people,
\par }{\QB to raise up the land, to make them inherit the desolate heritage,
\par }{\Q \VS{9}saying to those who are bound, ‘Come out!’;
\par }{\QB to those who are in darkness, ‘Show yourselves!’
\par }{\Q “They shall feed along the paths,
\par }{\QB and their pasture shall be on all treeless heights.
\par }{\Q \VS{10}They shall not hunger nor thirst;
\par }{\QB neither shall the heat nor sun strike them:
\par }{\QB for he who has mercy on them will lead them.
\par }{\QB He will guide them by springs of water.
\par }{\Q \VS{11}I will make all my mountains a road,
\par }{\QB and my highways shall be exalted.
\par }{\Q \VS{12}Behold, these shall come from afar,
\par }{\QB and behold, these from the north and from the west;
\par }{\QB and these from the land of Sinim.”
\par }{\Q \VS{13}Sing, heavens, and be joyful, earth!
\par }{\QB Break out into singing, mountains,
\par }{\Q for Yahweh has comforted his people,
\par }{\QB and will have compassion on his afflicted.
\par }{\BB \par }{\Q \VS{14}But Zion said, “Yahweh has forsaken me,
\par }{\QB and the Lord has forgotten me.”
\par }{\Q \VS{15}“Can a woman forget her nursing child,
\par }{\QB that she should not have compassion on the son of her womb?
\par }{\Q Yes, these may forget,
\par }{\QB yet I will not forget you!
\par }{\Q \VS{16}Behold, I have engraved you on the palms of my hands.
\par }{\QB Your walls are continually before me.
\par }{\Q \VS{17}Your children hurry.
\par }{\QB Your destroyers and those who devastated you will leave you.
\par }{\Q \VS{18}Lift up your eyes all around, and see:
\par }{\QB all these gather themselves together, and come to you.
\par }{\Q As I live,” says Yahweh, “you shall surely clothe yourself with them all as with an ornament,
\par }{\QB and dress yourself with them, like a bride.
\par }{\Q \VS{19}“For, as for your waste and your desolate places,
\par }{\QB and your land that has been destroyed,
\par }{\Q surely now that land will be too small for the inhabitants,
\par }{\QB and those who swallowed you up will be far away.
\par }{\Q \VS{20}The children of your bereavement will say in your ears,
\par }{\QB ‘This place is too small for me.
\par }{\QB Give me a place to live in.’
\par }{\Q \VS{21}Then you will say in your heart, ‘Who has conceived these for me, since I have been bereaved of my children
\par }{\QB and am alone, an exile, and wandering back and forth?
\par }{\Q Who has brought these up?
\par }{\QB Behold, I was left alone. Where were these?’ ”
\par }{\BB \par }{\Q \VS{22}The Lord Yahweh says, “Behold, I will lift up my hand to the nations,
\par }{\QB and lift up my banner to the peoples.
\par }{\Q They shall bring your sons in their bosom,
\par }{\QB and your daughters shall be carried on their shoulders.
\par }{\Q \VS{23}Kings shall be your foster fathers,
\par }{\QB and their queens your nursing mothers.
\par }{\Q They will bow down to you with their faces to the earth,
\par }{\QB and lick the dust of your feet.
\par }{\Q Then you will know that I am Yahweh;
\par }{\QB and those who wait for me shall not be disappointed.”
\par }{\BB \par }{\Q \VS{24}Shall the plunder be taken from the mighty,
\par }{\QB or the lawful captives be delivered?
\par }{\Q \VS{25}But Yahweh says, “Even the captives of the mighty shall be taken away,
\par }{\QB and the plunder retrieved from the fierce,
\par }{\Q for I will contend with him who contends with you
\par }{\QB and I will save your children.
\par }{\Q \VS{26}I will feed those who oppress you with their own flesh;
\par }{\QB and they will be drunk on their own blood, as with sweet wine.
\par }{\Q Then all flesh shall know that I, Yahweh, am your Savior
\par }{\QB and your Redeemer, the Mighty One of Jacob.”

\par }\PoetryChap{50}{\Q \VerseOne{1}Yahweh says, “Where is the bill of your mother’s divorce, with which I have put her away?
\par }{\QB Or to which of my creditors have I sold you?
\par }{\Q Behold, you were sold for your iniquities,
\par }{\QB and your mother was put away for your transgressions.
\par }{\Q \VS{2}Why, when I came, was there no one?
\par }{\QB When I called, why was there no one to answer?
\par }{\Q Is my hand shortened at all, that it can’t redeem?
\par }{\QB Or have I no power to deliver?
\par }{\Q Behold, at my rebuke I dry up the sea.
\par }{\QB I make the rivers a wilderness.
\par }{\QB Their fish stink because there is no water, and die of thirst.
\par }{\Q \VS{3}I clothe the heavens with blackness.
\par }{\QB I make sackcloth their covering.”
\par }{\BB \par }{\Q \VS{4}The Lord Yahweh has given me the tongue of those who are taught,
\par }{\QB that I may know how to sustain with words him who is weary.
\par }{\Q He awakens morning by morning,
\par }{\QB he awakens my ear to hear as those who are taught.
\par }{\Q \VS{5}The Lord Yahweh has opened my ear.
\par }{\QB I was not rebellious.
\par }{\QB I have not turned back.
\par }{\Q \VS{6}I gave my back to those who beat me,
\par }{\QB and my cheeks to those who plucked off the hair.
\par }{\QB I didn’t hide my face from shame and spitting.
\par }{\Q \VS{7}For the Lord Yahweh will help me.
\par }{\QB Therefore I have not been confounded.
\par }{\Q Therefore I have set my face like a flint,
\par }{\QB and I know that I won’t be disappointed.
\par }{\Q \VS{8}He who justifies me is near.
\par }{\QB Who will bring charges against me?
\par }{\Q Let us stand up together.
\par }{\QB Who is my adversary?
\par }{\QB Let him come near to me.
\par }{\Q \VS{9}Behold, the Lord Yahweh will help me!
\par }{\QB Who is he who will condemn me?
\par }{\Q Behold, they will all grow old like a garment.
\par }{\QB The moths will eat them up.
\par }{\BB \par }{\Q \VS{10}Who among you fears Yahweh
\par }{\QB and obeys the voice of his servant?
\par }{\Q He who walks in darkness
\par }{\QB and has no light,
\par }{\Q let him trust in Yahweh’s name,
\par }{\QB and rely on his God.
\par }{\Q \VS{11}Behold, all you who kindle a fire,
\par }{\QB who adorn yourselves with torches around yourselves,
\par }{\Q walk in the flame of your fire,
\par }{\QB and among the torches that you have kindled.
\par }{\Q You will have this from my hand:
\par }{\QB you will lie down in sorrow.

\par }\PoetryChap{51}{\Q \VerseOne{1}“Listen to me, you who follow after righteousness,
\par }{\QB you who seek Yahweh.
\par }{\Q Look to the rock you were cut from,
\par }{\QB and to the quarry you were dug from.
\par }{\Q \VS{2}Look to Abraham your father,
\par }{\QB and to Sarah who bore you;
\par }{\Q for when he was but one I called him,
\par }{\QB I blessed him,
\par }{\QB and made him many.
\par }{\Q \VS{3}For Yahweh has comforted Zion.
\par }{\QB He has comforted all her waste places,
\par }{\QB and has made her wilderness like Eden,
\par }{\QB and her desert like the garden of Yahweh.
\par }{\Q Joy and gladness will be found in them,
\par }{\QB thanksgiving, and the voice of melody.
\par }{\BB \par }{\Q \VS{4}“Listen to me, my people;
\par }{\QB and hear me, my nation,
\par }{\Q for a law will go out from me,
\par }{\QB and I will establish my justice for a light to the peoples.
\par }{\Q \VS{5}My righteousness is near.
\par }{\QB My salvation has gone out,
\par }{\QB and my arms will judge the peoples.
\par }{\Q The islands will wait for me,
\par }{\QB and they will trust my arm.
\par }{\Q \VS{6}Lift up your eyes to the heavens,
\par }{\QB and look at the earth beneath;
\par }{\Q for the heavens will vanish away like smoke,
\par }{\QB and the earth will wear out like a garment.
\par }{\Q Its inhabitants will die in the same way,
\par }{\QB but my salvation will be forever,
\par }{\QB and my righteousness will not be abolished.
\par }{\BB \par }{\Q \VS{7}“Listen to me, you who know righteousness,
\par }{\QB the people in whose heart is my law.
\par }{\Q Don’t fear the reproach of men,
\par }{\QB and don’t be dismayed at their insults.
\par }{\Q \VS{8}For the moth will eat them up like a garment,
\par }{\QB and the worm will eat them like wool;
\par }{\Q but my righteousness will be forever,
\par }{\QB and my salvation to all generations.”
\par }{\BB \par }{\Q \VS{9}Awake, awake, put on strength, arm of Yahweh!
\par }{\QB Awake, as in the days of old,
\par }{\QB the generations of ancient times.
\par }{\Q Isn’t it you who cut Rahab in pieces,
\par }{\QB who pierced the monster?
\par }{\Q \VS{10}Isn’t it you who dried up the sea,
\par }{\QB the waters of the great deep;
\par }{\QB who made the depths of the sea a way for the redeemed to pass over?
\par }{\Q \VS{11}Those ransomed by Yahweh will return,
\par }{\QB and come with singing to Zion.
\par }{\QB Everlasting joy shall be on their heads.
\par }{\Q They will obtain gladness and joy.
\par }{\QB Sorrow and sighing shall flee away.
\par }{\BB \par }{\Q \VS{12}“I, even I, am he who comforts you.
\par }{\QB Who are you, that you are afraid of man who shall die,
\par }{\QB and of the son of man who will be made as grass?
\par }{\Q \VS{13}Have you forgotten Yahweh your Maker,
\par }{\QB who stretched out the heavens,
\par }{\QB and laid the foundations of the earth?
\par }{\Q Do you live in fear continually all day because of the fury of the oppressor,
\par }{\QB when he prepares to destroy?
\par }{\QB Where is the fury of the oppressor?
\par }{\Q \VS{14}The captive exile will speedily be freed.
\par }{\QB He will not die and go down into the pit.
\par }{\QB His bread won’t fail.
\par }{\Q \VS{15}For I am Yahweh your God, who stirs up the sea
\par }{\QB so that its waves roar.
\par }{\QB Yahweh of Armies is his name.
\par }{\Q \VS{16}I have put my words in your mouth
\par }{\QB and have covered you in the shadow of my hand,
\par }{\Q that I may plant the heavens,
\par }{\QB and lay the foundations of the earth,
\par }{\QB and tell Zion, ‘You are my people.’ ”
\par }{\BB \par }{\Q \VS{17}Awake, awake!
\par }{\QB Stand up, Jerusalem,
\par }{\QB you who have drunk from Yahweh’s hand the cup of his wrath.
\par }{\Q You have drunken the bowl of the cup of staggering,
\par }{\QB and drained it.
\par }{\Q \VS{18}There is no one to guide her among all the sons to whom she has given birth;
\par }{\QB and there is no one who takes her by the hand among all the sons whom she has brought up.
\par }{\Q \VS{19}These two things have happened to you—
\par }{\QB who will grieve with you?—
\par }{\Q desolation and destruction,
\par }{\QB and famine and the sword.
\par }{\QB How can I comfort you?
\par }{\Q \VS{20}Your sons have fainted.
\par }{\QB They lie at the head of all the streets,
\par }{\QB like an antelope in a net.
\par }{\Q They are full of Yahweh’s wrath,
\par }{\QB the rebuke of your God.
\par }{\BB \par }{\Q \VS{21}Therefore now hear this, you afflicted,
\par }{\QB and drunken, but not with wine:
\par }{\Q \VS{22}Your Lord Yahweh,
\par }{\QB your God who pleads the cause of his people, says,
\par }{\Q “Behold, I have taken out of your hand the cup of staggering,
\par }{\QB even the bowl of the cup of my wrath.
\par }{\QB You will not drink it any more.
\par }{\Q \VS{23}I will put it into the hand of those who afflict you,
\par }{\QB who have said to your soul, ‘Bow down, that we may walk over you;’
\par }{\QB and you have laid your back as the ground,
\par }{\QB like a street to those who walk over.”

\par }\PoetryChap{52}{\Q \VerseOne{1}Awake, awake! Put on your strength, Zion.
\par }{\QB Put on your beautiful garments, Jerusalem, the holy city:
\par }{\QB for from now on the uncircumcised and the unclean will no more come into you.
\par }{\Q \VS{2}Shake yourself from the dust!
\par }{\QB Arise, sit up, Jerusalem!
\par }{\QB Release yourself from the bonds of your neck, captive daughter of Zion!
\par }{\BB \par }{\Q \VS{3}For Yahweh says, “You were sold for nothing;
\par }{\QB and you will be redeemed without money.”
\par }{\BB \par }{\PP \VS{4}For the Lord Yahweh says:
\par }{\Q “My people went down at the first into Egypt to live there:
\par }{\QB and the Assyrian has oppressed them without cause.
\par }{\BB \par }{\Q \VS{5}“Now therefore, what do I do here,” says Yahweh,
\par }{\QB “seeing that my people are taken away for nothing?
\par }{\Q Those who rule over them mock,” says Yahweh,
\par }{\QB “and my name is blasphemed continually all day long.
\par }{\Q \VS{6}Therefore my people shall know my name.
\par }{\QB Therefore they shall know in that day that I am he who speaks.
\par }{\QB Behold, it is I.”
\par }{\BB \par }{\Q \VS{7}How beautiful on the mountains are the feet of him who brings good news,
\par }{\QB who publishes peace,
\par }{\QB who brings good news,
\par }{\QB who proclaims salvation,
\par }{\QB who says to Zion, “Your God reigns!”
\par }{\Q \VS{8}Your watchmen lift up their voice.
\par }{\QB Together they sing;
\par }{\QB for they shall see eye to eye when Yahweh returns to Zion.
\par }{\Q \VS{9}Break out into joy!
\par }{\QB Sing together, you waste places of Jerusalem;
\par }{\QB for Yahweh has comforted his people.
\par }{\QB He has redeemed Jerusalem.
\par }{\Q \VS{10}Yahweh has made his holy arm bare in the eyes of all the nations.
\par }{\QB All the ends of the earth have seen the salvation of our God.
\par }{\BB \par }{\Q \VS{11}Depart! Depart! Go out from there! Touch no unclean thing!
\par }{\QB Go out from among her!
\par }{\QB Cleanse yourselves, you who carry Yahweh’s vessels.
\par }{\Q \VS{12}For you shall not go out in haste,
\par }{\QB neither shall you go by flight:
\par }{\Q for Yahweh will go before you;
\par }{\QB and the God of Israel will be your rear guard.
\par }{\BB \par }{\Q \VS{13}Behold, my servant will deal wisely.
\par }{\QB He will be exalted and lifted up,
\par }{\QB and will be very high.
\par }{\Q \VS{14}Just as many were astonished at you—
\par }{\QB his appearance was marred more than any man, and his form more than the sons of men—
\par }{\Q \VS{15}so he will cleanse\FTNT{*}{{\FR 52:15 }
or, sprinkle} many nations.
\par }{\QB Kings will shut their mouths at him;
\par }{\QB for they will see that which had not been told them,
\par }{\QB and they will understand that which they had not heard.

\par }\PoetryChap{53}{\Q \VerseOne{1}Who has believed our message?
\par }{\QB To whom has Yahweh’s arm been revealed?
\par }{\Q \VS{2}For he grew up before him as a tender plant,
\par }{\QB and as a root out of dry ground.
\par }{\Q He has no good looks or majesty.
\par }{\QB When we see him, there is no beauty that we should desire him.
\par }{\Q \VS{3}He was despised
\par }{\QB and rejected by men,
\par }{\Q a man of suffering
\par }{\QB and acquainted with disease.
\par }{\Q He was despised as one from whom men hide their face;
\par }{\QB and we didn’t respect him.
\par }{\BB \par }{\Q \VS{4}Surely he has borne our sickness
\par }{\QB and carried our suffering;
\par }{\Q yet we considered him plagued,
\par }{\QB struck by God, and afflicted.
\par }{\Q \VS{5}But he was pierced for our transgressions.
\par }{\QB He was crushed for our iniquities.
\par }{\Q The punishment that brought our peace was on him;
\par }{\QB and by his wounds we are healed.
\par }{\Q \VS{6}All we like sheep have gone astray.
\par }{\QB Everyone has turned to his own way;
\par }{\QB and Yahweh has laid on him the iniquity of us all.
\par }{\BB \par }{\Q \VS{7}He was oppressed,
\par }{\QB yet when he was afflicted he didn’t open his mouth.
\par }{\Q As a lamb that is led to the slaughter,
\par }{\QB and as a sheep that before its shearers is silent,
\par }{\QB so he didn’t open his mouth.
\par }{\Q \VS{8}He was taken away by oppression and judgment.
\par }{\QB As for his generation,
\par }{\QB who considered that he was cut off out of the land of the living
\par }{\QB and stricken for the disobedience of my people?
\par }{\Q \VS{9}They made his grave with the wicked,
\par }{\QB and with a rich man in his death,
\par }{\Q although he had done no violence,
\par }{\QB nor was any deceit in his mouth.
\par }{\BB \par }{\Q \VS{10}Yet it pleased Yahweh to bruise him.
\par }{\QB He has caused him to suffer.
\par }{\Q When you make his soul an offering for sin,
\par }{\QB he will see his offspring.
\par }{\Q He will prolong his days
\par }{\QB and Yahweh’s pleasure will prosper in his hand.
\par }{\Q \VS{11}After the suffering of his soul,
\par }{\QB he will see the light\FTNT{*}{{\FR 53:11 }
So read the Dead Sea Scrolls and Septuagint. Masoretic Text omits “the light”.} and be satisfied.
\par }{\Q My righteous servant will justify many by the knowledge of himself;
\par }{\QB and he will bear their iniquities.
\par }{\Q \VS{12}Therefore I will give him a portion with the great.
\par }{\QB He will divide the plunder with the strong;
\par }{\Q because he poured out his soul to death
\par }{\QB and was counted with the transgressors;
\par }{\Q yet he bore the sins of many
\par }{\QB and made intercession for the transgressors.

\par }\PoetryChap{54}{\Q \VerseOne{1}“Sing, barren, you who didn’t give birth;
\par }{\QB break out into singing, and cry aloud, you who didn’t travail with child:
\par }{\QB for more are the children of the desolate than the children of the married wife,” says Yahweh.
\par }{\Q \VS{2}“Enlarge the place of your tent,
\par }{\QB and let them stretch out the curtains of your habitations;
\par }{\QB don’t spare: lengthen your cords, and strengthen your stakes.
\par }{\Q \VS{3}For you will spread out on the right hand and on the left;
\par }{\QB and your offspring will possess the nations
\par }{\QB and settle in desolate cities.
\par }{\BB \par }{\Q \VS{4}“Don’t be afraid, for you will not be ashamed.
\par }{\QB Don’t be confounded, for you will not be disappointed.
\par }{\Q For you will forget the shame of your youth.
\par }{\QB You will remember the reproach of your widowhood no more.
\par }{\Q \VS{5}For your Maker is your husband; Yahweh of Armies is his name.
\par }{\QB The Holy One of Israel is your Redeemer.
\par }{\QB He will be called the God of the whole earth.
\par }{\Q \VS{6}For Yahweh has called you as a wife forsaken and grieved in spirit,
\par }{\QB even a wife of youth, when she is cast off,” says your God.
\par }{\BB \par }{\Q \VS{7}“For a small moment I have forsaken you,
\par }{\QB but I will gather you with great mercies.
\par }{\Q \VS{8}In overflowing wrath I hid my face from you for a moment,
\par }{\QB but with everlasting loving kindness I will have mercy on you,” says Yahweh your Redeemer.
\par }{\BB \par }{\Q \VS{9}“For this is like the waters of Noah to me;
\par }{\QB for as I have sworn that the waters of Noah will no more go over the earth,
\par }{\QB so I have sworn that I will not be angry with you, nor rebuke you.
\par }{\Q \VS{10}For the mountains may depart,
\par }{\QB and the hills be removed;
\par }{\Q but my loving kindness will not depart from you,
\par }{\QB and my covenant of peace will not be removed,”
\par }{\QB says Yahweh who has mercy on you.
\par }{\BB \par }{\Q \VS{11}“You afflicted, tossed with storms, and not comforted,
\par }{\QB behold, I will set your stones in beautiful colors,
\par }{\QB and lay your foundations with sapphires.
\par }{\Q \VS{12}I will make your pinnacles of rubies,
\par }{\QB your gates of sparkling jewels,
\par }{\QB and all your walls of precious stones.
\par }{\Q \VS{13}All your children will be taught by Yahweh;
\par }{\QB and your children’s peace will be great.
\par }{\Q \VS{14}You will be established in righteousness.
\par }{\QB You will be far from oppression,
\par }{\QB for you will not be afraid,
\par }{\QB and far from terror,
\par }{\QB for it shall not come near you.
\par }{\Q \VS{15}Behold, they may gather together, but not by me.
\par }{\QB Whoever gathers together against you will fall because of you.
\par }{\BB \par }{\Q \VS{16}“Behold, I have created the blacksmith who fans the coals into flame,
\par }{\QB and forges a weapon for his work;
\par }{\QB and I have created the destroyer to destroy.
\par }{\Q \VS{17}No weapon that is formed against you will prevail;
\par }{\QB and you will condemn every tongue that rises against you in judgment.
\par }{\Q This is the heritage of Yahweh’s servants,
\par }{\QB and their righteousness is of me,” says Yahweh.

\par }\PoetryChap{55}{\Q \VerseOne{1}“Hey! Come, everyone who thirsts, to the waters!
\par }{\QB Come, he who has no money, buy, and eat!
\par }{\QB Yes, come, buy wine and milk without money and without price.
\par }{\Q \VS{2}Why do you spend money for that which is not bread,
\par }{\QB and your labor for that which doesn’t satisfy?
\par }{\Q Listen diligently to me, and eat that which is good,
\par }{\QB and let your soul delight itself in richness.
\par }{\Q \VS{3}Turn your ear, and come to me.
\par }{\QB Hear, and your soul will live.
\par }{\QB I will make an everlasting covenant with you, even the sure mercies of David.
\par }{\Q \VS{4}Behold, I have given him for a witness to the peoples,
\par }{\QB a leader and commander to the peoples.
\par }{\Q \VS{5}Behold, you shall call a nation that you don’t know;
\par }{\QB and a nation that didn’t know you shall run to you,
\par }{\QB because of Yahweh your God,
\par }{\QB and for the Holy One of Israel;
\par }{\QB for he has glorified you.”
\par }{\BB \par }{\Q \VS{6}Seek Yahweh while he may be found.
\par }{\QB Call on him while he is near.
\par }{\Q \VS{7}Let the wicked forsake his way,
\par }{\QB and the unrighteous man his thoughts.
\par }{\Q Let him return to Yahweh, and he will have mercy on him,
\par }{\QB to our God, for he will freely pardon.
\par }{\BB \par }{\Q \VS{8}“For my thoughts are not your thoughts,
\par }{\QB and your ways are not my ways,” says Yahweh.
\par }{\Q \VS{9}“For as the heavens are higher than the earth,
\par }{\QB so are my ways higher than your ways,
\par }{\QB and my thoughts than your thoughts.
\par }{\Q \VS{10}For as the rain comes down and the snow from the sky,
\par }{\QB and doesn’t return there, but waters the earth,
\par }{\QB and makes it grow and bud,
\par }{\QB and gives seed to the sower and bread to the eater;
\par }{\Q \VS{11}so is my word that goes out of my mouth:
\par }{\QB it will not return to me void,
\par }{\QB but it will accomplish that which I please,
\par }{\QB and it will prosper in the thing I sent it to do.
\par }{\Q \VS{12}For you shall go out with joy,
\par }{\QB and be led out with peace.
\par }{\Q The mountains and the hills will break out before you into singing;
\par }{\QB and all the trees of the fields will clap their hands.
\par }{\Q \VS{13}Instead of the thorn the cypress tree will come up;
\par }{\QB and instead of the brier the myrtle tree will come up.
\par }{\Q It will make a name for Yahweh,
\par }{\QB for an everlasting sign that will not be cut off.”

\par }\Chap{56}{\PP \VerseOne{1}Yahweh says,
\par }{\Q “Maintain justice
\par }{\QB and do what is right,
\par }{\Q for my salvation is near
\par }{\QB and my righteousness will soon be revealed.
\par }{\Q \VS{2}Blessed is the man who does this,
\par }{\QB and the son of man who holds it fast;
\par }{\Q who keeps the Sabbath without profaning it
\par }{\QB and keeps his hand from doing any evil.”
\par }{\BB \par }{\Q \VS{3}Let no foreigner who has joined himself to Yahweh speak, saying,
\par }{\QB “Yahweh will surely separate me from his people.”
\par }{\QB Do not let the eunuch say, “Behold, I am a dry tree.”
\par }{\BB \par }{\Q \VS{4}For Yahweh says, “To the eunuchs who keep my Sabbaths,
\par }{\QB choose the things that please me,
\par }{\QB and hold fast to my covenant,
\par }{\Q \VS{5}I will give them in my house and within my walls a memorial and a name better than of sons and of daughters.
\par }{\QB I will give them an everlasting name that will not be cut off.
\par }{\BB \par }{\Q \VS{6}Also the foreigners who join themselves to Yahweh
\par }{\QB to serve him,
\par }{\Q and to love Yahweh’s name,
\par }{\QB to be his servants,
\par }{\Q everyone who keeps the Sabbath from profaning it,
\par }{\QB and holds fast my covenant,
\par }{\Q \VS{7}I will bring these to my holy mountain,
\par }{\QB and make them joyful in my house of prayer.
\par }{\Q Their burnt offerings and their sacrifices will be accepted on my altar;
\par }{\QB for my house will be called a house of prayer for all peoples.”
\par }{\Q \VS{8}The Lord Yahweh, who gathers the outcasts of Israel, says,
\par }{\QB “I will yet gather others to him,
\par }{\QB in addition to his own who are gathered.”
\par }{\BB \par }{\Q \VS{9}All you animals of the field,
\par }{\QB come to devour,
\par }{\QB all you animals in the forest.
\par }{\Q \VS{10}His watchmen are blind.
\par }{\QB They are all without knowledge.
\par }{\Q They are all mute dogs.
\par }{\QB They can’t bark—
\par }{\QB dreaming, lying down, loving to slumber.
\par }{\Q \VS{11}Yes, the dogs are greedy.
\par }{\QB They can never have enough.
\par }{\Q They are shepherds who can’t understand.
\par }{\QB They have all turned to their own way,
\par }{\QB each one to his gain, from every quarter.
\par }{\Q \VS{12}“Come,” they say, “I will get wine,
\par }{\QB and we will fill ourselves with strong drink;
\par }{\QB and tomorrow will be as today,
\par }{\QB great beyond measure.”

\par }\PoetryChap{57}{\Q \VerseOne{1}The righteous perish,
\par }{\QB and no one lays it to heart.
\par }{\Q Merciful men are taken away,
\par }{\QB and no one considers that the righteous is taken away from the evil.
\par }{\Q \VS{2}He enters into peace.
\par }{\QB They rest in their beds,
\par }{\QB each one who walks in his uprightness.
\par }{\BB \par }{\Q \VS{3}“But draw near here, you sons of a sorceress,
\par }{\QB you offspring of adulterers and prostitutes.
\par }{\Q \VS{4}Whom do you mock?
\par }{\QB Against whom do you make a wide mouth
\par }{\QB and stick out your tongue?
\par }{\Q Aren’t you children of disobedience
\par }{\QB and offspring of falsehood,
\par }{\Q \VS{5}you who inflame yourselves among the oaks,
\par }{\QB under every green tree;
\par }{\Q who kill the children in the valleys,
\par }{\QB under the clefts of the rocks?
\par }{\Q \VS{6}Among the smooth stones of the valley is your portion.
\par }{\QB They, they are your lot.
\par }{\Q You have even poured a drink offering to them.
\par }{\QB You have offered an offering.
\par }{\QB Shall I be appeased for these things?
\par }{\Q \VS{7}On a high and lofty mountain you have set your bed.
\par }{\QB You also went up there to offer sacrifice.
\par }{\Q \VS{8}You have set up your memorial behind the doors and the posts,
\par }{\QB for you have exposed yourself to someone besides me,
\par }{\QB and have gone up.
\par }{\Q You have enlarged your bed
\par }{\QB and made you a covenant with them.
\par }{\QB You loved what you saw on their bed.
\par }{\Q \VS{9}You went to the king with oil,
\par }{\QB increased your perfumes,
\par }{\QB sent your ambassadors far off,
\par }{\QB and degraded yourself even to Sheol.\FTNT{*}{{\FR 57:9 }
Sheol is the place of the dead.}
\par }{\Q \VS{10}You were wearied with the length of your ways;
\par }{\QB yet you didn’t say, ‘It is in vain.’
\par }{\Q You found a reviving of your strength;
\par }{\QB therefore you weren’t faint.
\par }{\BB \par }{\Q \VS{11}“Whom have you dreaded and feared,
\par }{\QB so that you lie,
\par }{\QB and have not remembered me, nor laid it to your heart?
\par }{\Q Haven’t I held my peace for a long time,
\par }{\QB and you don’t fear me?
\par }{\Q \VS{12}I will declare your righteousness;
\par }{\QB and as for your works, they will not benefit you.
\par }{\Q \VS{13}When you cry,
\par }{\QB let those whom you have gathered deliver you;
\par }{\Q but the wind will take them.
\par }{\QB a breath will carry them all away:
\par }{\Q but he who takes refuge in me will possess the land,
\par }{\QB and will inherit my holy mountain.”
\par }{\BB \par }{\Q \VS{14}He will say, “Build up, build up, prepare the way!
\par }{\QB Remove the stumbling-block out of the way of my people.”
\par }{\Q \VS{15}For the high and lofty One who inhabits eternity,
\par }{\QB whose name is Holy, says:
\par }{\Q “I dwell in the high and holy place, with him also who is of a contrite and humble spirit,
\par }{\QB to revive the spirit of the humble,
\par }{\QB and to revive the heart of the contrite.
\par }{\Q \VS{16}For I will not contend forever, neither will I always be angry;
\par }{\QB for the spirit would faint before me,
\par }{\QB and the souls whom I have made.
\par }{\Q \VS{17}I was angry because of the iniquity of his covetousness and struck him.
\par }{\QB I hid myself and was angry;
\par }{\QB and he went on backsliding in the way of his heart.
\par }{\Q \VS{18}I have seen his ways, and will heal him.
\par }{\QB I will lead him also,
\par }{\QB and restore comforts to him and to his mourners.
\par }{\Q \VS{19}I create the fruit of the lips:
\par }{\QB Peace, peace, to him who is far off and to him who is near,”
\par }{\QB says Yahweh; “and I will heal them.”
\par }{\Q \VS{20}But the wicked are like the troubled sea;
\par }{\QB for it can’t rest and its waters cast up mire and mud.
\par }{\Q \VS{21}“There is no peace”, says my God,
\par }{\QB “for the wicked.”

\par }\PoetryChap{58}{\Q \VerseOne{1}“Cry aloud! Don’t spare!
\par }{\QB Lift up your voice like a trumpet!
\par }{\Q Declare to my people their disobedience,
\par }{\QB and to the house of Jacob their sins.
\par }{\Q \VS{2}Yet they seek me daily,
\par }{\QB and delight to know my ways.
\par }{\Q As a nation that did righteousness,
\par }{\QB and didn’t forsake the ordinance of their God,
\par }{\Q they ask of me righteous judgments.
\par }{\QB They delight to draw near to God.
\par }{\Q \VS{3}‘Why have we fasted,’ they say, ‘and you don’t see?
\par }{\QB Why have we afflicted our soul, and you don’t notice?’
\par }{\BB \par }{\Q “Behold, in the day of your fast you find pleasure,
\par }{\QB and oppress all your laborers.
\par }{\Q \VS{4}Behold, you fast for strife and contention,
\par }{\QB and to strike with the fist of wickedness.
\par }{\QB You don’t fast today so as to make your voice to be heard on high.
\par }{\Q \VS{5}Is this the fast that I have chosen?
\par }{\QB A day for a man to humble his soul?
\par }{\Q Is it to bow down his head like a reed,
\par }{\QB and to spread sackcloth and ashes under himself?
\par }{\Q Will you call this a fast,
\par }{\QB and an acceptable day to Yahweh?
\par }{\BB \par }{\Q \VS{6}“Isn’t this the fast that I have chosen:
\par }{\QB to release the bonds of wickedness,
\par }{\QB to undo the straps of the yoke,
\par }{\QB to let the oppressed go free,
\par }{\QB and that you break every yoke?
\par }{\Q \VS{7}Isn’t it to distribute your bread to the hungry,
\par }{\QB and that you bring the poor who are cast out to your house?
\par }{\Q When you see the naked,
\par }{\QB that you cover him;
\par }{\QB and that you not hide yourself from your own flesh?
\par }{\Q \VS{8}Then your light will break out as the morning,
\par }{\QB and your healing will appear quickly;
\par }{\Q then your righteousness shall go before you,
\par }{\QB and Yahweh’s glory will be your rear guard.
\par }{\Q \VS{9}Then you will call, and Yahweh will answer.
\par }{\QB You will cry for help, and he will say, ‘Here I am.’
\par }{\BB \par }{\Q “If you take away from among you the yoke,
\par }{\QB finger pointing,
\par }{\QB and speaking wickedly;
\par }{\Q \VS{10}and if you pour out your soul to the hungry,
\par }{\QB and satisfy the afflicted soul,
\par }{\Q then your light will rise in darkness,
\par }{\QB and your obscurity will be as the noonday;
\par }{\Q \VS{11}and Yahweh will guide you continually,
\par }{\QB satisfy your soul in dry places,
\par }{\QB and make your bones strong.
\par }{\Q You will be like a watered garden,
\par }{\QB and like a spring of water
\par }{\QB whose waters don’t fail.
\par }{\Q \VS{12}Those who will be of you will build the old waste places.
\par }{\QB You will raise up the foundations of many generations.
\par }{\Q You will be called Repairer of the Breach,
\par }{\QB Restorer of Paths with Dwellings.
\par }{\BB \par }{\Q \VS{13}“If you turn away your foot from the Sabbath,
\par }{\QB from doing your pleasure on my holy day;
\par }{\Q and call the Sabbath a delight,
\par }{\QB and the holy of Yahweh honorable;
\par }{\QB and honor it,
\par }{\QB not doing your own ways,
\par }{\QB nor finding your own pleasure,
\par }{\QB nor speaking your own words,
\par }{\Q \VS{14}then you will delight yourself in Yahweh,
\par }{\QB and I will make you to ride on the high places of the earth,
\par }{\QB and I will feed you with the heritage of Jacob your father;”
\par }{\QB for Yahweh’s mouth has spoken it.

\par }\PoetryChap{59}{\Q \VerseOne{1}Behold, Yahweh’s hand is not shortened, that it can’t save;
\par }{\QB nor his ear dull, that it can’t hear.
\par }{\Q \VS{2}But your iniquities have separated you and your God,
\par }{\QB and your sins have hidden his face from you,
\par }{\QB so that he will not hear.
\par }{\Q \VS{3}For your hands are defiled with blood,
\par }{\QB and your fingers with iniquity.
\par }{\Q Your lips have spoken lies.
\par }{\QB Your tongue mutters wickedness.
\par }{\Q \VS{4}No one sues in righteousness,
\par }{\QB and no one pleads in truth.
\par }{\Q They trust in vanity,
\par }{\QB and speak lies.
\par }{\Q They conceive mischief,
\par }{\QB and give birth to iniquity.
\par }{\Q \VS{5}They hatch adders’ eggs,
\par }{\QB and weave the spider’s web.
\par }{\Q He who eats of their eggs dies;
\par }{\QB and that which is crushed breaks out into a viper.
\par }{\Q \VS{6}Their webs won’t become garments.
\par }{\QB They won’t cover themselves with their works.
\par }{\Q Their works are works of iniquity,
\par }{\QB and acts of violence are in their hands.
\par }{\Q \VS{7}Their feet run to evil,
\par }{\QB and they hurry to shed innocent blood.
\par }{\Q Their thoughts are thoughts of iniquity.
\par }{\QB Desolation and destruction are in their paths.
\par }{\Q \VS{8}They don’t know the way of peace;
\par }{\QB and there is no justice in their ways.
\par }{\Q They have made crooked paths for themselves;
\par }{\QB whoever goes in them doesn’t know peace.
\par }{\BB \par }{\Q \VS{9}Therefore justice is far from us,
\par }{\QB and righteousness doesn’t overtake us.
\par }{\Q We look for light, but see darkness;
\par }{\QB for brightness, but we walk in obscurity.
\par }{\Q \VS{10}We grope for the wall like the blind.
\par }{\QB Yes, we grope as those who have no eyes.
\par }{\Q We stumble at noon as if it were twilight.
\par }{\QB Among those who are strong, we are like dead men.
\par }{\Q \VS{11}We all roar like bears
\par }{\QB and moan bitterly like doves.
\par }{\Q We look for justice, but there is none,
\par }{\QB for salvation, but it is far off from us.
\par }{\BB \par }{\Q \VS{12}For our transgressions are multiplied before you,
\par }{\QB and our sins testify against us;
\par }{\Q for our transgressions are with us,
\par }{\QB and as for our iniquities, we know them:
\par }{\Q \VS{13}transgressing and denying Yahweh,
\par }{\QB and turning away from following our God,
\par }{\QB speaking oppression and revolt,
\par }{\QB conceiving and uttering from the heart words of falsehood.
\par }{\Q \VS{14}Justice is turned away backward,
\par }{\QB and righteousness stands far away;
\par }{\Q for truth has fallen in the street,
\par }{\QB and uprightness can’t enter.
\par }{\Q \VS{15}Yes, truth is lacking;
\par }{\QB and he who departs from evil makes himself a prey.
\par }{\BB \par }{\Q Yahweh saw it,
\par }{\QB and it displeased him that there was no justice.
\par }{\Q \VS{16}He saw that there was no man,
\par }{\QB and wondered that there was no intercessor.
\par }{\Q Therefore his own arm brought salvation to him;
\par }{\QB and his righteousness sustained him.
\par }{\Q \VS{17}He put on righteousness as a breastplate,
\par }{\QB and a helmet of salvation on his head.
\par }{\Q He put on garments of vengeance for clothing,
\par }{\QB and was clad with zeal as a mantle.
\par }{\Q \VS{18}According to their deeds,
\par }{\QB he will repay as appropriate,
\par }{\QB wrath to his adversaries,
\par }{\QB recompense to his enemies;
\par }{\QB he will repay the islands their due.
\par }{\Q \VS{19}So they will fear Yahweh’s name from the west,
\par }{\QB and his glory from the rising of the sun;
\par }{\Q for he will come as a rushing stream,
\par }{\QB which Yahweh’s breath drives.
\par }{\BB \par }{\Q \VS{20}“A Redeemer will come to Zion,
\par }{\QB and to those who turn from disobedience in Jacob,” says Yahweh.
\par }{\PP \VS{21}“As for me, this is my covenant with them,” says Yahweh. “My Spirit who is on you, and my words which I have put in your mouth shall not depart out of your mouth, nor out of the mouth of your offspring, nor out of the mouth of your offspring’s offspring,” says Yahweh, “from now on and forever.”

\par }\PoetryChap{60}{\Q \VerseOne{1}“Arise, shine; for your light has come,
\par }{\QB and Yahweh’s glory has risen on you.
\par }{\Q \VS{2}For, behold, darkness will cover the earth,
\par }{\QB and thick darkness the peoples;
\par }{\Q but Yahweh will arise on you,
\par }{\QB and his glory shall be seen on you.
\par }{\Q \VS{3}Nations will come to your light,
\par }{\QB and kings to the brightness of your rising.
\par }{\BB \par }{\Q \VS{4}“Lift up your eyes all around, and see:
\par }{\QB they all gather themselves together.
\par }{\QB They come to you.
\par }{\Q Your sons will come from far away,
\par }{\QB and your daughters will be carried in arms.
\par }{\Q \VS{5}Then you shall see and be radiant,
\par }{\QB and your heart will thrill and be enlarged;
\par }{\Q because the abundance of the sea will be turned to you.
\par }{\QB The wealth of the nations will come to you.
\par }{\Q \VS{6}A multitude of camels will cover you,
\par }{\QB the dromedaries of Midian and Ephah.
\par }{\Q All from Sheba will come.
\par }{\QB They will bring gold and frankincense,
\par }{\QB and will proclaim the praises of Yahweh.
\par }{\Q \VS{7}All the flocks of Kedar will be gathered together to you.
\par }{\QB The rams of Nebaioth will serve you.
\par }{\Q They will be accepted as offerings on my altar;
\par }{\QB and I will beautify my glorious house.
\par }{\BB \par }{\Q \VS{8}“Who are these who fly as a cloud,
\par }{\QB and as the doves to their windows?
\par }{\Q \VS{9}Surely the islands will wait for me,
\par }{\QB and the ships of Tarshish first,
\par }{\Q to bring your sons from far,
\par }{\QB their silver and their gold with them,
\par }{\Q for the name of Yahweh your God,
\par }{\QB and for the Holy One of Israel,
\par }{\QB because he has glorified you.
\par }{\BB \par }{\Q \VS{10}“Foreigners will build up your walls,
\par }{\QB and their kings will serve you:
\par }{\Q for in my wrath I struck you,
\par }{\QB but in my favor I have had mercy on you.
\par }{\Q \VS{11}Your gates also shall be open continually; they shall not be shut day nor night, that men may bring to you the wealth of the nations, and their kings led captive.
\par }{\Q \VS{12}For that nation and kingdom that will not serve you shall perish; yes, those nations shall be utterly wasted.
\par }{\BB \par }{\Q \VS{13}“The glory of Lebanon shall come to you, the cypress tree, the pine, and the box tree together, to beautify the place of my sanctuary; and I will make the place of my feet glorious.
\par }{\Q \VS{14}The sons of those who afflicted you will come bowing to you;
\par }{\QB and all those who despised you will bow themselves down at the soles of your feet.
\par }{\Q They will call you Yahweh’s City,
\par }{\QB the Zion of the Holy One of Israel.
\par }{\BB \par }{\Q \VS{15}“Whereas you have been forsaken and hated,
\par }{\QB so that no one passed through you,
\par }{\Q I will make you an eternal excellency,
\par }{\QB a joy of many generations.
\par }{\Q \VS{16}You will also drink the milk of the nations,
\par }{\QB and will nurse from royal breasts.
\par }{\Q Then you will know that I, Yahweh, am your Savior,
\par }{\QB your Redeemer,
\par }{\QB the Mighty One of Jacob.
\par }{\Q \VS{17}For bronze I will bring gold;
\par }{\QB for iron I will bring silver;
\par }{\QB for wood, bronze,
\par }{\QB and for stones, iron.
\par }{\Q I will also make peace your governor,
\par }{\QB and righteousness your ruler.
\par }{\Q \VS{18}Violence shall no more be heard in your land,
\par }{\QB nor desolation or destruction within your borders;
\par }{\Q but you will call your walls Salvation,
\par }{\QB and your gates Praise.
\par }{\Q \VS{19}The sun will be no more your light by day;
\par }{\QB nor will the brightness of the moon give light to you,
\par }{\Q but Yahweh will be your everlasting light,
\par }{\QB and your God will be your glory.
\par }{\Q \VS{20}Your sun will not go down any more,
\par }{\QB nor will your moon withdraw itself;
\par }{\Q for Yahweh will be your everlasting light,
\par }{\QB and the days of your mourning will end.
\par }{\Q \VS{21}Then your people will all be righteous.
\par }{\QB They will inherit the land forever,
\par }{\QB the branch of my planting,
\par }{\QB the work of my hands,
\par }{\QB that I may be glorified.
\par }{\Q \VS{22}The little one will become a thousand,
\par }{\QB and the small one a strong nation.
\par }{\QB I, Yahweh, will do this quickly in its time.”

\par }\PoetryChap{61}{\Q \VerseOne{1}The Lord Yahweh’s Spirit is on me,
\par }{\QB because Yahweh has anointed me to preach good news to the humble.
\par }{\Q He has sent me to bind up the broken hearted,
\par }{\QB to proclaim liberty to the captives
\par }{\QB and release to those who are bound,
\par }{\Q \VS{2}to proclaim the year of Yahweh’s favor
\par }{\QB and the day of vengeance of our God,
\par }{\QB to comfort all who mourn,
\par }{\Q \VS{3}to provide for those who mourn in Zion,
\par }{\QB to give to them a garland for ashes,
\par }{\QB the oil of joy for mourning,
\par }{\QB the garment of praise for the spirit of heaviness,
\par }{\Q that they may be called trees of righteousness,
\par }{\QB the planting of Yahweh,
\par }{\QB that he may be glorified.
\par }{\BB \par }{\Q \VS{4}They will rebuild the old ruins.
\par }{\QB They will raise up the former devastated places.
\par }{\Q They will repair the ruined cities
\par }{\QB that have been devastated for many generations.
\par }{\Q \VS{5}Strangers will stand and feed your flocks.
\par }{\QB Foreigners will work your fields and your vineyards.
\par }{\Q \VS{6}But you will be called Yahweh’s priests.
\par }{\QB Men will call you the servants of our God.
\par }{\Q You will eat the wealth of the nations.
\par }{\QB You will boast in their glory.
\par }{\Q \VS{7}Instead of your shame you will have double.
\par }{\QB Instead of dishonor, they will rejoice in their portion.
\par }{\Q Therefore in their land they will possess double.
\par }{\QB Everlasting joy will be to them.
\par }{\Q \VS{8}“For I, Yahweh, love justice.
\par }{\QB I hate robbery and iniquity.
\par }{\Q I will give them their reward in truth
\par }{\QB and I will make an everlasting covenant with them.
\par }{\Q \VS{9}Their offspring will be known among the nations,
\par }{\QB and their offspring among the peoples.
\par }{\Q All who see them will acknowledge them,
\par }{\QB that they are the offspring which Yahweh has blessed.”
\par }{\BB \par }{\Q \VS{10}I will greatly rejoice in Yahweh!
\par }{\QB My soul will be joyful in my God,
\par }{\Q for he has clothed me with the garments of salvation.
\par }{\QB He has covered me with the robe of righteousness,
\par }{\QB as a bridegroom decks himself with a garland
\par }{\QB and as a bride adorns herself with her jewels.
\par }{\Q \VS{11}For as the earth produces its bud,
\par }{\QB and as the garden causes the things that are sown in it to spring up,
\par }{\QB so the Lord Yahweh will cause righteousness and praise to spring up before all the nations.

\par }\PoetryChap{62}{\Q \VerseOne{1}For Zion’s sake I will not hold my peace,
\par }{\QB and for Jerusalem’s sake I will not rest,
\par }{\Q until her righteousness shines out like the dawn,
\par }{\QB and her salvation like a burning lamp.
\par }{\Q \VS{2}The nations will see your righteousness,
\par }{\QB and all kings your glory.
\par }{\Q You will be called by a new name,
\par }{\QB which Yahweh’s mouth will name.
\par }{\Q \VS{3}You will also be a crown of beauty in Yahweh‘s hand,
\par }{\QB and a royal diadem in your God’s hand.
\par }{\Q \VS{4}You will not be called Forsaken any more,
\par }{\QB nor will your land be called Desolate any more;
\par }{\Q but you will be called Hephzibah,\FTNT{*}{{\FR 62:4 }
Hephzibah means “I delight in her”.}
\par }{\QB and your land Beulah;\FTNT{†}{{\FR 62:4 }
Beulah means “married”}
\par }{\Q for Yahweh delights in you,
\par }{\QB and your land will be married.
\par }{\Q \VS{5}For as a young man marries a virgin,
\par }{\QB so your sons will marry you.
\par }{\Q As a bridegroom rejoices over his bride,
\par }{\QB so your God will rejoice over you.
\par }{\BB \par }{\Q \VS{6}I have set watchmen on your walls, Jerusalem.
\par }{\QB They will never be silent day nor night.
\par }{\Q You who call on Yahweh, take no rest,
\par }{\QB \VS{7}and give him no rest, until he establishes,
\par }{\QB and until he makes Jerusalem a praise in the earth.
\par }{\BB \par }{\Q \VS{8}Yahweh has sworn by his right hand,
\par }{\QB and by the arm of his strength,
\par }{\Q “Surely I will no more give your grain to be food for your enemies,
\par }{\QB and foreigners will not drink your new wine, for which you have labored,
\par }{\Q \VS{9}but those who have harvested it will eat it, and praise Yahweh.
\par }{\QB Those who have gathered it will drink it in the courts of my sanctuary.”
\par }{\BB \par }{\Q \VS{10}Go through, go through the gates!
\par }{\QB Prepare the way of the people!
\par }{\Q Build up, build up the highway!
\par }{\QB Gather out the stones!
\par }{\QB Lift up a banner for the peoples.
\par }{\Q \VS{11}Behold, Yahweh has proclaimed to the end of the earth,
\par }{\QB “Say to the daughter of Zion,
\par }{\QB ‘Behold, your salvation comes!
\par }{\Q Behold, his reward is with him,
\par }{\QB and his recompense before him!’ ”
\par }{\Q \VS{12}They will call them “The Holy People,
\par }{\QB Yahweh’s Redeemed”.
\par }{\Q You will be called “Sought Out,
\par }{\QB A City Not Forsaken”.

\par }\PoetryChap{63}{\Q \VerseOne{1}Who is this who comes from Edom,
\par }{\QB with dyed garments from Bozrah?
\par }{\Q Who is this who is glorious in his clothing,
\par }{\QB marching in the greatness of his strength?
\par }{\Q “It is I who speak in righteousness,
\par }{\QB mighty to save.”
\par }{\Q \VS{2}Why is your clothing red,
\par }{\QB and your garments like him who treads in the wine vat?
\par }{\BB \par }{\Q \VS{3}“I have trodden the wine press alone.
\par }{\QB Of the peoples, no one was with me.
\par }{\Q Yes, I trod them in my anger
\par }{\QB and trampled them in my wrath.
\par }{\Q Their lifeblood is sprinkled on my garments,
\par }{\QB and I have stained all my clothing.
\par }{\Q \VS{4}For the day of vengeance was in my heart,
\par }{\QB and the year of my redeemed has come.
\par }{\Q \VS{5}I looked, and there was no one to help;
\par }{\QB and I wondered that there was no one to uphold.
\par }{\Q Therefore my own arm brought salvation to me.
\par }{\QB My own wrath upheld me.
\par }{\Q \VS{6}I trod down the peoples in my anger
\par }{\QB and made them drunk in my wrath.
\par }{\QB I poured their lifeblood out on the earth.”
\par }{\BB \par }{\Q \VS{7}I will tell of the loving kindnesses of Yahweh
\par }{\QB and the praises of Yahweh,
\par }{\QB according to all that Yahweh has given to us,
\par }{\Q and the great goodness toward the house of Israel,
\par }{\QB which he has given to them according to his mercies,
\par }{\QB and according to the multitude of his loving kindnesses.
\par }{\Q \VS{8}For he said, “Surely, they are my people,
\par }{\QB children who will not deal falsely;”
\par }{\QB so he became their Savior.
\par }{\Q \VS{9}In all their affliction he was afflicted,
\par }{\QB and the angel of his presence saved them.
\par }{\Q In his love and in his pity he redeemed them.
\par }{\QB He bore them,
\par }{\QB and carried them all the days of old.
\par }{\BB \par }{\Q \VS{10}But they rebelled
\par }{\QB and grieved his Holy Spirit.
\par }{\Q Therefore he turned and became their enemy,
\par }{\QB and he himself fought against them.
\par }{\BB \par }{\Q \VS{11}Then he remembered the days of old,
\par }{\QB Moses and his people, saying,
\par }{\Q “Where is he who brought them up out of the sea with the shepherds of his flock?
\par }{\QB Where is he who put his Holy Spirit among them?”
\par }{\Q \VS{12}Who caused his glorious arm to be at Moses’ right hand?
\par }{\QB Who divided the waters before them, to make himself an everlasting name?
\par }{\Q \VS{13}Who led them through the depths,
\par }{\QB like a horse in the wilderness,
\par }{\QB so that they didn’t stumble?
\par }{\Q \VS{14}As the livestock that go down into the valley,
\par }{\QB Yahweh’s Spirit caused them to rest.
\par }{\QB So you led your people to make yourself a glorious name.
\par }{\BB \par }{\Q \VS{15}Look down from heaven,
\par }{\QB and see from the habitation of your holiness and of your glory.
\par }{\Q Where are your zeal and your mighty acts?
\par }{\QB The yearning of your heart and your compassion is restrained toward me.
\par }{\Q \VS{16}For you are our Father,
\par }{\QB though Abraham doesn’t know us,
\par }{\QB and Israel does not acknowledge us.
\par }{\Q You, Yahweh, are our Father.
\par }{\QB Our Redeemer from everlasting is your name.
\par }{\Q \VS{17}O Yahweh, why do you make us wander from your ways,
\par }{\QB and harden our heart from your fear?
\par }{\Q Return for your servants’ sake,
\par }{\QB the tribes of your inheritance.
\par }{\Q \VS{18}Your holy people possessed it but a little while.
\par }{\QB Our adversaries have trodden down your sanctuary.
\par }{\Q \VS{19}We have become like those over whom you never ruled,
\par }{\QB like those who were not called by your name.

\par }\PoetryChap{64}{\Q \VerseOne{1}Oh that you would tear the heavens,
\par }{\QB that you would come down,
\par }{\QB that the mountains might quake at your presence.
\par }{\Q \VS{2}As when fire kindles the brushwood,
\par }{\QB and the fire causes the water to boil;
\par }{\Q Make your name known to your adversaries,
\par }{\QB that the nations may tremble at your presence!
\par }{\Q \VS{3}When you did awesome things which we didn’t look for,
\par }{\QB you came down, and the mountains quaked at your presence.
\par }{\Q \VS{4}For from of old men have not heard,
\par }{\QB nor perceived by the ear,
\par }{\QB nor has the eye seen a God besides you,
\par }{\QB who works for him who waits for him.
\par }{\Q \VS{5}You meet him who rejoices and does righteousness,
\par }{\QB those who remember you in your ways.
\par }{\Q Behold, you were angry, and we sinned.
\par }{\QB We have been in sin for a long time.
\par }{\QB Shall we be saved?
\par }{\Q \VS{6}For we have all become like one who is unclean,
\par }{\QB and all our righteousness is like a polluted garment.
\par }{\Q We all fade like a leaf;
\par }{\QB and our iniquities, like the wind, take us away.
\par }{\Q \VS{7}There is no one who calls on your name,
\par }{\QB who stirs himself up to take hold of you;
\par }{\Q for you have hidden your face from us,
\par }{\QB and have consumed us by means of our iniquities.
\par }{\BB \par }{\Q \VS{8}But now, Yahweh, you are our Father.
\par }{\QB We are the clay and you our potter.
\par }{\QB We all are the work of your hand.
\par }{\Q \VS{9}Don’t be furious, Yahweh.
\par }{\QB Don’t remember iniquity forever.
\par }{\Q Look and see, we beg you,
\par }{\QB we are all your people.
\par }{\Q \VS{10}Your holy cities have become a wilderness.
\par }{\QB Zion has become a wilderness,
\par }{\QB Jerusalem a desolation.
\par }{\Q \VS{11}Our holy and our beautiful house where our fathers praised you
\par }{\QB is burned with fire.
\par }{\QB All our pleasant places are laid waste.
\par }{\Q \VS{12}Will you hold yourself back for these things, Yahweh?
\par }{\QB Will you keep silent and punish us very severely?

\par }\PoetryChap{65}{\Q \VerseOne{1}“I am inquired of by those who didn’t ask.
\par }{\QB I am found by those who didn’t seek me.
\par }{\QB I said, ‘See me, see me,’ to a nation that was not called by my name.
\par }{\Q \VS{2}I have spread out my hands all day to a rebellious people,
\par }{\QB who walk in a way that is not good,
\par }{\QB after their own thoughts;
\par }{\Q \VS{3}a people who provoke me to my face continually,
\par }{\QB sacrificing in gardens,
\par }{\QB and burning incense on bricks;
\par }{\Q \VS{4}who sit among the graves,
\par }{\QB and spend nights in secret places;
\par }{\Q who eat pig’s meat,
\par }{\QB and broth of abominable things is in their vessels;
\par }{\Q \VS{5}who say, ‘Stay by yourself,
\par }{\QB don’t come near to me,
\par }{\QB for I am holier than you.’
\par }{\Q These are smoke in my nose,
\par }{\QB a fire that burns all day.
\par }{\BB \par }{\Q \VS{6}“Behold, it is written before me:
\par }{\QB I will not keep silence,
\par }{\QB but will repay,
\par }{\QB yes, I will repay into their bosom,
\par }{\Q \VS{7}your own iniquities, and the iniquities of your fathers together”, says Yahweh,
\par }{\QB “who have burned incense on the mountains,
\par }{\QB and blasphemed me on the hills.
\par }{\QB Therefore I will first measure their work into their bosom.”
\par }{\BB \par }{\PP \VS{8}Yahweh says,
\par }{\Q “As the new wine is found in the cluster,
\par }{\QB and one says, ‘Don’t destroy it, for a blessing is in it:’
\par }{\Q so I will do for my servants’ sake,
\par }{\QB that I may not destroy them all.
\par }{\Q \VS{9}I will bring offspring out of Jacob,
\par }{\QB and out of Judah an inheritor of my mountains.
\par }{\Q My chosen will inherit it,
\par }{\QB and my servants will dwell there.
\par }{\Q \VS{10}Sharon will be a fold of flocks,
\par }{\QB and the valley of Achor a place for herds to lie down in,
\par }{\QB for my people who have sought me.
\par }{\BB \par }{\Q \VS{11}“But you who forsake Yahweh,
\par }{\QB who forget my holy mountain,
\par }{\QB who prepare a table for Fortune,
\par }{\QB and who fill up mixed wine to Destiny;
\par }{\Q \VS{12}I will destine you to the sword,
\par }{\QB and you will all bow down to the slaughter;
\par }{\Q because when I called, you didn’t answer.
\par }{\QB When I spoke, you didn’t listen;
\par }{\Q but you did that which was evil in my eyes,
\par }{\QB and chose that in which I didn’t delight.”
\par }{\BB \par }{\PP \VS{13}Therefore the Lord Yahweh says,
\par }{\Q “Behold, my servants will eat,
\par }{\QB but you will be hungry;
\par }{\Q behold, my servants will drink,
\par }{\QB but you will be thirsty.
\par }{\Q Behold, my servants will rejoice,
\par }{\QB but you will be disappointed;
\par }{\Q \VS{14}Behold, my servants will sing for joy of heart,
\par }{\QB but you will cry for sorrow of heart,
\par }{\QB and will wail for anguish of spirit.
\par }{\Q \VS{15}You will leave your name for a curse to my chosen;
\par }{\QB and the Lord Yahweh will kill you.
\par }{\Q He will call his servants by another name,
\par }{\QB \VS{16}so that he who blesses himself in the earth will bless himself in the God of truth;
\par }{\QB and he who swears in the earth will swear by the God of truth;
\par }{\Q because the former troubles are forgotten,
\par }{\QB and because they are hidden from my eyes.
\par }{\BB \par }{\Q \VS{17}“For, behold, I create new heavens and a new earth;
\par }{\QB and the former things will not be remembered,
\par }{\QB nor come into mind.
\par }{\Q \VS{18}But be glad and rejoice forever in that which I create;
\par }{\QB for, behold, I create Jerusalem to be a delight,
\par }{\QB and her people a joy.
\par }{\Q \VS{19}I will rejoice in Jerusalem,
\par }{\QB and delight in my people;
\par }{\Q and the voice of weeping and the voice of crying
\par }{\QB will be heard in her no more.
\par }{\BB \par }{\Q \VS{20}“No more will there be an infant who only lives a few days,
\par }{\QB nor an old man who has not filled his days;
\par }{\Q for the child will die one hundred years old,
\par }{\QB and the sinner being one hundred years old will be accursed.
\par }{\Q \VS{21}They will build houses and inhabit them.
\par }{\QB They will plant vineyards and eat their fruit.
\par }{\Q \VS{22}They will not build and another inhabit.
\par }{\QB They will not plant and another eat:
\par }{\Q for the days of my people will be like the days of a tree,
\par }{\QB and my chosen will long enjoy the work of their hands.
\par }{\Q \VS{23}They will not labor in vain
\par }{\QB nor give birth for calamity;
\par }{\Q for they are the offspring of Yahweh’s blessed
\par }{\QB and their descendants with them.
\par }{\Q \VS{24}It will happen that before they call, I will answer;
\par }{\QB and while they are yet speaking, I will hear.
\par }{\Q \VS{25}The wolf and the lamb will feed together.
\par }{\QB The lion will eat straw like the ox.
\par }{\QB Dust will be the serpent’s food.
\par }{\Q They will not hurt nor destroy in all my holy mountain,”
\par }{\QB says Yahweh.

\par }\Chap{66}{\PP \VerseOne{1}Yahweh says,
\par }{\Q “Heaven is my throne, and the earth is my footstool.
\par }{\QB What kind of house will you build to me?
\par }{\QB Where will I rest?
\par }{\Q \VS{2}For my hand has made all these things,
\par }{\QB and so all these things came to be,” says Yahweh:
\par }{\Q “but I will look to this man,
\par }{\QB even to he who is poor and of a contrite spirit,
\par }{\QB and who trembles at my word.
\par }{\Q \VS{3}He who kills an ox is as he who kills a man;
\par }{\QB he who sacrifices a lamb, as he who breaks a dog’s neck;
\par }{\QB he who offers an offering, as he who offers pig’s blood;
\par }{\QB he who burns frankincense, as he who blesses an idol.
\par }{\Q Yes, they have chosen their own ways,
\par }{\QB and their soul delights in their abominations:
\par }{\Q \VS{4}I also will choose their delusions,
\par }{\QB and will bring their fears on them;
\par }{\Q because when I called, no one answered;
\par }{\QB when I spoke, they didn’t listen;
\par }{\Q but they did that which was evil in my eyes,
\par }{\QB and chose that in which I didn’t delight.”
\par }{\BB \par }{\Q \VS{5}Hear Yahweh’s word,
\par }{\QB you who tremble at his word:
\par }{\Q “Your brothers who hate you,
\par }{\QB who cast you out for my name’s sake, have said,
\par }{\Q ‘Let Yahweh be glorified,
\par }{\QB that we may see your joy;’
\par }{\QB but it is those who shall be disappointed.
\par }{\Q \VS{6}A voice of tumult from the city,
\par }{\QB a voice from the temple,
\par }{\QB a voice of Yahweh that repays his enemies what they deserve.
\par }{\BB \par }{\Q \VS{7}“Before she travailed, she gave birth.
\par }{\QB Before her pain came, she delivered a son.
\par }{\Q \VS{8}Who has heard of such a thing?
\par }{\QB Who has seen such things?
\par }{\Q Shall a land be born in one day?
\par }{\QB Shall a nation be born at once?
\par }{\Q For as soon as Zion travailed,
\par }{\QB she gave birth to her children.
\par }{\Q \VS{9}Shall I bring to the birth, and not cause to be delivered?” says Yahweh.
\par }{\QB “Shall I who cause to give birth shut the womb?” says your God.
\par }{\BB \par }{\Q \VS{10}“Rejoice with Jerusalem, and be glad for her, all you who love her.
\par }{\QB Rejoice for joy with her, all you who mourn over her;
\par }{\Q \VS{11}that you may nurse and be satisfied at the comforting breasts;
\par }{\QB that you may drink deeply,
\par }{\QB and be delighted with the abundance of her glory.”
\par }{\BB \par }{\Q \VS{12}For Yahweh says, “Behold, I will extend peace to her like a river,
\par }{\QB and the glory of the nations like an overflowing stream;
\par }{\QB and you will nurse.
\par }{\Q You will be carried on her side,
\par }{\QB and will be dandled on her knees.
\par }{\Q \VS{13}As one whom his mother comforts,
\par }{\QB so I will comfort you.
\par }{\QB You will be comforted in Jerusalem.”
\par }{\BB \par }{\Q \VS{14}You will see it, and your heart shall rejoice,
\par }{\QB and your bones will flourish like the tender grass.
\par }{\Q Yahweh’s hand will be known among his servants;
\par }{\QB and he will have indignation against his enemies.
\par }{\BB \par }{\Q \VS{15}For, behold, Yahweh will come with fire,
\par }{\QB and his chariots will be like the whirlwind;
\par }{\Q to render his anger with fierceness,
\par }{\QB and his rebuke with flames of fire.
\par }{\Q \VS{16}For Yahweh will execute judgment by fire and by his sword on all flesh;
\par }{\QB and those slain by Yahweh will be many.
\par }{\PP \VS{17}“Those who sanctify themselves and purify themselves to go to the gardens, behind one in the middle, eating pig’s meat, abominable things, and the mouse, they shall come to an end together,” says Yahweh.
\par }{\PP \VS{18}“For I know their works and their thoughts. The time comes that I will gather all nations and languages, and they will come, and will see my glory.
\par }{\PP \VS{19}“I will set a sign among them, and I will send those who escape of them to the nations, to Tarshish, Pul, and Lud, who draw the bow, to Tubal and Javan, to far-away islands, who have not heard my fame, nor have seen my glory; and they shall declare my glory among the nations.
\VS{20}They shall bring all your brothers out of all the nations for an offering to Yahweh, on horses, in chariots, in litters, on mules, and on camels, to my holy mountain Jerusalem, says Yahweh, as the children of Israel bring their offering in a clean vessel into Yahweh’s house.
\VS{21}Of them I will also select priests and Levites,” says Yahweh.
\par }{\PP \VS{22}“For as the new heavens and the new earth, which I will make, shall remain before me,” says Yahweh, “so your offspring and your name shall remain.
\VS{23}It shall happen that from one new moon to another, and from one Sabbath to another, all flesh will come to worship before me,” says Yahweh.
\VS{24}“They will go out, and look at the dead bodies of the men who have transgressed against me; for their worm will not die, nor will their fire be quenched, and they will be loathsome to all mankind.”
\par }