\NormalFont\ShortTitle{Philemon}
{\MT Paul’s Letter to Philemon

\par }\OneChap {\PP \VerseOne{1}Paul, a prisoner of Christ Jesus, and Timothy our brother, to Philemon, our beloved fellow worker,
\VS{2}to the beloved Apphia, to Archippus our fellow soldier, and to the assembly in your house:
\VS{3}Grace to you and peace from God our Father and the Lord Jesus Christ.
\par }{\PP \VS{4}I thank my God always, making mention of you in my prayers,
\VS{5}hearing of your love and of the faith which you have toward the Lord Jesus, and toward all the saints,
\VS{6}that the fellowship of your faith may become effective in the knowledge of every good thing which is in us in Christ Jesus.
\VS{7}For we have much joy and comfort in your love, because the hearts of the saints have been refreshed through you, brother.
\par }{\PP \VS{8}Therefore though I have all boldness in Christ to command you that which is appropriate,
\VS{9}yet for love’s sake I rather beg, being such a one as Paul, the aged, but also a prisoner of Jesus Christ.
\VS{10}I beg you for my child, whom I have become the father of in my chains, Onesimus,
\FTNT{*}{{\FR 1:10 }
Onesimus means “useful”.}
\VS{11}who once was useless to you, but now is useful to you and to me.
\VS{12}I am sending him back. Therefore receive him, that is, my own heart,
\VS{13}whom I desired to keep with me, that on your behalf he might serve me in my chains for the Good News.
\VS{14}But I was willing to do nothing without your consent, that your goodness would not be as of necessity, but of free will.
\VS{15}For perhaps he was therefore separated from you for a while, that you would have him forever,
\VS{16}no longer as a slave, but more than a slave, a beloved brother, especially to me, but how much rather to you, both in the flesh and in the Lord.
\par }{\PP \VS{17}If then you count me a partner, receive him as you would receive me.
\VS{18}But if he has wronged you at all or owes you anything, put that to my account.
\VS{19}I, Paul, write this with my own hand: I will repay it (not to mention to you that you owe to me even your own self besides).
\VS{20}Yes, brother, let me have joy from you in the Lord. Refresh my heart in the Lord.
\par }{\PP \VS{21}Having confidence in your obedience, I write to you, knowing that you will do even beyond what I say.
\par }{\PP \VS{22}Also, prepare a guest room for me, for I hope that through your prayers I will be restored to you.
\par }{\PP \VS{23}Epaphras, my fellow prisoner in Christ Jesus, greets you,
\VS{24}as do Mark, Aristarchus, Demas, and Luke, my fellow workers.
\par }{\PP \VS{25}The grace of our Lord Jesus Christ be with your spirit. Amen.
\par }