\NormalFont\ShortTitle{Proverbs}
{\MT The Proverbs

\par }\ChapOne{1}{\PP \VerseOne{1}The proverbs of Solomon, the son of David, king of Israel:
\par }{\Q \VS{2}to know wisdom and instruction;
\par }{\QB to discern the words of understanding;
\par }{\Q \VS{3}to receive instruction in wise dealing,
\par }{\QB in righteousness, justice, and equity;
\par }{\Q \VS{4}to give prudence to the simple,
\par }{\QB knowledge and discretion to the young man:
\par }{\Q \VS{5}that the wise man may hear, and increase in learning;
\par }{\QB that the man of understanding may attain to sound counsel:
\par }{\Q \VS{6}to understand a proverb, and parables,
\par }{\QB the words and riddles of the wise.
\par }{\BB \par }{\Q \VS{7}The fear of Yahweh\FTNT{*}{{\FR 1:7 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.} is the beginning of knowledge;
\par }{\QB but the foolish despise wisdom and instruction.
\par }{\Q \VS{8}My son, listen to your father’s instruction,
\par }{\QB and don’t forsake your mother’s teaching:
\par }{\Q \VS{9}for they will be a garland to grace your head,
\par }{\QB and chains around your neck.
\par }{\Q \VS{10}My son, if sinners entice you,
\par }{\QB don’t consent.
\par }{\Q \VS{11}If they say, “Come with us.
\par }{\QB Let’s lay in wait for blood.
\par }{\QB Let’s lurk secretly for the innocent without cause.
\par }{\Q \VS{12}Let’s swallow them up alive like Sheol,\FTNT{†}{{\FR 1:12 }
Sheol is the place of the dead.}
\par }{\QB and whole, like those who go down into the pit.
\par }{\Q \VS{13}We’ll find all valuable wealth.
\par }{\QB We’ll fill our houses with plunder.
\par }{\Q \VS{14}You shall cast your lot among us.
\par }{\QB We’ll all have one purse.”
\par }{\Q \VS{15}My son, don’t walk on the path with them.
\par }{\QB Keep your foot from their path,
\par }{\Q \VS{16}for their feet run to evil.
\par }{\QB They hurry to shed blood.
\par }{\Q \VS{17}For the net is spread in vain in the sight of any bird;
\par }{\Q \VS{18}but these lay in wait for their own blood.
\par }{\QB They lurk secretly for their own lives.
\par }{\Q \VS{19}So are the ways of everyone who is greedy for gain.
\par }{\QB It takes away the life of its owners.
\par }{\BB \par }{\Q \VS{20}Wisdom calls aloud in the street.
\par }{\QB She utters her voice in the public squares.
\par }{\Q \VS{21}She calls at the head of noisy places.
\par }{\QB At the entrance of the city gates, she utters her words:
\par }{\Q \VS{22}“How long, you simple ones, will you love simplicity?
\par }{\QB How long will mockers delight themselves in mockery,
\par }{\QB and fools hate knowledge?
\par }{\Q \VS{23}Turn at my reproof.
\par }{\QB Behold,\FTNT{‡}{{\FR 1:23 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} I will pour out my spirit on you.
\par }{\QB I will make known my words to you.
\par }{\Q \VS{24}Because I have called, and you have refused;
\par }{\QB I have stretched out my hand, and no one has paid attention;
\par }{\Q \VS{25}but you have ignored all my counsel,
\par }{\QB and wanted none of my reproof;
\par }{\Q \VS{26}I also will laugh at your disaster.
\par }{\QB I will mock when calamity overtakes you,
\par }{\Q \VS{27}when calamity overtakes you like a storm,
\par }{\QB when your disaster comes on like a whirlwind,
\par }{\QB when distress and anguish come on you.
\par }{\Q \VS{28}Then they will call on me, but I will not answer.
\par }{\QB They will seek me diligently, but they will not find me;
\par }{\Q \VS{29}because they hated knowledge,
\par }{\QB and didn’t choose the fear of Yahweh.
\par }{\Q \VS{30}They wanted none of my counsel.
\par }{\QB They despised all my reproof.
\par }{\Q \VS{31}Therefore they will eat of the fruit of their own way,
\par }{\QB and be filled with their own schemes.
\par }{\Q \VS{32}For the backsliding of the simple will kill them.
\par }{\QB The careless ease of fools will destroy them.
\par }{\Q \VS{33}But whoever listens to me will dwell securely,
\par }{\QB and will be at ease, without fear of harm.”

\par }\PoetryChap{2}{\Q \VerseOne{1}My son, if you will receive my words,
\par }{\QB and store up my commandments within you,
\par }{\Q \VS{2}so as to turn your ear to wisdom,
\par }{\QB and apply your heart to understanding;
\par }{\Q \VS{3}yes, if you call out for discernment,
\par }{\QB and lift up your voice for understanding;
\par }{\Q \VS{4}if you seek her as silver,
\par }{\QB and search for her as for hidden treasures:
\par }{\Q \VS{5}then you will understand the fear of Yahweh,
\par }{\QB and find the knowledge of God.\FTNT{*}{{\FR 2:5 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).}
\par }{\Q \VS{6}For Yahweh gives wisdom.
\par }{\QB Out of his mouth comes knowledge and understanding.
\par }{\Q \VS{7}He lays up sound wisdom for the upright.
\par }{\QB He is a shield to those who walk in integrity,
\par }{\Q \VS{8}that he may guard the paths of justice,
\par }{\QB and preserve the way of his saints.
\par }{\Q \VS{9}Then you will understand righteousness and justice,
\par }{\QB equity and every good path.
\par }{\Q \VS{10}For wisdom will enter into your heart.
\par }{\QB Knowledge will be pleasant to your soul.
\par }{\Q \VS{11}Discretion will watch over you.
\par }{\QB Understanding will keep you,
\par }{\Q \VS{12}to deliver you from the way of evil,
\par }{\QB from the men who speak perverse things,
\par }{\Q \VS{13}who forsake the paths of uprightness,
\par }{\QB to walk in the ways of darkness,
\par }{\Q \VS{14}who rejoice to do evil,
\par }{\QB and delight in the perverseness of evil,
\par }{\Q \VS{15}who are crooked in their ways,
\par }{\QB and wayward in their paths,
\par }{\Q \VS{16}to deliver you from the strange woman,
\par }{\QB even from the foreigner who flatters with her words,
\par }{\Q \VS{17}who forsakes the friend of her youth,
\par }{\QB and forgets the covenant of her God;
\par }{\Q \VS{18}for her house leads down to death,
\par }{\QB her paths to the departed spirits.
\par }{\Q \VS{19}None who go to her return again,
\par }{\QB neither do they attain to the paths of life.
\par }{\Q \VS{20}So you may walk in the way of good men,
\par }{\QB and keep the paths of the righteous.
\par }{\Q \VS{21}For the upright will dwell in the land.
\par }{\QB The perfect will remain in it.
\par }{\Q \VS{22}But the wicked will be cut off from the land.
\par }{\QB The treacherous will be rooted out of it.

\par }\PoetryChap{3}{\Q \VerseOne{1}My son, don’t forget my teaching;
\par }{\QB but let your heart keep my commandments:
\par }{\Q \VS{2}for they will add to you length of days,
\par }{\QB years of life, and peace.
\par }{\Q \VS{3}Don’t let kindness and truth forsake you.
\par }{\QB Bind them around your neck.
\par }{\QB Write them on the tablet of your heart.
\par }{\Q \VS{4}So you will find favor,
\par }{\QB and good understanding in the sight of God and man.
\par }{\Q \VS{5}Trust in Yahweh with all your heart,
\par }{\QB and don’t lean on your own understanding.
\par }{\Q \VS{6}In all your ways acknowledge him,
\par }{\QB and he will make your paths straight.
\par }{\Q \VS{7}Don’t be wise in your own eyes.
\par }{\QB Fear Yahweh, and depart from evil.
\par }{\Q \VS{8}It will be health to your body,
\par }{\QB and nourishment to your bones.
\par }{\Q \VS{9}Honor Yahweh with your substance,
\par }{\QB with the first fruits of all your increase:
\par }{\Q \VS{10}so your barns will be filled with plenty,
\par }{\QB and your vats will overflow with new wine.
\par }{\Q \VS{11}My son, don’t despise Yahweh’s discipline,
\par }{\QB neither be weary of his correction;
\par }{\Q \VS{12}for whom Yahweh loves, he corrects,
\par }{\QB even as a father reproves the son in whom he delights.
\par }{\BB \par }{\Q \VS{13}Happy is the man who finds wisdom,
\par }{\QB the man who gets understanding.
\par }{\Q \VS{14}For her good profit is better than getting silver,
\par }{\QB and her return is better than fine gold.
\par }{\Q \VS{15}She is more precious than rubies.
\par }{\QB None of the things you can desire are to be compared to her.
\par }{\Q \VS{16}Length of days is in her right hand.
\par }{\QB In her left hand are riches and honor.
\par }{\Q \VS{17}Her ways are ways of pleasantness.
\par }{\QB All her paths are peace.
\par }{\Q \VS{18}She is a tree of life to those who lay hold of her.
\par }{\QB Happy is everyone who retains her.
\par }{\Q \VS{19}By wisdom Yahweh founded the earth.
\par }{\QB By understanding, he established the heavens.
\par }{\Q \VS{20}By his knowledge, the depths were broken up,
\par }{\QB and the skies drop down the dew.
\par }{\Q \VS{21}My son, let them not depart from your eyes.
\par }{\QB Keep sound wisdom and discretion:
\par }{\Q \VS{22}so they will be life to your soul,
\par }{\QB and grace for your neck.
\par }{\Q \VS{23}Then you shall walk in your way securely.
\par }{\QB Your foot won’t stumble.
\par }{\Q \VS{24}When you lie down, you will not be afraid.
\par }{\QB Yes, you will lie down, and your sleep will be sweet.
\par }{\Q \VS{25}Don’t be afraid of sudden fear,
\par }{\QB neither of the desolation of the wicked, when it comes;
\par }{\Q \VS{26}for Yahweh will be your confidence,
\par }{\QB and will keep your foot from being taken.
\par }{\BB \par }{\Q \VS{27}Don’t withhold good from those to whom it is due,
\par }{\QB when it is in the power of your hand to do it.
\par }{\Q \VS{28}Don’t say to your neighbor, “Go, and come again;
\par }{\QB tomorrow I will give it to you,”
\par }{\QB when you have it by you.
\par }{\Q \VS{29}Don’t devise evil against your neighbor,
\par }{\QB since he dwells securely by you.
\par }{\Q \VS{30}Don’t strive with a man without cause,
\par }{\QB if he has done you no harm.
\par }{\Q \VS{31}Don’t envy the man of violence.
\par }{\QB Choose none of his ways.
\par }{\Q \VS{32}For the perverse is an abomination to Yahweh,
\par }{\QB but his friendship is with the upright.
\par }{\Q \VS{33}Yahweh’s curse is in the house of the wicked,
\par }{\QB but he blesses the habitation of the righteous.
\par }{\Q \VS{34}Surely he mocks the mockers,
\par }{\QB but he gives grace to the humble.
\par }{\Q \VS{35}The wise will inherit glory,
\par }{\QB but shame will be the promotion of fools.

\par }\PoetryChap{4}{\Q \VerseOne{1}Listen, sons, to a father’s instruction.
\par }{\QB Pay attention and know understanding;
\par }{\Q \VS{2}for I give you sound learning.
\par }{\QB Don’t forsake my law.
\par }{\Q \VS{3}For I was a son to my father,
\par }{\QB tender and an only child in the sight of my mother.
\par }{\Q \VS{4}He taught me, and said to me:
\par }{\QB “Let your heart retain my words.
\par }{\QB Keep my commandments, and live.
\par }{\Q \VS{5}Get wisdom.
\par }{\QB Get understanding.
\par }{\QB Don’t forget, and don’t deviate from the words of my mouth.
\par }{\Q \VS{6}Don’t forsake her, and she will preserve you.
\par }{\QB Love her, and she will keep you.
\par }{\Q \VS{7}Wisdom is supreme.
\par }{\QB Get wisdom.
\par }{\QB Yes, though it costs all your possessions, get understanding.
\par }{\Q \VS{8}Esteem her, and she will exalt you.
\par }{\QB She will bring you to honor when you embrace her.
\par }{\Q \VS{9}She will give to your head a garland of grace.
\par }{\QB She will deliver a crown of splendor to you.”
\par }{\BB \par }{\Q \VS{10}Listen, my son, and receive my sayings.
\par }{\QB The years of your life will be many.
\par }{\Q \VS{11}I have taught you in the way of wisdom.
\par }{\QB I have led you in straight paths.
\par }{\Q \VS{12}When you go, your steps will not be hampered.
\par }{\QB When you run, you will not stumble.
\par }{\Q \VS{13}Take firm hold of instruction.
\par }{\QB Don’t let her go.
\par }{\QB Keep her, for she is your life.
\par }{\Q \VS{14}Don’t enter into the path of the wicked.
\par }{\QB Don’t walk in the way of evil men.
\par }{\Q \VS{15}Avoid it, and don’t pass by it.
\par }{\QB Turn from it, and pass on.
\par }{\Q \VS{16}For they don’t sleep unless they do evil.
\par }{\QB Their sleep is taken away, unless they make someone fall.
\par }{\Q \VS{17}For they eat the bread of wickedness
\par }{\QB and drink the wine of violence.
\par }{\Q \VS{18}But the path of the righteous is like the dawning light
\par }{\QB that shines more and more until the perfect day.
\par }{\Q \VS{19}The way of the wicked is like darkness.
\par }{\QB They don’t know what they stumble over.
\par }{\BB \par }{\Q \VS{20}My son, attend to my words.
\par }{\QB Turn your ear to my sayings.
\par }{\Q \VS{21}Let them not depart from your eyes.
\par }{\QB Keep them in the center of your heart.
\par }{\Q \VS{22}For they are life to those who find them,
\par }{\QB and health to their whole body.
\par }{\Q \VS{23}Keep your heart with all diligence,
\par }{\QB for out of it is the wellspring of life.
\par }{\Q \VS{24}Put away from yourself a perverse mouth.
\par }{\QB Put corrupt lips far from you.
\par }{\Q \VS{25}Let your eyes look straight ahead.
\par }{\QB Fix your gaze directly before you.
\par }{\Q \VS{26}Make the path of your feet level.
\par }{\QB Let all of your ways be established.
\par }{\Q \VS{27}Don’t turn to the right hand nor to the left.
\par }{\QB Remove your foot from evil.

\par }\PoetryChap{5}{\Q \VerseOne{1}My son, pay attention to my wisdom.
\par }{\QB Turn your ear to my understanding,
\par }{\Q \VS{2}that you may maintain discretion,
\par }{\QB that your lips may preserve knowledge.
\par }{\Q \VS{3}For the lips of an adulteress drip honey.
\par }{\QB Her mouth is smoother than oil,
\par }{\Q \VS{4}but in the end she is as bitter as wormwood,
\par }{\QB and as sharp as a two-edged sword.
\par }{\Q \VS{5}Her feet go down to death.
\par }{\QB Her steps lead straight to Sheol.\FTNT{*}{{\FR 5:5 }
Sheol is the place of the dead.
}
\par }{\Q \VS{6}She gives no thought to the way of life.
\par }{\QB Her ways are crooked, and she doesn’t know it.
\par }{\BB \par }{\Q \VS{7}Now therefore, my sons, listen to me.
\par }{\QB Don’t depart from the words of my mouth.
\par }{\Q \VS{8}Remove your way far from her.
\par }{\QB Don’t come near the door of her house,
\par }{\Q \VS{9}lest you give your honor to others,
\par }{\QB and your years to the cruel one;
\par }{\Q \VS{10}lest strangers feast on your wealth,
\par }{\QB and your labors enrich another man’s house.
\par }{\Q \VS{11}You will groan at your latter end,
\par }{\QB when your flesh and your body are consumed,
\par }{\Q \VS{12}and say, “How I have hated instruction,
\par }{\QB and my heart despised reproof;
\par }{\Q \VS{13}neither have I obeyed the voice of my teachers,
\par }{\QB nor turned my ear to those who instructed me!
\par }{\Q \VS{14}I have come to the brink of utter ruin,
\par }{\QB among the gathered assembly.”
\par }{\BB \par }{\Q \VS{15}Drink water out of your own cistern,
\par }{\QB running water out of your own well.
\par }{\Q \VS{16}Should your springs overflow in the streets,
\par }{\QB streams of water in the public squares?
\par }{\Q \VS{17}Let them be for yourself alone,
\par }{\QB not for strangers with you.
\par }{\Q \VS{18}Let your spring be blessed.
\par }{\QB Rejoice in the wife of your youth.
\par }{\Q \VS{19}A loving doe and a graceful deer—
\par }{\QB let her breasts satisfy you at all times.
\par }{\QB Be captivated always with her love.
\par }{\Q \VS{20}For why should you, my son, be captivated with an adulteress?
\par }{\QB Why embrace the bosom of another?
\par }{\Q \VS{21}For the ways of man are before Yahweh’s eyes.
\par }{\QB He examines all his paths.
\par }{\Q \VS{22}The evil deeds of the wicked ensnare him.
\par }{\QB The cords of his sin hold him firmly.
\par }{\Q \VS{23}He will die for lack of instruction.
\par }{\QB In the greatness of his folly, he will go astray.

\par }\PoetryChap{6}{\Q \VerseOne{1}My son, if you have become collateral for your neighbor,
\par }{\QB if you have struck your hands in pledge for a stranger,
\par }{\Q \VS{2}you are trapped by the words of your mouth;
\par }{\QB you are ensnared with the words of your mouth.
\par }{\Q \VS{3}Do this now, my son, and deliver yourself,
\par }{\QB since you have come into the hand of your neighbor.
\par }{\Q Go, humble yourself.
\par }{\QB Press your plea with your neighbor.
\par }{\Q \VS{4}Give no sleep to your eyes,
\par }{\QB nor slumber to your eyelids.
\par }{\Q \VS{5}Free yourself, like a gazelle from the hand of the hunter,
\par }{\QB like a bird from the snare of the fowler.
\par }{\BB \par }{\Q \VS{6}Go to the ant, you sluggard.
\par }{\QB Consider her ways, and be wise;
\par }{\Q \VS{7}which having no chief, overseer, or ruler,
\par }{\QB \VS{8}provides her bread in the summer,
\par }{\QB and gathers her food in the harvest.
\par }{\Q \VS{9}How long will you sleep, sluggard?
\par }{\QB When will you arise out of your sleep?
\par }{\Q \VS{10}A little sleep, a little slumber,
\par }{\QB a little folding of the hands to sleep:
\par }{\Q \VS{11}so your poverty will come as a robber,
\par }{\QB and your scarcity as an armed man.
\par }{\BB \par }{\Q \VS{12}A worthless person, a man of iniquity,
\par }{\QB is he who walks with a perverse mouth,
\par }{\Q \VS{13}who winks with his eyes, who signals with his feet,
\par }{\QB who motions with his fingers,
\par }{\Q \VS{14}in whose heart is perverseness,
\par }{\QB who devises evil continually,
\par }{\QB who always sows discord.
\par }{\Q \VS{15}Therefore his calamity will come suddenly.
\par }{\QB He will be broken suddenly, and that without remedy.
\par }{\BB \par }{\Q \VS{16}There are six things which Yahweh hates;
\par }{\QB yes, seven which are an abomination to him:
\par }{\Q \VS{17}arrogant eyes, a lying tongue,
\par }{\QB hands that shed innocent blood,
\par }{\Q \VS{18}a heart that devises wicked schemes,
\par }{\QB feet that are swift in running to mischief,
\par }{\Q \VS{19}a false witness who utters lies,
\par }{\QB and he who sows discord among brothers.
\par }{\BB \par }{\Q \VS{20}My son, keep your father’s commandment,
\par }{\QB and don’t forsake your mother’s teaching.
\par }{\Q \VS{21}Bind them continually on your heart.
\par }{\QB Tie them around your neck.
\par }{\Q \VS{22}When you walk, it will lead you.
\par }{\QB When you sleep, it will watch over you.
\par }{\QB When you awake, it will talk with you.
\par }{\Q \VS{23}For the commandment is a lamp,
\par }{\QB and the law is light.
\par }{\QB Reproofs of instruction are the way of life,
\par }{\Q \VS{24}to keep you from the immoral woman,
\par }{\QB from the flattery of the wayward wife’s tongue.
\par }{\Q \VS{25}Don’t lust after her beauty in your heart,
\par }{\QB neither let her captivate you with her eyelids.
\par }{\Q \VS{26}For a prostitute reduces you to a piece of bread.
\par }{\QB The adulteress hunts for your precious life.
\par }{\Q \VS{27}Can a man scoop fire into his lap,
\par }{\QB and his clothes not be burned?
\par }{\Q \VS{28}Or can one walk on hot coals,
\par }{\QB and his feet not be scorched?
\par }{\Q \VS{29}So is he who goes in to his neighbor’s wife.
\par }{\QB Whoever touches her will not be unpunished.
\par }{\Q \VS{30}Men don’t despise a thief
\par }{\QB if he steals to satisfy himself when he is hungry;
\par }{\Q \VS{31}but if he is found, he shall restore seven times.
\par }{\QB He shall give all the wealth of his house.
\par }{\Q \VS{32}He who commits adultery with a woman is void of understanding.
\par }{\QB He who does it destroys his own soul.
\par }{\Q \VS{33}He will get wounds and dishonor.
\par }{\QB His reproach will not be wiped away.
\par }{\Q \VS{34}For jealousy arouses the fury of the husband.
\par }{\QB He won’t spare in the day of vengeance.
\par }{\Q \VS{35}He won’t regard any ransom,
\par }{\QB neither will he rest content, though you give many gifts.

\par }\PoetryChap{7}{\Q \VerseOne{1}My son, keep my words.
\par }{\QB Lay up my commandments within you.
\par }{\Q \VS{2}Keep my commandments and live!
\par }{\QB Guard my teaching as the apple of your eye.
\par }{\Q \VS{3}Bind them on your fingers.
\par }{\QB Write them on the tablet of your heart.
\par }{\Q \VS{4}Tell wisdom, “You are my sister.”
\par }{\QB Call understanding your relative,
\par }{\Q \VS{5}that they may keep you from the strange woman,
\par }{\QB from the foreigner who flatters with her words.
\par }{\Q \VS{6}For at the window of my house,
\par }{\QB I looked out through my lattice.
\par }{\Q \VS{7}I saw among the simple ones.
\par }{\QB I discerned among the youths a young man void of understanding,
\par }{\Q \VS{8}passing through the street near her corner,
\par }{\QB he went the way to her house,
\par }{\Q \VS{9}in the twilight, in the evening of the day,
\par }{\QB in the middle of the night and in the darkness.
\par }{\Q \VS{10}Behold, there a woman met him with the attire of a prostitute,
\par }{\QB and with crafty intent.
\par }{\Q \VS{11}She is loud and defiant.
\par }{\QB Her feet don’t stay in her house.
\par }{\Q \VS{12}Now she is in the streets, now in the squares,
\par }{\QB and lurking at every corner.
\par }{\Q \VS{13}So she caught him, and kissed him.
\par }{\QB With an impudent face she said to him:
\par }{\Q \VS{14}“Sacrifices of peace offerings are with me.
\par }{\QB Today I have paid my vows.
\par }{\Q \VS{15}Therefore I came out to meet you,
\par }{\QB to diligently seek your face,
\par }{\QB and I have found you.
\par }{\Q \VS{16}I have spread my couch with carpets of tapestry,
\par }{\QB with striped cloths of the yarn of Egypt.
\par }{\Q \VS{17}I have perfumed my bed with myrrh, aloes, and cinnamon.
\par }{\Q \VS{18}Come, let’s take our fill of loving until the morning.
\par }{\QB Let’s solace ourselves with loving.
\par }{\Q \VS{19}For my husband isn’t at home.
\par }{\QB He has gone on a long journey.
\par }{\Q \VS{20}He has taken a bag of money with him.
\par }{\QB He will come home at the full moon.”
\par }{\Q \VS{21}With persuasive words, she led him astray.
\par }{\QB With the flattering of her lips, she seduced him.
\par }{\Q \VS{22}He followed her immediately,
\par }{\QB as an ox goes to the slaughter,
\par }{\QB as a fool stepping into a noose.
\par }{\Q \VS{23}Until an arrow strikes through his liver,
\par }{\QB as a bird hurries to the snare,
\par }{\QB and doesn’t know that it will cost his life.
\par }{\BB \par }{\Q \VS{24}Now therefore, sons, listen to me.
\par }{\QB Pay attention to the words of my mouth.
\par }{\Q \VS{25}Don’t let your heart turn to her ways.
\par }{\QB Don’t go astray in her paths,
\par }{\Q \VS{26}for she has thrown down many wounded.
\par }{\QB Yes, all her slain are a mighty army.
\par }{\Q \VS{27}Her house is the way to Sheol,\FTNT{*}{{\FR 7:27 }
Sheol is the place of the dead.
}
\par }{\QB going down to the rooms of death.

\par }\PoetryChap{8}{\Q \VerseOne{1}Doesn’t wisdom cry out?
\par }{\QB Doesn’t understanding raise her voice?
\par }{\Q \VS{2}On the top of high places by the way,
\par }{\QB where the paths meet, she stands.
\par }{\Q \VS{3}Beside the gates, at the entry of the city,
\par }{\QB at the entry doors, she cries aloud:
\par }{\Q \VS{4}“I call to you men!
\par }{\QB I send my voice to the sons of mankind.
\par }{\Q \VS{5}You simple, understand prudence!
\par }{\QB You fools, be of an understanding heart!
\par }{\Q \VS{6}Hear, for I will speak excellent things.
\par }{\QB The opening of my lips is for right things.
\par }{\Q \VS{7}For my mouth speaks truth.
\par }{\QB Wickedness is an abomination to my lips.
\par }{\Q \VS{8}All the words of my mouth are in righteousness.
\par }{\QB There is nothing crooked or perverse in them.
\par }{\Q \VS{9}They are all plain to him who understands,
\par }{\QB right to those who find knowledge.
\par }{\Q \VS{10}Receive my instruction rather than silver,
\par }{\QB knowledge rather than choice gold.
\par }{\Q \VS{11}For wisdom is better than rubies.
\par }{\QB All the things that may be desired can’t be compared to it.
\par }{\BB \par }{\Q \VS{12}“I, wisdom, have made prudence my dwelling.
\par }{\QB Find out knowledge and discretion.
\par }{\Q \VS{13}The fear of Yahweh is to hate evil.
\par }{\QB I hate pride, arrogance, the evil way, and the perverse mouth.
\par }{\Q \VS{14}Counsel and sound knowledge are mine.
\par }{\QB I have understanding and power.
\par }{\Q \VS{15}By me kings reign,
\par }{\QB and princes decree justice.
\par }{\Q \VS{16}By me princes rule,
\par }{\QB nobles, and all the righteous rulers of the earth.
\par }{\Q \VS{17}I love those who love me.
\par }{\QB Those who seek me diligently will find me.
\par }{\Q \VS{18}With me are riches, honor,
\par }{\QB enduring wealth, and prosperity.
\par }{\Q \VS{19}My fruit is better than gold, yes, than fine gold,
\par }{\QB my yield than choice silver.
\par }{\Q \VS{20}I walk in the way of righteousness,
\par }{\QB in the middle of the paths of justice,
\par }{\Q \VS{21}that I may give wealth to those who love me.
\par }{\QB I fill their treasuries.
\par }{\BB \par }{\Q \VS{22}“Yahweh possessed me in the beginning of his work,
\par }{\QB before his deeds of old.
\par }{\Q \VS{23}I was set up from everlasting, from the beginning,
\par }{\QB before the earth existed.
\par }{\Q \VS{24}When there were no depths, I was born,
\par }{\QB when there were no springs abounding with water.
\par }{\Q \VS{25}Before the mountains were settled in place,
\par }{\QB before the hills, I was born;
\par }{\Q \VS{26}while as yet he had not made the earth, nor the fields,
\par }{\QB nor the beginning of the dust of the world.
\par }{\Q \VS{27}When he established the heavens, I was there.
\par }{\QB When he set a circle on the surface of the deep,
\par }{\Q \VS{28}when he established the clouds above,
\par }{\QB when the springs of the deep became strong,
\par }{\Q \VS{29}when he gave to the sea its boundary,
\par }{\QB that the waters should not violate his commandment,
\par }{\QB when he marked out the foundations of the earth,
\par }{\Q \VS{30}then I was the craftsman by his side.
\par }{\QB I was a delight day by day,
\par }{\QB always rejoicing before him,
\par }{\Q \VS{31}rejoicing in his whole world.
\par }{\QB My delight was with the sons of men.
\par }{\BB \par }{\Q \VS{32}“Now therefore, my sons, listen to me,
\par }{\QB for blessed are those who keep my ways.
\par }{\Q \VS{33}Hear instruction, and be wise.
\par }{\QB Don’t refuse it.
\par }{\Q \VS{34}Blessed is the man who hears me,
\par }{\QB watching daily at my gates,
\par }{\QB waiting at my door posts.
\par }{\Q \VS{35}For whoever finds me, finds life,
\par }{\QB and will obtain favor from Yahweh.
\par }{\Q \VS{36}But he who sins against me wrongs his own soul.
\par }{\QB All those who hate me love death.”

\par }\PoetryChap{9}{\Q \VerseOne{1}Wisdom has built her house.
\par }{\QB She has carved out her seven pillars.
\par }{\Q \VS{2}She has prepared her meat.
\par }{\QB She has mixed her wine.
\par }{\QB She has also set her table.
\par }{\Q \VS{3}She has sent out her maidens.
\par }{\QB She cries from the highest places of the city:
\par }{\Q \VS{4}“Whoever is simple, let him turn in here!”
\par }{\QB As for him who is void of understanding, she says to him,
\par }{\Q \VS{5}“Come, eat some of my bread,
\par }{\QB Drink some of the wine which I have mixed!
\par }{\Q \VS{6}Leave your simple ways, and live.
\par }{\QB Walk in the way of understanding.”
\par }{\BB \par }{\Q \VS{7}One who corrects a mocker invites insult.
\par }{\QB One who reproves a wicked man invites abuse.
\par }{\Q \VS{8}Don’t reprove a scoffer, lest he hate you.
\par }{\QB Reprove a wise person, and he will love you.
\par }{\Q \VS{9}Instruct a wise person, and he will be still wiser.
\par }{\QB Teach a righteous person, and he will increase in learning.
\par }{\Q \VS{10}The fear of Yahweh is the beginning of wisdom.
\par }{\QB The knowledge of the Holy One is understanding.
\par }{\Q \VS{11}For by me your days will be multiplied.
\par }{\QB The years of your life will be increased.
\par }{\Q \VS{12}If you are wise, you are wise for yourself.
\par }{\QB If you mock, you alone will bear it.
\par }{\BB \par }{\Q \VS{13}The foolish woman is loud,
\par }{\QB undisciplined, and knows nothing.
\par }{\Q \VS{14}She sits at the door of her house,
\par }{\QB on a seat in the high places of the city,
\par }{\Q \VS{15}to call to those who pass by,
\par }{\QB who go straight on their ways,
\par }{\Q \VS{16}“Whoever is simple, let him turn in here.”
\par }{\QB as for him who is void of understanding, she says to him,
\par }{\Q \VS{17}“Stolen water is sweet.
\par }{\QB Food eaten in secret is pleasant.”
\par }{\Q \VS{18}But he doesn’t know that the departed spirits are there,
\par }{\QB that her guests are in the depths of Sheol.\FTNT{*}{{\FR 9:18 }
Sheol is the place of the dead.}

\par }\Chap{10}{\PP \VerseOne{1}The proverbs of Solomon.
\par }{\Q A wise son makes a glad father;
\par }{\QB but a foolish son brings grief to his mother.
\par }{\Q \VS{2}Treasures of wickedness profit nothing,
\par }{\QB but righteousness delivers from death.
\par }{\Q \VS{3}Yahweh will not allow the soul of the righteous to go hungry,
\par }{\QB but he thrusts away the desire of the wicked.
\par }{\Q \VS{4}He becomes poor who works with a lazy hand,
\par }{\QB but the hand of the diligent brings wealth.
\par }{\Q \VS{5}He who gathers in summer is a wise son,
\par }{\QB but he who sleeps during the harvest is a son who causes shame.
\par }{\Q \VS{6}Blessings are on the head of the righteous,
\par }{\QB but violence covers the mouth of the wicked.
\par }{\Q \VS{7}The memory of the righteous is blessed,
\par }{\QB but the name of the wicked will rot.
\par }{\Q \VS{8}The wise in heart accept commandments,
\par }{\QB but a chattering fool will fall.
\par }{\Q \VS{9}He who walks blamelessly walks surely,
\par }{\QB but he who perverts his ways will be found out.
\par }{\Q \VS{10}One winking with the eye causes sorrow,
\par }{\QB but a chattering fool will fall.
\par }{\Q \VS{11}The mouth of the righteous is a spring of life,
\par }{\QB but violence covers the mouth of the wicked.
\par }{\Q \VS{12}Hatred stirs up strife,
\par }{\QB but love covers all wrongs.
\par }{\Q \VS{13}Wisdom is found on the lips of him who has discernment,
\par }{\QB but a rod is for the back of him who is void of understanding.
\par }{\Q \VS{14}Wise men lay up knowledge,
\par }{\QB but the mouth of the foolish is near ruin.
\par }{\Q \VS{15}The rich man’s wealth is his strong city.
\par }{\QB The destruction of the poor is their poverty.
\par }{\Q \VS{16}The labor of the righteous leads to life.
\par }{\QB The increase of the wicked leads to sin.
\par }{\Q \VS{17}He is in the way of life who heeds correction,
\par }{\QB but he who forsakes reproof leads others astray.
\par }{\Q \VS{18}He who hides hatred has lying lips.
\par }{\QB He who utters a slander is a fool.
\par }{\Q \VS{19}In the multitude of words there is no lack of disobedience,
\par }{\QB but he who restrains his lips does wisely.
\par }{\Q \VS{20}The tongue of the righteous is like choice silver.
\par }{\QB The heart of the wicked is of little worth.
\par }{\Q \VS{21}The lips of the righteous feed many,
\par }{\QB but the foolish die for lack of understanding.
\par }{\Q \VS{22}Yahweh’s blessing brings wealth,
\par }{\QB and he adds no trouble to it.
\par }{\Q \VS{23}It is a fool’s pleasure to do wickedness,
\par }{\QB but wisdom is a man of understanding’s pleasure.
\par }{\Q \VS{24}What the wicked fear, will overtake them,
\par }{\QB but the desire of the righteous will be granted.
\par }{\Q \VS{25}When the whirlwind passes, the wicked is no more;
\par }{\QB but the righteous stand firm forever.
\par }{\Q \VS{26}As vinegar to the teeth, and as smoke to the eyes,
\par }{\QB so is the sluggard to those who send him.
\par }{\Q \VS{27}The fear of Yahweh prolongs days,
\par }{\QB but the years of the wicked shall be shortened.
\par }{\Q \VS{28}The prospect of the righteous is joy,
\par }{\QB but the hope of the wicked will perish.
\par }{\Q \VS{29}The way of Yahweh is a stronghold to the upright,
\par }{\QB but it is a destruction to the workers of iniquity.
\par }{\Q \VS{30}The righteous will never be removed,
\par }{\QB but the wicked will not dwell in the land.
\par }{\Q \VS{31}The mouth of the righteous produces wisdom,
\par }{\QB but the perverse tongue will be cut off.
\par }{\Q \VS{32}The lips of the righteous know what is acceptable,
\par }{\QB but the mouth of the wicked is perverse.

\par }\PoetryChap{11}{\Q \VerseOne{1}A false balance is an abomination to Yahweh,
\par }{\QB but accurate weights are his delight.
\par }{\Q \VS{2}When pride comes, then comes shame,
\par }{\QB but with humility comes wisdom.
\par }{\Q \VS{3}The integrity of the upright shall guide them,
\par }{\QB but the perverseness of the treacherous shall destroy them.
\par }{\Q \VS{4}Riches don’t profit in the day of wrath,
\par }{\QB but righteousness delivers from death.
\par }{\Q \VS{5}The righteousness of the blameless will direct his way,
\par }{\QB but the wicked shall fall by his own wickedness.
\par }{\Q \VS{6}The righteousness of the upright shall deliver them,
\par }{\QB but the unfaithful will be trapped by evil desires.
\par }{\Q \VS{7}When a wicked man dies, hope perishes,
\par }{\QB and expectation of power comes to nothing.
\par }{\Q \VS{8}A righteous person is delivered out of trouble,
\par }{\QB and the wicked takes his place.
\par }{\Q \VS{9}With his mouth the godless man destroys his neighbor,
\par }{\QB but the righteous will be delivered through knowledge.
\par }{\Q \VS{10}When it goes well with the righteous, the city rejoices.
\par }{\QB When the wicked perish, there is shouting.
\par }{\Q \VS{11}By the blessing of the upright, the city is exalted,
\par }{\QB but it is overthrown by the mouth of the wicked.
\par }{\Q \VS{12}One who despises his neighbor is void of wisdom,
\par }{\QB but a man of understanding holds his peace.
\par }{\Q \VS{13}One who brings gossip betrays a confidence,
\par }{\QB but one who is of a trustworthy spirit is one who keeps a secret.
\par }{\Q \VS{14}Where there is no wise guidance, the nation falls,
\par }{\QB but in the multitude of counselors there is victory.
\par }{\Q \VS{15}He who is collateral for a stranger will suffer for it,
\par }{\QB but he who refuses pledges of collateral is secure.
\par }{\Q \VS{16}A gracious woman obtains honor,
\par }{\QB but violent men obtain riches.
\par }{\Q \VS{17}The merciful man does good to his own soul,
\par }{\QB but he who is cruel troubles his own flesh.
\par }{\Q \VS{18}Wicked people earn deceitful wages,
\par }{\QB but one who sows righteousness reaps a sure reward.
\par }{\Q \VS{19}He who is truly righteous gets life.
\par }{\QB He who pursues evil gets death.
\par }{\Q \VS{20}Those who are perverse in heart are an abomination to Yahweh,
\par }{\QB but those whose ways are blameless are his delight.
\par }{\Q \VS{21}Most certainly, the evil man will not be unpunished,
\par }{\QB but the offspring\FTNT{*}{{\FR 11:21 }
or, seed} of the righteous will be delivered.
\par }{\Q \VS{22}Like a gold ring in a pig’s snout,
\par }{\QB is a beautiful woman who lacks discretion.
\par }{\Q \VS{23}The desire of the righteous is only good.
\par }{\QB The expectation of the wicked is wrath.
\par }{\Q \VS{24}There is one who scatters, and increases yet more.
\par }{\QB There is one who withholds more than is appropriate, but gains poverty.
\par }{\Q \VS{25}The liberal soul shall be made fat.
\par }{\QB He who waters shall be watered also himself.
\par }{\Q \VS{26}People curse someone who withholds grain,
\par }{\QB but blessing will be on the head of him who sells it.
\par }{\Q \VS{27}He who diligently seeks good seeks favor,
\par }{\QB but he who searches after evil, it shall come to him.
\par }{\Q \VS{28}He who trusts in his riches will fall,
\par }{\QB but the righteous shall flourish as the green leaf.
\par }{\Q \VS{29}He who troubles his own house shall inherit the wind.
\par }{\QB The foolish shall be servant to the wise of heart.
\par }{\Q \VS{30}The fruit of the righteous is a tree of life.
\par }{\QB He who is wise wins souls.
\par }{\Q \VS{31}Behold, the righteous shall be repaid in the earth,
\par }{\QB how much more the wicked and the sinner!

\par }\PoetryChap{12}{\Q \VerseOne{1}Whoever loves correction loves knowledge,
\par }{\QB but he who hates reproof is stupid.
\par }{\Q \VS{2}A good man shall obtain favor from Yahweh,
\par }{\QB but he will condemn a man of wicked plans.
\par }{\Q \VS{3}A man shall not be established by wickedness,
\par }{\QB but the root of the righteous shall not be moved.
\par }{\Q \VS{4}A worthy woman is the crown of her husband,
\par }{\QB but a disgraceful wife is as rottenness in his bones.
\par }{\Q \VS{5}The thoughts of the righteous are just,
\par }{\QB but the advice of the wicked is deceitful.
\par }{\Q \VS{6}The words of the wicked are about lying in wait for blood,
\par }{\QB but the speech of the upright rescues them.
\par }{\Q \VS{7}The wicked are overthrown, and are no more,
\par }{\QB but the house of the righteous shall stand.
\par }{\Q \VS{8}A man shall be commended according to his wisdom,
\par }{\QB but he who has a warped mind shall be despised.
\par }{\Q \VS{9}Better is he who is little known, and has a servant,
\par }{\QB than he who honors himself, and lacks bread.
\par }{\Q \VS{10}A righteous man respects the life of his animal,
\par }{\QB but the tender mercies of the wicked are cruel.
\par }{\Q \VS{11}He who tills his land shall have plenty of bread,
\par }{\QB but he who chases fantasies is void of understanding.
\par }{\Q \VS{12}The wicked desires the plunder of evil men,
\par }{\QB but the root of the righteous flourishes.
\par }{\Q \VS{13}An evil man is trapped by sinfulness of lips,
\par }{\QB but the righteous shall come out of trouble.
\par }{\Q \VS{14}A man shall be satisfied with good by the fruit of his mouth.
\par }{\QB The work of a man’s hands shall be rewarded to him.
\par }{\Q \VS{15}The way of a fool is right in his own eyes,
\par }{\QB but he who is wise listens to counsel.
\par }{\Q \VS{16}A fool shows his annoyance the same day,
\par }{\QB but one who overlooks an insult is prudent.
\par }{\Q \VS{17}He who is truthful testifies honestly,
\par }{\QB but a false witness lies.
\par }{\Q \VS{18}There is one who speaks rashly like the piercing of a sword,
\par }{\QB but the tongue of the wise heals.
\par }{\Q \VS{19}Truth’s lips will be established forever,
\par }{\QB but a lying tongue is only momentary.
\par }{\Q \VS{20}Deceit is in the heart of those who plot evil,
\par }{\QB but joy comes to the promoters of peace.
\par }{\Q \VS{21}No mischief shall happen to the righteous,
\par }{\QB but the wicked shall be filled with evil.
\par }{\Q \VS{22}Lying lips are an abomination to Yahweh,
\par }{\QB but those who do the truth are his delight.
\par }{\Q \VS{23}A prudent man keeps his knowledge,
\par }{\QB but the hearts of fools proclaim foolishness.
\par }{\Q \VS{24}The hands of the diligent ones shall rule,
\par }{\QB but laziness ends in slave labor.
\par }{\Q \VS{25}Anxiety in a man’s heart weighs it down,
\par }{\QB but a kind word makes it glad.
\par }{\Q \VS{26}A righteous person is cautious in friendship,
\par }{\QB but the way of the wicked leads them astray.
\par }{\Q \VS{27}The slothful man doesn’t roast his game,
\par }{\QB but the possessions of diligent men are prized.
\par }{\Q \VS{28}In the way of righteousness is life;
\par }{\QB in its path there is no death.

\par }\PoetryChap{13}{\Q \VerseOne{1}A wise son listens to his father’s instruction,
\par }{\QB but a scoffer doesn’t listen to rebuke.
\par }{\Q \VS{2}By the fruit of his lips, a man enjoys good things,
\par }{\QB but the unfaithful crave violence.
\par }{\Q \VS{3}He who guards his mouth guards his soul.
\par }{\QB One who opens wide his lips comes to ruin.
\par }{\Q \VS{4}The soul of the sluggard desires, and has nothing,
\par }{\QB but the desire of the diligent shall be fully satisfied.
\par }{\Q \VS{5}A righteous man hates lies,
\par }{\QB but a wicked man brings shame and disgrace.
\par }{\Q \VS{6}Righteousness guards the way of integrity,
\par }{\QB but wickedness overthrows the sinner.
\par }{\Q \VS{7}There are some who pretend to be rich, yet have nothing.
\par }{\QB There are some who pretend to be poor, yet have great wealth.
\par }{\Q \VS{8}The ransom of a man’s life is his riches,
\par }{\QB but the poor hear no threats.
\par }{\Q \VS{9}The light of the righteous shines brightly,
\par }{\QB but the lamp of the wicked is snuffed out.
\par }{\Q \VS{10}Pride only breeds quarrels,
\par }{\QB but wisdom is with people who take advice.
\par }{\Q \VS{11}Wealth gained dishonestly dwindles away,
\par }{\QB but he who gathers by hand makes it grow.
\par }{\Q \VS{12}Hope deferred makes the heart sick,
\par }{\QB but when longing is fulfilled, it is a tree of life.
\par }{\Q \VS{13}Whoever despises instruction will pay for it,
\par }{\QB but he who respects a command will be rewarded.
\par }{\Q \VS{14}The teaching of the wise is a spring of life,
\par }{\QB to turn from the snares of death.
\par }{\Q \VS{15}Good understanding wins favor,
\par }{\QB but the way of the unfaithful is hard.
\par }{\Q \VS{16}Every prudent man acts from knowledge,
\par }{\QB but a fool exposes folly.
\par }{\Q \VS{17}A wicked messenger falls into trouble,
\par }{\QB but a trustworthy envoy gains healing.
\par }{\Q \VS{18}Poverty and shame come to him who refuses discipline,
\par }{\QB but he who heeds correction shall be honored.
\par }{\Q \VS{19}Longing fulfilled is sweet to the soul,
\par }{\QB but fools detest turning from evil.
\par }{\Q \VS{20}One who walks with wise men grows wise,
\par }{\QB but a companion of fools suffers harm.
\par }{\Q \VS{21}Misfortune pursues sinners,
\par }{\QB but prosperity rewards the righteous.
\par }{\Q \VS{22}A good man leaves an inheritance to his children’s children,
\par }{\QB but the wealth of the sinner is stored for the righteous.
\par }{\Q \VS{23}An abundance of food is in poor people’s fields,
\par }{\QB but injustice sweeps it away.
\par }{\Q \VS{24}One who spares the rod hates his son,
\par }{\QB but one who loves him is careful to discipline him.
\par }{\Q \VS{25}The righteous one eats to the satisfying of his soul,
\par }{\QB but the belly of the wicked goes hungry.

\par }\PoetryChap{14}{\Q \VerseOne{1}Every wise woman builds her house,
\par }{\QB but the foolish one tears it down with her own hands.
\par }{\Q \VS{2}He who walks in his uprightness fears Yahweh,
\par }{\QB but he who is perverse in his ways despises him.
\par }{\Q \VS{3}The fool’s talk brings a rod to his back,
\par }{\QB but the lips of the wise protect them.
\par }{\Q \VS{4}Where no oxen are, the crib is clean,
\par }{\QB but much increase is by the strength of the ox.
\par }{\Q \VS{5}A truthful witness will not lie,
\par }{\QB but a false witness pours out lies.
\par }{\Q \VS{6}A scoffer seeks wisdom, and doesn’t find it,
\par }{\QB but knowledge comes easily to a discerning person.
\par }{\Q \VS{7}Stay away from a foolish man,
\par }{\QB for you won’t find knowledge on his lips.
\par }{\Q \VS{8}The wisdom of the prudent is to think about his way,
\par }{\QB but the folly of fools is deceit.
\par }{\Q \VS{9}Fools mock at making atonement for sins,
\par }{\QB but among the upright there is good will.
\par }{\Q \VS{10}The heart knows its own bitterness and joy;
\par }{\QB he will not share these with a stranger.
\par }{\Q \VS{11}The house of the wicked will be overthrown,
\par }{\QB but the tent of the upright will flourish.
\par }{\Q \VS{12}There is a way which seems right to a man,
\par }{\QB but in the end it leads to death.
\par }{\Q \VS{13}Even in laughter the heart may be sorrowful,
\par }{\QB and mirth may end in heaviness.
\par }{\Q \VS{14}The unfaithful will be repaid for his own ways;
\par }{\QB likewise a good man will be rewarded for his ways.
\par }{\Q \VS{15}A simple man believes everything,
\par }{\QB but the prudent man carefully considers his ways.
\par }{\Q \VS{16}A wise man fears and shuns evil,
\par }{\QB but the fool is hot headed and reckless.
\par }{\Q \VS{17}He who is quick to become angry will commit folly,
\par }{\QB and a crafty man is hated.
\par }{\Q \VS{18}The simple inherit folly,
\par }{\QB but the prudent are crowned with knowledge.
\par }{\Q \VS{19}The evil bow down before the good,
\par }{\QB and the wicked at the gates of the righteous.
\par }{\Q \VS{20}The poor person is shunned even by his own neighbor,
\par }{\QB but the rich person has many friends.
\par }{\Q \VS{21}He who despises his neighbor sins,
\par }{\QB but he who has pity on the poor is blessed.
\par }{\Q \VS{22}Don’t they go astray who plot evil?
\par }{\QB But love and faithfulness belong to those who plan good.
\par }{\Q \VS{23}In all hard work there is profit,
\par }{\QB but the talk of the lips leads only to poverty.
\par }{\Q \VS{24}The crown of the wise is their riches,
\par }{\QB but the folly of fools crowns them with folly.
\par }{\Q \VS{25}A truthful witness saves souls,
\par }{\QB but a false witness is deceitful.
\par }{\Q \VS{26}In the fear of Yahweh is a secure fortress,
\par }{\QB and he will be a refuge for his children.
\par }{\Q \VS{27}The fear of Yahweh is a fountain of life,
\par }{\QB turning people from the snares of death.
\par }{\Q \VS{28}In the multitude of people is the king’s glory,
\par }{\QB but in the lack of people is the destruction of the prince.
\par }{\Q \VS{29}He who is slow to anger has great understanding,
\par }{\QB but he who has a quick temper displays folly.
\par }{\Q \VS{30}The life of the body is a heart at peace,
\par }{\QB but envy rots the bones.
\par }{\Q \VS{31}He who oppresses the poor shows contempt for his Maker,
\par }{\QB but he who is kind to the needy honors him.
\par }{\Q \VS{32}The wicked is brought down in his calamity,
\par }{\QB but in death, the righteous has a refuge.
\par }{\Q \VS{33}Wisdom rests in the heart of one who has understanding,
\par }{\QB and is even made known in the inward part of fools.
\par }{\Q \VS{34}Righteousness exalts a nation,
\par }{\QB but sin is a disgrace to any people.
\par }{\Q \VS{35}The king’s favor is toward a servant who deals wisely,
\par }{\QB but his wrath is toward one who causes shame.

\par }\PoetryChap{15}{\Q \VerseOne{1}A gentle answer turns away wrath,
\par }{\QB but a harsh word stirs up anger.
\par }{\Q \VS{2}The tongue of the wise commends knowledge,
\par }{\QB but the mouth of fools gush out folly.
\par }{\Q \VS{3}Yahweh’s eyes are everywhere,
\par }{\QB keeping watch on the evil and the good.
\par }{\Q \VS{4}A gentle tongue is a tree of life,
\par }{\QB but deceit in it crushes the spirit.
\par }{\Q \VS{5}A fool despises his father’s correction,
\par }{\QB but he who heeds reproof shows prudence.
\par }{\Q \VS{6}In the house of the righteous is much treasure,
\par }{\QB but the income of the wicked brings trouble.
\par }{\Q \VS{7}The lips of the wise spread knowledge;
\par }{\QB not so with the heart of fools.
\par }{\Q \VS{8}The sacrifice made by the wicked is an abomination to Yahweh,
\par }{\QB but the prayer of the upright is his delight.
\par }{\Q \VS{9}The way of the wicked is an abomination to Yahweh,
\par }{\QB but he loves him who follows after righteousness.
\par }{\Q \VS{10}There is stern discipline for one who forsakes the way:
\par }{\QB whoever hates reproof shall die.
\par }{\Q \VS{11}Sheol\FTNT{*}{{\FR 15:11 }
Sheol is the place of the dead.} and Abaddon are before Yahweh—
\par }{\QB how much more then the hearts of the children of men!
\par }{\Q \VS{12}A scoffer doesn’t love to be reproved;
\par }{\QB he will not go to the wise.
\par }{\Q \VS{13}A glad heart makes a cheerful face,
\par }{\QB but an aching heart breaks the spirit.
\par }{\Q \VS{14}The heart of one who has understanding seeks knowledge,
\par }{\QB but the mouths of fools feed on folly.
\par }{\Q \VS{15}All the days of the afflicted are wretched,
\par }{\QB but one who has a cheerful heart enjoys a continual feast.
\par }{\Q \VS{16}Better is little, with the fear of Yahweh,
\par }{\QB than great treasure with trouble.
\par }{\Q \VS{17}Better is a dinner of herbs, where love is,
\par }{\QB than a fattened calf with hatred.
\par }{\Q \VS{18}A wrathful man stirs up contention,
\par }{\QB but one who is slow to anger appeases strife.
\par }{\Q \VS{19}The way of the sluggard is like a thorn patch,
\par }{\QB but the path of the upright is a highway.
\par }{\Q \VS{20}A wise son makes a father glad,
\par }{\QB but a foolish man despises his mother.
\par }{\Q \VS{21}Folly is joy to one who is void of wisdom,
\par }{\QB but a man of understanding keeps his way straight.
\par }{\Q \VS{22}Where there is no counsel, plans fail;
\par }{\QB but in a multitude of counselors they are established.
\par }{\Q \VS{23}Joy comes to a man with the reply of his mouth.
\par }{\QB How good is a word at the right time!
\par }{\Q \VS{24}The path of life leads upward for the wise,
\par }{\QB to keep him from going downward to Sheol.\FTNT{†}{{\FR 15:24 }
Sheol is the place of the dead.}
\par }{\Q \VS{25}Yahweh will uproot the house of the proud,
\par }{\QB but he will keep the widow’s borders intact.
\par }{\Q \VS{26}Yahweh detests the thoughts of the wicked,
\par }{\QB but the thoughts of the pure are pleasing.
\par }{\Q \VS{27}He who is greedy for gain troubles his own house,
\par }{\QB but he who hates bribes will live.
\par }{\Q \VS{28}The heart of the righteous weighs answers,
\par }{\QB but the mouth of the wicked gushes out evil.
\par }{\Q \VS{29}Yahweh is far from the wicked,
\par }{\QB but he hears the prayer of the righteous.
\par }{\Q \VS{30}The light of the eyes rejoices the heart.
\par }{\QB Good news gives health to the bones.
\par }{\Q \VS{31}The ear that listens to reproof lives,
\par }{\QB and will be at home among the wise.
\par }{\Q \VS{32}He who refuses correction despises his own soul,
\par }{\QB but he who listens to reproof gets understanding.
\par }{\Q \VS{33}The fear of Yahweh teaches wisdom.
\par }{\QB Before honor is humility.

\par }\PoetryChap{16}{\Q \VerseOne{1}The plans of the heart belong to man,
\par }{\QB but the answer of the tongue is from Yahweh.
\par }{\Q \VS{2}All the ways of a man are clean in his own eyes;
\par }{\QB but Yahweh weighs the motives.
\par }{\Q \VS{3}Commit your deeds to Yahweh,
\par }{\QB and your plans shall succeed.
\par }{\Q \VS{4}Yahweh has made everything for its own end—
\par }{\QB yes, even the wicked for the day of evil.
\par }{\Q \VS{5}Everyone who is proud in heart is an abomination to Yahweh:
\par }{\QB they shall certainly not be unpunished.
\par }{\Q \VS{6}By mercy and truth iniquity is atoned for.
\par }{\QB By the fear of Yahweh men depart from evil.
\par }{\Q \VS{7}When a man’s ways please Yahweh,
\par }{\QB he makes even his enemies to be at peace with him.
\par }{\Q \VS{8}Better is a little with righteousness,
\par }{\QB than great revenues with injustice.
\par }{\Q \VS{9}A man’s heart plans his course,
\par }{\QB but Yahweh directs his steps.
\par }{\Q \VS{10}Inspired judgments are on the lips of the king.
\par }{\QB He shall not betray his mouth.
\par }{\Q \VS{11}Honest balances and scales are Yahweh’s;
\par }{\QB all the weights in the bag are his work.
\par }{\Q \VS{12}It is an abomination for kings to do wrong,
\par }{\QB for the throne is established by righteousness.
\par }{\Q \VS{13}Righteous lips are the delight of kings.
\par }{\QB They value one who speaks the truth.
\par }{\Q \VS{14}The king’s wrath is a messenger of death,
\par }{\QB but a wise man will pacify it.
\par }{\Q \VS{15}In the light of the king’s face is life.
\par }{\QB His favor is like a cloud of the spring rain.
\par }{\Q \VS{16}How much better it is to get wisdom than gold!
\par }{\QB Yes, to get understanding is to be chosen rather than silver.
\par }{\Q \VS{17}The highway of the upright is to depart from evil.
\par }{\QB He who keeps his way preserves his soul.
\par }{\Q \VS{18}Pride goes before destruction,
\par }{\QB and an arrogant spirit before a fall.
\par }{\Q \VS{19}It is better to be of a lowly spirit with the poor,
\par }{\QB than to divide the plunder with the proud.
\par }{\Q \VS{20}He who heeds the Word finds prosperity.
\par }{\QB Whoever trusts in Yahweh is blessed.
\par }{\Q \VS{21}The wise in heart shall be called prudent.
\par }{\QB Pleasantness of the lips promotes instruction.
\par }{\Q \VS{22}Understanding is a fountain of life to one who has it,
\par }{\QB but the punishment of fools is their folly.
\par }{\Q \VS{23}The heart of the wise instructs his mouth,
\par }{\QB and adds learning to his lips.
\par }{\Q \VS{24}Pleasant words are a honeycomb,
\par }{\QB sweet to the soul, and health to the bones.
\par }{\Q \VS{25}There is a way which seems right to a man,
\par }{\QB but in the end it leads to death.
\par }{\Q \VS{26}The appetite of the laboring man labors for him;
\par }{\QB for his mouth urges him on.
\par }{\Q \VS{27}A worthless man devises mischief.
\par }{\QB His speech is like a scorching fire.
\par }{\Q \VS{28}A perverse man stirs up strife.
\par }{\QB A whisperer separates close friends.
\par }{\Q \VS{29}A man of violence entices his neighbor,
\par }{\QB and leads him in a way that is not good.
\par }{\Q \VS{30}One who winks his eyes to plot perversities,
\par }{\QB one who compresses his lips, is bent on evil.
\par }{\Q \VS{31}Gray hair is a crown of glory.
\par }{\QB It is attained by a life of righteousness.
\par }{\Q \VS{32}One who is slow to anger is better than the mighty;
\par }{\QB one who rules his spirit, than he who takes a city.
\par }{\Q \VS{33}The lot is cast into the lap,
\par }{\QB but its every decision is from Yahweh.

\par }\PoetryChap{17}{\Q \VerseOne{1}Better is a dry morsel with quietness,
\par }{\QB than a house full of feasting with strife.
\par }{\Q \VS{2}A servant who deals wisely will rule over a son who causes shame,
\par }{\QB and shall have a part in the inheritance among the brothers.
\par }{\Q \VS{3}The refining pot is for silver, and the furnace for gold,
\par }{\QB but Yahweh tests the hearts.
\par }{\Q \VS{4}An evildoer heeds wicked lips.
\par }{\QB A liar gives ear to a mischievous tongue.
\par }{\Q \VS{5}Whoever mocks the poor reproaches his Maker.
\par }{\QB He who is glad at calamity shall not be unpunished.
\par }{\Q \VS{6}Children’s children are the crown of old men;
\par }{\QB the glory of children are their parents.
\par }{\Q \VS{7}Arrogant speech isn’t fitting for a fool,
\par }{\QB much less do lying lips fit a prince.
\par }{\Q \VS{8}A bribe is a precious stone in the eyes of him who gives it;
\par }{\QB wherever he turns, he prospers.
\par }{\Q \VS{9}He who covers an offense promotes love;
\par }{\QB but he who repeats a matter separates best friends.
\par }{\Q \VS{10}A rebuke enters deeper into one who has understanding
\par }{\QB than a hundred lashes into a fool.
\par }{\Q \VS{11}An evil man seeks only rebellion;
\par }{\QB therefore a cruel messenger shall be sent against him.
\par }{\Q \VS{12}Let a bear robbed of her cubs meet a man,
\par }{\QB rather than a fool in his folly.
\par }{\Q \VS{13}Whoever rewards evil for good,
\par }{\QB evil shall not depart from his house.
\par }{\Q \VS{14}The beginning of strife is like breaching a dam,
\par }{\QB therefore stop contention before quarreling breaks out.
\par }{\Q \VS{15}He who justifies the wicked, and he who condemns the righteous,
\par }{\QB both of them alike are an abomination to Yahweh.
\par }{\Q \VS{16}Why is there money in the hand of a fool to buy wisdom,
\par }{\QB since he has no understanding?
\par }{\Q \VS{17}A friend loves at all times;
\par }{\QB and a brother is born for adversity.
\par }{\Q \VS{18}A man void of understanding strikes hands,
\par }{\QB and becomes collateral in the presence of his neighbor.
\par }{\Q \VS{19}He who loves disobedience loves strife.
\par }{\QB One who builds a high gate seeks destruction.
\par }{\Q \VS{20}One who has a perverse heart doesn’t find prosperity,
\par }{\QB and one who has a deceitful tongue falls into trouble.
\par }{\Q \VS{21}He who becomes the father of a fool grieves.
\par }{\QB The father of a fool has no joy.
\par }{\Q \VS{22}A cheerful heart makes good medicine,
\par }{\QB but a crushed spirit dries up the bones.
\par }{\Q \VS{23}A wicked man receives a bribe in secret,
\par }{\QB to pervert the ways of justice.
\par }{\Q \VS{24}Wisdom is before the face of one who has understanding,
\par }{\QB but the eyes of a fool wander to the ends of the earth.
\par }{\Q \VS{25}A foolish son brings grief to his father,
\par }{\QB and bitterness to her who bore him.
\par }{\Q \VS{26}Also to punish the righteous is not good,
\par }{\QB nor to flog officials for their integrity.
\par }{\Q \VS{27}He who spares his words has knowledge.
\par }{\QB He who is even tempered is a man of understanding.
\par }{\Q \VS{28}Even a fool, when he keeps silent, is counted wise.
\par }{\QB When he shuts his lips, he is thought to be discerning.

\par }\PoetryChap{18}{\Q \VerseOne{1}A man who isolates himself pursues selfishness,
\par }{\QB and defies all sound judgment.
\par }{\Q \VS{2}A fool has no delight in understanding,
\par }{\QB but only in revealing his own opinion.
\par }{\Q \VS{3}When wickedness comes, contempt also comes,
\par }{\QB and with shame comes disgrace.
\par }{\Q \VS{4}The words of a man’s mouth are like deep waters.
\par }{\QB The fountain of wisdom is like a flowing brook.
\par }{\Q \VS{5}To be partial to the faces of the wicked is not good,
\par }{\QB nor to deprive the innocent of justice.
\par }{\Q \VS{6}A fool’s lips come into strife,
\par }{\QB and his mouth invites beatings.
\par }{\Q \VS{7}A fool’s mouth is his destruction,
\par }{\QB and his lips are a snare to his soul.
\par }{\Q \VS{8}The words of a gossip are like dainty morsels:
\par }{\QB they go down into a person’s innermost parts.
\par }{\Q \VS{9}One who is slack in his work
\par }{\QB is brother to him who is a master of destruction.
\par }{\Q \VS{10}Yahweh’s name is a strong tower:
\par }{\QB the righteous run to him, and are safe.
\par }{\Q \VS{11}The rich man’s wealth is his strong city,
\par }{\QB like an unscalable wall in his own imagination.
\par }{\Q \VS{12}Before destruction the heart of man is proud,
\par }{\QB but before honor is humility.
\par }{\Q \VS{13}He who answers before he hears,
\par }{\QB that is folly and shame to him.
\par }{\Q \VS{14}A man’s spirit will sustain him in sickness,
\par }{\QB but a crushed spirit, who can bear?
\par }{\Q \VS{15}The heart of the discerning gets knowledge.
\par }{\QB The ear of the wise seeks knowledge.
\par }{\Q \VS{16}A man’s gift makes room for him,
\par }{\QB and brings him before great men.
\par }{\Q \VS{17}He who pleads his cause first seems right;
\par }{\QB until another comes and questions him.
\par }{\Q \VS{18}The lot settles disputes,
\par }{\QB and keeps strong ones apart.
\par }{\Q \VS{19}A brother offended is more difficult than a fortified city.
\par }{\QB Disputes are like the bars of a fortress.
\par }{\Q \VS{20}A man’s stomach is filled with the fruit of his mouth.
\par }{\QB With the harvest of his lips he is satisfied.
\par }{\Q \VS{21}Death and life are in the power of the tongue;
\par }{\QB those who love it will eat its fruit.
\par }{\Q \VS{22}Whoever finds a wife finds a good thing,
\par }{\QB and obtains favor of Yahweh.
\par }{\Q \VS{23}The poor plead for mercy,
\par }{\QB but the rich answer harshly.
\par }{\Q \VS{24}A man of many companions may be ruined,
\par }{\QB but there is a friend who sticks closer than a brother.

\par }\PoetryChap{19}{\Q \VerseOne{1}Better is the poor who walks in his integrity
\par }{\QB than he who is perverse in his lips and is a fool.
\par }{\Q \VS{2}It isn’t good to have zeal without knowledge,
\par }{\QB nor being hasty with one’s feet and missing the way.
\par }{\Q \VS{3}The foolishness of man subverts his way;
\par }{\QB his heart rages against Yahweh.
\par }{\Q \VS{4}Wealth adds many friends,
\par }{\QB but the poor is separated from his friend.
\par }{\Q \VS{5}A false witness shall not be unpunished.
\par }{\QB He who pours out lies shall not go free.
\par }{\Q \VS{6}Many will entreat the favor of a ruler,
\par }{\QB and everyone is a friend to a man who gives gifts.
\par }{\Q \VS{7}All the relatives of the poor shun him:
\par }{\QB how much more do his friends avoid him!
\par }{\QB He pursues them with pleas, but they are gone.
\par }{\Q \VS{8}He who gets wisdom loves his own soul.
\par }{\QB He who keeps understanding shall find good.
\par }{\Q \VS{9}A false witness shall not be unpunished.
\par }{\QB He who utters lies shall perish.
\par }{\Q \VS{10}Delicate living is not appropriate for a fool,
\par }{\QB much less for a servant to have rule over princes.
\par }{\Q \VS{11}The discretion of a man makes him slow to anger.
\par }{\QB It is his glory to overlook an offense.
\par }{\Q \VS{12}The king’s wrath is like the roaring of a lion,
\par }{\QB but his favor is like dew on the grass.
\par }{\Q \VS{13}A foolish son is the calamity of his father.
\par }{\QB A wife’s quarrels are a continual dripping.
\par }{\Q \VS{14}House and riches are an inheritance from fathers,
\par }{\QB but a prudent wife is from Yahweh.
\par }{\Q \VS{15}Slothfulness casts into a deep sleep.
\par }{\QB The idle soul shall suffer hunger.
\par }{\Q \VS{16}He who keeps the commandment keeps his soul,
\par }{\QB but he who is contemptuous in his ways shall die.
\par }{\Q \VS{17}He who has pity on the poor lends to Yahweh;
\par }{\QB he will reward him.
\par }{\Q \VS{18}Discipline your son, for there is hope;
\par }{\QB don’t be a willing party to his death.
\par }{\Q \VS{19}A hot-tempered man must pay the penalty,
\par }{\QB for if you rescue him, you must do it again.
\par }{\Q \VS{20}Listen to counsel and receive instruction,
\par }{\QB that you may be wise in your latter end.
\par }{\Q \VS{21}There are many plans in a man’s heart,
\par }{\QB but Yahweh’s counsel will prevail.
\par }{\Q \VS{22}That which makes a man to be desired is his kindness.
\par }{\QB A poor man is better than a liar.
\par }{\Q \VS{23}The fear of Yahweh leads to life, then contentment;
\par }{\QB he rests and will not be touched by trouble.
\par }{\Q \VS{24}The sluggard buries his hand in the dish;
\par }{\QB he will not so much as bring it to his mouth again.
\par }{\Q \VS{25}Flog a scoffer, and the simple will learn prudence;
\par }{\QB rebuke one who has understanding, and he will gain knowledge.
\par }{\Q \VS{26}He who robs his father and drives away his mother,
\par }{\QB is a son who causes shame and brings reproach.
\par }{\Q \VS{27}If you stop listening to instruction, my son,
\par }{\QB you will stray from the words of knowledge.
\par }{\Q \VS{28}A corrupt witness mocks justice,
\par }{\QB and the mouth of the wicked gulps down iniquity.
\par }{\Q \VS{29}Penalties are prepared for scoffers,
\par }{\QB and beatings for the backs of fools.
\par }{\BB \par }
\PoetryChap{20}{\Q \VerseOne{1}Wine is a mocker and beer is a brawler.
\par }{\QB Whoever is led astray by them is not wise.
\par }{\Q \VS{2}The terror of a king is like the roaring of a lion.
\par }{\QB He who provokes him to anger forfeits his own life.
\par }{\Q \VS{3}It is an honor for a man to keep aloof from strife,
\par }{\QB but every fool will be quarreling.
\par }{\Q \VS{4}The sluggard will not plow by reason of the winter;
\par }{\QB therefore he shall beg in harvest, and have nothing.
\par }{\Q \VS{5}Counsel in the heart of man is like deep water,
\par }{\QB but a man of understanding will draw it out.
\par }{\Q \VS{6}Many men claim to be men of unfailing love,
\par }{\QB but who can find a faithful man?
\par }{\Q \VS{7}A righteous man walks in integrity.
\par }{\QB Blessed are his children after him.
\par }{\Q \VS{8}A king who sits on the throne of judgment
\par }{\QB scatters away all evil with his eyes.
\par }{\Q \VS{9}Who can say, “I have made my heart pure.
\par }{\QB I am clean and without sin?”
\par }{\Q \VS{10}Differing weights and differing measures,
\par }{\QB both of them alike are an abomination to Yahweh.
\par }{\Q \VS{11}Even a child makes himself known by his doings,
\par }{\QB whether his work is pure, and whether it is right.
\par }{\Q \VS{12}The hearing ear, and the seeing eye,
\par }{\QB Yahweh has made even both of them.
\par }{\Q \VS{13}Don’t love sleep, lest you come to poverty.
\par }{\QB Open your eyes, and you shall be satisfied with bread.
\par }{\Q \VS{14}“It’s no good, it’s no good,” says the buyer;
\par }{\QB but when he is gone his way, then he boasts.
\par }{\Q \VS{15}There is gold and abundance of rubies,
\par }{\QB but the lips of knowledge are a rare jewel.
\par }{\Q \VS{16}Take the garment of one who puts up collateral for a stranger;
\par }{\QB and hold him in pledge for a wayward woman.
\par }{\Q \VS{17}Fraudulent food is sweet to a man,
\par }{\QB but afterwards his mouth is filled with gravel.
\par }{\Q \VS{18}Plans are established by advice;
\par }{\QB by wise guidance you wage war!
\par }{\Q \VS{19}He who goes about as a tale-bearer reveals secrets;
\par }{\QB therefore don’t keep company with him who opens wide his lips.
\par }{\Q \VS{20}Whoever curses his father or his mother,
\par }{\QB his lamp shall be put out in blackness of darkness.
\par }{\Q \VS{21}An inheritance quickly gained at the beginning,
\par }{\QB won’t be blessed in the end.
\par }{\Q \VS{22}Don’t say, “I will pay back evil.”
\par }{\QB Wait for Yahweh, and he will save you.
\par }{\Q \VS{23}Yahweh detests differing weights,
\par }{\QB and dishonest scales are not pleasing.
\par }{\Q \VS{24}A man’s steps are from Yahweh;
\par }{\QB how then can man understand his way?
\par }{\Q \VS{25}It is a snare to a man to make a rash dedication,
\par }{\QB then later to consider his vows.
\par }{\Q \VS{26}A wise king winnows out the wicked,
\par }{\QB and drives the threshing wheel over them.
\par }{\Q \VS{27}The spirit of man is Yahweh’s lamp,
\par }{\QB searching all his innermost parts.
\par }{\Q \VS{28}Love and faithfulness keep the king safe.
\par }{\QB His throne is sustained by love.
\par }{\Q \VS{29}The glory of young men is their strength.
\par }{\QB The splendor of old men is their gray hair.
\par }{\Q \VS{30}Wounding blows cleanse away evil,
\par }{\QB and beatings purge the innermost parts.
\par }{\BB \par }
\PoetryChap{21}{\Q \VerseOne{1}The king’s heart is in Yahweh’s hand like the watercourses.
\par }{\QB He turns it wherever he desires.
\par }{\Q \VS{2}Every way of a man is right in his own eyes,
\par }{\QB but Yahweh weighs the hearts.
\par }{\Q \VS{3}To do righteousness and justice
\par }{\QB is more acceptable to Yahweh than sacrifice.
\par }{\Q \VS{4}A high look and a proud heart,
\par }{\QB the lamp of the wicked, is sin.
\par }{\Q \VS{5}The plans of the diligent surely lead to profit;
\par }{\QB and everyone who is hasty surely rushes to poverty.
\par }{\Q \VS{6}Getting treasures by a lying tongue
\par }{\QB is a fleeting vapor for those who seek death.
\par }{\Q \VS{7}The violence of the wicked will drive them away,
\par }{\QB because they refuse to do what is right.
\par }{\Q \VS{8}The way of the guilty is devious,
\par }{\QB but the conduct of the innocent is upright.
\par }{\Q \VS{9}It is better to dwell in the corner of the housetop
\par }{\QB than to share a house with a contentious woman.
\par }{\Q \VS{10}The soul of the wicked desires evil;
\par }{\QB his neighbor finds no mercy in his eyes.
\par }{\Q \VS{11}When the mocker is punished, the simple gains wisdom.
\par }{\QB When the wise is instructed, he receives knowledge.
\par }{\Q \VS{12}The Righteous One considers the house of the wicked,
\par }{\QB and brings the wicked to ruin.
\par }{\Q \VS{13}Whoever stops his ears at the cry of the poor,
\par }{\QB he will also cry out, but shall not be heard.
\par }{\Q \VS{14}A gift in secret pacifies anger,
\par }{\QB and a bribe in the cloak, strong wrath.
\par }{\Q \VS{15}It is joy to the righteous to do justice;
\par }{\QB but it is a destruction to the workers of iniquity.
\par }{\Q \VS{16}The man who wanders out of the way of understanding
\par }{\QB shall rest in the assembly of the departed spirits.
\par }{\Q \VS{17}He who loves pleasure will be a poor man.
\par }{\QB He who loves wine and oil won’t be rich.
\par }{\Q \VS{18}The wicked is a ransom for the righteous,
\par }{\QB the treacherous for the upright.
\par }{\Q \VS{19}It is better to dwell in a desert land,
\par }{\QB than with a contentious and fretful woman.
\par }{\Q \VS{20}There is precious treasure and oil in the dwelling of the wise;
\par }{\QB but a foolish man swallows it up.
\par }{\Q \VS{21}He who follows after righteousness and kindness
\par }{\QB finds life, righteousness, and honor.
\par }{\Q \VS{22}A wise man scales the city of the mighty,
\par }{\QB and brings down the strength of its confidence.
\par }{\Q \VS{23}Whoever guards his mouth and his tongue
\par }{\QB keeps his soul from troubles.
\par }{\Q \VS{24}The proud and arrogant man—“Scoffer” is his name—
\par }{\QB he works in the arrogance of pride.
\par }{\Q \VS{25}The desire of the sluggard kills him,
\par }{\QB for his hands refuse to labor.
\par }{\Q \VS{26}There are those who covet greedily all day long;
\par }{\QB but the righteous give and don’t withhold.
\par }{\Q \VS{27}The sacrifice of the wicked is an abomination—
\par }{\QB how much more, when he brings it with a wicked mind!
\par }{\Q \VS{28}A false witness will perish.
\par }{\QB A man who listens speaks to eternity.
\par }{\Q \VS{29}A wicked man hardens his face;
\par }{\QB but as for the upright, he establishes his ways.
\par }{\Q \VS{30}There is no wisdom nor understanding
\par }{\QB nor counsel against Yahweh.
\par }{\Q \VS{31}The horse is prepared for the day of battle;
\par }{\QB but victory is with Yahweh.
\par }{\BB \par }
\PoetryChap{22}{\Q \VerseOne{1}A good name is more desirable than great riches,
\par }{\QB and loving favor is better than silver and gold.
\par }{\Q \VS{2}The rich and the poor have this in common:
\par }{\QB Yahweh is the maker of them all.
\par }{\Q \VS{3}A prudent man sees danger and hides himself;
\par }{\QB but the simple pass on, and suffer for it.
\par }{\Q \VS{4}The result of humility and the fear of Yahweh
\par }{\QB is wealth, honor, and life.
\par }{\Q \VS{5}Thorns and snares are in the path of the wicked:
\par }{\QB whoever guards his soul stays from them.
\par }{\Q \VS{6}Train up a child in the way he should go,
\par }{\QB and when he is old he will not depart from it.
\par }{\Q \VS{7}The rich rule over the poor.
\par }{\QB The borrower is servant to the lender.
\par }{\Q \VS{8}He who sows wickedness reaps trouble,
\par }{\QB and the rod of his fury will be destroyed.
\par }{\Q \VS{9}He who has a generous eye will be blessed;
\par }{\QB for he shares his food with the poor.
\par }{\Q \VS{10}Drive out the mocker, and strife will go out;
\par }{\QB yes, quarrels and insults will stop.
\par }{\Q \VS{11}He who loves purity of heart and speaks gracefully
\par }{\QB is the king’s friend.
\par }{\Q \VS{12}Yahweh’s eyes watch over knowledge;
\par }{\QB but he frustrates the words of the unfaithful.
\par }{\Q \VS{13}The sluggard says, “There is a lion outside!
\par }{\QB I will be killed in the streets!”
\par }{\Q \VS{14}The mouth of an adulteress is a deep pit.
\par }{\QB He who is under Yahweh’s wrath will fall into it.
\par }{\Q \VS{15}Folly is bound up in the heart of a child:
\par }{\QB the rod of discipline drives it far from him.
\par }{\Q \VS{16}Whoever oppresses the poor for his own increase and whoever gives to the rich,
\par }{\QB both come to poverty.
\par }{\BB \par }{\Q \VS{17}Turn your ear, and listen to the words of the wise.
\par }{\QB Apply your heart to my teaching.
\par }{\Q \VS{18}For it is a pleasant thing if you keep them within you,
\par }{\QB if all of them are ready on your lips.
\par }{\Q \VS{19}I teach you today, even you,
\par }{\QB So that your trust may be in Yahweh.
\par }{\Q \VS{20}Haven’t I written to you thirty excellent things
\par }{\QB of counsel and knowledge,
\par }{\Q \VS{21}To teach you truth, reliable words,
\par }{\QB to give sound answers to the ones who sent you?
\par }{\BB \par }{\Q \VS{22}Don’t exploit the poor, because he is poor;
\par }{\QB and don’t crush the needy in court;
\par }{\Q \VS{23}for Yahweh will plead their case,
\par }{\QB and plunder the life of those who plunder them.
\par }{\BB \par }{\Q \VS{24}Don’t befriend a hot-tempered man,
\par }{\QB and don’t associate with one who harbors anger:
\par }{\Q \VS{25}lest you learn his ways,
\par }{\QB and ensnare your soul.
\par }{\BB \par }{\Q \VS{26}Don’t you be one of those who strike hands,
\par }{\QB of those who are collateral for debts.
\par }{\Q \VS{27}If you don’t have means to pay,
\par }{\QB why should he take away your bed from under you?
\par }{\BB \par }{\Q \VS{28}Don’t move the ancient boundary stone
\par }{\QB which your fathers have set up.
\par }{\BB \par }{\Q \VS{29}Do you see a man skilled in his work?
\par }{\QB He will serve kings.
\par }{\QB He won’t serve obscure men.
\par }{\BB \par }
\PoetryChap{23}{\Q \VerseOne{1}When you sit to eat with a ruler,
\par }{\QB consider diligently what is before you;
\par }{\Q \VS{2}put a knife to your throat,
\par }{\QB if you are a man given to appetite.
\par }{\Q \VS{3}Don’t be desirous of his dainties,
\par }{\QB since they are deceitful food.
\par }{\Q \VS{4}Don’t weary yourself to be rich.
\par }{\QB In your wisdom, show restraint.
\par }{\Q \VS{5}Why do you set your eyes on that which is not?
\par }{\QB For it certainly sprouts wings like an eagle and flies in the sky.
\par }{\Q \VS{6}Don’t eat the food of him who has a stingy eye,
\par }{\QB and don’t crave his delicacies:
\par }{\QB \VS{7}for as he thinks about the cost, so he is.
\par }{\QB “Eat and drink!” he says to you,
\par }{\QB but his heart is not with you.
\par }{\Q \VS{8}The morsel which you have eaten you shall vomit up,
\par }{\QB and lose your good words.
\par }{\BB \par }{\Q \VS{9}Don’t speak in the ears of a fool,
\par }{\QB for he will despise the wisdom of your words.
\par }{\BB \par }{\Q \VS{10}Don’t move the ancient boundary stone.
\par }{\QB Don’t encroach on the fields of the fatherless,
\par }{\Q \VS{11}for their Defender is strong.
\par }{\QB He will plead their case against you.
\par }{\BB \par }{\Q \VS{12}Apply your heart to instruction,
\par }{\QB and your ears to the words of knowledge.
\par }{\Q \VS{13}Don’t withhold correction from a child.
\par }{\QB If you punish him with the rod, he will not die.
\par }{\Q \VS{14}Punish him with the rod,
\par }{\QB and save his soul from Sheol.\FTNT{*}{{\FR 23:14 }
Sheol is the place of the dead.}
\par }{\BB \par }{\Q \VS{15}My son, if your heart is wise,
\par }{\QB then my heart will be glad, even mine.
\par }{\Q \VS{16}Yes, my heart will rejoice
\par }{\QB when your lips speak what is right.
\par }{\Q \VS{17}Don’t let your heart envy sinners,
\par }{\QB but rather fear Yahweh all day long.
\par }{\Q \VS{18}Indeed surely there is a future hope,
\par }{\QB and your hope will not be cut off.
\par }{\Q \VS{19}Listen, my son, and be wise,
\par }{\QB and keep your heart on the right path!
\par }{\Q \VS{20}Don’t be among ones drinking too much wine,
\par }{\QB or those who gorge themselves on meat:
\par }{\Q \VS{21}for the drunkard and the glutton shall become poor;
\par }{\QB and drowsiness clothes them in rags.
\par }{\Q \VS{22}Listen to your father who gave you life,
\par }{\QB and don’t despise your mother when she is old.
\par }{\Q \VS{23}Buy the truth, and don’t sell it.
\par }{\QB Get wisdom, discipline, and understanding.
\par }{\Q \VS{24}The father of the righteous has great joy.
\par }{\QB Whoever fathers a wise child delights in him.
\par }{\Q \VS{25}Let your father and your mother be glad!
\par }{\QB Let her who bore you rejoice!
\par }{\Q \VS{26}My son, give me your heart;
\par }{\QB and let your eyes keep in my ways.
\par }{\Q \VS{27}For a prostitute is a deep pit;
\par }{\QB and a wayward wife is a narrow well.
\par }{\Q \VS{28}Yes, she lies in wait like a robber,
\par }{\QB and increases the unfaithful among men.
\par }{\BB \par }{\Q \VS{29}Who has woe?
\par }{\QB Who has sorrow?
\par }{\QB Who has strife?
\par }{\QB Who has complaints?
\par }{\QB Who has needless bruises?
\par }{\QB Who has bloodshot eyes?
\par }{\Q \VS{30}Those who stay long at the wine;
\par }{\QB those who go to seek out mixed wine.
\par }{\Q \VS{31}Don’t look at the wine when it is red,
\par }{\QB when it sparkles in the cup,
\par }{\QB when it goes down smoothly.
\par }{\Q \VS{32}In the end, it bites like a snake,
\par }{\QB and poisons like a viper.
\par }{\Q \VS{33}Your eyes will see strange things,
\par }{\QB and your mind will imagine confusing things.
\par }{\Q \VS{34}Yes, you will be as he who lies down in the middle of the sea,
\par }{\QB or as he who lies on top of the rigging:
\par }{\Q \VS{35}“They hit me, and I was not hurt!
\par }{\QB They beat me, and I don’t feel it!
\par }{\QB When will I wake up? I can do it again.
\par }{\QB I can find another.”
\par }{\BB \par }
\PoetryChap{24}{\Q \VerseOne{1}Don’t be envious of evil men,
\par }{\QB neither desire to be with them;
\par }{\Q \VS{2}for their hearts plot violence
\par }{\QB and their lips talk about mischief.
\par }{\Q \VS{3}Through wisdom a house is built;
\par }{\QB by understanding it is established;
\par }{\Q \VS{4}by knowledge the rooms are filled
\par }{\QB with all rare and beautiful treasure.
\par }{\Q \VS{5}A wise man has great power;
\par }{\QB and a knowledgeable man increases strength;
\par }{\Q \VS{6}for by wise guidance you wage your war;
\par }{\QB and victory is in many advisors.
\par }{\Q \VS{7}Wisdom is too high for a fool.
\par }{\QB He doesn’t open his mouth in the gate.
\par }{\Q \VS{8}One who plots to do evil
\par }{\QB will be called a schemer.
\par }{\Q \VS{9}The schemes of folly are sin.
\par }{\QB The mocker is detested by men.
\par }{\Q \VS{10}If you falter in the time of trouble,
\par }{\QB your strength is small.
\par }{\Q \VS{11}Rescue those who are being led away to death!
\par }{\QB Indeed, hold back those who are staggering to the slaughter!
\par }{\Q \VS{12}If you say, “Behold, we didn’t know this,”
\par }{\QB doesn’t he who weighs the hearts consider it?
\par }{\Q He who keeps your soul, doesn’t he know it?
\par }{\QB Shall he not render to every man according to his work?
\par }{\Q \VS{13}My son, eat honey, for it is good,
\par }{\QB the droppings of the honeycomb, which are sweet to your taste;
\par }{\Q \VS{14}so you shall know wisdom to be to your soul.
\par }{\QB If you have found it, then there will be a reward:
\par }{\QB Your hope will not be cut off.
\par }{\Q \VS{15}Don’t lay in wait, wicked man, against the habitation of the righteous.
\par }{\QB Don’t destroy his resting place;
\par }{\Q \VS{16}for a righteous man falls seven times and rises up again;
\par }{\QB but the wicked are overthrown by calamity.
\par }{\Q \VS{17}Don’t rejoice when your enemy falls.
\par }{\QB Don’t let your heart be glad when he is overthrown,
\par }{\Q \VS{18}lest Yahweh see it, and it displease him,
\par }{\QB and he turn away his wrath from him.
\par }{\Q \VS{19}Don’t fret yourself because of evildoers,
\par }{\QB neither be envious of the wicked;
\par }{\Q \VS{20}for there will be no reward to the evil man.
\par }{\QB The lamp of the wicked will be snuffed out.
\par }{\Q \VS{21}My son, fear Yahweh and the king.
\par }{\QB Don’t join those who are rebellious;
\par }{\Q \VS{22}for their calamity will rise suddenly.
\par }{\QB Who knows what destruction may come from them both?
\par }{\BB \par }{\PP \VS{23}These also are sayings of the wise.
\par }{\BB \par }{\Q To show partiality in judgment is not good.
\par }{\Q \VS{24}He who says to the wicked, “You are righteous,”
\par }{\QB peoples will curse him, and nations will abhor him—
\par }{\Q \VS{25}but it will go well with those who convict the guilty,
\par }{\QB and a rich blessing will come on them.
\par }{\Q \VS{26}An honest answer
\par }{\QB is like a kiss on the lips.
\par }{\Q \VS{27}Prepare your work outside,
\par }{\QB and get your fields ready.
\par }{\QB Afterwards, build your house.
\par }{\Q \VS{28}Don’t be a witness against your neighbor without cause.
\par }{\QB Don’t deceive with your lips.
\par }{\Q \VS{29}Don’t say, “I will do to him as he has done to me;
\par }{\QB I will repay the man according to his work.”
\par }{\Q \VS{30}I went by the field of the sluggard,
\par }{\QB by the vineyard of the man void of understanding:
\par }{\Q \VS{31}Behold, it was all grown over with thorns.
\par }{\QB Its surface was covered with nettles,
\par }{\QB and its stone wall was broken down.
\par }{\Q \VS{32}Then I saw, and considered well.
\par }{\QB I saw, and received instruction:
\par }{\Q \VS{33}a little sleep, a little slumber,
\par }{\QB a little folding of the hands to sleep,
\par }{\Q \VS{34}so your poverty will come as a robber
\par }{\QB and your want as an armed man.

\par }\Chap{25}{\PP \VerseOne{1}These also are proverbs of Solomon, which the men of Hezekiah king of Judah copied out.
\par }{\Q \VS{2}It is the glory of God to conceal a thing,
\par }{\QB but the glory of kings is to search out a matter.
\par }{\Q \VS{3}As the heavens for height, and the earth for depth,
\par }{\QB so the hearts of kings are unsearchable.
\par }{\Q \VS{4}Take away the dross from the silver,
\par }{\QB and material comes out for the refiner;
\par }{\Q \VS{5}Take away the wicked from the king’s presence,
\par }{\QB and his throne will be established in righteousness.
\par }{\Q \VS{6}Don’t exalt yourself in the presence of the king,
\par }{\QB or claim a place among great men;
\par }{\Q \VS{7}for it is better that it be said to you, “Come up here,”
\par }{\QB than that you should be put lower in the presence of the prince,
\par }{\QB whom your eyes have seen.
\par }{\Q \VS{8}Don’t be hasty in bringing charges to court.
\par }{\QB What will you do in the end when your neighbor shames you?
\par }{\Q \VS{9}Debate your case with your neighbor,
\par }{\QB and don’t betray the confidence of another,
\par }{\QB \VS{10}lest one who hears it put you to shame,
\par }{\QB and your bad reputation never depart.
\par }{\BB \par }{\Q \VS{11}A word fitly spoken
\par }{\QB is like apples of gold in settings of silver.
\par }{\Q \VS{12}As an earring of gold, and an ornament of fine gold,
\par }{\QB so is a wise reprover to an obedient ear.
\par }{\Q \VS{13}As the cold of snow in the time of harvest,
\par }{\QB so is a faithful messenger to those who send him;
\par }{\QB for he refreshes the soul of his masters.
\par }{\Q \VS{14}As clouds and wind without rain,
\par }{\QB so is he who boasts of gifts deceptively.
\par }{\Q \VS{15}By patience a ruler is persuaded.
\par }{\QB A soft tongue breaks the bone.
\par }{\Q \VS{16}Have you found honey?
\par }{\QB Eat as much as is sufficient for you,
\par }{\QB lest you eat too much, and vomit it.
\par }{\Q \VS{17}Let your foot be seldom in your neighbor’s house,
\par }{\QB lest he be weary of you, and hate you.
\par }{\Q \VS{18}A man who gives false testimony against his neighbor
\par }{\QB is like a club, a sword, or a sharp arrow.
\par }{\Q \VS{19}Confidence in someone unfaithful in time of trouble
\par }{\QB is like a bad tooth or a lame foot.
\par }{\Q \VS{20}As one who takes away a garment in cold weather,
\par }{\QB or vinegar on soda,
\par }{\QB so is one who sings songs to a heavy heart.
\par }{\Q \VS{21}If your enemy is hungry, give him food to eat.
\par }{\QB If he is thirsty, give him water to drink;
\par }{\Q \VS{22}for you will heap coals of fire on his head,
\par }{\QB and Yahweh will reward you.
\par }{\Q \VS{23}The north wind produces rain;
\par }{\QB so a backbiting tongue brings an angry face.
\par }{\Q \VS{24}It is better to dwell in the corner of the housetop
\par }{\QB than to share a house with a contentious woman.
\par }{\Q \VS{25}Like cold water to a thirsty soul,
\par }{\QB so is good news from a far country.
\par }{\Q \VS{26}Like a muddied spring and a polluted well,
\par }{\QB so is a righteous man who gives way before the wicked.
\par }{\Q \VS{27}It is not good to eat much honey,
\par }{\QB nor is it honorable to seek one’s own honor.
\par }{\Q \VS{28}Like a city that is broken down and without walls
\par }{\QB is a man whose spirit is without restraint.

\par }\PoetryChap{26}{\Q \VerseOne{1}Like snow in summer, and as rain in harvest,
\par }{\QB so honor is not fitting for a fool.
\par }{\Q \VS{2}Like a fluttering sparrow,
\par }{\QB like a darting swallow,
\par }{\QB so the undeserved curse doesn’t come to rest.
\par }{\Q \VS{3}A whip is for the horse,
\par }{\QB a bridle for the donkey,
\par }{\QB and a rod for the back of fools!
\par }{\Q \VS{4}Don’t answer a fool according to his folly,
\par }{\QB lest you also be like him.
\par }{\Q \VS{5}Answer a fool according to his folly,
\par }{\QB lest he be wise in his own eyes.
\par }{\Q \VS{6}One who sends a message by the hand of a fool
\par }{\QB is cutting off feet and drinking violence.
\par }{\Q \VS{7}Like the legs of the lame that hang loose,
\par }{\QB so is a parable in the mouth of fools.
\par }{\Q \VS{8}As one who binds a stone in a sling,
\par }{\QB so is he who gives honor to a fool.
\par }{\Q \VS{9}Like a thorn bush that goes into the hand of a drunkard,
\par }{\QB so is a parable in the mouth of fools.
\par }{\Q \VS{10}As an archer who wounds all,
\par }{\QB so is he who hires a fool
\par }{\QB or he who hires those who pass by.
\par }{\Q \VS{11}As a dog that returns to his vomit,
\par }{\QB so is a fool who repeats his folly.
\par }{\Q \VS{12}Do you see a man wise in his own eyes?
\par }{\QB There is more hope for a fool than for him.
\par }{\Q \VS{13}The sluggard says, “There is a lion in the road!
\par }{\QB A fierce lion roams the streets!”
\par }{\Q \VS{14}As the door turns on its hinges,
\par }{\QB so does the sluggard on his bed.
\par }{\Q \VS{15}The sluggard buries his hand in the dish.
\par }{\QB He is too lazy to bring it back to his mouth.
\par }{\Q \VS{16}The sluggard is wiser in his own eyes
\par }{\QB than seven men who answer with discretion.
\par }{\Q \VS{17}Like one who grabs a dog’s ears
\par }{\QB is one who passes by and meddles in a quarrel not his own.
\par }{\Q \VS{18}Like a madman who shoots torches, arrows, and death,
\par }{\QB \VS{19}is the man who deceives his neighbor and says, “Am I not joking?”
\par }{\Q \VS{20}For lack of wood a fire goes out.
\par }{\QB Without gossip, a quarrel dies down.
\par }{\Q \VS{21}As coals are to hot embers,
\par }{\QB and wood to fire,
\par }{\QB so is a contentious man to kindling strife.
\par }{\Q \VS{22}The words of a whisperer are as dainty morsels,
\par }{\QB they go down into the innermost parts.
\par }{\Q \VS{23}Like silver dross on an earthen vessel
\par }{\QB are the lips of a fervent one with an evil heart.
\par }{\Q \VS{24}A malicious man disguises himself with his lips,
\par }{\QB but he harbors evil in his heart.
\par }{\Q \VS{25}When his speech is charming, don’t believe him,
\par }{\QB for there are seven abominations in his heart.
\par }{\Q \VS{26}His malice may be concealed by deception,
\par }{\QB but his wickedness will be exposed in the assembly.
\par }{\Q \VS{27}Whoever digs a pit shall fall into it.
\par }{\QB Whoever rolls a stone, it will come back on him.
\par }{\Q \VS{28}A lying tongue hates those it hurts;
\par }{\QB and a flattering mouth works ruin.

\par }\PoetryChap{27}{\Q \VerseOne{1}Don’t boast about tomorrow;
\par }{\QB for you don’t know what a day may bring.
\par }{\Q \VS{2}Let another man praise you,
\par }{\QB and not your own mouth;
\par }{\QB a stranger, and not your own lips.
\par }{\Q \VS{3}A stone is heavy,
\par }{\QB and sand is a burden;
\par }{\QB but a fool’s provocation is heavier than both.
\par }{\Q \VS{4}Wrath is cruel,
\par }{\QB and anger is overwhelming;
\par }{\QB but who is able to stand before jealousy?
\par }{\Q \VS{5}Better is open rebuke
\par }{\QB than hidden love.
\par }{\Q \VS{6}The wounds of a friend are faithful,
\par }{\QB although the kisses of an enemy are profuse.
\par }{\Q \VS{7}A full soul loathes a honeycomb;
\par }{\QB but to a hungry soul, every bitter thing is sweet.
\par }{\Q \VS{8}As a bird that wanders from her nest,
\par }{\QB so is a man who wanders from his home.
\par }{\Q \VS{9}Perfume and incense bring joy to the heart;
\par }{\QB so does earnest counsel from a man’s friend.
\par }{\Q \VS{10}Don’t forsake your friend and your father’s friend.
\par }{\QB Don’t go to your brother’s house in the day of your disaster.
\par }{\QB A neighbor who is near is better than a distant brother.
\par }{\Q \VS{11}Be wise, my son,
\par }{\QB and bring joy to my heart,
\par }{\QB then I can answer my tormentor.
\par }{\Q \VS{12}A prudent man sees danger and takes refuge;
\par }{\QB but the simple pass on, and suffer for it.
\par }{\Q \VS{13}Take his garment when he puts up collateral for a stranger.
\par }{\QB Hold it for a wayward woman!
\par }{\Q \VS{14}He who blesses his neighbor with a loud voice early in the morning,
\par }{\QB it will be taken as a curse by him.
\par }{\Q \VS{15}A continual dropping on a rainy day
\par }{\QB and a contentious wife are alike:
\par }{\Q \VS{16}restraining her is like restraining the wind,
\par }{\QB or like grasping oil in his right hand.
\par }{\BB \par }{\Q \VS{17}Iron sharpens iron;
\par }{\QB so a man sharpens his friend’s countenance.
\par }{\Q \VS{18}Whoever tends the fig tree shall eat its fruit.
\par }{\QB He who looks after his master shall be honored.
\par }{\Q \VS{19}Like water reflects a face,
\par }{\QB so a man’s heart reflects the man.
\par }{\Q \VS{20}Sheol\FTNT{*}{{\FR 27:20 }
Sheol is the place of the dead.} and Abaddon are never satisfied;
\par }{\QB and a man’s eyes are never satisfied.
\par }{\Q \VS{21}The crucible is for silver,
\par }{\QB and the furnace for gold;
\par }{\QB but man is refined by his praise.
\par }{\Q \VS{22}Though you grind a fool in a mortar with a pestle along with grain,
\par }{\QB yet his foolishness will not be removed from him.
\par }{\BB \par }{\Q \VS{23}Know well the state of your flocks,
\par }{\QB and pay attention to your herds:
\par }{\Q \VS{24}for riches are not forever,
\par }{\QB nor does the crown endure to all generations.
\par }{\Q \VS{25}The hay is removed, and the new growth appears,
\par }{\QB the grasses of the hills are gathered in.
\par }{\Q \VS{26}The lambs are for your clothing,
\par }{\QB and the goats are the price of a field.
\par }{\Q \VS{27}There will be plenty of goats’ milk for your food,
\par }{\QB for your family’s food,
\par }{\QB and for the nourishment of your servant girls.

\par }\PoetryChap{28}{\Q \VerseOne{1}The wicked flee when no one pursues;
\par }{\QB but the righteous are as bold as a lion.
\par }{\Q \VS{2}In rebellion, a land has many rulers,
\par }{\QB but order is maintained by a man of understanding and knowledge.
\par }{\Q \VS{3}A needy man who oppresses the poor
\par }{\QB is like a driving rain which leaves no crops.
\par }{\Q \VS{4}Those who forsake the law praise the wicked;
\par }{\QB but those who keep the law contend with them.
\par }{\Q \VS{5}Evil men don’t understand justice;
\par }{\QB but those who seek Yahweh understand it fully.
\par }{\Q \VS{6}Better is the poor who walks in his integrity,
\par }{\QB than he who is perverse in his ways, and he is rich.
\par }{\Q \VS{7}Whoever keeps the law is a wise son;
\par }{\QB but he who is a companion of gluttons shames his father.
\par }{\Q \VS{8}He who increases his wealth by excessive interest
\par }{\QB gathers it for one who has pity on the poor.
\par }{\Q \VS{9}He who turns away his ear from hearing the law,
\par }{\QB even his prayer is an abomination.
\par }{\Q \VS{10}Whoever causes the upright to go astray in an evil way,
\par }{\QB he will fall into his own trap;
\par }{\QB but the blameless will inherit good.
\par }{\Q \VS{11}The rich man is wise in his own eyes;
\par }{\QB but the poor who has understanding sees through him.
\par }{\Q \VS{12}When the righteous triumph, there is great glory;
\par }{\QB but when the wicked rise, men hide themselves.
\par }{\Q \VS{13}He who conceals his sins doesn’t prosper,
\par }{\QB but whoever confesses and renounces them finds mercy.
\par }{\Q \VS{14}Blessed is the man who always fears;
\par }{\QB but one who hardens his heart falls into trouble.
\par }{\Q \VS{15}As a roaring lion or a charging bear,
\par }{\QB so is a wicked ruler over helpless people.
\par }{\Q \VS{16}A tyrannical ruler lacks judgment.
\par }{\QB One who hates ill-gotten gain will have long days.
\par }{\Q \VS{17}A man who is tormented by life blood will be a fugitive until death;
\par }{\QB no one will support him.
\par }{\Q \VS{18}Whoever walks blamelessly is kept safe;
\par }{\QB but one with perverse ways will fall suddenly.
\par }{\Q \VS{19}One who works his land will have an abundance of food;
\par }{\QB but one who chases fantasies will have his fill of poverty.
\par }{\Q \VS{20}A faithful man is rich with blessings;
\par }{\QB but one who is eager to be rich will not go unpunished.
\par }{\Q \VS{21}To show partiality is not good;
\par }{\QB yet a man will do wrong for a piece of bread.
\par }{\Q \VS{22}A stingy man hurries after riches,
\par }{\QB and doesn’t know that poverty waits for him.
\par }{\Q \VS{23}One who rebukes a man will afterward find more favor
\par }{\QB than one who flatters with the tongue.
\par }{\Q \VS{24}Whoever robs his father or his mother and says, “It’s not wrong,”
\par }{\QB is a partner with a destroyer.
\par }{\Q \VS{25}One who is greedy stirs up strife;
\par }{\QB but one who trusts in Yahweh will prosper.
\par }{\Q \VS{26}One who trusts in himself is a fool;
\par }{\QB but one who walks in wisdom is kept safe.
\par }{\Q \VS{27}One who gives to the poor has no lack;
\par }{\QB but one who closes his eyes will have many curses.
\par }{\Q \VS{28}When the wicked rise, men hide themselves;
\par }{\QB but when they perish, the righteous thrive.

\par }\PoetryChap{29}{\Q \VerseOne{1}He who is often rebuked and stiffens his neck
\par }{\QB will be destroyed suddenly, with no remedy.
\par }{\Q \VS{2}When the righteous thrive, the people rejoice;
\par }{\QB but when the wicked rule, the people groan.
\par }{\Q \VS{3}Whoever loves wisdom brings joy to his father;
\par }{\QB but a companion of prostitutes squanders his wealth.
\par }{\Q \VS{4}The king by justice makes the land stable,
\par }{\QB but he who takes bribes tears it down.
\par }{\Q \VS{5}A man who flatters his neighbor
\par }{\QB spreads a net for his feet.
\par }{\Q \VS{6}An evil man is snared by his sin,
\par }{\QB but the righteous can sing and be glad.
\par }{\Q \VS{7}The righteous care about justice for the poor.
\par }{\QB The wicked aren’t concerned about knowledge.
\par }{\Q \VS{8}Mockers stir up a city,
\par }{\QB but wise men turn away anger.
\par }{\Q \VS{9}If a wise man goes to court with a foolish man,
\par }{\QB the fool rages or scoffs, and there is no peace.
\par }{\Q \VS{10}The bloodthirsty hate a man of integrity;
\par }{\QB and they seek the life of the upright.
\par }{\Q \VS{11}A fool vents all of his anger,
\par }{\QB but a wise man brings himself under control.
\par }{\Q \VS{12}If a ruler listens to lies,
\par }{\QB all of his officials are wicked.
\par }{\Q \VS{13}The poor man and the oppressor have this in common:
\par }{\QB Yahweh gives sight to the eyes of both.
\par }{\Q \VS{14}The king who fairly judges the poor,
\par }{\QB his throne shall be established forever.
\par }{\Q \VS{15}The rod of correction gives wisdom,
\par }{\QB but a child left to himself causes shame to his mother.
\par }{\Q \VS{16}When the wicked increase, sin increases;
\par }{\QB but the righteous will see their downfall.
\par }{\Q \VS{17}Correct your son, and he will give you peace;
\par }{\QB yes, he will bring delight to your soul.
\par }{\Q \VS{18}Where there is no revelation, the people cast off restraint;
\par }{\QB but one who keeps the law is blessed.
\par }{\Q \VS{19}A servant can’t be corrected by words.
\par }{\QB Though he understands, yet he will not respond.
\par }{\Q \VS{20}Do you see a man who is hasty in his words?
\par }{\QB There is more hope for a fool than for him.
\par }{\Q \VS{21}He who pampers his servant from youth
\par }{\QB will have him become a son in the end.
\par }{\Q \VS{22}An angry man stirs up strife,
\par }{\QB and a wrathful man abounds in sin.
\par }{\Q \VS{23}A man’s pride brings him low,
\par }{\QB but one of lowly spirit gains honor.
\par }{\Q \VS{24}Whoever is an accomplice of a thief is an enemy of his own soul.
\par }{\QB He takes an oath, but dares not testify.
\par }{\Q \VS{25}The fear of man proves to be a snare,
\par }{\QB but whoever puts his trust in Yahweh is kept safe.
\par }{\Q \VS{26}Many seek the ruler’s favor,
\par }{\QB but a man’s justice comes from Yahweh.
\par }{\Q \VS{27}A dishonest man detests the righteous,
\par }{\QB and the upright in their ways detest the wicked.

\par }\Chap{30}{\PP \VerseOne{1}The words of Agur the son of Jakeh; the revelation:
\par }{\Q the man says to Ithiel,
\par }{\QB to Ithiel and Ucal:
\par }{\Q \VS{2}“Surely I am the most ignorant man,
\par }{\QB and don’t have a man’s understanding.
\par }{\Q \VS{3}I have not learned wisdom,
\par }{\QB neither do I have the knowledge of the Holy One.
\par }{\Q \VS{4}Who has ascended up into heaven, and descended?
\par }{\QB Who has gathered the wind in his fists?
\par }{\QB Who has bound the waters in his garment?
\par }{\QB Who has established all the ends of the earth?
\par }{\QB What is his name, and what is his son’s name, if you know?
\par }{\BB \par }{\Q \VS{5}“Every word of God is flawless.
\par }{\QB He is a shield to those who take refuge in him.
\par }{\Q \VS{6}Don’t you add to his words,
\par }{\QB lest he reprove you, and you be found a liar.
\par }{\BB \par }{\Q \VS{7}“Two things I have asked of you.
\par }{\QB Don’t deny me before I die.
\par }{\Q \VS{8}Remove far from me falsehood and lies.
\par }{\QB Give me neither poverty nor riches.
\par }{\QB Feed me with the food that is needful for me,
\par }{\Q \VS{9}lest I be full, deny you, and say, ‘Who is Yahweh?’
\par }{\QB or lest I be poor, and steal,
\par }{\QB and so dishonor the name of my God.
\par }{\BB \par }{\Q \VS{10}“Don’t slander a servant to his master,
\par }{\QB lest he curse you, and you be held guilty.
\par }{\BB \par }{\Q \VS{11}There is a generation that curses their father,
\par }{\QB and doesn’t bless their mother.
\par }{\Q \VS{12}There is a generation that is pure in their own eyes,
\par }{\QB yet are not washed from their filthiness.
\par }{\Q \VS{13}There is a generation, oh how lofty are their eyes!
\par }{\QB Their eyelids are lifted up.
\par }{\Q \VS{14}There is a generation whose teeth are like swords,
\par }{\QB and their jaws like knives,
\par }{\QB to devour the poor from the earth, and the needy from among men.
\par }{\BB \par }{\Q \VS{15}“The leech has two daughters:
\par }{\QB ‘Give, give.’
\par }{\BB \par }{\Q “There are three things that are never satisfied;
\par }{\QB four that don’t say, ‘Enough:’
\par }{\QB \VS{16}Sheol,\FTNT{*}{{\FR 30:16 }
Sheol is the place of the dead.}
\par }{\QB the barren womb;
\par }{\QB the earth that is not satisfied with water;
\par }{\QB and the fire that doesn’t say, ‘Enough.’
\par }{\BB \par }{\Q \VS{17}“The eye that mocks at his father,
\par }{\QB and scorns obedience to his mother:
\par }{\QB the ravens of the valley shall pick it out,
\par }{\QB the young eagles shall eat it.
\par }{\BB \par }{\Q \VS{18}“There are three things which are too amazing for me,
\par }{\QB four which I don’t understand:
\par }{\QB \VS{19}The way of an eagle in the air,
\par }{\QB the way of a serpent on a rock,
\par }{\QB the way of a ship in the middle of the sea,
\par }{\QB and the way of a man with a maiden.
\par }{\BB \par }{\Q \VS{20}“So is the way of an adulterous woman:
\par }{\QB She eats and wipes her mouth,
\par }{\QB and says, ‘I have done nothing wrong.’
\par }{\BB \par }{\Q \VS{21}“For three things the earth trembles,
\par }{\QB and under four, it can’t bear up:
\par }{\QB \VS{22}For a servant when he is king,
\par }{\QB a fool when he is filled with food,
\par }{\QB \VS{23}for an unloved woman when she is married,
\par }{\QB and a servant who is heir to her mistress.
\par }{\BB \par }{\Q \VS{24}“There are four things which are little on the earth,
\par }{\QB but they are exceedingly wise:
\par }{\QB \VS{25}The ants are not a strong people,
\par }{\QB yet they provide their food in the summer.
\par }{\QB \VS{26}The hyraxes are but a feeble folk,
\par }{\QB yet make they their houses in the rocks.
\par }{\QB \VS{27}The locusts have no king,
\par }{\QB yet they advance in ranks.
\par }{\QB \VS{28}You can catch a lizard with your hands,
\par }{\QB yet it is in kings’ palaces.
\par }{\BB \par }{\Q \VS{29}“There are three things which are stately in their march,
\par }{\QB four which are stately in going:
\par }{\QB \VS{30}The lion, which is mightiest among animals,
\par }{\QB and doesn’t turn away for any;
\par }{\QB \VS{31}the greyhound;
\par }{\QB the male goat;
\par }{\QB and the king against whom there is no rising up.
\par }{\BB \par }{\Q \VS{32}“If you have done foolishly in lifting up yourself,
\par }{\QB or if you have thought evil,
\par }{\Q put your hand over your mouth.
\par }{\QB \VS{33}For as the churning of milk produces butter,
\par }{\QB and the wringing of the nose produces blood;
\par }{\QB so the forcing of wrath produces strife.”

\par }\Chap{31}{\PP \VerseOne{1}The words of king Lemuel; the revelation which his mother taught him.
\par }{\BB \par }{\Q \VS{2}“Oh, my son!
\par }{\QB Oh, son of my womb!
\par }{\QB Oh, son of my vows!
\par }{\Q \VS{3}Don’t give your strength to women,
\par }{\QB nor your ways to that which destroys kings.
\par }{\Q \VS{4}It is not for kings, Lemuel,
\par }{\QB it is not for kings to drink wine,
\par }{\QB nor for princes to say, ‘Where is strong drink?’
\par }{\Q \VS{5}lest they drink, and forget the law,
\par }{\QB and pervert the justice due to anyone who is afflicted.
\par }{\Q \VS{6}Give strong drink to him who is ready to perish,
\par }{\QB and wine to the bitter in soul.
\par }{\Q \VS{7}Let him drink, and forget his poverty,
\par }{\QB and remember his misery no more.
\par }{\Q \VS{8}Open your mouth for the mute,
\par }{\QB in the cause of all who are left desolate.
\par }{\Q \VS{9}Open your mouth, judge righteously,
\par }{\QB and serve justice to the poor and needy.”
\par }{\BB \par }{\Q \VS{10}\FTNT{*}{{\FR 31:10 }
Proverbs 31:10-31 form an acrostic, with each verse starting with each letter of the Hebrew alphabet, in order.}Who can find a worthy woman?
\par }{\QB For her price is far above rubies.
\par }{\Q \VS{11}The heart of her husband trusts in her.
\par }{\QB He shall have no lack of gain.
\par }{\Q \VS{12}She does him good, and not harm,
\par }{\QB all the days of her life.
\par }{\Q \VS{13}She seeks wool and flax,
\par }{\QB and works eagerly with her hands.
\par }{\Q \VS{14}She is like the merchant ships.
\par }{\QB She brings her bread from afar.
\par }{\Q \VS{15}She rises also while it is yet night,
\par }{\QB gives food to her household,
\par }{\QB and portions for her servant girls.
\par }{\Q \VS{16}She considers a field, and buys it.
\par }{\QB With the fruit of her hands, she plants a vineyard.
\par }{\Q \VS{17}She arms her waist with strength,
\par }{\QB and makes her arms strong.
\par }{\Q \VS{18}She perceives that her merchandise is profitable.
\par }{\QB Her lamp doesn’t go out by night.
\par }{\Q \VS{19}She lays her hands to the distaff,
\par }{\QB and her hands hold the spindle.
\par }{\Q \VS{20}She opens her arms to the poor;
\par }{\QB yes, she extends her hands to the needy.
\par }{\Q \VS{21}She is not afraid of the snow for her household;
\par }{\QB for all her household are clothed with scarlet.
\par }{\Q \VS{22}She makes for herself carpets of tapestry.
\par }{\QB Her clothing is fine linen and purple.
\par }{\Q \VS{23}Her husband is respected in the gates,
\par }{\QB when he sits among the elders of the land.
\par }{\Q \VS{24}She makes linen garments and sells them,
\par }{\QB and delivers sashes to the merchant.
\par }{\Q \VS{25}Strength and dignity are her clothing.
\par }{\QB She laughs at the time to come.
\par }{\Q \VS{26}She opens her mouth with wisdom.
\par }{\QB Kind instruction is on her tongue.
\par }{\Q \VS{27}She looks well to the ways of her household,
\par }{\QB and doesn’t eat the bread of idleness.
\par }{\Q \VS{28}Her children rise up and call her blessed.
\par }{\QB Her husband also praises her:
\par }{\Q \VS{29}“Many women do noble things,
\par }{\QB but you excel them all.”
\par }{\Q \VS{30}Charm is deceitful, and beauty is vain;
\par }{\QB but a woman who fears Yahweh, she shall be praised.
\par }{\Q \VS{31}Give her of the fruit of her hands!
\par }{\QB Let her works praise her in the gates!
\par }