\NormalFont\ShortTitle{Titus}
{\MT Paul’s Letter to Titus

\par }\ChapOne{1}{\PP \VerseOne{1}Paul, a servant of God, and an apostle of Jesus Christ, according to the faith of God’s chosen ones, and the knowledge of the truth which is according to godliness,
\VS{2}in hope of eternal life, which God, who can’t lie, promised before time began;
\VS{3}but in his own time revealed his word in the message with which I was entrusted according to the commandment of God our Savior;
\VS{4}to Titus, my true child according to a common faith: Grace, mercy, and peace from God the Father and the Lord Jesus Christ our Savior.
\par }{\PP \VS{5}I left you in Crete for this reason, that you would set in order the things that were lacking and appoint elders in every city, as I directed you,
\VS{6}if anyone is blameless, the husband of one wife, having children who believe, who are not accused of loose or unruly behavior.
\VS{7}For the overseer must be blameless, as God’s steward, not self-pleasing, not easily angered, not given to wine, not violent, not greedy for dishonest gain;
\VS{8}but given to hospitality, a lover of good, sober minded, fair, holy, self-controlled,
\VS{9}holding to the faithful word which is according to the teaching, that he may be able to exhort in the sound doctrine, and to convict those who contradict him.
\par }{\PP \VS{10}For there are also many unruly men, vain talkers and deceivers, especially those of the circumcision,
\VS{11}whose mouths must be stopped: men who overthrow whole houses, teaching things which they ought not, for dishonest gain’s sake.
\VS{12}One of them, a prophet of their own, said, “Cretans are always liars, evil beasts, and idle gluttons.”
\VS{13}This testimony is true. For this cause, reprove them sharply, that they may be sound in the faith,
\VS{14}not paying attention to Jewish fables and commandments of men who turn away from the truth.
\VS{15}To the pure, all things are pure; but to those who are defiled and unbelieving, nothing is pure; but both their mind and their conscience are defiled.
\VS{16}They profess that they know God, but by their deeds they deny him, being abominable, disobedient, and unfit for any good work.

\par }\Chap{2}{\PP \VerseOne{1}But say the things which fit sound doctrine,
\VS{2}that older men should be temperate, sensible, sober minded, sound in faith, in love, and in perseverance:
\VS{3}and that older women likewise be reverent in behavior, not slanderers nor enslaved to much wine, teachers of that which is good,
\VS{4}that they may train the young wives to love their husbands, to love their children,
\VS{5}to be sober minded, chaste, workers at home, kind, being in subjection to their own husbands, that God’s word may not be blasphemed.
\VS{6}Likewise, exhort the younger men to be sober minded.
\VS{7}In all things be showing yourself an example of good works. In your teaching, be showing integrity, seriousness, incorruptibility,
\VS{8}and soundness of speech that can’t be condemned, that he who opposes you may be ashamed, having no evil thing to say about us.
\VS{9}Exhort servants to be in subjection to their own masters and to be well-pleasing in all things, not contradicting,
\VS{10}not stealing, but showing all good fidelity, that they may adorn the doctrine of God, our Savior, in all things.
\VS{11}For the grace of God has appeared, bringing salvation to all men,
\VS{12}instructing us to the intent that, denying ungodliness and worldly lusts, we would live soberly, righteously, and godly in this present age;
\VS{13}looking for the blessed hope and appearing of the glory of our great God and Savior, Jesus Christ,
\VS{14}who gave himself for us, that he might redeem us from all iniquity, and purify for himself a people for his own possession, zealous for good works.
\par }{\PP \VS{15}Say these things and exhort and reprove with all authority. Let no one despise you.

\par }\Chap{3}{\PP \VerseOne{1}Remind them to be in subjection to rulers and to authorities, to be obedient, to be ready for every good work,
\VS{2}to speak evil of no one, not to be contentious, to be gentle, showing all humility toward all men.
\VS{3}For we were also once foolish, disobedient, deceived, serving various lusts and pleasures, living in malice and envy, hateful, and hating one another.
\VS{4}But when the kindness of God our Savior and his love toward mankind appeared,
\VS{5}not by works of righteousness which we did ourselves, but according to his mercy, he saved us through the washing of regeneration and renewing by the Holy Spirit,
\VS{6}whom he poured out on us richly, through Jesus Christ our Savior;
\VS{7}that being justified by his grace, we might be made heirs according to the hope of eternal life.
\VS{8}This saying is faithful, and concerning these things I desire that you affirm confidently, so that those who have believed God may be careful to maintain good works. These things are good and profitable to men;
\VS{9}but shun foolish questionings, genealogies, strife, and disputes about the law; for they are unprofitable and vain.
\VS{10}Avoid a factious man after a first and second warning,
\VS{11}knowing that such a one is perverted and sins, being self-condemned.
\par }{\PP \VS{12}When I send Artemas to you, or Tychicus, be diligent to come to me to Nicopolis, for I have determined to winter there.
\VS{13}Send Zenas, the lawyer, and Apollos on their journey speedily, that nothing may be lacking for them.
\VS{14}Let our people also learn to maintain good works for necessary uses, that they may not be unfruitful.
\par }{\PP \VS{15}All who are with me greet you. Greet those who love us in faith. Grace be with you all. Amen.
\par }