\NormalFont\ShortTitle{2 Timothy}
{\MT Paul’s Second Letter to Timothy

\par }\ChapOne{1}{\PP \VerseOne{1}Paul, an apostle of Jesus Christ through the will of God, according to the promise of the life which is in Christ Jesus,
\VS{2}to Timothy, my beloved child: Grace, mercy, and peace, from God the Father and Christ Jesus our Lord.
\par }{\PP \VS{3}I thank God, whom I serve as my forefathers did, with a pure conscience. How unceasing is my memory of you in my petitions, night and day
\VS{4}longing to see you, remembering your tears, that I may be filled with joy;
\VS{5}having been reminded of the sincere faith that is in you, which lived first in your grandmother Lois, and your mother Eunice, and, I am persuaded, in you also.
\par }{\PP \VS{6}For this cause, I remind you that you should stir up the gift of God which is in you through the laying on of my hands.
\VS{7}For God didn’t give us a spirit of fear, but of power, love, and self-control.
\VS{8}Therefore don’t be ashamed of the testimony of our Lord, nor of me his prisoner; but endure hardship for the Good News according to the power of God,
\VS{9}who saved us and called us with a holy calling, not according to our works, but according to his own purpose and grace, which was given to us in Christ Jesus before times eternal,
\VS{10}but has now been revealed by the appearing of our Savior, Christ Jesus, who abolished death, and brought life and immortality to light through the Good News.
\VS{11}For this I was appointed as a preacher, an apostle, and a teacher of the Gentiles.
\VS{12}For this cause I also suffer these things.
\par }{\PP Yet I am not ashamed, for I know him whom I have believed, and I am persuaded that he is able to guard that which I have committed to him against that day.
\par }{\PP \VS{13}Hold the pattern of sound words which you have heard from me, in faith and love which is in Christ Jesus.
\VS{14}That good thing which was committed to you, guard through the Holy Spirit who dwells in us.
\par }{\PP \VS{15}This you know, that all who are in Asia turned away from me, of whom are Phygelus and Hermogenes.
\VS{16}May the Lord grant mercy to the house of Onesiphorus, for he often refreshed me, and was not ashamed of my chain,
\VS{17}but when he was in Rome, he sought me diligently and found me
\VS{18}(the Lord grant to him to find the Lord’s mercy in that day); and in how many things he served at Ephesus, you know very well.

\par }\Chap{2}{\PP \VerseOne{1}You therefore, my child, be strengthened in the grace that is in Christ Jesus.
\VS{2}The things which you have heard from me among many witnesses, commit the same things to faithful men, who will be able to teach others also.
\VS{3}You therefore must endure hardship as a good soldier of Christ Jesus.
\VS{4}No soldier on duty entangles himself in the affairs of life, that he may please him who enrolled him as a soldier.
\VS{5}Also, if anyone competes in athletics, he isn’t crowned unless he has competed by the rules.
\VS{6}The farmer who labors must be the first to get a share of the crops.
\VS{7}Consider what I say, and may the Lord give you understanding in all things.
\par }{\PP \VS{8}Remember Jesus Christ, risen from the dead, of the offspring\FTNT{*}{{\FR 2:8 }
or, seed} of David, according to my Good News,
\VS{9}in which I suffer hardship to the point of chains as a criminal. But God’s word isn’t chained.
\VS{10}Therefore I endure all things for the chosen ones’ sake, that they also may obtain the salvation which is in Christ Jesus with eternal glory.
\VS{11}This saying is trustworthy:
\par }{\Q “For if we died with him,
\par }{\QB we will also live with him.
\par }{\Q \VS{12}If we endure,
\par }{\QB we will also reign with him.
\par }{\Q If we deny him,
\par }{\QB he also will deny us.
\par }{\Q \VS{13}If we are faithless,
\par }{\QB he remains faithful;
\par }{\QB for he can’t deny himself.”
\par }{\PP \VS{14}Remind them of these things, charging them in the sight of the Lord, that they don’t argue about words, to no profit, to the subverting of those who hear.
\par }{\PP \VS{15}Give diligence to present yourself approved by God, a workman who doesn’t need to be ashamed, properly handling the Word of Truth.
\VS{16}But shun empty chatter, for it will go further in ungodliness,
\VS{17}and those words will consume like gangrene, of whom is Hymenaeus and Philetus:
\VS{18}men who have erred concerning the truth, saying that the resurrection is already past, and overthrowing the faith of some.
\VS{19}However God’s firm foundation stands, having this seal, “The Lord knows those who are his,”\XREF{✡}{{\XO 2:19 }
Numbers 16:5} and, “Let every one who names the name of the Lord\FTNT{†}{{\FR 2:19 }
TR reads “Christ” instead of “the Lord”} depart from unrighteousness.”
\VS{20}Now in a large house there are not only vessels of gold and of silver, but also of wood and of clay. Some are for honor, and some for dishonor.
\VS{21}If anyone therefore purges himself from these, he will be a vessel for honor, sanctified, and suitable for the master’s use, prepared for every good work.
\par }{\PP \VS{22}Flee from youthful lusts; but pursue righteousness, faith, love, and peace with those who call on the Lord out of a pure heart.
\VS{23}But refuse foolish and ignorant questionings, knowing that they generate strife.
\VS{24}The Lord’s servant must not quarrel, but be gentle toward all, able to teach, patient,
\VS{25}in gentleness correcting those who oppose him: perhaps God may give them repentance leading to a full knowledge of the truth,
\VS{26}and they may recover themselves out of the devil’s snare, having been taken captive by him to his will.

\par }\Chap{3}{\PP \VerseOne{1}But know this: that in the last days, grievous times will come.
\VS{2}For men will be lovers of self, lovers of money, boastful, arrogant, blasphemers, disobedient to parents, unthankful, unholy,
\VS{3}without natural affection, unforgiving, slanderers, without self-control, fierce, not lovers of good,
\VS{4}traitors, headstrong, conceited, lovers of pleasure rather than lovers of God,
\VS{5}holding a form of godliness, but having denied its power. Turn away from these, also.
\VS{6}For some of these are people who creep into houses and take captive gullible women loaded down with sins, led away by various lusts,
\VS{7}always learning, and never able to come to the knowledge of the truth.
\VS{8}Even as Jannes and Jambres opposed Moses, so these also oppose the truth, men corrupted in mind, who concerning the faith are rejected.
\VS{9}But they will proceed no further. For their folly will be evident to all men, as theirs also came to be.
\VS{10}But you followed my teaching, conduct, purpose, faith, patience, love, steadfastness,
\VS{11}persecutions, and sufferings: those things that happened to me at Antioch, Iconium, and Lystra. I endured those persecutions. The Lord delivered me out of them all.
\VS{12}Yes, and all who desire to live godly in Christ Jesus will suffer persecution.
\VS{13}But evil men and impostors will grow worse and worse, deceiving and being deceived.
\VS{14}But you remain in the things which you have learned and have been assured of, knowing from whom you have learned them.
\VS{15}From infancy, you have known the holy Scriptures which are able to make you wise for salvation through faith, which is in Christ Jesus.
\VS{16}Every Scripture is God-breathed and\FTNT{*}{{\FR 3:16 }
or, Every writing inspired by God is} profitable for teaching, for reproof, for correction, and for instruction in righteousness,
\VS{17}that each person who belongs to God may be complete, thoroughly equipped for every good work.

\par }\Chap{4}{\PP \VerseOne{1}I command you therefore before God and the Lord Jesus Christ, who will judge the living and the dead at his appearing and his Kingdom:
\VS{2}preach the word; be urgent in season and out of season; reprove, rebuke, and exhort with all patience and teaching.
\VS{3}For the time will come when they will not listen to the sound doctrine, but having itching ears, will heap up for themselves teachers after their own lusts,
\VS{4}and will turn away their ears from the truth, and turn away to fables.
\VS{5}But you be sober in all things, suffer hardship, do the work of an evangelist, and fulfill your ministry.
\par }{\PP \VS{6}For I am already being offered, and the time of my departure has come.
\VS{7}I have fought the good fight. I have finished the course. I have kept the faith.
\VS{8}From now on, the crown of righteousness is stored up for me, which the Lord, the righteous judge, will give to me on that day; and not to me only, but also to all those who have loved his appearing.
\par }{\PP \VS{9}Be diligent to come to me soon,
\VS{10}for Demas left me, having loved this present world, and went to Thessalonica; Crescens to Galatia; and Titus to Dalmatia.
\VS{11}Only Luke is with me. Take Mark, and bring him with you, for he is useful to me for service.
\VS{12}But I sent Tychicus to Ephesus.
\VS{13}Bring the cloak that I left at Troas with Carpus when you come, and the books, especially the parchments.
\VS{14}Alexander, the coppersmith, did much evil to me. The Lord will repay him according to his deeds,
\VS{15}of whom you also must beware; for he greatly opposed our words.
\par }{\PP \VS{16}At my first defense, no one came to help me, but all left me. May it not be held against them.
\VS{17}But the Lord stood by me and strengthened me, that through me the message might be fully proclaimed, and that all the Gentiles might hear. So I was delivered out of the mouth of the lion.
\VS{18}And the Lord will deliver me from every evil work, and will preserve me for his heavenly Kingdom. To him be the glory forever and ever. Amen.
\par }{\PP \VS{19}Greet Prisca and Aquila, and the house of Onesiphorus.
\VS{20}Erastus remained at Corinth, but I left Trophimus at Miletus sick.
\VS{21}Be diligent to come before winter. Eubulus salutes you, as do Pudens, Linus, Claudia, and all the brothers.
\par }{\PP \VS{22}The Lord Jesus Christ be with your spirit. Grace be with you. Amen.
\par }