\NormalFont\ShortTitle{Esther (Greek)}
{\MT ESTHER
\par }{\MTB translated from the Greek Septuagint
\par }{\IS Introduction
\par }{\IP The book of Esther in the Greek Septuagint contains 5 additions that the traditional Hebrew text doesn’t have. These additions are recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches. Those additions are enclosed in [square brackets]. Because the additions by themselves make little sense without the broader context of the book, we present here a translation of the whole book of Esther from the Greek.
\par }{\IP We have chosen not to distract the reader with confusing out-of-order chapter numbers that would result from using the KJV versification, but rather merge these 5 additions as extensions at the beginning of 1:1 and after 3:13, 4:17, 8:12, and 10:3. This makes some verses (1:1, 5:1, and 8:12) really long, but it also makes the verses line up with the same verse numbers in Esther as translated from the traditional Hebrew text. Some of the proper names in this book have been changed to the more familiar Hebrew form instead of the direct transliteration from the Greek.

\par }\ChapOne{1}{\PP \VerseOne{1}[In the second year of the reign of Ahasuerus the great king, on the first day of Nisan, Mordecai the son of Jair, the son of Shimei, the son of Kish, of the tribe of Benjamin, a Jew dwelling in the city Susa, a great man, serving in the king’s palace, saw a vision. Now he was of the captivity which Nebuchadnezzar king of Babylon had carried captive from Jerusalem, with Jeconiah the king of Judea. This was his dream: Behold, voices and a noise, thunders and earthquake, tumult upon the earth. And, behold, two great serpents came out, both ready for conflict. A great voice came from them. Every nation was prepared for battle by their voice, even to fight against the nation of the just. Behold, a day of darkness and blackness, suffering and anguish, affection and tumult upon the earth. And all the righteous nation was troubled, fearing their own afflictions. They prepared to die, and cried to God. Something like a great river from a little spring with much water, came from their cry. Light and the sun arose, and the lowly were exalted, and devoured the honorable.
\par }{\PP Mordecai, who had seen this vision and what God desired to do, having arisen, kept it in his heart, and desired by all means to interpret it, even until night.
\par }{\PP Mordecai rested quietly in the palace with Gabatha and Tharrha the king’s two chamberlains, eunuchs who guarded the palace. He heard their conversation and searched out their plans. He learned that they were preparing to lay hands on king Ahasuerus; and he informed the king concerning them. The king examined the two chamberlains. They confessed, and were led away and executed. The king wrote these things for a record. Mordecai also wrote concerning these matters. The king commanded Mordecai to serve in the palace, and gave gifts for this service. But Haman the son of Hammedatha the Bougean was honored in the sight of the king, and he endeavored to harm Mordecai and his people, because of the king’s two chamberlains.]\FTNT{*}{{\FR 1:1 }
Note:
In the
Hebrew
and some copies of LXX, Esther begins here.} And it came to pass after these things\FTNT{†}{{\FR 1:1 }
Greek
words.} in the days of Ahasuerus, —(this Ahasuerus ruled over one hundred twenty-seven provinces from India)—
\VS{2}in those days, when king Ahasuerus was on the throne in the city of Susa.
\VS{3}In the third year of his reign, he made a feast for his friends, and the other nations, and for the nobles of the Persians and Medes, and the chief of the local governors.
\VS{4}After this, after he had shown them the wealth of his kingdom and the abundant glory of his wealth during one hundred eighty days,
\VS{5}when the days of the wedding feast were completed, the king made a banquet for the people of the nations who were present in the city for six days, in the court of the king’s house,
\VS{6}which was adorned with fine linen and flax on cords of fine linen and purple, fastened to golden and silver studs, on pillars of white marble and stone. There were golden and silver couches on a pavement of emerald stone, and of pearl, and of white marble, with transparent coverings variously flowered, having roses worked around it.
\VS{7}There were gold and silver cups, and a small cup of carbuncle set out of the value of thirty thousand talents, with abundant and sweet wine, which the king himself drank.
\VS{8}This banquet was not according to the appointed law; but as the king desired to have it. He charged the stewards to perform his will and that of the company.
\par }{\PP \VS{9}Also Vashti the queen made a banquet for the women in the palace where king Ahasuerus lived.
\VS{10}Now on the seventh day, the king, being merry, told Haman, Bazan, Tharrha, Baraze, Zatholtha, Abataza, and Tharaba, the seven chamberlains, servants of king Ahasuerus,
\VS{11}to bring in the queen to him, to
\FTNT{‡}{{\FR 1:11 }
Greek
to make her queen.
} enthrone her, and crown her with the diadem, and to show her to the princes, and her beauty to the nations: for she was beautiful.
\VS{12}But queen Vashti didn’t listen to him to come with the chamberlains; so the king was grieved and angered.
\VS{13}And he said to his friends, “This is what Vashti said. Therefore pronounce your legal judgement on this case.”
\par }{\PP \VS{14}So Arkesaeus, Sarsathaeus, and Malisear, the princes of the Persians and Medes, who were near the king, who sat chief in rank by the king, drew near to him,
\VS{15}and reported to him according to the laws what it was proper to do to queen Vashti, because she had not done the things commanded by the king through the chamberlains.
\VS{16}And Memucan said to the king and to the princes, “Queen Vashti has not wronged the king only, but also all the king’s rulers and princes:
\VS{17}for he has told them the words of the queen, and how she
\FTNT{§}{{\FR 1:17 }
Greek
contradicted.
} disobeyed the king. As she then refused to obey king Ahasuerus,
\VS{18}so this day the other wives of the chiefs of the Persians and Medes, having heard what she said to the king, will dare in the same way to dishonor their husbands.
\VS{19}If then it seems good to the king, let him make a royal decree, and let it be written according to the laws of the Medes and Persians, and let him not alter it: ‘Don’t allow the queen to come in to him any more. Let the king give her royalty to a woman better than she.’
\VS{20}Let the law of the king which he will have made be widely proclaimed in his kingdom. Then all the women will give honor to their husbands, from the poor even to the rich.”
\VS{21}This advice pleased the king and the princes; and the king did as Memucan had said,
\VS{22}and sent into all his kingdom through the several provinces, according to their language, so that men might be feared in their own houses.

\par }\Chap{2}{\PP \VerseOne{1}After this, the king’s anger was pacified, and he no more mentioned Vashti, bearing in mind what she had said, and how he had condemned her.
\VS{2}Then the servants of the king said, “Let chaste, beautiful young virgins be sought for the king.
\VS{3}Let the king appoint local governors in all the provinces of his kingdom, and let them select fair, chaste young ladies and bring them to the city Susa, into the women’s apartment. Let them be consigned to the king’s chamberlain, the keeper of the women. Then let things for purification and other needs be given to them.
\VS{4}Let the woman who pleases the king be queen instead of Vashti.”
\par }{\PP This thing pleased the king; and he did so.
\par }{\PP \VS{5}Now there was a Jew in the city Susa, and his name was Mordecai, the son of Jairus, the son of Shimei, the son of Kish, of the tribe of Benjamin.
\VS{6}He had been brought as a prisoner from Jerusalem, which Nebuchadnezzar king of Babylon had carried into captivity.
\VS{7}He had a foster child, daughter of Aminadab his father’s brother. Her name was Esther. When her parents died, he brought her up to womanhood as his own. This lady was beautiful.
\VS{8}And because the king’s ordinance was published, many ladies were gathered to the city of Susa under the hand of Hegai; and Esther was brought to Hegai, the keeper of the women.
\VS{9}The lady pleased him, and she found favor in his sight. He hurried to give her the things for purification, her portion, and the seven maidens appointed her out of the palace. He treated her and her maidens well in the women’s apartment.
\VS{10}But Esther didn’t reveal her family or her kindred; for Mordecai had charged her not to tell.
\VS{11}But Mordecai used to walk every day by the women’s court, to see what would become of Esther.
\VS{12}Now this was the time for a virgin to go into the king, when she should have fulfilled twelve months; for so are the days of purification fulfilled, six months while they are anointing themselves with oil of myrrh, and six months with spices and women’s purifications.
\VS{13}And then the lady goes in to the king. The officer that he commands to do so will bring her to come in with him from the women’s apartment to the king’s chamber.
\VS{14}She enters in the evening, and in the morning she departs to the second women’s apartment, where Hegai the king’s chamberlain is keeper of the women. She doesn’t go in to the king again, unless she is called by name.
\VS{15}And when the time was fulfilled for Esther the daughter of Aminadab the brother of Mordecai’s father to go in to the king, she neglected nothing which the chamberlain, the women’s keeper, commanded; for Esther found grace in the sight of all who looked at her.
\VS{16}So Esther went in to king Ahasuerus in the twelfth month, which is Adar, in the seventh year of his reign.
\VS{17}The king loved Esther, and she found favor beyond all the other virgins. He put the queen’s crown on her.
\VS{18}The king made a banquet for all his friends and great men for seven days, and he highly celebrated the marriage of Esther; and he granted a remission of taxes to those who were under his dominion.
\par }{\PP \VS{19}Meanwhile, Mordecai served in the palace.
\VS{20}Now Esther had not revealed her
\FTNT{*}{{\FR 2:20 }
Greek
country.
} kindred; for so Mordecai commanded her, to fear God, and perform his commandments, as when she was with him. Esther didn’t change her manner of life.
\par }{\PP \VS{21}Two chamberlains of the king, the chiefs of the body-guard, were grieved, because Mordecai was promoted; and they sought to kill king Ahasuerus.
\VS{22}And the matter was discovered by Mordecai, and he made it known to Esther, and she declared to the king the matter of the conspiracy.
\VS{23}And the king examined the two chamberlains, and hanged them. Then the king gave orders to make a note for a memorial in the royal
\FTNT{†}{{\FR 2:23 }
Greek
library.
} records of the good offices of Mordecai, as a commendation.

\par }\Chap{3}{\PP \VerseOne{1}After this, king Ahasuerus highly honored Haman the son of Hammedatha, the Bugaean. He exalted him and set his seat above all his friends.
\VS{2}All in the palace bowed down to him, for so the king had given orders to do; but Mordecai didn’t bow down to him.
\VS{3}And they in the king’s palace said to Mordecai, “Mordecai, why do you transgress the commands of the king?”
\VS{4}They questioned him daily, but he didn’t listen to them; so they reported to Haman that Mordecai resisted the commands of the king; and Mordecai had shown to them that he was a Jew.
\VS{5}When Haman understood that Mordecai didn’t bow down to him, he was greatly enraged,
\VS{6}and took counsel to utterly destroy all the Jews who were under the rule of Ahasuerus.
\par }{\PP \VS{7}In the twelfth year of the reign of Ahasuerusin the twelfth year of the reign of Ahasuerus, he made a decision by casting lots by day and month, to kill the race of Mordecai in one day. The lot fell on the fourteenth day of the month of Adar.
\VS{8}So he spoke to king Ahasuerus, saying, “There is a nation scattered among the nations in all your kingdom, and their laws differ from all the other nations. They disobey the king’s laws. It is not expedient for the king to tolerate them.
\VS{9}If it seem good to the king, let him make a decree to destroy them, and I will remit into the king’s treasury ten thousand talents of silver.”
\par }{\PP \VS{10}So the king took off his ring, and gave it into the hands of Haman, to seal the decrees against the Jews.
\VS{11}The king said to Haman, “Keep the silver, and treat the nation as you will.”
\VS{12}So the king’s recorders were called in the first month, on the thirteenth day, and they wrote as Haman commanded to the captains and governors in every province, from India even to Ethiopia, to one hundred twenty-seven provinces; and to the rulers of the nations according to their languages, in the name of king Ahasuerus.
\VS{13}The message was sent by couriers throughout the kingdom of Ahasuerus, to utterly destroy the race of the Jews on the first day of the twelfth month, which is Adar, and to plunder their goods.
\FTNT{*}{{\FR 3:13 }
Note:
The part in brackets is not in
Hebrew}[The following is the copy of the letter. “From the great king Ahasuerus to the rulers and the governors under them of one hundred twenty-seven provinces, from India even to Ethiopia, who hold authority under him:
\par }{\PP “Ruling over many nations and having obtained dominion over the whole world, I was determined (not elated by the confidence of power, but ever conducting myself with great moderation and gentleness) to make the lives of my subjects continually tranquil, desiring both to maintain the kingdom quiet and orderly to its utmost limits, and to restore the peace desired by all men. When I had asked my counselors how this should be brought to pass, Haman, who excels in soundness of judgment among us, and has been manifestly well inclined without wavering and with unshaken fidelity, and had obtained the second post in the kingdom, informed us that a certain ill-disposed people is mixed up with all the tribes throughout the world, opposed in their law to every other nation, and continually neglecting the commands of the king, so that the united government blamelessly administered by us is not quietly established. Having then conceived that this nation is continually set in opposition to every man, introducing as a change a foreign code of laws, and injuriously plotting to accomplish the worst of evils against our interests, and against the happy establishment of the monarchy; we instruct you in the letter written by Haman, who is set over the public affairs and is our second governor, to destroy them all utterly with their wives and children by the swords of the enemies, without pitying or sparing any, on the fourteenth day of the twelfth month Adar, of the present year; that the people aforetime and now ill-disposed to us having been violently consigned to death in one day, may hereafter secure to us continually a well constituted and quiet state of affairs.”]
\VS{14}Copies of the letters were published in every province; and an order was given to all the nations to be ready for that day.
\VS{15}This business was hastened also in Susa. The king and Haman began to drink, but the city was troubled.

\par }\Chap{4}{\PP \VerseOne{1}But Mordecai, having perceived what was done, tore his garments, put on sackcloth, and sprinkled dust upon himself. Having rushed forth through the open street of the city, he cried with a loud voice, “A nation that has done no wrong is going to be destroyed!”
\VS{2}He came to the king’s gate, and stood; for it was not lawful for him to enter into the palace wearing sackcloth and ashes.
\VS{3}And in every province where the letters were published, there was crying, lamentation, and great mourning on the part of the Jews. They wore sackcloth and ashes.
\VS{4}The queen’s maids and chamberlains went in and told her; and when she had heard what was done, she was disturbed. She sent to clothe Mordecai, and take away his sackcloth; but he refused.
\VS{5}So Esther called for her chamberlain Hathach, who waited upon her; and she sent to learn the truth from Mordecai.
\VS{7}Mordecai showed him what was done, and the promise which Haman had made the king of ten thousand talents to be paid into the treasury, that he might destroy the Jews.
\VS{8}And he gave him the copy of what was published in Susa concerning their destruction to show to Esther; and told him to charge her to go in and entreat the king, and to beg him for the people. “Remember, he said, the days of your humble condition, how you were nursed by my hand; because Haman who holds the next place to the king has spoken against us for death. Call upon the Lord, and speak to the king concerning us, to deliver us from death.”
\par }{\PP \VS{9}So Hathach went in and told her all these words.
\VS{10}Esther said to Hathach, “Go to Mordecai, and say,
\VS{11}‘All the nations of the empire know than any man or woman who goes in to the king into the inner court without being called, that person can’t live; except to whomever the king stretches out his golden sceptre, he shall live. I haven’t been called to go into the king for thirty days.’ ”
\par }{\PP \VS{12}So Hathach reported to Mordecai all the words of Esther.
\VS{13}Then Mordecai said to Hathach, “Go, and say to her, ‘Esther, don’t say to yourself that you alone will escape in the kingdom, more than all the other Jews.
\VS{14}For if you refuse to listen on this occasion, help and protection will come to the Jews from another place; but you and your father’s house will perish. Who knows if you have been made queen for this occasion?’ ”
\par }{\PP \VS{15}And Esther sent the man that came to her to Mordecai, saying,
\VS{16}“Go and assemble the Jews that are in Susa, and all of you fast for me, and don’t eat or drink for three days, night and day. My maidens and I will also fast. Then I will go in to the king contrary to the law, even if I must die.”
\par }{\PP \VS{17}So Mordecai went and did all that Esther commanded him.
\FTNT{*}{{\FR 4:17 }
Note:
The part between brackets,
i.e.
to the end of chap.
iv. is not in the
Hebrew
}
\VS{18}[He prayed to the Lord, making mention of all the works of the Lord.
\VS{19}He said, “Lord
\FTNT{†}{{\FR 4:19 }
See 3 Kings 8.53.
Note.
} God, you are king ruling over all, for all things are in your power, and there is no one who can oppose you in your purpose to save Israel;
\VS{20}for you have made the heaven and the earth and every wonderful thing under heaven.
\VS{21}You are Lord of all, and there is no one who can resist you, Lord.
\VS{22}You know all things. You know, Lord, that it is not in insolence, nor arrogance, nor love of glory, that I have done this, to refuse to bow down to the arrogant Haman.
\VS{23}For I would gladly have kissed the soles of his feet for the safety of Israel.
\VS{24}But I have done this that I might not set the glory of man above the glory of God. I will not worship anyone except you, my Lord, and I will not do these things in arrogance.
\VS{25}And now, O Lord God, the King, the God of Abraham, spare your people, for our enemies are planning our destruction, and they have desired to destroy your ancient inheritance.
\VS{26}Do not overlook your people, whom you have redeemed for yourself out of the land of Egypt.
\VS{27}Listen to my prayer. Have mercy on your inheritance and turn our mourning into gladness, that we may live and sing praise to your name, O Lord. Don’t utterly destroy the mouth of those who praise you, O Lord.”
\par }{\PP \VS{28}All Israel cried with all their might, for death was before their eyes.
\VS{29}And queen Esther took refuge in the Lord, being taken as it were in the agony of death.
\VS{30}Having taken off her glorious apparel, she put on garments of distress and mourning. Instead of grand perfumes she filled her head with ashes and dung. She greatly brought down her body, and she filled every place of her glad adorning with the her tangled hair.
\VS{31}She implored the Lord God of Israel, and said, “O my Lord, you alone are our king. Help me. I am destitute, and have no helper but you,
\VS{32}for my danger is
\FTNT{‡}{{\FR 4:32 }
Greek
in my hand.
}near at hand.
\VS{33}I have heard from my birth, in the tribe of my kindred that you, Lord, took Israel out of all the nations, and our fathers out of all their kindred for a perpetual inheritance, and have done for them all that you have said.
\VS{34}And now we have sinned before you, and you have delivered us into the hands of our enemies,
\VS{35}because we honored their gods. You are righteous, O Lord.
\VS{36}But now they have not been content with the bitterness of our slavery, but have laid their hands on the hands of their idols
\VS{37}to abolish the decree of your mouth, and utterly to destroy your inheritance, and to stop the mouth of those who praise you, and to extinguish the glory of your house and your alter,
\VS{38}and to open the mouth of the Gentiles to
{\ADD{speak}} the
\FTNT{§}{{\FR 4:38 }
Greek
virtues.
} praises of vanities, and that a mortal king should be admired forever.
\VS{39}O Lord, don’t resign your sceptre to them that are not, and don’t let them laugh at our fall, but turn their counsel against themselves, and make an example of him who has
\FTNT{*}{{\FR 4:39 }
Greek
begun against us.
} begun to injure us.
\VS{40}Remember us, O Lord, manifest yourself in the time of our affliction. Encourage me, O King of gods, and ruler of all dominion!
\VS{41}Put harmonious speech into my mouth before the lion, and turn his heart to hate him who fights against us, to the utter destruction of him who agrees with him.
\VS{42}But deliver us by your hand, and help me who am destitute, and have none but you, O Lord.
\VS{43}You know all things, and know that I hate the
\FTNT{†}{{\FR 4:43 }
Or,
opinion.
}glory of transgressors, and that I abhor the couch of the uncircumcised, and of every stranger.
\VS{44}You know my necessity, for I abhor the symbol of my proud station, which is upon my head in the days of my
\FTNT{‡}{{\FR 4:44 }
Greek
vision.} splendor. I abhor it as a menstruous cloth, and I don’t wear it in the days of my tranquility.
\VS{45}Your handmaid has not eaten at Haman’s table Haman, and I have not honored the banquet of the king, neither have I drunk wine of libations.
\VS{46}Neither has your handmaid rejoiced since the day of my promotion until now, except in you, O Lord God of Abraham.
\VS{47}O god, who has power over all, listen to the voice of the desperate, and deliver us from the hand of those who devise mischief. Deliver me from my fear.]

\par }\Chap{5}{\PP \VerseOne{1}\FTNT{*}{{\FR 5:1 }
From the first verse to the third, theGreek widely differs from the
Hebrew
} It came to pass on the third day, when she had ceased praying, that she took off her servant’s dress, and put on her glorious apparel. Being splendidly dressed and having called upon God the Overseer and Preserver of all things, she took her two maids, and she leaned upon one, as a delicate female, and the other followed bearing her train. She was blooming in the perfection of her beauty. Her face was cheerful and looked lovely, but her heart was filled with fear. Having passed through all the doors, she stood before the king. He was sitting on his royal throne. He had put on all his glorious apparel, covered all over with gold and precious stones, and was very terrifying. And having raised his face resplendent with glory, he looked with intense anger. The queen fell, and changed her color as she fainted. She bowed herself upon the head of the maid who went before her. But God changed the spirit of the king to gentleness, and in intense feeling, he sprang from off his throne, and took her into his arms, until she recovered. He comforted her with peaceful words, and said to her, “What is the matter, Esther? I am your relative. Cheer up! You shall not die, for our command is openly declared to you: ‘Draw near.’ ”
\par }{\PP \VS{2}And having raised the golden sceptre, he laid it upon her neck, and embraced her. He said, “Speak to me.”
\par }{\PP So she said to him, “I saw you, my lord, as an angel of God, and my heart was troubled for fear of your glory; for you, my lord, are to be wondered at, and your face is full of grace.” While she was speaking, she fainted and fell.
\par }{\PP Then the king was troubled, and all his servants comforted her.
\VS{3}The king said, “What do you desire, Esther? What is your request? Ask even to the half of my kingdom, and it shall be yours.”
\par }{\PP \VS{4}Esther said, “Today is a special day. So if it seems good to the king, let both him and Haman come to the feast which I will prepare this day.”
\par }{\PP \VS{5}The king said, “Hurry and bring Haman here, that we may do as Esther said.” So they both come to the feast about which Esther had spoken.
\VS{6}At the banquet, the king said to Esther, “What is your request, queen Esther? You shall have all that you require.”
\par }{\PP \VS{7}She said, “My request and my petition are:
\VS{8}if I have found favor in the king’s sight, let the king and Haman come again tomorrow to the feast which I shall prepare for them, and tomorrow I will do the same.”
\par }{\PP \VS{9}So Haman went out from the king very glad and merry; but when Haman saw Mordecai the Jew in the court, he was greatly enraged.
\VS{10}Having gone into his own house, he called his friends, and his wife Zeresh.
\VS{11}He showed them his wealth and the glory with which the king had invested him, and how he had promoted him to be chief ruler in the kingdom.
\VS{12}Haman said, “The queen has called no one to the feast with the king but me, and I am invited tomorrow.
\VS{13}But these things don’t please me while I see Mordecai the Jew in the court.
\par }{\PP \VS{14}Then Zeresh his wife and his friends said to him, “Let a fifty cubit tall
\FTNT{†}{{\FR 5:14 }
Greek
a tree cut.
} gallows be made for you. In the morning you speak to the king, and let Mordecai be hanged on the gallows; but you go in to the feast with the king, and be merry.”
\par }{\PP The saying pleased Haman, and the gallows was prepared.

\par }\Chap{6}{\PP \VerseOne{1}The Lord removed sleep from the king that night; so he told his servant to bring in the
\FTNT{*}{{\FR 6:1 }
Greek
letters.
} books, the registers of daily events, to read to him.
\VS{2}And he found the
\FTNT{†}{{\FR 6:2 }
Greek
letters.
} records written concerning Mordecai, how he had told the king about the king’s two chamberlains, when they were keeping guard, and sought to lay hands on Ahasuerus.
\VS{3}The king said, “What honor or favor have we done for Mordecai?”
\par }{\PP The king’s servants said, “You haven’t done anything for him.”
\par }{\PP \VS{4}And while the king was enquiring about the kindness of Mordecai, behold, Haman was in the court. The king said, “Who is in the court? Now Haman had come in to speak to the king about hanging Mordecai on the gallows which he had prepared.
\VS{5}The king’s servants said, “Behold, Haman stands in the court.”
\par }{\PP And the king said, “Call him!”
\par }{\PP \VS{6}The king said to Haman, “What should I do for the man whom I wish to honor?”
\par }{\PP Haman said within himself, “Whom would the king honor but myself?”
\VS{7}He said to the king, “As for the man whom the king wishes to honor,
\VS{8}let the king’s servants bring the robe of fine linen which the king puts on, and the horse on which the king rides,
\VS{9}and let him give it to one of the king’s noble friends, and let him dress the man whom the king loves. Let him mount him on the horse, and proclaim through the
\FTNT{‡}{{\FR 6:9 }
Or,
wide space.
} streets of the city, saying, “This is what will be done for every man whom the king honors!”
\par }{\PP \VS{10}Then the king said to Haman, “You have spoken well. Do so for Mordecai the Jew, who waits in the palace, and let not a word of what you have spoken be neglected!”
\par }{\PP \VS{11}So Haman took the robe and the horse, dressed Mordecai, mounted him on the horse, and went through the streets of the city, and proclaimed, saying, “This is what will be done for every man whom the king wishes to honor.”
\VS{12}Then Mordecai returned to the palace; but Haman went home mourning, and having his head covered.
\par }{\PP \VS{13}Haman related the events that had happened him to Zeresh his wife, and to his friends. His friends and his wife said to him,
\FTNT{§}{{\FR 6:13 }
Or,
if it be M. etc. before whom.
}“If Mordecai is of the race of the Jews, and you have begun to be humbled before him, you will assuredly fall, and you will not be able to withstand him, for the living God is with him.”
\VS{14}While they were still speaking, the chamberlains arrived, to rush Haman to the banquet which Esther had prepared.

\par }\Chap{7}{\PP \VerseOne{1}So the king and Haman went in to drink with the queen.
\VS{2}The king said to Esther at the banquet on the second day, “What is it, queen Esther? What is your request? What is your petition? It shall be done for you, up to half of my kingdom.”
\par }{\PP \VS{3}She answered and said, “If I have found favor in the sight of the king, let my life be granted as my petition, and my people as my request.
\VS{4}For both I and my people are sold for destruction, pillage, and genocide. If both we and our children were sold for male and female slaves, I would not have bothered you, for this
\FTNT{*}{{\FR 7:4 }
see
Heb. slanderer
} isn’t worthy of the king’s palace.”
\par }{\PP \VS{5}The king said, “Who has dared to do this thing?”
\par }{\PP \VS{6}Esther said, “The enemy is Haman, this wicked man!”
\par }{\PP Then Haman was troubled before the king and the queen.
\VS{7}The king rose up from the banquet to go into the garden. Haman began to beg the queen for mercy; for he saw that he was in trouble.
\VS{8}The king returned from the garden; and Haman had fallen upon the bed, begging the queen for mercy. The king said, “Will you even assault my wife in my house?”
\par }{\PP And when Haman heard it, he changed countenance.
\VS{9}And Bugathan, one of the chamberlains, said to the king, “Behold, Haman has also prepared a gallows for Mordecai, who spoke concerning the king, and a fifty cubit high gallows has been set up on Haman’s property.”
\par }{\PP The king said, “Let him be
\FTNT{†}{{\FR 7:9 }
Or,
impaled.
} hanged on it!”
\VS{10}So Haman was hanged on the gallows that had been prepared for Mordecai. Then the king’s wrath was appeased.

\par }\Chap{8}{\PP \VerseOne{1}On that day, king Ahasuerus gave to Esther all that belonged to Haman the slanderer. The king called Mordecai; for Esther had shown that he was related to her.
\VS{2}The king took the ring which he had taken away from Haman and gave it to Mordecai. Esther appointed Mordecai over all that had been Haman’s.
\VS{3}She spoke yet again to the king, and fell at his feet, and implored him to undo Haman’s mischief and all that he had done against the Jews.
\VS{4}Then the king extended the golden sceptre to Esther; and Esther arose to stand near the king.
\VS{5}Esther said, “If it seems good to you, and I have found favor in your sight, let an order be sent that the letters sent by Haman may be reversed, that were written for the destruction of the Jews who are in your kingdom.
\VS{6}For how could I see the affliction of my people, and how could I survive the destruction of my
\FTNT{*}{{\FR 8:6 }
Greek
country.
} kindred?”
\par }{\PP \VS{7}Then the king said to Esther, “If I have given and freely granted you all that was Haman’s, and hanged him on a gallows, because he laid his hands upon the Jews, what more do you seek?
\VS{8}Write in my name whatever seems good to you, and seal it with my ring; for whatever is written at the command of the king, and sealed with my ring, cannot be countermanded.
\VS{9}So the scribes were called in the first month, which is Nisan, on the twenty-third day of the same year; and orders were written to the Jews, whatever the king had commanded to the
\FTNT{†}{{\FR 8:9 }
Greek
stewards.} local governors and chiefs of the local governors, from India even to Ethiopia, one hundred twenty-seven local governors, according to the several provinces, in their own languages.
\VS{10}They were written by order of the king, sealed with his ring, and the letters were sent by the couriers.
\VS{11}In them, he charged them to use their own laws in every city, to help each other, and to treat their adversaries and those who attacked them as they pleased,
\VS{12}on one day in all the kingdom of Ahasuerus, on the thirteenth day of the twelfth month, which is Adar.
\VS{13}Let the copies be posted in conspicuous places throughout the kingdom. Let all the Jews be ready against this day, to fight against their enemies. The following is a copy of the letter containing orders.
\par }{\PP \FTNT{‡}{{\FR 8:13 }
the passages in brackets are not in the
Hebrew.
} [The great king Ahasuerus sends greetings to the rulers of provinces in one hundred twenty-seven local governance regions, from India to Ethiopia, even to those who are faithful to our interests. Many who have been frequently honored by the most abundant kindness of their
\FTNT{§}{{\FR 8:13 }
perhaps rulers, see Luke 22. 25.
} benefactors have conceived ambitious designs, and not only endeavor to hurt our subjects, but moreover, not being able to bear prosperity, they also endeavor to plot against their own benefactors. They not only would utterly abolish gratitude from among men, but also, elated by the boastings of men who are strangers to all that is good, they supposed that they would escape the sin-hating vengeance of the ever-seeing God. And oftentimes evil exhortation has made partakers of the guilt of shedding innocent blood, and has involved in irremediable calamities, many of those who had been appointed to offices of authority, who had been entrusted with the management of their friends’ affairs; while men, by the false sophistry of an evil disposition, have deceived the simple candour of the ruling powers. And it is possible to see this, not so much from more ancient traditionary accounts, as it is immediately in your power to see it by examining what things have been wickedly
\FTNT{*}{{\FR 8:13 }
Greek
contrived.
} perpetrated by the baseness of men unworthily holding power. It is right to take heed with regard to the future, that we may maintain the government in undisturbed peace for all men, adopting needful changes, and ever judging those cases which come under our notice with truly equitable decisions. For whereas Haman, a Macedonian, the son of Hammedatha, in reality an alien from the blood of the Persians, and differing widely from our mild course of government, having been hospitable entertained by us, obtained so large a share of our universal kindness as to be called our father, and to continue the person next to the royal throne, reverenced of all; he however,
\FTNT{†}{{\FR 8:13 }
Greek
not having borne.} overcome by pride, endeavored to deprive us of our dominion, and our
\FTNT{‡}{{\FR 8:13 }
Greek
spirit.
}life; having by various and subtle artifices demanded for destruction both Mordecai our deliverer and perpetual benefactor, and Esther the blameless consort of our kingdom, along with their whole nation. For by these methods he thought, having surprised us in a defenseless state, to transfer the dominion of the Persians to the Macedonians. But we find that the Jews, who have been consigned to destruction by the
\FTNT{§}{{\FR 8:13 }
Greek
thrice guilty.
} most abominable of men, are not malefactors, but living according to the justest laws, and being the sons of the living God, the most high and
\FTNT{*}{{\FR 8:13 }
Greek
greatest.
} mighty, who maintains the kingdom, to us as well as to our forefathers, in the most excellent order. You will therefore do well in refusing to obey the letter sent by Haman the son of Hammedatha, because he who has done these things has been hanged with his whole family at the gates of Susa, Almighty God having swiftly returned to him a worthy punishment. We enjoin you then, having openly published a copy of this letter in every place, to give the Jews permission to use their own lawful customs, and to strengthen them, that on the thirteenth of the twelfth month Adar, on the self-same day, they may defend themselves against those who attack them in a time of affliction. For in the place of the destruction of the chosen race, Almighty God has granted them this time of gladness. Therefore you also, among your notable feasts, must keep a distinct day with all festivity, that both now and hereafter it may be a day of deliverance to us and who are well disposed toward the Persians, but to those that plotted against us a memorial of destruction. And every city and province collectively, which shall not do accordingly, shall be consumed with vengeance by spear and fire. It shall be made not only inaccessible to men, but most hateful to wild beasts and birds forever.] Let the copies be posted in conspicuous places throughout the kingdom and let all the Jews be ready against this day, to fight against their enemies.
\VS{14}So the horsemen went forth with haste to perform the king’s commands. The ordinance was also published in Susa.
\par }{\PP \VS{15}Mordecai went out robed in royal apparel, wearing a golden crown and a diadem of fine purple linen. The people in Susa saw it and rejoiced.
\VS{16}The Jews had light and gladness
\VS{17}in every city and province where the ordinance was published. Wherever the proclamation took place, the Jews had joy and gladness, feasting and mirth. Many of the Gentiles were circumcised and became Jews, for fear of the Jews.

\par }\Chap{9}{\PP \VerseOne{1}Now in the twelfth month, on the thirteenth day of the month which is Adar, the letters written by the king arrived.
\VS{2}In that day, the adversaries of the Jews perished; for no one resisted, through fear of them.
\VS{3}For the chiefs of the local governors, and the princes and the royal scribes, honored the Jews; for the fear of Mordecai was upon them.
\VS{4}For the order of the king was in force, that he should be celebrated in all the kingdom.
\VS{6}In the city Susa the Jews killed five hundred men,
\VS{7}including Pharsannes, Delphon, Phasga,
\VS{8}Pharadatha, Barea, Sarbaca,
\VS{9}Marmasima, Ruphaeus, Arsaeus, and Zabuthaeus,
\VS{10}the ten sons of Haman the son of Hammedatha the Bugaean, the enemy of the Jews, and they plundered their property on the same day.
\VS{11}The number of those who perished in Susa was reported to the king.
\par }{\PP \VS{12}Then the king said to Esther, “The Jews have slain five hundred men in the city Susa. What do you think they have done in the rest of the country? What more do you ask, that it may be done for you?”
\par }{\PP \VS{13}Esther said to the king, “Let it be granted to the Jews to do the same to them tomorrow. Also hang the bodies of the ten sons of Haman.”
\par }{\PP \VS{14}He permitted it to be done; and he gave up to the Jews of the city the bodies of the sons of Haman to hang.
\VS{15}The Jews assembled in Susa on the fourteenth day of Adar and killed three hundred men, but plundered no property.
\VS{16}The rest of the Jews who were in the kingdom assembled, and helped one another, and obtained rest from their enemies; for they destroyed fifteen thousand of them on the thirteenth day of Adar, but took no spoil.
\VS{17}They rested on the fourteenth of the same month, and kept it as a day of rest with joy and gladness.
\VS{18}The Jews in the city Susa assembled also on the fourteenth day and rested; and they also observed the fifteenth with joy and gladness.
\VS{19}On this account then it is that the Jews dispersed in every foreign land keep the fourteenth of Adar as a
\FTNT{*}{{\FR 9:19 }
Greek
good day.
} holy day with joy, each sending gifts of food to his neighbor.
\VS{20}Mordecai wrote these things in a book, and sent them to the Jews, as many as were in the kingdom of Ahasuerus, both those who were near and those who were far away,
\VS{21}to establish these as joyful days, and to keep the fourteenth and fifteenth of Adar;
\VS{22}for on these days the Jews obtained rest from their enemies; and in that month, which was Adar, in which a change was made for them, from mourning to joy, and from sorrow to a good day, to spend the whole of it in good days of
\FTNT{†}{{\FR 9:22 }
Greek
weddings.
} feasting and gladness, sending portions to their friends and to the poor.
\VS{23}And the Jews consented to this as Mordecai wrote to them,
\VS{24}showing how Haman the son of Hammedatha the Macedonian fought against them, how he made a decree and cast
\FTNT{‡}{{\FR 9:24 }
Greek
lot.
} lots to destroy them utterly;
\VS{25}also how he went in to the king, telling him to hang Mordecai; but all the calamities he tried to bring upon the Jews came upon himself, and he was hanged, along with his children.
\VS{26}Therefore these days were called Purim, because of the lots (for in their language they are called Purim) because of the words of this letter, and because of all they suffered on this account, and all that happened to them.
\VS{27}Mordecai established it, and the Jews took upon themselves, and upon their seed, and upon those that were joined to them to observe it, neither would they on any account behave differently; but these days were to be a memorial kept in every generation, city, family, and province.
\VS{28}These days of the Purim shall be kept forever, and their memorial shall not fail in any generation.
\par }{\PP \VS{29}Queen Esther the daughter of Aminadab, and Mordecai the Jew, wrote all that they had done, and the confirmation of the letter of Purim.
\VS{31}Mordecai and Esther the queen established this decision on their own, pledging their own health to their plan.
\VS{32}And Esther established it by a command forever, and it was written for a memorial.

\par }\Chap{10}{\PP \VerseOne{1}The king levied a tax upon his kingdom both by land and sea.
\VS{2}As for his strength and valour, and the wealth and glory of his kingdom, behold, they are written in the book of the Persians and Medes for a memorial.
\par }{\PP \VS{3}Mordecai
\FTNT{*}{{\FR 10:3 }
Greek
succeded to
Or, came into the place of.
} was viceroy to king Ahasuerus, and was a great man in the kingdom, honored by the Jews, and lived his life loved by all his nation.
\FTNT{†}{{\FR 10:3 }
the passages in brackets are not in the
Hebrew.
}
\VS{4}[Mordecai said, “These things have come from God.
\VS{5}For I remember the dream which I had concerning these matters; for not one detail of them has failed.
\VS{6}There was the little spring which became a river, and there was light, and the sun and much water. The river is Esther, whom the king married, and made queen.
\VS{7}The two serpents are Haman and me.
\VS{8}The nations are those
{\ADD{nations}} who combined to destroy the name of the Jews.
\VS{9}But as for my nation, this is Israel, even those who cried to God and were delivered; for the Lord delivered his people. The Lord rescued us out of all these calamities; and God worked such signs and great wonders as have not been done among the nations.
\VS{10}Therefore he ordained two lots. One for the people of God, and one for all the other nations.
\VS{11}And these two lots came for an appointed season, and for a day of judgment, before God, and for all the nations.
\VS{12}God remembered his people and vindicated his inheritance.
\VS{13}They shall observe these days in the month Adar, on the fourteenth and on the fifteenth day of the month, with an assembly, joy, and gladness before God, throughout the generations forever among his people Israel.
\VS{14}In the fourth year of the reign of Ptolemeus and Cleopatra, Dositheus, who said he was a priest and Levite, and Ptolemeus his son, brought this letter of Purim, which they said was the same, and that Lysimachus the son of Ptolemeus, who was in Jerusalem, had interpreted.]
\par }