\NormalFont\ShortTitle{Joel}
{\MTB The Book of
\par }{\MT Joel

\par }\ChapOne{1}{\PP \VerseOne{1}Yahweh’s\FTNT{*}{{\FR 1:1 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.} word that came to Joel, the son of Pethuel.
\par }{\Q \VS{2}Hear this, you elders,
\par }{\QB And listen, all you inhabitants of the land.
\par }{\Q Has this ever happened in your days,
\par }{\QB or in the days of your fathers?
\par }{\Q \VS{3}Tell your children about it,
\par }{\QB and have your children tell their children,
\par }{\QB and their children, another generation.
\par }{\Q \VS{4}What the swarming locust has left, the great locust has eaten.
\par }{\QB What the great locust has left, the grasshopper has eaten.
\par }{\QB What the grasshopper has left, the caterpillar has eaten.
\par }{\Q \VS{5}Wake up, you drunkards, and weep!
\par }{\QB Wail, all you drinkers of wine, because of the sweet wine;
\par }{\QB for it is cut off from your mouth.
\par }{\Q \VS{6}For a nation has come up on my land, strong, and without number.
\par }{\QB His teeth are the teeth of a lion,
\par }{\QB and he has the fangs of a lioness.
\par }{\Q \VS{7}He has laid my vine waste,
\par }{\QB and stripped my fig tree.
\par }{\QB He has stripped its bark, and thrown it away.
\par }{\QB Its branches are made white.
\par }{\Q \VS{8}Mourn like a virgin dressed in sackcloth for the husband of her youth!
\VS{9}The meal offering and the drink offering are cut off from Yahweh’s house.
\par }{\QB The priests, Yahweh’s ministers, mourn.
\par }{\Q \VS{10}The field is laid waste.
\par }{\QB The land mourns, for the grain is destroyed,
\par }{\QB The new wine has dried up,
\par }{\QB and the oil languishes.
\par }{\Q \VS{11}Be confounded, you farmers!
\par }{\QB Wail, you vineyard keepers;
\par }{\QB for the wheat and for the barley;
\par }{\QB for the harvest of the field has perished.
\par }{\Q \VS{12}The vine has dried up, and the fig tree withered;
\par }{\QB the pomegranate tree, the palm tree also, and the apple tree,
\par }{\QB even all of the trees of the field are withered;
\par }{\QB for joy has withered away from the sons of men.
\par }{\Q \VS{13}Put on sackcloth and mourn, you priests!
\par }{\QB Wail, you ministers of the altar.
\par }{\Q Come, lie all night in sackcloth, you ministers of my God,\FTNT{†}{{\FR 1:13 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).}
\par }{\QB for the meal offering and the drink offering are withheld from your God’s house.
\par }{\Q \VS{14}Sanctify a fast.
\par }{\QB Call a solemn assembly.
\par }{\QB Gather the elders,
\par }{\QB and all the inhabitants of the land, to the house of Yahweh, your God,
\par }{\QB and cry to Yahweh.
\par }{\Q \VS{15}Alas for the day!
\par }{\QB For the day of Yahweh is at hand,
\par }{\QB and it will come as destruction from the Almighty.
\par }{\Q \VS{16}Isn’t the food cut off before our eyes;
\par }{\QB joy and gladness from the house of our God?
\par }{\Q \VS{17}The seeds rot under their clods.
\par }{\QB The granaries are laid desolate.
\par }{\QB The barns are broken down, for the grain has withered.
\par }{\Q \VS{18}How the animals groan!
\par }{\QB The herds of livestock are perplexed, because they have no pasture.
\par }{\QB Yes, the flocks of sheep are made desolate.
\par }{\Q \VS{19}Yahweh, I cry to you,
\par }{\QB For the fire has devoured the pastures of the wilderness,
\par }{\QB and the flame has burned all the trees of the field.
\par }{\Q \VS{20}Yes, the animals of the field pant to you,
\par }{\QB for the water brooks have dried up,
\par }{\QB And the fire has devoured the pastures of the wilderness.
\par }{\BB \par }
\PoetryChap{2}{\Q \VerseOne{1}Blow the trumpet in Zion,
\par }{\QB and sound an alarm in my holy mountain!
\par }{\Q Let all the inhabitants of the land tremble,
\par }{\QB for the day of Yahweh comes,
\par }{\QB for it is close at hand:
\par }{\Q \VS{2}A day of darkness and gloominess,
\par }{\QB a day of clouds and thick darkness.
\par }{\Q As the dawn spreading on the mountains,
\par }{\QB a great and strong people;
\par }{\QB there has never been the like,
\par }{\QB neither will there be any more after them,
\par }{\QB even to the years of many generations.
\par }{\Q \VS{3}A fire devours before them,
\par }{\QB and behind them, a flame burns.
\par }{\Q The land is as the garden of Eden before them,
\par }{\QB and behind them, a desolate wilderness.
\par }{\Q Yes, and no one has escaped them.
\par }{\Q \VS{4}Their appearance is as the appearance of horses,
\par }{\QB and they run as horsemen.
\par }{\Q \VS{5}Like the noise of chariots on the tops of the mountains, they leap,
\par }{\QB like the noise of a flame of fire that devours the stubble,
\par }{\QB like a strong people set in battle array.
\par }{\Q \VS{6}At their presence the peoples are in anguish.
\par }{\QB All faces have grown pale.
\par }{\Q \VS{7}They run like mighty men.
\par }{\QB They climb the wall like warriors.
\par }{\QB They each march in his line, and they don’t swerve off course.
\par }{\Q \VS{8}Neither does one jostle another;
\par }{\QB they march everyone in his path,
\par }{\QB and they burst through the defenses,
\par }{\QB and don’t break ranks.
\par }{\Q \VS{9}They rush on the city.
\par }{\QB They run on the wall.
\par }{\QB They climb up into the houses.
\par }{\QB They enter in at the windows like thieves.
\par }{\Q \VS{10}The earth quakes before them.
\par }{\QB The heavens tremble.
\par }{\QB The sun and the moon are darkened,
\par }{\QB and the stars withdraw their shining.
\par }{\Q \VS{11}Yahweh thunders his voice before his army;
\par }{\QB for his forces are very great;
\par }{\QB for he is strong who obeys his command;
\par }{\QB for the day of Yahweh is great and very awesome,
\par }{\QB and who can endure it?
\par }{\Q \VS{12}“Yet even now,” says Yahweh, “turn to me with all your heart,
\par }{\QB and with fasting, and with weeping, and with mourning.”
\par }{\Q \VS{13}Tear your heart, and not your garments,
\par }{\QB and turn to Yahweh, your God;
\par }{\QB for he is gracious and merciful,
\par }{\QB slow to anger, and abundant in loving kindness,
\par }{\QB and relents from sending calamity.
\par }{\Q \VS{14}Who knows? He may turn and relent,
\par }{\QB and leave a blessing behind him,
\par }{\QB even a meal offering and a drink offering to Yahweh, your God.
\par }{\Q \VS{15}Blow the trumpet in Zion!
\par }{\QB Sanctify a fast.
\par }{\QB Call a solemn assembly.
\par }{\Q \VS{16}Gather the people.
\par }{\QB Sanctify the assembly.
\par }{\QB Assemble the elders.
\par }{\QB Gather the children, and those who nurse from breasts.
\par }{\Q Let the bridegroom go out of his room,
\par }{\QB and the bride out of her room.
\par }{\Q \VS{17}Let the priests, the ministers of Yahweh, weep between the porch and the altar,
\par }{\QB and let them say, “Spare your people, Yahweh,
\par }{\QB and don’t give your heritage to reproach,
\par }{\QB that the nations should rule over them.
\par }{\Q Why should they say among the peoples,
\par }{\QB ‘Where is their God?’ ”
\par }{\Q \VS{18}Then Yahweh was jealous for his land,
\par }{\QB And had pity on his people.
\par }{\Q \VS{19}Yahweh answered his people,
\par }{\QB “Behold,\FTNT{*}{{\FR 2:19 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} I will send you grain, new wine, and oil,
\par }{\QB and you will be satisfied with them;
\par }{\QB and I will no more make you a reproach among the nations.
\par }{\Q \VS{20}But I will remove the northern army far away from you,
\par }{\QB and will drive it into a barren and desolate land,
\par }{\QB its front into the eastern sea,
\par }{\QB and its back into the western sea;
\par }{\QB and its stench will come up,
\par }{\QB and its bad smell will rise.”
\par }{\Q Surely he has done great things.
\par }{\Q \VS{21}Land, don’t be afraid.
\par }{\QB Be glad and rejoice, for Yahweh has done great things.
\par }{\Q \VS{22}Don’t be afraid, you animals of the field;
\par }{\QB for the pastures of the wilderness spring up,
\par }{\QB for the tree bears its fruit.
\par }{\QB The fig tree and the vine yield their strength.
\par }{\BB \par }{\Q \VS{23}“Be glad then, you children of Zion,
\par }{\QB and rejoice in Yahweh, your God;
\par }{\QB for he gives you the early rain in just measure,
\par }{\QB and he causes the rain to come down for you,
\par }{\QB the early rain and the latter rain,
\par }{\QB as before.
\par }{\Q \VS{24}The threshing floors will be full of wheat,
\par }{\QB and the vats will overflow with new wine and oil.
\par }{\Q \VS{25}I will restore to you the years that the swarming locust has eaten,
\par }{\QB the great locust, the grasshopper, and the caterpillar,
\par }{\QB my great army, which I sent among you.
\par }{\Q \VS{26}You will have plenty to eat, and be satisfied,
\par }{\QB and will praise the name of Yahweh, your God,
\par }{\QB who has dealt wondrously with you;
\par }{\QB and my people will never again be disappointed.
\par }{\Q \VS{27}You will know that I am among Israel,
\par }{\QB and that I am Yahweh, your God, and there is no one else;
\par }{\QB and my people will never again be disappointed.
\par }{\BB \par }{\Q \VS{28}“It will happen afterward, that I will pour out my Spirit on all flesh;
\par }{\QB and your sons and your daughters will prophesy.
\par }{\QB Your old men will dream dreams.
\par }{\QB Your young men will see visions.
\par }{\Q \VS{29}And also on the servants and on the handmaids in those days,
\par }{\QB I will pour out my Spirit.
\par }{\Q \VS{30}I will show wonders in the heavens and in the earth:
\par }{\QB blood, fire, and pillars of smoke.
\par }{\Q \VS{31}The sun will be turned into darkness,
\par }{\QB and the moon into blood,
\par }{\QB before the great and terrible day of Yahweh comes.
\par }{\Q \VS{32}It will happen that whoever will call on Yahweh’s name shall be saved;
\par }{\QB for in Mount Zion and in Jerusalem there will be those who escape,
\par }{\QB as Yahweh has said,
\par }{\QB and among the remnant, those whom Yahweh calls.
\par }{\BB \par }
\PoetryChap{3}{\Q \VerseOne{1}“For, behold, in those days,
\par }{\QB and in that time,
\par }{\QB when I restore the fortunes of Judah and Jerusalem,
\par }{\Q \VS{2}I will gather all nations,
\par }{\QB and will bring them down into the valley of Jehoshaphat;
\par }{\QB and I will execute judgment on them there for my people,
\par }{\QB and for my heritage, Israel, whom they have scattered among the nations.
\par }{\QB They have divided my land,
\par }{\QB \VS{3}and have cast lots for my people,
\par }{\QB and have given a boy for a prostitute,
\par }{\QB and sold a girl for wine, that they may drink.
\par }{\BB \par }{\Q \VS{4}“Yes, and what are you to me, Tyre, and Sidon,
\par }{\QB and all the regions of Philistia?
\par }{\Q Will you repay me?
\par }{\QB And if you repay me,
\par }{\QB I will swiftly and speedily return your repayment on your own head.
\par }{\Q \VS{5}Because you have taken my silver and my gold,
\par }{\QB and have carried my finest treasures into your temples,
\par }{\QB \VS{6}and have sold the children of Judah and the children of Jerusalem to the sons of the Greeks,
\par }{\QB that you may remove them far from their border.
\par }{\Q \VS{7}Behold, I will stir them up out of the place where you have sold them,
\par }{\QB and will return your repayment on your own head;
\par }{\Q \VS{8}and I will sell your sons and your daughters into the hands of the children of Judah,
\par }{\QB and they will sell them to the men of Sheba,
\par }{\QB to a faraway nation,
\par }{\QB for Yahweh has spoken it.”
\par }{\BB \par }{\Q \VS{9}Proclaim this among the nations:
\par }{\QB “Prepare for war!
\par }{\QB Stir up the mighty men.
\par }{\Q Let all the warriors draw near.
\par }{\QB Let them come up.
\VS{10}Beat your plowshares into swords,
\par }{\QB and your pruning hooks into spears.
\par }{\QB Let the weak say, ‘I am strong.’
\par }{\Q \VS{11}Hurry and come, all you surrounding nations,
\par }{\QB and gather yourselves together.”
\par }{\Q Cause your mighty ones to come down there, Yahweh.
\par }{\Q \VS{12}“Let the nations arouse themselves,
\par }{\QB and come up to the valley of Jehoshaphat;
\par }{\QB for there I will sit to judge all the surrounding nations.
\par }{\Q \VS{13}Put in the sickle;
\par }{\QB for the harvest is ripe.
\par }{\QB Come, tread, for the wine press is full,
\par }{\QB the vats overflow, for their wickedness is great.”
\par }{\Q \VS{14}Multitudes, multitudes in the valley of decision!
\par }{\QB For the day of Yahweh is near, in the valley of decision.
\par }{\Q \VS{15}The sun and the moon are darkened,
\par }{\QB and the stars withdraw their shining.
\par }{\Q \VS{16}Yahweh will roar from Zion,
\par }{\QB and thunder from Jerusalem;
\par }{\QB and the heavens and the earth will shake;
\par }{\QB but Yahweh will be a refuge to his people,
\par }{\QB and a stronghold to the children of Israel.
\par }{\Q \VS{17}“So you will know that I am Yahweh, your God,
\par }{\QB dwelling in Zion, my holy mountain.
\par }{\Q Then Jerusalem will be holy,
\par }{\QB and no strangers will pass through her any more.
\par }{\Q \VS{18}It will happen in that day,
\par }{\QB that the mountains will drop down sweet wine,
\par }{\QB the hills will flow with milk,
\par }{\QB all the brooks of Judah will flow with waters,
\par }{\QB and a fountain will flow out from Yahweh’s house,
\par }{\QB and will water the valley of Shittim.
\par }{\Q \VS{19}Egypt will be a desolation,
\par }{\QB and Edom will be a desolate wilderness,
\par }{\QB for the violence done to the children of Judah,
\par }{\QB because they have shed innocent blood in their land.
\par }{\Q \VS{20}But Judah will be inhabited forever,
\par }{\QB and Jerusalem from generation to generation.
\par }{\Q \VS{21}I will cleanse their blood,
\par }{\QB that I have not cleansed:
\par }{\QB for Yahweh dwells in Zion.”
\par }