\NormalFont\ShortTitle{Lamentations}
{\MTB The
\par }{\MT Lamentations
\par }{\MTB of Jeremiah

\par }\ChapOne{1}{\PP \VerseOne{1}How the city sits solitary,
\par }{\QB that was full of people!
\par }{\Q She has become as a widow,
\par }{\QB who was great among the nations!
\par }{\Q She who was a princess among the provinces
\par }{\QB has become a slave!
\par }{\BB \par }{\Q \VS{2}She weeps bitterly in the night.
\par }{\QB Her tears are on her cheeks.
\par }{\Q Among all her lovers
\par }{\QB she has no one to comfort her.
\par }{\Q All her friends have dealt treacherously with her.
\par }{\QB They have become her enemies.
\par }{\BB \par }{\Q \VS{3}Judah has gone into captivity because of affliction,
\par }{\QB and because of great servitude.
\par }{\Q She dwells among the nations.
\par }{\QB She finds no rest.
\par }{\QB All her persecutors overtook her within the straits.
\par }{\BB \par }{\Q \VS{4}The roads to Zion mourn,
\par }{\QB because no one comes to the solemn assembly.
\par }{\Q All her gates are desolate.
\par }{\QB Her priests sigh.
\par }{\Q Her virgins are afflicted,
\par }{\QB and she herself is in bitterness.
\par }{\BB \par }{\Q \VS{5}Her adversaries have become the head.
\par }{\QB Her enemies prosper;
\par }{\Q for Yahweh\FTNT{*}{{\FR 1:5 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.} has afflicted her for the multitude of her transgressions.
\par }{\QB Her young children have gone into captivity before the adversary.
\par }{\BB \par }{\Q \VS{6}All majesty has departed from the daughter of Zion.
\par }{\QB Her princes have become like deer that find no pasture.
\par }{\QB They have gone without strength before the pursuer.
\par }{\BB \par }{\Q \VS{7}Jerusalem remembers in the days of her affliction and of her miseries
\par }{\QB all her pleasant things that were from the days of old;
\par }{\Q when her people fell into the hand of the adversary,
\par }{\QB and no one helped her.
\par }{\Q The adversaries saw her.
\par }{\QB They mocked at her desolations.
\par }{\BB \par }{\Q \VS{8}Jerusalem has grievously sinned.
\par }{\QB Therefore she has become unclean.
\par }{\Q All who honored her despise her,
\par }{\QB because they have seen her nakedness.
\par }{\QB Yes, she sighs, and turns backward.
\par }{\BB \par }{\Q \VS{9}Her filthiness was in her skirts.
\par }{\QB She didn’t remember her latter end.
\par }{\Q Therefore she has come down astoundingly.
\par }{\QB She has no comforter.
\par }{\Q “See, Yahweh, my affliction;
\par }{\QB for the enemy has magnified himself.”
\par }{\BB \par }{\Q \VS{10}The adversary has spread out his hand on all her pleasant things;
\par }{\QB for she has seen that the nations have entered into her sanctuary,
\par }{\QB concerning whom you commanded that they should not enter into your assembly.
\par }{\BB \par }{\Q \VS{11}All her people sigh.
\par }{\QB They seek bread.
\par }{\QB They have given their pleasant things for food to refresh their soul.
\par }{\Q “Look, Yahweh, and see;
\par }{\QB for I have become despised.”
\par }{\BB \par }{\Q \VS{12}“Is it nothing to you, all you who pass by?
\par }{\QB Look, and see if there is any sorrow like my sorrow,
\par }{\QB which is brought on me,
\par }{\QB with which Yahweh has afflicted me in the day of his fierce anger.
\par }{\BB \par }{\Q \VS{13}“From on high has he sent fire into my bones,
\par }{\QB and it prevails against them.
\par }{\Q He has spread a net for my feet.
\par }{\QB He has turned me back.
\par }{\QB He has made me desolate and I faint all day long.
\par }{\BB \par }{\Q \VS{14}“The yoke of my transgressions is bound by his hand.
\par }{\QB They are knit together.
\par }{\QB They have come up on my neck.
\par }{\QB He made my strength fail.
\par }{\Q The Lord\FTNT{†}{{\FR 1:14 }
The word translated “Lord” is “Adonai.”} has delivered me into their hands,
\par }{\QB against whom I am not able to stand.
\par }{\BB \par }{\Q \VS{15}“The Lord has set at nothing all my mighty men within me.
\par }{\QB He has called a solemn assembly against me to crush my young men.
\par }{\QB The Lord has trodden the virgin daughter of Judah as in a wine press.
\par }{\BB \par }{\Q \VS{16}“For these things I weep.
\par }{\QB My eye, my eye runs down with water,
\par }{\QB because the comforter who should refresh my soul is far from me.
\par }{\Q My children are desolate,
\par }{\QB because the enemy has prevailed.”
\par }{\BB \par }{\Q \VS{17}Zion spreads out her hands.
\par }{\QB There is no one to comfort her.
\par }{\Q Yahweh has commanded concerning Jacob,
\par }{\QB that those who are around him should be his adversaries.
\par }{\QB Jerusalem is among them as an unclean thing.
\par }{\BB \par }{\Q \VS{18}“Yahweh is righteous;
\par }{\QB for I have rebelled against his commandment.
\par }{\Q Please hear all you peoples,
\par }{\QB and see my sorrow.
\par }{\QB My virgins and my young men have gone into captivity.
\par }{\BB \par }{\Q \VS{19}“I called for my lovers,
\par }{\QB but they deceived me.
\par }{\Q My priests and my elders gave up the spirit in the city,
\par }{\QB while they sought food for themselves to refresh their souls.
\par }{\BB \par }{\Q \VS{20}“Look, Yahweh; for I am in distress.
\par }{\QB My heart is troubled.
\par }{\Q My heart turns over within me,
\par }{\QB for I have grievously rebelled.
\par }{\Q Abroad, the sword bereaves.
\par }{\QB At home, it is like death.
\par }{\BB \par }{\Q \VS{21}“They have heard that I sigh.
\par }{\QB There is no one to comfort me.
\par }{\Q All my enemies have heard of my trouble.
\par }{\QB They are glad that you have done it.
\par }{\Q You will bring the day that you have proclaimed,
\par }{\QB and they will be like me.
\par }{\BB \par }{\Q \VS{22}“Let all their wickedness come before you.
\par }{\QB Do to them as you have done to me for all my transgressions.
\par }{\Q For my sighs are many,
\par }{\QB and my heart is faint.

\par }\Chap{2}{\PP \VerseOne{1}How has the Lord covered the daughter of Zion with a cloud in his anger!
\par }{\QB He has cast the beauty of Israel down from heaven to the earth,
\par }{\QB and hasn’t remembered his footstool in the day of his anger.
\par }{\BB \par }{\Q \VS{2}The Lord has swallowed up all the dwellings of Jacob
\par }{\QB without pity.
\par }{\Q He has thrown down in his wrath the strongholds of the daughter of Judah.
\par }{\QB He has brought them down to the ground.
\par }{\QB He has profaned the kingdom and its princes.
\par }{\BB \par }{\Q \VS{3}He has cut off all the horn of Israel in fierce anger.
\par }{\QB He has drawn back his right hand from before the enemy.
\par }{\Q He has burned up Jacob like a flaming fire,
\par }{\QB which devours all around.
\par }{\BB \par }{\Q \VS{4}He has bent his bow like an enemy.
\par }{\QB He has stood with his right hand as an adversary.
\par }{\Q Has killed all that were pleasant to the eye.
\par }{\QB In the tent of the daughter of Zion, he has poured out his wrath like fire.
\par }{\BB \par }{\Q \VS{5}The Lord has become as an enemy.
\par }{\QB He has swallowed up Israel.
\par }{\Q He has swallowed up all her palaces.
\par }{\QB He has destroyed his strongholds.
\par }{\QB He has multiplied mourning and lamentation in the daughter of Judah.
\par }{\BB \par }{\Q \VS{6}He has violently taken away his tabernacle,
\par }{\QB as if it were a garden.
\par }{\Q He has destroyed his place of assembly.
\par }{\QB Yahweh has caused solemn assembly and Sabbath to be forgotten in Zion.
\par }{\QB In the indignation of his anger, he has despised the king and the priest.
\par }{\BB \par }{\Q \VS{7}The Lord has cast off his altar.
\par }{\QB He has abhorred his sanctuary.
\par }{\Q He has given the walls of her palaces into the hand of the enemy.
\par }{\QB They have made a noise in Yahweh’s house,
\par }{\QB as in the day of a solemn assembly.
\par }{\BB \par }{\Q \VS{8}Yahweh has purposed to destroy the wall of the daughter of Zion.
\par }{\QB He has stretched out the line.
\par }{\QB He has not withdrawn his hand from destroying;
\par }{\Q He has made the rampart and wall lament.
\par }{\QB They languish together.
\par }{\BB \par }{\Q \VS{9}Her gates have sunk into the ground.
\par }{\QB He has destroyed and broken her bars.
\par }{\Q Her king and her princes are among the nations where the law is not.
\par }{\QB Yes, her prophets find no vision from Yahweh.
\par }{\BB \par }{\Q \VS{10}The elders of the daughter of Zion sit on the ground.
\par }{\QB They keep silence.
\par }{\Q They have cast up dust on their heads.
\par }{\QB They have clothed themselves with sackcloth.
\par }{\QB The virgins of Jerusalem hang down their heads to the ground.
\par }{\BB \par }{\Q \VS{11}My eyes fail with tears.
\par }{\QB My heart is troubled.
\par }{\Q My liver is poured on the earth,
\par }{\QB because of the destruction of the daughter of my people,
\par }{\QB because the young children and the infants swoon in the streets of the city.
\par }{\BB \par }{\Q \VS{12}They ask their mothers,
\par }{\QB “Where is grain and wine?”
\par }{\QB when they swoon as the wounded in the streets of the city,
\par }{\QB when their soul is poured out into their mothers’ bosom.
\par }{\BB \par }{\Q \VS{13}What shall I testify to you?
\par }{\QB What shall I liken to you, daughter of Jerusalem?
\par }{\Q What shall I compare to you,
\par }{\QB that I may comfort you, virgin daughter of Zion?
\par }{\Q For your breach is as big as the sea.
\par }{\QB Who can heal you?
\par }{\BB \par }{\Q \VS{14}Your prophets have seen false and foolish visions for you.
\par }{\QB They have not uncovered your iniquity,
\par }{\QB to reverse your captivity,
\par }{\QB but have seen for you false revelations and causes of banishment.
\par }{\BB \par }{\Q \VS{15}All that pass by clap their hands at you.
\par }{\QB They hiss and wag their head at the daughter of Jerusalem, saying,
\par }{\Q “Is this the city that men called ‘The perfection of beauty,
\par }{\QB the joy of the whole earth’?”
\par }{\BB \par }{\Q \VS{16}All your enemies have opened their mouth wide against you.
\par }{\QB They hiss and gnash their teeth.
\par }{\QB They say, “We have swallowed her up.
\par }{\Q Certainly this is the day that we looked for.
\par }{\QB We have found it.
\par }{\QB We have seen it.”
\par }{\BB \par }{\Q \VS{17}Yahweh has done that which he planned.
\par }{\QB He has fulfilled his word that he commanded in the days of old.
\par }{\Q He has thrown down,
\par }{\QB and has not pitied.
\par }{\Q He has caused the enemy to rejoice over you.
\par }{\QB He has exalted the horn of your adversaries.
\par }{\BB \par }{\Q \VS{18}Their heart cried to the Lord.
\par }{\QB O wall of the daughter of Zion,
\par }{\QB let tears run down like a river day and night.
\par }{\Q Give yourself no relief.
\par }{\QB Don’t let the your eyes rest.
\par }{\BB \par }{\Q \VS{19}Arise, cry out in the night,
\par }{\QB at the beginning of the watches!
\par }{\Q Pour out your heart like water before the face of the Lord.
\par }{\QB Lift up your hands toward him for the life of your young children,
\par }{\QB who faint for hunger at the head of every street.
\par }{\BB \par }{\Q \VS{20}“Look, Yahweh, and see to whom you have done thus!
\par }{\QB Should the women eat their offspring,
\par }{\QB the children that they held and bounced on their knees?
\par }{\QB Should the priest and the prophet be killed in the sanctuary of the Lord?
\par }{\BB \par }{\Q \VS{21}“The youth and the old man lie on the ground in the streets.
\par }{\QB My virgins and my young men have fallen by the sword.
\par }{\Q You have killed them in the day of your anger.
\par }{\QB You have slaughtered, and not pitied.
\par }{\BB \par }{\Q \VS{22}“You have called, as in the day of a solemn assembly, my terrors on every side.
\par }{\QB There was no one that escaped or remained in the day of Yahweh’s anger.
\par }{\QB My enemy has consumed those whom I have cared for and brought up.

\par }\PoetryChap{3}{\Q \VerseOne{1}I am the man who has seen affliction
\par }{\QB by the rod of his wrath.
\par }{\Q \VS{2}He has led me and caused me to walk in darkness,
\par }{\QB and not in light.
\par }{\Q \VS{3}Surely he turns his hand against me
\par }{\QB again and again all day long.
\par }{\BB \par }{\Q \VS{4}He has made my flesh and my skin old.
\par }{\QB He has broken my bones.
\par }{\Q \VS{5}He has built against me,
\par }{\QB and surrounded me with bitterness and hardship.
\par }{\Q \VS{6}He has made me dwell in dark places,
\par }{\QB as those who have been long dead.
\par }{\BB \par }{\Q \VS{7}He has walled me about, so that I can’t go out.
\par }{\QB He has made my chain heavy.
\par }{\Q \VS{8}Yes, when I cry, and call for help,
\par }{\QB he shuts out my prayer.
\par }{\Q \VS{9}He has walled up my ways with cut stone.
\par }{\QB He has made my paths crooked.
\par }{\BB \par }{\Q \VS{10}He is to me as a bear lying in wait,
\par }{\QB as a lion in secret places.
\par }{\Q \VS{11}He has turned away my ways,
\par }{\QB and pulled me in pieces.
\par }{\QB He has made me desolate.
\par }{\Q \VS{12}He has bent his bow,
\par }{\QB and set me as a mark for the arrow.
\par }{\BB \par }{\Q \VS{13}He has caused the shafts of his quiver to enter into my kidneys.
\par }{\QB \VS{14}I have become a derision to all my people,
\par }{\QB and their song all day long.
\par }{\Q \VS{15}He has filled me with bitterness.
\par }{\QB He has stuffed me with wormwood.
\par }{\BB \par }{\Q \VS{16}He has also broken my teeth with gravel.
\par }{\QB He has covered me with ashes.
\par }{\Q \VS{17}You have removed my soul far away from peace.
\par }{\QB I forgot prosperity.
\par }{\Q \VS{18}I said, “My strength has perished,
\par }{\QB along with my expectation from Yahweh.”
\par }{\BB \par }{\Q \VS{19}Remember my affliction and my misery,
\par }{\QB the wormwood and the bitterness.
\par }{\Q \VS{20}My soul still remembers them,
\par }{\QB and is bowed down within me.
\par }{\Q \VS{21}This I recall to my mind;
\par }{\QB therefore I have hope.
\par }{\BB \par }{\Q \VS{22}It is because of Yahweh’s loving kindnesses that we are not consumed,
\par }{\QB because his compassion doesn’t fail.
\par }{\Q \VS{23}They are new every morning.
\par }{\QB Great is your faithfulness.
\par }{\Q \VS{24}“Yahweh is my portion,” says my soul.
\par }{\QB “Therefore I will hope in him.”
\par }{\BB \par }{\Q \VS{25}Yahweh is good to those who wait for him,
\par }{\QB to the soul who seeks him.
\par }{\Q \VS{26}It is good that a man should hope
\par }{\QB and quietly wait for the salvation of Yahweh.
\par }{\QB \VS{27}It is good for a man that he bear the yoke in his youth.
\par }{\BB \par }{\Q \VS{28}Let him sit alone and keep silence,
\par }{\QB because he has laid it on him.
\par }{\Q \VS{29}Let him put his mouth in the dust,
\par }{\QB if it is so that there may be hope.
\par }{\Q \VS{30}Let him give his cheek to him who strikes him.
\par }{\QB Let him be filled full of reproach.
\par }{\BB \par }{\Q \VS{31}For the Lord will not cast off forever.
\par }{\QB \VS{32}For though he causes grief,
\par }{\QB yet he will have compassion according to the multitude of his loving kindnesses.
\par }{\Q \VS{33}For he does not afflict willingly,
\par }{\QB nor grieve the children of men.
\par }{\BB \par }{\Q \VS{34}To crush under foot all the prisoners of the earth,
\par }{\QB \VS{35}to turn away the right of a man before the face of the Most High,
\par }{\QB \VS{36}to subvert a man in his cause, the Lord doesn’t approve.
\par }{\BB \par }{\Q \VS{37}Who is he who says, and it comes to pass,
\par }{\QB when the Lord doesn’t command it?
\par }{\Q \VS{38}Doesn’t evil and good come out of the mouth of the Most High?
\par }{\QB \VS{39}Why does a living man complain,
\par }{\QB a man for the punishment of his sins?
\par }{\BB \par }{\Q \VS{40}Let us search and try our ways,
\par }{\QB and turn again to Yahweh.
\par }{\Q \VS{41}Let’s lift up our heart with our hands to God\FTNT{*}{{\FR 3:41 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).} in the heavens.
\par }{\QB \VS{42}“We have transgressed and have rebelled.
\par }{\QB You have not pardoned.
\par }{\BB \par }{\Q \VS{43}“You have covered us with anger and pursued us.
\par }{\QB You have killed.
\par }{\QB You have not pitied.
\par }{\Q \VS{44}You have covered yourself with a cloud,
\par }{\QB so that no prayer can pass through.
\par }{\Q \VS{45}You have made us an off-scouring and refuse
\par }{\QB in the middle of the peoples.
\par }{\BB \par }{\Q \VS{46}“All our enemies have opened their mouth wide against us.
\par }{\QB \VS{47}Terror and the pit have come on us,
\par }{\QB devastation and destruction.”
\par }{\BB \par }{\Q \VS{48}My eye runs down with streams of water,
\par }{\QB for the destruction of the daughter of my people.
\par }{\Q \VS{49}My eye pours down
\par }{\QB and doesn’t cease,
\par }{\QB without any intermission,
\par }{\Q \VS{50}until Yahweh looks down,
\par }{\QB and sees from heaven.
\par }{\Q \VS{51}My eye affects my soul,
\par }{\QB because of all the daughters of my city.
\par }{\BB \par }{\Q \VS{52}They have chased me relentlessly like a bird,
\par }{\QB those who are my enemies without cause.
\par }{\Q \VS{53}They have cut off my life in the dungeon,
\par }{\QB and have cast a stone on me.
\par }{\Q \VS{54}Waters flowed over my head.
\par }{\QB I said, “I am cut off.”
\par }{\BB \par }{\Q \VS{55}I called on your name, Yahweh,
\par }{\QB out of the lowest dungeon.
\par }{\Q \VS{56}You heard my voice:
\par }{\QB “Don’t hide your ear from my sighing,
\par }{\QB and my cry.”
\par }{\BB \par }{\Q \VS{57}You came near in the day that I called on you.
\par }{\QB You said, “Don’t be afraid.”
\par }{\BB \par }{\Q \VS{58}Lord, you have pleaded the causes of my soul.
\par }{\QB You have redeemed my life.
\par }{\Q \VS{59}Yahweh, you have seen my wrong.
\par }{\QB Judge my cause.
\par }{\Q \VS{60}You have seen all their vengeance
\par }{\QB and all their plans against me.
\par }{\BB \par }{\Q \VS{61}You have heard their reproach, Yahweh,
\par }{\QB and all their plans against me,
\par }{\Q \VS{62}the lips of those that rose up against me,
\par }{\QB and their plots against me all day long.
\par }{\Q \VS{63}You see their sitting down and their rising up.
\par }{\QB I am their song.
\par }{\BB \par }{\Q \VS{64}You will pay them back, Yahweh,
\par }{\QB according to the work of their hands.
\par }{\Q \VS{65}You will give them hardness of heart,
\par }{\QB your curse to them.
\par }{\Q \VS{66}You will pursue them in anger,
\par }{\QB and destroy them from under the heavens of Yahweh.

\par }\PoetryChap{4}{\Q \VerseOne{1}How the gold has become dim!
\par }{\QB The most pure gold has changed!
\par }{\Q The stones of the sanctuary are poured out
\par }{\QB at the head of every street.
\par }{\BB \par }{\Q \VS{2}The precious sons of Zion,
\par }{\QB comparable to fine gold,
\par }{\Q how they are esteemed as earthen pitchers,
\par }{\QB the work of the hands of the potter!
\par }{\BB \par }{\Q \VS{3}Even the jackals offer their breast.
\par }{\QB They nurse their young ones.
\par }{\Q But the daughter of my people has become cruel,
\par }{\QB like the ostriches in the wilderness.
\par }{\BB \par }{\Q \VS{4}The tongue of the nursing child clings to the roof of his mouth for thirst.
\par }{\QB The young children ask bread,
\par }{\QB and no one breaks it for them.
\par }{\BB \par }{\Q \VS{5}Those who ate delicacies are desolate in the streets.
\par }{\QB Those who were brought up in purple embrace dunghills.
\par }{\BB \par }{\Q \VS{6}For the iniquity of the daughter of my people is greater than the sin of Sodom,
\par }{\QB which was overthrown as in a moment.
\par }{\QB No hands were laid on her.
\par }{\BB \par }{\Q \VS{7}Her nobles were purer than snow.
\par }{\QB They were whiter than milk.
\par }{\Q They were more ruddy in body than rubies.
\par }{\QB Their polishing was like sapphire.
\par }{\BB \par }{\Q \VS{8}Their appearance is blacker than a coal.
\par }{\QB They are not known in the streets.
\par }{\Q Their skin clings to their bones.
\par }{\QB It is withered.
\par }{\QB It has become like a stick.
\par }{\BB \par }{\Q \VS{9}Those who are killed with the sword are better than those who are killed with hunger;
\par }{\QB For these pine away, stricken through,
\par }{\QB for lack of the fruits of the field.
\par }{\BB \par }{\Q \VS{10}The hands of the pitiful women have boiled their own children.
\par }{\QB They were their food in the destruction of the daughter of my people.
\par }{\BB \par }{\Q \VS{11}Yahweh has accomplished his wrath.
\par }{\QB He has poured out his fierce anger.
\par }{\Q He has kindled a fire in Zion,
\par }{\QB which has devoured its foundations.
\par }{\BB \par }{\Q \VS{12}The kings of the earth didn’t believe,
\par }{\QB neither did all the inhabitants of the world,
\par }{\QB that the adversary and the enemy would enter into the gates of Jerusalem.
\par }{\BB \par }{\Q \VS{13}It is because of the sins of her prophets
\par }{\QB and the iniquities of her priests,
\par }{\QB That have shed the blood of the just in the middle of her.
\par }{\BB \par }{\Q \VS{14}They wander as blind men in the streets.
\par }{\QB They are polluted with blood,
\par }{\QB So that men can’t touch their garments.
\par }{\BB \par }{\Q \VS{15}“Go away!” they cried to them.
\par }{\QB “Unclean! Go away! Go away! Don’t touch!
\par }{\Q When they fled away and wandered, men said among the nations,
\par }{\QB “They can’t live here any more.”
\par }{\BB \par }{\Q \VS{16}Yahweh’s anger has scattered them.
\par }{\QB He will not pay attention to them any more.
\par }{\Q They didn’t respect the persons of the priests.
\par }{\QB They didn’t favor the elders.
\par }{\BB \par }{\Q \VS{17}Our eyes still fail,
\par }{\QB looking in vain for our help.
\par }{\QB In our watching we have watched for a nation that could not save.
\par }{\BB \par }{\Q \VS{18}They hunt our steps,
\par }{\QB so that we can’t go in our streets.
\par }{\Q Our end is near.
\par }{\QB Our days are fulfilled,
\par }{\QB for our end has come.
\par }{\BB \par }{\Q \VS{19}Our pursuers were swifter than the eagles of the sky.
\par }{\QB They chased us on the mountains.
\par }{\QB They set an ambush for us in the wilderness.
\par }{\BB \par }{\Q \VS{20}The breath of our nostrils,
\par }{\QB the anointed of Yahweh,
\par }{\QB was taken in their pits;
\par }{\Q of whom we said,
\par }{\QB under his shadow we will live among the nations.
\par }{\BB \par }{\Q \VS{21}Rejoice and be glad, daughter of Edom,
\par }{\QB that dwells in the land of Uz.
\par }{\Q The cup will pass through to you also.
\par }{\QB You will be drunken,
\par }{\QB and will make yourself naked.
\par }{\BB \par }{\Q \VS{22}The punishment of your iniquity is accomplished, daughter of Zion.
\par }{\QB He will no more carry you away into captivity.
\par }{\Q He will visit your iniquity, daughter of Edom.
\par }{\QB He will uncover your sins.

\par }\PoetryChap{5}{\Q \VerseOne{1}Remember, Yahweh, what has come on us.
\par }{\QB Look, and see our reproach.
\par }{\Q \VS{2}Our inheritance has been turned over to strangers,
\par }{\QB our houses to aliens.
\par }{\Q \VS{3}We are orphans and fatherless.
\par }{\QB Our mothers are as widows.
\par }{\Q \VS{4}We have drunken our water for money.
\par }{\QB Our wood is sold to us.
\par }{\Q \VS{5}Our pursuers are on our necks.
\par }{\QB We are weary, and have no rest.
\par }{\Q \VS{6}We have given our hands to the Egyptians,
\par }{\QB and to the Assyrians, to be satisfied with bread.
\par }{\Q \VS{7}Our fathers sinned, and are no more.
\par }{\QB We have borne their iniquities.
\par }{\Q \VS{8}Servants rule over us.
\par }{\QB There is no one to deliver us out of their hand.
\par }{\Q \VS{9}We get our bread at the peril of our lives,
\par }{\QB because of the sword of the wilderness.
\par }{\Q \VS{10}Our skin is black like an oven,
\par }{\QB because of the burning heat of famine.
\par }{\Q \VS{11}They ravished the women in Zion,
\par }{\QB the virgins in the cities of Judah.
\par }{\Q \VS{12}Princes were hanged up by their hands.
\par }{\QB The faces of elders were not honored.
\par }{\Q \VS{13}The young men carry millstones.
\par }{\QB The children stumbled under loads of wood.
\par }{\Q \VS{14}The elders have ceased from the gate,
\par }{\QB and the young men from their music.
\par }{\Q \VS{15}The joy of our heart has ceased.
\par }{\QB Our dance is turned into mourning.
\par }{\Q \VS{16}The crown has fallen from our head.
\par }{\QB Woe to us, for we have sinned!
\par }{\Q \VS{17}For this our heart is faint.
\par }{\QB For these things our eyes are dim.
\par }{\Q \VS{18}For the mountain of Zion, which is desolate.
\par }{\QB The foxes walk on it.
\par }{\BB \par }{\Q \VS{19}You, Yahweh, remain forever.
\par }{\QB Your throne is from generation to generation.
\par }{\Q \VS{20}Why do you forget us forever,
\par }{\QB and forsake us for so long a time?
\par }{\Q \VS{21}Turn us to yourself, Yahweh, and we will be turned.
\par }{\QB Renew our days as of old.
\par }{\Q \VS{22}But you have utterly rejected us.
\par }{\QB You are very angry against us.
\par }