\NormalFont\ShortTitle{Man}
{\MTC THE
\par }{\MT PRAYER OF MANASSES
\par }{\MTB KING OF JUDAH,
\par }{\MTC WHEN HE WAS HELD CAPTIVE IN BABYLON
\par }{\IP {\BK{The Prayer of Manasses}} is recognized as Deuterocanonical Scripture by the Greek Orthodox and Russian Orthodox Churches. It is included in an appendix to the Latin Vulgate Bible.

\par }\OneChap {\PP \VerseOne{1}O LORD Almighty, that are in heaven, you God of our fathers, of Abraham, and Isaac, and Jacob, and of their righteous seed;
\VS{2}who have made heaven and earth, with all the
\FTNT{*}{{\FR 1:2 }
Or,
order
Or,
array
Genesis 2:1 (Sept.).} ornament thereof;
\VS{3}who have bound the sea by the word of your commandment; who have shut up the deep, and sealed it by your terrible and glorious name;
\VS{4}whom all things fear, yes, tremble before your power;
\VS{5}for the majesty of your glory can’t be borne, and the anger of your threatening toward sinners is importable:
\VS{6}your merciful promise is unmeasurable and unsearchable;
\VS{7}for you are the Lord Most High, of great compassion, patient and abundant in mercy, and repent
\FTNT{†}{{\FR 1:7 }
Gr.
of the evils of men.} of bringing evils upon men.
\VS{8}\FTNT{‡}{{\FR 1:8 }
The Alex. MS. omits
You... saved.}You, O Lord, according to your great goodness have promised repentance and forgiveness to those who have sinned against you: and of your infinite mercies have appointed repentance to sinners, that they may be saved. You therefore, O Lord, that are the God of the just, have not appointed repentance to the just, to Abraham, and Isaac, and Jacob, which have not sinned against you; but you have appointed repentance to me that am a sinner:
\VS{9}for I have sinned above the number of the sands of the sea. My transgressions are multiplied,
\FTNT{§}{{\FR 1:9 }
The Alex. MS. omits
O Lord... multiplied.} O Lord: my transgressions are multiplied, and I am not worthy to behold and see the height of heaven for the multitude of my iniquities.
\VS{10}I am bowed down with many iron bands, that I can’t lift up my head
\FTNT{*}{{\FR 1:10 }
Some authorities omit
by reason of my sins.}by reason of my sins, neither have I any respite: for I have provoked your wrath, and done that which is evil before you:
\FTNT{†}{{\FR 1:10 }
The Alex. MS. omits
I did... commandments.}I didn’t do your will, neither did I keep your commandments: I have set up abominations, and have multiplied
\FTNT{‡}{{\FR 1:10 }
Or,
stumbling blocks}detestable things.
\VS{11}Now therefore I bow the knee of my heart, beseeching you of grace.
\VS{12}I have sinned, O Lord, I have sinned, and I acknowledge my iniquities:
\VS{13}but, I humbly beseech you, forgive me, O Lord, forgive me, and destroy me not with my iniquities. Be not angry with me forever, by reserving evil for me; neither condemn me into the lower parts of the earth. For you,
\FTNT{§}{{\FR 1:13 }
Another reading is,
O God.} O Lord, are the God of those who repent;
\VS{14}and in me you will show all your goodness: for you will save me, that am unworthy, according to your great mercy.
\VS{15}And I will praise you forever all the days of my life: for all the army of heaven sings your praise, and yours is the glory forever and ever. Amen.
\par }