\NormalFont\ShortTitle{Micah}
{\MTB The Book of
\par }{\MT Micah

\par }\ChapOne{1}{\PP \VerseOne{1}Yahweh’s\FTNT{*}{{\FR 1:1 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.} word that came to Micah the Morashtite in the days of Jotham, Ahaz, and Hezekiah, kings of Judah, which he saw concerning Samaria and Jerusalem.
\par }{\Q \VS{2}Hear, you peoples, all of you.
\par }{\QB Listen, O earth, and all that is therein:
\par }{\Q and let the Lord\FTNT{†}{{\FR 1:2 }
The word translated “Lord” is “Adonai.”} Yahweh be witness against you,
\par }{\QB the Lord from his holy temple.
\par }{\Q \VS{3}For, behold,\FTNT{‡}{{\FR 1:3 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} Yahweh comes out of his place,
\par }{\QB and will come down and tread on the high places of the earth.
\par }{\Q \VS{4}The mountains melt under him,
\par }{\QB and the valleys split apart,
\par }{\QB like wax before the fire,
\par }{\QB like waters that are poured down a steep place.
\par }{\BB \par }{\Q \VS{5}“All this is for the disobedience of Jacob,
\par }{\QB and for the sins of the house of Israel.
\par }{\Q What is the disobedience of Jacob?
\par }{\QB Isn’t it Samaria?
\par }{\Q And what are the high places of Judah?
\par }{\QB Aren’t they Jerusalem?
\par }{\Q \VS{6}Therefore I will make Samaria like a rubble heap of the field,
\par }{\QB like places for planting vineyards;
\par }{\Q and I will pour down its stones into the valley,
\par }{\QB and I will uncover its foundations.
\par }{\Q \VS{7}All her idols will be beaten to pieces,
\par }{\QB and all her temple gifts will be burned with fire,
\par }{\QB and all her images I will destroy;
\par }{\Q for of the hire of a prostitute has she gathered them,
\par }{\QB and to the hire of a prostitute shall they return.”
\par }{\BB \par }{\Q \VS{8}For this I will lament and wail;
\par }{\QB I will go stripped and naked;
\par }{\QB I will howl like the jackals,
\par }{\QB and moan like the daughters of owls.
\par }{\Q \VS{9}For her wounds are incurable;
\par }{\QB for it has come even to Judah.
\par }{\Q It reaches to the gate of my people,
\par }{\QB even to Jerusalem.
\par }{\Q \VS{10}Don’t tell it in Gath.
\par }{\QB Don’t weep at all.
\par }{\QB At Beth Ophrah\FTNT{§}{{\FR 1:10 }
Beth Ophrah means literally “House of Dust.”} I have rolled myself in the dust.
\par }{\Q \VS{11}Pass on, inhabitant of Shaphir, in nakedness and shame.
\par }{\QB The inhabitant of Zaanan won’t come out.
\par }{\QB The wailing of Beth Ezel will take from you his protection.
\par }{\Q \VS{12}For the inhabitant of Maroth waits anxiously for good,
\par }{\QB because evil has come down from Yahweh to the gate of Jerusalem.
\par }{\Q \VS{13}Harness the chariot to the swift steed, inhabitant of Lachish.
\par }{\QB She was the beginning of sin to the daughter of Zion;
\par }{\QB For the transgressions of Israel were found in you.
\par }{\Q \VS{14}Therefore you will give a parting gift to Moresheth Gath.
\par }{\QB The houses of Achzib will be a deceitful thing to the kings of Israel.
\par }{\Q \VS{15}I will yet bring to you, inhabitant of Mareshah, him who will possess you.
\par }{\QB He who is the glory of Israel will come to Adullam.
\par }{\Q \VS{16}Shave your heads,
\par }{\QB and cut off your hair for the children of your delight.
\par }{\Q Enlarge your baldness like the vulture;
\par }{\QB for they have gone into captivity from you!
\par }{\BB \par }
\PoetryChap{2}{\Q \VerseOne{1}Woe to those who devise iniquity
\par }{\QB and work evil on their beds!
\par }{\Q When the morning is light, they practice it,
\par }{\QB because it is in the power of their hand.
\par }{\Q \VS{2}They covet fields, and seize them;
\par }{\QB and houses, and take them away:
\par }{\QB and they oppress a man and his house,
\par }{\QB even a man and his heritage.
\par }{\Q \VS{3}Therefore Yahweh says:
\par }{\Q “Behold, I am planning against these people a disaster,
\par }{\QB from which you will not remove your necks,
\par }{\QB neither will you walk haughtily;
\par }{\QB for it is an evil time.
\par }{\Q \VS{4}In that day they will take up a parable against you,
\par }{\QB and lament with a doleful lamentation, saying,
\par }{\QB ‘We are utterly ruined!
\par }{\QB My people’s possession is divided up.
\par }{\QB Indeed he takes it from me and assigns our fields to traitors!’ ”
\par }{\Q \VS{5}Therefore you will have no one who divides the land by lot in Yahweh’s assembly.
\par }{\Q \VS{6}“Don’t prophesy!”
\par }{\QB They prophesy.
\par }{\Q “Don’t prophesy about these things.
\par }{\QB Disgrace won’t overtake us.”
\par }{\Q \VS{7}Shall it be said, O house of Jacob:
\par }{\QB “Is Yahweh’s Spirit angry?
\par }{\QB Are these his doings?
\par }{\QB Don’t my words do good to him who walks blamelessly?”
\par }{\Q \VS{8}But lately my people have risen up as an enemy.
\par }{\QB You strip the robe and clothing from those who pass by without a care, returning from battle.
\par }{\Q \VS{9}You drive the women of my people out from their pleasant houses;
\par }{\QB from their young children you take away my blessing forever.
\par }{\Q \VS{10}Arise, and depart!
\par }{\QB For this is not your resting place,
\par }{\QB because of uncleanness that destroys,
\par }{\QB even with a grievous destruction.
\par }{\Q \VS{11}If a man walking in a spirit of falsehood lies:
\par }{\QB “I will prophesy to you of wine and of strong drink;”
\par }{\QB he would be the prophet of this people.
\par }{\Q \VS{12}I will surely assemble, Jacob, all of you;
\par }{\QB I will surely gather the remnant of Israel;
\par }{\Q I will put them together as the sheep of Bozrah,
\par }{\QB as a flock in the middle of their pasture;
\par }{\QB they will swarm with people.
\par }{\Q \VS{13}He who breaks open the way goes up before them.
\par }{\QB They break through the gate, and go out.
\par }{\QB And their king passes on before them,
\par }{\QB with Yahweh at their head.

\par }\Chap{3}{\PP \VerseOne{1}I said,
\par }{\Q “Please listen, you heads of Jacob,
\par }{\QB and rulers of the house of Israel:
\par }{\QB Isn’t it for you to know justice?
\par }{\Q \VS{2}You who hate the good,
\par }{\QB and love the evil;
\par }{\QB who tear off their skin,
\par }{\QB and their flesh from off their bones;
\par }{\QB \VS{3}who also eat the flesh of my people,
\par }{\QB and peel their skin from off them,
\par }{\QB and break their bones,
\par }{\QB and chop them in pieces, as for the pot,
\par }{\QB and as meat within the cauldron.
\par }{\Q \VS{4}Then they will cry to Yahweh,
\par }{\QB but he will not answer them.
\par }{\Q Yes, he will hide his face from them at that time,
\par }{\QB because they made their deeds evil.”
\par }{\PP \VS{5}Yahweh says concerning the prophets who lead my people astray; for those who feed their teeth, they proclaim, “Peace!” and whoever doesn’t provide for their mouths, they prepare war against him:
\par }{\Q \VS{6}“Therefore night is over you, with no vision,
\par }{\QB and it is dark to you, that you may not divine;
\par }{\QB and the sun will go down on the prophets,
\par }{\QB and the day will be black over them.
\par }{\Q \VS{7}The seers shall be disappointed,
\par }{\QB and the diviners confounded.
\par }{\Q Yes, they shall all cover their lips;
\par }{\QB for there is no answer from God.”\FTNT{*}{{\FR 3:7 }
The Hebrew word rendered “God” is “אֱלֹהִ֑ים” (Elohim).}
\par }{\Q \VS{8}But as for me, I am full of power by Yahweh’s Spirit,
\par }{\QB and of judgment, and of might,
\par }{\QB to declare to Jacob his disobedience,
\par }{\QB and to Israel his sin.
\par }{\Q \VS{9}Please listen to this, you heads of the house of Jacob,
\par }{\QB and rulers of the house of Israel,
\par }{\QB who abhor justice,
\par }{\QB and pervert all equity.
\par }{\Q \VS{10}They build up Zion with blood,
\par }{\QB and Jerusalem with iniquity.
\par }{\Q \VS{11}Her leaders judge for bribes,
\par }{\QB and her priests teach for a price,
\par }{\QB and her prophets of it tell fortunes for money:
\par }{\Q yet they lean on Yahweh, and say,
\par }{\QB “Isn’t Yahweh among us?
\par }{\QB No disaster will come on us.”
\par }{\Q \VS{12}Therefore Zion for your sake will be plowed like a field,
\par }{\QB and Jerusalem will become heaps of rubble,
\par }{\QB and the mountain of the temple like the high places of a forest.
\par }{\BB \par }
\PoetryChap{4}{\Q \VerseOne{1}But in the latter days,
\par }{\QB it will happen that the mountain of Yahweh’s temple will be established on the top of the mountains,
\par }{\QB and it will be exalted above the hills;
\par }{\QB and peoples will stream to it.
\par }{\Q \VS{2}Many nations will go and say,
\par }{\QB “Come! Let’s go up to the mountain of Yahweh,
\par }{\QB and to the house of the God of Jacob;
\par }{\QB and he will teach us of his ways,
\par }{\QB and we will walk in his paths.”
\par }{\Q For the law will go out of Zion,
\par }{\QB and Yahweh’s word from Jerusalem;
\par }{\Q \VS{3}and he will judge between many peoples,
\par }{\QB and will decide concerning strong nations afar off.
\par }{\QB They will beat their swords into plowshares,
\par }{\QB and their spears into pruning hooks.
\par }{\Q Nation will not lift up sword against nation,
\par }{\QB neither will they learn war any more.
\par }{\Q \VS{4}But they will sit every man under his vine and under his fig tree;
\par }{\QB and no one will make them afraid:
\par }{\QB For the mouth of Yahweh of Armies has spoken.
\VS{5}Indeed all the nations may walk in the name of their gods;
\par }{\QB but we will walk in the name of Yahweh our God forever and ever.
\par }{\Q \VS{6}“In that day,” says Yahweh,
\par }{\QB “I will assemble that which is lame,
\par }{\QB and I will gather that which is driven away,
\par }{\QB and that which I have afflicted;
\par }{\QB \VS{7}and I will make that which was lame a remnant,
\par }{\QB and that which was cast far off a strong nation:
\par }{\QB and Yahweh will reign over them on Mount Zion from then on, even forever.”
\par }{\Q \VS{8}You, tower of the flock, the hill of the daughter of Zion,
\par }{\QB to you it will come,
\par }{\QB yes, the former dominion will come,
\par }{\QB the kingdom of the daughter of Jerusalem.
\par }{\Q \VS{9}Now why do you cry out aloud?
\par }{\QB Is there no king in you?
\par }{\QB Has your counselor perished,
\par }{\QB that pains have taken hold of you as of a woman in travail?
\par }{\Q \VS{10}Be in pain, and labor to give birth, daughter of Zion,
\par }{\QB like a woman in travail;
\par }{\QB for now you will go out of the city,
\par }{\QB and will dwell in the field,
\par }{\QB and will come even to Babylon.
\par }{\Q There you will be rescued.
\par }{\QB There Yahweh will redeem you from the hand of your enemies.
\par }{\Q \VS{11}Now many nations have assembled against you, that say,
\par }{\QB “Let her be defiled,
\par }{\QB and let our eye gloat over Zion.”
\par }{\Q \VS{12}But they don’t know the thoughts of Yahweh,
\par }{\QB neither do they understand his counsel;
\par }{\QB for he has gathered them like the sheaves to the threshing floor.
\par }{\Q \VS{13}Arise and thresh, daughter of Zion;
\par }{\QB for I will make your horn iron,
\par }{\QB and I will make your hoofs bronze;
\par }{\Q and you will beat in pieces many peoples:
\par }{\Q and I will devote their gain to Yahweh,
\par }{\QB and their substance to the Lord of the whole earth.
\par }{\BB \par }
\PoetryChap{5}{\Q \VerseOne{1}Now you shall gather yourself in troops,
\par }{\QB daughter of troops.
\par }{\Q He has laid siege against us.
\par }{\QB They will strike the judge of Israel with a rod on the cheek.
\par }{\Q \VS{2}But you, Bethlehem Ephrathah,
\par }{\QB being small among the clans of Judah,
\par }{\QB out of you one will come out to me that is to be ruler in Israel;
\par }{\QB whose goings out are from of old, from ancient times.
\par }{\Q \VS{3}Therefore he will abandon them until the time that she who is in labor gives birth.
\par }{\QB Then the rest of his brothers will return to the children of Israel.
\par }{\Q \VS{4}He shall stand, and shall shepherd in the strength of Yahweh,
\par }{\QB in the majesty of the name of Yahweh his God:
\par }{\QB and they will live, for then he will be great to the ends of the earth.
\par }{\Q \VS{5}He will be our peace when Assyria invades our land,
\par }{\QB and when he marches through our fortresses,
\par }{\QB then we will raise against him seven shepherds,
\par }{\QB and eight leaders of men.
\par }{\Q \VS{6}They will rule the land of Assyria with the sword,
\par }{\QB and the land of Nimrod in its gates.
\par }{\Q He will deliver us from the Assyrian,
\par }{\QB when he invades our land,
\par }{\QB and when he marches within our border.
\par }{\Q \VS{7}The remnant of Jacob will be among many peoples,
\par }{\QB like dew from Yahweh,
\par }{\QB like showers on the grass,
\par }{\QB that don’t wait for man,
\par }{\QB nor wait for the sons of men.
\par }{\Q \VS{8}The remnant of Jacob will be among the nations,
\par }{\QB among many peoples,
\par }{\QB like a lion among the animals of the forest,
\par }{\QB like a young lion among the flocks of sheep;
\par }{\QB who, if he goes through, treads down and tears in pieces,
\par }{\QB and there is no one to deliver.
\par }{\Q \VS{9}Let your hand be lifted up above your adversaries,
\par }{\QB and let all of your enemies be cut off.
\par }{\Q \VS{10}“It will happen in that day”, says Yahweh,
\par }{\QB “that I will cut off your horses out from among you,
\par }{\QB and will destroy your chariots.
\par }{\Q \VS{11}I will cut off the cities of your land,
\par }{\QB and will tear down all your strongholds.
\par }{\Q \VS{12}I will destroy witchcraft from your hand;
\par }{\QB and you shall have no soothsayers.
\par }{\Q \VS{13}I will cut off your engraved images and your pillars out from among you;
\par }{\QB and you shall no more worship the work of your hands.
\par }{\Q \VS{14}I will uproot your Asherah poles out from among you;
\par }{\QB and I will destroy your cities.
\par }{\Q \VS{15}I will execute vengeance in anger,
\par }{\QB and wrath on the nations that didn’t listen.”

\par }\Chap{6}{\PP \VerseOne{1}Listen now to what Yahweh says:
\par }{\Q “Arise, plead your case before the mountains,
\par }{\QB and let the hills hear what you have to say.
\par }{\Q \VS{2}Hear, you mountains, Yahweh’s controversy,
\par }{\QB and you enduring foundations of the earth;
\par }{\QB for Yahweh has a controversy with his people,
\par }{\QB and he will contend with Israel.
\par }{\Q \VS{3}My people, what have I done to you?
\par }{\QB How have I burdened you?
\par }{\QB Answer me!
\par }{\Q \VS{4}For I brought you up out of the land of Egypt,
\par }{\QB and redeemed you out of the house of bondage.
\par }{\QB I sent before you Moses, Aaron, and Miriam.
\par }{\Q \VS{5}My people, remember now what Balak king of Moab devised,
\par }{\QB and what Balaam the son of Beor answered him from Shittim to Gilgal,
\par }{\QB that you may know the righteous acts of Yahweh.”
\par }{\BB \par }{\Q \VS{6}How shall I come before Yahweh,
\par }{\QB and bow myself before the exalted God?
\par }{\Q Shall I come before him with burnt offerings,
\par }{\QB with calves a year old?
\par }{\Q \VS{7}Will Yahweh be pleased with thousands of rams?
\par }{\QB With tens of thousands of rivers of oil?
\par }{\Q Shall I give my firstborn for my disobedience?
\par }{\QB The fruit of my body for the sin of my soul?
\par }{\Q \VS{8}He has shown you, O man, what is good.
\par }{\QB What does Yahweh require of you, but to act justly,
\par }{\QB to love mercy, and to walk humbly with your God?
\par }{\BB \par }{\Q \VS{9}Yahweh’s voice calls to the city,
\par }{\QB and wisdom sees your name:
\par }{\Q “Listen to the rod,
\par }{\QB and he who appointed it.
\par }{\Q \VS{10}Are there yet treasures of wickedness in the house of the wicked,
\par }{\QB and a short ephah\FTNT{*}{{\FR 6:10 }
An ephah is a measure of volume (about 22 liters or about 2/3 of a bushel), and a short ephah is made smaller than a full ephah for the purpose of cheating customers.} that is accursed?
\par }{\Q \VS{11}Shall I be pure with dishonest scales,
\par }{\QB and with a bag of deceitful weights?
\par }{\Q \VS{12}Her rich men are full of violence,
\par }{\QB her inhabitants speak lies,
\par }{\QB and their tongue is deceitful in their speech.
\par }{\Q \VS{13}Therefore I also have struck you with a grievous wound.
\par }{\QB I have made you desolate because of your sins.
\par }{\Q \VS{14}You shall eat, but not be satisfied.
\par }{\QB Your humiliation will be within you.
\par }{\QB You will store up, but not save;
\par }{\QB and that which you save I will give up to the sword.
\par }{\Q \VS{15}You will sow, but won’t reap.
\par }{\QB You will tread the olives, but won’t anoint yourself with oil;
\par }{\QB and crush grapes, but won’t drink the wine.
\par }{\Q \VS{16}For the statutes of Omri are kept,
\par }{\QB and all the works of Ahab’s house.
\par }{\QB You walk in their counsels,
\par }{\QB that I may make you a ruin,
\par }{\QB and her inhabitants a hissing;
\par }{\QB And you will bear the reproach of my people.”
\par }{\BB \par }
\PoetryChap{7}{\Q \VerseOne{1}Misery is mine!
\par }{\QB Indeed, I am like one who gathers the summer fruits, as gleanings of the vineyard:
\par }{\QB There is no cluster of grapes to eat.
\par }{\QB My soul desires to eat the early fig.
\par }{\Q \VS{2}The godly man has perished out of the earth,
\par }{\QB and there is no one upright among men.
\par }{\QB They all lie in wait for blood;
\par }{\QB every man hunts his brother with a net.
\par }{\Q \VS{3}Their hands are on that which is evil to do it diligently.
\par }{\QB The ruler and judge ask for a bribe;
\par }{\QB and the powerful man dictates the evil desire of his soul.
\par }{\QB Thus they conspire together.
\par }{\Q \VS{4}The best of them is like a brier.
\par }{\QB The most upright is worse than a thorn hedge.
\par }{\Q The day of your watchmen,
\par }{\QB even your visitation, has come;
\par }{\QB now is the time of their confusion.
\par }{\Q \VS{5}Don’t trust in a neighbor.
\par }{\QB Don’t put confidence in a friend.
\par }{\QB With the woman lying in your embrace,
\par }{\QB be careful of the words of your mouth!
\par }{\Q \VS{6}For the son dishonors the father,
\par }{\QB the daughter rises up against her mother,
\par }{\QB the daughter-in-law against her mother-in-law;
\par }{\QB a man’s enemies are the men of his own house.
\par }{\Q \VS{7}But as for me, I will look to Yahweh.
\par }{\QB I will wait for the God of my salvation.
\par }{\QB My God will hear me.
\par }{\Q \VS{8}Don’t rejoice against me, my enemy.
\par }{\QB When I fall, I will arise.
\par }{\QB When I sit in darkness, Yahweh will be a light to me.
\par }{\Q \VS{9}I will bear the indignation of Yahweh,
\par }{\QB because I have sinned against him,
\par }{\QB until he pleads my case, and executes judgment for me.
\par }{\QB He will bring me out to the light.
\par }{\QB I will see his righteousness.
\par }{\Q \VS{10}Then my enemy will see it,
\par }{\QB and shame will cover her who said to me,
\par }{\QB where is Yahweh your God?
\par }{\Q Then my enemy will see me and will cover her shame.
\par }{\QB Now she will be trodden down like the mire of the streets.
\par }{\Q \VS{11}A day to build your walls—
\par }{\QB In that day, he will extend your boundary.
\par }{\Q \VS{12}In that day they will come to you from Assyria and the cities of Egypt,
\par }{\QB and from Egypt even to the River,
\par }{\QB and from sea to sea,
\par }{\QB and mountain to mountain.
\par }{\Q \VS{13}Yet the land will be desolate because of those who dwell therein,
\par }{\QB for the fruit of their doings.
\par }{\Q \VS{14}Shepherd your people with your staff,
\par }{\QB the flock of your heritage,
\par }{\QB who dwell by themselves in a forest,
\par }{\QB in the middle of fertile pasture land, let them feed;
\par }{\QB in Bashan and Gilead, as in the days of old.
\par }{\Q \VS{15}“As in the days of your coming out of the land of Egypt,
\par }{\QB I will show them marvelous things.”
\par }{\Q \VS{16}The nations will see and be ashamed of all their might.
\par }{\QB They will lay their hand on their mouth.
\par }{\QB Their ears will be deaf.
\par }{\Q \VS{17}They will lick the dust like a serpent.
\par }{\QB Like crawling things of the earth they shall come trembling out of their dens.
\par }{\QB They will come with fear to Yahweh our God,
\par }{\QB and will be afraid because of you.
\par }{\Q \VS{18}Who is a God like you, who pardons iniquity,
\par }{\QB and passes over the disobedience of the remnant of his heritage?
\par }{\Q He doesn’t retain his anger forever,
\par }{\QB because he delights in loving kindness.
\par }{\Q \VS{19}He will again have compassion on us.
\par }{\QB He will tread our iniquities under foot;
\par }{\QB and you will cast all their sins into the depths of the sea.
\par }{\Q \VS{20}You will give truth to Jacob,
\par }{\QB and mercy to Abraham,
\par }{\QB as you have sworn to our fathers from the days of old.
\par }