\NormalFont\ShortTitle{Sirach}
{\MT THE WISDOM OF JESUS THE SON OF SIRACH,
\par }{\MTB OR
\par }{\MT ECCLESIASTICUS
\par }{\IP {\BK{The Wisdom of Jesus the Son of Sirach}}, also called
{\BK{Ecclesiasticus}}, is recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches.
\par }{\IS The Prologue of the Wisdom of Jesus the Son of Sirach.
\par }{\IP WHEREAS many and great things have been delivered to us by the law and the prophets, and by the others that have followed in their steps, for which we must give Israel the praise for instruction and wisdom; and since not only the readers need to become skillful themselves, but also those who love learning must be able to profit those who are outside, both by speaking and writing; my grandfather Jesus, having much given himself to the reading of the law, and the prophets, and the other books of our fathers, and having gained great familiarity with them, was also drawn on himself to write somewhat pertaining to instruction and wisdom, in order that those who love learning, and are devoted to these things, might make progress much more by living according to the law. You are entreated therefore to read with favor and attention, and to pardon us, if in any parts of what we have labored to interpret, we may seem to fail in some of the phrases. For things originally spoken in Hebrew don’t have the same force in them when they are translated into another language. Not only these, but the law itself, and the prophecies, and the rest of the books, have no small difference, when they are spoken in their original language. For having come into Egypt in the thirty-eighth year of Energetes the king, and having continued there some time, I found a copy giving no small instruction. I thought it therefore most necessary for me to apply some diligence and travail to translate this book, applying indeed much watchfulness and skill in that space of time to bring the book to an end and publish for them also, who in the land of their travels are desiring to learn, preparing their character in advance, so as to live according to the law.

\par }{\BB \par }\PoetryChap{1}{\Q \VerseOne{1}All wisdom comes from the Lord,
\par }{\QB and is with him forever.
\par }{\Q \VS{2}Who can count the sand of the seas,
\par }{\QB the drops of rain,
\par }{\QB and the days of eternity?
\par }{\Q \VS{3}Who will search out the height of the sky,
\par }{\QB the breadth of the earth, the deep,
\par }{\QB and wisdom?
\par }{\Q \VS{4}Wisdom has been created before all things,
\par }{\QB and the understanding of prudence from everlasting.
\VS{5}\FTNT{*}{{\FR 1:5 }
Verse 5 is omitted by the best authorities:
The source of wisdom is God's word in the highest heaven, and her ways are the eternal commandments.}
\par }{\Q \VS{6}To whom has the root of wisdom been revealed?
\par }{\QB Who has known her shrewd counsels?
\VS{7}\FTNT{†}{{\FR 1:7 }
Verse 7 is omitted by the best authorities:
To whom was the knowledge of wisdom manifested? Who has understood her abundant experience?}
\par }{\Q \VS{8}There is one wise, greatly to be feared,
\par }{\QB sitting upon his throne: the Lord.
\par }{\Q \VS{9}He created her.
\par }{\QB He saw and measured her.
\par }{\QB He poured her out upon all his works.
\par }{\Q \VS{10}She is with all flesh according to his gift.
\par }{\QB He gave her freely to those who love him.
\par }{\Q \VS{11}The fear of the Lord is glory, exultation,
\par }{\QB and gladness, and a crown of rejoicing.
\par }{\Q \VS{12}The fear of the Lord will delight the heart,
\par }{\QB and will give gladness, joy, and length of days.
\par }{\Q \VS{13}Whoever fears the Lord, it will go well with him at the last.
\par }{\QB He will be blessed in the day of his death.
\par }{\Q \VS{14}To fear the Lord is the beginning of wisdom.
\par }{\QB It was created together with the faithful in the womb.
\par }{\Q \VS{15}She
\FTNT{‡}{{\FR 1:15 }
Gr.
nested.
} laid an eternal foundation with men.
\par }{\QB She will be trusted among their offspring.
\par }{\Q \VS{16}To fear the Lord is the fullness of wisdom.
\par }{\QB She inebriates men with her fruits.
\par }{\Q \VS{17}She will fill all her house with desirable things,
\par }{\QB and her storehouses with her produce.
\par }{\Q \VS{18}The fear of the Lord is the crown of wisdom,
\par }{\QB making peace and
\FTNT{§}{{\FR 1:18 }
Gr.
health of cure.
} perfect health to flourish.
\FTNT{*}{{\FR 1:18 }
The remainder of this verse is omitted by the best authorities:
Both are gifts of God for peace; glory opens out for those who love him. He saw her and took her measure.}
\par }{\Q \VS{19}He both saw and measured her.
\par }{\QB He rained down skill and knowledge of understanding,
\par }{\QB and exalted the honor of those who hold her fast.
\par }{\Q \VS{20}To fear the Lord is the root of wisdom.
\par }{\QB Her branches are length of days.
\VS{21}\FTNT{†}{{\FR 1:21 }
Verse 21 is omitted by the best authorities:
The fear of the Lord drives away sins. Where it resides, it will turn away all anger.
}
\par }{\BB \par }{\Q \VS{22}Unjust wrath can never be justified,
\par }{\QB for the sway of his wrath is his downfall.
\par }{\Q \VS{23}A man that is patient will resist for a season,
\par }{\QB and afterward gladness will spring up to him.
\par }{\Q \VS{24}He will hide his words for a season,
\par }{\QB and the lips of many will tell of his understanding.
\par }{\Q \VS{25}A parable of knowledge is in the treasures of wisdom;
\par }{\QB but godliness is an abomination to a sinner.
\par }{\Q \VS{26}If you desire wisdom, keep the commandments
\par }{\QB and the Lord will give her to you freely;
\par }{\Q \VS{27}for the fear of the Lord is wisdom and instruction.
\par }{\QB Faith and humility are his good pleasure.
\par }{\Q \VS{28}Don’t disobey the fear of the Lord.
\par }{\QB Don’t come to him with a double heart.
\par }{\Q \VS{29}Don’t be a hypocrite in the mouths of men.
\par }{\QB Keep watch over your lips.
\par }{\Q \VS{30}Don’t exalt yourself,
\par }{\QB lest you fall and bring dishonor upon your soul.
\par }{\Q The Lord will reveal your secrets
\par }{\QB and will cast you down in the midst of the congregation,
\par }{\QB because you didn’t come to the fear of the Lord
\par }{\QB and your heart was full of deceit.
\par }{\BB \par }
\PoetryChap{2}{\Q \VerseOne{1}My son, if you come to serve the Lord,
\par }{\QB prepare your soul for temptation.
\par }{\Q \VS{2}Set your heart aright, constantly endure,
\par }{\QB and don’t make haste in time of calamity.
\par }{\Q \VS{3}Cling to him, and don’t depart,
\par }{\QB that you may be increased at your latter end.
\par }{\Q \VS{4}Accept whatever is brought upon you,
\par }{\QB and be patient when you suffer humiliation.
\par }{\Q \VS{5}For gold is tried in the fire,
\par }{\QB and acceptable men in the furnace of humiliation.
\par }{\Q \VS{6}Put your trust in him, and he will help you.
\par }{\QB Make your ways straight, and set your hope on him.
\par }{\BB \par }{\Q \VS{7}All you who fear the Lord, wait for his mercy.
\par }{\QB Don’t turn aside, lest you fall.
\par }{\Q \VS{8}All you who fear the Lord, put your trust in him,
\par }{\QB and your reward will not fail.
\par }{\Q \VS{9}All you who fear the Lord, hope for good things,
\par }{\QB and for eternal gladness and mercy.
\par }{\Q \VS{10}Look at the generations of old, and see:
\par }{\QB Who ever put his trust in the Lord, and was ashamed?
\par }{\QB Or who remained in his fear, and was forsaken?
\par }{\QB Or who called upon him, and he despised him?
\par }{\Q \VS{11}For the Lord is full of compassion and mercy.
\par }{\QB He forgives sins and saves in time of affliction.
\par }{\BB \par }{\Q \VS{12}Woe to fearful hearts, to faint hands,
\par }{\QB and to the sinner that goes two ways!
\par }{\Q \VS{13}Woe to the faint heart! For it doesn’t believe.
\par }{\QB Therefore it won’t be defended.
\par }{\Q \VS{14}Woe to you who have lost your patience!
\par }{\QB And what will you all do when the Lord visits you?
\par }{\Q \VS{15}Those who fear the Lord will not disobey his words.
\par }{\QB Those who love him will keep his ways.
\par }{\Q \VS{16}Those who fear the Lord will seek his good pleasure.
\par }{\QB Those who love him will be filled with the law.
\par }{\Q \VS{17}Those who fear the Lord will prepare their hearts,
\par }{\QB and will humble their souls in his sight.
\par }{\Q \VS{18}We will fall into the hands of the Lord,
\par }{\QB and not into the hands of men;
\par }{\Q for as his majesty is,
\par }{\QB so also is his mercy.
\par }{\BB \par }
\PoetryChap{3}{\Q \VerseOne{1}Hear me, your father, O my children,
\par }{\QB and do what you hear, that you all may be saved.
\par }{\Q \VS{2}For the Lord has given the father glory concerning the children,
\par }{\QB and has confirmed the judgement of the mother concerning the sons.
\par }{\Q \VS{3}He who honors his father will make atonement for sins.
\par }{\QB \VS{4}He that gives glory to his mother is as one who lays up treasure.
\par }{\Q \VS{5}Whoever honors his father will have joy in his own children.
\par }{\QB He will be heard in the day of his prayer.
\par }{\Q \VS{6}He who gives glory to his father will have length of days.
\par }{\QB He who listens to the Lord will bring rest to his mother,
\par }{\Q \VS{7}\FTNT{*}{{\FR 3:7 }
Some manuscripts add
those who fear the Lord honor their father,} and will serve under his parents, as to masters.
\par }{\Q \VS{8}Honor your father in deed and word,
\par }{\QB that a blessing may come upon you from him.
\par }{\Q \VS{9}For the blessing of the father establishes the houses of children,
\par }{\QB but the curse of the mother roots out the foundations.
\par }{\BB \par }{\Q \VS{10}Don’t glorify yourself in the dishonor of your father,
\par }{\QB for your father’s dishonor is no glory to you.
\par }{\Q \VS{11}For the glory of a man is from the honor of his father,
\par }{\QB and a mother in dishonor is a reproach to her children.
\par }{\Q \VS{12}My son, help your father in his old age,
\par }{\QB and don’t grieve him as long as he lives.
\par }{\Q \VS{13}If he fails in understanding, have patience with him.
\par }{\QB Don’t dishonor him in your full strength.
\par }{\Q \VS{14}For the relieving of your father will not be forgotten.
\par }{\QB Instead of sins it will be added to build you up.
\par }{\Q \VS{15}In the day of your affliction it will remember you,
\par }{\QB as fair weather upon ice,
\par }{\QB so will your sins also melt away.
\par }{\Q \VS{16}He who forsakes his father is as a blasphemer.
\par }{\QB He who provokes his mother is cursed by the Lord.
\par }{\BB \par }{\Q \VS{17}My son, go on with your business in humility;
\par }{\QB so will you be loved by an acceptable man.
\par }{\Q \VS{18}The greater you are, humble yourself the more,
\par }{\QB and you will find favor before the Lord.
\VS{19}\FTNT{†}{{\FR 3:19 }
Some manuscripts add
Many are lofty and renowned, but he reveals his secrets to the humble.}
\par }{\Q \VS{20}For the power of the Lord is great,
\par }{\QB and he is glorified by those who are lowly.
\par }{\Q \VS{21}Don’t seek things that are too hard for you,
\par }{\QB and don’t search out things that are above your strength.
\par }{\Q \VS{22}Think about the things that have been commanded you,
\par }{\QB for you have no need of the things that are secret.
\par }{\Q \VS{23}Don’t be overly busy in your superfluous works,
\par }{\QB for more things are showed to you than men can understand.
\par }{\Q \VS{24}For the conceit of many has led them astray.
\par }{\QB Evil opinion has caused their judgement to slip.
\par }{\Q \VS{25}\FTNT{‡}{{\FR 3:25 }
Some manuscripts omit verse 25.}There is no light without eyes.
\par }{\QB There is no wisdom without knowledge.
\par }{\BB \par }{\Q \VS{26}A stubborn heart will do badly at the end.
\par }{\QB He who loves danger will perish in it.
\par }{\Q \VS{27}A stubborn heart will be burdened with troubles.
\par }{\QB The sinner will heap sin upon sins.
\par }{\Q \VS{28}The calamity of the proud is no healing,
\par }{\QB for a weed of wickedness has taken root in him.
\par }{\Q \VS{29}The heart of the prudent will understand a parable.
\par }{\QB A wise man desires the ear of a listener.
\par }{\Q \VS{30}Water will quench a flaming fire;
\par }{\QB almsgiving will make atonement for sins.
\par }{\Q \VS{31}He who repays good turns is mindful of that which comes afterward.
\par }{\QB In the time of his falling he will find a support.
\par }{\BB \par }
\PoetryChap{4}{\Q \VerseOne{1}My son, don’t deprive the poor of his living.
\par }{\QB Don’t make the needy eyes wait long.
\par }{\Q \VS{2}Don’t make a hungry soul sorrowful,
\par }{\QB or provoke a man in his distress.
\par }{\Q \VS{3}Don’t add more trouble to a heart that is provoked.
\par }{\QB Don’t put off giving to him who is in need.
\par }{\Q \VS{4}Don’t reject a suppliant in his affliction.
\par }{\QB Don’t turn your face away from a poor man.
\par }{\Q \VS{5}Don’t turn your eye away from one who asks.
\par }{\QB Give no occasion to a man to curse you.
\par }{\Q \VS{6}For if he curses you in the bitterness of his soul,
\par }{\QB he who made him will hear his supplication.
\par }{\BB \par }{\Q \VS{7}Endear yourself to the assembly.
\par }{\QB Bow your head to a great man.
\par }{\Q \VS{8}Incline your ear to a poor man.
\par }{\QB Answer him with peaceful words in humility.
\par }{\Q \VS{9}Deliver him who is wronged from the hand of him that wrongs him;
\par }{\QB Don’t be faint-hearted in giving judgement.
\par }{\Q \VS{10}Be as a father to the fatherless,
\par }{\QB and like a husband to their mother.
\par }{\Q So you will be as a son of the Most High,
\par }{\QB and he will love you more than your mother does.
\par }{\BB \par }{\Q \VS{11}Wisdom exalts her sons,
\par }{\QB and takes hold of those who seek her.
\par }{\Q \VS{12}He who loves her loves life.
\par }{\QB Those who seek her early will be filled with gladness.
\par }{\Q \VS{13}He who holds her fast will inherit glory.
\par }{\QB Where
\FTNT{*}{{\FR 4:13 }
Or,
she
} he enters, the Lord will bless.
\par }{\Q \VS{14}Those who serve her minister to the Holy One.
\par }{\QB The Lord loves those who love her.
\par }{\Q \VS{15}He who gives ear to her will judge the nations.
\par }{\QB He who heeds her will dwell securely.
\par }{\Q \VS{16}If he trusts her, he will inherit her,
\par }{\QB and his generations will possess her.
\par }{\Q \VS{17}For at the first she will walk with him in crooked ways,
\par }{\QB and will bring fear and dread upon him,
\par }{\QB and torment him with her discipline,
\par }{\QB until she may trust his soul, and try him by her judgments.
\par }{\Q \VS{18}Then she will return him again to the straight way,
\par }{\QB and will gladden him, and reveal to him her secrets.
\par }{\Q \VS{19}If he goes astray, she will forsake him,
\par }{\QB and hand him over to his fall.
\par }{\BB \par }{\Q \VS{20}Observe the opportunity, and beware of evil.
\par }{\QB Don’t be ashamed of your soul.
\par }{\Q \VS{21}For there is a shame that brings sin,
\par }{\QB and there is a shame that is glory and grace.
\par }{\Q \VS{22}Don’t show partiality against your soul.
\par }{\QB Don’t revere any man to your falling.
\par }{\Q \VS{23}Don’t refrain from speaking when it is for safety.
\par }{\QB \FTNT{†}{{\FR 4:23 }
Some manuscripts omit this line.
}Don’t hide your wisdom for the sake of seeming fair.
\par }{\Q \VS{24}For wisdom will be known by speech,
\par }{\QB and instruction by the word of the tongue.
\par }{\Q \VS{25}Don’t speak against the truth
\par }{\Q and be shamed for your ignorance.
\par }{\Q \VS{26}Don’t be ashamed to confess your sins.
\par }{\QB Don’t fight the river’s current.
\par }{\Q \VS{27}Don’t lay yourself down for a fool to tread upon.
\par }{\QB Don’t be partial to one that is mighty.
\par }{\Q \VS{28}Strive for the truth to death,
\par }{\QB and the Lord God will fight for you.
\par }{\BB \par }{\Q \VS{29}Don’t be hasty with your tongue,
\par }{\QB or slack and negligent in your deeds.
\par }{\Q \VS{30}Don’t be like a lion in your house,
\par }{\QB or suspicious of your servants.
\par }{\Q \VS{31}Don’t let your hand be stretched out to receive,
\par }{\QB and closed when you should repay.

\par }\PoetryChap{5}{\Q \VerseOne{1}Don’t set your heart upon your goods.
\par }{\QB Don’t say, “They are sufficient for me.”
\par }{\Q \VS{2}Don’t follow your own mind and your strength
\par }{\QB to walk in the desires of your heart.
\par }{\Q \VS{3}Don’t say, “Who will have dominion over me?”
\par }{\QB for the Lord will surely take vengeance on you.
\par }{\BB \par }{\Q \VS{4}Don’t say, “I sinned, and what happened to me?”
\par }{\QB for the Lord is patient.
\par }{\Q \VS{5}Don’t be so confident of atonement
\par }{\QB that you add sin upon sins.
\par }{\Q \VS{6}Don’t say, “His compassion is great.
\par }{\QB He will be pacified for the multitude of my sins,”
\par }{\QB for mercy and wrath are with him,
\par }{\QB and his indignation will rest on sinners.
\par }{\Q \VS{7}Don’t wait to turn to the Lord.
\par }{\QB Don’t put off from day to day;
\par }{\QB for suddenly the wrath of the Lord will come on you,
\par }{\QB and you will perish in the time of vengeance.
\par }{\BB \par }{\Q \VS{8}Don’t set your heart upon unrighteous gains,
\par }{\QB for you will profit nothing in the day of calamity.
\par }{\Q \VS{9}Don’t winnow with every wind.
\par }{\QB Don’t walk in every path.
\par }{\QB This is what the sinner who has a double tongue does.
\par }{\Q \VS{10}Be steadfast in your understanding.
\par }{\QB Let your speech be consistent.
\par }{\BB \par }{\Q \VS{11}Be swift to hear
\par }{\QB and answer with patience.
\par }{\Q \VS{12}If you have understanding, answer your neighbor;
\par }{\QB but if not, put your hand over your mouth.
\par }{\Q \VS{13}Glory and dishonor is in talk.
\par }{\QB A man’s tongue may be his downfall.
\par }{\Q \VS{14}Don’t be called a whisperer.
\par }{\QB Don’t lie in wait with your tongue;
\par }{\QB for shame is on the thief,
\par }{\QB and an evil condemnation is on him who has a double tongue.
\par }{\Q \VS{15}Don’t be ignorant in a great or small matter.

\par }\PoetryChap{6}{\Q \VerseOne{1}Don’t become an enemy instead of a friend;
\par }{\QB for an evil name will inherit shame and reproach.
\par }{\QB So it is with the sinner who has a double tongue.
\par }{\BB \par }{\Q \VS{2}Don’t exalt yourself in the counsel of your soul,
\par }{\QB that your soul be not torn in pieces as a bull.
\par }{\Q \VS{3}You will eat up your leaves, destroy your fruit,
\par }{\QB and leave yourself as a dry tree.
\par }{\Q \VS{4}A wicked soul will destroy him who has it,
\par }{\QB and will make him a laughing stock to his enemies.
\par }{\BB \par }{\Q \VS{5}Sweet words will multiply a man’s friends.
\par }{\QB A gracious tongue will multiply courtesies.
\par }{\Q \VS{6}Let those that are at peace with you be many,
\par }{\QB but your advisers one of a thousand.
\par }{\Q \VS{7}If you want to gain a friend, get him in a time of testing,
\par }{\QB and don’t be in a hurry to trust him.
\par }{\Q \VS{8}For there is a friend just for an occasion.
\par }{\QB He won’t continue in the day of your affliction.
\par }{\Q \VS{9}And there is a friend who turns into an enemy.
\par }{\QB He will discover strife to your reproach.
\par }{\Q \VS{10}And there is a friend who is a companion at the table,
\par }{\QB but he won’t continue in the day of your affliction.
\par }{\Q \VS{11}In your prosperity he will be as yourself,
\par }{\QB and will be bold over your servants.
\par }{\Q \VS{12}If you are brought low, he will be against you,
\par }{\QB and will hide himself from your face.
\par }{\Q \VS{13}Separate yourself from your enemies,
\par }{\QB and beware of your friends.
\par }{\BB \par }{\Q \VS{14}A faithful friend is a strong defense.
\par }{\QB He who has found him has found a treasure.
\par }{\Q \VS{15}There is nothing that can be taken in exchange for a faithful friend.
\par }{\QB His excellency is beyond price.
\par }{\Q \VS{16}A faithful friend is a life-saving medicine.
\par }{\QB Those who fear the Lord will find him.
\par }{\Q \VS{17}He who fears the Lord directs his friendship properly;
\par }{\QB for as he is, so is his neighbor also.
\par }{\BB \par }{\Q \VS{18}My son, gather instruction from your youth up.
\par }{\QB Even when you have gray hair you will find wisdom.
\par }{\Q \VS{19}Come to her as one who plows and sows
\par }{\QB and wait for her good fruit;
\par }{\Q for your toil will be little in her tillage,
\par }{\QB and you will soon eat of her fruit.
\par }{\Q \VS{20}How exceedingly harsh she is to the unlearned!
\par }{\QB He who is without understanding will not remain in her.
\par }{\Q \VS{21}She will rest upon him as a mighty stone of trial.
\par }{\QB He won’t hesitate to cast her from him.
\par }{\Q \VS{22}For wisdom is according to her name.
\par }{\Q She isn’t manifest to many.
\par }{\BB \par }{\Q \VS{23}Give ear, my son, and accept my judgement.
\par }{\QB Don’t refuse my counsel.
\par }{\Q \VS{24}Bring your feet into her fetters,
\par }{\QB and your neck into her chain.
\par }{\Q \VS{25}Put your shoulder under her and bear her.
\par }{\QB Don’t be grieved with her bonds.
\par }{\Q \VS{26}Come to her with all your soul.
\par }{\QB Keep her ways with your whole power.
\par }{\Q \VS{27}Search and seek, and she will be made known to you.
\par }{\QB When you get hold of her, don’t let her go.
\par }{\Q \VS{28}For at the last you will find her rest;
\par }{\QB and she will be turned for you into gladness.
\par }{\Q \VS{29}Her fetters will be to you for a covering of strength,
\par }{\QB and her chains for a robe of glory.
\par }{\Q \VS{30}For there is a golden ornament upon her,
\par }{\QB and her bands are
\XREF{✡}{{\XO 6:30 }
Numbers 15:38} a blue ribbon.
\par }{\Q \VS{31}You shall put her on as a robe of glory,
\par }{\QB and shall put her on as a crown of rejoicing.
\par }{\BB \par }{\Q \VS{32}My son, if you are willing, you will be instructed.
\par }{\QB If you will yield your soul, you will be prudent.
\par }{\Q \VS{33}If you love to hear, you will receive.
\par }{\QB If you incline your ear, you will be wise.
\par }{\Q \VS{34}Stand in the multitude of the elders.
\par }{\QB Attach yourself to whoever is wise.
\par }{\Q \VS{35}Be willing to listen to every godly discourse.
\par }{\QB Don’t let the proverbs of understanding escape you.
\par }{\Q \VS{36}If you see a man of understanding, get to him early.
\par }{\QB Let your foot wear out the steps of his doors.
\par }{\Q \VS{37}Let your mind dwell on the ordinances of the Lord
\par }{\QB and meditate continually on his commandments.
\par }{\Q He will establish your heart
\par }{\QB and your desire for wisdom will be given to you.
\par }{\BB \par }
\PoetryChap{7}{\Q \VerseOne{1}Do no evil,
\par }{\QB so will no evil overtake you.
\par }{\Q \VS{2}Depart from wrong,
\par }{\QB and it will turn away from you.
\par }{\Q \VS{3}My son, don’t sow upon the furrows of unrighteousness,
\par }{\QB and you won’t reap them sevenfold.
\par }{\BB \par }{\Q \VS{4}Don’t seek preeminence from the Lord,
\par }{\QB nor the seat of honor from the king.
\par }{\Q \VS{5}Don’t justify yourself in the presence of the Lord,
\par }{\QB and don’t display your wisdom before the king.
\par }{\Q \VS{6}Don’t seek to be a judge,
\par }{\QB lest you not be able to take away iniquities,
\par }{\QB lest perhaps you fear the person of a mighty man,
\par }{\QB and lay a stumbling block in the way of your uprightness.
\par }{\BB \par }{\Q \VS{7}Don’t sin against the multitude of the city.
\par }{\QB Don’t cast yourself down in the crowd.
\par }{\Q \VS{8}Don’t commit a sin twice,
\par }{\QB for even in one you will not be unpunished.
\par }{\Q \VS{9}Don’t say, “He will look upon the multitude of my gifts.
\par }{\QB When I make an offering to the Most High God, he will accept it.”
\par }{\Q \VS{10}Don’t be faint-hearted in your prayer.
\par }{\QB Don’t neglect to give alms.
\par }{\BB \par }{\Q \VS{11}Don’t laugh a man to scorn when he is in the bitterness of his soul,
\par }{\QB for there is one who humbles and exalts.
\par }{\Q \VS{12}\FTNT{*}{{\FR 7:12 }
Gr.
Don’t plow} Don’t devise a lie against your brother,
\par }{\QB or do the same to a friend.
\par }{\Q \VS{13}Don’t love to make any manner of lie,
\par }{\QB for that is not a good habit.
\par }{\Q \VS{14}Don’t babble in the multitude of elders.
\par }{\QB Don’t repeat your words in your prayer.
\par }{\BB \par }{\Q \VS{15}Don’t hate hard labor
\par }{\QB or farm work, which the Most High has created.
\par }{\Q \VS{16}Don’t number yourself among the multitude of sinners.
\par }{\QB Remember that wrath will not wait.
\par }{\Q \VS{17}Humble your soul greatly,
\par }{\QB for the punishment of the ungodly man is fire and the worm.
\par }{\BB \par }{\Q \VS{18}Don’t exchange a friend for something,
\par }{\QB neither a true brother for the gold of Ophir.
\par }{\Q \VS{19}Don’t deprive yourself of a wise and good wife,
\par }{\QB for her grace is worth more than gold.
\par }{\Q \VS{20}Don’t abuse a servant that works faithfully,
\par }{\QB or a hireling who gives you his life.
\par }{\Q \VS{21}Let your soul love a wise servant.
\par }{\QB Don’t defraud him of liberty.
\par }{\BB \par }{\Q \VS{22}Do you have cattle? Look after them;
\par }{\QB if they are profitable to you, let them stay by you.
\par }{\Q \VS{23}Do you have children? Correct them,
\par }{\QB and bow down their necks from their youth.
\par }{\Q \VS{24}Do you have daughters? Take care of their bodies,
\par }{\QB and don’t be overly indulgent toward them.
\par }{\Q \VS{25}Give your daughter in marriage, and you will have accomplished a great matter.
\par }{\QB Give her to a man of understanding.
\par }{\BB \par }{\Q \VS{26}Do you have a wife after your mind? Don’t cast her out.
\par }{\QB \FTNT{†}{{\FR 7:26 }
Many authorities omit this line
} But don’t trust yourself to one who is hateful.
\par }{\Q \VS{27}Give glory to your father with your whole heart,
\par }{\QB and don’t forget the birth pangs of your mother.
\par }{\Q \VS{28}Remember that you were born of them.
\par }{\QB What will you repay them for the things that they have done for you?
\par }{\BB \par }{\Q \VS{29}Fear the Lord with all your soul;
\par }{\QB and revere his priests.
\par }{\Q \VS{30}With all your strength love him who made you.
\par }{\QB Don’t forsake his ministers.
\par }{\Q \VS{31}Fear the Lord and honor the priest.
\par }{\QB Give him his portion, even as it is commanded you:
\par }{\QB the first fruits, the trespass offering, the gift of the shoulders,
\par }{\QB the sacrifice of sanctification, and the first fruits of holy things.
\par }{\BB \par }{\Q \VS{32}Also stretch out your hand to the poor man,
\par }{\QB that your blessing may be perfected.
\par }{\Q \VS{33}A gift has grace in the sight of every living man.
\par }{\QB Don’t withhold grace for a dead man.
\par }{\Q \VS{34}Don’t be lacking to those who weep,
\par }{\QB and mourn with those who mourn.
\par }{\Q \VS{35}Don’t be slow to visit a sick man,
\par }{\QB for by such things you will gain love.
\par }{\Q \VS{36}In all your words, remember eternity,
\par }{\QB and you will never sin.
\par }{\BB \par }
\PoetryChap{8}{\Q \VerseOne{1}Don’t contend with a mighty man,
\par }{\QB lest perhaps you fall into his hands.
\par }{\Q \VS{2}Don’t strive with a rich man, lest perhaps he overpower you;
\par }{\QB for gold has destroyed many,
\par }{\QB and turned away the hearts of kings.
\par }{\Q \VS{3}Don’t contend with a talkative man.
\par }{\QB Don’t heap wood upon his fire.
\par }{\BB \par }{\Q \VS{4}Don’t jest with a rude man,
\par }{\QB Lest your ancestors be dishonored.
\par }{\Q \VS{5}Don’t reproach a man when he turns from sin.
\par }{\QB Remember that we are all worthy of punishment.
\par }{\Q \VS{6}Don’t dishonor a man in his old age;
\par }{\QB for some of us are also growing old.
\par }{\Q \VS{7}Don’t rejoice over one who is dead.
\par }{\QB Remember that we all die.
\par }{\BB \par }{\Q \VS{8}Don’t neglect the discourse of the wise.
\par }{\QB Be conversant with their proverbs;
\par }{\QB for from them you will learn instruction
\par }{\QB and how to minister to great men.
\par }{\Q \VS{9}Don’t miss the discourse of the aged,
\par }{\QB for they also learned from their fathers,
\par }{\QB because from them you will learn understanding,
\par }{\QB and to give an answer in time of need.
\par }{\BB \par }{\Q \VS{10}Don’t kindle the coals of a sinner,
\par }{\QB lest you be burned with the flame of his fire.
\par }{\Q \VS{11}Don’t rise up from the presence of an insolent man,
\par }{\QB lest he lie in wait as an ambush for your mouth.
\par }{\Q \VS{12}Don’t lend to a man who is mightier than yourself;
\par }{\QB and if you lend, be as one who has lost.
\par }{\Q \VS{13}Don’t be surety above your power.
\par }{\QB If you are surety, think as one who will have to pay.
\par }{\BB \par }{\Q \VS{14}Don’t go to law with a judge;
\par }{\QB for according to his honor they will give judgement for him.
\par }{\Q \VS{15}Don’t go in the way with a rash man,
\par }{\QB lest he be burdensome to you;
\par }{\QB for he will do according to his own will,
\par }{\QB and you will perish with his folly.
\par }{\Q \VS{16}Don’t fight with a wrathful man.
\par }{\QB Don’t travel with him through the desert,
\par }{\QB for blood is as nothing in his sight.
\par }{\QB Where there is no help, he will overthrow you.
\par }{\Q \VS{17}Don’t take counsel with a fool,
\par }{\QB for he will not be able to conceal the matter.
\par }{\Q \VS{18}Do no secret thing before a stranger,
\par }{\QB For you don’t know what it will cause.
\par }{\Q \VS{19}Don’t open your heart to every man.
\par }{\QB Don’t let him return you a favor.
\par }{\BB \par }
\PoetryChap{9}{\Q \VerseOne{1}Don’t be jealous over the wife of your bosom,
\par }{\QB and don’t teach her an evil lesson against yourself.
\par }{\Q \VS{2}Don’t give your soul to a woman,
\par }{\QB that she should set her foot upon your strength.
\par }{\Q \VS{3}Don’t go to meet a woman who plays the prostitute,
\par }{\QB lest perhaps you fall into her snares.
\par }{\Q \VS{4}Don’t use the company of a woman who is a singer,
\par }{\QB lest perhaps you be caught by her attempts.
\par }{\Q \VS{5}Don’t gaze at a maid,
\par }{\QB lest perhaps you be trapped in her penalties.
\par }{\Q \VS{6}Don’t give your soul to prostitutes,
\par }{\QB that you not lose your inheritance.
\par }{\Q \VS{7}Don’t look around you in the streets of the city,
\par }{\QB neither wander in its solitary places.
\par }{\Q \VS{8}Turn your eye away from a beautiful woman,
\par }{\QB and don’t gaze at another’s beauty.
\par }{\Q Many have been led astray by the beauty of a woman;
\par }{\QB and with this, affection is kindled as a fire.
\par }{\Q \VS{9}Don’t sit at all with a woman who has a husband,
\par }{\QB or revel with her at the wine,\FTNT{*}{{\FR 9:9 }
The preceding line of this verse is omitted by the best authorities.
}
\par }{\QB lest perhaps your soul turn away to her,
\par }{\QB and with your spirit you slide into destruction.
\par }{\BB \par }{\Q \VS{10}Don’t forsake an old friend;
\par }{\QB for the new is not comparable to him.
\par }{\Q A new friend is like new wine:
\par }{\QB if it becomes old, you will drink it with gladness.
\par }{\BB \par }{\Q \VS{11}Don’t envy the glory of a sinner;
\par }{\QB for you don’t know what his overthrow will be.
\par }{\Q \VS{12}Don’t delight in the delights of the ungodly.
\par }{\QB Remember they will not go unpunished to
\FTNT{†}{{\FR 9:12 }
Gr.
Hades.
} the grave.
\par }{\BB \par }{\Q \VS{13}Keep yourself far from the man who has
\FTNT{‡}{{\FR 9:13 }
Or,
authority
} power to kill,
\par }{\QB and you will have no suspicion of the fear of death.
\par }{\Q If you come to him, commit no fault,
\par }{\QB lest he take away your life.
\par }{\Q Know surely that you go about in the midst of snares,
\par }{\QB and walk upon the battlements of a city.
\par }{\BB \par }{\Q \VS{14}As well as you can, aim to know your neighbors,
\par }{\QB and take counsel with the wise.
\par }{\Q \VS{15}Let your conversation be with men of understanding.
\par }{\QB Let all your discourse be in the law of the Most High.
\par }{\Q \VS{16}Let just men be companions at your table.
\par }{\QB Let your glorying be in the fear of the Lord.
\par }{\BB \par }{\Q \VS{17}A work is commended because of the hand of the artisan;
\par }{\QB so he who rules the people will be considered wise for his speech.
\par }{\Q \VS{18}A talkative man is dangerous in his city.
\par }{\QB He who is headlong in his speech will be hated.
\par }{\BB \par }
\PoetryChap{10}{\Q \VerseOne{1}A wise judge will instruct his people.
\par }{\QB The government of a man of understanding will be well ordered.
\par }{\Q \VS{2}As is the judge of his people, so are his ministers.
\par }{\QB As the city’s ruler is, so are all those who dwell in it.
\par }{\Q \VS{3}An uninstructed king will destroy his people.
\par }{\QB A city will be established through the understanding of the powerful.
\par }{\Q \VS{4}The authority of the earth is in the Lord’s hand.
\par }{\QB In due time, he will raise up over it one who is profitable.
\par }{\Q \VS{5}A man’s prosperity is in the Lord’s hand.
\par }{\QB He will lay his honor upon the person of the scribe.
\par }{\BB \par }{\Q \VS{6}Don’t be angry with your neighbor for every wrong.
\par }{\QB Do nothing by works of violence.
\par }{\Q \VS{7}Pride is hateful before the Lord and men.
\par }{\QB Unrighteousness is abhorrent in the judgement of both.
\par }{\Q \VS{8}Sovereignty is transferred from nation to nation
\par }{\QB because of iniquities, deeds of violence, and greed for money.
\par }{\Q \VS{9}Why are dirt and ashes proud?
\FTNT{*}{{\FR 10:9 }
Two lines of this verse are here omitted by the best authorities.}
\par }{\QB Because in life, my body decays.
\par }{\Q \VS{10}A long disease mocks the physician.
\par }{\QB He is a king today, and tomorrow he will die.
\par }{\Q \VS{11}For when a man is dead,
\par }{\QB he will inherit creeping things, and beasts, and worms.
\par }{\Q \VS{12}It is the beginning of pride when a man departs from the Lord.
\par }{\QB His heart has departed from him who made him.
\par }{\Q \VS{13}For the beginning of pride is sin.
\par }{\QB He who keeps it will pour out abomination.
\par }{\Q For this cause the Lord brought upon them strange calamities
\par }{\QB and utterly overthrew them.
\par }{\Q \VS{14}The Lord cast down the thrones of rulers
\par }{\QB and set the meek in their place.
\par }{\Q \VS{15}The Lord plucked up the roots of nations
\par }{\QB and planted the lowly in their place.
\par }{\Q \VS{16}The Lord overthrew the lands of nations
\par }{\QB and destroyed them to the foundations of the earth.
\par }{\Q \VS{17}He took some of them away and destroyed them,
\par }{\QB and made their memorial to cease from the earth.
\par }{\Q \VS{18}Pride has not been created for men,
\par }{\QB nor wrathful anger for the offspring of women.
\par }{\BB \par }{\Q \VS{19}What manner of seed has honor?
\par }{\QB The seed of man, those who fear the Lord.
\par }{\Q What manner of seed has no honor?
\par }{\QB The seed of man, those who transgress the commandments.
\par }{\Q \VS{20}In the midst of kindred he who rules them has honor.
\par }{\QB Those who fear the Lord have honor in his eyes.
\VS{21}\FTNT{†}{{\FR 10:21 }
Verse 21 is omitted by the best authorities: Fear of the Lord is the beginning of acceptance, but obstinance and pride are the beginning of rejection.}
\par }{\Q \VS{22}The rich man, the honorable, and the poor
\par }{\QB all glory in the fear of the Lord.
\par }{\Q \VS{23}It is not right to dishonor a poor man who has understanding.
\par }{\QB It is not fitting to glorify a man who is a sinner.
\par }{\Q \VS{24}The great man, the judge, and the mighty man will be glorified.
\par }{\QB There is not one of them greater than he who fears the Lord.
\par }{\Q \VS{25}Free men will minister to a wise servant.
\par }{\QB A man who has knowledge will not complain.
\par }{\BB \par }{\Q \VS{26}Don’t flaunt your wisdom in doing your work.
\par }{\QB Don’t glorify yourself in the time of your distress.
\par }{\Q \VS{27}\FTNT{‡}{{\FR 10:27 }
The Greek text of this verse is uncertain.
} Better is he who labors and abounds in all things,
\par }{\QB Than he who glorifies himself and lacks bread.
\par }{\Q \VS{28}My son, glorify your soul in humility,
\par }{\QB and esteem yourself honor according to your worthiness.
\par }{\Q \VS{29}Who will justify him who sins against his own soul?
\par }{\QB Who will glorify him who dishonors his own life?
\par }{\BB \par }{\Q \VS{30}A poor man is glorified for his knowledge.
\par }{\QB A rich man is glorified for his riches.
\par }{\Q \VS{31}But he who is glorified in poverty, how much more in riches?
\par }{\QB And he who is dishonored in riches, how much more in poverty?
\par }{\BB \par }
\PoetryChap{11}{\Q \VerseOne{1}The wisdom of the lowly will lift up his head,
\par }{\QB and make him sit in the midst of great men.
\par }{\BB \par }{\Q \VS{2}Don’t commend a man for his beauty.
\par }{\QB Don’t abhor a man for his outward appearance.
\par }{\Q \VS{3}The bee is little among flying creatures,
\par }{\QB but what it produces is the best of confections.
\par }{\Q \VS{4}Don’t boast about the clothes you wear,
\par }{\QB and don’t exalt yourself in the day of honor;
\par }{\Q for the Lord’s works are wonderful,
\par }{\QB and his works are hidden among men.
\par }{\Q \VS{5}Many
\FTNT{*}{{\FR 11:5 }
Gr.
tyrants
} kings have sat down upon the ground,
\par }{\QB but one who was never thought of has worn a crown.
\par }{\Q \VS{6}Many mighty men have been greatly disgraced.
\par }{\QB Men of renown have been delivered into other men’s hands.
\par }{\BB \par }{\Q \VS{7}Don’t blame before you investigate.
\par }{\QB Understand first, and then rebuke.
\par }{\Q \VS{8}Don’t answer before you have heard.
\par }{\QB Don’t interrupt while someone else is speaking.
\par }{\Q \VS{9}Don’t argue about a matter that doesn’t concern you.
\par }{\QB Don’t sit with sinners when they judge.
\par }{\BB \par }{\Q \VS{10}My son, don’t be busy about many matters;
\par }{\QB for if you meddle much, you will not be unpunished.
\par }{\Q If you pursue, you will not overtake,
\par }{\QB and you will not escape by fleeing.
\par }{\Q \VS{11}There is one who toils, labors, and makes haste,
\par }{\QB and is even more behind.
\par }{\Q \VS{12}There is one who is sluggish, and needs help,
\par }{\QB lacking in strength, and who abounds in poverty;
\par }{\Q But the Lord’s eyes looked upon him for good,
\par }{\QB and he raised him up from his low condition,
\par }{\Q \VS{13}and lifted up his head
\par }{\QB so that many marveled at him.
\par }{\BB \par }{\Q \VS{14}Good things and evil, life and death,
\par }{\QB poverty and riches, are from the Lord.
\VS{15-16}\FTNT{†}{{\FR 11:15-16 }
Verses 15 and 16 are omitted by the best authorities.
}
\par }{\Q \VS{17}The Lord’s gift remains with the godly.
\par }{\QB His good pleasure will prosper forever.
\par }{\Q \VS{18}One grows rich by his diligence and self-denial,
\par }{\QB and this is the portion of his reward:
\par }{\Q \VS{19}when he says, “I have found rest,
\par }{\QB and now I will eat of my goods!”
\par }{\Q he doesn’t know how much time will pass
\par }{\QB until he leaves them to others and dies.
\par }{\Q \VS{20}Be steadfast in your covenant and be doing it,
\par }{\QB and grow old in your work.
\par }{\BB \par }{\Q \VS{21}Don’t marvel at the works of a sinner,
\par }{\QB but trust the Lord and stay in your labor;
\par }{\Q for it is an easy thing in the sight of the Lord
\par }{\QB to swiftly and suddenly make a poor man rich.
\par }{\Q \VS{22}The Lord’s blessing is in the reward of the godly.
\par }{\QB He makes his blessing flourish in an hour that comes swiftly.
\par }{\Q \VS{23}Don’t say, “What use is there of me?
\par }{\QB What further good things can be mine?”
\par }{\Q \VS{24}Don’t say, “I have enough.
\par }{\QB What harm could happen to me now?”
\par }{\Q \VS{25}In the day of good things, evil things are forgotten.
\par }{\QB In the day of evil things, a man will not remember things that are good.
\par }{\Q \VS{26}For it is an easy thing in the sight of the Lord
\par }{\QB to reward a man in the day of death according to his ways.
\par }{\Q \VS{27}The affliction of an hour causes delights to be forgotten.
\par }{\QB In the end, a man’s deeds are revealed.
\par }{\Q \VS{28}Call no man happy before his death.
\par }{\QB A man will be known in his children.
\par }{\BB \par }{\Q \VS{29}Don’t bring every man into your house,
\par }{\QB for many are the plots of a deceitful man.
\par }{\Q \VS{30}Like a decoy partridge in a cage, so is the heart of a proud man.
\par }{\QB Like a spy, he looks for your weakness.
\par }{\Q \VS{31}For he lies in wait to turn things that are good into evil,
\par }{\QB and assigns blame in things that are praiseworthy.
\par }{\Q \VS{32}From a spark of fire, a heap of many coals is kindled,
\par }{\QB and a sinful man lies in wait for blood.
\par }{\Q \VS{33}Take heed of an evil-doer, for he plans wicked things,
\par }{\QB lest perhaps he ruin your reputation forever.
\par }{\Q \VS{34}Receive a stranger into your house, and he will distract you with arguments
\par }{\QB and estrange you from your own.
\par }{\BB \par }
\PoetryChap{12}{\Q \VerseOne{1}If you do good, know to whom you do it,
\par }{\QB and your good deeds will have thanks.
\par }{\Q \VS{2}Do good to a godly man, and you will find a reward—
\par }{\QB if not from him, then from the Most High.
\par }{\Q \VS{3}No good will come to him who continues to do evil,
\par }{\QB nor to him who gives no alms.
\par }{\Q \VS{4}Give to the godly man,
\par }{\QB and don’t help the sinner.
\par }{\Q \VS{5}Do good to one who is lowly.
\par }{\QB Don’t give to an ungodly man.
\par }{\Q Keep back his bread, and don’t give it to him,
\par }{\QB lest he subdue you with it;
\par }{\Q for you would receive twice as much evil
\par }{\QB for all the good you would have done to him.
\par }{\Q \VS{6}For the Most High also hates sinners,
\par }{\QB and will repay vengeance to the ungodly.
\FTNT{*}{{\FR 12:6 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{7}Give to the good man,
\par }{\QB and don’t help the sinner.
\par }{\BB \par }{\Q \VS{8}A man’s friend won’t be
\FTNT{†}{{\FR 12:8 }
Or,
punished
} fully tried in prosperity.
\par }{\QB His enemy won’t be hidden in adversity.
\par }{\Q \VS{9}In a man’s prosperity, his enemies are grieved.
\par }{\QB In his adversity, even his friend leaves.
\par }{\Q \VS{10}Never trust your enemy,
\par }{\QB for his wickedness is like corrosion in copper.
\par }{\Q \VS{11}Though he humbles himself and walks bowed down,
\par }{\QB still be careful and beware of him.
\par }{\Q You will be to him as one who has wiped a mirror,
\par }{\QB to be sure it doesn’t completely tarnish.
\par }{\Q \VS{12}Don’t set him next to you,
\par }{\QB lest he overthrow you and stand in your place.
\par }{\Q Don’t let him sit on your right hand,
\par }{\QB lest he seek to take your seat,
\par }{\Q and at the last you acknowledge my words,
\par }{\QB and be pricked with my sayings.
\par }{\Q \VS{13}Who will pity a charmer that is bitten by a snake,
\par }{\QB or any who come near wild beasts?
\par }{\Q \VS{14}Even so, who will pity him who goes to a sinner,
\par }{\QB and is associated with him in his sins?
\par }{\Q \VS{15}For a while he will stay with you,
\par }{\QB and if you falter, he will not stay.
\par }{\Q \VS{16}The enemy will speak sweetly with his lips,
\par }{\QB and in his heart plan to throw you into a pit.
\par }{\Q The enemy may weep with his eyes,
\par }{\QB but if he finds opportunity, he will want more blood.
\par }{\Q \VS{17}If adversity meets you, you will find him there before you.
\par }{\QB Pretending to help you, he will trip you.
\par }{\Q \VS{18}He will shake his head, clap his hands,
\par }{\QB whisper much, and change his countenance.
\par }{\BB \par }
\PoetryChap{13}{\Q \VerseOne{1}He who touches pitch will be defiled.
\par }{\QB He who has fellowship with a proud man will become like to him.
\par }{\Q \VS{2}Don’t take up a burden above your strength.
\par }{\QB Have no fellowship with one that is mightier and richer than yourself.
\par }{\Q What fellowship would the earthen pot have with the kettle?
\par }{\QB The kettle will strike, and the pot will be dashed in pieces.
\par }{\Q \VS{3}The rich man does a wrong and threatens.
\par }{\QB The poor is wronged and apologizes.
\par }{\Q \VS{4}If you are profitable, he will make merchandise of you.
\par }{\QB If you are in lack, he will forsake you.
\par }{\Q \VS{5}If you own something, he will live with you.
\par }{\QB He will drain you bare and will not be sorry.
\par }{\Q \VS{6}Does he need you? Then he will deceive you,
\par }{\QB smile at you, and give you hope.
\par }{\QB He will speak kindly to you and say, “What do you need?”
\par }{\Q \VS{7}He will shame you by his meats
\par }{\QB until he has made you bare twice or thrice,
\par }{\QB and in the end he will laugh you to scorn.
\par }{\Q Afterward he will see you, will forsake you,
\par }{\QB and shake his head at you.
\par }{\Q \VS{8}Beware that you are not deceived
\par }{\QB and brought low in your mirth.
\par }{\Q \VS{9}If a mighty man invites you, be reserved,
\par }{\QB and he will invite you more.
\par }{\Q \VS{10}Don’t press him, lest you be thrust back.
\par }{\QB Don’t stand far off, lest you be forgotten.
\par }{\Q \VS{11}Don’t try to speak with him as an equal,
\par }{\QB and don’t believe his many words;
\par }{\Q For he will test you with much talk,
\par }{\QB and will examine you in a smiling manner.
\par }{\Q \VS{12}He who doesn’t keep secrets to himself is unmerciful.
\par }{\QB He won’t hesitate to harm and to bind.
\par }{\Q \VS{13}Keep them to yourself and be careful,
\par }{\QB for you walk
\FTNT{*}{{\FR 13:13 }
Gr.
along with.
} in danger of falling.
\VS{14}\FTNT{†}{{\FR 13:14 }
The remainder of verse 13, and verse 14, are omitted by the best authorities.
}
\par }{\Q \VS{15}Every living creature loves his own kind,
\par }{\QB and every man loves his neighbor.
\par }{\Q \VS{16}All flesh associates with their own kind.
\par }{\QB A man will stick to people like himself.
\par }{\Q \VS{17}What fellowship would the wolf have with the lamb?
\par }{\QB So is the sinner to the godly.
\par }{\Q \VS{18}What peace is there between a hyena and a dog?
\par }{\QB What peace is there between a rich man and the poor?
\par }{\Q \VS{19}Wild donkeys are the prey of lions in the wilderness;
\par }{\QB likewise poor men are feeding grounds for the rich.
\par }{\Q \VS{20}Lowliness is an abomination to a proud man;
\par }{\QB likewise a poor man is an abomination to the rich.
\par }{\BB \par }{\Q \VS{21}When a rich man is shaken, he is supported by his friends,
\par }{\QB but when one of low degree is down, he is pushed away even by his friends.
\par }{\Q \VS{22}When a rich man falls, there are many helpers.
\par }{\QB He speaks things not to be spoken, and men justify him.
\par }{\Q A man of low degree falls, and men rebuke him.
\par }{\QB He utters wisdom, and is not listened to.
\par }{\Q \VS{23}A rich man speaks, and all keep silence.
\par }{\QB They extol what he says extol to the clouds.
\par }{\Q A poor man speaks, and they say, “Who is this?”
\par }{\QB If he stumbles, they will help to overthrow him.
\par }{\BB \par }{\Q \VS{24}Riches are good if they have no sin.
\par }{\QB Poverty is evil in the mouth of the ungodly.
\par }{\Q \VS{25}The heart of a man changes his countenance,
\par }{\QB whether it is for good or for evil.
\FTNT{‡}{{\FR 13:25 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{26}A cheerful countenance is a sign of a prosperous heart.
\par }{\QB Devising proverbs takes strenuous thinking.

\par }\PoetryChap{14}{\Q \VerseOne{1}Blessed is the man who has not slipped with his mouth,
\par }{\QB and doesn’t suffer from sorrow for sins.
\par }{\Q \VS{2}Blessed is he whose soul does not condemn him,
\par }{\QB and who has not given up hope.
\par }{\BB \par }{\Q \VS{3}Riches are not appropriate for a stingy person.
\par }{\QB What would an envious man do with money?
\par }{\Q \VS{4}He who gathers by denying himself gathers for others.
\par }{\QB Others will revel in his goods.
\par }{\Q \VS{5}If one is mean to himself, to whom will he be good?
\par }{\QB He won’t enjoy his possessions.
\par }{\Q \VS{6}There is none more evil than he who is stingy with himself.
\par }{\QB This is a punishment for his wickedness.
\par }{\Q \VS{7}Even if he does good, he does it in forgetfulness.
\par }{\QB In the end, he reveals his wickedness.
\par }{\Q \VS{8}A miser is evil.
\par }{\QB He turns away and disregards souls.
\par }{\Q \VS{9}A covetous man’s eye is not satisfied with his portion.
\par }{\QB Wicked injustice dries up his soul.
\par }{\Q \VS{10}A miser begrudges bread,
\par }{\QB and it is lacking at his table.
\par }{\BB \par }{\Q \VS{11}My son, according to what you have, treat yourself well,
\par }{\QB and bring worthy offerings to the Lord.
\par }{\Q \VS{12}Remember that death will not wait,
\par }{\QB and that the covenant of Hades hasn’t been shown to you.
\par }{\Q \VS{13}Do good to your friend before you die.
\par }{\QB According to your ability, reach out and give to him.
\par }{\Q \VS{14}Don’t defraud yourself of a good day.
\par }{\QB Don’t let the portion of a good desire pass you by.
\par }{\Q \VS{15}Won’t you leave your labors to another,
\par }{\QB and your toils be divided by lot?
\par }{\Q \VS{16}Give, take, and treat yourself well,
\par }{\QB because there is no seeking of luxury in Hades.
\par }{\Q \VS{17}All flesh grows old like a garment,
\par }{\QB for the covenant from the beginning is, “You must die.”
\par }{\Q \VS{18}Like the leaves flourishing on a thick tree,
\par }{\QB some it sheds, and some grow,
\par }{\Q so also are the generations of flesh and blood:
\par }{\QB one comes to an end and another is born.
\par }{\Q \VS{19}Every work rots and falls away,
\par }{\QB and its builder will depart with it.
\par }{\BB \par }{\Q \VS{20}Blessed is the man who meditates on wisdom,
\par }{\QB and who reasons by his understanding.
\par }{\Q \VS{21}He who considers her ways in his heart
\par }{\QB will also have knowledge of her secrets.
\par }{\Q \VS{22}Go after her like one who tracks,
\par }{\QB and lie in wait in her ways.
\par }{\Q \VS{23}He who pries in at her windows
\par }{\QB will also listen at her doors.
\par }{\Q \VS{24}He who lodges close to her house
\par }{\QB will also fasten a nail in her walls.
\par }{\Q \VS{25}He will pitch his tent near at hand to her,
\par }{\QB and will lodge in a lodging where good things are.
\par }{\Q \VS{26}He will set his children under her shelter,
\par }{\QB and will rest under her branches.
\par }{\Q \VS{27}By her he will be covered from heat,
\par }{\QB and will lodge in her glory.

\par }\PoetryChap{15}{\Q \VerseOne{1}He who fears the Lord will do this.
\par }{\QB He who has possession of the law will obtain her.
\par }{\Q \VS{2}As a mother shall she meet him,
\par }{\QB and receive him as a wife married in her virginity.
\par }{\Q \VS{3}She will feed him with bread of understanding
\par }{\QB and give him water of wisdom to drink.
\par }{\Q \VS{4}He will be stayed upon her, and will not be moved.
\par }{\QB He will rely upon her, and will not be confounded.
\par }{\Q \VS{5}She will exalt him above his neighbors.
\par }{\QB She will open his mouth in the midst of the congregation.
\par }{\Q \VS{6}He will inherit joy, a crown of gladness,
\par }{\QB and an everlasting name.
\par }{\Q \VS{7}Foolish men will not obtain her.
\par }{\QB Sinners will not see her.
\par }{\Q \VS{8}She is far from pride.
\par }{\QB Liars will not remember her.
\par }{\Q \VS{9}Praise is not attractive in the mouth of a sinner;
\par }{\QB for it was not sent to him from the Lord.
\par }{\Q \VS{10}For praise will be spoken in wisdom;
\par }{\QB The Lord will prosper it.
\par }{\BB \par }{\Q \VS{11}Don’t say, “It is through the Lord that I fell away;”
\par }{\QB for you shall not do the things that he hates.
\par }{\Q \VS{12}Don’t say, “It is he that caused me to err;”
\par }{\QB for he has no need of a sinful man.
\par }{\Q \VS{13}The Lord hates every abomination;
\par }{\QB and those who fear him don’t love them.
\par }{\Q \VS{14}He himself made man from the beginning
\par }{\QB and left him in the hand of his own counsel.
\par }{\Q \VS{15}If you will, you shall keep the commandments.
\par }{\QB To be faithful is good pleasure.
\par }{\Q \VS{16}He has set fire and water before you.
\par }{\QB You will stretch forth your hand to whichever you desire.
\par }{\Q \VS{17}Before man is life and death.
\par }{\QB Whichever he likes, it will be given to him.
\par }{\Q \VS{18}For great is the wisdom of the Lord.
\par }{\QB He is mighty in power, and sees all things.
\par }{\Q \VS{19}His eyes are upon those who fear him.
\par }{\QB He knows every work of man.
\par }{\Q \VS{20}He has not commanded any man to be ungodly.
\par }{\QB He has not given any man license to sin.
\par }{\BB \par }
\PoetryChap{16}{\Q \VerseOne{1}Don’t desire a multitude of unprofitable children,
\par }{\QB neither delight in ungodly sons.
\par }{\Q \VS{2}If they multiply, delight not in them,
\par }{\Q Except the fear of the Lord be with them.
\par }{\Q \VS{3}Trust not you in their life,
\par }{\Q Neither rely on their condition:
\par }{\Q For one is better than a thousand;
\par }{\Q And to die childless than to have ungodly children.
\par }{\Q \VS{4}For from one that has understanding will a city be populated;
\par }{\Q But a race of wicked men will be made desolate.
\par }{\Q \VS{5}Many such things I have seen with my eyes;
\par }{\Q And my ear has heard mightier things than these.
\par }{\BB \par }{\Q \VS{6}In the congregation of sinners will a fire be kindled;
\par }{\Q And in a disobedient nation wrath is kindled.
\par }{\Q \VS{7}He was not pacified toward the giants of old time,
\par }{\Q Who revolted in their strength.
\par }{\Q \VS{8}He spared not those with whom Lot sojourned,
\par }{\Q Whom he abhorred for their pride.
\par }{\Q \VS{9}He pitied not the people of perdition,
\par }{\Q Who were taken away in their sins.
\par }{\Q \VS{10}And in like manner the six hundred thousand footmen,
\par }{\Q Who were gathered together in the hardness of their hearts.
\par }{\Q \VS{11}Even if there is one stiff-necked person,
\par }{\Q It is a marvel if he will be unpunished:
\par }{\Q For mercy and wrath are with him;
\par }{\Q He is mighty to forgive, and he pours out wrath.
\par }{\Q \VS{12}As his mercy is great, so is his correction also:
\par }{\Q He judges a man according to his works.
\par }{\Q \VS{13}The sinner will not escape with
{\ADD{his
}} plunder;
\par }{\Q And the perseverance of the godly will not be frustrated.
\par }{\Q \VS{14}He will make room for every work of mercy;
\par }{\Q Each man will find according to his works.
\VS{15-16}\FTNT{*}{{\FR 16:15-16 }
Verses 15 and 16 are omitted by the best authorities.
}
\par }{\BB \par }{\Q \VS{17}Don’t say, I shall be hidden from the Lord;
\par }{\Q And who will remember me from on high?
\par }{\Q I will not be known among so many people;
\par }{\Q For what is my soul in a boundless creation?
\par }{\Q \VS{18}Behold, the heaven, and the heaven of heavens,
\par }{\Q The deep, and the earth, will be moved when he shall visit.
\par }{\Q \VS{19}The mountains and the foundations of the earth together
\par }{\Q Are shaken with trembling, when he looks upon them.
\par }{\Q \VS{20}And no heart will think upon these things:
\par }{\Q And who shall conceive his ways?
\par }{\Q \VS{21}And
{\ADD{there is
}} a tempest which no man will see;
\par }{\Q Yes, the more part of his works are
\FTNT{†}{{\FR 16:21 }
Gr.
among hidden things.
} hid.
\par }{\Q \VS{22}Who will declare the works of
{\ADD{his
}} righteousness?
\par }{\Q Or who will endure them?
\par }{\Q For his covenant is afar off.
\FTNT{‡}{{\FR 16:22 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{23}He that is lacking in
\FTNT{§}{{\FR 16:23 }
Gr.
heart.
} understanding thinks upon these things;
\par }{\Q And an unwise and erring man thinks follies.
\par }{\Q \VS{24}My son, listen to me, and learn knowledge,
\par }{\Q And give heed to my words with your heart.
\par }{\Q \VS{25}I will show forth instruction by weight,
\par }{\Q And declare knowledge exactly.
\par }{\Q \VS{26}In the judgement of the Lord are his works from the beginning;
\par }{\Q And from the making of them he disposed the parts thereof.
\par }{\Q \VS{27}He garnished his works forever,
\par }{\Q And the beginnings of them to their generations:
\par }{\Q They neither hunger, nor are weary,
\par }{\Q And they cease not from their works.
\par }{\Q \VS{28}No one thrusts aside his neighbor;
\par }{\Q And they will never disobey his word.
\par }{\Q \VS{29}After this also the Lord looked upon the earth,
\par }{\Q And filled it with his blessings.
\par }{\Q \VS{30}\FTNT{*}{{\FR 16:30 }
The Greek text of this line is uncertain.
}
\FTNT{†}{{\FR 16:30 }
Gr.
The soul of every living thing.
} All manner of living things covered the face thereof;
\par }{\Q And into it is their return.
\par }{\BB \par }
\PoetryChap{17}{\Q \VerseOne{1}The Lord created man of the earth,
\par }{\Q And turned him back to it again.
\par }{\Q \VS{2}He gave them days by number, and a set time,
\par }{\Q And gave them authority over the things that are on it.
\par }{\Q \VS{3}He endowed them with strength proper to them;
\par }{\Q And made them according to his own image.
\par }{\Q \VS{4}He put the fear of
\FTNT{*}{{\FR 17:4 }
Gr.
him.
} man upon all flesh,
\par }{\Q And
{\ADD{gave him
}} to have dominion over beasts and fowls.
\VS{5}\FTNT{†}{{\FR 17:5 }
Verse 5 is omitted by the best authorities.
}
\par }{\Q \VS{6}Counsel, and tongue, and eyes,
\par }{\Q Ears, and heart, gave he them to understand withal.
\par }{\Q \VS{7}He filled them with the knowledge of wisdom,
\par }{\Q And showed them good and evil.
\par }{\Q \VS{8}He set his eye upon their hearts,
\par }{\Q To show them the majesty of his works.
\VS{9}\FTNT{‡}{{\FR 17:9 }
Verse 9 is omitted by the best authorities.
}
\par }{\Q \VS{10}And they will praise the name of
{\ADD{his
}} holiness,
\par }{\Q \FTNT{§}{{\FR 17:10 }
This line is added by the best authorities.
} That they may declare the majesty of his works.
\par }{\Q \VS{11}He added to them knowledge,
\par }{\Q And gave them a law of life for a heritage.
\par }{\Q \VS{12}He made an everlasting covenant with them,
\par }{\Q And showed them his judgments.
\par }{\Q \VS{13}Their eyes saw the majesty of
{\ADD{his
}} glory;
\par }{\Q And their ear heard the glory of
\FTNT{*}{{\FR 17:13 }
Some ancient authorities read
their.
} his voice.
\par }{\Q \VS{14}And he said to them, Beware of all unrighteousness;
\par }{\Q And he gave them commandment, each man concerning his neighbor.
\par }{\BB \par }{\Q \VS{15}Their ways are ever before him;
\par }{\Q They will not be hid from his eyes.
\VS{16}\FTNT{†}{{\FR 17:16 }
Verses 16, 18, and 21 are omitted by the best authorities.
}
\par }{\Q \VS{17}\FTNT{‡}{{\FR 17:17 }
The preceding part of this verse is omitted by the best authorities.
} For every nation he appointed a ruler;
\par }{\Q And Israel is the Lord’s portion.
\VS{18}\FTNT{§}{{\FR 17:18 }
Verses 16, 18, and 21 are omitted by the best authorities.
}
\par }{\Q \VS{19}All their works are as the sun before him;
\par }{\Q And his eyes are continually upon their ways.
\par }{\Q \VS{20}Their iniquities are not hid from him;
\par }{\Q And all their sins are before the Lord.
\VS{21}\FTNT{*}{{\FR 17:21 }
Verses 16, 18, and 21 are omitted by the best authorities.
}
\par }{\Q \VS{22}With him the alms of a man is as a signet;
\par }{\Q And he will keep the bounty of a man as the apple of the eye.
\FTNT{†}{{\FR 17:22 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{23}Afterwards he will rise up and recompense them,
\par }{\Q And render their recompense upon their head.
\par }{\Q \VS{24}However to those who repent he grants a return;
\par }{\Q And he comforts those who are losing patience.
\par }{\BB \par }{\Q \VS{25}Return to the Lord, and forsake sins:
\par }{\Q Make your prayer before
{\ADD{his
}} face, and lessen the offence.
\par }{\Q \VS{26}Turn again to the Most High, and turn away from iniquity;
\FTNT{‡}{{\FR 17:26 }
A line is here omitted by the best authorities.
}
\par }{\Q And greatly hate the abominable thing.
\par }{\Q \VS{27}Who will give praise to the Most High in
\FTNT{§}{{\FR 17:27 }
Gr.
Hades.
} the grave,
\par }{\Q Instead of them which live and return thanks?
\par }{\Q \VS{28}Thanksgiving perishes from the dead, as from one that is not:
\par }{\Q He that is in life and health will praise the Lord.
\par }{\Q \VS{29}How great is the mercy of the Lord,
\par }{\Q And his forgiveness to those who turn to him!
\par }{\Q \VS{30}For all things can’t be in men,
\par }{\Q Because the son of man is not immortal.
\par }{\Q \VS{31}What is brighter than the sun? yet this fails:
\par }{\Q And an evil man will think on flesh and blood.
\par }{\Q \VS{32}He looks upon the power of the height of heaven:
\par }{\Q And all men are earth and ashes.
\par }{\BB \par }
\PoetryChap{18}{\Q \VerseOne{1}He that lives forever created all things in common.
\par }{\Q \VS{2}The Lord alone will be justified.
\VS{3}\FTNT{*}{{\FR 18:3 }
The remainder of verse 2, and verse 3, are omitted by the best authorities.
}
\par }{\Q \VS{4}To none has he given power to declare his works:
\par }{\Q And who will trace out his mighty deeds?
\par }{\Q \VS{5}Who shall number the strength of his majesty?
\par }{\Q And who shall also tell out his mercies?
\par }{\Q \VS{6}As for the wondrous works of the Lord, it is not possible to take from them nor add to them,
\par }{\Q Neither is it possible to track them out.
\par }{\Q \VS{7}When a man has finished, then he is but at the beginning;
\par }{\Q And when he ceases, then he will be in perplexity.
\par }{\Q \VS{8}What is man, and what purpose do they serve?
\par }{\Q What is his good, and what is his evil?
\par }{\Q \VS{9}The number of man’s days at the most are a hundred years.
\par }{\Q \VS{10}As a drop of water from the sea, and a pebble from the sand;
\par }{\Q So are a few years in the day of eternity.
\par }{\Q \VS{11}For this cause the Lord was patient over them,
\par }{\Q And poured out his mercy upon them.
\par }{\Q \VS{12}He saw and perceived their end, that it is evil;
\par }{\Q Therefore he multiplied his forgiveness.
\par }{\Q \VS{13}The mercy of a man is upon his neighbor;
\par }{\Q But the mercy of the Lord is upon all flesh;
\par }{\Q Reproving, and chastening, and teaching,
\par }{\Q And bringing again, as a shepherd does his flock.
\par }{\Q \VS{14}He has mercy on those who accept chastening,
\par }{\Q And that diligently seek after his judgments.
\par }{\BB \par }{\Q \VS{15}My son, to your good deeds add no blemish;
\par }{\Q And no grief of words in any of your giving.
\par }{\Q \VS{16}Shall not the dew assuage the scorching heat?
\par }{\Q So is a word better than a gift.
\par }{\Q \VS{17}Behold, is not a word better than a gift?
\par }{\Q And both are with a gracious man.
\par }{\Q \VS{18}A fool will upbrade ungraciously;
\par }{\Q And the gift of an envious man consumes the eyes.
\par }{\BB \par }{\Q \VS{19}Learn before you speak;
\par }{\Q And take care of your health before you are sick.
\par }{\Q \VS{20}Before judgement examine yourself;
\par }{\Q And in the hour of visitation you will find forgiveness.
\par }{\Q \VS{21}Humble yourself before you become sick;
\par }{\Q And in the time of sins show repentance.
\par }{\Q \VS{22}Let nothing hinder you to pay your vow in due time;
\par }{\Q And don’t wait until death to be justified.
\par }{\Q \VS{23}Before you make a vow, prepare yourself;
\par }{\Q And be not as a man that tempts the Lord.
\par }{\Q \VS{24}Think upon the wrath
{\ADD{that will be
}} in the days of the end,
\par }{\Q And the time of vengeance, when he turns away his face.
\par }{\Q \VS{25}In the days of fullness remember the time of hunger,
\par }{\Q {\ADD{And
}} poverty and lack in the days of wealth.
\par }{\Q \VS{26}From morning until evening the time changes;
\par }{\Q And all things are speedy before the Lord.
\par }{\Q \VS{27}A wise man will fear in everything;
\par }{\Q And in days of sinning he will beware of offence.
\FTNT{†}{{\FR 18:27 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{28}Every man of understanding knows wisdom;
\par }{\Q And he will give thanks to him that found her.
\par }{\Q \VS{29}They that were of understanding in sayings became also wise themselves,
\par }{\Q And poured forth apt proverbs.
\par }{\BB \par }{\Q \VS{30}Go not after your lusts;
\par }{\Q And refrain yourself from your appetites.
\par }{\Q \VS{31}If you give fully to your soul the delight of her desire,
\par }{\Q She will make you
\FTNT{‡}{{\FR 18:31 }
Or,
a rejoicing to
} the laughing stock of your enemies.
\par }{\Q \VS{32}Make not merry in much luxury;
\par }{\Q Neither be tied to the expense thereof.
\par }{\Q \VS{33}Be not made a beggar by banqueting upon borrowing,
\par }{\Q When you have nothing in your purse.
\FTNT{§}{{\FR 18:33 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\BB \par }
\PoetryChap{19}{\Q \VerseOne{1}A workman that is a drunkard will not become rich:
\par }{\Q He that despises small things will fall by little and little.
\par }{\Q \VS{2}Wine and women will make men of understanding to fall away:
\par }{\Q And he that joins with prostitutes will be the more reckless.
\par }{\Q \VS{3}Moths and worms will have him to heritage;
\par }{\Q And a reckless soul will be taken away.
\par }{\BB \par }{\Q \VS{4}He that is hasty to trust is light-minded;
\par }{\Q And he that sins will offend against his own soul.
\par }{\Q \VS{5}He that makes merry in his heart will be condemned:
\FTNT{*}{{\FR 19:5 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{6}\FTNT{†}{{\FR 19:6 }
The preceding part of this verse is omitted by the best authorities.
} And he that hates talk has the less wickedness.
\par }{\Q \VS{7}Never repeat what is told you,
\par }{\Q And you will fare never the worse.
\par }{\Q \VS{8}Whether it be of friend or foe, tell it not;
\par }{\Q And unless it is a sin to you, reveal it not.
\par }{\Q \VS{9}For he has heard you, and observed you,
\par }{\Q And when the time comes he will hate you.
\par }{\Q \VS{10}Have you heard a word? let it die with you:
\par }{\Q Be of good courage, it will not burst you.
\par }{\Q \VS{11}A fool will travail in pain with a word,
\par }{\Q As a woman in labor with a child.
\par }{\Q \VS{12}As an arrow that sticks in the
\FTNT{‡}{{\FR 19:12 }
Gr.
thigh of flesh.
} flesh of the thigh,
\par }{\Q So is a word in a fool’s belly.
\par }{\BB \par }{\Q \VS{13}Reprove a friend; it may be he did it not:
\par }{\Q And if he did something, that he may do it no more.
\par }{\Q \VS{14}Reprove your neighbor; it may be he said it not:
\par }{\Q And if he has said it, that he may not say it again.
\par }{\Q \VS{15}Reprove a friend; for many times there is slander:
\par }{\Q And trust not every word.
\par }{\Q \VS{16}There is one that slips, and not from the heart:
\par }{\Q And who is he that has not sinned with his tongue?
\par }{\Q \VS{17}Reprove your neighbor before you threaten him;
\par }{\Q And give place to the law of the Most High.
\VS{18-19}\FTNT{§}{{\FR 19:18-19 }
Verses 18 and 19 are omitted by the best authorities.
}
\par }{\BB \par }{\Q \VS{20}All wisdom is the fear of the Lord;
\par }{\Q And in all wisdom is the doing of the law.
\VS{21}\FTNT{*}{{\FR 19:21 }
The remainder of verse 20, and verse 21, are omitted by the best authorities.
}
\par }{\Q \VS{22}And the knowledge of wickedness is not wisdom;
\par }{\Q And the prudence of sinners is not counsel.
\par }{\Q \VS{23}There is a wickedness, and it is abomination;
\par }{\Q And there is a fool lacking in wisdom.
\par }{\Q \VS{24}Better is one that has small understanding, and fears,
\par }{\Q Than one that has much prudence, and transgresses the law.
\par }{\Q \VS{25}There is an exquisite subtlety, and it is unjust;
\par }{\Q And there is one that perverts favor
\FTNT{†}{{\FR 19:25 }
Gr.
to bring to light.
} to gain a judgement.
\FTNT{‡}{{\FR 19:25 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{26}There is one that does wickedly, that hangs down his head with mourning;
\par }{\Q But inwardly he is full of deceit,
\par }{\Q \VS{27}Bowing down his face, and making as if he were deaf of one ear:
\par }{\Q Where he is not known, he will be beforehand with you.
\par }{\Q \VS{28}And if for lack of power he be hindered from sinning,
\par }{\Q If he find opportunity, he will do mischief.
\par }{\Q \VS{29}A man will be known by his look,
\par }{\Q And one that has understanding will be known by his face, when you meet him.
\par }{\Q \VS{30}A man’s attire, and
\FTNT{§}{{\FR 19:30 }
Gr.
laughter of the teeth.
} grinning laughter,
\par }{\Q And gait, show what he is.
\par }{\BB \par }
\PoetryChap{20}{\Q \VerseOne{1}There is a reproof that is not comely;
\par }{\Q And there is a man that keeps silence, and he is wise.
\par }{\Q \VS{2}How good is it to reprove, rather than to be angry;
\par }{\Q And he that makes confession will be kept back from hurt.
\VS{3}\FTNT{*}{{\FR 20:3 }
Verse 3 is omitted by the best authorities.
}
\par }{\Q \VS{4}As is the lust of an eunuch to deflower a virgin;
\par }{\Q So is he that executes judgments with violence.
\par }{\Q \VS{5}There is one that keeps silence, and is found wise;
\par }{\Q And there is one that is hated for his much talk.
\par }{\Q \VS{6}There is one that keeps silence, for he has no answer to make;
\par }{\Q And there is that keeps silence, as knowing his time.
\par }{\Q \VS{7}A wise man will be silent till his time come;
\par }{\Q But the braggart and fool will overpass his time.
\par }{\Q \VS{8}He that uses many words will be abhorred;
\par }{\Q And he that takes to himself authority therein will be hated.
\par }{\BB \par }{\Q \VS{9}There is a prosperity that a man finds in misfortunes;
\par }{\Q And there is a gain that turns to loss.
\par }{\Q \VS{10}There is a gift that will not profit you;
\par }{\Q And there is a gift whose recompense is double.
\par }{\Q \VS{11}There is an abasement because of glory;
\par }{\Q And there is that has lifted up his head from a low estate.
\par }{\Q \VS{12}There is that buys much for a little,
\par }{\Q And pays for it again sevenfold.
\par }{\Q \VS{13}He that is wise in words will make himself beloved;
\par }{\Q But the pleasantries of fools will be wasted.
\par }{\Q \VS{14}The gift of a fool will not profit you;\FTNT{†}{{\FR 20:14 }
A line of this verse is here omitted by the best authorities.
}
\par }{\Q For his eyes are many instead of one.
\par }{\Q \VS{15}He will give little, and upbraid much;
\par }{\Q And he will open his mouth like a crier:
\par }{\Q Today he will lend, and tomorrow he will ask it again:
\par }{\Q Such an one is a hateful man.
\par }{\Q \VS{16}The fool will say, I have no friend,
\par }{\Q And I have no thanks for my good deeds;
\par }{\Q They that eat my bread are of evil tongue.
\par }{\Q \VS{17}How often, and of how many, will he be laughed to scorn!
\FTNT{‡}{{\FR 20:17 }
The latter part of verse 17 is omitted by the best authorities.
}
\par }{\BB \par }{\Q \VS{18}A slip on a pavement is better than
{\ADD{a slip
}} with the tongue;
\par }{\Q So the fall of the wicked will come speedily.
\par }{\Q \VS{19}A man without grace is
{\ADD{as
}} a tale out of season:
\par }{\Q It will be continually in the mouth of the ignorant.
\par }{\Q \VS{20}A
\FTNT{§}{{\FR 20:20 }
Gr.
parable.
} wise sentence from a fool’s mouth will be rejected;
\par }{\Q For he will not speak it in its season.
\par }{\BB \par }{\Q \VS{21}There is that is hindered from sinning through lack;
\par }{\Q And when he takes rest, he will not be troubled.
\par }{\Q \VS{22}There is that destroys his soul through bashfulness;
\par }{\Q And by a foolish countenance he will destroy it.
\par }{\Q \VS{23}There is that for bashfulness promises to his friend;
\par }{\Q And he makes him his enemy for nothing.
\par }{\BB \par }{\Q \VS{24}A lie is a foul blot in a man:
\par }{\Q It will be continually in the mouth of the ignorant.
\par }{\Q \VS{25}A thief is better than a man that is continually lying:
\par }{\Q But they both will inherit destruction.
\par }{\Q \VS{26}The disposition of a liar is dishonor;
\par }{\Q And his shame is with him continually.
\par }{\BB \par }{\Q \VS{27}He that is wise in words will advance himself;
\par }{\Q And one that is prudent will please great men.
\par }{\Q \VS{28}He that tills his land will raise his heap high;
\par }{\Q And he that pleases great men will get pardon for iniquity.
\par }{\Q \VS{29}Presents and gifts blind the eyes of the wise,
\par }{\Q And as a muzzle on the mouth, turn away reproofs.
\par }{\Q \VS{30}Wisdom that is hid, and treasure that is out of sight,
\par }{\Q What profit is in them both?
\par }{\Q \VS{31}Better is a man that hides his folly
\par }{\Q Than a man that hides his wisdom.
\VS{32}\FTNT{*}{{\FR 20:32 }
Verse 32 is omitted by the best authorities.}
\par }{\BB \par }
\PoetryChap{21}{\Q \VerseOne{1}My son, have you sinned? add no more thereto;
\par }{\Q And make supplication for your former
{\ADD{sins.
}}
\par }{\Q \VS{2}Flee from sin as from the face of a serpent;
\par }{\Q For if you draw near it will bite you:
\par }{\Q The teeth thereof are the teeth of a lion,
\par }{\Q Slaying the souls of men.
\par }{\Q \VS{3}All iniquity is as a two-edged sword;
\par }{\Q Its stroke has no healing.
\par }{\BB \par }{\Q \VS{4}Terror and violence will lay waste riches;
\par }{\Q So the house of an arrogant man will be laid waste.
\par }{\Q \VS{5}Supplication from a poor man’s mouth reaches to the ears of
\FTNT{*}{{\FR 21:5 }
Gr.
him.
} God,
\par }{\Q And his judgement comes speedily.
\par }{\Q \VS{6}One that hates reproof is in the path of the sinner;
\par }{\Q And he that fears the Lord will turn again in his heart.
\par }{\Q \VS{7}He that is mighty in tongue is known afar off:
\par }{\Q But the man of understanding knows when he slips.
\par }{\BB \par }{\Q \VS{8}He that builds his house with other men’s money
\par }{\Q Is like one that gathers himself stones against winter.
\par }{\Q \VS{9}The congregation of wicked men is
{\ADD{as
}} tow wrapped together;
\par }{\Q And the end of them is a flame of fire.
\par }{\Q \VS{10}The way of sinners is made smooth with stones;
\par }{\Q And at the last end thereof is the pit of Hades.
\par }{\BB \par }{\Q \VS{11}He that keeps the law becomes master of the intent thereof;
\par }{\Q And the end of the fear of the Lord is wisdom.
\par }{\Q \VS{12}He that is not clever will not be instructed;
\par }{\Q And there is a cleverness which makes bitterness to abound.
\par }{\Q \VS{13}The knowledge of a wise man will be made to abound as a flood;
\par }{\Q And his counsel as a fountain of life.
\par }{\Q \VS{14}The inward parts of a fool are like a broken vessel;
\par }{\Q And he will hold no knowledge.
\par }{\BB \par }{\Q \VS{15}If a man of knowledge hear a wise word,
\par }{\Q He will commend it, and add to it:
\par }{\Q The dissolute man hears it, and it displeases him,
\par }{\Q And he puts it away behind his back.
\par }{\Q \VS{16}The discourse of a fool is like a burden in the way;
\par }{\Q But grace will be found on the lips of the wise.
\par }{\Q \VS{17}The mouth of the prudent man will be sought for in the congregation;
\par }{\Q And they will ponder his words in their heart.
\par }{\BB \par }{\Q \VS{18}As a house that is destroyed, so is wisdom to a fool;
\par }{\Q And the knowledge of an unwise man is
{\ADD{as
}}
\FTNT{†}{{\FR 21:18 }
Gr.
unexamined words.
} talk without sense.
\par }{\Q \VS{19}Instruction is
{\ADD{as
}} fetters on the feet of an unwise man,
\par }{\Q And as manacles on the right hand.
\par }{\Q \VS{20}A fool lifts up his voice with laughter;
\par }{\Q But a clever man will scarce smile quietly.
\par }{\Q \VS{21}Instruction is to a prudent man as an ornament of gold,
\par }{\Q And as a bracelet upon his right arm.
\par }{\BB \par }{\Q \VS{22}The foot of a fool is soon in
{\ADD{another man’s
}} house;
\par }{\Q But a man of experience will be ashamed of entering.
\par }{\Q \VS{23}A foolish man peeps in from the door of
{\ADD{another man’s
}} house;
\par }{\Q But a man that is instructed will stand without.
\par }{\Q \VS{24}It is a lack of instruction in a man to listen at the door;
\par }{\Q But the prudent man will be grieved with the disgrace.
\par }{\Q \VS{25}\FTNT{‡}{{\FR 21:25 }
The text seems to be uncertain.
} The lips of strangers will be grieved at these things;
\par }{\Q But the words of prudent men will be weighed in the balance.
\par }{\BB \par }{\Q \VS{26}The heart of fools is in their mouth;
\par }{\Q But the mouth of wise men is their heart.
\par }{\Q \VS{27}When the ungodly curses Satan,
\par }{\Q He curses his own soul.
\par }{\Q \VS{28}A whisperer defiles his own soul,
\par }{\Q And will be hated wherever he travels.
\par }{\BB \par }
\PoetryChap{22}{\Q \VerseOne{1}A slothful man is compared to a stone that is defiled;
\par }{\Q And every one will hiss him out in his disgrace.
\par }{\Q \VS{2}A slothful man is compared to the filth of a dunghill:
\par }{\Q Every man that takes it up will shake out his hand.
\par }{\BB \par }{\Q \VS{3}A father has shame in having begotten an uninstructed
{\ADD{son;
}}
\par }{\Q And a
{\ADD{foolish
}} daughter is born to his loss.
\par }{\Q \VS{4}A prudent daughter will inherit a husband of her own;
\par }{\Q And she that brings shame is the grief of her father.
\par }{\Q \VS{5}She that is bold brings shame upon father and husband;
\par }{\Q And she will be despised of them both.
\par }{\Q \VS{6}Unseasonable discourse is
{\ADD{as
}} music in mourning;
\par }{\Q But stripes and correction are wisdom at every season.
\par }{\BB \par }{\Q \VS{7}He that teaches a fool is
{\ADD{as
}} one that glues a potsherd together;
\par }{\Q {\ADD{Even as
}} one that wakes a sleeper out of a deep sleep.
\par }{\Q \VS{8}He who teaches a fool is
{\ADD{as
}} one who teaches a man who slumbers;
\par }{\Q And at the end he will say, What is it?
\VS{9-10}\FTNT{*}{{\FR 22:9-10 }
Verses 9 and 10 are omitted by the best authorities.
}
\par }{\Q \VS{11}Weep for the dead, for light has failed
{\ADD{him;
}}
\par }{\Q And weep for a fool, for understanding has failed
{\ADD{him:
}}
\par }{\Q Weep more sweetly for the dead, because he has found rest;
\par }{\Q But the life of the fool is worse than death.
\par }{\Q \VS{12}Seven days are
{\ADD{the days of
}} mourning for the dead;
\par }{\Q But for a fool and an ungodly man, all the days of his life.
\par }{\BB \par }{\Q \VS{13}Talk not much with a foolish man,
\par }{\Q And go not to one that has no understanding:
\par }{\Q Beware of him, lest you have trouble;
\par }{\Q And
{\ADD{so
}} you shall not be
\FTNT{†}{{\FR 22:13 }
Or,
defiled: in his onslaught turn
} defiled in his onslaught:
\par }{\Q Turn away from him, and you will find rest;
\par }{\Q And
{\ADD{so
}} you shall not be wearied in his madness.
\par }{\Q \VS{14}What shall be heavier than lead?
\par }{\Q And what is the name thereof, but a fool?
\par }{\Q \VS{15}Sand, and salt, and a mass of iron, is easier to bear,
\par }{\Q Than a man without understanding.
\par }{\BB \par }{\Q \VS{16}Timber girded and bound into a building will not be released with shaking:
\par }{\Q So a heart established in due season on well advised counsel will not be afraid.
\par }{\Q \VS{17}A heart settled upon a thoughtful understanding
\par }{\Q Is as an ornament of plaster on a polished wall.
\par }{\Q \VS{18}Pales set on a high place will not stand against the wind:
\par }{\Q So a fearful heart in the imagination of a fool will not stand against any fear.
\par }{\BB \par }{\Q \VS{19}He that pricks the eye will make tears to fall;
\par }{\Q And he that pricks the heart makes it to show feeling.
\par }{\Q \VS{20}Whoso casts a stone at birds frays them away;
\par }{\Q And he that upbraids a friend will dissolve friendship.
\par }{\Q \VS{21}If you have drawn a sword against a friend, despair not;
\par }{\Q For there may be a returning.
\par }{\Q \VS{22}If you have opened your mouth against a friend, fear not;
\par }{\Q For there may be a reconciling;
\par }{\Q Except it be for upbraiding, and arrogance, and disclosing of a secret, and a treacherous blow:
\par }{\Q For these things every friend will flee.
\par }{\BB \par }{\Q \VS{23}Gain trust with your neighbor in his poverty,
\par }{\Q That in his prosperity you may have gladness:
\par }{\Q Stay steadfast to him in the time of his affliction,
\par }{\Q That you may be heir with him in his inheritance.
\FTNT{‡}{{\FR 22:23 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{24}Before fire is the vapor and smoke of a furnace;
\par }{\Q So revilings before bloodshed.
\par }{\Q \VS{25}I will not be ashamed to shelter a friend;
\par }{\Q And I will not hide myself from his face:
\par }{\Q \VS{26}And if any evil happen to me because of him,
\par }{\Q Every one that hears it will beware of him.
\par }{\BB \par }{\Q \VS{27}Who shall set a watch over my mouth,
\par }{\Q And a seal of shrewdness upon my lips,
\par }{\Q That I fall not from it, and that my tongue destroy me not?

\par }\PoetryChap{23}{\Q \VerseOne{1}O Lord, Father and Master of my life,
\par }{\Q Abandon me not to their counsel:
\par }{\Q Suffer me not to fall by them.
\par }{\Q \VS{2}Who will set scourges over my thought,
\par }{\Q And a discipline of wisdom over my heart?
\par }{\Q That they spare me not for my ignorances,
\par }{\Q And
{\ADD{my heart
}} pass not by their sins:
\par }{\Q \VS{3}That my ignorances be not multiplied,
\par }{\Q And my sins abound not;
\par }{\Q And I will fall before my adversaries,
\par }{\Q And my enemy rejoice over me.
\FTNT{*}{{\FR 23:3 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{4}O Lord, Father and God of my life,
\par }{\Q Give me not a
\FTNT{†}{{\FR 23:4 }
Gr.
lifting up of eyes.
} proud look,
\FTNT{‡}{{\FR 23:4 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{5}And turn away concupiscense from me.
\FTNT{§}{{\FR 23:5 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{6}Let not
\FTNT{*}{{\FR 23:6 }
Gr.
appetite of the belly.
} greediness and chambering overtake me;
\par }{\Q And give me not over to a shameless mind.
\par }{\BB \par }{\Q \VS{7}Hear you°, my children, the discipline of the mouth;
\par }{\Q And he that keeps it will not be taken.
\par }{\Q \VS{8}The sinner will be
\FTNT{†}{{\FR 23:8 }
Most authorities read
left.
} overtaken in his lips;
\par }{\Q And the reviler and the proud man will stumble therein.
\par }{\Q \VS{9}Accustom not your mouth to an oath;
\par }{\Q And be not accustomed to the naming of the Holy One.
\par }{\Q \VS{10}For as a servant that is continually scourged will not lack a bruise,
\par }{\Q So he also that swears and names
{\ADD{God
}} continually will not be cleansed from sin.
\par }{\Q \VS{11}A man of many oaths will be filled with iniquity;
\par }{\Q And the scourge will not depart from his house:
\par }{\Q If he shall offend, his sin will be upon him;
\par }{\Q And if he disregard it, he has sinned doubly;
\par }{\Q And if he has sworn in vain, he will not be justified;
\par }{\Q For his house will be filled with calamities.
\par }{\Q \VS{12}There is a manner of speech that is clothed about with death:
\par }{\Q Let it not be found in the heritage of Jacob;
\par }{\Q For all these things will be far from the godly,
\par }{\Q And they will not wallow in sins.
\par }{\Q \VS{13}Accustom not your mouth to gross rudeness,
\par }{\Q For therein is the word of sin.
\par }{\Q \VS{14}Remember your father and your mother,
\par }{\Q For you sit in the midst of great men;
\par }{\Q That you be not forgetful before them,
\par }{\Q And become a fool by your custom;
\par }{\Q So will you wish that you had not been born,
\par }{\Q And curse the day of your nativity.
\par }{\Q \VS{15}A man that is accustomed to words of reproach
\par }{\Q Will not be corrected all the days of his life.
\par }{\BB \par }{\Q \VS{16}Two sorts
{\ADD{of men
}} multiply sins,
\par }{\Q And the third will bring wrath:
\par }{\Q A hot mind, as a burning fire, will not be quenched till it be consumed:
\par }{\Q A fornicator in the body of his flesh will never cease till he has
\FTNT{‡}{{\FR 23:16 }
Or,
made a fire blaze up
} burned out the fire.
\par }{\Q \VS{17}All bread is sweet to a fornicator:
\par }{\Q He will not leave off till he die.
\par }{\Q \VS{18}A man that goes astray from his own bed,
\par }{\Q Saying in his heart, Who sees me?
\par }{\Q Darkness is round about me, and the walls hide me,
\par }{\Q And no man sees me; of whom am I afraid?
\par }{\Q The Most High will not remember my sins;
\par }{\Q \VS{19}—And the eyes of men are his terror,
\par }{\Q And he knows not that the eyes of the Lord are ten thousand times brighter than the sun,
\par }{\Q Beholding all the ways of men,
\par }{\Q And looking into secret places.
\par }{\Q \VS{20}All things were known to him or ever they were created;
\par }{\Q And in like manner also after they were perfected.
\par }{\Q \VS{21}This man will be punished in the streets of the city;
\par }{\Q And where he suspected not he will be taken.
\par }{\Q \VS{22}So also a wife that leaves her husband,
\par }{\Q And brings in an heir by a stranger.
\par }{\Q \VS{23}For first, she was disobedient in the law of the Most High;
\par }{\Q And secondly, she trespassed against her own husband;
\par }{\Q And thirdly, she played the adulteress in whoredom,
\par }{\Q And brought in children by a stranger.
\par }{\Q \VS{24}She shall be brought out into the congregation;
\par }{\Q And upon her children will there be visitation.
\par }{\Q \VS{25}Her children will not spread into roots,
\par }{\Q And her branches will bear no fruit.
\par }{\Q \VS{26}She will leave her memory for a curse;
\par }{\Q And her reproach will not be blotted out.
\par }{\Q \VS{27}And those who are left behind will know that there is nothing better than the fear of the Lord,
\par }{\Q And nothing sweeter than to take heed to the commandments of the Lord.
\VS{28}\FTNT{§}{{\FR 23:28 }
Verse 28 is omitted by the best authorities.}

\par }\PoetryChap{24}{\Q \VerseOne{1}Wisdom will praise
\FTNT{*}{{\FR 24:1 }
Gr.
her own soul.
} herself,
\par }{\Q And will glory in the midst of her people.
\par }{\Q \VS{2}She will open her mouth in the congregation of the Most High,
\par }{\Q And glory in the presence of his power.
\par }{\Q \VS{3}I came forth from the mouth of the Most High,
\par }{\Q And covered the earth as a mist.
\par }{\Q \VS{4}I lived in high places,
\par }{\Q And my throne is in the pillar of the cloud.
\par }{\Q \VS{5}Alone I compassed the circuit of heaven,
\par }{\Q And walked in the depth of the abyss.
\par }{\Q \VS{6}In the waves of the sea, and in all the earth,
\par }{\Q And in every people and nation, I got a possession.
\par }{\Q \VS{7}With all these I sought rest;
\par }{\Q And in whose inheritance shall I lodge?
\par }{\Q \VS{8}Then the Creator of all things gave me a commandment;
\par }{\Q And he that created me made my tabernacle to rest,
\par }{\Q And said, Let your tabernacle be in Jacob,
\par }{\Q And your inheritance in Israel.
\par }{\Q \VS{9}He created me from the beginning before the world;
\par }{\Q And to the end I will not fail.
\par }{\Q \VS{10}In the holy tabernacle I ministered before him;
\par }{\Q And so was I established in Sion.
\par }{\Q \VS{11}In the beloved city likewise he gave me rest;
\par }{\Q And in Jerusalem was my authority.
\par }{\Q \VS{12}And I took root in a people that was glorified,
\par }{\Q Even in the portion of the Lord’s own inheritance.
\par }{\Q \VS{13}I was exalted like a cedar in Libanus,
\par }{\Q And as a cypress tree on the mountains of Hermon.
\par }{\Q \VS{14}I was exalted like a palm tree on the sea shore,
\par }{\Q And as rose plants in Jericho,
\par }{\Q And as a fair olive tree in the plain;
\par }{\Q And I was exalted as a plane tree.
\par }{\Q \VS{15}As cinnamon and aspalathus, I have given a scent of perfumes;
\par }{\Q And as choice myrrh, I spread abroad a pleasant odour;
\par }{\Q As
\FTNT{†}{{\FR 24:15 }
See Exodus 30:34.
} galbanum, and onyx, and stacte,
\par }{\Q And as the fume of frankincense in the tabernacle.
\par }{\Q \VS{16}As the terebinth I stretched out my branches;
\par }{\Q And my branches are branches of glory and grace.
\par }{\Q \VS{17}As the vine I put forth grace;
\par }{\Q And my flowers are the fruit of glory and riches.
\VS{18}\FTNT{‡}{{\FR 24:18 }
Verse 18 is omitted by the best authorities.}
\par }{\Q \VS{19}Come to me, you° that are desirous of me,
\par }{\Q And be you° filled with my produce.
\par }{\Q \VS{20}For my memorial is sweeter than honey,
\par }{\Q And my inheritance than the honeycomb.
\par }{\Q \VS{21}They that eat me will yet be hungry;
\par }{\Q And those who drink me will yet be thirsty.
\par }{\Q \VS{22}He that obeys me will not be ashamed;
\par }{\Q And those who work in me will not do amiss.
\par }{\BB \par }{\Q \VS{23}All these things are the book of the covenant of the Most High God,
\par }{\Q {\ADD{Even
}} the law which Moses commanded us for a heritage to the assemblies of Jacob.
\VS{24}\FTNT{§}{{\FR 24:24 }
Verse 24 is omitted by the best authorities.
}
\par }{\Q \VS{25}It is he that makes wisdom abundant, as Pishon,
\par }{\Q And as Tigris in the days of new
{\ADD{fruits;
}}
\par }{\Q \VS{26}That makes understanding full as Euphrates,
\par }{\Q And as Jordan in the days of harvest;
\par }{\Q \VS{27}That makes instruction to shine forth as the light,
\par }{\Q As Gihon in the days of vintage.
\par }{\Q \VS{28}The first man knew her not perfectly;
\par }{\Q And in like manner the last has not traced her out.
\par }{\Q \VS{29}For her thoughts are filled from the sea,
\par }{\Q And her counsels from the great deep.
\par }{\BB \par }{\Q \VS{30}And I came out as a
\FTNT{*}{{\FR 24:30 }
Gr.
canal.
} stream from a river,
\par }{\Q And as a conduit into a garden.
\par }{\Q \VS{31}I said, I will water my garden,
\par }{\Q And will water abundantly my garden bed;
\par }{\Q And, behold, my stream became a river,
\par }{\Q And my river became a sea.
\par }{\Q \VS{32}I will yet bring instruction to light as the morning,
\par }{\Q And will make
\FTNT{†}{{\FR 24:32 }
Gr.
them.
} these things to shine forth afar off.
\par }{\Q \VS{33}I will yet pour out doctrine as prophecy,
\par }{\Q And leave it to generations of ages.
\par }{\Q \VS{34}Behold that I have not labored for myself only,
\par }{\Q But for all those who diligently seek her.
\par }{\BB \par }
\PoetryChap{25}{\Q \VerseOne{1}In three things I was beautified,
\par }{\Q And stood up beautiful before the Lord and men:
\par }{\Q The concord of kindred, and friendship of neighbors,
\par }{\Q And a woman and her husband that walk together in agreement.
\par }{\Q \VS{2}But three sorts
{\ADD{of men
}} my soul hates,
\par }{\Q And I am greatly offended at their life:
\par }{\Q A poor man that is arrogant, and a rich man that is a liar,
\par }{\Q {\ADD{And
}} an old man that is an adulterer lacking understanding.
\par }{\BB \par }{\Q \VS{3}In
{\ADD{your
}} youth you have not gathered,
\par }{\Q And how should you find in your old age?
\par }{\Q \VS{4}How beautiful a thing is judgement for gray hairs,
\par }{\Q And for elders to know counsel!
\par }{\Q \VS{5}How beautiful is the wisdom of old men,
\par }{\Q And thought and counsel to men that are in honor!
\par }{\Q \VS{6}Much experience is the crown of old men;
\par }{\Q And their glorying is the fear of the Lord.
\par }{\BB \par }{\Q \VS{7}There be nine things that I have thought of, and in my heart counted happy;
\par }{\Q And the tenth I will utter with my tongue:
\par }{\Q A man that has joy of his children;
\par }{\Q A man that lives and looks upon the fall of his enemies:
\par }{\Q \VS{8}Happy is he that dwells with a wife of understanding;
\par }{\Q And he that has not slipped with his tongue;
\par }{\Q And he that has not served a man that is unworthy of him:
\par }{\Q \VS{9}Happy is he that has found prudence;
\par }{\Q And he that speaks in the ears of those who listen.
\par }{\Q \VS{10}How great is he that has found wisdom!
\par }{\Q Yet is there none above him that fears the Lord.
\par }{\Q \VS{11}The fear of the Lord passes all things:
\par }{\Q He that holds it, to whom shall he be likened?
\VS{12}\FTNT{*}{{\FR 25:12 }
Verse 12 is omitted by the best authorities.
}
\par }{\BB \par }{\Q \VS{13}{\ADD{Give me
}} any plague but the plague of the heart;
\par }{\Q And any wickedness but the wickedness of a woman;
\par }{\Q \VS{14}Any calamity, but a calamity from those who hate me;
\par }{\Q And any vengeance, but the vengeance of enemies.
\par }{\Q \VS{15}There is no head above the head of a serpent;
\par }{\Q And there is no wrath above the wrath of an enemy.
\par }{\BB \par }{\Q \VS{16}I will rather dwell with a lion and a dragon,
\par }{\Q Than keep house with a wicked woman.
\par }{\Q \VS{17}The wickedness of a woman changes her look,
\par }{\Q And darkens her countenance as a bear does.
\par }{\Q \VS{18}Her husband will sit at meat among his neighbors,
\par }{\Q And when he hears it he sighs bitterly.
\par }{\Q \VS{19}All malice is but little to the malice of a woman:
\par }{\Q Let the portion of a sinner fall on her.
\par }{\Q \VS{20}{\ADD{As
}} the going up a sandy way
{\ADD{is
}} to the feet of the aged,
\par }{\Q So is a wife full of words to a quiet man.
\par }{\Q \VS{21}Throw not yourself upon the beauty of a woman;
\par }{\Q And desire not a woman for her beauty.
\par }{\Q \VS{22}There is anger, and impudence, and great reproach,
\par }{\Q If a woman maintain her husband.
\par }{\Q \VS{23}A wicked woman is abasement of heart,
\par }{\Q And sadness of countenance, and a wounded heart:
\par }{\Q A woman that will not make her husband happy
\par }{\Q Is
{\ADD{as
}} hands that hang down, and palsied knees.
\par }{\Q \VS{24}From a woman
{\ADD{was
}} the beginning of sin;
\par }{\Q And because of her we all die.
\par }{\Q \VS{25}Give not water an outlet;
\par }{\Q Neither to a wicked woman freedom of speech.
\par }{\Q \VS{26}If she go not
\FTNT{†}{{\FR 25:26 }
Gr.
according to your hand.
} as you would have her,
\par }{\Q Cut her off from your flesh.
\FTNT{‡}{{\FR 25:26 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\BB \par }
\PoetryChap{26}{\Q \VerseOne{1}Happy is the husband of a good wife;
\par }{\Q And the number of his days will be doubled.
\par }{\Q \VS{2}A brave woman rejoices her husband;
\par }{\Q And he will fulfill his years in peace.
\par }{\Q \VS{3}A good wife is a good portion:
\par }{\Q She will be given in the portion of such as fear the Lord.
\par }{\Q \VS{4}Whether a man be rich or poor,
\par }{\Q A good heart
{\ADD{makes
}} at all times a cheerful countenance.
\par }{\BB \par }{\Q \VS{5}Of three things my heart was afraid;
\par }{\Q And concerning the fourth
\FTNT{*}{{\FR 26:5 }
Gr.
countenance.
} kind I made supplication:
\par }{\Q The slander of a city, and the assembly of a multitude, and a false accusation:
\par }{\Q All these are more grievous than death.
\par }{\Q \VS{6}A grief of heart and sorrow is a woman that is jealous of
{\ADD{another
}} woman,
\par }{\Q And the scourge of a tongue communicating to all.
\par }{\Q \VS{7}A wicked woman is
{\ADD{as
}} a yoke of oxen shaken to and fro:
\par }{\Q He that takes hold of her is as one that grasps a scorpion.
\par }{\Q \VS{8}A drunken woman
{\ADD{causes
}} great wrath;
\par }{\Q And she will not cover her own shame.
\par }{\Q \VS{9}The whoredom of a woman is in the lifting up of her eyes;
\par }{\Q And it will be known by her eyelids.
\par }{\Q \VS{10}Keep strict watch on a headstrong daughter,
\par }{\Q Lest she find liberty for herself, and use it.
\par }{\Q \VS{11}Look well after an impudent eye;
\par }{\Q And marvel not if it trespass against you.
\par }{\Q \VS{12}She will open her mouth, as a thirsty traveller,
\par }{\Q And drink of every water that is near:
\par }{\Q She will sit down at every post,
\par }{\Q And open her quiver against
{\ADD{any
}} arrow.
\par }{\BB \par }{\Q \VS{13}The grace of a wife will delight her husband;
\par }{\Q And her knowledge will fatten his bones.
\par }{\Q \VS{14}A silent woman is a gift of the Lord;
\par }{\Q And there is nothing worth so much as a well-instructed soul.
\par }{\Q \VS{15}A shamefast woman is grace upon grace;
\par }{\Q And there is no
\FTNT{†}{{\FR 26:15 }
Gr.
weight.
} price worthy of a continent soul.
\par }{\Q \VS{16}As the sun when it arises in the highest places of the Lord,
\par }{\Q So is the beauty of a good wife in the ordering of
\FTNT{‡}{{\FR 26:16 }
Gr.
his.
} a man’s house.
\par }{\Q \VS{17}As the lamp that shines upon the holy candlestick,
\par }{\Q So is the beauty of the face in ripe age.
\par }{\Q \VS{18}As the golden pillars are upon a base of silver,
\par }{\Q So are beautiful feet with the breasts of one that is steadfast.
\VS{19-27}\FTNT{§}{{\FR 26:19-27 }
Verses 19-27 are omitted by the best authorities.
}
\par }{\BB \par }{\Q \VS{28}For two things my heart is grieved;
\par }{\Q And for the third anger comes upon me:
\par }{\Q A man of war that suffers for poverty;
\par }{\Q And men of understanding that are counted as refuse:
\par }{\Q One that turns back from righteousness to sin;
\par }{\Q The Lord will prepare him for the sword.
\par }{\BB \par }{\Q \VS{29}A merchant shall hardly keep himself from wrong doing;
\par }{\Q And a huckster will not be acquitted of sin.

\par }\PoetryChap{27}{\Q \VerseOne{1}Many have sinned for a thing indifferent;
\par }{\Q And he that seeks to multiply
{\ADD{gain
}} will turn his eye away.
\par }{\Q \VS{2}A nail will stick fast between the joinings of stones;
\par }{\Q And sin will
\FTNT{*}{{\FR 27:2 }
Gr.
rub.
} thrust itself in between buying and selling.
\par }{\Q \VS{3}Unless
{\ADD{a man
}} hold on diligently in the fear of the Lord,
\par }{\Q His house will soon be overthrown.
\par }{\BB \par }{\Q \VS{4}In the shaking of a sieve, the refuse remains;
\par }{\Q So the filth of man in his reasoning.
\par }{\Q \VS{5}The furnace will prove the potter’s vessels;
\par }{\Q And the trial of a man is in his reasoning.
\par }{\Q \VS{6}The fruit of a tree declares the husbandry thereof;
\par }{\Q So is the utterance of the thought of the heart of a man.
\par }{\Q \VS{7}Praise no man before
{\ADD{you hear him
}} reason;
\par }{\Q For this is the trial of men.
\par }{\BB \par }{\Q \VS{8}If you follow righteousness, you will obtain her,
\par }{\Q And put her on, as a long robe of glory.
\par }{\Q \VS{9}Birds will return to their like;
\par }{\Q And truth will return to those who practice her.
\par }{\Q \VS{10}The lion lies in wait for prey;
\par }{\Q So does sin for those who work iniquity.
\par }{\BB \par }{\Q \VS{11}The discourse of a godly man is always wisdom:
\par }{\Q But the foolish man changes as the moon.
\par }{\Q \VS{12}Among men void of understanding observe the opportunity;
\par }{\Q But stay continually among the thoughtful.
\par }{\Q \VS{13}The discourse of fools is an offence;
\par }{\Q And their laughter is in the wantonness of sin.
\par }{\Q \VS{14}The talk of a man of many oaths will make the hair stand upright;
\par }{\Q And their strife makes one stop his ears.
\par }{\Q \VS{15}The strife of the proud is a shedding of blood;
\par }{\Q And their reviling of each other is a grievous thing to hear.
\par }{\BB \par }{\Q \VS{16}He that reveals secrets destroys credit,
\par }{\Q And will not find a friend to his mind.
\par }{\Q \VS{17}Love a friend, and keep faith with him:
\par }{\Q But if you reveal his secrets,
\par }{\Q You shall not pursue after him;
\par }{\Q \VS{18}For as a man has destroyed his enemy,
\par }{\Q So have you destroyed the friendship of your neighbor.
\par }{\Q \VS{19}And as a bird which you have released out of your hand,
\par }{\Q So have you let your neighbor go, and you will not catch him again:
\par }{\Q \VS{20}Pursue him not, for he is gone far away,
\par }{\Q And has escaped as a gazelle out of the snare.
\par }{\Q \VS{21}For a wound may be bound up, and after reviling there may be a reconcilement;
\par }{\Q But he that reveals secrets has lost hope.
\par }{\BB \par }{\Q \VS{22}One that winks with the eye contrives evil things;
\par }{\Q And no man will remove him from it.
\par }{\Q \VS{23}When you are present, he will speak sweetly,
\par }{\Q And will admire your words;
\par }{\Q But afterward he will writhe his mouth,
\par }{\Q And set a trap
{\ADD{for you
}} in your words.
\par }{\Q \VS{24}I have hated many things, but nothing like him;
\par }{\Q And the Lord will hate him.
\par }{\BB \par }{\Q \VS{25}One that casts a stone on high casts it on his own head;
\par }{\Q And a deceitful stroke will open wounds.
\par }{\Q \VS{26}He that digs a pit will fall into it;
\par }{\Q And he that sets a snare will be taken therein.
\par }{\Q \VS{27}He that does evil things, they will roll upon him,
\par }{\Q And he will not know whence they have come to him.
\par }{\Q \VS{28}Mockery and reproach are from the arrogant;
\par }{\Q And vengeance, as a lion, will lie in wait for him.
\par }{\Q \VS{29}They that rejoice at the fall of the godly will be taken in a snare;
\par }{\Q And anguish will consume them before they die.
\par }{\BB \par }{\Q \VS{30}Wrath and anger, these also are abominations;
\par }{\Q And a sinful man will possess them.

\par }\PoetryChap{28}{\Q \VerseOne{1}He that takes vengeance will find vengeance from the Lord;
\par }{\Q And he will surely make firm his sins.
\par }{\Q \VS{2}Forgive your neighbor the hurt that he has done
{\ADD{you;
}}
\par }{\Q And then your sins will be pardoned when you pray.
\par }{\Q \VS{3}Man cherishes anger against man;
\par }{\Q And does he seek healing from the Lord?
\par }{\Q \VS{4}Upon a man like himself he has no mercy;
\par }{\Q And does he make supplication for his own sins?
\par }{\Q \VS{5}He being himself flesh nourishes wrath:
\par }{\Q Who will make atonement for his sins?
\par }{\Q \VS{6}Remember your last end, and cease from enmity:
\par }{\Q {\ADD{Remember
}} corruption and death, and remain in the commandments.
\par }{\Q \VS{7}Remember the commandments, and be not angry with your neighbor;
\par }{\Q And
{\ADD{remember
}} the covenant of the Highest, and wink at ignorance.
\par }{\BB \par }{\Q \VS{8}Abstain from strife, and you will diminish your sins:
\par }{\Q For a passionate man will kindle strife;
\par }{\Q \VS{9}And a man that is a sinner will trouble friends,
\par }{\Q And will make debate among those who are at peace.
\par }{\Q \VS{10}\FTNT{*}{{\FR 28:10 }
The order of the lines in this verse is uncertain.
} As is the fuel of the fire, so will it burn;
\par }{\Q And as the stoutness of the strife is,
{\ADD{so
}} will it burn:
\par }{\Q As is the strength of the man,
{\ADD{so
}} will be his wrath;
\par }{\Q And as is his wealth,
{\ADD{so
}} he will exalt his anger.
\par }{\Q \VS{11}A contention begun in haste kindles a fire;
\par }{\Q And a hasty fighting sheds blood.
\par }{\Q \VS{12}If you blow a spark, it will burn;
\par }{\Q And if you spit upon it, it will be quenched:
\par }{\Q And both these shall come out of your mouth.
\par }{\BB \par }{\Q \VS{13}Curse the whisperer and double-tongued:
\par }{\Q For he has destroyed many that were at peace.
\par }{\Q \VS{14}A third person’s tongue has shaken many,
\par }{\Q And dispersed them from nation to nation;
\par }{\Q And it has pulled down strong cities,
\par }{\Q And overthrown the houses of great men.
\par }{\Q \VS{15}A third person’s tongue has cast out brave women,
\par }{\Q And deprived them of their labors.
\par }{\Q \VS{16}He that listens to it will not find rest,
\par }{\Q Nor shall he dwell quietly.
\par }{\Q \VS{17}The stroke of a whip makes a mark in the flesh;
\par }{\Q But the stroke of a tongue will break bones.
\par }{\Q \VS{18}Many have fallen by the edge of the sword:
\par }{\Q Yet not so many as those who have fallen because of the tongue.
\par }{\Q \VS{19}Happy is he that is sheltered from it,
\par }{\Q That has not passed through the wrath thereof;
\par }{\Q That has not drawn its yoke,
\par }{\Q And has not been bound with its bands.
\par }{\Q \VS{20}For the yoke thereof is a yoke of iron,
\par }{\Q And the bands thereof are bands of brass.
\par }{\Q \VS{21}The death thereof is an evil death;
\par }{\Q And Hades were better than it.
\par }{\Q \VS{22}It shall not have rule over godly men;
\par }{\Q And they shall not be burned in its flame.
\par }{\Q \VS{23}They that forsake the Lord shall fall into it;
\par }{\Q And it shall burn among them, and shall not be quenched:
\par }{\Q It shall be sent forth upon them as a lion;
\par }{\Q And as a leopard it shall destroy them.
\par }{\Q \VS{24}Look that you hedge your possession about with thorns;
\par }{\Q Bind up your silver and your gold;
\par }{\Q \VS{25}And make a balance and a weight for your words;
\par }{\Q And make a door and a bar for your mouth.
\par }{\Q \VS{26}Take heed lest you slip therein;
\par }{\Q Lest you fall before one that lies in wait.
\par }{\BB \par }
\PoetryChap{29}{\Q \VerseOne{1}He that shows mercy will lend to his neighbor;
\par }{\Q And he that strengthens him with his hand keeps the commandments.
\par }{\Q \VS{2}Lend to your neighbor in time of his need;
\par }{\Q And pay you your neighbor again in due season.
\par }{\Q \VS{3}Confirm your word, and keep faith with him;
\par }{\Q And at all seasons you shall find what you need.
\par }{\Q \VS{4}Many have reckoned a loan as a windfall,
\par }{\Q And have given trouble to those that helped them.
\par }{\Q \VS{5}Till he has received, he will kiss a man’s hands;
\par }{\Q And for his neighbor’s money he will speak submissly:
\par }{\Q And when payment is due, he will prolong the time,
\par }{\Q And return words of heaviness, and complain of
\FTNT{*}{{\FR 29:5 }
Gr.
the season.
} the times.
\par }{\Q \VS{6}If he prevail, he shall hardly receive the half;
\par }{\Q And he will count it as a windfall:
\par }{\Q If not, he has deprived him of his money,
\par }{\Q And he has gotten him for an enemy without cause:
\par }{\Q He will pay him with cursing and railing;
\par }{\Q And for honor he will pay him disgrace.
\par }{\Q \VS{7}\FTNT{†}{{\FR 29:7 }
Some ancient authorities read
Many therefore.
} Many on account of
{\ADD{men’s
}} ill-dealing have turned away;
\par }{\Q They have feared to be defrauded for nothing.
\par }{\Q \VS{8}However with a man in poor estate be patient;
\par }{\Q And let him not wait for
{\ADD{your
}} alms.
\par }{\Q \VS{9}Help a poor man for the commandment’s sake;
\par }{\Q And according to his need send him not empty away.
\par }{\Q \VS{10}Lose
{\ADD{your
}} money for a brother and a friend;
\par }{\Q And let it not rust under the stone to be lost.
\par }{\Q \VS{11}Bestow your treasure according to the commandments of the Most High;
\par }{\Q And it shall profit you more than gold.
\par }{\Q \VS{12}Shut up alms in your store-chambers;
\par }{\Q And it shall deliver you out of all affliction:
\par }{\Q \VS{13}It shall fight for you against your enemy
\par }{\Q Better than a mighty shield and a ponderous spear.
\par }{\BB \par }{\Q \VS{14}A good man will be surety for his neighbor;
\par }{\Q And he that has lost shame will fail him.
\par }{\Q \VS{15}Forget not the good offices of your surety;
\par }{\Q For he has given his life for you.
\par }{\Q \VS{16}A sinner will overthrow the good estate of his surety;
\par }{\Q \VS{17}And he that is of an unthankful mind will fail him that delivered him.
\par }{\Q \VS{18}Suretiship has undone many that were prospering,
\par }{\Q And shaken them as a wave of the sea:
\par }{\Q Mighty men has it driven from their homes;
\par }{\Q And they wandered among strange nations.
\par }{\Q \VS{19}A sinner that falls into suretiship,
\par }{\Q And undertakes contracts for work, shall fall into lawsuits.
\par }{\Q \VS{20}Help your neighbor according to your power,
\par }{\Q And take heed to yourself that you fall not
{\ADD{the same way.
}}
\par }{\BB \par }{\Q \VS{21}The chief thing for life is water, and bread,
\par }{\Q And a garment, and a house to cover shame.
\par }{\Q \VS{22}Better is the life of a poor man under a shelter of logs,
\par }{\Q Than sumptuous fare in another man’s house.
\par }{\Q \VS{23}With little or with much, be well satisfied.
\FTNT{‡}{{\FR 29:23 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{24}It is a miserable life to go from house to house:
\par }{\Q And where you are a sojourner, you shall not
{\ADD{dare to
}} open your mouth.
\par }{\Q \VS{25}You shall entertain, and give to drink, and have no thanks:
\par }{\Q And in addition to this you shall hear bitter words.
\par }{\Q \VS{26}Come here, you sojourner, furnish a table,
\par }{\Q And if you have anything in your hand, feed me with it.
\par }{\Q \VS{27}Go forth, you sojourner, from the face of honor;
\par }{\Q My brother is come to be my guest; I have need of my house.
\par }{\Q \VS{28}These things are grievous to a man of understanding;
\par }{\Q The upbraiding of house-room, and the reproaching of the money-lender.
\par }{\BB \par }
\PoetryChap{30}{\Q \VerseOne{1}He that loves his son will continue to lay stripes upon him,
\par }{\Q That he may have joy of him in the end.
\par }{\Q \VS{2}He that chastises his son shall have profit of him,
\par }{\Q And shall glory of him among his acquaintance.
\par }{\Q \VS{3}He that teaches his son shall provoke his enemy to jealousy;
\par }{\Q And before friends he shall rejoice of him.
\par }{\Q \VS{4}His father dies, and is as though he had not died;
\par }{\Q For he has left one behind him like himself.
\par }{\Q \VS{5}In his life, he saw and rejoiced
{\ADD{in him;
}}
\par }{\Q And when he died, he sorrowed not:
\par }{\Q \VS{6}He left behind him an avenger against his enemies,
\par }{\Q And one to requite kindness to his friends.
\par }{\Q \VS{7}He that makes too much of his son shall bind up his wounds;
\par }{\Q And his heart will be troubled at every cry.
\par }{\Q \VS{8}An unbroken horse becomes stubborn;
\par }{\Q And a son left at large becomes headstrong.
\par }{\Q \VS{9}Cocker your child, and he shall make you afraid:
\par }{\Q Play with him, and he will grieve you.
\par }{\Q \VS{10}Laugh not with him, lest you have sorrow with him;
\par }{\Q And you shall gnash your teeth in the end.
\par }{\Q \VS{11}Give him no liberty in his youth,
\par }{\Q \FTNT{*}{{\FR 30:11 }
These three lines are absent from the oldest MSS.
} And wink not at his follies.
\par }{\Q \VS{12}\FTNT{†}{{\FR 30:12 }
These three lines are absent from the oldest MSS.
} Bow down his neck in his youth,
\par }{\Q And beat him on the sides while he is a child,
\par }{\Q Lest he wax stubborn, and be disobedient to you;
\par }{\Q \FTNT{‡}{{\FR 30:12 }
These three lines are absent from the oldest MSS.
} And there shall be sorrow to your soul.
\par }{\Q \VS{13}Chastise your son, and take pains with him,
\par }{\Q Lest his shameless behavior be an offence to you.
\par }{\BB \par }{\Q \VS{14}Better is a poor man, being sound and strong of constitution,
\par }{\Q Than a rich man that is plagued in his body.
\par }{\Q \VS{15}Health and a good constitution are better than all gold;
\par }{\Q And a strong body than wealth without measure.
\par }{\Q \VS{16}There is no riches better than health of body;
\par }{\Q And there is no gladness above the joy of the heart.
\par }{\Q \VS{17}Death is better than a bitter life,
\par }{\Q \FTNT{§}{{\FR 30:17 }
The oldest MSS. omit
And eternal rest.
} And
\FTNT{*}{{\FR 30:17 }
Gr.
age-long rest.
} eternal rest than a continual sickness.
\par }{\Q \VS{18}Good things poured out upon a mouth that is closed
\par }{\Q Are
{\ADD{as
}} messes of meat laid upon a grave.
\par }{\Q \VS{19}What does an offering profit an idol?
\par }{\Q For neither shall it eat nor smell:
\par }{\Q So is he that is afflicted of the Lord,
\par }{\Q \VS{20}Seeing with his eyes and groaning,
\par }{\Q As an eunuch embracing a virgin and groaning.
\par }{\BB \par }{\Q \VS{21}Give not over your soul to sorrow;
\par }{\Q And afflict not yourself in your own counsel.
\par }{\Q \VS{22}Gladness of heart is the life of a man;
\par }{\Q And the joyfulness of a man is length of days.
\par }{\Q \VS{23}Love your own soul, and comfort your heart:
\par }{\Q And remove sorrow far from you;
\par }{\Q For sorrow has destroyed many,
\par }{\Q And there is no profit therein.
\par }{\Q \VS{24}Envy and wrath shorten
{\ADD{a man’s
}} days;
\par }{\Q And care brings old age before the time.
\par }{\Q \VS{25}A cheerful and good heart
\par }{\Q Will have a care of his meat and diet.

\par }\PoetryChap{31}{\Q \VerseOne{1}Wakefulness that comes of riches consumes the flesh,
\par }{\Q And the anxiety thereof puts away sleep.
\par }{\Q \VS{2}Wakeful anxiety will crave slumber;
\par }{\Q And
\FTNT{*}{{\FR 31:2 }
This appears to be the meaning; but the Greek text here is probably corrupt.
} in a sore disease sleep will be broken.
\par }{\BB \par }{\Q \VS{3}A rich man toils in gathering money together;
\par }{\Q And when he rests, he is filled with his good things.
\par }{\Q \VS{4}A poor man toils in lack of substance;
\par }{\Q And when he rests, he becomes needy.
\par }{\Q \VS{5}He that loves gold shall not be justified;
\par }{\Q And he that follows destruction shall himself have his fill
{\ADD{of it.
}}
\par }{\Q \VS{6}Many have been given over to ruin for the sake of gold;
\par }{\Q And their perdition
\FTNT{†}{{\FR 31:6 }
Or,
comes to pass in their faces
} meets them face to face.
\par }{\Q \VS{7}It is a stumbling block to those who sacrifice to it;
\par }{\Q And every fool shall be taken therewith.
\par }{\Q \VS{8}Blessed is the rich that is found without blemish,
\par }{\Q And that goes not after gold.
\par }{\Q \VS{9}Who is he? and we will call him blessed:
\par }{\Q For wonderful things has he done among his people.
\par }{\Q \VS{10}Who has been tried by it, and found perfect?
\par }{\Q Then let him glory.
\par }{\Q Who has had the power to transgress, and has not transgressed?
\par }{\Q And to do evil, and has not done it?
\par }{\Q \VS{11}His goods shall be made sure,
\par }{\Q And the congregation shall declare his alms.
\par }{\Q \VS{12}Sittest you at a great table?
\FTNT{‡}{{\FR 31:12 }
Gr.
open not your throat upon it.} be not greedy upon it,
\par }{\Q And say not, Many are the things upon it.
\par }{\Q \VS{13}Remember that an evil eye is a wicked thing:
\par }{\Q What has been created more evil than an eye?
\par }{\Q Therefore it sheds tears from every face.
\par }{\Q \VS{14}Stretch not your hand wherever it looks,
\par }{\Q And thrust not yourself with it into the dish.
\par }{\Q \VS{15}Consider your neighbor’s
{\ADD{liking
}} by your own;
\par }{\Q And be discreet in every point.
\par }{\Q \VS{16}Eat, as
{\ADD{becomes
}} a man, those things which are set before you;
\par }{\Q And eat not greedily, lest you be hated.
\par }{\Q \VS{17}Be first to leave off for manners’ sake;
\par }{\Q And be not insatiable, lest you offend.
\par }{\Q \VS{18}And if you sit among many,
\par }{\Q Reach not out your hand before them.
\par }{\BB \par }{\Q \VS{19}How sufficient to a well-mannered man is a very little,
\par }{\Q And he does not breathe hard upon his bed.
\par }{\Q \VS{20}Healthy sleep comes of moderate eating;
\par }{\Q He rises early, and his wits are with him:
\par }{\Q The pain of wakefulness, and colic,
\par }{\Q And griping, are with an insatiable man.
\par }{\Q \VS{21}And if you have been forced to eat,
\par }{\Q Rise up in the midst thereof, and you shall have rest.
\par }{\Q \VS{22}Hear me, my son, and despise me not,
\par }{\Q And at the last you shall find my words
{\ADD{true:
}}
\par }{\Q In all your works be quick,
\par }{\Q And no disease shall come to you.
\par }{\BB \par }{\Q \VS{23}Him that is liberal of his meat the lips shall bless;
\par }{\Q And the testimony of his excellence shall be believed.
\par }{\Q \VS{24}The city will murmer at he who is a stingy with his meat;
\par }{\Q And the testimony of his stinginess will be sure.
\par }{\BB \par }{\Q \VS{25}Show not yourself valiant in wine;
\par }{\Q For wine has destroyed many.
\par }{\Q \VS{26}The furnace proves the temper
{\ADD{of steel
}} by dipping;
\par }{\Q So does wine
{\ADD{prove
}} hearts in the quarreling of the proud.
\par }{\Q \VS{27}Wine is as good as life to men,
\par }{\Q If you drink it in its measure:
\par }{\Q What life is there to a man that is without wine?
\par }{\Q And it has been created to make men glad.
\par }{\Q \VS{28}Wine drunk in season
{\ADD{and
}} to satisfy
\par }{\Q Is joy of heart, and gladness of soul:
\par }{\Q \VS{29}Wine drunk largely is bitterness of soul,
\par }{\Q With provocation and conflict.
\par }{\Q \VS{30}Drunkenness increases the rage of a fool to his hurt;
\par }{\Q It diminishes strength, and adds wounds.
\par }{\BB \par }{\Q \VS{31}Rebuke not your neighbor at a banquet of wine,
\par }{\Q Neither set him at nothing in his mirth:
\par }{\Q Speak not to him a word of reproach,
\par }{\Q And press not upon him by asking back
{\ADD{a debt.
}}

\par }\PoetryChap{32}{\Q \VerseOne{1}Have they made you ruler
{\ADD{of a feast?
}}
\par }{\Q Be not lifted up,
\par }{\Q Be you among them as one of them;
\par }{\Q Take thought for them, and so sit down.
\par }{\Q \VS{2}And when you have done all your office, take your place,
\par }{\Q That you may be gladdened on their account,
\par }{\Q And receive a crown for your well ordering.
\par }{\BB \par }{\Q \VS{3}Speak, you that are the elder, for it becomes you,
{\ADD{but
}} with sound knowledge;
\par }{\Q And hinder not music.
\par }{\Q \VS{4}Pour not out talk where there is a performance of music,
\par }{\Q And display not your wisdom out of season.
\par }{\Q \VS{5}{\ADD{As
}} a signet of carbuncle in a setting of gold,
\par }{\Q {\ADD{So
}} is a concert of music in a banquet of wine.
\par }{\Q \VS{6}{\ADD{As
}} a signet of emerald in a work of gold,
\par }{\Q {\ADD{So
}} is a strain of music with pleasant wine.
\par }{\BB \par }{\Q \VS{7}Speak, young man, if there be need of you;
\par }{\Q {\ADD{Yet
}} scarcely if you be twice asked.
\par }{\Q \VS{8}Sum up your speech, many things in few words;
\par }{\Q Be as one that knows and yet holds his tongue.
\par }{\Q \VS{9}{\ADD{If you be
}} among great men, behave not as their equal;
\par }{\Q And when another is speaking, make not much babbling.
\par }{\Q \VS{10}Lightning speeds before thunder;
\par }{\Q And before a shamefast man favor shall go forth.
\par }{\Q \VS{11}Rise up betimes, and be not the last;
\par }{\Q Get you home quickly and loiter not:
\par }{\Q \VS{12}There take your pastime, and do what is in your heart;
\par }{\Q And sin not by proud speech:
\par }{\Q \VS{13}And for these things bless him that made you,
\par }{\Q And gives you to drink freely of his good things.
\par }{\BB \par }{\Q \VS{14}He that fears the Lord will receive
{\ADD{his
}} discipline;
\par }{\Q And those who seek
{\ADD{him
}} early shall find favor.
\par }{\Q \VS{15}He that seeks the law shall be filled therewith:
\par }{\Q But the hypocrite shall stumble thereat.
\par }{\Q \VS{16}They that fear the Lord shall find judgement,
\par }{\Q And shall kindle righteous acts as a light.
\par }{\Q \VS{17}A sinful man shuns reproof,
\par }{\Q And will find a judgement according to his will.
\par }{\BB \par }{\Q \VS{18}A man of counsel will not neglect a thought;
\par }{\Q A strange and proud man will not crouch in fear,
\par }{\Q Even after he has done a thing by himself without counsel.
\par }{\Q \VS{19}Do nothing without counsel;
\par }{\Q And when you have once done, repent not.
\par }{\Q \VS{20}Go not in a way of conflict;
\par }{\Q And stumble not in stony places.
\par }{\Q \VS{21}Be not confident in a
\FTNT{*}{{\FR 32:21 }
Gr.
without stumbling blocks.} smooth way.
\par }{\Q \VS{22}And beware of your own children.
\par }{\Q \VS{23}In every work trust your own soul;
\par }{\Q For this is the keeping of the commandments.
\par }{\BB \par }{\Q \VS{24}He that believes the law gives heed to the commandment;
\par }{\Q And he that trusts in the Lord shall suffer no loss.

\par }\PoetryChap{33}{\Q \VerseOne{1}There shall no evil happen to him that fears the Lord;
\par }{\Q But in temptation once and again he will deliver him.
\par }{\Q \VS{2}A wise man will not hate the law;
\par }{\Q But he that is a hypocrite therein is as a ship in a storm.
\par }{\Q \VS{3}A man of understanding will put his trust in the law;
\par }{\Q And the law is faithful to him, as when one asks at the oracle.
\par }{\BB \par }{\Q \VS{4}Prepare
{\ADD{your
}} speech, and so shall you be heard;
\par }{\Q Bind up instruction, and make your answer.
\par }{\Q \VS{5}The heart of a fool is
{\ADD{as
}} a cartwheel;
\par }{\Q And his thoughts like a rolling axle tree.
\par }{\Q \VS{6}A stallion horse is as a mocking friend;
\par }{\Q He neighs under every one that sits upon him.
\par }{\BB \par }{\Q \VS{7}Why does one day excel another,
\par }{\Q When all the light of every day in the year is of the sun?
\par }{\Q \VS{8}By the knowledge of the Lord they were distinguished;
\par }{\Q And he varied seasons and feasts:
\par }{\Q \VS{9}Some of them he exalted and hallowed,
\par }{\Q And some of them has he made ordinary days.
\par }{\Q \VS{10}And all men are from the ground,
\par }{\Q And Adam was created of earth.
\par }{\Q \VS{11}In the abundance of his knowledge the Lord distinguished them,
\par }{\Q And made their ways various:
\par }{\Q \VS{12}Some of them he blessed and exalted,
\par }{\Q And some of them he hallowed and brought near to himself:
\par }{\Q Some of them he cursed and brought low,
\par }{\Q And overthrew them from their place.
\par }{\Q \VS{13}As the clay of the potter in his hand,
\par }{\Q All his ways are according to his good pleasure;
\par }{\Q So men are in the hand of him that made them,
\par }{\Q To render to them according to his judgement.
\par }{\Q \VS{14}Good is the opposite of evil,
\par }{\Q And life is the opposite of death:
\par }{\Q So\FTNT{*}{{\FR 33:14 }
A line of this verse is here omitted by the best authorities.
} the sinner is the opposite of the godly.
\par }{\Q \VS{15}And thus look upon all the works of the Most High;
\par }{\Q Two and two, one against another.
\par }{\BB \par }{\Q \VS{16}And I waked up last,
\par }{\Q As one that gleans after the grape gatherers:
\par }{\Q By the blessing of the Lord I got before them,
\par }{\Q And filled my winepress as one that gathers grapes.
\par }{\Q \VS{17}Consider that I labored not for myself alone,
\par }{\Q But for all those who seek instruction.
\par }{\Q \VS{18}Hear me, you° great men of the people,
\par }{\Q And listen with your ears, you° rulers of the congregation.
\par }{\BB \par }{\Q \VS{19}To son and wife, to brother and friend,
\par }{\Q Give not power over you while you live;
\par }{\Q And give not your goods to another,
\par }{\Q Lest you repent and make supplication for them
{\ADD{again.
}}
\par }{\Q \VS{20}Whilst you yet live, and breath is in you,
\par }{\Q Give not yourself over to anybody.
\par }{\Q \VS{21}For better it is that your children should supplicate you,
\par }{\Q Than that you should look to the hand of your sons.
\par }{\Q \VS{22}In all your works keep the upper hand;
\par }{\Q Bring not a stain on your honor.
\par }{\Q \VS{23}In the day that you end the days of your life,
\par }{\Q And in the time of death, distribute your inheritance.
\par }{\BB \par }{\Q \VS{24}Fodder, a stick, and burdens, for an ass;
\par }{\Q Bread, and discipline, and work, for a servant.
\par }{\Q \VS{25}Set your servant to work, and you shall find rest:
\par }{\Q Leave his hands idle, and he will seek liberty.
\par }{\Q \VS{26}Yoke and thong will bow the neck:
\par }{\Q And for an evil servant there are racks and tortures.
\par }{\Q \VS{27}Send him to labor, that he be not idle;
\par }{\Q For idleness teaches much mischief.
\par }{\Q \VS{28}Set him to work, as is fit for him;
\par }{\Q And if he obey not, make his fetters heavy.
\par }{\Q \VS{29}And be not excessive toward any;
\par }{\Q And without judgement do nothing.
\par }{\BB \par }{\Q \VS{30}If you have a servant, let him be as yourself,
\par }{\Q Because you have bought him with blood.
\par }{\Q \VS{31}If you have a servant, treat him as yourself;
\par }{\Q \FTNT{†}{{\FR 33:31 }
The Greek text of this line is probably corrupt.
} For as your own soul will you have need of him:
\par }{\Q If you treat him ill, and he depart and run away,
\par }{\Q Which way will you go to seek him?
\par }{\BB \par }
\PoetryChap{34}{\Q \VerseOne{1}Vain and false hopes are for a man void of understanding;
\par }{\Q And dreams give wings to fools.
\par }{\Q \VS{2}As one that catches at a shadow, and follows after the wind,
\par }{\Q So is he that sets his mind on dreams.
\par }{\Q \VS{3}The vision of dreams is
{\ADD{as
}} this thing against that,
\par }{\Q The likeness of a face near a face.
\par }{\Q \VS{4}Of an unclean thing what shall be cleansed?
\par }{\Q And of that which is false what shall be true?
\par }{\Q \VS{5}Divinations, and soothsayings, and dreams, are vain:
\par }{\Q And the heart fancies, as a woman’s in travail.
\par }{\Q \VS{6}If they be not sent from the Most High in
{\ADD{your
}} visitation,
\par }{\Q Give not your heart to them.
\par }{\Q \VS{7}For dreams have led many astray:
\par }{\Q And they have failed by putting their hope in them.
\par }{\Q \VS{8}Without lying shall the law be accomplished;
\par }{\Q And wisdom is perfection to a faithful mouth.
\par }{\BB \par }{\Q \VS{9}A well-instructed man knows many things;
\par }{\Q And he that has much experience will declare understanding.
\par }{\Q \VS{10}He that has no experience knows few things:
\par }{\Q But he that has wandered shall increase
{\ADD{his
}} skill.
\par }{\Q \VS{11}In my wandering I have seen many things;
\par }{\Q And more than my words is my understanding.
\par }{\Q \VS{12}Ofttimes was I in danger even to death;
\par }{\Q And I was preserved because of these things.
\par }{\Q \VS{13}The spirit of those that fear the Lord shall live;
\par }{\Q For their hope is upon him that saves them.
\par }{\Q \VS{14}Whoso fears the Lord shall not be afraid, and shall not play the coward;
\par }{\Q For he is his hope.
\par }{\Q \VS{15}Blessed is the soul of him that fears the Lord:
\par }{\Q To whom does he give heed? and who is his stay?
\par }{\Q \VS{16}The eyes of the Lord are upon those who love him,
\par }{\Q A mighty protection and strong stay,
\par }{\Q A cover from the hot blast, and a cover from the noonday,
\par }{\Q A guard from stumbling, and a help from falling.
\par }{\Q \VS{17}He raises up the soul, and enlightens the eyes:
\par }{\Q He gives healing, life, and blessing.
\par }{\BB \par }{\Q \VS{18}He that sacrifices of a thing wrongfully gotten, his offering is made in mockery;
\par }{\Q And the mockeries of wicked men are not well-pleasing.
\par }{\Q \VS{19}The Most High has no pleasure in the offerings of the ungodly;
\par }{\Q Neither is he pacified for sins by the multitude of sacrifices.
\par }{\Q \VS{20}{\ADD{As
}} one that kills the son before his father’s eyes
\par }{\Q Is he that brings a sacrifice from the goods of the poor.
\par }{\Q \VS{21}The bread of the needy is the life of the poor:
\par }{\Q He that deprives him thereof is a man of blood.
\par }{\Q \VS{22}{\ADD{As
}} one that slays his neighbor is he that takes away his living;
\par }{\Q And
{\ADD{as
}} a shedder of blood is he that deprives a hireling of his hire.
\par }{\Q \VS{23}One building, and another pulling down,
\par }{\Q What profit have they had but toil?
\par }{\Q \VS{24}One praying, and another cursing,
\par }{\Q Whose voice will the Lord listen to?
\par }{\Q \VS{25}He that washes himself after
{\ADD{touching
}} a dead body, and touches it again,
\par }{\Q What profit has he in his washing?
\par }{\Q \VS{26}Even so a man fasting for his sins,
\par }{\Q And going again, and doing the same;
\par }{\Q Who will listen to his prayer?
\par }{\Q And what profit has he in his humiliation?
\par }{\BB \par }
\PoetryChap{35}{\Q \VerseOne{1}He that keeps the law multiplies offerings;
\par }{\Q He that takes heed to the commandments sacrifices a peace offering.
\par }{\Q \VS{2}He that requites a good turn offers fine flour;
\par }{\Q And he that gives alms sacrifices a thank offering.
\par }{\Q \VS{3}To depart from wickedness is a thing pleasing to the Lord;
\par }{\Q And to depart from unrighteousness is an atoning sacrifice.
\par }{\Q \VS{4}See that you appear not in the presence of the Lord empty.
\par }{\Q \VS{5}For all these things
{\ADD{are to be done
}} because of the commandment.
\par }{\Q \VS{6}The offering of the righteous makes the altar fat;
\par }{\Q And the sweet savor thereof
{\ADD{is
}} before the Most High.
\par }{\Q \VS{7}The sacrifice of a righteous man is acceptable;
\par }{\Q And the memorial thereof shall not be forgotten.
\par }{\Q \VS{8}Glorify the Lord with a good eye,
\par }{\Q And stint not the first fruits of your hands.
\par }{\Q \VS{9}In every gift show a cheerful countenance,
\par }{\Q And dedicate your tithe with gladness.
\par }{\Q \VS{10}Give to the Most High according as he has given;
\par }{\Q And as your hand has found,
{\ADD{give
}} with a good eye.
\par }{\Q \VS{11}For the Lord recompenses,
\par }{\Q And he will recompense you sevenfold.
\par }{\BB \par }{\Q \VS{12}Think not to corrupt with gifts; for he will not receive them:
\par }{\Q And set not your mind on an unrighteous sacrifice;
\par }{\Q For the Lord is judge,
\par }{\Q And with him is no respect of persons.
\par }{\Q \VS{13}He will not accept any person against a poor man;
\par }{\Q And he will listen to the prayer of him that is wronged.
\par }{\Q \VS{14}He will in no wise despise the supplication of the fatherless;
\par }{\Q Nor the widow,
\FTNT{*}{{\FR 35:14 }
Gr.
if
} when she pours out her tale.
\par }{\Q \VS{15}Do not the tears of the widow run down her cheek?
\par }{\Q And is not her cry against him that has caused them to fall?
\par }{\Q \VS{16}He that serves
{\ADD{God
}} according to his good pleasure shall be accepted,
\par }{\Q And his supplication shall reach to the clouds.
\par }{\Q \VS{17}The prayer of the humble pierces the clouds;
\par }{\Q And till it come near, he will not be comforted;
\par }{\Q And he will not depart, till the Most High shall visit;
\par }{\Q And he shall judge
\FTNT{†}{{\FR 35:17 }
Some ancient authorities read
for the righteous.
} righteously, and execute judgement.
\par }{\Q \VS{18}And the Lord will not be slack, neither will he be patient toward them,
\par }{\Q Till he have crushed the loins of the unmerciful;
\par }{\Q And he shall repay vengeance to the heathen;
\par }{\Q Till he have taken away the multitude of the arrogant,
\par }{\Q And broken in pieces the sceptres of the unrighteous;
\par }{\Q \VS{19}Till he have rendered to
{\ADD{every
}} man according to his doings,
\par }{\Q And
{\ADD{to
}} the works of men according to their plans;
\par }{\Q Till he have judged the cause of his people;
\par }{\Q And he shall make them to rejoice in his mercy.
\par }{\Q \VS{20}Mercy is seasonable in the time of his afflicting
{\ADD{them,
}}
\par }{\Q As clouds of rain in the time of drought.
\par }{\BB \par }
\PoetryChap{36}{\Q \VerseOne{1}Have mercy upon us, O Lord the God of all, and behold;
\par }{\Q \VS{2}And send your fear upon all the nations:
\FTNT{*}{{\FR 36:2 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{3}Lift up your hand against the strange nations;
\par }{\Q And let them see your mighty power.
\par }{\Q \VS{4}As you were sanctified in us before them,
\par }{\Q So be you magnified in them before us.
\par }{\Q \VS{5}And let them know you, as we also have known you,
\par }{\Q That there is no God but only you, O God.
\par }{\Q \VS{6}Show new signs, and work various wonders;
\par }{\Q Glorify your hand and your right arm.
\FTNT{†}{{\FR 36:6 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{7}Raise up indignation, and pour out wrath;
\par }{\Q Take away the adversary, and destroy the enemy.
\par }{\Q \VS{8}Hasten the time, and remember the oath;
\par }{\Q And let them declare your mighty works.
\par }{\Q \VS{9}Let him that escapes be devoured by the rage of fire;
\par }{\Q And may those who harm your people find destruction.
\par }{\Q \VS{10}Crush the heads of the rulers of the enemies,
\par }{\Q That say, There is none but we.
\par }{\Q \VS{11}Gather all the tribes of Jacob together,
\par }{\Q And
\FTNT{‡}{{\FR 36:11 }
The ancient authorities read
I took them for my inheritance:
but the Greek text is here very confused.
} take them for your inheritance, as from the beginning.
\par }{\Q \VS{12}O Lord, have mercy upon the people that is called by your name,
\par }{\Q And upon Israel, whom you did liken to a firstborn.
\par }{\Q \VS{13}Have compassion upon the city of your sanctuary,
\par }{\Q Jerusalem, the place of your rest.
\par }{\Q \VS{14}Fill Sion; exalt your oracles,
\par }{\Q And
{\ADD{fill
}} your people with your glory.
\par }{\Q \VS{15}Give testimony to those that were your creatures in the beginning,
\par }{\Q And raise up the prophecies that have been in your name.
\par }{\Q \VS{16}Give reward to those who wait for you:
\par }{\Q And men shall put their trust in your prophets.
\par }{\Q \VS{17}Listen, O Lord, to the prayer of your suppliants,
\par }{\Q According to the blessing of Aaron concerning your people;
\par }{\Q And all those who are on the earth shall know
\par }{\Q That you are the Lord, the
\FTNT{§}{{\FR 36:17 }
Gr.
God of the ages.
} eternal God.
\par }{\BB \par }{\Q \VS{18}The belly will eat any meat;
\par }{\Q Yet is one meat better than another.
\par }{\Q \VS{19}The
\FTNT{*}{{\FR 36:19 }
Gr.
throat.
} mouth tastes meats taken in hunting:
\par }{\Q So does an understanding heart false speeches.
\par }{\Q \VS{20}A contrary heart will cause heaviness:
\par }{\Q And a man of experience will recompense him.
\par }{\Q \VS{21}A woman will receive any man;
\par }{\Q But one daughter is better than another.
\par }{\Q \VS{22}The beauty of a woman cheers the countenance;
\par }{\Q And a man desires nothing so much.
\par }{\Q \VS{23}If there is on her tongue mercy and meekness,
\par }{\Q Her husband is not like the sons of men.
\par }{\Q \VS{24}He that gets a wife enters upon a possession:
\par }{\Q A help meet for him, and a pillar of rest.
\par }{\Q \VS{25}Where no hedge is, the possession will be laid waste:
\par }{\Q And he that has no wife will mourn as he wanders up and down.
\par }{\Q \VS{26}For who will trust a nimble robber, that skips from city to city?
\par }{\Q Even so
{\ADD{who shall trust
}} a man that has no nest, and lodges wherever he finds himself at nightfall?
\par }{\BB \par }
\PoetryChap{37}{\Q \VerseOne{1}Every friend will say, I also am his friend:
\par }{\Q But there is a friend, which is only a friend in name.
\par }{\Q \VS{2}Is there not a grief in it even to death,
\par }{\Q When a companion and friend is turned to enmity?
\par }{\Q \VS{3}O wicked imagination, whence came you rolling in
\par }{\Q To cover the dry land with deceitfulness?
\par }{\Q \VS{4}There is a companion, which rejoices in the gladness of a friend,
\par }{\Q But in time of affliction will be against him.
\par }{\Q \VS{5}There is a companion, which for the belly’s sake labors with his friend,
\par }{\Q In the face of battle will take up the buckler.
\par }{\Q \VS{6}Forget not a friend in your soul;
\par }{\Q And be not unmindful of him in your riches.
\par }{\BB \par }{\Q \VS{7}Every counselor extols counsel;
\par }{\Q But there is that counsels for himself.
\par }{\Q \VS{8}Let your soul beware of a counselor,
\par }{\Q And know you before what is his interest
\par }{\Q (For he will take counsel for himself);
\par }{\Q Lest he cast the lot upon you,
\par }{\Q \VS{9}And say to you, Your way is good:
\par }{\Q And he will stand near you, to see what will happen to you.
\par }{\Q \VS{10}Take not counsel with one that looks askance at you;
\par }{\Q And hide your counsel from such as are jealous of you.
\par }{\Q \VS{11}{\ADD{Take not counsel
}} with a woman about her rival;
\par }{\Q Neither with a coward about war;
\par }{\Q Nor with a merchant about exchange;
\par }{\Q Nor with a buyer about selling;
\par }{\Q Nor with an envious man about thankfulness;
\par }{\Q Nor with an unmerciful man about kindliness;
\par }{\Q Nor with a sluggard about any kind of work;
\par }{\Q Nor with a hireling in your house about finishing
{\ADD{his work;
}}
\par }{\Q Nor with an idle servant about much business:
\par }{\Q Give not heed to these in any matter of counsel.
\par }{\Q \VS{12}But rather be continually with a godly man,
\par }{\Q Whom you shall have known to be a keeper of the commandments,
\par }{\Q Who in his soul is as your own soul,
\par }{\Q And who will grieve with you, if you shall miscarry.
\par }{\Q \VS{13}And make the counsel of your heart to stand;
\par }{\Q For there is none more faithful to you than it.
\par }{\Q \VS{14}For a man’s soul is sometimes wont to bring him tidings,
\par }{\Q More than seven watchmen that sit on high on a watch-tower.
\par }{\Q \VS{15}And above all this entreat the Most High,
\par }{\Q That he may direct your way in truth.
\par }{\BB \par }{\Q \VS{16}Let reason be the beginning of every work,
\par }{\Q And let counsel go before every action.
\par }{\Q \VS{17}As a token of the changing of the heart,
\par }{\Q \VS{18}four manner of things do rise up,
\par }{\Q Good and evil, life and death;
\par }{\Q And that which rules over them continually is the tongue.
\par }{\Q \VS{19}There is one that is shrewd
{\ADD{and
}} the instructor of many,
\par }{\Q And yet is unprofitable to his own soul.
\par }{\Q \VS{20}There is
{\ADD{one
}} that is subtle in words, and is hated;
\par }{\Q He shall be destitute of all food:
\par }{\Q \VS{21}For grace was not given him from the Lord;
\par }{\Q Because he is deprived of all wisdom.
\par }{\Q \VS{22}There is one that is wise to his own soul;
\par }{\Q And the fruits of his understanding are trustworthy in the mouth.
\par }{\Q \VS{23}A wise man will instruct his own people;
\par }{\Q And the fruits of his understanding are trustworthy.
\par }{\Q \VS{24}A wise man shall be filled with blessing;
\par }{\Q And all those who see him shall call him happy.
\par }{\Q \VS{25}The life of man is counted by days;
\par }{\Q And the days of Israel are innumerable.
\par }{\Q \VS{26}The wise man shall inherit confidence among his people,
\par }{\Q And his name shall live forever.
\par }{\BB \par }{\Q \VS{27}My son, prove your soul in your life,
\par }{\Q And see what is evil for it, and give not that to it.
\par }{\Q \VS{28}For all things are not profitable for all men,
\par }{\Q Neither has every soul pleasure in every thing.
\par }{\Q \VS{29}Don’t be insatiable in any luxury,
\par }{\Q And don’t be greedy in the things that you eat.
\par }{\Q \VS{30}For in multitude of meats there shall be disease,
\par }{\Q And surfeiting shall come near to colic.
\par }{\Q \VS{31}Because of surfeiting have many perished;
\par }{\Q But he that takes heed shall prolong his life.
\par }{\BB \par }
\PoetryChap{38}{\Q \VerseOne{1}Honor a physician according to your need
{\ADD{of him
}} with the honors due to him:
\par }{\Q For truly the Lord has created him.
\par }{\Q \VS{2}For from the Most High comes healing;
\par }{\Q And from the king he shall receive a gift.
\par }{\Q \VS{3}The skill of the physician shall lift up his head;
\par }{\Q And in the sight of great men he shall be admired.
\par }{\Q \VS{4}The Lord created medicines out of the earth;
\par }{\Q And a prudent man will have no disgust at them.
\par }{\Q \VS{5}Was not water made sweet with wood,
\par }{\Q That the virtue thereof might be known?
\par }{\Q \VS{6}And he gave men skill,
\par }{\Q That
\FTNT{*}{{\FR 38:6 }
Or,
he
} they might be glorified in his marvelous works.
\par }{\Q \VS{7}With them does he heal
{\ADD{a man,
}}
\par }{\Q And takes away his pain.
\par }{\Q \VS{8}With these will the apothecary make a confection;
\par }{\Q And his works shall not be brought to an end;
\par }{\Q And from him is peace upon the face of the earth.
\par }{\BB \par }{\Q \VS{9}My son, in your sickness be not negligent;
\par }{\Q But pray to the Lord, and he shall heal you.
\par }{\Q \VS{10}Put away wrong doing, and order your hands aright,
\par }{\Q And cleanse your heart from all manner of sin.
\par }{\Q \VS{11}Give a sweet savor, and a memorial of fine flour;
\par }{\Q And make fat your offering, as one that is not.
\par }{\Q \VS{12}Then give place to the physician, for truly the Lord has created him;
\par }{\Q And let him not go from you, for you have need of him.
\par }{\Q \VS{13}There is a time when in their very hands is the issue for good.
\par }{\Q \VS{14}For they also shall beseech the Lord,
\par }{\Q That he may prosper them in
{\ADD{giving
}} relief and in healing for the maintenance of life.
\par }{\Q \VS{15}He that sins before his Maker,
\par }{\Q Let him fall into the hands of the physician.
\par }{\BB \par }{\Q \VS{16}My son, let your tears fall over the dead,
\par }{\Q And as one that suffers grievously begin lamentation;
\par }{\Q And wind up his body according to his due,
\par }{\Q And neglect not his burial.
\par }{\Q \VS{17}Make bitter weeping, and make passionate wailing,
\par }{\Q And let your mourning be according to his merit,
\par }{\Q For one day or two, lest you be spoken evil of:
\par }{\Q And so be comforted for your sorrow.
\par }{\Q \VS{18}For of sorrow comes death,
\par }{\Q And sorrow of heart will bow down the strength.
\par }{\Q \VS{19}In calamity sorrow also remains:
\par }{\Q And the poor man’s life is
\FTNT{†}{{\FR 38:19 }
Gr.
against the heart.
} grievous to the heart.
\par }{\Q \VS{20}Don’t give your heart to sorrow.
\par }{\Q Put it away, remembering the last end.
\par }{\Q \VS{21}Don’t forget it, for there is no returning again:
\par }{\Q Him you shall not profit, and you will hurt yourself.
\par }{\Q \VS{22}Remember the sentence upon him; for so also shall yours be;
\par }{\Q Yesterday for me, and today for you.
\par }{\Q \VS{23}When the dead is at rest, let his remembrance rest;
\par }{\Q And be comforted for him, when his spirit departs from him.
\par }{\BB \par }{\Q \VS{24}The wisdom of the scribe comes by opportunity of leisure;
\par }{\Q And
\FTNT{‡}{{\FR 38:24 }
Gr.
he that is lessened in his business.
} he that has little business shall become wise.
\par }{\Q \VS{25}How shall he become wise that holds the plow,
\par }{\Q That glories in the shaft of the goad,
\par }{\Q That drives oxen, and is occupied in their labors,
\par }{\Q And whose discourse is of the stock of bulls?
\par }{\Q \VS{26}He will set his heart upon turning his furrows;
\par }{\Q And his wakefulness is to give his heifers their fodder.
\par }{\Q \VS{27}So is every craftsman and workmaster,
\par }{\Q That passes his time by night as by day;
\par }{\Q They that cut gravings of signets,
\par }{\Q And his diligence is to make great variety;
\par }{\Q He will set his heart to preserve likeness in his portraiture,
\par }{\Q And will be wakeful to finish his work.
\par }{\Q \VS{28}So is the smith sitting by the anvil,
\par }{\Q And considering the unwrought iron:
\par }{\Q The vapor of the fire will waste his flesh;
\par }{\Q And in the heat of the furnace he will wrestle
{\ADD{with his work:
}}
\par }{\Q The noise of the hammer will
\FTNT{§}{{\FR 38:28 }
Gr.
renew.
} be ever in his ear,
\par }{\Q And his eyes are upon the pattern of the vessel;
\par }{\Q He will set his heart upon perfecting his works,
\par }{\Q And he will be wakeful to adorn them perfectly.
\par }{\Q \VS{29}So is the potter sitting at his work,
\par }{\Q And turning the wheel about with his feet,
\par }{\Q Who is always anxiously set at his work,
\par }{\Q And all his handiwork is by number;
\par }{\Q \VS{30}He will fashion the clay with his arm,
\par }{\Q And will bend its strength in front of his feet;
\par }{\Q He will apply his heart to finish the glazing;
\par }{\Q And he will be wakeful to make clean the furnace.
\par }{\BB \par }{\Q \VS{31}All these put their trust in their hands;
\par }{\Q And each becomes wise in his own work.
\par }{\Q \VS{32}Without these shall not a city be inhabited,
\par }{\Q And men shall not sojourn nor walk up and down
{\ADD{therein.
}}
\par }{\Q \VS{33}\FTNT{*}{{\FR 38:33 }
This line is absent from the oldest MSS.
} They shall not be sought for in the council of the people,
\par }{\Q And in the assembly they shall not mount on high;
\par }{\Q They shall not sit on the seat of the judge,
\par }{\Q And they shall not understand the covenant of judgement:
\par }{\Q Neither shall they declare instruction and judgement;
\par }{\Q And where parables are they shall not be found.
\par }{\Q \VS{34}But they will maintain the fabric of the
\FTNT{†}{{\FR 38:34 }
Gr.
age.
} world;
\par }{\Q And in the handiwork of their craft is their prayer.
\par }{\BB \par }
\PoetryChap{39}{\Q \VerseOne{1}\FTNT{*}{{\FR 39:1 }
Gr.
Except him that.
} Not so he that has applied his soul,
\par }{\Q And meditates in the law of the Most High;
\par }{\Q He will seek out the wisdom of all the ancients,
\par }{\Q And will be occupied in prophecies.
\par }{\Q \VS{2}He will keep the discourse of the men of renown,
\par }{\Q And will enter in amidst the subtleties of parables.
\par }{\Q \VS{3}He will seek out the hidden meaning of proverbs,
\par }{\Q And be conversant in the dark sayings of parables.
\par }{\Q \VS{4}He will serve among great men,
\par }{\Q And appear before him that rules:
\par }{\Q He will travel through the land of strange nations;
\par }{\Q For he has tried good things and evil among men.
\par }{\Q \VS{5}He will apply his heart to return early to the Lord that made him,
\par }{\Q And will make supplication before the Most High,
\par }{\Q And will open his mouth in prayer,
\par }{\Q And will make supplication for his sins.
\par }{\BB \par }{\Q \VS{6}If the great Lord will,
\par }{\Q He shall be filled with the spirit of understanding:
\par }{\Q He shall pour forth the words of his wisdom,
\par }{\Q And in prayer give thanks to the Lord.
\par }{\Q \VS{7}He shall direct his counsel and knowledge,
\par }{\Q And in his secrets shall he meditate.
\par }{\Q \VS{8}He shall show forth the instruction which he has been taught,
\par }{\Q And shall glory in the law of the covenant of the Lord.
\par }{\Q \VS{9}Many shall commend his understanding;
\par }{\Q And so long as the world endures, it shall not be blotted out:
\par }{\Q His memorial shall not depart,
\par }{\Q And his name shall live from generation to generation.
\par }{\Q \VS{10}Nations shall declare his wisdom,
\par }{\Q And the congregation shall tell out his praise.
\par }{\Q \VS{11}If he continue, he shall leave a greater name than a thousand:
\par }{\Q And if he
\FTNT{†}{{\FR 39:11 }
Gr.
cease.
} die, he adds thereto.
\par }{\BB \par }{\Q \VS{12}Yet more I will utter, which I have thought upon;
\par }{\Q And I am filled as the moon at the full.
\par }{\Q \VS{13}Listen to me, you° holy children,
\par }{\Q And bud forth as a rose growing by a brook of water:
\par }{\Q \VS{14}And give you° a sweet savor as frankincense,
\par }{\Q And put forth flowers as a lily,
\par }{\Q Spread abroad a sweet smell, and sing a song of praise;
\par }{\Q Bless you° the Lord for all his works.
\par }{\Q \VS{15}Magnify his name,
\par }{\Q And give utterance to his praise
\par }{\Q With the songs of your lips, and with harps;
\par }{\Q And thus shall you° say when you° utter
{\ADD{his praise:
}}
\par }{\BB \par }{\Q \VS{16}All the works of the Lord are exceedingly good,
\par }{\Q And every command shall be
{\ADD{accomplished
}} in his season.
\par }{\Q \VS{17}None can say, What is this? wherefore is that?
\par }{\Q For in his season they shall all be sought out.
\par }{\Q At his word the waters stood as a heap,
\par }{\Q And the receptacles of waters at the word of his mouth.
\par }{\Q \VS{18}At his command is all his good pleasure
{\ADD{done;
}}
\par }{\Q And there is none that shall hinder his salvation.
\par }{\Q \VS{19}The works of all flesh are before him;
\par }{\Q And it is not possible to be hid from his eyes.
\par }{\Q \VS{20}He sees from everlasting to everlasting;
\par }{\Q And there is nothing wonderful before him.
\par }{\Q \VS{21}None can say, What is this? wherefore is that?
\par }{\Q For all things are created for their uses.
\par }{\BB \par }{\Q \VS{22}His blessing covered the dry land as a river,
\par }{\Q And saturated it as a flood.
\par }{\Q \VS{23}As he has made the waters salty;
\par }{\Q So shall the heathen inherit his wrath.
\par }{\Q \VS{24}His ways are plain to the holy;
\par }{\Q So are they stumbling blocks to the wicked.
\par }{\Q \VS{25}Good things are created from the beginning for the good;
\par }{\Q So are evil things for sinners.
\par }{\Q \VS{26}The chief of all things necessary for the life of man
\par }{\Q Are water, and fire, and iron, and salt,
\par }{\Q And flour of wheat, and honey, and milk,
\par }{\Q The blood of the grape, and oil, and clothing.
\par }{\Q \VS{27}All these things are for good to the godly;
\par }{\Q So to the sinners they shall be turned into evil.
\par }{\BB \par }{\Q \VS{28}There be winds that are created for vengeance,
\par }{\Q And in their fury lay on their scourges heavily;
\par }{\Q In the time of consummation they pour out their strength,
\par }{\Q And shall appease the wrath of him that made them.
\par }{\Q \VS{29}Fire, and hail, and famine, and death,
\par }{\Q All these are created for vengeance;
\par }{\Q \VS{30}Teeth of wild beasts, and scorpions and adders,
\par }{\Q And a sword punishing the ungodly to destruction.
\par }{\Q \VS{31}They shall rejoice in his commandment,
\par }{\Q And shall be made ready upon earth, when need is;
\par }{\Q And in their seasons they shall not transgress
{\ADD{his
}} word.
\par }{\BB \par }{\Q \VS{32}Therefore from the beginning I was resolved,
\par }{\Q And I thought
{\ADD{this,
}} and left it in writing;
\par }{\Q \VS{33}All the works of the Lord are good:
\par }{\Q And he will supply every need in its season.
\par }{\Q \VS{34}And none can say, This is worse than that:
\par }{\Q For they shall all be well approved in their season.
\par }{\Q \VS{35}And now with all your heart and mouth sing you° praises,
\par }{\Q And bless the name of the Lord.
\par }{\BB \par }
\PoetryChap{40}{\Q \VerseOne{1}Great travail is created for every man,
\par }{\Q And a heavy yoke is upon the sons of Adam,
\par }{\Q From the day of their coming forth from their mother’s womb,
\par }{\Q Until the day for their burial in the mother of all things.
\par }{\Q \VS{2}The expectation of things to come, and the day of death,
\par }{\Q \FTNT{*}{{\FR 40:2 }
The Greek text of this line is probably corrupt.
}
{\ADD{Trouble
}} their thoughts, and
{\ADD{cause
}} fear of heart;
\par }{\Q \VS{3}From him that sits on a throne of glory,
\par }{\Q Even to him that is humbled in earth and ashes;
\par }{\Q \VS{4}From him that wears purple and a crown,
\par }{\Q Even to him that is clothed with a hempen frock.
\par }{\Q \VS{5}{\ADD{There is
}} wrath, and jealousy, and trouble, and disquiet,
\par }{\Q And fear of death, and anger, and strife;
\par }{\Q And in the time of rest upon his bed
\par }{\Q His night sleep does change his knowledge.
\par }{\Q \VS{6}A little or nothing is his resting,
\par }{\Q And afterward in his sleep, as in a day of keeping watch,
\par }{\Q He is troubled in the vision of his heart,
\par }{\Q As one that has escaped from the front of battle.
\par }{\Q \VS{7}In the very time of his deliverance he awakens,
\par }{\Q And marvels that the fear is nothing.
\par }{\BB \par }{\Q \VS{8}{\ADD{It is thus
}} with all flesh, from man to beast,
\par }{\Q And upon sinners sevenfold more.
\par }{\Q \VS{9}Death, and bloodshed, and strife, and sword,
\par }{\Q Calamities, famine, suffering, and the scourge;
\par }{\Q \VS{10}All these things were created for the wicked,
\par }{\Q And because of them came the flood.
\par }{\Q \VS{11}All things that are of the earth turn to the earth again:
\par }{\Q And
{\ADD{all things that are
}} of the waters return into the sea.
\par }{\BB \par }{\Q \VS{12}All bribery and injustice shall be blotted out;
\par }{\Q And good faith shall stand forever.
\par }{\Q \VS{13}The goods of the unjust shall be dried up like a river,
\par }{\Q And like a great thunder in rain shall go off in noise.
\par }{\Q \VS{14}In opening his hands
{\ADD{a man
}} shall be made glad:
\par }{\Q So shall transgressors utterly fail.
\par }{\Q \VS{15}The children of the ungodly shall not put forth many branches;
\par }{\Q And
{\ADD{are as
}} unclean roots upon a sheer rock.
\par }{\Q \VS{16}The sedge
{\ADD{that grows
}} upon every water and bank of a river
\par }{\Q Shall be plucked up before all grass.
\par }{\Q \VS{17}Bounty is as a garden
\FTNT{†}{{\FR 40:17 }
Gr.
in.
} of blessings,
\par }{\Q And almsgiving endures forever.
\par }{\BB \par }{\Q \VS{18}The life of one that labors, and is content, shall be made sweet;
\par }{\Q And he that finds a treasure is above both.
\par }{\Q \VS{19}Children and the building of a city establish a
{\ADD{man’s
}} name;
\par }{\Q And a blameless wife is counted above both.
\par }{\Q \VS{20}Wine and music rejoice the heart;
\par }{\Q And the love of wisdom is above both.
\par }{\Q \VS{21}The pipe and the lute make pleasant melody;
\par }{\Q And a pleasant tongue is above both.
\par }{\Q \VS{22}Your eye shall desire grace and beauty;
\par }{\Q And above both the green blade of corn.
\par }{\Q \VS{23}A friend and a companion never meet amiss;
\par }{\Q And a wife with her husband is above both.
\par }{\Q \VS{24}Brethren and help are for a time of affliction;
\par }{\Q And almsgiving is a deliverer above both.
\par }{\Q \VS{25}Gold and silver will make the foot stand sure;
\par }{\Q And counsel is esteemed above them both.
\par }{\Q \VS{26}Riches and strength will lift up the heart;
\par }{\Q And the fear of the Lord is above both:
\par }{\Q There is nothing lacking in the fear of the Lord,
\par }{\Q And there is no need to seek help therein.
\par }{\Q \VS{27}The fear of the Lord is as a garden of blessing,
\par }{\Q And covers
\FTNT{‡}{{\FR 40:27 }
Gr.
him.
} a man above all glory.
\par }{\BB \par }{\Q \VS{28}My son, lead not a beggar’s life;
\par }{\Q Better it is to die than to beg.
\par }{\Q \VS{29}A man that looks to the table of another,
\par }{\Q His life is not to be counted for a life;
\par }{\Q He will pollute his soul with another man’s meats:
\par }{\Q But a man wise and well-instructed will beware thereof.
\par }{\Q \VS{30}In the mouth of the shameless begging will be sweet;
\par }{\Q And in his belly a fire shall be kindled.
\par }{\BB \par }
\PoetryChap{41}{\Q \VerseOne{1}O death, how bitter is the remembrance of you to a man that is at peace in his possessions,
\par }{\Q To the man that has nothing to distract him, and has prosperity in all things,
\par }{\Q And that still has strength to receive meat!
\par }{\Q \VS{2}O death, acceptable is your sentence to a man that is needy, and that fails in strength,
\par }{\Q That is in extreme old age, and is distracted about all things,
\par }{\Q And is perverse, and has lost patience!
\par }{\Q \VS{3}Don’t be afraid of the sentence of death;
\par }{\Q Remember those who have been before you, and that come after:
\par }{\Q This is the sentence from the Lord over all flesh.
\par }{\Q \VS{4}And why do you refuse, when it is the good pleasure of the Most High?
\par }{\Q Whether it be ten, or a hundred, or a thousand years,
\par }{\Q There is no inquisition of life in
\FTNT{*}{{\FR 41:4 }
Gr.
hades
} the grave.
\par }{\BB \par }{\Q \VS{5}The children of sinners are abominable children,
\par }{\Q And they frequent the dwellings of the ungodly.
\par }{\Q \VS{6}The inheritance of sinners’ children shall perish,
\par }{\Q And with their posterity shall be a perpetual reproach.
\par }{\Q \VS{7}Children will complain of an ungodly father,
\par }{\Q Because they shall be reproached for his sake.
\par }{\Q \VS{8}Woe to you, ungodly men,
\par }{\Q Which have forsaken the law of the Most High God!
\FTNT{†}{{\FR 41:8 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{9}If you° be born, you° shall be born to a curse;
\par }{\Q If you° die, a curse shall be your portion.
\par }{\Q \VS{10}All things that are of the earth shall go back to the earth:
\par }{\Q So the ungodly shall go from a curse to perdition.
\par }{\Q \VS{11}The mourning of men is about their bodies:
\par }{\Q But the name of sinners being evil shall be blotted out.
\par }{\Q \VS{12}Have regard to your name;
\par }{\Q For it continues with you longer than a thousand great treasures of gold.
\par }{\Q \VS{13}A good life has its number of days;
\par }{\Q And a good name continues forever.
\par }{\BB \par }{\Q \VS{14}My children, keep instruction in peace:
\par }{\Q But wisdom that is hid, and a treasure that is not seen,
\par }{\Q What profit is in them both?
\par }{\Q \VS{15}Better is a man that hides his foolishness
\par }{\Q Than a man that hides his wisdom.
\par }{\Q \VS{16}Wherefore show reverence to my word:
\par }{\Q For it is not good to retain every kind of shame;
\par }{\Q And not all things are approved by all in good faith.
\par }{\BB \par }{\Q \VS{17}Be ashamed of whoredom before father and mother:
\par }{\Q And of a lie before a prince and a mighty man;
\par }{\Q \VS{18}Of an offence before a judge and ruler;
\par }{\Q Of iniquity before the congregation and the people;
\par }{\Q Of unjust dealing before a partner and friend;
\par }{\Q \VS{19}And of theft in regard of the place where you sojourn,
\par }{\Q And in regard of the truth of God and his covenant;
\par }{\Q And of leaning with your elbow at meat;
\par }{\Q And of scurrility in the matter of giving and taking;
\par }{\Q \VS{20}And of silence before those who salute you;
\par }{\Q And of looking upon a woman that is a prostitute;
\par }{\Q \VS{21}And of turning away
\FTNT{‡}{{\FR 41:21 }
Or,
the face of a kinsman
} your face from a kinsman;
\par }{\Q Of taking away a portion or a gift;
\par }{\Q And of gazing upon a woman that has a husband;
\par }{\Q \VS{22}Of being over busy with his maid; and come not near her bed;
\par }{\Q Of upbraiding speeches before friends;
\par }{\Q And after you have given, upbraid not;
\par }{\Q \VS{23}Of repeating and speaking what you have heard;
\par }{\Q And of revealing of secrets.
\par }{\Q \VS{24}So shall you be truly shamefast,
\par }{\Q And find favor in the sight of every man.
\par }{\BB \par }
\PoetryChap{42}{\Q \VerseOne{1}Of these things be not ashamed,
\par }{\Q And accept no man’s person to sin
{\ADD{by it:
}}
\par }{\Q \VS{2}Of the law of the Most High, and his covenant;
\par }{\Q And of judgement to do justice to the ungodly;
\par }{\Q \VS{3}Of reckoning with a partner and with travellers;
\par }{\Q And of a gift from the heritage of friends;
\par }{\Q \VS{4}Of exactness of balance and weights;
\par }{\Q And of getting much or little;
\par }{\Q \VS{5}Of indifferent selling of merchants;
\par }{\Q And of much correction of children;
\par }{\Q And of making the side of an evil servant to bleed.
\par }{\Q \VS{6}\FTNT{*}{{\FR 42:6 }
Or,
A seal
} Sure keeping is good, where an evil wife is;
\par }{\Q And where many hands are, shut you close.
\par }{\Q \VS{7}Whatsoever you hand over, let it be by number and weight;
\par }{\Q And in giving and receiving let all be in writing.
\par }{\Q \VS{8}{\ADD{Be not ashamed
}} to instruct the unwise and foolish,
\par }{\Q And one of extreme old age that contends with those that are young;
\par }{\Q And so shall you be well instructed indeed,
\par }{\Q And approved in the sight of every man living.
\par }{\BB \par }{\Q \VS{9}A daughter is a secret cause of wakefulness to a father;
\par }{\Q And the care for her puts away sleep;
\par }{\Q In her youth, lest she pass the flower of her age;
\par }{\Q And when she is married, lest she should be hated:
\par }{\Q \VS{10}In her virginity, lest she should be defiled
\par }{\Q And be with child in her father’s house;
\par }{\Q And when she has a husband, lest she should transgress;
\par }{\Q And when she is married, lest she should be barren.
\par }{\Q \VS{11}Keep a strict watch over a headstrong daughter,
\par }{\Q Lest she make you
\FTNT{†}{{\FR 42:11 }
Or,
a rejoicing
} a laughing stock to your enemies,
\par }{\Q A byword in the city and
\FTNT{‡}{{\FR 42:11 }
Gr.
called forth.
} notorious among the people,
\par }{\Q And shame you before the multitude.
\par }{\BB \par }{\Q \VS{12}Look not upon every body in regard of beauty,
\par }{\Q And sit not in the midst of women;
\par }{\Q \VS{13}For from garments comes a moth,
\par }{\Q And from a woman a woman’s wickedness.
\par }{\Q \VS{14}Better is the wickedness of a man than a pleasant-dealing woman,
\par }{\Q And a woman which puts you to shameful reproach.
\par }{\BB \par }{\Q \VS{15}I will make mention now of the works of the Lord,
\par }{\Q And will declare the things that I have seen:
\par }{\Q In the words of the Lord are his works.
\par }{\Q \VS{16}The sun that gives light looks upon all things;
\par }{\Q And the work of the Lord is full of his glory.
\par }{\Q \VS{17}The Lord has not given power to the saints to declare all his marvelous works;
\par }{\Q Which the Almighty Lord firmly settled,
\par }{\Q That whatever is might be established in his glory.
\par }{\Q \VS{18}He searches out the deep, and the heart,
\par }{\Q And he has understanding of their cunning plans:
\par }{\Q For the Most High knows all knowledge,
\par }{\Q And he looks into the
\FTNT{§}{{\FR 42:18 }
Gr.
sign.
} signs of the world,
\par }{\Q \VS{19}Declaring the things that are past, and the things that shall be,
\par }{\Q And revealing the traces of hidden things.
\par }{\Q \VS{20}No thought escapes him;
\par }{\Q There is not a word hid from him.
\par }{\Q \VS{21}The mighty works of his wisdom he has ordered,
\par }{\Q Who is from everlasting to everlasting:
\par }{\Q Nothing has been added to them, nor diminished from them;
\par }{\Q And he had no need of any counselor.
\par }{\Q \VS{22}How desirable are all his works!
\par }{\Q One may behold
{\ADD{this
}} even to a spark.
\par }{\Q \VS{23}All these things live and remain forever in all manner of uses,
\par }{\Q And they are all obedient.
\par }{\Q \VS{24}All things are double one against another:
\par }{\Q And he has made nothing imperfect.
\par }{\Q \VS{25}One thing establishes the good things of another:
\par }{\Q And who shall be filled with beholding his glory?
\par }{\BB \par }
\PoetryChap{43}{\Q \VerseOne{1}The pride of the height is the firmament
\FTNT{*}{{\FR 43:1 }
Gr.
of clearness.
} in its clearness,
\par }{\Q The appearance of heaven, in the spectacle of its glory.
\par }{\Q \VS{2}The sun when he appears, bringing tidings as he goes forth,
\par }{\Q Is a marvelous instrument, the work of the Most High:
\par }{\Q \VS{3}At his noon he dries up the country,
\par }{\Q And who shall stand against his burning heat?
\par }{\Q \VS{4}A man blowing a furnace is in works of heat,
\par }{\Q {\ADD{But
}} the sun three times more, burning up the mountains:
\par }{\Q Breathing out fiery vapours,
\par }{\Q And sending forth bright beams, he dims the eyes.
\par }{\Q \VS{5}Great is the Lord that made him;
\par }{\Q And at his word he hastens his course.
\par }{\BB \par }{\Q \VS{6}The moon also is in all things for her season,
\par }{\Q For a declaration of times, and a sign of the world.
\par }{\Q \VS{7}From the moon is the sign of the feast day;
\par }{\Q A light that wanes when she is come to the full.
\par }{\Q \VS{8}The month is called after her name,
\par }{\Q Increasing wonderfully in her changing;
\par }{\Q An instrument of the army on high,
\par }{\Q Shining forth in the firmament of heaven;
\par }{\Q \VS{9}The beauty of heaven, the glory of the stars,
\par }{\Q An ornament giving light in the highest places of the Lord.
\par }{\Q \VS{10}At the word of the Holy One they will stand in
\FTNT{†}{{\FR 43:10 }
Gr.
judgement.
} due order,
\par }{\Q And they will not faint in their watches.
\par }{\Q \VS{11}Look upon the rainbow, and praise him that made it;
\par }{\Q Exceeding beautiful in the brightness thereof.
\par }{\Q \VS{12}It encircles the heaven round about with a circle of glory;
\par }{\Q The hands of the Most High have stretched it.
\par }{\BB \par }{\Q \VS{13}By his commandment he makes the snow to fall apace,
\par }{\Q And sends swiftly the lightnings of his judgement.
\par }{\Q \VS{14}By reason thereof the treasure-houses are opened;
\par }{\Q And clouds fly forth as fowls.
\par }{\Q \VS{15}By his mighty power he makes strong the clouds,
\par }{\Q And the hailstones are broken small:
\par }{\Q \VS{16}And at his appearing the mountains will be shaken,
\par }{\Q And at his will the south wind will blow.
\par }{\Q \VS{17}The voice of his thunder makes the earth to travail;
\par }{\Q So does the northern storm and the whirlwind:
\par }{\Q As birds flying down he sprinkles the snow;
\par }{\Q And as the lighting of the locust is the falling down thereof:
\par }{\Q \VS{18}The eye will marvel at the beauty of its whiteness,
\par }{\Q And the heart will be astonished at the raining of it.
\par }{\Q \VS{19}The hoar frost also he pours on the earth as salt;
\par }{\Q And when it is congealed, it is
{\ADD{as
}} points of thorns.
\par }{\BB \par }{\Q \VS{20}The cold north wind shall blow,
\par }{\Q And the ice shall be congealed on the water:
\par }{\Q It shall lodge upon every gathering together of water,
\par }{\Q And the water shall put on as it were a breastplate.
\par }{\Q \VS{21}It shall devour the mountains, and burn up the wilderness,
\par }{\Q And consume the green herb as fire.
\par }{\Q \VS{22}A mist coming speedily is the healing of all things;
\par }{\Q A dew coming after heat shall bring cheerfulness.
\par }{\Q \VS{23}By his counsel he has stilled the deep,
\par }{\Q And
\FTNT{‡}{{\FR 43:23 }
The most ancient authorities read
Jesus planted it.
} planted islands therein.
\par }{\Q \VS{24}They that sail on the sea tell of the danger thereof;
\par }{\Q And when we hear it with our ears, we marvel.
\par }{\Q \VS{25}Therein be also those strange and wondrous works,
\par }{\Q Variety of all that has life, the
\FTNT{§}{{\FR 43:25 }
Gr.
creation.
Several ancient authorities read
possession of cattle.
} race of sea-monsters.
\par }{\Q \VS{26}By reason of him his end has success,
\par }{\Q And by his word all things consist.
\par }{\BB \par }{\Q \VS{27}We may say many things, yet shall we not attain;
\par }{\Q And the sum of our words is, He is all.
\par }{\Q \VS{28}How shall we have strength to glorify him?
\par }{\Q For he is himself the great one above all his works.
\par }{\Q \VS{29}The Lord is terrible and exceedingly great;
\par }{\Q And marvelous is his power.
\par }{\Q \VS{30}When you° glorify the Lord, exalt him as much as you° can;
\par }{\Q For even yet he will exceed:
\par }{\Q And when you° exalt him, put forth your full strength:
\par }{\Q Be not weary; for you° will never attain.
\par }{\Q \VS{31}Who has seen him, that he may declare him?
\par }{\Q And who shall magnify him as he is?
\par }{\Q \VS{32}Many things are hidden greater than these;
\par }{\Q For we have seen but a few of his works.
\par }{\Q \VS{33}For the Lord made all things;
\par }{\Q And to the godly gave he wisdom.
\par }{\BB \par }
\PoetryChap{44}{\Q \VerseOne{1}Let us now praise famous men,
\par }{\Q And our fathers that became the father of us.
\par }{\Q \VS{2}The Lord
\FTNT{*}{{\FR 44:2 }
Gr.
created.
} manifested
{\ADD{in them
}} great glory,
\par }{\Q {\ADD{Even
}} his mighty power from the beginning.
\par }{\Q \VS{3}Such as did bear rule in their kingdoms,
\par }{\Q And were men renowned for their power,
\par }{\Q \FTNT{†}{{\FR 44:3 }
Most authorities read
They shall take counsel.
} Giving counsel by their understanding,
\par }{\Q Such as have brought tidings in prophecies:
\par }{\Q \VS{4}Leaders of the people by their counsels,
\par }{\Q And by their understanding
\FTNT{‡}{{\FR 44:4 }
The Greek text is here corrupt.
}
{\ADD{men of
}} learning for the people;
\par }{\Q Wise
{\ADD{were
}} their words in their instruction:
\par }{\Q \VS{5}Such as sought out musical tunes,
\par }{\Q And set forth verses in writing:
\par }{\Q \VS{6}Rich men furnished with ability,
\par }{\Q Living peaceably in their habitations:
\par }{\Q \VS{7}All these were honored in their generations,
\par }{\Q And were a glory in their days.
\par }{\Q \VS{8}There be of them, that have left a name behind them,
\par }{\Q To declare their praises.
\par }{\Q \VS{9}And some there be, which have no memorial;
\par }{\Q Who are perished as though they had not been,
\par }{\Q And are become as though they had not been born;
\par }{\Q And their children after them.
\par }{\Q \VS{10}But these were men of mercy,
\par }{\Q Whose righteous deeds have not been forgotten.
\par }{\Q \VS{11}With their seed shall remain continually a good inheritance;
\par }{\Q Their children
{\ADD{are
}}
\FTNT{§}{{\FR 44:11 }
Or,
In their testaments
} within the covenants.
\par }{\Q \VS{12}Their seed stands fast,
\par }{\Q And their children for their sakes.
\par }{\Q \VS{13}Their seed shall remain forever,
\par }{\Q And their glory shall not be blotted out.
\par }{\Q \VS{14}Their bodies were buried in peace,
\par }{\Q And their name lives to all generations.
\par }{\Q \VS{15}Peoples will declare their wisdom,
\par }{\Q And the congregation tells out their praise.
\par }{\BB \par }{\Q \VS{16}Enoch pleased the Lord, and was translated,
\par }{\Q {\ADD{Being
}} an example of repentance to all generations.
\par }{\Q \VS{17}Noah was found perfect
{\ADD{and
}} righteous;
\par }{\Q In the season of wrath he was taken in exchange
{\ADD{for the world;
}}
\par }{\Q Therefore was there left a remnant to the earth,
\par }{\Q When the flood came.
\par }{\Q \VS{18}Everlasting covenants were made with him,
\par }{\Q That all flesh should no more be blotted out by a flood.
\par }{\BB \par }{\Q \VS{19}Abraham was a great father of a multitude of nations;
\par }{\Q And there was none found like him in glory;
\par }{\Q \VS{20}Who kept the law of the Most High,
\par }{\Q And was taken into covenant with him:
\par }{\Q In his flesh he established the covenant;
\par }{\Q And when he was proved, he was found faithful.
\par }{\BB \par }{\Q \VS{21}Therefore he assured him by an oath,
\par }{\Q That the nations should be blessed in his seed;
\par }{\Q That he would multiply him as the dust of the earth,
\par }{\Q And exalt his seed as the stars,
\par }{\Q And cause them to inherit from sea to sea,
\par }{\Q And from the River to the utmost part of the earth.
\par }{\Q \VS{22}In Isaac also did he establish likewise, for Abraham his father’s sake,
\par }{\Q The blessing of all men, and the covenant:
\par }{\Q \VS{23}And he made it rest upon the head of Jacob;
\par }{\Q He acknowledged him in his blessings,
\par }{\Q And gave to him by inheritance,
\par }{\Q And divided his portions;
\par }{\Q Among twelve tribes did he part them.
\par }{\BB \par }
\PoetryChap{45}{\Q \VerseOne{1}And he brought out of him a man of mercy,
\par }{\Q Which found favor in the sight of all flesh;
\par }{\Q A man beloved of God and men, even Moses,
\par }{\Q Whose memorial is blessed.
\par }{\Q \VS{2}He made him like to the glory of the saints,
\par }{\Q And magnified him in the fears of his enemies.
\par }{\Q \VS{3}By his words he caused the wonders to cease;
\par }{\Q He glorified him in the sight of kings;
\par }{\Q He gave him commandment for his people,
\par }{\Q And showed him part of his glory.
\par }{\Q \VS{4}He sanctified him in his faithfulness and meekness;
\par }{\Q He chose him out of all flesh.
\par }{\Q \VS{5}He made him to hear his voice,
\par }{\Q And led him into the thick darkness,
\par }{\Q And gave him commandments face to face,
\par }{\Q Even the law of life and knowledge,
\par }{\Q That he might teach Jacob the covenant,
\par }{\Q And Israel his judgments.
\par }{\BB \par }{\Q \VS{6}He exalted Aaron, a holy man like to him,
\par }{\Q Even his brother, of the tribe of Levi.
\par }{\Q \VS{7}He established for him an everlasting covenant,
\par }{\Q And gave him the priesthood of the people;
\par }{\Q He
\FTNT{*}{{\FR 45:7 }
Gr.
blessed.
} beautified him with comely ornaments,
\par }{\Q And girded him about with a robe of glory.
\par }{\Q \VS{8}He clothed him with the perfection of exultation;
\par }{\Q And strengthened him with apparel of
\FTNT{†}{{\FR 45:8 }
Gr.
strength.
} honor,
\par }{\Q The linen trousers, the long robe, and the ephod.
\par }{\Q \VS{9}And he compassed him with pomegranates of gold,
\par }{\Q {\ADD{And
}} with many bells round about,
\par }{\Q To send forth a sound as he went,
\par }{\Q To make a sound that might be heard in the temple,
\par }{\Q For a memorial to the children of his people;
\par }{\Q \VS{10}With a holy garment, with gold and blue and purple, the work of the embroiderer,
\par }{\Q With an oracle of judgement,
{\ADD{even
}} with the Urim and Thummim;
\par }{\Q \VS{11}With twisted scarlet, the work of the craftsman;
\par }{\Q With precious stones graven like a signet, in a setting of gold, the work of the jeweller,
\par }{\Q For a memorial engraved in writing, after the number of the tribes of Israel;
\par }{\Q \VS{12}With a crown of gold upon the mitre,
\FTNT{‡}{{\FR 45:12 }
Gr.
an engraving of a signet of holiness.
} having graven on it, as on a signet, HOLINESS,
\par }{\Q An ornament of honor, a work of might,
\par }{\Q The desires of the eyes, goodly and beautiful.
\par }{\Q \VS{13}Before him there never have been any such;
\par }{\Q No stranger put them on, but his sons only, and his offspring perpertually.
\par }{\Q \VS{14}His sacrifices shall be wholly consumed
\par }{\Q Every day twice continually.
\par }{\Q \VS{15}Moses
\FTNT{§}{{\FR 45:15 }
Gr.
filled his hands.
} consecrated him,
\par }{\Q And anointed him with holy oil:
\par }{\Q It was to him for an everlasting covenant,
\par }{\Q And to his seed, all the days of heaven,
\par }{\Q To minister to him, and to execute also the priest’s office,
\par }{\Q And bless his people in his name.
\par }{\Q \VS{16}He chose him out of all living
\par }{\Q To offer sacrifice to the Lord,
\par }{\Q Incense, and a sweet savor, for a memorial,
\par }{\Q To make reconciliation for your people.
\par }{\Q \VS{17}He gave to him in his commandments,
\par }{\Q {\ADD{Yes,
}} authority in the covenants of judgments,
\par }{\Q To teach Jacob the testimonies,
\par }{\Q And to enlighten Israel in his law.
\par }{\Q \VS{18}Strangers gathered themselves together against him,
\par }{\Q And envied him in the wilderness,
\par }{\Q {\ADD{Even
}} Dathan and Abiram with their company,
\par }{\Q And the congregation of Korah, with wrath and anger.
\par }{\Q \VS{19}The Lord saw it, and it displeased him;
\par }{\Q And in the wrath of his anger they were destroyed:
\par }{\Q He did wonders upon them,
\par }{\Q To consume them with flaming fire.
\par }{\Q \VS{20}And he added glory to Aaron,
\par }{\Q And gave him a heritage:
\par }{\Q He divided to him the first fruits of the increase;
\par }{\Q {\ADD{And
}} first did he prepare bread in abundance:
\par }{\Q \VS{21}For they shall eat the sacrifices of the Lord,
\par }{\Q Which he gave to him and to his seed.
\par }{\Q \VS{22}However in the land of the people he shall have no inheritance,
\par }{\Q And he has no portion among the people:
\par }{\Q For he himself is your portion
{\ADD{and
}} inheritance.
\par }{\BB \par }{\Q \VS{23}And Phinehas the son of Eleazar is the third in glory,
\par }{\Q In that he was zealous in the fear of the Lord,
\par }{\Q And stood fast in the good forwardness of his soul when the people turned away,
\par }{\Q And he made reconciliation for Israel.
\par }{\Q \VS{24}Therefore was there a covenant of peace established for him,
\par }{\Q {\ADD{That he should be
}} leader of the
\FTNT{*}{{\FR 45:24 }
Or,
sanctuary
} saints and of his people;
\par }{\Q That he and his seed
\par }{\Q Should have the dignity of the priesthood forever.
\par }{\Q \VS{25}Also
{\ADD{he made
}} a covenant with David the son of Jesse, of the tribe of Judah;
\par }{\Q The inheritance of the king is his alone from son to son;
\par }{\Q So the inheritance of Aaron is also to his seed.
\par }{\Q \VS{26}{\ADD{God
}} give you wisdom in your heart
\par }{\Q To judge his people in righteousness,
\par }{\Q That their good things be not abolished,
\par }{\Q And
{\ADD{that
}} their glory
{\ADD{endure
}} for all their generations.
\par }{\BB \par }
\PoetryChap{46}{\Q \VerseOne{1}Joshua the son of
\FTNT{*}{{\FR 46:1 }
Gr.
Nave.
} Nun was valiant in war,
\par }{\Q And was the successor of Moses in prophecies:
\par }{\Q Who according to his name was made great
\par }{\Q For the saving of
\FTNT{†}{{\FR 46:1 }
Gr.
his.
} God’s elect,
\par }{\Q To take vengeance of the enemies that rose up against them,
\par }{\Q That he might give Israel their inheritance.
\par }{\Q \VS{2}How was he glorified in the lifting up his hands,
\par }{\Q And in stretching out his sword against the cities!
\par }{\Q \VS{3}Who before him so stood fast?
\par }{\Q For the Lord himself brought his enemies to him.
\par }{\Q \VS{4}Did not the sun go back by his hand?
\par }{\Q And didn’t one day become as two?
\par }{\Q \VS{5}He called upon the Most High
{\ADD{and
}} Mighty One,
\par }{\Q When his foes pressed him round about;
\par }{\Q And the great Lord heard him.
\par }{\Q \VS{6}With hailstones of mighty power
\par }{\Q He caused war to break violently upon the nation,
\par }{\Q And
\FTNT{‡}{{\FR 46:6 }
See Joshua 10:11
} in the going down he destroyed those who resisted;
\par }{\Q That the nations might know his
\FTNT{§}{{\FR 46:6 }
Gr.
panoply.
} armor,
\par }{\Q How that he fought in the sight of the Lord;
\par }{\Q For he followed after the Mighty One.
\par }{\Q \VS{7}Also in the time of Moses he did a work of mercy,
\par }{\Q He and Caleb the son of Jephunneh,
\par }{\Q In that they withstood the adversary,
\par }{\Q Hindered the people from sin,
\par }{\Q And stilled the murmuring of wickedness.
\par }{\Q \VS{8}And of six hundred thousand people on foot, they two alone were preserved
\par }{\Q To bring them into the heritage,
\par }{\Q Even into a land flowing with milk and honey.
\par }{\Q \VS{9}Also the Lord gave strength to Caleb,
\par }{\Q And it remained with him to his old age;
\par }{\Q So that he entered upon the height of the land,
\par }{\Q And his seed obtained it for a heritage:
\par }{\Q \VS{10}That all the children of Israel might see
\par }{\Q That it is good to walk after the Lord.
\par }{\BB \par }{\Q \VS{11}Also the judges, every one by his name,
\par }{\Q All whose hearts went not a whoring,
\par }{\Q And who turned not away from the Lord,
\par }{\Q May their memorial be blessed.
\par }{\Q \VS{12}May their bones flourish again out of their place,
\par }{\Q And may the name of those who have been honored be renewed upon their children.
\par }{\BB \par }{\Q \VS{13}Samuel, the prophet of the Lord, beloved of his Lord,
\par }{\Q Established a kingdom, and anointed princes over his people.
\par }{\Q \VS{14}By the law of the Lord he judged the congregation,
\par }{\Q And the Lord visited Jacob.
\par }{\Q \VS{15}By his faithfulness he was proved to be a prophet,
\par }{\Q And by his words he was known to be faithful in vision.
\par }{\Q \VS{16}Also when his enemies pressed him round about
\par }{\Q He called upon the Lord, the Mighty One,
\par }{\Q With the offering of the sucking lamb.
\par }{\Q \VS{17}And the Lord thundered from heaven,
\par }{\Q And with a mighty sound made his voice to be heard.
\par }{\Q \VS{18}And he utterly destroyed the rulers of the Tyrians,
\par }{\Q And all the princes of the Philistines.
\par }{\Q \VS{19}Also before the time of his
\FTNT{*}{{\FR 46:19 }
Gr.
age-long sleep.
} long sleep
\par }{\Q He made protestations in the sight of the lord and
{\ADD{his
}} anointed,
\par }{\Q I have not taken any man’s goods, so much as a sandal:
\par }{\Q And no man did accuse him.
\par }{\Q \VS{20}And after he fell asleep he prophesied,
\par }{\Q And showed the king his end,
\par }{\Q And lifted up his voice from the earth in prophecy,
\par }{\Q To blot out the wickedness of the people.
\par }{\BB \par }
\PoetryChap{47}{\Q \VerseOne{1}And after him rose up Nathan
\par }{\Q To prophesy in the days of David.
\par }{\Q \VS{2}As is the fat when it is separated from the peace offering,
\par }{\Q So was David
{\ADD{separated
}} from the children of Israel.
\par }{\Q \VS{3}He played with lions as with kids,
\par }{\Q And with bears as with lambs of the flock.
\par }{\Q \VS{4}In his youth did he not kill a giant,
\par }{\Q And take away reproach from the people,
\par }{\Q When he lifted up his hand with a sling stone,
\par }{\Q And beat down the boasting of Goliath?
\par }{\Q \VS{5}For he called upon the Most High Lord;
\par }{\Q And he gave him strength in his right hand,
\par }{\Q To kill a man mighty in war,
\par }{\Q To exalt the horn of his people.
\par }{\Q \VS{6}So they glorified him for
{\ADD{his
}} ten thousands,
\par }{\Q And praised him for the blessings of the Lord,
\par }{\Q In that there was given him a diadem of glory.
\par }{\Q \VS{7}For he destroyed the enemies on every side,
\par }{\Q And brought to nothing the Philistines his adversaries,
\par }{\Q Brake their horn in pieces to this day.
\par }{\Q \VS{8}In every work of his he gave thanks to the Holy One Most High with words of glory;
\par }{\Q With his whole heart he sang praise,
\par }{\Q And loved him that made him.
\par }{\Q \VS{9}Also he set singers before the altar,
\par }{\Q And to make sweet melody by their music.
\FTNT{*}{{\FR 47:9 }
The remainder of this verse is omitted by the best authorities.
}
\par }{\Q \VS{10}He gave beauty to the feasts,
\par }{\Q And set in order the seasons to perfection,
\par }{\Q While they praised his holy name,
\par }{\Q And the sanctuary sounded from early morning.
\par }{\Q \VS{11}The Lord took away his sins,
\par }{\Q And exalted his horn forever;
\par }{\Q And gave him a covenant of kings,
\par }{\Q And a throne of glory in Israel.
\par }{\Q \VS{12}After him rose up a son, a man of understanding;
\par }{\Q And for his sake he lived at large.
\par }{\Q \VS{13}Solomon reigned in days of peace;
\par }{\Q And to him God gave rest round about,
\par }{\Q That he might set up a house for his name,
\par }{\Q And prepare a sanctuary forever.
\par }{\Q \VS{14}How wise you were made in your youth,
\par }{\Q And filled as a river with understanding!
\par }{\Q \VS{15}Your soul covered the earth,
\par }{\Q And you filled it with
\FTNT{†}{{\FR 47:15 }
Gr.
parables of riddles.
} dark parables.
\par }{\Q \VS{16}Your name reached to the isles afar off;
\par }{\Q And for your peace you were beloved.
\par }{\Q \VS{17}For your songs and proverbs and parables,
\par }{\Q And for your interpretations, the countries marveled at you.
\par }{\Q \VS{18}By the name of the Lord God,
\par }{\Q Which is called the God of Israel,
\par }{\Q You did gather gold as tin,
\par }{\Q And did multiply silver as lead.
\par }{\Q \VS{19}You did bow your loins to women,
\par }{\Q And in your body you were brought into subjection.
\par }{\Q \VS{20}You did blemish your honor,
\par }{\Q And profane your seed,
\par }{\Q To bring wrath upon your children;
\par }{\Q And I was grieved for your folly:
\par }{\Q \VS{21}So that the sovereignty was divided,
\par }{\Q And out of Ephraim ruled a disobedient kingdom.
\par }{\Q \VS{22}But the Lord will never forsake his mercy;
\par }{\Q And he will not destroy any of his works,
\par }{\Q Nor blot out the posterity of his elect;
\par }{\Q And the seed of him that loved him he will not take away;
\par }{\Q And he gave a remnant to Jacob,
\par }{\Q And to David a root out of him.
\par }{\BB \par }{\Q \VS{23}And
{\ADD{so
}} rested Solomon with his fathers;
\par }{\Q And of his seed he left behind him Rehoboam,
\par }{\Q {\ADD{Even
}} the foolishness of the people, and one that lacked understanding,
\par }{\Q Who made the people to revolt by his counsel.
\par }{\Q Also Jeroboam the son of Nebat,
\par }{\Q Who made Israel to sin,
\par }{\Q And gave to Ephraim a way of sin.
\par }{\Q \VS{24}And their sins were multiplied exceedingly,
\par }{\Q To remove them from their land.
\par }{\Q \VS{25}For they sought out all manner of wickedness,
\par }{\Q Till vengeance should come upon them.
\par }{\BB \par }
\PoetryChap{48}{\Q \VerseOne{1}Also there arose Elijah the prophet as fire,
\par }{\Q And his word burned like a torch:
\par }{\Q \VS{2}Who brought a famine upon them,
\par }{\Q And by his zeal made them few in number.
\par }{\Q \VS{3}By the word of the Lord he shut up the heaven:
\par }{\Q Thrice did he thus bring down fire.
\par }{\Q \VS{4}How you were glorified, O Elijah, in your wondrous deeds!
\par }{\Q And who shall glory like to you?
\par }{\Q \VS{5}Who did raise up a dead man from death,
\par }{\Q And from
\FTNT{*}{{\FR 48:5 }
Gr.
Hades.
} the place of the dead, by the word of the Most High:
\par }{\Q \VS{6}Who brought down kings to destruction,
\par }{\Q And honorable men from their bed:
\par }{\Q \VS{7}Who heard rebuke in Sinai,
\par }{\Q And judgments of vengeance in Horeb:
\par }{\Q \VS{8}Who anointed kings for retribution,
\par }{\Q And prophets to succeed after him:
\par }{\Q \VS{9}Who was taken up in a tempest of fire,
\par }{\Q In a chariot of fiery horses:
\par }{\Q \VS{10}Who was recorded for reproofs in their seasons,
\par }{\Q To pacify anger, before it brake forth into wrath;
\par }{\Q To turn the heart of the father to the son,
\par }{\Q And to restore the tribes of Jacob.
\par }{\Q \VS{11}Blessed are those who saw you,
\par }{\Q And those who have been beautified with love;
\par }{\Q For we also shall surely live.
\par }{\BB \par }{\Q \VS{12}Elijah
{\ADD{it was,
}} who was wrapped in a tempest:
\par }{\Q And Elisha was filled with his spirit;
\par }{\Q And in
{\ADD{all
}} his days he was not moved by
{\ADD{the fear of
}} any ruler,
\par }{\Q And no one brought him into subjection.
\par }{\Q \VS{13}\FTNT{†}{{\FR 48:13 }
Or,
No man’s word overcame him
} Nothing was too high for him;
\par }{\Q And when he was laid on sleep his body prophesied.
\par }{\Q \VS{14}As in his life he did wonders,
\par }{\Q So in death were his works marvelous.
\par }{\Q \VS{15}For all this the people repented not,
\par }{\Q And they departed not from their sins,
\par }{\Q Till they were carried away as a plunder from their land,
\par }{\Q And were scattered through all the earth;
\par }{\Q And the people was left very few in number,
\par }{\Q And a ruler
{\ADD{was left
}} in the house of David.
\par }{\Q \VS{16}Some of them did that which was pleasing
{\ADD{to God,
}}
\par }{\Q And some multiplied sins.
\par }{\BB \par }{\Q \VS{17}Hezekiah fortified his city,
\par }{\Q And brought in
\FTNT{‡}{{\FR 48:17 }
Some ancient authorities, apparently by a confusion, read
Gog.
} water into the midst of them:
\par }{\Q He digged the sheer rock with iron,
\par }{\Q And built up wells for waters.
\par }{\Q \VS{18}In his days Sennacherib came up,
\par }{\Q And sent Rabshakeh, and departed;
\par }{\Q And he lifted up his hand against Sion,
\par }{\Q And boasted great things in his arrogance.
\par }{\Q \VS{19}Then were their hearts and their hands shaken,
\par }{\Q And they were in pain, as women in travail;
\par }{\Q \VS{20}And they called upon the Lord which is merciful,
\par }{\Q Spreading forth their hands to him:
\par }{\Q And the Holy One heard them speedily out of Heaven,
\par }{\Q And delivered them by the hand of Isaiah.
\par }{\Q \VS{21}He struck the camp of the Assyrians,
\par }{\Q And his angel utterly destroyed them.
\par }{\Q \VS{22}For Hezekiah did that which was pleasing to the Lord,
\par }{\Q And was strong in the ways of David his father,
\par }{\Q Which Isaiah the prophet commanded,
\par }{\Q Who was great and faithful in his vision.
\par }{\Q \VS{23}In his days the sun went backward;
\par }{\Q And he added life to the king.
\par }{\Q \VS{24}He saw by an excellent spirit what should come to pass at the last;
\par }{\Q And he comforted those who mourned in Sion.
\par }{\Q \VS{25}He showed the things that should be to the end of time,
\par }{\Q And the hidden things or ever they came.
\par }{\BB \par }
\PoetryChap{49}{\Q \VerseOne{1}The memorial of Josiah is like the composition of incense
\par }{\Q Prepared by the work of the apothecary:
\par }{\Q It shall be sweet as honey in every mouth,
\par }{\Q And as music at a banquet of wine.
\par }{\Q \VS{2}He behaved himself uprightly in the conversion of the people,
\par }{\Q And took away the abominations of iniquity.
\par }{\Q \VS{3}He set his heart right toward the Lord;
\par }{\Q In the days of wicked men he made godliness to prevail.
\par }{\BB \par }{\Q \VS{4}Except David and Hezekiah and Josiah,
\par }{\Q All committed trespass:
\par }{\Q For they forsook the law of the Most High;
\par }{\Q The kings of Judah failed.
\par }{\Q \VS{5}For they gave their
\FTNT{*}{{\FR 49:5 }
Gr.
horn.
} power to others,
\par }{\Q And their glory to a strange nation.
\par }{\Q \VS{6}They set on fire the chosen city of the sanctuary,
\par }{\Q And made her streets desolate,
{\ADD{as it was written
}} by the hand of Jeremiah.
\par }{\Q \VS{7}For they mistreated him;
\par }{\Q And yet he was sanctified in the womb to be a prophet,
\par }{\Q To root out, and to afflict, and to destroy;
\par }{\Q {\ADD{And
}} in like manner to build and to plant.
\par }{\Q \VS{8}{\ADD{It was
}} Ezekiel who saw the vision of glory,
\par }{\Q Which
{\ADD{God
}} showed him upon the chariot of the cherubim.
\par }{\Q \VS{9}For truly he remembered the enemies in
\FTNT{†}{{\FR 49:9 }
Gr.
rain.
} storm,
\par }{\Q And to do good to those who directed their ways aright.
\par }{\Q \VS{10}Also of the twelve prophets
\FTNT{‡}{{\FR 49:10 }
The remainder of this line is omitted by the best authorities.
}
\par }{\Q May the bones flourish again out of their place.
\par }{\Q And he comforted Jacob,
\par }{\Q And delivered them by confidence of hope.
\par }{\BB \par }{\Q \VS{11}How shall we magnify Zerubbabel?
\par }{\Q And he was as a signet on the right hand:
\par }{\Q \VS{12}So was Jesus the son of Josedek:
\par }{\Q Who in their days built the house,
\par }{\Q And exalted a
\FTNT{§}{{\FR 49:12 }
Some ancient authorities read
temple.
} people holy to the Lord,
\par }{\Q Prepared for everlasting glory.
\par }{\Q \VS{13}Also of Nehemiah the memorial is great;
\par }{\Q Who raised up for us the walls that were fallen,
\par }{\Q And set up the gates and bars,
\par }{\Q And raised up our homes again.
\par }{\BB \par }{\Q \VS{14}No man was created upon the earth such as was Enoch;
\par }{\Q For he was taken up from the earth.
\par }{\Q \VS{15}Neither was there a man born like to Joseph,
\par }{\Q A governor of his kindred, a stay of the people:
\par }{\Q Yes, his bones were visited.
\par }{\Q \VS{16}Shem and Seth were glorified among men;
\par }{\Q And above every living thing in the creation is Adam.
\par }{\BB \par }
\PoetryChap{50}{\Q \VerseOne{1}{\ADD{It was
}} Simon, the son of Onias, the
\FTNT{*}{{\FR 50:1 }
Or,
high
} great priest,
\par }{\Q Who in his life repaired the house,
\par }{\Q And in his days strengthened the temple:
\par }{\Q \VS{2}And by him was built from the foundation the height of the double
{\ADD{wall,
}}
\par }{\Q The lofty underworks of the inclosure of the temple:
\par }{\Q \VS{3}In his days the cistern of waters was
\FTNT{†}{{\FR 50:3 }
The text here seems to be corrupt.
} diminished,
\par }{\Q The brazen vessel in compass as the sea.
\par }{\Q \VS{4}{\ADD{It was
}} he that took thought for his people that they should not fall,
\par }{\Q And fortified the city
\FTNT{‡}{{\FR 50:4 }
Gr.
to beseige:
that is,
as to besieging it.
} against besieging:
\par }{\Q \VS{5}How glorious was he when the people gathered round him
\par }{\Q At his coming forth out of the
\FTNT{§}{{\FR 50:5 }
Gr.
house of the veil.
} sanctuary!
\par }{\Q \VS{6}As the morning star in the midst of a cloud,
\par }{\Q As the moon at the full:
\par }{\Q \VS{7}As the sun shining forth upon the temple of the Most High,
\par }{\Q And as the rainbow giving light in clouds of glory:
\par }{\Q \VS{8}As the flower of roses in the days of new
{\ADD{fruits,
}}
\par }{\Q As lilies at the water spring,
\par }{\Q As the shoot of the frankincense tree in the time of summer:
\par }{\Q \VS{9}As fire and incense in the censer,
\par }{\Q As a vessel all of beaten gold
\par }{\Q Adorned with all manner of precious stones:
\par }{\Q \VS{10}As an olive tree budding forth fruits,
\par }{\Q And as a cypress growing high among the clouds.
\par }{\Q \VS{11}When he took up the robe of glory,
\par }{\Q And put on the perfection of exultation,
\par }{\Q In the ascent of the holy altar,
\par }{\Q He made glorious the precinct of the sanctuary.
\par }{\Q \VS{12}And when he received the portions out of the priests’ hands,
\par }{\Q Himself also standing by the hearth of the altar,
\par }{\Q His kindred as a garland round about him,
\par }{\Q He was as a young cedar in Libanus;
\par }{\Q And as stems of palm trees compassed they him round about,
\par }{\Q \VS{13}And all the sons of Aaron in their glory,
\par }{\Q And the Lord’s offering in their hands, before all the congregation of Israel.
\par }{\Q \VS{14}And finishing the service at the altars,
\par }{\Q That he might adorn the offering of the Most High, the Almighty,
\par }{\Q \VS{15}He stretched out his hand to the
\FTNT{*}{{\FR 50:15 }
Gr.
cup of libation.
} cup,
\par }{\Q And poured out the cup of the grape;
\par }{\Q He poured out at the foot of the altar
\par }{\Q A sweet smelling savor to the Most High, the King of all.
\par }{\Q \VS{16}Then shouted the sons of Aaron,
\par }{\Q They sounded the trumpets of beaten work,
\par }{\Q They made a great noise to be heard,
\par }{\Q For a remembrance before the Most High.
\par }{\Q \VS{17}Then all the people together hurried,
\par }{\Q And fell down upon the earth on their faces
\par }{\Q To worship their Lord, the Almighty, God Most High.
\par }{\Q \VS{18}The singers also praised him with their voices;
\par }{\Q In the whole house was there made sweet melody.
\par }{\Q \VS{19}And the people implored the Lord Most High,
\par }{\Q In prayer before him that is merciful.
\par }{\Q Till the
\FTNT{†}{{\FR 50:19 }
Gr.
adornment.
} worship of the Lord should be ended;
\par }{\Q And so they accomplished his service.
\par }{\Q \VS{20}Then he went down, and lifted up his hands
\par }{\Q Over the whole congregation of the children of Israel,
\par }{\Q To give blessing to the Lord with his lips,
\par }{\Q And to glory in his name.
\par }{\Q \VS{21}And he bowed himself down in worship the second time,
\par }{\Q To declare the blessing from the Most High.
\par }{\BB \par }{\Q \VS{22}And now bless the God of all,
\par }{\QB who everywhere does great things,
\par }{\QB who exalts our days from the womb,
\par }{\QB and deals with us according to his mercy.
\par }{\Q \VS{23}May he grant us joyfulness of heart,
\par }{\Q And that peace may be in our days in Israel for the days of eternity:
\par }{\Q \VS{24}To entrust his mercy with us;
\par }{\Q And let him deliver us in his time!
\par }{\BB \par }{\Q \VS{25}With two nations is my soul vexed,
\par }{\Q And the third is no nation:
\par }{\Q \VS{26}They that sit upon the mountain of
\FTNT{‡}{{\FR 50:26 }
According to some ancient versions,
Seir.
} Samaria,
{\ADD{and
}} the Philistines,
\par }{\Q And that foolish people that dwells in Sichem.
\par }{\BB \par }{\Q \VS{27}I have
\FTNT{§}{{\FR 50:27 }
Gr.
graven.
} written in this book the instruction of understanding and knowledge,
\par }{\Q I Jesus, the son of Sirach Eleazar, of Jerusalem,
\par }{\Q Who out of his heart poured forth wisdom.
\par }{\Q \VS{28}Blessed is he that shall be exercised in these things;
\par }{\Q And he that lays them up in his heart shall become wise.
\par }{\Q \VS{29}For if he do them, he shall be strong to all things:
\par }{\Q For the light of the Lord is his
\FTNT{*}{{\FR 50:29 }
Gr.
footstep.
} guide.
\FTNT{†}{{\FR 50:29 }
The remainder of this verse is omitted by the best authorities.
}

\par }\Chap{51}{\PP A Prayer of Jesus the son of Sirach.
\par }{\Q \VerseOne{1}I will give thanks to you, O Lord, O King,
\par }{\Q And will praise you, O God my Savior:
\par }{\Q I do give thanks to your name:
\par }{\Q \VS{2}For you were my protector and helper,
\par }{\Q And did deliver my body out of destruction,
\par }{\Q And out of the snare of a slanderous tongue,
\par }{\Q From lips that forge lies,
\par }{\Q And was my helper before those who stood by;
\par }{\Q \VS{3}And did deliver me, according to the abundance of your mercy, and
{\ADD{greatness
}} of your name,
\par }{\Q From the gnashings
{\ADD{of teeth
}} ready to devour,
\par }{\Q Out of the hand of such as sought my life,
\par }{\Q Out of the manifold afflictions which I had;
\par }{\Q \VS{4}From the choking of a fire on every side,
\par }{\Q And out of the midst of fire which I kindled not;
\par }{\Q \VS{5}Out of the depth of the belly of
\FTNT{*}{{\FR 51:5 }
Gr.
Hades.
} the grave,
\par }{\Q And from an unclean tongue,
\par }{\Q And from lying words,
\par }{\Q \VS{6}The slander of an unrighteous tongue to the king.
\par }{\Q My soul drew near even to death,
\par }{\Q And my life was near to
\FTNT{†}{{\FR 51:6 }
Gr.
Hades
} the grave beneath.
\par }{\Q \VS{7}They compassed me on every side,
\par }{\Q And there was none to help
{\ADD{me.
}}
\par }{\Q {\ADD{I was
}} looking for the help of men,
\par }{\Q And it was not.
\par }{\Q \VS{8}And I remembered your mercy, O Lord,
\par }{\Q And your working which has been from everlasting,
\par }{\Q How you deliver those who wait for you,
\par }{\Q And save them out of the hand of the enemies.
\par }{\Q \VS{9}And I lifted up my supplication from the earth,
\par }{\Q And prayed for deliverance from death.
\par }{\Q \VS{10}I called upon the Lord, the Father of my Lord,
\par }{\Q That he would not forsake me in the days of affliction,
\par }{\Q In the time when there was no help against the proud.
\par }{\Q \VS{11}I will praise your name continually,
\par }{\Q And will sing praise with thanksgiving;
\par }{\Q And my supplication was heard:
\par }{\Q \VS{12}For you saved me from destruction,
\par }{\Q And delivered me from the evil time:
\par }{\Q Therefore I will give thanks and praise to you,
\par }{\Q And bless the name of the Lord.
\par }{\BB \par }{\Q \VS{13}When I was yet young,
\par }{\Q Or ever I went abroad,
\par }{\Q I sought wisdom openly in my prayer.
\par }{\Q \VS{14}Before the temple I asked for her,
\par }{\Q And I will seek her out even to the end.
\par }{\Q \VS{15}From
{\ADD{her
}} flower as from the ripening grape my heart delighted in her:
\par }{\Q My foot trod in uprightness,
\par }{\Q From my youth I tracked her out.
\par }{\Q \VS{16}I bowed down my ear a little, and received her,
\par }{\Q And found for myself much instruction.
\par }{\Q \VS{17}I profited in her:
\par }{\Q To him that gives me wisdom I will give glory.
\par }{\Q \VS{18}For I purposed to practice her,
\par }{\Q And I was zealous for that which is good;
\par }{\Q And I shall never be put to shame.
\par }{\Q \VS{19}My soul has wrestled in her,
\par }{\Q And in my doing I was exact:
\par }{\Q I spread forth my hands to the heaven above,
\par }{\Q And bewailed my ignorances of her.
\par }{\Q \VS{20}I set my soul aright to her,
\par }{\Q And in pureness I found her.
\par }{\Q I got me a heart
{\ADD{joined
}} with her from the beginning:
\par }{\Q Therefore shall I not be forsaken.
\par }{\Q \VS{21}My inward part also was troubled to seek her:
\par }{\Q Therefore I have gotten a good possession.
\par }{\Q \VS{22}The Lord gave me a tongue for my reward;
\par }{\Q And I will praise him therewith.
\par }{\BB \par }{\Q \VS{23}Draw near to me, you° unlearned,
\par }{\Q And lodge in the house of instruction.
\par }{\Q \VS{24}Say, wherefore are you° lacking in these things,
\par }{\Q And your souls are very thirsty?
\par }{\Q \VS{25}I opened my mouth, and spoke,
\par }{\Q Get her for yourselves without money.
\par }{\Q \VS{26}Put your neck under the yoke,
\par }{\Q And let your soul receive instruction:
\par }{\Q She is hard at hand to find.
\par }{\Q \VS{27}Behold with your eyes,
\par }{\Q How that I labored but a little,
\par }{\Q And found for myself much rest.
\par }{\Q \VS{28}Get you instruction with a great sum of silver,
\par }{\Q And gain much gold by her.
\par }{\Q \VS{29}May your soul rejoice in his mercy,
\par }{\Q And may you° not be put to shame in praising him.
\par }{\Q \VS{30}Work your work before the time comes,
\par }{\Q And in his time he will give you your reward.
\par }