\NormalFont\ShortTitle{Song of Solomon}
{\MT The Song of Solomon

\par }\ChapOne{1}{\PP \VerseOne{1}The Song of songs, which is Solomon’s.
\par }{\SP Beloved
\par }{\Q \VS{2}Let him kiss me with the kisses of his mouth;
\par }{\QB for your love is better than wine.
\par }{\Q \VS{3}Your oils have a pleasing fragrance.
\par }{\QB Your name is oil poured out,
\par }{\QB therefore the virgins love you.
\par }{\Q \VS{4}Take me away with you.
\par }{\QB Let’s hurry.
\par }{\QB The king has brought me into his rooms.
\par }{\SP Friends
\par }{\Q We will be glad and rejoice in you.
\par }{\QB We will praise your love more than wine!
\par }{\SP Beloved
\par }{\Q They are right to love you.
\par }{\Q \VS{5}I am dark, but lovely,
\par }{\QB you daughters of Jerusalem,
\par }{\QB like Kedar’s tents,
\par }{\QB like Solomon’s curtains.
\par }{\Q \VS{6}Don’t stare at me because I am dark,
\par }{\QB because the sun has scorched me.
\par }{\Q My mother’s sons were angry with me.
\par }{\QB They made me keeper of the vineyards.
\par }{\QB I haven’t kept my own vineyard.
\par }{\Q \VS{7}Tell me, you whom my soul loves,
\par }{\QB where you graze your flock,
\par }{\QB where you rest them at noon;
\par }{\QB for why should I be as one who is veiled
\par }{\QB beside the flocks of your companions?
\par }{\SP Lover
\par }{\Q \VS{8}If you don’t know, most beautiful among women,
\par }{\QB follow the tracks of the sheep.
\par }{\QB Graze your young goats beside the shepherds’ tents.
\par }{\BB \par }{\Q \VS{9}I have compared you, my love,
\par }{\QB to a steed in Pharaoh’s chariots.
\par }{\Q \VS{10}Your cheeks are beautiful with earrings,
\par }{\QB your neck with strings of jewels.
\par }{\SP Friends
\par }{\Q \VS{11}We will make you earrings of gold,
\par }{\QB with studs of silver.
\par }{\SP Beloved
\par }{\Q \VS{12}While the king sat at his table,
\par }{\QB my perfume spread its fragrance.
\par }{\Q \VS{13}My beloved is to me a sachet of myrrh,
\par }{\QB that lies between my breasts.
\par }{\Q \VS{14}My beloved is to me a cluster of henna blossoms
\par }{\QB from the vineyards of En Gedi.
\par }{\SP Lover
\par }{\Q \VS{15}Behold,\FTNT{*}{{\FR 1:15 }
“Behold”, from “הִנֵּה”, means look at, take notice, observe, see, or gaze at. It is often used as an interjection.} you are beautiful, my love.
\par }{\QB Behold, you are beautiful.
\par }{\QB Your eyes are like doves.
\par }{\SP Beloved
\par }{\Q \VS{16}Behold, you are beautiful, my beloved, yes, pleasant;
\par }{\QB and our couch is verdant.
\par }{\SP Lover
\par }{\Q \VS{17}The beams of our house are cedars.
\par }{\QB Our rafters are firs.

\par }\Chap{2}{\SP Beloved
\par }{\Q \VerseOne{1}I am a rose of Sharon,
\par }{\QB a lily of the valleys.
\par }{\SP Lover
\par }{\Q \VS{2}As a lily among thorns,
\par }{\QB so is my love among the daughters.
\par }{\SP Beloved
\par }{\Q \VS{3}As the apple tree among the trees of the wood,
\par }{\QB so is my beloved among the sons.
\par }{\Q I sat down under his shadow with great delight,
\par }{\QB his fruit was sweet to my taste.
\par }{\Q \VS{4}He brought me to the banquet hall.
\par }{\QB His banner over me is love.
\par }{\Q \VS{5}Strengthen me with raisins,
\par }{\QB refresh me with apples;
\par }{\QB for I am faint with love.
\par }{\Q \VS{6}His left hand is under my head.
\par }{\QB His right hand embraces me.
\par }{\Q \VS{7}I adjure you, daughters of Jerusalem,
\par }{\QB by the roes, or by the hinds of the field,
\par }{\QB that you not stir up, nor awaken love,
\par }{\QB until it so desires.
\par }{\BB \par }{\Q \VS{8}The voice of my beloved!
\par }{\QB Behold, he comes,
\par }{\QB leaping on the mountains,
\par }{\QB skipping on the hills.
\par }{\Q \VS{9}My beloved is like a roe or a young deer.
\par }{\QB Behold, he stands behind our wall!
\par }{\Q He looks in at the windows.
\par }{\QB He glances through the lattice.
\par }{\BB \par }{\Q \VS{10}My beloved spoke, and said to me,
\par }{\QB “Rise up, my love, my beautiful one, and come away.
\par }{\Q \VS{11}For behold, the winter is past.
\par }{\QB The rain is over and gone.
\par }{\Q \VS{12}The flowers appear on the earth.
\par }{\QB The time of the singing has come,
\par }{\QB and the voice of the turtledove is heard in our land.
\par }{\Q \VS{13}The fig tree ripens her green figs.
\par }{\QB The vines are in blossom.
\par }{\QB They give out their fragrance.
\par }{\Q Arise, my love, my beautiful one,
\par }{\QB and come away.”
\par }{\SP Lover
\par }{\Q \VS{14}My dove in the clefts of the rock,
\par }{\QB in the hiding places of the mountainside,
\par }{\QB let me see your face.
\par }{\QB let me hear your voice;
\par }{\QB for your voice is sweet and your face is lovely.
\par }{\Q \VS{15}Catch for us the foxes,
\par }{\QB the little foxes that plunder the vineyards;
\par }{\QB for our vineyards are in blossom.
\par }{\SP Beloved
\par }{\Q \VS{16}My beloved is mine, and I am his.
\par }{\QB He browses among the lilies.
\par }{\Q \VS{17}Until the day is cool, and the shadows flee away,
\par }{\QB turn, my beloved,
\par }{\QB and be like a roe or a young deer on the mountains of Bether.

\par }\Chap{3}{\PP \VerseOne{1}By night on my bed,
\par }{\QB I sought him whom my soul loves.
\par }{\QB I sought him, but I didn’t find him.
\par }{\Q \VS{2}I will get up now, and go about the city;
\par }{\QB in the streets and in the squares I will seek him whom my soul loves.
\par }{\QB I sought him, but I didn’t find him.
\par }{\Q \VS{3}The watchmen who go about the city found me;
\par }{\QB “Have you seen him whom my soul loves?”
\par }{\Q \VS{4}I had scarcely passed from them,
\par }{\QB when I found him whom my soul loves.
\par }{\Q I held him, and would not let him go,
\par }{\QB until I had brought him into my mother’s house,
\par }{\QB into the room of her who conceived me.
\par }{\BB \par }{\Q \VS{5}I adjure you, daughters of Jerusalem,
\par }{\QB by the roes, or by the hinds of the field,
\par }{\QB that you not stir up nor awaken love,
\par }{\QB until it so desires.
\par }{\BB \par }{\Q \VS{6}Who is this who comes up from the wilderness like pillars of smoke,
\par }{\QB perfumed with myrrh and frankincense,
\par }{\QB with all spices of the merchant?
\par }{\Q \VS{7}Behold, it is Solomon’s carriage!
\par }{\QB Sixty mighty men are around it,
\par }{\QB of the mighty men of Israel.
\par }{\Q \VS{8}They all handle the sword, and are expert in war.
\par }{\QB Every man has his sword on his thigh,
\par }{\QB because of fear in the night.
\par }{\BB \par }{\Q \VS{9}King Solomon made himself a carriage
\par }{\QB of the wood of Lebanon.
\par }{\Q \VS{10}He made its pillars of silver,
\par }{\QB its bottom of gold, its seat of purple,
\par }{\QB the middle of it being paved with love,
\par }{\QB from the daughters of Jerusalem.
\par }{\Q \VS{11}Go out, you daughters of Zion, and see king Solomon,
\par }{\QB with the crown with which his mother has crowned him,
\par }{\QB in the day of his weddings,
\par }{\QB in the day of the gladness of his heart.

\par }\Chap{4}{\SP Lover
\par }{\PP \VerseOne{1}Behold, you are beautiful, my love.
\par }{\QB Behold, you are beautiful.
\par }{\Q Your eyes are like doves behind your veil.
\par }{\QB Your hair is as a flock of goats,
\par }{\QB that descend from Mount Gilead.
\par }{\Q \VS{2}Your teeth are like a newly shorn flock,
\par }{\QB which have come up from the washing,
\par }{\QB where every one of them has twins.
\par }{\QB None is bereaved among them.
\par }{\Q \VS{3}Your lips are like scarlet thread.
\par }{\QB Your mouth is lovely.
\par }{\QB Your temples are like a piece of a pomegranate behind your veil.
\par }{\Q \VS{4}Your neck is like David’s tower built for an armory,
\par }{\QB on which a thousand shields hang,
\par }{\QB all the shields of the mighty men.
\par }{\Q \VS{5}Your two breasts are like two fawns
\par }{\QB that are twins of a roe,
\par }{\QB which feed among the lilies.
\par }{\BB \par }{\Q \VS{6}Until the day is cool, and the shadows flee away,
\par }{\QB I will go to the mountain of myrrh,
\par }{\QB to the hill of frankincense.
\par }{\BB \par }{\Q \VS{7}You are all beautiful, my love.
\par }{\QB There is no spot in you.
\par }{\Q \VS{8}Come with me from Lebanon, my bride,
\par }{\QB with me from Lebanon.
\par }{\QB Look from the top of Amana,
\par }{\QB from the top of Senir and Hermon,
\par }{\QB from the lions’ dens,
\par }{\QB from the mountains of the leopards.
\par }{\BB \par }{\Q \VS{9}You have ravished my heart, my sister, my bride.
\par }{\QB You have ravished my heart with one of your eyes,
\par }{\QB with one chain of your neck.
\par }{\Q \VS{10}How beautiful is your love, my sister, my bride!
\par }{\QB How much better is your love than wine,
\par }{\QB the fragrance of your perfumes than all kinds of spices!
\par }{\Q \VS{11}Your lips, my bride, drip like the honeycomb.
\par }{\QB Honey and milk are under your tongue.
\par }{\QB The smell of your garments is like the smell of Lebanon.
\par }{\Q \VS{12}My sister, my bride, is a locked up garden;
\par }{\QB a locked up spring,
\par }{\QB a sealed fountain.
\par }{\Q \VS{13}Your shoots are an orchard of pomegranates, with precious fruits,
\par }{\QB henna with spikenard plants,
\par }{\QB \VS{14}spikenard and saffron,
\par }{\QB calamus and cinnamon, with every kind of incense tree;
\par }{\QB myrrh and aloes, with all the best spices,
\par }{\QB \VS{15}a fountain of gardens,
\par }{\QB a well of living waters,
\par }{\QB flowing streams from Lebanon.
\par }{\SP Beloved
\par }{\Q \VS{16}Awake, north wind, and come, you south!
\par }{\QB Blow on my garden, that its spices may flow out.
\par }{\Q Let my beloved come into his garden,
\par }{\QB and taste his precious fruits.

\par }\Chap{5}{\SP Lover
\par }{\Q \VerseOne{1}I have come into my garden, my sister, my bride.
\par }{\QB I have gathered my myrrh with my spice;
\par }{\QB I have eaten my honeycomb with my honey;
\par }{\QB I have drunk my wine with my milk.
\par }{\SP Friends
\par }{\Q Eat, friends!
\par }{\QB Drink, yes, drink abundantly, beloved.
\par }{\SP Beloved
\par }{\Q \VS{2}I was asleep, but my heart was awake.
\par }{\QB It is the voice of my beloved who knocks:
\par }{\QB “Open to me, my sister, my love, my dove, my undefiled;
\par }{\QB for my head is filled with dew,
\par }{\QB and my hair with the dampness of the night.”
\par }{\Q \VS{3}I have taken off my robe. Indeed, must I put it on?
\par }{\QB I have washed my feet. Indeed, must I soil them?
\par }{\Q \VS{4}My beloved thrust his hand in through the latch opening.
\par }{\QB My heart pounded for him.
\par }{\Q \VS{5}I rose up to open for my beloved.
\par }{\QB My hands dripped with myrrh,
\par }{\QB my fingers with liquid myrrh,
\par }{\QB on the handles of the lock.
\par }{\Q \VS{6}I opened to my beloved;
\par }{\QB but my beloved left, and had gone away.
\par }{\Q My heart went out when he spoke.
\par }{\QB I looked for him, but I didn’t find him.
\par }{\QB I called him, but he didn’t answer.
\par }{\Q \VS{7}The watchmen who go about the city found me.
\par }{\QB They beat me.
\par }{\QB They bruised me.
\par }{\QB The keepers of the walls took my cloak away from me.
\par }{\BB \par }{\Q \VS{8}I adjure you, daughters of Jerusalem,
\par }{\QB If you find my beloved,
\par }{\QB that you tell him that I am faint with love.
\par }{\SP Friends
\par }{\Q \VS{9}How is your beloved better than another beloved,
\par }{\QB you fairest among women?
\par }{\Q How is your beloved better than another beloved,
\par }{\Q that you do so adjure us?
\par }{\SP Beloved
\par }{\Q \VS{10}My beloved is white and ruddy.
\par }{\QB The best among ten thousand.
\par }{\Q \VS{11}His head is like the purest gold.
\par }{\QB His hair is bushy, black as a raven.
\par }{\Q \VS{12}His eyes are like doves beside the water brooks,
\par }{\QB washed with milk, mounted like jewels.
\par }{\Q \VS{13}His cheeks are like a bed of spices with towers of perfumes.
\par }{\QB His lips are like lilies, dropping liquid myrrh.
\par }{\Q \VS{14}His hands are like rings of gold set with beryl.
\par }{\QB His body is like ivory work overlaid with sapphires.
\par }{\Q \VS{15}His legs are like pillars of marble set on sockets of fine gold.
\par }{\QB His appearance is like Lebanon, excellent as the cedars.
\par }{\Q \VS{16}His mouth is sweetness;
\par }{\QB yes, he is altogether lovely.
\par }{\Q This is my beloved, and this is my friend,
\par }{\QB daughters of Jerusalem.

\par }\Chap{6}{\SP Friends
\par }{\Q \VerseOne{1}Where has your beloved gone, you fairest among women?
\par }{\QB Where has your beloved turned, that we may seek him with you?
\par }{\SP Beloved
\par }{\Q \VS{2}My beloved has gone down to his garden,
\par }{\QB to the beds of spices,
\par }{\QB to pasture his flock in the gardens, and to gather lilies.
\par }{\Q \VS{3}I am my beloved’s, and my beloved is mine.
\par }{\QB He browses among the lilies.
\par }{\BB \par }{\SP Lover
\par }{\Q \VS{4}You are beautiful, my love, as Tirzah,
\par }{\QB lovely as Jerusalem,
\par }{\QB awesome as an army with banners.
\par }{\Q \VS{5}Turn away your eyes from me,
\par }{\QB for they have overcome me.
\par }{\Q Your hair is like a flock of goats,
\par }{\QB that lie along the side of Gilead.
\par }{\Q \VS{6}Your teeth are like a flock of ewes,
\par }{\QB which have come up from the washing,
\par }{\QB of which every one has twins;
\par }{\QB not one is bereaved among them.
\par }{\Q \VS{7}Your temples are like a piece of a pomegranate behind your veil.
\par }{\BB \par }{\Q \VS{8}There are sixty queens, eighty concubines,
\par }{\QB and virgins without number.
\par }{\Q \VS{9}My dove, my perfect one, is unique.
\par }{\QB She is her mother’s only daughter.
\par }{\QB She is the favorite one of her who bore her.
\par }{\Q The daughters saw her, and called her blessed.
\par }{\QB The queens and the concubines saw her, and they praised her.
\par }{\BB \par }{\Q \VS{10}Who is she who looks out as the morning,
\par }{\QB beautiful as the moon,
\par }{\QB clear as the sun,
\par }{\QB and awesome as an army with banners?
\par }{\BB \par }{\Q \VS{11}I went down into the nut tree grove,
\par }{\QB to see the green plants of the valley,
\par }{\QB to see whether the vine budded,
\par }{\QB and the pomegranates were in flower.
\par }{\Q \VS{12}Without realizing it,
\par }{\QB my desire set me with my royal people’s chariots.
\par }{\SP Friends
\par }{\Q \VS{13}Return, return, Shulammite!
\par }{\QB Return, return, that we may gaze at you.
\par }{\SP Lover
\par }{\Q Why do you desire to gaze at the Shulammite,
\par }{\QB as at the dance of Mahanaim?

\par }\Chap{7}{\PP \VerseOne{1}How beautiful are your feet in sandals, prince’s daughter!
\par }{\QB Your rounded thighs are like jewels,
\par }{\QB the work of the hands of a skillful workman.
\par }{\Q \VS{2}Your body is like a round goblet,
\par }{\QB no mixed wine is wanting.
\par }{\Q Your waist is like a heap of wheat,
\par }{\QB set about with lilies.
\par }{\Q \VS{3}Your two breasts are like two fawns,
\par }{\QB that are twins of a roe.
\par }{\Q \VS{4}Your neck is like an ivory tower.
\par }{\QB Your eyes are like the pools in Heshbon by the gate of Bathrabbim.
\par }{\QB Your nose is like the tower of Lebanon which looks toward Damascus.
\par }{\Q \VS{5}Your head on you is like Carmel.
\par }{\QB The hair of your head like purple.
\par }{\QB The king is held captive in its tresses.
\par }{\Q \VS{6}How beautiful and how pleasant you are,
\par }{\QB love, for delights!
\par }{\Q \VS{7}This, your stature, is like a palm tree,
\par }{\QB your breasts like its fruit.
\par }{\Q \VS{8}I said, “I will climb up into the palm tree.
\par }{\QB I will take hold of its fruit.”
\par }{\Q Let your breasts be like clusters of the vine,
\par }{\QB the smell of your breath like apples.
\par }{\Q \VS{9}Your mouth is like the best wine,
\par }{\QB that goes down smoothly for my beloved,
\par }{\QB gliding through the lips of those who are asleep.
\par }{\SP Beloved
\par }{\Q \VS{10}I am my beloved’s.
\par }{\QB His desire is toward me.
\par }{\Q \VS{11}Come, my beloved! Let’s go out into the field.
\par }{\QB Let’s lodge in the villages.
\par }{\Q \VS{12}Let’s go early up to the vineyards.
\par }{\QB Let’s see whether the vine has budded,
\par }{\QB its blossom is open,
\par }{\QB and the pomegranates are in flower.
\par }{\QB There I will give you my love.
\par }{\Q \VS{13}The mandrakes produce fragrance.
\par }{\QB At our doors are all kinds of precious fruits, new and old,
\par }{\QB which I have stored up for you, my beloved.

\par }\Chap{8}{\PP \VerseOne{1}Oh that you were like my brother,
\par }{\QB who nursed from the breasts of my mother!
\par }{\Q If I found you outside, I would kiss you;
\par }{\QB yes, and no one would despise me.
\par }{\Q \VS{2}I would lead you, bringing you into the house of my mother,
\par }{\QB who would instruct me.
\par }{\Q I would have you drink spiced wine,
\par }{\QB of the juice of my pomegranate.
\par }{\Q \VS{3}His left hand would be under my head.
\par }{\QB His right hand would embrace me.
\par }{\BB \par }{\Q \VS{4}I adjure you, daughters of Jerusalem,
\par }{\QB that you not stir up, nor awaken love,
\par }{\QB until it so desires.
\par }{\SP Friends
\par }{\Q \VS{5}Who is this who comes up from the wilderness,
\par }{\QB leaning on her beloved?
\par }{\BB \par }{\SP Beloved
\par }{\Q Under the apple tree I aroused you.
\par }{\QB There your mother conceived you.
\par }{\QB There she was in labor and bore you.
\par }{\BB \par }{\Q \VS{6}Set me as a seal on your heart,
\par }{\QB as a seal on your arm;
\par }{\QB for love is strong as death.
\par }{\QB Jealousy is as cruel as Sheol.\FTNT{*}{{\FR 8:6 }
Sheol is the place of the dead.
}
\par }{\QB Its flashes are flashes of fire,
\par }{\QB a very flame of Yahweh.\FTNT{†}{{\FR 8:6 }
“Yahweh” is God’s proper Name, sometimes rendered “LORD” (all caps) in other translations.}
\par }{\Q \VS{7}Many waters can’t quench love,
\par }{\QB neither can floods drown it.
\par }{\Q If a man would give all the wealth of his house for love,
\par }{\QB he would be utterly scorned.
\par }{\SP Brothers
\par }{\Q \VS{8}We have a little sister.
\par }{\QB She has no breasts.
\par }{\Q What shall we do for our sister
\par }{\QB in the day when she is to be spoken for?
\par }{\BB \par }{\Q \VS{9}If she is a wall,
\par }{\QB we will build on her a turret of silver.
\par }{\Q If she is a door,
\par }{\QB we will enclose her with boards of cedar.
\par }{\SP Beloved
\par }{\Q \VS{10}I am a wall, and my breasts like towers,
\par }{\QB then I was in his eyes like one who found peace.
\par }{\Q \VS{11}Solomon had a vineyard at Baal Hamon.
\par }{\QB He leased out the vineyard to keepers.
\par }{\QB Each was to bring a thousand shekels\FTNT{‡}{{\FR 8:11 }
A shekel is about 10 grams or about 0.35 ounces, so 1000 shekels is about 10 kilograms or about 22 pounds.} of silver for its fruit.
\par }{\Q \VS{12}My own vineyard is before me.
\par }{\QB The thousand are for you, Solomon,
\par }{\QB two hundred for those who tend its fruit.
\par }{\SP Lover
\par }{\Q \VS{13}You who dwell in the gardens, with friends in attendance,
\par }{\QB let me hear your voice!
\par }{\SP Beloved
\par }{\Q \VS{14}Come away, my beloved!
\par }{\QB Be like a gazelle or a young stag on the mountains of spices!
\par }