\NormalFont\ShortTitle{Wisdom of Solomon}
{\MT THE WISDOM OF SOLOMON
\par }{\IP The
{\BK{Wisdom of Solomon}} is recognized as Deuterocanonical Scripture by the Roman Catholic, Greek Orthodox, and Russian Orthodox Churches.

\par }\PoetryChap{1}{\Q \VerseOne{1}Love righteousness, all you who are judges of the earth.
\par }{\QB Think of the Lord
\FTNT{*}{{\FR 1:1 }
Gr.
in goodness.
} with a good mind.
\par }{\Q Seek him in singleness of heart,
\par }{\QB \VS{2}because he is found by those who don’t tempt him,
\par }{\QB and is manifested to those who trust him.
\par }{\Q \VS{3}For crooked thoughts separate from God.
\par }{\QB His Power convicts when it is tested,
\par }{\QB and exposes the foolish.
\par }{\Q \VS{4}Because wisdom will not enter into a soul that devises evil,
\par }{\QB nor dwell in a body that is enslaved by sin.
\par }{\Q \VS{5}For a holy spirit of discipline will flee deceit,
\par }{\QB and will depart from thoughts that are without understanding,
\par }{\QB and will be ashamed when unrighteousness has come in.
\par }{\Q \VS{6}For
\FTNT{†}{{\FR 1:6 }
Some authorities read
the spirit of wisdom is loving to man.
} wisdom is a spirit who loves man,
\par }{\QB and she will not hold a
\FTNT{‡}{{\FR 1:6 }
Or,
reviler
} blasphemer guiltless for his lips;
\par }{\QB because God is witness of his inmost self,
\par }{\QB and is a true overseer of his heart,
\par }{\QB and a hearer of his tongue:
\par }{\Q \VS{7}Because the spirit of the Lord has filled
\FTNT{§}{{\FR 1:7 }
Gr.
the inhabited earth.
} the world,
\par }{\QB and that which holds all things together knows what is said.
\par }{\Q \VS{8}Therefore no one who utters unrighteous things will be unseen;
\par }{\QB \FTNT{*}{{\FR 1:8 }
Some authorities read
nor indeed.
} neither will Justice, when it convicts, pass him by.
\par }{\Q \VS{9}For in his counsels the ungodly will be searched out,
\par }{\QB and the sound of his words will come to the Lord to bring his lawless deeds to conviction;
\par }{\Q \VS{10}because a jealous ear listens to all things,
\par }{\QB and the noise of murmurings is not hidden.
\par }{\Q \VS{11}Beware then of unprofitable murmuring,
\par }{\QB and keep your tongue from slander;
\par }{\QB because no secret utterance will go on its way void,
\par }{\QB and a lying mouth destroys a soul.
\par }{\Q \VS{12}Don’t court death in the error of your life;
\par }{\QB and don’t draw destruction upon yourselves by the works of your hands,
\par }{\QB \VS{13}because God didn’t make death;
\par }{\QB neither does he delight when the living perish.
\par }{\Q \VS{14}For he created all things that they might have being.
\par }{\QB \FTNT{†}{{\FR 1:14 }
Or, all
the races
of creatures
in the world
}The generative powers of the world are wholesome,
\par }{\QB and there is no poison of destruction in them,
\par }{\QB nor has Hades
\FTNT{‡}{{\FR 1:14 }
Or,
a royal house
} royal dominion upon earth;
\par }{\Q \VS{15}For righteousness is immortal,
\par }{\QB \VS{16}but ungodly men by their hands and their words summon
\FTNT{§}{{\FR 1:16 }
Or,
Hades
Gr.
him.
}death;
\par }{\QB deeming him a friend they
\FTNT{*}{{\FR 1:16 }
Or,
were consumed
with love of him
} pined away.
\par }{\Q They made a covenant with him,
\par }{\QB because they are worthy to belong with him.
\par }{\BB \par }
\PoetryChap{2}{\Q \VerseOne{1}For they said
\FTNT{*}{{\FR 2:1 }
Or,
among
} within themselves, with unsound reasoning,
\par }{\QB “Our life is short and sorrowful.
\par }{\QB There is no healing when a man comes to his end,
\par }{\QB and no one was ever known that
\FTNT{†}{{\FR 2:1 }
Or,
returned out of Hades
} was released from Hades.
\par }{\Q \VS{2}Because we were born by mere chance,
\par }{\QB and hereafter we will be as though we had never been;
\par }{\QB because the breath in our nostrils is smoke,
\par }{\QB and reason is a spark kindled by the beating of our heart;
\par }{\Q \VS{3}which being extinguished, the body will be turned into ashes,
\par }{\QB and the spirit will be dispersed as thin air.
\par }{\Q \VS{4}Our name will be forgotten in time.
\par }{\QB No one will remember our works.
\par }{\QB Our life will pass away as the traces of a cloud,
\par }{\QB and will be scattered as is a mist,
\par }{\QB when it is chased by the rays of the sun,
\par }{\QB and
\FTNT{‡}{{\FR 2:4 }
Gr.
weighed down.
} overcome by its heat.
\par }{\Q \VS{5}For our allotted time is the passing of a shadow,
\par }{\QB and
\FTNT{§}{{\FR 2:5 }
Or,
there is no putting back of our end
} our end doesn’t retreat;
\par }{\QB because it is securely sealed, and no one
\FTNT{*}{{\FR 2:5 }
Or,
comes again
} turns it back.
\par }{\BB \par }{\Q \VS{6}“Come therefore and let’s enjoy the good things
\FTNT{†}{{\FR 2:6 }
Or,
that are
} that
{\ADD{now
}} exist.
\par }{\QB Let’s use the creation earnestly as in our youth.
\par }{\Q \VS{7}Let’s fill ourselves with costly wine and perfumes,
\par }{\QB and let no Spring flower pass us by.
\par }{\Q \VS{8}Let’s crown ourselves with rosebuds before they wither.
\par }{\Q \VS{9}Let none of us go without his share in our proud revelry.
\par }{\QB Let’s leave tokens of mirth everywhere,
\par }{\QB because this is our portion, and this is our lot.
\par }{\Q \VS{10}Let’s oppress the righteous poor.
\par }{\QB Let’s not spare the widow,
\par }{\QB nor reverence the gray hair of the old man.
\par }{\Q \VS{11}But let our strength be a law of righteousness;
\par }{\QB for that which is weak is proven useless.
\par }{\Q \VS{12}But let’s lie in wait for the righteous man,
\par }{\QB because he annoys us,
\par }{\QB is contrary to our works,
\par }{\QB reproaches us with sins against the law,
\par }{\QB and charges us with sins against our training.
\par }{\Q \VS{13}He professes to have knowledge of God,
\par }{\QB and calls himself a child of the Lord.
\par }{\Q \VS{14}He became to us a reproof of our thoughts.
\par }{\Q \VS{15}He is grievous to us even to look at,
\par }{\QB because his life is unlike other men’s,
\par }{\QB and his paths are strange.
\par }{\Q \VS{16}We were regarded by him as worthless metal,
\par }{\QB and he abstains from our ways as from uncleanness.
\par }{\QB He calls the latter end of the righteous happy.
\par }{\QB He boasts that God is his father.
\par }{\Q \VS{17}Let’s see if his words are true.
\par }{\QB Let’s test what will happen at the end of his life.
\par }{\Q \VS{18}For if the righteous man is God’s son, he will uphold him,
\par }{\QB and he will deliver him out of the hand of his adversaries.
\par }{\Q \VS{19}Let’s test him with outrage and torture,
\par }{\QB that we may find out how gentle he is,
\par }{\QB and test his patience.
\par }{\Q \VS{20}Let’s condemn him to a shameful death,
\par }{\QB for he will be overseen according to his words.”
\par }{\BB \par }{\Q \VS{21}Thus they reasoned, and they were led astray;
\par }{\QB for their wickedness blinded them,
\par }{\Q \VS{22}and they didn’t know the mysteries of God,
\par }{\QB neither did they hope for wages of holiness,
\par }{\QB nor did they discern that there is a prize for blameless souls.
\par }{\Q \VS{23}Because God created man for incorruption,
\par }{\QB and made him an image of his own everlastingness;
\par }{\Q \VS{24}but death entered into the world by the envy of the devil,
\par }{\QB and those who belong to him experience it.

\par }\PoetryChap{3}{\Q \VerseOne{1}But the souls of the righteous are in the hand of God,
\par }{\QB and no torment will touch them.
\par }{\Q \VS{2}In the eyes of the foolish they seemed to have died.
\par }{\QB Their departure was considered affliction,
\par }{\QB \VS{3}and their travel away from us ruin;
\par }{\QB But they are in peace.
\par }{\Q \VS{4}For even if in the sight of men they are punished,
\par }{\QB their hope is full of immortality.
\par }{\Q \VS{5}Having borne a little chastening, they will receive great good;
\par }{\QB because God tested them, and found them worthy of himself.
\par }{\Q \VS{6}He tested them like gold in the furnace,
\par }{\QB and he accepted them as a whole burnt offering.
\par }{\Q \VS{7}In the time of their visitation they will shine.
\par }{\QB They will run back and forth like sparks among stubble.
\par }{\Q \VS{8}They will judge nations and have dominion over peoples.
\par }{\QB The Lord will reign over them forever.
\par }{\QB \VS{9}Those who trust him will understand truth.
\par }{\QB \FTNT{*}{{\FR 3:9 }
Or,
Those who are faithful through love will live with him
} The faithful will live with him in love,
\par }{\QB because grace and mercy are with his chosen ones.
\par }{\BB \par }{\Q \VS{10}But the ungodly will be punished even as they reasoned,
\par }{\QB those who neglected righteousness and revolted from the Lord;
\par }{\QB \VS{11}for he who despises wisdom and discipline is miserable.
\par }{\Q Their hope is void and their toils unprofitable.
\par }{\QB Their works are useless.
\par }{\Q \VS{12}Their wives are foolish and their children are wicked.
\par }{\QB \VS{13}Their descendants are cursed,
\par }{\QB because the barren woman who is undefiled is happy,
\par }{\QB she who has not conceived in transgression.
\par }{\QB She will have fruit in the visitation of souls.
\par }{\Q \VS{14}So is the eunuch which has done no lawless deed with his hands,
\par }{\QB nor imagined wicked things against the Lord;
\par }{\Q for a precious gift will be given to him for his faithfulness, a special favor,
\par }{\QB and a delightful inheritance in the Lord’s sanctuary.
\par }{\Q \VS{15}For good labors have fruit of great renown.
\par }{\QB The root of understanding can’t fail.
\par }{\Q \VS{16}But children of adulterers will not come to maturity.
\par }{\QB The seed of an unlawful bed will vanish away.
\par }{\Q \VS{17}For if they live long, they will not be esteemed,
\par }{\QB and in the end, their old age will be without honor.
\par }{\Q \VS{18}If they die quickly, they will have no hope,
\par }{\QB nor consolation in the day of decision.
\par }{\Q \VS{19}For the end of an unrighteous generation is always grievous.

\par }\PoetryChap{4}{\Q \VerseOne{1}It is better to be childless with virtue,
\par }{\QB for immortality is in the memory of virtue,
\par }{\QB because it is recognized both before God and before men.
\par }{\Q \VS{2}When it is present, people imitate it.
\par }{\QB They long after it when it has departed.
\par }{\Q \FTNT{*}{{\FR 4:2 }
Gr.
in the age.}Throughout all time it marches crowned in triumph,
\par }{\QB victorious in the competition for the prizes that are undefiled.
\par }{\Q \VS{3}But the multiplying brood of the ungodly will be of no profit,
\par }{\QB and their illegitimate offshoots won’t take deep root,
\par }{\QB nor will they establish a sure hold.
\par }{\Q \VS{4}For even if they grow branches and flourish for a season,
\par }{\QB standing unsure, they will be shaken by the wind.
\par }{\QB They will be uprooted by the violence of winds.
\par }{\Q \VS{5}Their branches will be broken off before they come to maturity.
\par }{\QB Their fruit will be useless,
\par }{\QB never ripe to eat, and fit for nothing.
\par }{\Q \VS{6}For unlawfully conceived children are witnesses of wickedness
\par }{\QB against parents when they are investigated.
\par }{\BB \par }{\Q \VS{7}But a righteous man, even if he dies before his time, will be at rest.
\par }{\Q \VS{8}For honorable old age is not that which stands in length of time,
\par }{\QB nor is its measure given by number of years:
\par }{\Q \VS{9}But understanding is gray hair to men,
\par }{\QB and an unspotted life is ripe old age.
\par }{\Q \VS{10}Being found well-pleasing to God, he was loved.
\par }{\QB While living among sinners he was transported.
\par }{\Q \VS{11}He was caught away, lest evil should change his understanding,
\par }{\QB or guile deceive his soul.
\par }{\Q \VS{12}For the witchcraft of worthlessness obscures the things which are good,
\par }{\QB and the whirl of desire perverts an innocent mind.
\par }{\Q \VS{13}Being made perfect quickly,
\par }{\QB he filled a long time;
\par }{\QB \VS{14}for his soul was pleasing to the Lord.
\par }{\QB Therefore he hurried out of the midst of wickedness.
\par }{\Q \VS{15}But as for the peoples seeing and not understanding,
\par }{\QB not considering this, that grace and mercy are with his chosen,
\par }{\QB and that he visits his holy ones—
\par }{\Q \VS{16}but a righteous man who is dead will condemn the ungodly who are living,
\par }{\QB and youth that is quickly perfected will condemn the many years of an unrighteous man’s old age.
\par }{\Q \VS{17}For the ungodly will see a wise man’s end,
\par }{\QB and won’t understand what the Lord planned for him,
\par }{\QB and why he safely kept him.
\par }{\Q \VS{18}They will see, and they will despise;
\par }{\QB but the Lord will laugh them to scorn.
\par }{\Q After this, they will become a dishonored carcass
\par }{\QB and a reproach among the dead forever;
\par }{\Q \VS{19}because he will dash them speechless to the ground,
\par }{\QB and will shake them from the foundations.
\par }{\Q They will lie utterly waste.
\par }{\QB They will be in anguish
\par }{\QB and their memory will perish.
\par }{\BB \par }{\Q \VS{20}They will come with coward fear when their sins are counted.
\par }{\QB Their lawless deeds will convict them to their face.

\par }\PoetryChap{5}{\Q \VerseOne{1}Then the righteous man will stand in great boldness
\par }{\QB before the face of those who afflicted him,
\par }{\QB and those who make his labors of no account.
\par }{\Q \VS{2}When they see him, they will be troubled with terrible fear,
\par }{\QB and will be amazed at the marvel of salvation.
\par }{\Q \VS{3}They will speak among themselves repenting,
\par }{\QB and for distress of spirit they will groan,
\par }{\QB “This was he whom we used to hold in derision,
\par }{\QB as a parable of reproach.
\par }{\Q \VS{4}We fools considered his life madness,
\par }{\QB and his end without honor.
\par }{\Q \VS{5}How was he counted among sons of God?
\par }{\QB How is his lot among saints?
\par }{\Q \VS{6}Truly we went astray from the way of truth.
\par }{\QB The light of righteousness didn’t shine for us.
\par }{\QB The sun didn’t rise for us.
\par }{\Q \VS{7}We
\FTNT{*}{{\FR 5:7 }
See Proverbs 14:14.
} took our fill of the paths of lawlessness and destruction.
\par }{\QB We traveled through trackless deserts,
\par }{\QB but we didn’t know the Lord’s way.
\par }{\Q \VS{8}What did our arrogance profit us?
\par }{\QB What good have riches and boasting brought us?
\par }{\Q \VS{9}Those things all passed away as a shadow,
\par }{\QB like a message that runs by,
\par }{\QB \VS{10}like a ship passing through the billowy water,
\par }{\QB which, when it has gone by, there is no trace to be found,
\par }{\QB no pathway of its keel in the billows.
\par }{\Q \VS{11}Or it is like when a bird flies through the air,
\par }{\QB no evidence of its passage is found,
\par }{\QB but the light wind, lashed with the stroke of its pinions,
\par }{\QB and torn apart with the violent rush of the moving wings, is passed through.
\par }{\QB Afterwards no sign of its coming remains.
\par }{\Q \VS{12}Or it is like when an arrow is shot at a mark,
\par }{\QB the air disparted closes up again immediately,
\par }{\QB so that men don’t know where it passed through.
\par }{\Q \VS{13}So we also, as soon as we were born, ceased to be;
\par }{\QB and we had no sign of virtue to show,
\par }{\QB but we were utterly consumed in our wickedness.”
\par }{\Q \VS{14}Because the hope of the ungodly man is like chaff carried by the wind,
\par }{\QB and
\FTNT{†}{{\FR 5:14 }
Gr.
like foam chased to thinness:
or,
as thin foam chased.
} as
\FTNT{‡}{{\FR 5:14 }
Most Greek authorities read
hoar frost:
some authorities, perhaps rightly,
a spider’s web.} foam vanishing before a tempest;
\par }{\QB and is scattered like smoke by the wind,
\par }{\QB and passes by as the remembrance of a guest that waits but a day.
\par }{\BB \par }{\Q \VS{15}But the righteous live forever.
\par }{\QB Their reward is in the Lord,
\par }{\QB and the care for them with the Most High.
\par }{\Q \VS{16}Therefore they will receive the crown of royal dignity
\par }{\QB and the diadem of beauty from the Lord’s hand;
\par }{\Q because he will cover them with his right hand,
\par }{\QB and he will shield them with his arm.
\par }{\Q \VS{17}He will take his jealousy as complete armor,
\par }{\QB and will make the whole creation his weapons to punish his enemies:
\par }{\Q \VS{18}He will put on righteousness as a breastplate,
\par }{\QB and will wear impartial judgement as a helmet.
\par }{\Q \VS{19}He will take holiness as an invincible shield.
\par }{\QB \VS{20}He will sharpen stern wrath for a sword.
\par }{\QB The world will go with him to fight against his frenzied foes.
\par }{\Q \VS{21}Shafts of lightning will fly with true aim.
\par }{\QB They will leap to the mark from the clouds, as from a well-drawn bow.
\par }{\Q \VS{22}Hailstones full of wrath will be hurled from an engine of war.
\par }{\QB The water of the sea will be angered against them.
\par }{\QB Rivers will sternly overwhelm them.
\par }{\Q \VS{23}A mighty blast will encounter them.
\par }{\QB It will winnow them away like a tempest.
\par }{\Q So lawlessness will make all the land desolate.
\par }{\QB Their evil-doing will overturn the thrones of princes.

\par }\PoetryChap{6}{\Q \VerseOne{1}Hear therefore, you kings, and understand.
\par }{\QB Learn, you judges of the ends of the earth.
\par }{\QB \VS{2}Give ear, you rulers who have dominion over many people,
\par }{\QB and make your boast
\FTNT{*}{{\FR 6:2 }
Or,
in the multitudes of your nations
} in multitudes of nations,
\par }{\Q \VS{3}because your dominion was given to you from the Lord,
\par }{\QB and your sovereignty from the Most High.
\par }{\Q He will search out your works,
\par }{\QB and will inquire about your plans;
\par }{\Q \VS{4}because being officers of his kingdom, you didn’t judge rightly,
\par }{\QB nor did you keep\FTNT{†}{{\FR 6:4 }
Or,
the law
} law,
\par }{\QB nor did you walk according to God’s counsel.
\par }{\Q \VS{5}He will come upon you awfully and swiftly,
\par }{\QB because a stern judgement comes on those who are in high place.
\par }{\Q \VS{6}For the man of low estate may be pardoned in mercy,
\par }{\QB but mighty men will be mightily tested.
\par }{\Q \VS{7}For the Sovereign Lord of all will not be impressed with anyone,
\par }{\QB neither will he show deference to greatness;
\par }{\Q because it is he who made both small and great,
\par }{\QB and cares about them all;
\par }{\Q \VS{8}but the scrutiny that comes upon the powerful is strict.
\par }{\Q \VS{9}Therefore, my words are to you, O princes,
\par }{\QB that you may learn wisdom and not fall away.
\par }{\Q \VS{10}For those who have kept the things that are holy in holiness will be made holy.
\par }{\QB Those who have been taught them will find what to say in defense.
\par }{\Q \VS{11}Therefore set your desire on my words.
\par }{\QB Long for them, and you princes will be instructed.
\par }{\BB \par }{\Q \VS{12}Wisdom is radiant and doesn’t fade away;
\par }{\QB and is easily seen by those who love her,
\par }{\QB and found by those who seek her.
\par }{\Q \VS{13}She anticipates those who desire her, making herself known.
\par }{\Q \VS{14}He who rises up early to seek her won’t have difficulty,
\par }{\QB for he will find her sitting at his gates.
\par }{\QB \VS{15}For to think upon her is perfection of understanding,
\par }{\QB and he who watches for her will quickly be free from care;
\par }{\Q \VS{16}because she herself goes around, seeking those who are worthy of her,
\par }{\QB and in their paths she appears to them graciously,
\par }{\QB and in every purpose she meets them.
\par }{\Q \VS{17}For her true beginning is desire for instruction;
\par }{\QB and desire for instruction is love.
\par }{\Q \VS{18}And love is observance of her laws.
\par }{\QB To give heed to her laws confirms immortality.
\par }{\Q \VS{19}Immortality brings closeness to God.
\par }{\QB \VS{20}So then desire for wisdom promotes to a kingdom.
\par }{\Q \VS{21}If therefore you delight in thrones and sceptres, you princes of peoples,
\par }{\QB honor wisdom, that you may reign forever.
\par }{\Q \VS{22}But what wisdom is, and how she came into being, I will declare.
\par }{\QB I won’t hide mysteries from you;
\par }{\QB but I will explore from her first beginning,
\par }{\QB bring the knowledge of her into clear light,
\par }{\QB and I will not pass by the truth.
\par }{\Q \VS{23}Indeed, I won’t go with consuming envy,
\par }{\QB because envy will have no fellowship with wisdom.
\par }{\Q \VS{24}But a multitude of wise men is salvation to the world,
\par }{\QB and an understanding king is stability for his people.
\par }{\Q \VS{25}Therefore be instructed by my words, and you will profit.

\par }\PoetryChap{7}{\Q \VerseOne{1}I myself am also
\FTNT{*}{{\FR 7:1 }
Many authorities read
a mortal man.} mortal, like everyone else,
\par }{\QB and am a descendant of one formed first and born of the earth.
\par }{\Q \VS{2}I molded into flesh in the time of ten months in my mother’s womb,
\par }{\QB being compacted in blood from the seed of man and pleasure that came with sleep.
\par }{\Q \VS{3}I also, when I was born, drew in the common air,
\par }{\QB and fell upon the kindred earth,
\par }{\Q uttering, like all, for my first voice, the same cry.
\par }{\Q \VS{4}I was nursed with care in swaddling clothes.
\par }{\Q \VS{5}For no king had any other first beginning;
\par }{\QB \VS{6}but all men have one entrance into life, and a common departure.
\par }{\Q \VS{7}For this cause I prayed, and understanding was given to me.
\par }{\QB I asked a spirit of wisdom came to me.
\par }{\Q \VS{8}I preferred her before sceptres and thrones.
\par }{\QB I considered riches nothing in comparison to her.
\par }{\Q \VS{9}Neither did I liken to her any priceless gem,
\par }{\QB because all gold in her presence is a little sand,
\par }{\QB and silver will be considered as clay before her.
\par }{\Q \VS{10}I loved her more than health and beauty,
\par }{\QB and I chose to have her rather than light,
\par }{\QB because her bright shining is never laid to sleep.
\par }{\Q \VS{11}All good things came to me with her,
\par }{\QB and innumerable riches are in her hands.
\par }{\Q \VS{12}And I rejoiced over them all because wisdom leads them;
\par }{\QB although I didn’t know that she was their mother.
\par }{\Q \VS{13}As I learned without guile, I impart without grudging.
\par }{\QB I don’t hide her riches.
\par }{\Q \VS{14}For she is a treasure for men that doesn’t fail,
\par }{\QB and those who use it obtain friendship with God,
\par }{\QB commended by the gifts which they present through discipline.
\par }{\BB \par }{\Q \VS{15}But may God grant that I may speak his judgement,
\par }{\QB and to conceive thoughts worthy of what has been given me;
\par }{\QB because he is one who guides even wisdom and who corrects the wise.
\par }{\Q \VS{16}For both we and our words are in his hand,
\par }{\QB with all understanding and skill in various crafts.
\par }{\Q \VS{17}For he himself gave me an unerring knowledge of the things that are,
\par }{\QB to know the structure of the universe and the operation of the elements;
\par }{\QB \VS{18}the beginning, end, and middle of times;
\par }{\QB the alternations of the solstices and the changes of seasons;
\par }{\QB \VS{19}the circuits of years and the positions of stars;
\par }{\QB \VS{20}the natures of living creatures and the raging of wild beasts;
\par }{\QB The violence of
\FTNT{†}{{\FR 7:20 }
Or,
spirits
} winds and the thoughts of men;
\par }{\QB the diversities of plants and the virtues of roots.
\par }{\Q \VS{21}All things that are either secret or manifest I learned,
\par }{\QB \VS{22}for wisdom, that is the architect of all things, taught me.
\par }{\BB \par }{\Q For there is in her a spirit that is quick to understand, holy,
\par }{\QB unique, manifold, subtle, freely moving, clear in utterance, unpolluted,
\par }{\QB distinct, unharmed, loving what is good, keen, unhindered,
\par }{\QB \VS{23}beneficent, loving toward man, steadfast, sure, free from care, all-powerful, all-surveying,
\par }{\QB and penetrating through all spirits that are quick to understand, pure, most subtle:
\par }{\Q \VS{24}For wisdom is more mobile than any motion.
\par }{\QB Yes, she pervades and penetrates all things by reason of her purity.
\par }{\Q \VS{25}For she is a breath of the power of God,
\par }{\QB and a clear effluence of the glory of the Almighty.
\par }{\Q Therefore nothing defiled can find entrance into her.
\par }{\Q \VS{26}For she is a reflection of everlasting light,
\par }{\QB an unspotted mirror of the working of God,
\par }{\QB and an image of his goodness.
\par }{\Q \VS{27}She, being one, has power to do all things.
\par }{\QB Remaining in herself, she renews all things.
\par }{\Q From generation to generation passing into holy souls,
\par }{\QB she makes friends of God and prophets.
\par }{\QB \VS{28}For God loves nothing as much as one who dwells with wisdom.
\par }{\Q \VS{29}For she is fairer than the sun,
\par }{\QB and above all the constellations of the stars.
\par }{\QB She is better than light.
\par }{\Q \VS{30}For daylight succeeds night,
\par }{\QB but evil does not prevail against wisdom.

\par }\PoetryChap{8}{\Q \VerseOne{1}But she reaches from one end to the other with full strength,
\par }{\QB and orders all things well.
\par }{\BB \par }{\Q \VS{2}I loved her and sought her from my youth.
\par }{\QB I sought to take her for my bride.
\par }{\QB I became enamoured by her beauty.
\par }{\Q \VS{3}She glorifies her noble birth by living with God.
\par }{\QB The Sovereign Lord of all loves her.
\par }{\Q \VS{4}For she is initiated into the knowledge of God,
\par }{\QB and she chooses his works.
\par }{\Q \VS{5}But if riches are a desired possession in life,
\par }{\QB what is richer than wisdom, which makes all things?
\par }{\Q \VS{6}And if understanding works,
\par }{\QB who more than
\FTNT{*}{{\FR 8:6 }
Gr.
she.
} wisdom is an architect of the things that exist?
\par }{\Q \VS{7}If a man loves righteousness,
\par }{\QB \FTNT{†}{{\FR 8:7 }
Gr.
Her labors are.}the fruits of wisdom’s labor are virtues,
\par }{\QB for she teaches soberness, understanding, righteousness, and courage.
\par }{\QB There is nothing in life more profitable for people than these.
\par }{\Q \VS{8}And if anyone longs for wide experience,
\par }{\QB she knows the things of old, and infers the things to come.
\par }{\QB She understands subtleties of speeches and interpretations of dark sayings.
\par }{\QB She foresees signs and wonders, and the issues of seasons and times.
\par }{\Q \VS{9}Therefore I determined to take her to live with me,
\par }{\QB knowing that she is one who would give me good counsel,
\par }{\QB and encourage me in cares and grief.
\par }{\Q \VS{10}Because of her, I will have glory among multitudes,
\par }{\QB and honor in the sight of elders, though I am young.
\par }{\Q \VS{11}I will be found keen when I give judgement.
\par }{\QB I will be admired in the presence of rulers.
\par }{\Q \VS{12}When I am silent, they will wait for me.
\par }{\QB When I open my lips, they will heed what I say.
\par }{\QB If I continue speaking, they will put their hands on their mouths.
\par }{\Q \VS{13}Because of her, I will have immortality,
\par }{\QB and leave behind an eternal memory to those who come after me.
\par }{\Q \VS{14}I will govern peoples.
\par }{\QB Nations will be subjected to me.
\par }{\Q \VS{15}Dreaded monarchs will fear me when they hear of me.
\par }{\QB Among the people, I will show myself to be good, and courageous in war.
\par }{\Q \VS{16}When I come into my house, I will find rest with her.
\par }{\QB For conversation with her has no bitterness,
\par }{\QB and living with her has no pain, but gladness and joy.
\par }{\Q \VS{17}When I considered these things in myself,
\par }{\QB and thought in my heart how immortality is in kinship to wisdom,
\par }{\Q \VS{18}and in her friendship is good delight,
\par }{\QB and in the labors of her hands is wealth that doesn’t fail,
\par }{\QB and understanding is in her companionship,
\par }{\QB and great renown in having fellowship with her words,
\par }{\QB I went about seeking how to take her to myself.
\par }{\Q \VS{19}Now I was a clever child, and received a good soul fell.
\par }{\QB \VS{20}Or rather, being good, I came into an undefiled body.
\par }{\Q \VS{21}But perceiving that I could not otherwise possess wisdom unless God gave her to me—
\par }{\QB yes, and to know and understand by whom the grace is given—
\par }{\QB I pleaded with the Lord and implored him, and with my whole heart I said,

\par }\PoetryChap{9}{\Q \VerseOne{1}“O God of the fathers, and Lord of mercy,
\par }{\QB who made all things by your word;
\par }{\Q \VS{2}and by your wisdom you formed man,
\par }{\QB that he should have dominion over the creatures that were made by you,
\par }{\Q \VS{3}and rule the world in holiness and righteousness,
\par }{\QB and execute judgement in uprightness of soul;
\par }{\QB \VS{4}give me wisdom, her who sits by you on your thrones.
\par }{\Q Don’t reject me from among your
\FTNT{*}{{\FR 9:4 }
Or,
children
} servants,
\par }{\QB \VS{5}because I am your servant and the son of your handmaid,
\par }{\QB a weak and short-lived man,
\par }{\QB with little power to understand judgement and laws.
\par }{\Q \VS{6}For even if a man is perfect among the sons of men,
\par }{\QB if the wisdom that comes from you is not with him, he will count for nothing.
\par }{\Q \VS{7}You chose me to be king of your people,
\par }{\QB and a judge for your sons and daughters.
\par }{\Q \VS{8}You gave a command to build a sanctuary on your holy mountain,
\par }{\QB and
\FTNT{†}{{\FR 9:8 }
Or,
a place of sacrifice
} an altar in the city where you pitch your tent,
\par }{\QB a copy of the holy tent which you prepared from the beginning.
\par }{\Q \VS{9}Wisdom is with you and knows your works,
\par }{\QB and was present when you were making the world,
\par }{\QB and understands what is pleasing in your eyes,
\par }{\QB and what is right according to your commandments.
\par }{\Q \VS{10}Send her from the holy heavens,
\par }{\QB and ask her to come from the throne of your glory,
\par }{\QB that being present with me she may work,
\par }{\QB and I may learn what is pleases you well.
\par }{\Q \VS{11}For she knows all things and understands,
\par }{\QB and she will guide me soberly in my actions.
\par }{\QB She will guard me in her glory.
\par }{\Q \VS{12}So my works will be acceptible.
\par }{\QB I will judge your people righteously,
\par }{\QB and I will be worthy of my father’s
\FTNT{‡}{{\FR 9:12 }
Gr.
thrones.
} throne.
\par }{\Q \VS{13}For what man will know the counsel of God?
\par }{\QB Or who will conceive what the Lord wills?
\par }{\Q \VS{14}For the thoughts of mortals are unstable,
\par }{\QB and our plans are prone to fail.
\par }{\Q \VS{15}For a corruptible body weighs down the soul.
\par }{\QB The earthy frame lies heavy on a mind that is full of cares.
\par }{\Q \VS{16}We can hardly guess the things that are on earth,
\par }{\QB and we find the things that are close at hand with labor;
\par }{\QB but who has traced out the things that are in the heavens?
\par }{\Q \VS{17}Who gained knowledge of your counsel, unless you gave wisdom,
\par }{\QB and sent your holy spirit from the highest?
\par }{\Q \VS{18}It was thus that the ways of those who are on earth were corrected,
\par }{\QB and men were taught the things that are pleasing to you.
\par }{\QB They were saved through wisdom.”
\par }{\BB \par }
\PoetryChap{10}{\Q \VerseOne{1}\FTNT{*}{{\FR 10:1 }
Gr.
She.
} Wisdom guarded to the end the first formed father of the world, that was created alone,
\par }{\QB And delivered him out of his own transgression,
\par }{\Q \VS{2}And gave him strength to get dominion over all things.
\par }{\Q \VS{3}But when an unrighteous man fell away from her in his anger,
\par }{\QB He perished himself in the rage wherewith he killed his brother.
\par }{\Q \VS{4}And when for his cause the earth was drowning with a flood,
\par }{\Q Wisdom again saved it,
\par }{\Q Guiding the righteous man’s course by a poor piece of wood.
\par }{\BB \par }{\Q \VS{5}Moreover, when nations consenting together in wickedness had been confounded,
\par }{\QB \FTNT{†}{{\FR 10:5 }
Gr.
She
} Wisdom knew the righteous man, and preserved him blameless to God,
\par }{\QB And kept him strong when his heart yearned toward his child.
\par }{\BB \par }{\Q \VS{6}While the ungodly were perishing,
\FTNT{‡}{{\FR 10:6 }
Gr.
she
} wisdom delivered a righteous man,
\par }{\QB When he fled from the fire that descended out of heaven on
\FTNT{§}{{\FR 10:6 }
That is,
the region of the five cities.
} Pentapolis.
\par }{\Q \VS{7}To whose wickedness a smoking waste still witnesses,
\par }{\QB And plants bearing fair fruit that comes not to ripeness;
\par }{\QB {\ADD{Yes and
}} a
\FTNT{*}{{\FR 10:7 }
Or,
distrustful
} disbelieving soul has a memorial
{\ADD{there,
}} a pillar of salt
{\ADD{still
}} standing.
\par }{\Q \VS{8}For having passed wisdom by,
\par }{\QB Not only were they disabled from recognising the things which are good,
\par }{\QB But they also left behind them
\FTNT{†}{{\FR 10:8 }
Or,
by their life
} for
{\ADD{human
}} life a monument of their folly;
\par }{\QB To the end that
\FTNT{‡}{{\FR 10:8 }
Gr.
wherein.
} where they
\FTNT{§}{{\FR 10:8 }
Gr.
stumbled.
} went astray they might fail even to be unseen:
\par }{\Q \VS{9}But wisdom delivered out of troubles those that waited on her.
\par }{\BB \par }{\Q \VS{10}When a righteous man was a fugitive from a brother’s wrath,
\FTNT{*}{{\FR 10:10 }
Gr.
she.
} wisdom guided him in straight paths;
\par }{\QB She showed him God’s kingdom, and gave him knowledge of holy things;
\par }{\QB She prospered him in his toils, and multiplied the fruits of his labor;
\par }{\Q \VS{11}When in their covetousness
{\ADD{men
}} dealt hardly with him,
\par }{\QB She stood by him and made him rich;
\par }{\Q \VS{12}She guarded him from enemies,
\par }{\QB And from those that lay in wait she kept him safe,
\par }{\QB And over his sore conflict she watched as judge,
\par }{\QB That he might know that godliness is more powerful than
\FTNT{†}{{\FR 10:12 }
Gr.
every one.
} all.
\par }{\BB \par }{\Q \VS{13}When a righteous man was sold,
\FTNT{‡}{{\FR 10:13 }
Gr.
she.
} wisdom forsook him not,
\par }{\QB But
\FTNT{§}{{\FR 10:13 }
Or,
from the sin
of his kindred...into a pit
} from sin she delivered him;
\par }{\QB She went down with him into a dungeon,
\par }{\Q \VS{14}And in bonds she left him not,
\par }{\QB Till she brought him the sceptre of a kingdom,
\par }{\QB And authority over those that dealt tyrannously with him;
\par }{\QB She showed them also to be false that had mockingly accused him,
\par }{\QB And gave him eternal glory.
\par }{\BB \par }{\Q \VS{15}\FTNT{*}{{\FR 10:15 }
Gr.
she.
} Wisdom delivered a holy people and a blameless seed from a nation of oppressors.
\par }{\Q \VS{16}She entered into the soul of a servant of the Lord,
\par }{\QB And withstood terrible kings in wonders and signs.
\par }{\Q \VS{17}She rendered to holy men a reward of their toils;
\par }{\QB She guided them along a marvelous way,
\par }{\QB And became to them a covering in the day-time,
\par }{\QB And a flame of stars through the night.
\par }{\Q \VS{18}She brought them over the Red sea,
\par }{\QB And led them through much water;
\par }{\Q \VS{19}But their enemies she drowned,
\par }{\QB And out of the bottom of the deep she cast them up.
\par }{\Q \VS{20}Therefore the righteous plundered the ungodly;
\par }{\QB And they sang praise to your holy name, O Lord,
\par }{\QB And extolled with one accord your hand that fought for them:
\par }{\Q \VS{21}Because wisdom opened the mouth of the mute,
\par }{\Q And made the tongues of babes to speak clearly.
\par }{\BB \par }
\PoetryChap{11}{\Q \VerseOne{1}She prospered their works in the hand of a holy prophet.
\par }{\BB \par }{\Q \VS{2}They traveled through a desert without inhabitant,
\par }{\QB And in trackless regions they pitched their tents.
\par }{\Q \VS{3}They withstood enemies, and
\FTNT{*}{{\FR 11:3 }
Or,
took vengeance on foes
} repelled foes.
\par }{\Q \VS{4}They thirsted, and they called upon you,
\par }{\QB And there was given them water out of
\FTNT{†}{{\FR 11:4 }
Or,
the steep rock
} the
\FTNT{‡}{{\FR 11:4 }
See Deuteronomy 8:15; Psalm 114:8.
} flinty rock,
\par }{\QB And healing of their thirst out of the hard stone.
\par }{\Q \VS{5}For by what things their foes were punished,
\par }{\QB By these they in their need were benefited.
\par }{\Q \VS{6}When enemies were troubled with clotted blood
\par }{\QB instead of a river’s ever-flowing fountain,
\par }{\Q \VS{7}to rebuke the decree for the slaying of babies,
\par }{\QB you gave them abundant water beyond all hope,
\par }{\Q \VS{8}having shown by the thirst which they had suffered
\par }{\QB how you punished the adversaries.
\par }{\Q \VS{9}For when they were tried, although chastened in mercy,
\par }{\QB they learned how the ungodly were tormented, being judged with wrath.
\par }{\Q \VS{10}For you tested these as a father admonishing them;
\par }{\QB but you searched out those as a stern king condemning them.
\par }{\Q \VS{11}Yes and whether they were far off or near,
\par }{\QB they were equally distressed;
\par }{\Q \VS{12}for a double grief seized them,
\par }{\QB and a groaning at the memory of things past.
\par }{\Q \VS{13}For when they heard that through their own punishments the others benefited,
\par }{\QB they recognized the Lord.
\par }{\Q \VS{14}For him who long before was thrown out and exposed they stopped mocking.
\par }{\QB In the end of what happened, they marveled,
\par }{\QB having thirsted in another manner than the righteous.
\par }{\Q \VS{15}But in return for the senseless imaginings of their unrighteousness,
\par }{\QB wherein they were led astray to worship irrational reptiles and wretched vermin,
\par }{\QB you sent upon them a multitude of irrational creatures for vengeance;
\par }{\QB \VS{16}that they might learn that by what things a man sins, by these he is punished.
\par }{\Q \VS{17}For your all-powerful hand
\par }{\QB that created the world out of formless matter
\par }{\QB didn’t lack means to send upon them a multitude of bears, fierce lions,
\par }{\QB \VS{18}or newly-created and unknown wild beasts, full of rage,
\par }{\QB either breathing out a blast of fiery breath,
\par }{\QB or belching out smoke,
\par }{\QB or flashing dreadful sparks from their eyes;
\par }{\Q \VS{19}which had power not only to consume them by their violence,
\par }{\QB but to destroy them even by the terror of their sight.
\par }{\Q \VS{20}Yes and without these they might have fallen by a single breath,
\par }{\QB being pursued by Justice, and scattered abroad by the breath of your power;
\par }{\QB but you arranged all things by measure, number, and weight.
\par }{\BB \par }{\Q \VS{21}For to be greatly strong is yours at all times.
\par }{\QB Who could withstand the might of your arm?
\par }{\Q \VS{22}Because the whole world before you is as a grain in a balance,
\par }{\QB and as a drop of dew that comes down upon the earth in the morning.
\par }{\Q \VS{23}But you have mercy on all men, because you have power to do all things,
\par }{\QB and you overlook the sins of men to the end that they may repent.
\par }{\Q \VS{24}For you love all things that are,
\par }{\QB and abhor none of the things which you made;
\par }{\QB For you never would have formed anything if you hated it.
\par }{\Q \VS{25}How would anything have endured unless you had willed it?
\par }{\QB Or that which was not called by you, how would it have been preserved?
\par }{\Q \VS{26}But you spare all things, because they are yours,
\par }{\QB O Sovereign Lord, you lover of lives.

\par }\PoetryChap{12}{\Q \VerseOne{1}For your incorruptible spirit is in all things.
\par }{\Q \VS{2}Wherefore you convict little by little those who fall from the right way,
\par }{\QB and, putting them in remembrance by the things wherein they sin, you admonish them,
\par }{\QB that escaping from their wickedness they may believe in you, O Lord.
\par }{\BB \par }{\Q \VS{3}For truly the old inhabitants of your holy land,
\par }{\QB \VS{4}hating them because they practiced detestable works of enchantments and unholy rites—
\par }{\QB \VS{5}\FTNT{*}{{\FR 12:5 }
The words rendered
slaughters
and
impious
in verses 5 and 6 differ but slightly from the readings of the Greek text, which here yield no sense.
}merciless slaughters of children
\par }{\QB and sacrificial banquets of men’s flesh and of blood—
\par }{\QB \VS{6}allies in an impious fellowship,
\par }{\QB and murderers of their own helpless babes,
\par }{\QB it was your counsel to destroy by the hands of our fathers;
\par }{\QB \VS{7}that the land which in your sight is most precious of all
\par }{\QB might receive a worthy colony of God’s servants.\FTNT{†}{{\FR 12:7 }
Or,
children
}
\par }{\Q \VS{8}Nevertheless you even spared these as men,
\par }{\QB and you sent hornets\FTNT{‡}{{\FR 12:8 }
Or,
wasps
} as forerunners of your army,
\par }{\QB to cause them to perish little by little;
\par }{\Q \VS{9}Not that you were unable to subdue the ungodly under the hand of the righteous in battle,
\par }{\QB or by terrible beasts or by a stern word to make away with them at once;
\par }{\Q \VS{10}but judging them little by little you gave them a chance to repent,
\par }{\QB not being ignorant that their nature by birth was evil,
\par }{\QB their wickedness inborn,
\par }{\QB and that their manner of thought would never be changed.
\par }{\Q \VS{11}For they were a cursed seed from the beginning.
\par }{\QB It wasn’t through fear of any that you left them unpunished for their sins.
\par }{\BB \par }{\Q \VS{12}For who will say, “What have you done?”
\par }{\QB Or “Who will withstand your judgement?”
\par }{\Q Who will accuse you for the perishing of nations which you caused?
\par }{\QB Or who will come and stand before you as an avenger for unrighteous men?
\par }{\Q \VS{13}For there isn’t any God beside you that cares for all,
\par }{\QB that you might show that you didn’t judge unrighteously.
\par }{\Q \VS{14}No king or prince will be able to look you in the face
\par }{\QB for those whom you have punished.
\par }{\Q \VS{15}But being righteous, you rule all things righteously,
\par }{\QB deeming it a thing alien from your power
\par }{\QB to condemn one that does not deserve to be punished.
\par }{\Q \VS{16}For your strength is the beginning of righteousness,
\par }{\QB and your sovereignty over all makes you to forbear all.
\par }{\Q \VS{17}For when men don’t believe that you are perfect in power, you show your strength,
\par }{\QB \FTNT{§}{{\FR 12:17 }
The Greek text here is perhaps corrupt.
}and in dealing with those who think this, you confuse their boldness.
\par }{\Q \VS{18}But you, being sovereign over strength, judge in gentleness,
\par }{\QB and with great forbearance do you govern us;
\par }{\QB for the power is yours whenever you desire it.
\par }{\BB \par }{\Q \VS{19}But you taught your people by such works as these,
\par }{\QB how the righteous must be a lover of men.
\par }{\Q You made your sons to have good hope,
\par }{\QB because you give repentance when men have sinned.
\par }{\Q \VS{20}For if on those who were enemies of your servants\FTNT{*}{{\FR 12:20 }
Or,
children
} and due to death
\par }{\QB you took vengeance with so great deliberation and indulgence,
\par }{\QB giving them times and opportunities when they might escape from their wickedness;
\par }{\Q \VS{21}with how great carefulness you judged your sons,
\par }{\QB to whose fathers you gave oaths and covenants of good promises!
\par }{\Q \VS{22}Therefore while you chasten us, you scourge our enemies ten thousand times more,
\par }{\QB to the intent that we may ponder your goodness when we judge,
\par }{\QB and when we are judged may look for mercy.
\par }{\Q \VS{23}Wherefore also the unrighteous that lived in folly of life,
\par }{\QB you tormented through their own abominations.
\par }{\Q \VS{24}For truly they went astray very far in the ways of error,
\par }{\QB Taking as gods those animals\FTNT{†}{{\FR 12:24 }
Gr.
living creatures:
and so elsewhere in this book.
} which even among their enemies were held in dishonor,
\par }{\QB deceived like foolish babes.
\par }{\Q \VS{25}Therefore, as to unreasoning children, you sent your judgement to mock them.
\par }{\Q \VS{26}But those who would not be admonished by a mocking correction as of children
\par }{\QB will experience a judgement worthy of God.
\par }{\Q \VS{27}For through the sufferings they were indignant of,
\par }{\QB being punished in these creatures which they supposed to be gods,
\par }{\QB they saw and recognized as the true God him whom they refused to know.
\par }{\QB Therefore also the result of condemnation came upon them.
\par }{\BB \par }
\PoetryChap{13}{\Q \VerseOne{1}For truly all men who had no perception of God were vain by nature,
\par }{\QB and didn’t gain power to know him who exists from the good things that are seen.
\par }{\QB They didn’t recognize the architect from his works.
\par }{\Q \VS{2}But they thought that either fire, or wind, or swift air,
\par }{\QB or circling stars, or raging water, or luminaries of heaven
\par }{\QB were gods that rule the world.
\par }{\Q \VS{3}If it was through delight in their beauty that they took them to be gods,
\par }{\QB let them know how much better their Sovereign Lord is than these,
\par }{\QB for the first author of beauty created them.
\par }{\Q \VS{4}But if it was through astonishment at their power and influence,
\par }{\QB then let them understand from them how much more powerful he who formed them is.
\par }{\Q \VS{5}For from the greatness of the beauty of created things,
\par }{\QB mankind forms the corresponding image of their Maker.\FTNT{*}{{\FR 13:5 }
Gr.
is the first maker of them seen.}
\par }{\Q \VS{6}But yet for these men there is but small blame,
\par }{\QB for they too perhaps go astray
\par }{\QB while they are seeking God and desiring to find him.
\par }{\Q \VS{7}For they diligently search while living among his works,
\par }{\QB and they trust their sight that the things that they look at are beautiful.
\par }{\Q \VS{8}But again even they are not to be excused.
\par }{\Q \VS{9}For if they had power to know so much,
\par }{\QB that they should be able to explore the world,
\par }{\QB how is it that they didn’t find the Sovereign Lord sooner?
\par }{\BB \par }{\Q \VS{10}But miserable
\FTNT{†}{{\FR 13:10 }
Or, are
}
{\ADD{were
}} they, and
\FTNT{‡}{{\FR 13:10 }
Or,
among
} in dead things
\FTNT{§}{{\FR 13:10 }
Or, are
}
{\ADD{were
}} their hopes,
\par }{\QB Who called them gods which are works of men’s hands,
\par }{\QB gold and silver, skillfully made, and likenesses of animals,
\par }{\QB or a useless stone, the work of an ancient hand.
\par }{\Q \VS{11}Yes and if some
\FTNT{*}{{\FR 13:11 }
Gr.
carpenter who is a woodcutter.
} woodcutter, having sawn down a
\FTNT{†}{{\FR 13:11 }
Gr.
plant.
The Greek word, slightly changed, would mean
trunk
} tree that is easily moved,
\par }{\QB skillfully strips away all its bark,
\par }{\QB and fashioning it in attractive form, makes a useful vessel to serve his life’s needs.
\par }{\Q \VS{12}Burning the scraps from his handiwork to cook his food,
\par }{\QB he eats his fill.
\par }{\Q \VS{13}Taking a discarded scrap which served no purpose,
\par }{\QB a crooked piece of wood and full of knots,
\par }{\QB carves it with the diligence of his idleness,
\par }{\QB and shapes it by the skill of his
\FTNT{‡}{{\FR 13:13 }
Or,
leisure
} idleness.
\par }{\Q He shapes it in the image of a man,
\par }{\QB \VS{14}or makes it like some paltry animal,
\par }{\QB smearing it with something red, painting it red,
\par }{\QB and smearing over every stain in it.
\par }{\Q \VS{15}Having made a worthy chamber for it,
\par }{\QB he sets it in a wall, securing it with iron.
\par }{\Q \VS{16}He plans for it that it may not fall down,
\par }{\QB knowing that it is unable to help itself
\par }{\QB (for truly it is an image, and needs help).
\par }{\Q \VS{17}When he makes his prayer concerning goods and his marriage and children,
\par }{\QB he is not ashamed to speak to that which has no life.
\par }{\Q \VS{18}Yes, for health, he calls upon that which is weak.
\par }{\QB For life, he implores that which is dead.
\par }{\QB For aid, he supplicates that which has no experience.
\par }{\QB For a good journey, he asks that which can’t so much as move a step.
\par }{\Q \VS{19}And for profit in business and good success of his hands,
\par }{\QB he asks ability from that which has hands with no ability.

\par }\PoetryChap{14}{\Q \VerseOne{1}Again, one preparing to sail, and about to journey over raging waves,
\par }{\QB calls upon a piece of wood more rotten than the vessel that carries him.
\par }{\Q \VS{2}For the hunger for profit planned it,
\par }{\QB and wisdom was the craftsman who built it.
\par }{\Q \VS{3}Your providence, O Father, guides it along,
\par }{\QB because even in the sea you gave a way,
\par }{\QB and in the waves a sure path,
\par }{\Q \VS{4}showing that you can save out of every danger,
\par }{\QB that even a man without skill may put to sea.
\par }{\Q \VS{5}It is your will that the works of your wisdom should be not idle.
\par }{\QB Therefore men also entrust their lives to a little piece of wood,
\par }{\QB and passing through the surge on a raft come safely to land.
\par }{\Q \VS{6}For
\FTNT{*}{{\FR 14:6 }
The Greek text here may be corrupt.
} in the old time also, when proud giants were perishing,
\par }{\QB the hope of the world, taking refuge on a raft,
\par }{\QB your hand guided the seed of generations of the race of men.
\par }{\Q \VS{7}For blessed is wood through which comes righteousness;
\par }{\QB \VS{8}but the idol made with hands is accursed, itself and he that made it;
\par }{\QB because his was the working, and the corruptible thing was called a god:
\par }{\Q \VS{9}For both the ungodly and his ungodliness are alike hateful to God;
\par }{\QB \VS{10}for truly the deed will be punished together with him who committed it.
\par }{\QB \VS{11}Therefore also there will be a visitation among the idols of the nation,
\par }{\QB because, though formed of things which God created, they were made an abomination,
\par }{\QB stumbling blocks to the souls of men,
\par }{\QB and a snare to the feet of the foolish.
\par }{\BB \par }{\Q \VS{12}For the devising of idols was the beginning of fornication,
\par }{\QB and the invention of them the corruption of life.
\par }{\Q \VS{13}For they didn’t exist from the beginning, and they won’t exist forever.
\par }{\Q \VS{14}For by the vaingloriousness of men they entered into the world,
\par }{\QB and therefore was a speedy end is planned for them.
\par }{\Q \VS{15}For a father worn with untimely grief,
\par }{\QB making an image of the child quickly taken away,
\par }{\QB now honored him as a god which was then a dead man,
\par }{\QB and delivered to those that were under him mysteries and solemn rites.
\par }{\Q \VS{16}Afterward the ungodly custom, in process of time grown strong, was kept as a law,
\par }{\QB and the engraved images received worship by the commandments of princes.
\par }{\Q \VS{17}And when men could not honor them in presence because they lived far off,
\par }{\QB imagining the likeness from afar,
\par }{\QB they made a visible image of the king whom they honored,
\par }{\QB that by their zeal they might flatter the absent as if present.
\par }{\Q \VS{18}But worship was raised to a yet higher pitch, even by those who didn’t know him,
\par }{\QB urged forward by the ambition of the architect;
\par }{\Q \VS{19}for he, wishing perhaps to please one in authority,
\par }{\QB used his art to force the likeness toward a greater beauty.
\par }{\Q \VS{20}So the multitude, allured by reason of the grace of his handiwork,
\par }{\QB now consider an object of devotion him that a little before was honored as a man.
\par }{\Q \VS{21}And this became an ambush,
\par }{\QB because men, in bondage either to calamity or to tyranny,
\par }{\QB invested stones and stocks with the incommunicable Name.
\par }{\BB \par }{\Q \VS{22}Afterward it was not enough for them to go astray concerning the knowledge of God;
\par }{\QB but also, while they live in a great war of ignorance that they call a multitude of evils peace.
\par }{\Q \VS{23}For either slaughtering children in solemn rites, or celebrating secret mysteries,
\par }{\QB or holding frantic revels of strange ordinances,
\par }{\Q \VS{24}no longer do they guard either life or purity of marriage,
\par }{\QB but one brings upon another either death by treachery, or anguish by adultery.
\par }{\Q \VS{25}And all things confusedly are filled with blood and murder, theft and deceit,
\par }{\QB corruption, faithlessness, tumult, perjury,
\par }{\Q \VS{26}confusion about what is good, forgetfulness of favors,
\par }{\QB ingratitude for benefits,
\par }{\QB defiling of souls, confusion of sex,
\par }{\QB disorder in marriage, adultery and wantonness.
\par }{\QB \VS{27}For the worship of idols that may not be named
\XREF{✡}{{\XO 14:27 }
Exodus 23:13; Psalm 16:4; Hosea 2:17; Wisdom 14:21}
\par }{\QB is a beginning and cause and end of every evil.
\par }{\Q \VS{28}For their worshipers either make merry to madness, or prophesy lies,
\par }{\QB or live unrighteously, or lightly commit perjury.
\par }{\Q \VS{29}For putting their trust in lifeless idols,
\par }{\QB when they have sworn a wicked oath, they expect not to suffer harm.
\par }{\Q \VS{30}But for both, the just doom will pursue them,
\par }{\QB because they had evil thoughts of God by giving heed to idols,
\par }{\QB and swore unrighteously in deceit through contempt for holiness.
\par }{\Q \VS{31}For it is not the power of them by whom men swear,
\par }{\QB but it is the just penalty for those who sin
\par }{\QB that always visits the transgression of the unrighteous.
\par }{\BB \par }
\PoetryChap{15}{\Q \VerseOne{1}But you, our God, are gracious and true,
\par }{\QB patient, and in mercy ordering all things.
\par }{\Q \VS{2}For even if we sin, we are yours, knowing your dominion;
\par }{\QB but we will not sin, knowing that we have been accounted yours.
\par }{\Q \VS{3}For to be acquainted with you is
\FTNT{*}{{\FR 15:3 }
Gr.
entire.
} perfect righteousness,
\par }{\QB and to know your dominion is the root of immortality.
\par }{\Q \VS{4}For we weren’t led astray by any evil plan of men’s,
\par }{\QB nor yet by painters’ fruitless labor,
\par }{\QB a form stained with varied colors,
\par }{\Q \VS{5}the sight of which leads fools into
\FTNT{†}{{\FR 15:5 }
Some authorities read
reproach.
} lust.
\par }{\QB Their desire is for the breathless form of a dead image.
\par }{\Q \VS{6}Lovers of evil things, and worthy of such hopes,
\par }{\QB are those who make, desire, and worship them.
\par }{\BB \par }{\Q \VS{7}For a potter, kneading soft earth,
\par }{\QB laboriously molds each article for our service.
\par }{\Q He fashions out of the same clay
\par }{\QB both the vessels that minister to clean uses, and those of a contrary sort,
\par }{\QB all in like manner.
\par }{\Q What shall be the use of each article of either sort,
\par }{\QB the potter is the judge.
\par }{\Q \VS{8}Also, laboring to an evil end, he molds a vain god out of the same clay,
\par }{\QB he who, having but a little before been made of earth,
\par }{\QB after a short space goes his way to the earth out of which he was taken,
\par }{\QB when he is required to render back the
\FTNT{‡}{{\FR 15:8 }
Or,
life
} soul which was lent him.
\par }{\Q \VS{9}However he has anxious care,
\par }{\QB not because his powers must fail,
\par }{\QB nor because his span of life is short;
\par }{\Q But he compares himself with goldsmiths and silversmiths,
\par }{\QB and he imitates molders in
\FTNT{§}{{\FR 15:9 }
Or,
copper
} brass,
\par }{\QB and esteems it glory that he molds counterfeits.
\par }{\Q \VS{10}His heart is ashes.
\par }{\QB His hope is of less value than earth.
\par }{\QB His life is of less honor than clay;
\par }{\Q \VS{11}because he was ignorant of him who molded him,
\par }{\QB and of him that inspired into him
\FTNT{*}{{\FR 15:11 }
Gr.
a soul that moves to activity.
} an active
\FTNT{†}{{\FR 15:11 }
Or,
life
} soul,
\par }{\QB and breathed into him a vital spirit.
\par }{\Q \VS{12}But
\FTNT{‡}{{\FR 15:12 }
Some authorities read
they accounted.
} he accounted our life to be a game,
\par }{\QB and our
\FTNT{§}{{\FR 15:12 }
Or,
way of life
} lifetime a festival for profit;
\par }{\Q for, he says, one must get gain however one can, even if it is by evil.
\par }{\Q \VS{13}For this man beyond all others knows that he sins,
\par }{\QB out of earthy matter making brittle vessels and graven images.
\par }{\Q \VS{14}But most foolish and more miserable than a baby,
\par }{\QB are the enemies of your people, who oppressed them;
\par }{\Q \VS{15}because they even considered all the idols of the nations to be gods,
\par }{\QB which have neither the use of eyes for seeing,
\par }{\QB nor nostrils for drawing breath,
\par }{\QB nor ears to hear,
\par }{\QB nor fingers for handling,
\par }{\QB and their feet are helpless for walking.
\par }{\Q \VS{16}For a man made them,
\par }{\QB and one whose own spirit is borrowed molded them;
\par }{\QB for no one has power as a man to mold a god like to himself.
\par }{\Q \VS{17}But, being mortal, he makes a dead thing by the work of lawless hands;
\par }{\QB for he is better than the objects of his worship,
\par }{\QB since he indeed had life, but they never did.
\par }{\BB \par }{\Q \VS{18}Yes, and they worship the creatures that are most hateful,
\par }{\QB for, being compared as to lack of sense, these are worse than all others;
\par }{\Q \VS{19}Neither, as seen beside other creatures, are they beautiful, so that one should desire them,
\par }{\QB but they have escaped both the praise of God and his blessing.

\par }\PoetryChap{16}{\Q \VerseOne{1}For this cause, they were deservedly punished through creatures like those which they worship,
\par }{\QB and tormented through a multitude of vermin.
\par }{\Q \VS{2}Instead of this punishment, you, giving benefits to your people,
\par }{\QB prepared quails for food,
\par }{\Q a delicacy to satisfy the desire of their appetite,
\par }{\Q \VS{3}to the end that your enemies, desiring food,
\par }{\QB might for the hideousness of the creatures sent among them,
\par }{\QB loathe even the necessary appetite;
\par }{\Q but these, your people, having for a short time suffered lack,
\par }{\QB might even partake of delicacies.
\par }{\Q \VS{4}For it was necessary that inescapable lack should come upon those oppressors,
\par }{\QB but that to these it should only be showed how their enemies were tormented.
\par }{\Q \VS{5}For even when terrible raging of wild beasts came upon your people,
\par }{\QB and they were perishing by the bites of crooked serpents,
\par }{\QB your wrath didn’t continue to the uttermost;
\par }{\Q \VS{6}but for admonition were they troubled for a short time,
\par }{\QB having a token of salvation
\par }{\QB to put them in remembrance of the commandment of your law;
\par }{\Q \VS{7}for he who turned toward it was not saved because of that which was seen,
\par }{\QB but because of you, the Savior of all.
\par }{\Q \VS{8}Yes, and in this you persuaded our enemies
\par }{\QB that you are he who delivers out of every evil.
\par }{\Q \VS{9}For the bites of locusts and flies truly killed them.
\par }{\QB No healing for their life was found,
\par }{\QB because they were worthy to be punished by such things.
\par }{\Q \VS{10}But your children weren’t overcome by the very teeth of venomous dragons,
\par }{\QB for your mercy passed by where they were and healed them.
\par }{\Q \VS{11}For they were bitten to put them in remembrance of your oracles,
\par }{\QB and were quickly saved, lest, falling into deep forgetfulness,
\par }{\QB they should become unable to respond to your kindness.
\par }{\Q \VS{12}For truly it was neither herb nor mollifying plaster that cured them,
\par }{\QB but your word, O Lord, which heals all things.
\par }{\Q \VS{13}For you have authority over life and death,
\par }{\QB and you lead down to the gates of Hades, and lead up again.
\par }{\Q \VS{14}But though a man kills by his wickedness,
\par }{\QB he can’t retrieve the spirit that has departed
\par }{\QB or release the imprisoned soul.
\par }{\BB \par }{\Q \VS{15}But it is not possible to escape your hand;
\par }{\Q \VS{16}for ungodly men, refusing to know you, were scourged in the strength of your arm,
\par }{\QB pursued with strange rains and hails and relentless storms,
\par }{\QB and utterly consumed with fire.
\par }{\Q \VS{17}For, what was most marvelous,
\par }{\QB in the water which quenches all things, the fire burned hotter;
\par }{\QB for the world fights for the righteous.
\par }{\Q \VS{18}For at one time the flame lost its fierceness,
\par }{\QB that it might not burn up the creatures sent against the ungodly,
\par }{\QB but that these themselves as they looked might see that they were chased through the judgement of God.
\par }{\Q \VS{19}At another time even in the midst of water it burns above the power of fire,
\par }{\QB that it may destroy the produce of an unrighteous land.
\par }{\Q \VS{20}Instead of these things, you gave your people angels’ food to eat,
\par }{\QB and you provided ready-to-eat bread for them from heaven without toil,
\par }{\QB having the virtue of every pleasant flavor,
\par }{\QB and agreeable to every taste.
\par }{\Q \VS{21}For your nature showed your sweetness toward your children,
\par }{\QB while that bread, serving the desire of the eater,
\par }{\QB changed itself according to every man’s choice.
\par }{\Q \VS{22}But snow and ice endured fire, and didn’t melt,
\par }{\QB that people might know that fire was destroying the fruits of the enemies,
\par }{\QB burning in the hail and flashing in the rains;
\par }{\Q \VS{23}and that this fire again, in order that righteous people may be nourished,
\par }{\QB has even forgotten its own power.
\par }{\Q \VS{24}For the creation, ministering to you its maker,
\par }{\QB strains its force against the unrighteous for punishment
\par }{\QB and slackens it on behalf of those who trust in you, for kindness.
\par }{\Q \VS{25}Therefore at that time also, converting itself into all forms,
\par }{\QB it ministered to your all-nourishing bounty,
\par }{\QB according to the desire of those who had need,
\par }{\Q \VS{26}that your sons, whom you loved, O Lord, might learn
\par }{\QB that it is not the growth of crops that nourishes a man,
\par }{\QB but that your word preserves those who trust you.
\par }{\Q \VS{27}For that which was not marred by fire,
\par }{\QB when it was simply warmed by a faint sunbeam melted away,
\par }{\Q \VS{28}that it might be known that we must rise before the sun to give you thanks,
\par }{\QB and must plead with you at the dawning of the light;
\par }{\Q \VS{29}for the hope of the unthankful will melt as the winter’s hoar frost,
\par }{\QB and will flow away as water that has no use.
\par }{\BB \par }
\PoetryChap{17}{\Q \VerseOne{1}For your judgments are great, and hard to interpret;
\par }{\QB therefore undisciplined souls went astray.
\par }{\Q \VS{2}For when lawless men had supposed that they held a holy nation in their power,
\par }{\QB they, prisoners of darkness, and bound in the fetters of a long night,
\par }{\QB kept close beneath their roofs,
\par }{\QB lay exiled from the eternal providence.
\par }{\Q \VS{3}For while they thought that they were unseen in
{\ADD{their
}} secret sins,
\par }{\QB they were divided from one another by a dark curtain of forgetfulness,
\par }{\QB stricken with terrible awe, and very troubled by apparitions.
\par }{\Q \VS{4}For neither did the dark recesses that held them guard them from fears,
\par }{\QB But terrifying sounds rang around them,
\par }{\QB and dismal phantoms appeared with unsmiling faces.
\par }{\Q \VS{5}And no force of fire prevailed to give light,
\par }{\QB neither were the brightest flames of the stars strong enough to illuminate that gloomy night;
\par }{\Q \VS{6}but only the glimmering of a self-kindled fire appeared to them, full of fear.
\par }{\QB In terror, they considered the things which they saw
\par }{\QB to be worse than that sight, on which they could not gaze.
\par }{\Q \VS{7}The mockeries of their magic arts were powerless, now,
\par }{\QB and a shameful rebuke of their boasted understanding:
\par }{\Q \VS{8}For those who promised to drive away terrors and disorders from a sick soul,
\par }{\QB these were sick with a ludicrous fearfulness.
\par }{\Q \VS{9}For even if no troubling thing frighted them,
\par }{\QB yet, scared with the creeping of vermin and hissing of serpents,
\par }{\Q \VS{10}they perished trembling in fear,
\par }{\QB refusing even to look at the air, which could not be escaped on any side.
\par }{\Q \VS{11}For wickedness, condemned by a witness within, is a coward thing,
\par }{\QB and, being pressed hard by conscience, always has added forecasts of the worst.
\par }{\QB \VS{12}For fear is nothing else but a surrender of the help which reason offers;
\par }{\QB \VS{13}and from within, the expectation of being less
\par }{\QB makes of greater account the ignorance of the cause that brings the torment.
\par }{\Q \VS{14}But they, all through the night which was powerless indeed,
\par }{\QB and which came upon them out of the recesses of powerless Hades,
\par }{\QB sleeping the same sleep,
\par }{\QB \VS{15}now were haunted by monstrous apparitions,
\par }{\QB and now were paralyzed by their soul’s surrendering;
\par }{\QB for sudden and unexpected fear came upon them.
\par }{\Q \VS{16}So then whoever it might be, sinking down in his place,
\par }{\QB was kept captive, shut up in that prison which was not barred with iron;
\par }{\Q \VS{17}for whether he was a farmer, or a shepherd,
\par }{\QB or a laborer whose toils were in the wilderness,
\par }{\QB he was overtaken, and endured that inevitable necessity;
\par }{\QB for they were all bound with one chain of darkness.
\par }{\Q \VS{18}Whether there was a whistling wind,
\par }{\QB or a melodious sound of birds among the spreading branches,
\par }{\QB or a measured fall of water running violently,
\par }{\QB \VS{19}or a harsh crashing of rocks hurled down,
\par }{\QB or the swift course of animals bounding along unseen,
\par }{\QB or the voice of wild beasts harshly roaring,
\par }{\QB or an echo rebounding from the hollows of the mountains,
\par }{\QB all these things paralyzed them with terror.
\par }{\Q \VS{20}For the whole world was illuminated with clear light,
\par }{\QB and was occupied with unhindered works,
\par }{\QB \VS{21}while over them alone was spread a heavy night,
\par }{\QB an image of the darkness that should afterward receive them;
\par }{\QB but to themselves, they were heavier than darkness.

\par }\PoetryChap{18}{\Q \VerseOne{1}But for your holy ones there was great light.
\par }{\QB Their enemies, hearing their voice but seeing not their form,
\par }{\QB counted it a happy thing that they too had suffered,
\par }{\Q \VS{2}yet for that they do not hurt them, though wronged by them before, they are thankful;
\par }{\QB and because they had been at variance with them, they begged for pardon.
\par }{\Q \VS{3}Whereas you provided a burning pillar of fire,
\par }{\QB to be a guide for your people’s unknown journey,
\par }{\QB and a harmless sun for their glorious exile.
\par }{\Q \VS{4}For the Egyptians well deserved to be deprived of light and imprisoned by darkness,
\par }{\QB they who had imprisoned your children,
\par }{\QB through whom the incorruptible light of the law was to be given to the race of men.
\par }{\BB \par }{\Q \VS{5}After they had taken counsel to kill the babes of the holy ones,
\par }{\QB and when a single child had been abandoned and saved to convict them of their sin,
\par }{\QB you took away from them their multitude of children,
\par }{\QB and destroyed all their army together in a mighty flood.
\par }{\Q \VS{6}Our fathers were made aware of that night beforehand,
\par }{\QB that, having sure knowledge, they might be cheered by the oaths which they had trusted.
\par }{\Q \VS{7}Salvation of the righteous and destruction of the enemies was expected by your people.
\par }{\Q \VS{8}For as you took vengeance on the adversaries,
\par }{\QB by the same means, calling us to yourself, you glorified us.
\par }{\Q \VS{9}For holy children of good men offered sacrifice in secret,
\par }{\QB and with one consent they took upon themselves the covenant of the divine law,
\par }{\QB that they would partake alike in the same good things and the same perils,
\par }{\QB the fathers already leading the sacred songs of praise.
\par }{\Q \VS{10}But the discordant cry of the enemies echoed back,
\par }{\QB and a pitiful voice of lamentation for children was spread abroad.
\par }{\Q \VS{11}Both servant and master were punished with the same just doom,
\par }{\QB and the commoner suffering the same as king;
\par }{\Q \VS{12}Yes, they all together, under one form of death,
\par }{\QB had corpses without number.
\par }{\Q For the living were not sufficient even to bury them,
\par }{\QB Since at a single stroke, their most cherished offspring was consumed.
\par }{\Q \VS{13}For while they were disbelieving all things by reason of the enchantments,
\par }{\QB upon the destruction of the firstborn they confessed the people to be God’s son.
\par }{\Q \VS{14}For while peaceful silence wrapped all things,
\par }{\QB and night in her own swiftness was in mid course,
\par }{\Q \VS{15}your all-powerful word leaped from heaven out of the royal thrones,
\par }{\QB a stern warrior, into the midst of the doomed land,
\par }{\Q \VS{16}bearing as a sharp sword your authentic commandment,
\par }{\QB and standing, it filled all things with death,
\par }{\QB and while it touched the heaven it stood upon the earth.
\par }{\Q \VS{17}Then immediately apparitions in dreams terribly troubled them,
\par }{\QB and unexpected fears came upon them.
\par }{\Q \VS{18}And each, one thrown here half dead, another there,
\par }{\QB made known why he was dying;
\par }{\Q \VS{19}for the dreams, disturbing them, forewarned them of this,
\par }{\QB that they might not perish without knowing why they were afflicted.
\par }{\BB \par }{\Q \VS{20}But experience of death also touched the righteous,
\par }{\QB and a multitude were destroyed in the wilderness,
\par }{\QB however the wrath didn’t last long.
\par }{\Q \VS{21}For a blameless man hurried to be their champion,
\par }{\QB bringing the weapon of his own ministry,
\par }{\QB prayer and the atoning sacrifice of incense,
\par }{\QB he withstood the indignation, and set an end to the calamity,
\par }{\QB showing that he was your servant.
\par }{\Q \VS{22}And he overcame the anger,
\par }{\QB not by strength of body, not by force of weapons,
\par }{\QB but he subdued him who was punishing by word,
\par }{\QB by bringing to remembrance oaths and covenants made with the fathers.
\par }{\Q \VS{23}For when the dead had already fallen in heaps one upon another,
\par }{\QB standing between he stopped the wrath,
\par }{\QB and cut off the way to the living.
\par }{\Q \VS{24}For the whole world was upon his long robe,
\par }{\QB and the glories of the fathers were upon the engraving of the four rows of precious stones,
\par }{\QB and your majesty was upon the diadem of his head.
\par }{\Q \VS{25}The destroyer yielded to these, and they feared;
\par }{\QB for it was enough only to test the wrath.
\par }{\BB \par }
\PoetryChap{19}{\Q \VerseOne{1}But indignation without mercy came upon the ungodly to the end;
\par }{\QB For God also foreknew their future,
\par }{\Q \VS{2}how, having changed their minds to let your people go,
\par }{\QB and having sped them eagerly on their way,
\par }{\QB they would change their minds and pursue them.
\par }{\Q \VS{3}For while they were yet in the midst of their mourning,
\par }{\QB and lamenting at the graves of the dead,
\par }{\QB they drew upon themselves another counsel of folly,
\par }{\QB and pursued as fugitives those whom they had begged to leave and driven out.
\par }{\Q \VS{4}For the doom which they deserved was drawing them to this end,
\par }{\QB and it made them forget the things that had happened to them,
\par }{\QB that they might fill up the punishment which was yet lacking to their torments,
\par }{\QB \VS{5}and that your people might journey on by a marvelous road,
\par }{\QB but they themselves might find a strange death.
\par }{\BB \par }{\Q \VS{6}For the whole creation, each part in its diverse kind, was made new again,
\par }{\QB complying with your commandments,
\par }{\QB that your servants might be kept unharmed.
\par }{\Q \VS{7}Then the cloud that overshadowed the camp was seen,
\par }{\QB and dry land rising up out of what had been water,
\par }{\QB out of the Red sea an unhindered highway,
\par }{\QB and a grassy plain out of the violent surge,
\par }{\QB \VS{8}by which they passed over with all their army,
\par }{\QB these who were covered with your hand,
\par }{\QB having seen strange marvels.
\par }{\Q \VS{9}For like horses they roamed at large,
\par }{\QB and they skipped about like lambs,
\par }{\QB praising you, O Lord, who was their deliverer.
\par }{\Q \VS{10}For they still remembered the things that happened in the time of their sojourning,
\par }{\QB how instead of bearing cattle, the land brought forth lice,
\par }{\QB and instead of fish, the river spewed out a multitude of frogs.
\par }{\Q \VS{11}But afterwards, they also saw a new kind of birds,
\par }{\QB when, led on by desire, they asked for luxurious dainties;
\par }{\Q \VS{12}For, to comfort them, quails came up for them from the sea.
\par }{\BB \par }{\Q \VS{13}Punishments came upon the sinners,
\par }{\QB not without the signs that were given beforehand by the force of the thunders;
\par }{\QB for they justly suffered through their own wickednesses,
\par }{\QB for the hatred which they practiced toward guests was grievous indeed.
\par }{\Q \VS{14}For whereas the others didn’t receive the strangers when they came to them,
\par }{\QB the Egyptians made slaves of guests who were their benefactors.
\par }{\Q \VS{15}And not only so, but God will visit the men of Sodom another way,
\par }{\QB since they received as enemies those who were aliens;
\par }{\QB \VS{16}whereas these first welcomed with feastings,
\par }{\QB and then afflicted with dreadful toils,
\par }{\QB those who had already shared with them in the same rights.
\par }{\Q \VS{17}And moreover they were stricken with loss of sight
\par }{\QB (even as were those others at the righteous man’s doors),
\par }{\QB when, being surrounded with yawning darkness,
\par }{\QB they each looked for the passage through his own door.
\par }{\Q \VS{18}For as the notes of a lute vary the character of the rhythm,
\par }{\QB even so the elements, changing their order one with another,
\par }{\QB continuing always in its sound,
\par }{\QB as may clearly be conjectured from the sight of the things that have happened.
\par }{\Q \VS{19}For creatures of the dry land were turned into creatures of the waters,
\par }{\QB and creatures that swim moved upon the land.
\par }{\Q \VS{20}Fire kept the mastery of its own power in water,
\par }{\QB and water forgot its quenching nature.
\par }{\Q \VS{21}On the contrary, flames didn’t consume flesh of perishable creatures that walked among them,
\par }{\QB neither did they melt the ice-like grains of ambrosial food that were apt to melt.
\par }{\Q \VS{22}For in all things, O Lord, you magnified your people,
\par }{\QB and you glorified them and didn’t lightly regard them,
\par }{\QB standing by their side in every time and place.
\par }